### 2018-08-08 logistic.Rmd
initial commit

---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
use_python("/usr/local/bin/python3", required = TRUE)
py_config()
```

```{python,engine.path = '/usr/local/bin/python3'}
## Let's make sure we are using the Python version we want
import sys
print(sys.version)
## For me this is Python3, which is what I want
## We want to be working from our project directory so make sure you are in the right place
import os
print(os.getcwd())
## I'm in my DeepLearning project directory.

###### IMPORT MODULES #### ###
import matplotlib
matplotlib.use('TkAgg')
import matplotlib.pyplot as plt
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import train_test_split
from sklearn.model_selection import KFold, cross_val_score, validation_curve
from sklearn import linear_model
import numpy as np # linear algebra
import pandas as pd # data processing, CSV file I/O (e.g. pd.read_csv)
from sympy import *
from sklearn.model_selection import GridSearchCV
from sklearn.neural_network import MLPClassifier
import numpy as np
from scipy import interp
from itertools import cycle
from sklearn import svm, datasets
from sklearn.metrics import roc_curve, auc
from sklearn.model_selection import StratifiedKFold
import numpy as np
import pandas as pd
# dependencies for statistic analysis
from scipy import stats
#importing our parameter tuning dependencies
from sklearn.neighbors import KNeighborsClassifier
from sklearn.model_selection import (cross_val_score, GridSearchCV, StratifiedKFold, ShuffleSplit )
#importing our dependencies for Feature Selection
from sklearn.feature_selection import (SelectKBest, chi2, RFE, RFECV)
from sklearn.linear_model import LogisticRegression, RandomizedLogisticRegression
from sklearn.ensemble import ExtraTreesClassifier
from sklearn.cross_validation import ShuffleSplit
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import f1_score
# Importing our sklearn dependencies for the modeling
from sklearn.ensemble import RandomForestClassifier
from sklearn.neighbors import NearestNeighbors
from sklearn.preprocessing import StandardScaler
from sklearn.cross_validation import KFold
from sklearn.model_selection import train_test_split
from sklearn import metrics
from sklearn.metrics import (accuracy_score, confusion_matrix, classification_report, roc_curve, auc)
from sklearn.neural_network import MLPClassifier
from itertools import cycle
from scipy import interp
import warnings
from sklearn.model_selection import StratifiedKFold
warnings.filterwarnings('ignore')

## Read in the data
shared = pd.read_table("data/baxter.0.03.subsample.shared")
shared.head()
meta = pd.read_table("data/metadata.tsv")

## Check and visualize the data
meta.head()
shared.head()

## Remove unnecessary columns from meta
meta = meta[['sample','dx']]

## Rename the column name "Group" to match the "sample" in meta
shared = shared.rename(index=str, columns={"Group":"sample"})

## Merge the 2 datasets on sample
data=pd.merge(meta,shared,on=['sample'])

## Remove adenoma samples
data= data[data.dx.str.contains("adenoma") == False]

## Drop all except OTU columns for x
x = data.drop(["sample", "dx", "numOtus", "label"], axis=1)
## Cancer =1 Normal =0
diagnosis = { "cancer":1, "normal":0}
##Generate y which only has diagnosis as 0 and 1
y = data["dx"].replace(diagnosis)
# y = np.eye(2, dtype='uint8')[y]
## Drop if NA elements
y.dropna()
x.dropna()

## Split the data to generate training and test sets %80-20
x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, random_state=82089, shuffle=True)

######### Logistic Regression ###################

## Define L2 regularized logistic classifier
logreg = linear_model.LogisticRegression(C=0.001)
kfold = StratifiedKFold(n_splits=5)
cv_results = cross_val_score(logreg, x_train, y_train, cv=kfold)
print (cv_results.mean()*100, "%")

## Fit to training sets
logreg.fit(x_train, y_train)

## Plot ROC curve for Logistic Regression training cross validation
cv = StratifiedKFold(n_splits=5)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)

X=x_train.values
Y=y_train.values

#logistic_plot = plt.figure()
i = 0
for train, test in cv.split(X,Y):
    probas_ = logreg.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
    fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
    tprs.append(interp(mean_fpr, fpr, tpr))
    tprs[-1][0] = 0.0
    roc_auc = auc(fpr, tpr)
    aucs.append(roc_auc)
    plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (i, roc_auc))
    i += 1

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Logistic Regression ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
#logistic_plot.savefig('results/figures/Logit_Baxter.png', dpi=1000)

## Predict using the Logistic Regression classifier on the test set
y_pred = logreg.predict(x_test)
print('Accuracy of logistic regression classifier on test set: {:.2f}'.format(logreg.score(x_test, y_test)*100), "%")
```
Output image:
![output](results/figures/Logit_Baxter.png)

### 2018-08-08 sklearn_baxter.Rmd
first commit

---
title: "Logistic Regression, MLP and Random Forest on Baxter Dataset"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
use_python("/Library/Frameworks/Python.framework/Versions/3.6/bin/python3", required = TRUE)
py_config()
```

#### Let's make sure we are using the Python version we want
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import sys
print(sys.version)
```
#### Let's make sure we are in our project directory
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import os
print(os.getcwd())
```
#### Start by importing modules that will be necessary 
```{python, echo=T, results='hide', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
###### IMPORT MODULES #######
# We need to use a backend for matplotlib
# https://stackoverflow.com/questions/21784641/installation-issue-with-matplotlib-python/21789908#21789908
import matplotlib as mpl
mpl.use('TkAgg')
import matplotlib.pyplot as plt
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import train_test_split
from sklearn.model_selection import KFold, cross_val_score, validation_curve
from sklearn import linear_model
import numpy as np # linear algebra
import pandas as pd # data processing, CSV file I/O (e.g. pd.read_csv)
from sympy import *
from sklearn.model_selection import GridSearchCV
from sklearn.neural_network import MLPClassifier
from scipy import interp
from itertools import cycle
from sklearn import svm, datasets
from sklearn.metrics import roc_curve, auc
from sklearn.model_selection import StratifiedKFold
# dependencies for statistic analysis
from scipy import stats
#importing our parameter tuning dependencies
from sklearn.neighbors import KNeighborsClassifier
from sklearn.model_selection import (cross_val_score, GridSearchCV, StratifiedKFold, ShuffleSplit )
#importing our dependencies for Feature Selection
from sklearn.feature_selection import (SelectKBest, chi2, RFE, RFECV)
from sklearn.linear_model import LogisticRegression, RandomizedLogisticRegression
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import f1_score
# Importing our sklearn dependencies for the modeling
from sklearn.ensemble import RandomForestClassifier
from sklearn.neighbors import NearestNeighbors
from sklearn.preprocessing import StandardScaler
from sklearn.model_selection import train_test_split
from sklearn import metrics
from sklearn.metrics import (accuracy_score, confusion_matrix, classification_report, roc_curve, auc)
from sklearn.neural_network import MLPClassifier
from itertools import cycle
from scipy import interp
import warnings
from sklearn.model_selection import StratifiedKFold
warnings.filterwarnings('ignore')
```
#### Read and prepare data 
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
## Read in the data
shared = pd.read_table("data/baxter.0.03.subsample.shared")
shared.head()
meta = pd.read_table("data/metadata.tsv")

## Check and visualize the data
print(meta.head())
print(shared.head())

## Remove unnecessary columns from meta
meta = meta[['sample','dx']]

## Rename the column name "Group" to match the "sample" in meta
shared = shared.rename(index=str, columns={"Group":"sample"})

## Merge the 2 datasets on sample
data=pd.merge(meta,shared,on=['sample'])

## Remove adenoma samples
data= data[data.dx.str.contains("adenoma") == False]

## Drop all except OTU columns for x
x = data.drop(["sample", "dx", "numOtus", "label"], axis=1)
## Cancer =1 Normal =0
diagnosis = { "cancer":1, "normal":0}
##Generate y which only has diagnosis as 0 and 1
y = data["dx"].replace(diagnosis)
# y = np.eye(2, dtype='uint8')[y]
## Drop if NA elements
y.dropna()
x.dropna()
print(x.head())
print(y.head())
print(len(x))
print(len(y))
```
#### Split the data to generate training and testing set %80-20
Here we also want to shuffle the data and set a random state
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, random_state=82089, shuffle=True)
```
#### Now let's define a L2 regularized logistic regression model
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

## Define L2 regularized logistic classifier
logreg = linear_model.LogisticRegression(C=0.01)

```
#### Plot ROC curve for Logistic Regression training set
We split the training set to 10 to cross-validate
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

cv = StratifiedKFold(n_splits=10)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)

## Convert from pandas dataframe to numpy
X=x_train.values
Y=y_train.values

# Plot the ROC curve over 10 iterations for each split
#logistic_plot = plt.figure()
i = 0
for train, test in cv.split(X,Y):
    probas_ = logreg.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
    fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
    tprs.append(interp(mean_fpr, fpr, tpr))
    tprs[-1][0] = 0.0
    roc_auc = auc(fpr, tpr)
    aucs.append(roc_auc)
    plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (i, roc_auc))
    i += 1

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Logistic Regression ROC\n')
plt.legend(loc="lower right", fontsize=6)
plt.show()
#logistic_plot.savefig('results/figures/Logit_Baxter.png', dpi=1000)
```

####Predict using the Logistic Regression model on the test set

```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
y_pred = logreg.predict(x_test)
print('Accuracy of logistic regression classifier on test set: {:.2f}'.format(logreg.score(x_test, y_test)*100), "%")
```

#### Now let's define a Multi-layer Perceptron Neural Network Model

```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
clf = MLPClassifier(activation='logistic', alpha=0.001, batch_size='auto',
       beta_1=0.9, beta_2=0.999, early_stopping=True, epsilon=1e-08,
       hidden_layer_sizes=(100,), learning_rate='adaptive',
       learning_rate_init=0.001, max_iter=200, momentum=0.9,
       nesterovs_momentum=True, power_t=0.5, random_state=1, shuffle=True,
       solver='sgd', tol=0.0001, validation_fraction=0.1, verbose=False,
       warm_start=False)

```
#### Plot ROC curve for MLP on training set
We split the training set to 10 to cross-validate
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

cv = StratifiedKFold(n_splits=10)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
mlp_plot = plt.figure()

i = 0
for train, test in cv.split(X,Y):
    probas_ = clf.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
    fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
    tprs.append(interp(mean_fpr, fpr, tpr))
    tprs[-1][0] = 0.0
    roc_auc = auc(fpr, tpr)
    aucs.append(roc_auc)
    plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (i, roc_auc))
    i += 1

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Neural Network ROC\n')
plt.legend(loc="lower right", fontsize=7)
plt.show()
#mlp_plot.savefig('results/figures/MLP_Baxter.png', dpi=1000)
```

####Predict using the MLP model on the test set

```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
y_pred = clf.predict(x_test)
## Print accuracy
print("Performance Accuracy on the Testing data:", round(clf.score(x_test, y_test) *100))
print("Number of correct classifiers:", round(accuracy_score(y_test, y_pred, normalize=False)))
```
#### Now let's define a Random Forest model

```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

rfc = RandomForestClassifier(bootstrap=True, class_weight=None, criterion='gini',
            max_depth=None, max_features='auto', max_leaf_nodes=None,
            min_impurity_split=None, min_samples_leaf=1,
            min_samples_split=2, min_weight_fraction_leaf=0.0,
            n_estimators=150, n_jobs=1, oob_score=True, random_state=None,
            verbose=0, warm_start=False)
```
#### Plot ROC curve for Random Forest on training set
We split the training set to 10 to cross-validate
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

RF_plot = plt.figure()
cv = StratifiedKFold(n_splits=10)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)

i = 0
for train, test in cv.split(X,Y):
    probas_ = rfc.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
    fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
    tprs.append(interp(mean_fpr, fpr, tpr))
    tprs[-1][0] = 0.0
    roc_auc = auc(fpr, tpr)
    aucs.append(roc_auc)
    plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (i, roc_auc))
    i += 1

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Random Forest ROC\n')
plt.legend(loc="lower right", fontsize=7)
plt.show()
#RF_plot.savefig('results/figures/Random_Forest_Baxter.png', dpi=1000)
```

####Predict using the Random Forest model on the test set

```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
y_pred = rfc.predict(x_test)
print("Performance Accuracy on the Testing data:", round(rfc.score(x_test, y_test) *100))
print("Number of correct classifiers:", round(accuracy_score(y_test, y_pred, normalize=False)))
```

### 2018-08-08 sklearn_pyTorch_baxter.Rmd
added pytorch code

---
title: "Logistic Regression, MLP and Random Forest with Sklearn and Feed Forward Neural Net with PyTorch on Baxter Dataset"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
library(reticulate)
use_python("/Library/Frameworks/Python.framework/Versions/3.6/bin/python3", required = TRUE)
py_config()

```

#### Let's make sure we are using the Python version we want
```{python,tidy = TRUE, comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import sys
print(sys.version)
```
#### Let's make sure we are in our project directory
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import os
print(os.getcwd())
```
#### Start by importing modules that will be necessary 
```{python,tidy = TRUE, echo=T, results='hide', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
###### IMPORT MODULES #######
# We need to use a backend for matplotlib
# https://stackoverflow.com/questions/21784641/installation-issue-with-matplotlib-python/21789908#21789908
import matplotlib as mpl
mpl.use('TkAgg')
import matplotlib.pyplot as plt
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import train_test_split
from sklearn.model_selection import KFold, cross_val_score, validation_curve
from sklearn import linear_model
import numpy as np # linear algebra
import pandas as pd # data processing, CSV file I/O (e.g. pd.read_csv)
from sympy import *
from sklearn.model_selection import GridSearchCV
from sklearn.neural_network import MLPClassifier
from scipy import interp
from itertools import cycle
from sklearn import svm, datasets
from sklearn.metrics import roc_curve, auc
from sklearn.model_selection import StratifiedKFold
# dependencies for statistic analysis
from scipy import stats
#importing our parameter tuning dependencies
from sklearn.neighbors import KNeighborsClassifier
from sklearn.model_selection import (cross_val_score, GridSearchCV, StratifiedKFold, ShuffleSplit )
#importing our dependencies for Feature Selection
from sklearn.feature_selection import (SelectKBest, chi2, RFE, RFECV)
from sklearn.linear_model import LogisticRegression, RandomizedLogisticRegression
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import f1_score
# Importing our sklearn dependencies for the modeling
from sklearn.ensemble import RandomForestClassifier
from sklearn.neighbors import NearestNeighbors
from sklearn.preprocessing import StandardScaler
from sklearn.model_selection import train_test_split
from sklearn import metrics
from sklearn.metrics import (accuracy_score, confusion_matrix, classification_report, roc_curve, auc)
from sklearn.neural_network import MLPClassifier
from itertools import cycle
from scipy import interp
import warnings
from sklearn.model_selection import StratifiedKFold
import torch
from torch.autograd import Variable
import torch.utils.data as data_utils
import torch.nn.init as init
import torch.nn as nn
import torch.nn.functional as F
from torch.utils.data import Dataset, DataLoader
from sklearn.model_selection import train_test_split
from sklearn.preprocessing import StandardScaler
warnings.filterwarnings('ignore')
```
#### Read and prepare data 
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
## Read in the data
shared = pd.read_table("data/baxter.0.03.subsample.shared")
shared.head()
meta = pd.read_table("data/metadata.tsv")

## Check and visualize the data
print(meta.head())
print(shared.head())

## Remove unnecessary columns from meta
meta = meta[['sample','dx']]

## Rename the column name "Group" to match the "sample" in meta
shared = shared.rename(index=str, columns={"Group":"sample"})

## Merge the 2 datasets on sample
data=pd.merge(meta,shared,on=['sample'])

## Remove adenoma samples
data= data[data.dx.str.contains("adenoma") == False]

## Drop all except OTU columns for x
x = data.drop(["sample", "dx", "numOtus", "label"], axis=1)
## Cancer =1 Normal =0
diagnosis = { "cancer":1, "normal":0}
##Generate y which only has diagnosis as 0 and 1
y = data["dx"].replace(diagnosis)
# y = np.eye(2, dtype='uint8')[y]
## Drop if NA elements
y.dropna()
x.dropna()
print(x.head())
print(y.head())
print(len(x))
print(len(y))
```
#### Split the data to generate training and testing set %80-20
Here we also want to shuffle the data and set a random state
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, random_state=82089, shuffle=True)
```
#### Now let's define a L2 regularized logistic regression model
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

## Define L2 regularized logistic classifier
logreg = linear_model.LogisticRegression(C=0.01)

```
#### Plot ROC curve for Logistic Regression training set
We split the training set to 10 to cross-validate
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

cv = StratifiedKFold(n_splits=10)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)

## Convert from pandas dataframe to numpy
X=x_train.values
Y=y_train.values

# Plot the ROC curve over 10 iterations for each split
#logistic_plot = plt.figure()
i = 0
for train, test in cv.split(X,Y):
    probas_ = logreg.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
    fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
    tprs.append(interp(mean_fpr, fpr, tpr))
    tprs[-1][0] = 0.0
    roc_auc = auc(fpr, tpr)
    aucs.append(roc_auc)
    plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (i, roc_auc))
    i += 1

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Logistic Regression ROC\n')
plt.legend(loc="lower right", fontsize=6)
plt.show()
#logistic_plot.savefig('results/figures/Logit_Baxter.png', dpi=1000)
```

#### Predict using the Logistic Regression model on the test set

```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
y_pred = logreg.predict(x_test)
print('Accuracy of logistic regression classifier on test set: {:.2f}'.format(logreg.score(x_test, y_test)*100), "%")
```

#### Now let's define a Multi-layer Perceptron Neural Network Model

```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
clf = MLPClassifier(activation='logistic', alpha=0.001, batch_size='auto',
       beta_1=0.9, beta_2=0.999, early_stopping=True, epsilon=1e-08,
       hidden_layer_sizes=(100,), learning_rate='adaptive',
       learning_rate_init=0.001, max_iter=200, momentum=0.9,
       nesterovs_momentum=True, power_t=0.5, random_state=1, shuffle=True,
       solver='sgd', tol=0.0001, validation_fraction=0.1, verbose=False,
       warm_start=False)

```
#### Plot ROC curve for MLP on training set
We split the training set to 10 to cross-validate
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

cv = StratifiedKFold(n_splits=10)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
mlp_plot = plt.figure()

i = 0
for train, test in cv.split(X,Y):
    probas_ = clf.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
    fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
    tprs.append(interp(mean_fpr, fpr, tpr))
    tprs[-1][0] = 0.0
    roc_auc = auc(fpr, tpr)
    aucs.append(roc_auc)
    plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (i, roc_auc))
    i += 1

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Neural Network ROC\n')
plt.legend(loc="lower right", fontsize=7)
plt.show()
#mlp_plot.savefig('results/figures/MLP_Baxter.png', dpi=1000)
```

#### Predict using the MLP model on the test set

```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
y_pred = clf.predict(x_test)
## Print accuracy
print("Performance Accuracy on the Testing data:", round(clf.score(x_test, y_test) *100))
print("Number of correct classifiers:", round(accuracy_score(y_test, y_pred, normalize=False)))
```
#### Now let's define a Random Forest model

```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

rfc = RandomForestClassifier(bootstrap=True, class_weight=None, criterion='gini',
            max_depth=None, max_features='auto', max_leaf_nodes=None,
            min_impurity_split=None, min_samples_leaf=1,
            min_samples_split=2, min_weight_fraction_leaf=0.0,
            n_estimators=150, n_jobs=1, oob_score=True, random_state=None,
            verbose=0, warm_start=False)
```
#### Plot ROC curve for Random Forest on training set
We split the training set to 10 to cross-validate
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

RF_plot = plt.figure()
cv = StratifiedKFold(n_splits=10)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)

i = 0
for train, test in cv.split(X,Y):
    probas_ = rfc.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
    fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
    tprs.append(interp(mean_fpr, fpr, tpr))
    tprs[-1][0] = 0.0
    roc_auc = auc(fpr, tpr)
    aucs.append(roc_auc)
    plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (i, roc_auc))
    i += 1

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Random Forest ROC\n')
plt.legend(loc="lower right", fontsize=7)
plt.show()
#RF_plot.savefig('results/figures/Random_Forest_Baxter.png', dpi=1000)
```

#### Predict using the Random Forest model on the test set

```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
y_pred = rfc.predict(x_test)
print("Performance Accuracy on the Testing data:", round(rfc.score(x_test, y_test) *100))
print("Number of correct classifiers:", round(accuracy_score(y_test, y_pred, normalize=False)))
```

#### Now let's define a 1 hidden layer(n=100) Feed Forward Neural Net
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

class Net(nn.Module):
    def __init__(self):
        super(Net, self).__init__()
        self.fc1 = nn.Linear(6920, 100)
        self.fc2 = nn.Linear(100, 2)
    def forward(self, x):
        x = self.fc1(x)
        x = F.dropout(x, p=0.1)
        x = F.relu(x)
        x = self.fc2(x)
        x = F.sigmoid(x)
        return x

net = Net()

## Batch size allows for random sampling of the dataset during training
batch_size = 50
num_epochs = 50
learning_rate = 0.0001
batch_no = len(x_train) // batch_size

criterion = nn.CrossEntropyLoss()
optimizer = torch.optim.Adam(net.parameters(), lr=learning_rate)

from sklearn.utils import shuffle
from torch.autograd import Variable
from scipy import interp
from sklearn.metrics import (accuracy_score, confusion_matrix, classification_report, roc_curve, auc)

## ROC plot for training
pyTorch_plot = plt.figure()
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
for epoch in range(num_epochs):
    x_train, y_train = shuffle(x_train, y_train)
    # Mini batch learning
    for i in range(batch_no):
        start = i * batch_size
        end = start + batch_size
        x_var = Variable(torch.FloatTensor(x_train.values[start:end]))
        y_var = Variable(torch.LongTensor(y_train.values[start:end]))
        # Forward + Backward + Optimize
        ypred_var = net(x_var)
        loss =criterion(ypred_var, y_var)
        correct_num = 0
        ## The outputs of the model (ypred_var) are energies for the 10 classes. Higher the energy for a class, the more the network thinks that the image is of the particular class. So, let’s get the index of the highest energy:
        values, labels = torch.max(ypred_var, 1)
        correct_num = np.sum(labels.data.numpy() == y_var.numpy())
        fpr, tpr, thresholds = roc_curve(y_var.numpy(), labels.data.numpy())
        tprs.append(interp(mean_fpr, fpr, tpr))
        tprs[-1][0] = 0.0
        roc_auc = auc(fpr, tpr)
        aucs.append(roc_auc)
        #plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (epoch, roc_auc))
        optimizer.zero_grad()
        loss.backward()
        optimizer.step()
        print('Epoch [%d], Loss:%.4f, Accuracy:%.4f' % (epoch, loss.data[0], correct_num/len(labels)))

plt.plot([0, 1], [0, 1], linestyle='--', color='r', label='Luck', alpha=.8)
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='grey', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('PyTorch Neural Network ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
#pyTorch_plot.savefig('results/figures/pyTorch_Baxter.png', dpi=1000)

# Evaluate the model on test set
net.eval()
pred = net(torch.from_numpy(x_test.values).float())
pred = torch.max(pred,1)[1]
len(pred)
pred = pred.data.numpy()
print(accuracy_score(y_test, pred))
print(confusion_matrix(y_test, pred))
```

### 2018-08-08 sklearn_pyTorch_baxter.Rmd
export as pdf

@@
 title: "Logistic Regression, MLP and Random Forest with Sklearn and Feed Forward Neural Net with PyTorch on Baxter Dataset"
-output: html_document
+geometry: "left=1cm,right=1cm,top=1cm,bottom=1cm"
+fontsize: 10pt
+output: pdf_document
 ---
@@
 library(knitr)
-opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
+opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
 library(reticulate)
@@
 #### Predict using the Random Forest model on the test set
-
 ```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
@@
 ```
-
 #### Now let's define a 1 hidden layer(n=100) Feed Forward Neural Net
@@
         ## The outputs of the model (ypred_var) are energies for the 10 classes. 
-Higher the energy for a class, the more the network thinks that the image is of the particular class.
-So, let’s get the index of the highest energy:
+        #Higher the energy for a class, the more the network thinks that the image is of the         #particular class. So, let’s get the index of the highest energy:
         values, labels = torch.max(ypred_var, 1)

### 2018-08-13 baxter.Rmd
first commit

---
title: "Logistic Regression, Multi-layer Perceptron Neural Network, Random Forest, XGBoost and PyTorchon Baxter Dataset"
geometry: "left=1cm,right=1cm,top=1cm,bottom=1cm"
fontsize: 10pt
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
library(reticulate)
use_python("/Library/Frameworks/Python.framework/Versions/3.6/bin/python3", required = TRUE)
py_config()

```

#### Let's make sure we are using the Python version we want
```{python,tidy = TRUE, comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import sys
print(sys.version)
```
#### Let's make sure we are in our project directory
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import os
print(os.getcwd())
```
#### Start by importing modules that will be necessary 
```{python,tidy = TRUE, include=FALSE, echo=F, results='hide', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
###### IMPORT MODULES #######
# We need to use a backend for matplotlib
# https://stackoverflow.com/questions/21784641/installation-issue-with-matplotlib-python/21789908#21789908
import matplotlib as mpl
mpl.use('TkAgg')
import matplotlib.pyplot as plt
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import train_test_split
from sklearn.model_selection import KFold, cross_val_score, validation_curve
from sklearn import linear_model
import numpy as np # linear algebra
import pandas as pd # data processing, CSV file I/O (e.g. pd.read_csv)
from sympy import *
from sklearn.model_selection import GridSearchCV
from sklearn.neural_network import MLPClassifier
from scipy import interp
from itertools import cycle
from sklearn import svm, datasets
from sklearn.metrics import roc_curve, auc
from sklearn.model_selection import StratifiedKFold
# dependencies for statistic analysis
from scipy import stats
#importing our parameter tuning dependencies
from sklearn.neighbors import KNeighborsClassifier
from sklearn.model_selection import (cross_val_score, GridSearchCV, StratifiedKFold, ShuffleSplit )
#importing our dependencies for Feature Selection
from sklearn.feature_selection import (SelectKBest, chi2, RFE, RFECV)
from sklearn.linear_model import LogisticRegression, RandomizedLogisticRegression
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import f1_score
# Importing our sklearn dependencies for the modeling
from sklearn.ensemble import RandomForestClassifier
from sklearn.neighbors import NearestNeighbors
from sklearn.preprocessing import StandardScaler
from sklearn.model_selection import train_test_split
from sklearn import metrics
from sklearn.metrics import (accuracy_score, confusion_matrix, classification_report, roc_curve, auc)
from sklearn.neural_network import MLPClassifier
from itertools import cycle
from scipy import interp
import warnings
from sklearn.model_selection import StratifiedKFold
import torch
from torch.autograd import Variable
import torch.utils.data as data_utils
import torch.nn.init as init
import torch.nn as nn
import torch.nn.functional as F
from torch.utils.data import Dataset, DataLoader
from sklearn.model_selection import train_test_split
from sklearn.preprocessing import StandardScaler
import xgboost
from xgboost import XGBClassifier
from sklearn.model_selection import train_test_split
from sklearn.grid_search import GridSearchCV, RandomizedSearchCV
from sklearn.datasets import make_classification
from sklearn.cross_validation import StratifiedKFold
import xgboost as xgb
import xgboost
from xgboost import XGBClassifier
from sklearn.model_selection import train_test_split
from sklearn.grid_search import GridSearchCV, RandomizedSearchCV
from sklearn.datasets import make_classification
from sklearn.cross_validation import StratifiedKFold
from sklearn.externals import joblib
from sklearn.metrics import precision_score
from sklearn import preprocessing
from sklearn import cross_validation
from sklearn.model_selection import KFold
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import train_test_split
from sklearn.model_selection import KFold, cross_val_score, validation_curve
from sklearn import linear_model
import numpy as np # linear algebra
import pandas as pd # data processing, CSV file I/O (e.g. pd.read_csv)
from sympy import *
import matplotlib.pyplot as plt
from sklearn.model_selection import GridSearchCV
from sklearn.neural_network import MLPClassifier
import numpy as np
from scipy import interp
from itertools import cycle
from sklearn import svm, datasets
from sklearn.metrics import roc_curve, auc
from sklearn.model_selection import StratifiedKFold
from pylab import rcParams
# dependencies for statistic analysis
from scipy import stats
#importing our parameter tuning dependencies
from sklearn.neighbors import KNeighborsClassifier
from sklearn.model_selection import (cross_val_score, GridSearchCV, StratifiedKFold, ShuffleSplit )
#importing our dependencies for Feature Selection
from sklearn.feature_selection import (SelectKBest, chi2, RFE, RFECV)
from sklearn.linear_model import LogisticRegression, RandomizedLogisticRegression
from sklearn.ensemble import ExtraTreesClassifier
from sklearn.cross_validation import ShuffleSplit
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import f1_score
from collections import defaultdict
# Importing our sklearn dependencies for the modeling
from sklearn.ensemble import RandomForestClassifier
from sklearn.neighbors import NearestNeighbors
from sklearn.preprocessing import StandardScaler
from sklearn.cross_validation import KFold
from sklearn.model_selection import train_test_split
from sklearn import metrics
from sklearn.metrics import (accuracy_score, confusion_matrix, classification_report, roc_curve, auc)
from sklearn.neural_network import MLPClassifier
from itertools import cycle
from scipy import interp
import warnings
warnings.filterwarnings('ignore')
```
#### Read and prepare data 
We will use OTUs for each sample as our features and diagnosis column in the metadata as our classifier.
Cancer is 1 and Normal is 0.

```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
## Read in the data
shared = pd.read_table("data/baxter.0.03.subsample.shared")
shared.head()
meta = pd.read_table("data/metadata.tsv")

## Check and visualize the data
print(meta.head())
print(shared.head())

## Remove unnecessary columns from meta
meta = meta[['sample','dx']]

## Rename the column name "Group" to match the "sample" in meta
shared = shared.rename(index=str, columns={"Group":"sample"})

## Merge the 2 datasets on sample
data=pd.merge(meta,shared,on=['sample'])

## Remove adenoma samples
data= data[data.dx.str.contains("adenoma") == False]

## Drop all except OTU columns for x
x = data.drop(["sample", "dx", "numOtus", "label"], axis=1)
## Cancer =1 Normal =0
diagnosis = { "cancer":1, "normal":0}
##Generate y which only has diagnosis as 0 and 1
y = data["dx"].replace(diagnosis)
# y = np.eye(2, dtype='uint8')[y]
## Drop if NA elements
y.dropna()
x.dropna()
print(x.head())
print(y.head())
print(len(x))
print(len(y))
```
#### Now let's define a L2 regularized logistic regression model
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

## Define L2 regularized logistic classifier
logreg = linear_model.LogisticRegression(C=0.01)

```
#### Plot ROC curve for Logistic Regression model on training set and predict test set
We split training and test set 80-20%. We further split the training set to 5 to cross-validate in a loop.
We predict test set in the same loop. 
We generate mean AUC for cross-validation with training set upon 50 epochs and mean AUC for test set upon 50 epochs. 
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

cv = StratifiedKFold(n_splits=5)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
tprs_test = []
aucs_test = []
mean_fpr_test = np.linspace(0, 1, 100)

# Plot the ROC curve over 10 iterations for each split
Logit_plot = plt.figure()
epochs = 50
for epoch in range(epochs):
    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, shuffle=True)
    X=x_train.values
    Y=y_train.values
    X_test= x_test.values
    Y_test= y_test.values
    i = 0
    for train, test in cv.split(X,Y):
        probas_ = logreg.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
        fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
        tprs.append(interp(mean_fpr, fpr, tpr))
        tprs[-1][0] = 0.0
        roc_auc = auc(fpr, tpr)
        aucs.append(roc_auc)
        #plt.plot(fpr, tpr, lw=1, alpha=0.3, label='Epoch %d, ROC fold %d (AUC = %0.2f)' % (epoch, i, roc_auc))
        #i += 1
    probas_ = logreg.predict_proba(X_test)
    # Compute ROC curve and area the curve
    fpr_test, tpr_test, thresholds_test = roc_curve(Y_test, probas_[:, 1])
    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
    tprs_test[-1][0] = 0.0
    roc_auc_test = auc(fpr_test, tpr_test)
    aucs_test.append(roc_auc_test)

plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)
mean_tpr_test = np.mean(tprs_test, axis=0)
mean_tpr_test[-1] = 1.0
mean_auc_test = auc(mean_fpr_test, mean_tpr_test)
std_auc_test = np.std(aucs_test)
plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Mean test ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
std_tpr_test = np.std(tprs_test, axis=0)
tprs_upper_test = np.minimum(mean_tpr_test + std_tpr_test, 1)
tprs_lower_test = np.maximum(mean_tpr_test - std_tpr_test, 0)
plt.fill_between(mean_fpr_test, tprs_lower_test, tprs_upper_test, color='tomato', alpha=.2, label=r'$\pm$ 1 std. dev.')
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='dodgerblue', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('L2 Regularized Logistic Regression ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
#Logit_plot.savefig('results/figures/Logit_Baxter.png', dpi=1000)
```

#### Now let's define a Multi-layer Perceptron Neural Network Model
Here the activation function is a logistic regression. There is 1 hidden layer with 100 nodes. 
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
clf = MLPClassifier(activation='logistic', alpha=0.001, batch_size='auto',
       beta_1=0.9, beta_2=0.999, early_stopping=True, epsilon=1e-08,
       hidden_layer_sizes=(100,), learning_rate='adaptive',
       learning_rate_init=0.001, max_iter=200, momentum=0.9,
       nesterovs_momentum=True, power_t=0.5, random_state=1, shuffle=True,
       solver='sgd', tol=0.0001, validation_fraction=0.1, verbose=False,
       warm_start=False)

```
#### Plot ROC curve for MLP on training set
We split training and test set 80-20%. We further split the training set to 5 to cross-validate in a loop.
We predict test set in the same loop. 
We generate mean AUC for cross-validation with training set upon 50 epochs and mean AUC for test set upon 50 epochs. 
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

cv = StratifiedKFold(n_splits=5)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
tprs_test = []
aucs_test = []
mean_fpr_test = np.linspace(0, 1, 100)

MLP_plot = plt.figure()
epochs = 50
for epoch in range(epochs):
    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, shuffle=True)
    X=x_train.values
    Y=y_train.values
    X_test= x_test.values
    Y_test= y_test.values
    i = 0
    for train, test in cv.split(X,Y):
        probas_ = clf.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
        fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
        tprs.append(interp(mean_fpr, fpr, tpr))
        tprs[-1][0] = 0.0
        roc_auc = auc(fpr, tpr)
        aucs.append(roc_auc)
        #plt.plot(fpr, tpr, lw=1, alpha=0.3, label='Epoch %d, ROC fold %d (AUC = %0.2f)' % (epoch, i, roc_auc))
        #i += 1
    probas_ = clf.predict_proba(X_test)
    # Compute ROC curve and area the curve
    fpr_test, tpr_test, thresholds_test = roc_curve(Y_test, probas_[:, 1])
    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
    tprs_test[-1][0] = 0.0
    roc_auc_test = auc(fpr_test, tpr_test)
    aucs_test.append(roc_auc_test)

plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)
mean_tpr_test = np.mean(tprs_test, axis=0)
mean_tpr_test[-1] = 1.0
mean_auc_test = auc(mean_fpr_test, mean_tpr_test)
std_auc_test = np.std(aucs_test)
plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Mean test ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
std_tpr_test = np.std(tprs_test, axis=0)
tprs_upper_test = np.minimum(mean_tpr_test + std_tpr_test, 1)
tprs_lower_test = np.maximum(mean_tpr_test - std_tpr_test, 0)
plt.fill_between(mean_fpr_test, tprs_lower_test, tprs_upper_test, color='tomato', alpha=.2, label=r'$\pm$ 1 std. dev.')
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='dodgerblue', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Multi-layer Perceptron Neural Network Model ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
#MLP_plot.savefig('results/figures/MLP_Baxter.png', dpi=1000)
```

#### Now let's define a Random Forest model
Previous cross validation step allowed us to choose n_estimators(decision trees). Here we have 150 decision trees.
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

rfc = RandomForestClassifier(bootstrap=True, class_weight=None, criterion='gini',
            max_depth=None, max_features='auto', max_leaf_nodes=None,
            min_impurity_split=None, min_samples_leaf=1,
            min_samples_split=2, min_weight_fraction_leaf=0.0,
            n_estimators=150, n_jobs=1, oob_score=True, random_state=None,
            verbose=0, warm_start=False)
```
#### Plot ROC curve for Random Forest on training set and predict test set
We split training and test set 80-20%. We further split the training set to 5 to cross-validate in a loop.
We predict test set in the same loop. 
We generate mean AUC for cross-validation with training set upon 50 epochs and mean AUC for test set upon 50 epochs. 
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
tprs_test = []
aucs_test = []
mean_fpr_test = np.linspace(0, 1, 100)

RF_plot = plt.figure()
epochs = 50
for epoch in range(epochs):
    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, shuffle=True)
    X=x_train.values
    Y=y_train.values
    X_test= x_test.values
    Y_test= y_test.values
    i = 0
    for train, test in cv.split(X,Y):
        probas_ = rfc.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
        fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
        tprs.append(interp(mean_fpr, fpr, tpr))
        tprs[-1][0] = 0.0
        roc_auc = auc(fpr, tpr)
        aucs.append(roc_auc)
        #plt.plot(fpr, tpr, lw=1, alpha=0.3, label='Epoch %d, ROC fold %d (AUC = %0.2f)' % (epoch, i, roc_auc))
        #i += 1
    probas_ = rfc.predict_proba(X_test)
    # Compute ROC curve and area the curve
    fpr_test, tpr_test, thresholds_test = roc_curve(Y_test, probas_[:, 1])
    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
    tprs_test[-1][0] = 0.0
    roc_auc_test = auc(fpr_test, tpr_test)
    aucs_test.append(roc_auc_test)

plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)
mean_tpr_test = np.mean(tprs_test, axis=0)
mean_tpr_test[-1] = 1.0
mean_auc_test = auc(mean_fpr_test, mean_tpr_test)
std_auc_test = np.std(aucs_test)
plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Mean test ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
std_tpr_test = np.std(tprs_test, axis=0)
tprs_upper_test = np.minimum(mean_tpr_test + std_tpr_test, 1)
tprs_lower_test = np.maximum(mean_tpr_test - std_tpr_test, 0)
plt.fill_between(mean_fpr_test, tprs_lower_test, tprs_upper_test, color='tomato', alpha=.2, label=r'$\pm$ 1 std. dev.')
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='dodgerblue', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('Random Forest ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
#RF_plot.savefig('results/figures/RF_Baxter.png', dpi=1000)
```

#### Now let's define a XGBoost model
Previous cross validation step allowed us to choose n_estimators(decision trees), max_depth and learning rate.
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

clf = xgb.XGBClassifier(base_score=0.5, booster='gbtree', colsample_bylevel=1,
       colsample_bytree=1, gamma=0, learning_rate=0.01, max_delta_step=0,
       max_depth=1, min_child_weight=1, missing=None, n_estimators=250,
       n_jobs=1, nthread=None, objective='binary:logistic', random_state=0,
       reg_alpha=0, reg_lambda=0.1, scale_pos_weight=1, seed=None,
       silent=True, subsample=1)
```
#### Plot ROC curve for XGBoost on training set and predict test set
We split training and test set 80-20%. We further split the training set to 5 to cross-validate in a loop.
We predict test set in the same loop. 
We generate mean AUC for cross-validation with training set upon 50 epochs and mean AUC for test set upon 50 epochs. 
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

cv = StratifiedKFold(n_splits=5)
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
tprs_test = []
aucs_test = []
mean_fpr_test = np.linspace(0, 1, 100)

XGboost_plot = plt.figure()
epochs = 50
for epoch in range(epochs):
    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, shuffle=True)
    X=x_train.values
    Y=y_train.values
    X_test= x_test.values
    Y_test= y_test.values
    i = 0
    for train, test in cv.split(X,Y):
        probas_ = clf.fit(X[train], Y[train]).predict_proba(X[test])
    # Compute ROC curve and area the curve
        fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
        tprs.append(interp(mean_fpr, fpr, tpr))
        tprs[-1][0] = 0.0
        roc_auc = auc(fpr, tpr)
        aucs.append(roc_auc)
        #plt.plot(fpr, tpr, lw=1, alpha=0.3, label='Epoch %d, ROC fold %d (AUC = %0.2f)' % (epoch, i, roc_auc))
        #i += 1
    probas_ = clf.predict_proba(X_test)
    # Compute ROC curve and area the curve
    fpr_test, tpr_test, thresholds_test = roc_curve(Y_test, probas_[:, 1])
    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
    tprs_test[-1][0] = 0.0
    roc_auc_test = auc(fpr_test, tpr_test)
    aucs_test.append(roc_auc_test)

plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)
mean_tpr_test = np.mean(tprs_test, axis=0)
mean_tpr_test[-1] = 1.0
mean_auc_test = auc(mean_fpr_test, mean_tpr_test)
std_auc_test = np.std(aucs_test)
plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Mean test ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
std_tpr_test = np.std(tprs_test, axis=0)
tprs_upper_test = np.minimum(mean_tpr_test + std_tpr_test, 1)
tprs_lower_test = np.maximum(mean_tpr_test - std_tpr_test, 0)
plt.fill_between(mean_fpr_test, tprs_lower_test, tprs_upper_test, color='tomato', alpha=.2, label=r'$\pm$ 1 std. dev.')
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='dodgerblue', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('XGBoost ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
#XGboost_plot.savefig('results/figures/XGBoost_Baxter.png', dpi=1000)
```

#### Now let's define a 1 hidden layer(n=100) Feed Forward Neural Net
PyTorch allows for forward computation. One epoch will mean getting the output values(the way our model defines them),  one backward pass where weights in the model are updated and then optimize the hyperparameters for the next epoch.
```{python,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

class Net(nn.Module):
    def __init__(self):
        super(Net, self).__init__()
        self.fc1 = nn.Linear(6920, 100)
        self.fc2 = nn.Linear(100, 2)
    def forward(self, x):
        x = self.fc1(x)
        x = F.dropout(x, p=0.1)
        x = F.relu(x)
        x = self.fc2(x)
        x = F.sigmoid(x)
        return x

net = Net()

## Batch size allows for random sampling of the dataset during training
batch_size = 50
num_epochs = 40
learning_rate = 0.0001
batch_no = 233 // batch_size

# Define Loss function and optimizer function
criterion = nn.CrossEntropyLoss()
optimizer = torch.optim.Adam(net.parameters(), lr=learning_rate)

## Plot ROC for training (with minibatches) and evaluate test sets
pyTorch_plot = plt.figure()
tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)
tprs_test = []
aucs_test = []
mean_fpr_test = np.linspace(0, 1, 100)

for epoch in range(num_epochs):
    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, shuffle=True)
    # Mini batch learning
    for i in range(batch_no):
        start = i * batch_size
        end = start + batch_size
        x_var = Variable(torch.FloatTensor(x_train.values[start:end]))
        y_var = Variable(torch.LongTensor(y_train.values[start:end]))
        # Forward + Backward + Optimize
        ypred_var = net(x_var)
        loss =criterion(ypred_var, y_var)
        correct_num = 0
        ## The outputs of the model (ypred_var) are energies for the 10 classes. Higher the energy for a class, the more the network thinks that the image is of the particular class. So, let’s get the index of the highest energy:
        values, labels = torch.max(ypred_var, 1)
        correct_num = np.sum(labels.data.numpy() == y_var.numpy())
        fpr, tpr, thresholds = roc_curve(y_var.numpy(), labels.data.numpy())
        tprs.append(interp(mean_fpr, fpr, tpr))
        tprs[-1][0] = 0.0
        roc_auc = auc(fpr, tpr)
        aucs.append(roc_auc)
        optimizer.zero_grad()
        loss.backward()
        optimizer.step()
        print('Epoch [%d], Loss:%.4f, Accuracy:%.4f' % (epoch, loss.data[0], correct_num/len(labels)))
        #plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (epoch, roc_auc))
   #Predict test set and plot mean AUC
    x_var_test = Variable(torch.FloatTensor(x_test.values))
    y_var_test = Variable(torch.LongTensor(y_test.values))
    net.eval()
    ypred_var_test = net(x_var_test)
    correct_num_test = 0
    values_test, labels_test = torch.max(ypred_var_test, 1)
    correct_num_test = np.sum(labels_test.data.numpy() == y_var_test.numpy())
    fpr_test, tpr_test, thresholds_test = roc_curve(y_var_test.numpy(), labels_test.data.numpy())
    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
    tprs_test[-1][0] = 0.0
    roc_auc_test = auc(fpr_test, tpr_test)
    aucs_test.append(roc_auc_test)

plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)
mean_tpr_test = np.mean(tprs_test, axis=0)
mean_tpr_test[-1] = 1.0
mean_auc_test = auc(mean_fpr_test, mean_tpr_test)
std_auc_test = np.std(aucs_test)
plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Mean test ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
std_tpr_test = np.std(tprs_test, axis=0)
tprs_upper_test = np.minimum(mean_tpr_test + std_tpr_test, 1)
tprs_lower_test = np.maximum(mean_tpr_test - std_tpr_test, 0)
plt.fill_between(mean_fpr_test, tprs_lower_test, tprs_upper_test, color='tomato', alpha=.2, label=r'$\pm$ 1 std. dev.')
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='dodgerblue', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('PyTorch 1 hidden layer with 100 nodes Neural Network ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
```

### 2018-08-23 LogisticRegression_BaxterData.Rmd
initial commit

---
title: "Logistic Regression on Baxter colorectal cancer dataset"
output: pdf_document

---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
library(reticulate)
use_python("/Library/Frameworks/Python.framework/Versions/3.6/bin/python3", required = TRUE)
py_config()
```

#### Let's make sure we are using the Python version we want
```{python,tidy = TRUE, comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import sys
print(sys.version)
```

#### Let's make sure we are in our project directory
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import os
print(os.getcwd())
```

#### Start by importing modules that will be necessary 
```{python,tidy = TRUE, cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
############## IMPORT MODULES ######################
from sklearn.linear_model import LogisticRegression
from sklearn import linear_model
from sklearn.model_selection import train_test_split
import numpy as np # linear algebra
import pandas as pd # data processing, CSV file I/O (e.g. pd.read_csv)
from sympy import *
import matplotlib.pyplot as plt
from scipy import interp
from sklearn.metrics import roc_curve, auc
from sklearn.model_selection import StratifiedKFold
```

#### Pre-process the data

We have 6920 OTUs that are the features we put in.
We have 292 samples with labels:
- 120 Cancer colon
- 172 Normal colon

I would like to do a classification model to predict discrete values (diagnosis 1 or 0)

```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
## Read in the data

## shared has our input features(OTUs)
shared = pd.read_table("data/baxter.0.03.subsample.shared")
## meta has the labels(diagnosis)
meta = pd.read_table("data/metadata.tsv")

## Check and visualize the data
meta.head()
shared.head()

## Remove unnecessary columns from meta and only keep label for classification(diagnosis) and the sample name
meta = meta[['sample','dx']]

## Rename the column name "Group" to match the "sample" in meta
shared = shared.rename(index=str, columns={"Group":"sample"})

## Merge the 2 datasets on sample
data=pd.merge(meta,shared,on=['sample'])

## Remove adenoma samples. We will do a 2 classification model, just for cancer and normal colon samples.
data= data[data.dx.str.contains("adenoma") == False]

## Drop all except OTU columns for x
x = data.drop(["sample", "dx", "numOtus", "label"], axis=1)

## Cancer =1 Normal =0
diagnosis = { "cancer":1, "normal":0}

## Generate y which only has diagnosis as 0 and 1
y = data["dx"].replace(diagnosis)
# y = np.eye(2, dtype='uint8')[y]

## Drop if NA elements
y.dropna()
x.dropna()

```

#### Now let's define a L2 regularized logistic regression model
```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}

logreg = linear_model.LogisticRegression(C=0.001) # coefficient is chosen to increase penalty for weights (overfitting).
```

#### Plot ROC curve for cross validation of training set and then test set 
1. Data is split to 80-20 training-test 100 times in a loop. 
2. Each training set then is split to 80-20; 5 times to do a cross-validation ROC. A mean cross validation ROC is generated.
3. For each epoch, the test set that is generated in the beginning of the loop is predicted by the model and a mean test ROC is generated.

```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
cv = StratifiedKFold(n_splits=5, shuffle=True)

tprs = []
aucs = []
mean_fpr = np.linspace(0, 1, 100)

tprs_test = []
aucs_test = []
mean_fpr_test = np.linspace(0, 1, 100)

epochs = 100

for epoch in range(epochs):
    # For each epoch, generate new train and test sets by splitting randomly, 80/20 ratios
    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, shuffle=True)

    ## Converting to numpy array from pandas
    X=x_train.values
    Y=y_train.values
    X_test= x_test.values
    Y_test= y_test.values

    ## Plot mean ROC curve for 100 epochs of cross-validation with n_splits=5. The train set is used here for cross-validation.
    for train, test in cv.split(X,Y):
        probas_ = logreg.fit(X[train], Y[train]).predict_proba(X[test])
        fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
        tprs.append(interp(mean_fpr, fpr, tpr))
        tprs[-1][0] = 0.0
        roc_auc = auc(fpr, tpr)
        aucs.append(roc_auc)

    ## Plot mean ROC curve for 100 epochs of test set that is generated from splitting in the beginning of the loop.
    probas_ = logreg.predict_proba(X_test)
    # Compute ROC curve and area the curve
    fpr_test, tpr_test, thresholds_test = roc_curve(Y_test, probas_[:, 1])
    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
    tprs_test[-1][0] = 0.0
    roc_auc_test = auc(fpr_test, tpr_test)
    aucs_test.append(roc_auc_test)

## Now let's plot this
plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)
mean_tpr_test = np.mean(tprs_test, axis=0)
mean_tpr_test[-1] = 1.0
mean_auc_test = auc(mean_fpr_test, mean_tpr_test)
std_auc_test = np.std(aucs_test)
plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Mean test ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
std_tpr_test = np.std(tprs_test, axis=0)
tprs_upper_test = np.minimum(mean_tpr_test + std_tpr_test, 1)
tprs_lower_test = np.maximum(mean_tpr_test - std_tpr_test, 0)
plt.fill_between(mean_fpr_test, tprs_lower_test, tprs_upper_test, color='tomato', alpha=.2, label=r'$\pm$ 1 std. dev.')
mean_tpr = np.mean(tprs, axis=0)
mean_tpr[-1] = 1.0
mean_auc = auc(mean_fpr, mean_tpr)
std_auc = np.std(aucs)
plt.plot(mean_fpr, mean_tpr, color='b', label=r'Mean cross-val ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc, std_auc), lw=2, alpha=.8)
std_tpr = np.std(tprs, axis=0)
tprs_upper = np.minimum(mean_tpr + std_tpr, 1)
tprs_lower = np.maximum(mean_tpr - std_tpr, 0)
plt.fill_between(mean_fpr, tprs_lower, tprs_upper, color='dodgerblue', alpha=.2, label=r'$\pm$ 1 std. dev.')
plt.xlim([-0.05, 1.05])
plt.ylim([-0.05, 1.05])
plt.xlabel('False Positive Rate')
plt.ylabel('True Positive Rate')
plt.title('L2 Logistic Regression ROC\n')
plt.legend(loc="lower right", fontsize=8)
plt.show()
```

### 2018-08-27 LogisticRegression_BaxterData.Rmd
submit to Jenna

@@
 title: "Logistic Regression on Baxter colorectal cancer dataset"
-output: pdf_document
+output: html_document
 
@@
 from sklearn.metrics import roc_curve, auc
-from sklearn.model_selection import StratifiedKFold
+from sklearn.model_selection import RepeatedStratifiedKFold
+from sklearn.model_selection import GridSearchCV
 ```
@@
 ```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
+############## PRE-PROCESS DATA ######################
+
 ## Read in the data
@@
 
+## Split dataset to 80% training 20% test sets.
+
+x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2,shuffle=True)
+
 ```
@@
 #### Now let's define a L2 regularized logistic regression model
+- We tune our hyper-parameter(regualarization strength) here with 80-20 splits and 1000 repeats. 
+- Once we have our "C" value tuned, we define our best model.
 ```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
+######################### Logistic Regression ##############################
 
-logreg = linear_model.LogisticRegression(C=0.001) # coefficient is chosen to increase penalty for weights (overfitting).
+## Define L2 regularized logistic classifier
+logreg = linear_model.LogisticRegression()
+
+## Define the n-folds for hyper-parameter optimization on training set.
+cv = RepeatedStratifiedKFold(n_splits=5, n_repeats=1000, random_state=200889)
+## We will try these regularization strength coefficients to optimize our model
+C = {"C": [0.0000001, 0.000001, 0.00001, 0.0001]}
+## Define the best model:
+clf = GridSearchCV(logreg, C, cv=cv, verbose=0, scoring='roc_auc')
+best_model = clf.fit(x_train, y_train)
+print('Best C:', best_model.best_estimator_.get_params()['C'])
+print('Best model:', best_model.best_estimator_)
+## The best model we pick here will be used for predicting test set.
+best_model = best_model.best_estimator_
+
 ```
@@
 #### Plot ROC curve for cross validation of training set and then test set 
-1. Data is split to 80-20 training-test 100 times in a loop. 
-2. Each training set then is split to 80-20; 5 times to do a cross-validation ROC. A mean cross validation ROC is generated.
-3. For each epoch, the test set that is generated in the beginning of the loop is predicted by the model and a mean test ROC is generated.
+1. Plot mean ROC curve for cross-validation with n_splits=5 (80-20 stratified) and n_repeats=1000 to evaluate the variation of prediction in our training set.
+2. Plot ROC curve for never before seen test set.
 
 ```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
-cv = StratifiedKFold(n_splits=5, shuffle=True)
-
+## Now let's plot some ROC curves
 tprs = []
@@
 
-epochs = 100
-
-for epoch in range(epochs):
-    # For each epoch, generate new train and test sets by splitting randomly, 80/20 ratios
-    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2, shuffle=True)
+Logit_plot = plt.figure()
 
@@
 
-    ## Plot mean ROC curve for 100 epochs of cross-validation with n_splits=5. The train set is used here for cross-validation.
+## Plot mean ROC curve for cross-validation with n_splits=5 and n_repeats=1000 to evaluate the variation of prediction in our training set.
 for train, test in cv.split(X,Y):
-        probas_ = logreg.fit(X[train], Y[train]).predict_proba(X[test])
+    probas_ = best_model.fit(X[train], Y[train]).predict_proba(X[test])
     fpr, tpr, thresholds = roc_curve(Y[test], probas_[:, 1])
@@
 
-    ## Plot mean ROC curve for 100 epochs of test set that is generated from splitting in the beginning of the loop.
-    probas_ = logreg.predict_proba(X_test)
+## Predict and plot ROC curve for never before seen test set.
+probas_ = best_model.predict_proba(x_test)
 # Compute ROC curve and area the curve
-    fpr_test, tpr_test, thresholds_test = roc_curve(Y_test, probas_[:, 1])
+fpr_test, tpr_test, thresholds_test = roc_curve(y_test, probas_[:, 1])
 tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
@@
 
-## Now let's plot this
 plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)
@@
 std_auc_test = np.std(aucs_test)
-plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Mean test ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
+plt.plot(mean_fpr_test, mean_tpr_test, color='r', label=r'Never-before-seen test set ROC (AUC = %0.2f $\pm$ %0.2f)' % (mean_auc_test, std_auc_test), lw=2, alpha=.8)
 std_tpr_test = np.std(tprs_test, axis=0)
@@
 plt.show()
+#Logit_plot.savefig('results/figures/Logit_Baxter.png', dpi=1000)
+
 ```

### 2018-08-30 LogisticRegression_BaxterData.Rmd
epochs=100 will run the whole thing x100 times

@@
 from sklearn.model_selection import GridSearchCV
+from sklearn.externals import joblib
 ```
@@
 We have 292 samples with labels:
+
   - 120 Cancer colon
@@
 x.dropna()
-
-## Split dataset to 80% training 20% test sets.
-
-x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2,shuffle=True)
-
 ```
@@
 #### Now let's define a L2 regularized logistic regression model
-- We tune our hyper-parameter(regualarization strength) here with 80-20 splits and 1000 repeats. 
-- Once we have our "C" value tuned, we define our best model.
+1. We will select hyper-parameters and an optimal model during training with 80% of the data and test that best model on the remanining %20 x100 times.
+  - The only hyper-parameter in Logit here will be the strength of the L2-regularization
+2. For each epoch, the chosen best model will be tested on the %20 test set that was not seen before during trainig. We will report mean AUC values +/- sd for the testing/predicting.
+3. For each epoch, we will also be splitting the 80% training data 5-fold again, another 100 times and cross validate (inner loop). We will report mean AUC values +/- sd for each cross-validation.
+- 
 ```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
+## Generate empty lists to fill with AUC values for test-set
+tprs_test = []
+aucs_test = []
+mean_fpr_test = np.linspace(0, 1, 100)
+
 ######################### Logistic Regression ##############################
+
+epochs= 100
+for epoch in range(epochs):
+    ## Split dataset to 80% training 20% test sets.
+    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2,shuffle=True)
 
@@
     ## Define the n-folds for hyper-parameter optimization on training set.
-cv = RepeatedStratifiedKFold(n_splits=5, n_repeats=1000, random_state=200889)
+    cv = RepeatedStratifiedKFold(n_splits=5, n_repeats=100, random_state=200889)
     ## We will try these regularization strength coefficients to optimize our model
@@
 
-```
-
-#### Plot ROC curve for cross validation of training set and then test set 
-1. Plot mean ROC curve for cross-validation with n_splits=5 (80-20 stratified) and n_repeats=1000 to evaluate the variation of prediction in our training set.
-2. Plot ROC curve for never before seen test set.
-
-```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
-## Now let's plot some ROC curves
+    ## Generate empty lists to fill with AUC values for train-set cv
     tprs = []
@@
     mean_fpr = np.linspace(0, 1, 100)
-
-tprs_test = []
-aucs_test = []
-mean_fpr_test = np.linspace(0, 1, 100)
 
@@
 
-## Plot mean ROC curve for cross-validation with n_splits=5 and n_repeats=1000 to evaluate the variation of prediction in our training set.
+    ## Predict never-before-seen test-set
+    probas_ = best_model.predict_proba(x_test)
+    # Compute ROC curve and area the curve
+    fpr_test, tpr_test, thresholds_test = roc_curve(y_test, probas_[:, 1])
+    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
+    tprs_test[-1][0] = 0.0
+    roc_auc_test = auc(fpr_test, tpr_test)
+    aucs_test.append(roc_auc_test)
+    print("Test", roc_auc_test)
+
+    ## Plot mean ROC curve for cross-validation with n_splits=5 and n_repeats=100 to evaluate the variation of prediction in our training set.
     for train, test in cv.split(X,Y):
@@
         aucs.append(roc_auc)
+        print("Train", roc_auc)
 
-## Predict and plot ROC curve for never before seen test set.
-probas_ = best_model.predict_proba(x_test)
-# Compute ROC curve and area the curve
-fpr_test, tpr_test, thresholds_test = roc_curve(y_test, probas_[:, 1])
-tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
-tprs_test[-1][0] = 0.0
-roc_auc_test = auc(fpr_test, tpr_test)
-aucs_test.append(roc_auc_test)
+```
 
+#### Plot ROC curve for cross validation of training set and test set
+
+```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
 plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)

### 2018-08-30 code/LogisticRegression.py

@@ -52,62 +52,73 @@ y = data["dx"].replace(diagnosis)
 y.dropna()
 x.dropna()
 
-## Split dataset to 80% training 20% test sets.
-
-x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2,shuffle=True)
-
-######################### Logistic Regression ##############################
-
-## Define L2 regularized logistic classifier
-logreg = linear_model.LogisticRegression()
-
-## Define the n-folds for hyper-parameter optimization on training set.
-cv = RepeatedStratifiedKFold(n_splits=5, n_repeats=1000, random_state=200889)
-## We will try these regularization strength coefficients to optimize our model
-C = {"C": [0.0000001, 0.000001, 0.00001, 0.0001]}
-## Define the best model:
-clf = GridSearchCV(logreg, C, cv=cv, verbose=0, scoring='roc_auc')
-best_model = clf.fit(x_train, y_train)
-print('Best C:', best_model.best_estimator_.get_params()['C'])
-print('Best model:', best_model.best_estimator_)
-## The best model we pick here will be used for predicting test set.
-best_model = best_model.best_estimator_
-
-## Now let's plot some ROC curves
-tprs = []
-aucs = []
-mean_fpr = np.linspace(0, 1, 100)
-
+## Generate empty lists to fill with AUC values for test-set
 tprs_test = []
 aucs_test = []
 mean_fpr_test = np.linspace(0, 1, 100)
 
-Logit_plot = plt.figure()
+################## Logistic Regression ###############
+
+## We will select hyper-parameters and an optimal model during trainign with 80% of the data and test that best model on the remanining %20 x100 times.
+
+# For each epoch, the chosen best model will be tested on the %20 test set that was not seen before during trainig. We will report mean AUC values +/- sd for the testing/predicting.
+
+# For each epoch, we will also be splitting the 80% training data 5-fold again, another 100 times and cross validate (inner loop). We will report mean AUC values +/- sd for each cross-validation.
+
+epochs= 100
+for epoch in range(epochs):
+    ## Split dataset to 80% training 20% test sets.
+    x_train, x_test, y_train, y_test = train_test_split(x, y, test_size=0.2,shuffle=True)
+
+## Define L2 regularized logistic classifier
[…]

### 2018-09-05 LogisticRegression_BaxterData.Rmd

@@
 3. For each epoch, we will also be splitting the 80% training data 5-fold again, another 100 times and cross validate (inner loop). We will report mean AUC values +/- sd for each cross-validation.
-- 
+
 ```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
@@
     best_model = clf.fit(x_train, y_train)
-    print('Best C:', best_model.best_estimator_.get_params()['C'])
-    print('Best model:', best_model.best_estimator_)
     ## The best model we pick here will be used for predicting test set.
@@
     roc_auc_test = auc(fpr_test, tpr_test)
-    aucs_test.append(roc_auc_test)
-    print("Test", roc_auc_test)
 
@@
         aucs.append(roc_auc)
-        print("Train", roc_auc)
 

### 2018-09-05 LogisticRegression_BaxterData.Rmd
intro-methods-results

@@
 ---
+### Backgorund
+Colorectal cancer is one of the leading cause of death among cancers in the United States.
+Early diagnosis highly increases the chance of survival.
+However the current diagnostic methods are expensive and invasive.
+As a less invasive tool, numerous studies have used gut microbiome composition to predict disease progression.
+Each study uses different machine learning pipelines to classify disease progression which negatively impacts the reproducibility, robustness and actionability of this tool.
+### Goal of this project
+We want to evaluate the classification success of different machine learning methods.
+We also want to discuss the reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting of each method.
+### Methods
+We sequenced the 16S rRNA genes from the stool samples of 292 patients, 120 with colorectal cancer
+and 172 with healthy colons.
+We used the relative abundances of the bacterial populations within each sample to develop a L2-regularized logistic regression, SVM(linear), SVM(non-linear), decision tree, random forest, XGBoost and feed forward neural net(1 hidden layer) classification models that detects colonic lesions using the relative abundance of gut microbiota.
+### Results
+So far, I have used L2 logistic, Random Forest, PyTorch and XGboost to classify. I've also used the TPOT algorithm to see if this can be used by our scientific community when choosing ML pipelines.
+
+Here I'm sharing my code and my ROC plot for L2 Logistic Regression.
+It would be great to get feedback on my pipeline for this (the way I split the data, cross-validate and report AUC) so that I can implement that for all the other methods.
 
@@
 py_config()
-```
-
-#### Let's make sure we are using the Python version we want
-```{python,tidy = TRUE, comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
-import sys
-print(sys.version)
-```
-
-#### Let's make sure we are in our project directory
-```{python,tidy = TRUE,comment='[Output]', cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
-import os
-print(os.getcwd())
 ```
@@
 
-  - 120 Cancer colon
-  - 172 Normal colon
+* 120 Cancer colon
+* 172 Normal colon
 
-I would like to do a classification model to predict discrete values (diagnosis 1 or 0)
+We would like to do a classification model to predict discrete values (diagnosis 1 or 0)
 
@@
 #### Now let's define a L2 regularized logistic regression model
-1. We will select hyper-parameters and an optimal model during training with 80% of the data and test that best model on the remanining %20 x100 times.
-  - The only hyper-parameter in Logit here will be the strength of the L2-regularization
-2. For each epoch, the chosen best model will be tested on the %20 test set that was not seen before during trainig. We will report mean AUC values +/- sd for the testing/predicting.
-3. For each epoch, we will also be splitting the 80% training data 5-fold again, another 100 times and cross validate (inner loop). We will report mean AUC values +/- sd for each cross-validation.
+1.
+First we split the data 80-20%.
+We select hyper-parameters and an optimal model during training with the 80% and test that best model on the 20%.
+We repeat this(splitting,tuning,learning and testing x100 epochs)
+  * The only hyper-parameter that we tune will be the strength of the L2-regularization (C value).
+  * For each epoch there will be a new C value. (We get a distribution of C values over the 100 epochs)
+2. For each epoch, the chosen best model is being tested on the %20 test set that was not seen before during learning. We report mean AUC values +/- sd for this testing.
+3.
+For each epoch, we will also be splitting the 80% learning data(that we use to tune hyperparameter) 5-fold again, another 100 times and cross validate (inner loop).
+We will report mean AUC values +/- sd for each cross-validation.
 
@@
 
-    Logit_plot = plt.figure()
-
     ## Converting to numpy array from pandas
@@
     roc_auc_test = auc(fpr_test, tpr_test)
+    aucs_test.append(roc_auc_test)
 
@@
         aucs.append(roc_auc)
-
 ```
@@
 plt.show()
-#Logit_plot.savefig('results/figures/Logit_Baxter.png', dpi=1000)
-
 ```

### 2018-09-19 results/ML_outline.Rmd
outline

---
title: "Machine Learning Manuscript Outline"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

##### General Context of the work
As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning  methods to analyze microbiome sequence read count data and to determine association with a continuous or categorical phenotype of interest.

##### Narrower research area and statement of its importance
Colorectal cancer is one of the leading cause of death among cancers in the United States. Early diagnosis increases the chance of survival. However the current diagnostic methods are expensive and invasive. As a less invasive tool, numerous studies use relative abundances of the gut bacteria populations to predict disease progression.

##### Identification of a gap or other need for research
* Do we know what machine learning methods are available to us as microbial ecologists interested in gut microbiome?
* Each study that models colorectal cancer and microbiome association uses different machine learning pipelines to classify disease progression which negatively impacts the reproducibility, robustness and actionability of models.

##### Summary of appraoch to answer the research question
* We perform L2-regularized logistic regression, SVM, Decision tree, Random forest, XGBoost and Feed Forward Neural Net classification models.
* We evaluate the classification success of different machine learning methods. We also want to discuss the reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting of each method.

##### Announcement of principal findings
* Is there a maximum threshold of prediction with all these methods?
* What gives the best AUC value? 
* Does an increase in model complexity improve predictibility?
* Synthesis statement regarding modeling 16S microbiome data

## Methods
##### Brief explanation of study design/patient sampling and 16S rRNA gene sequencing/curation
* Refer to Baxter Dataset:
We sequenced the 16S rRNA genes from the stool samples of 292 patients, 120 with colorectal cancer
and 172 with healthy colons. We used the relative abundances of the bacterial populations within each sample to develop classification models that detects colonic lesions using the relative abundance of gut microbiota.

##### Analysis of data
* Pre-proccessing of the data
* Machine Learning pipeline backbone. How do I split, train, validate and test?
* Cross-validation and hyper-parameter tuning methods for each modeling method
* Programming languages and packages/modules utilized
* Statistical methods for comparison of model performance

## Results
##### Results of modeling in text, tables and figures
##### Comparisons among modeling approaches

## Discussion
#### Interpretation of modeling results in terms of reproducibility, robustness, actionability, interpretibility and susceptibility
#### Consideration of possible weaknesses for each model 
#### Consideration of possible weaknesses for our approach and chosen dataset
#### Relationship of results to previous literature and broader implications of this work
#### Prospects of future progress

Notes from Pat:
For the general context, I think it’s very important to hit a few things hard…

1. Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small. It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features. So tools like linear models (e.g. metastats, lefse, wilcoxon, etc) are likely worthless.

2. The interactions between the biomarkers may be nonlinear. Obviously, the linear models will not incorporate this because they are linear.

3. ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc. There is a need for guidance on how to properly implement these different methods

4. There is likely to be a tradeoff between interpretability and performance (perhaps this should go in a paragraph about ML methods that would also indicate which methods are linear/non-linear.

### 2018-09-19 code/learning/pyTorch_baxter.py
playing with pyTorch

@@ -198,16 +198,16 @@ for epoch in range(num_epochs):
         print('Epoch [%d], Loss:%.4f, Accuracy:%.4f' % (epoch, loss.data[0], correct_num/len(labels)))
         #plt.plot(fpr, tpr, lw=1, alpha=0.3, label='ROC fold %d (AUC = %0.2f)' % (epoch, roc_auc))
 
-    with torch.no_grad():
-        ypred_var_test = net(x_test)
-    correct_num_test = 0
-    values_test, labels_test = torch.max(ypred_var_test, 1)
-    correct_num_test = np.sum(labels_test.data.numpy() == ypred_var_test.numpy())
-    fpr_test, tpr_test, thresholds_test = roc_curve(y_test.numpy(), labels_test.data.numpy())
-    tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
-    tprs_test[-1][0] = 0.0
-    roc_auc_test = auc(fpr_test, tpr_test)
-    aucs_test.append(roc_auc_test)
+        with torch.no_grad():
+            ypred_var_test = net(x_test)
+            correct_num_test = 0
+            values_test, labels_test = torch.max(ypred_var_test, 1)
+            correct_num_test = np.sum(labels_test.data.numpy() == ypred_var_test.numpy())
+            fpr_test, tpr_test, thresholds_test = roc_curve(y_test.numpy(), labels_test.data.numpy())
+            tprs_test.append(interp(mean_fpr_test, fpr_test, tpr_test))
+            tprs_test[-1][0] = 0.0
+            roc_auc_test = auc(fpr_test, tpr_test)
+            aucs_test.append(roc_auc_test)
 
 
 plt.plot([0, 1], [0, 1], linestyle='--', color='green', label='Luck', alpha=.8)

### 2018-09-19 results/ML_outline.Rmd
shared with Jenna

@@
 ##### Narrower research area and statement of its importance
-Colorectal cancer is one of the leading cause of death among cancers in the United States.
+* Colorectal cancer is one of the leading cause of death among cancers in the United States.
 Early diagnosis increases the chance of survival.
@@
 As a less invasive tool, numerous studies use relative abundances of the gut bacteria populations to predict disease progression.
+* Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small.
+It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features.
 
 ##### Identification of a gap or other need for research
-* Do we know what machine learning methods are available to us as microbial ecologists interested in gut microbiome?
-* Each study that models colorectal cancer and microbiome association uses different machine learning pipelines to classify disease progression which negatively impacts the reproducibility, robustness and actionability of models.
+* ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc.
+There is a need for guidance on how to properly implement these different methods.
+* We need to emphasize good machine learning practices and pipelines and discuss the reproducibility, robustness and actionability of models.
 
 ##### Summary of appraoch to answer the research question
-* We perform L2-regularized logistic regression, SVM, Decision tree, Random forest, XGBoost and Feed Forward Neural Net classification models.
+* We establish a non-leaky pipeline. 
+* We perform L1 and L2-regularized logistic regression, Linear SVM, Non-Linear SVM, Decision tree, Random forest, XGBoost and Feed Forward Neural Net classification models.
 * We evaluate the classification success of different machine learning methods.
@@
 ##### Announcement of principal findings
+* Generalisation Perfomance of each model. 
 * Is there a maximum threshold of prediction with all these methods?
-* What gives the best AUC value? 
 * Does an increase in model complexity improve predictibility?
@@
 ##### Analysis of data
+* What is the data (temporal or not)? Assumptions we make when we use a dataset. What will the future data look like?
 * Pre-proccessing of the data
@@
 * Statistical methods for comparison of model performance
+* There is likely to be a tradeoff between interpretability and performance (perhaps this should go in a paragraph about ML methods that would also indicate which methods are linear/non-linear.)
 
@@
 #### Consideration of possible weaknesses for each model 
+* The interactions between the biomarkers may be nonlinear.
+Obviously, the linear models will not incorporate this because they are linear.
+Tools like linear models (e.g.
+metastats, lefse, wilcoxon, etc) are likely worthless.
+
 #### Consideration of possible weaknesses for our approach and chosen dataset
@@
 #### Prospects of future progress
-
-Notes from Pat:
-For the general context, I think it’s very important to hit a few things hard…
-
-1.
-Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small.
-It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features.
-So tools like linear models (e.g.
-metastats, lefse, wilcoxon, etc) are likely worthless.
-
-2. The interactions between the biomarkers may be nonlinear. Obviously, the linear models will not incorporate this because they are linear.
-
-3.
-ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc.
-There is a need for guidance on how to properly implement these different methods
-
-4. There is likely to be a tradeoff between interpretability and performance (perhaps this should go in a paragraph about ML methods that would also indicate which methods are linear/non-linear.

### 2018-10-18 submission/manuscript.Rmd
initial commit

---
title: "**NAME OF THIS STUDY**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
# Load useful functions
source('../code/learning/functions.R')
# Read in necessary libraries
deps = c("doParallel","pROC", "caret","knitr","rmarkdown","vegan","gtools", "tidyverse");
for (dep in deps){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
# Read in metadata and select only sample Id and diagnosis columns
meta <- read.delim('../data/metadata.tsv', header=T, sep='\t') %>%
  select(sample, dx)

# Read in OTU table and remove label and numOtus columns
shared <- read.delim('../data/baxter.0.03.subsample.shared', header=T, sep='\t') %>%
   select(-label, -numOtus)
```

```{r BinaryPreProcess, eval=TRUE, echo=FALSE, cache=FALSE}
# Merge metadata and OTU table and remove all the samples that are diagnosed with adenomas. Keep only cancer and normal.
# Then remove the sample ID column
data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
  filter(dx != 'adenoma') %>%
  select(-sample)

# We want the diagnosis column to a factor
data$dx <- factor(data$dx, labels=c("normal", "cancer"))
```

```{r BinaryClassification, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
cl <- makePSOCKcluster(4)
registerDoParallel(cl)
# Create
all.test.response <- all.test.predictor <- test_aucs <- c()
for (i in 1:50) {
  inTraining <- createDataPartition(data$dx, p = .80, list = FALSE)
  training <- data[ inTraining,]
  testing  <- data[-inTraining,]
  x_train <- training %>% select(-dx)
  y_train <- training$dx
  y_train <- as.factor(y_train)
  # L2 regularization is used for Logistic regression
  # cost is the inverse of penalty and we provide a grid
  grid <-  expand.grid(cost = c(0.0000001, 0.000001, 0.00001, 0.0001),
                       loss = "L2_dual",
                       epsilon = 0.1)
  cv <- trainControl(method="repeatedcv",
                     repeats = 50,
                     number=5,
                     returnResamp="final",
                     classProbs=TRUE,
                     summaryFunction=twoClassSummary,
                     indexFinal=NULL,
                     preProc = "scale",
                     savePredictions = TRUE)

  L2Logit <- train(x_train, y_train,
                               method = "regLogistic",
                               trControl = cv,
                               metric = "ROC",
                               tuneGrid = grid,
                               preProcess = c("center", "scale"), 
                               family = "binomial")
  # Mean AUC value of the best lambda parameter training over repeats
  cv_auc <- getTrainPerf(L2Logit)$TrainROC
  # Predict on the test set and get predicted probabilities
  probs <- predict(L2Logit, testing, type="prob")
  # Test AUC calculation
  test_roc <- roc(ifelse(testing$dx == "cancer", 1, 0), probs[[2]])
  test_auc <- test_roc$auc
  # Save all the test AUCs over iterations in test_aucs
  test_aucs <- c(test_aucs, test_auc)
  # Save the test set labels in all.test.response. Labels converted to 0 for normal and 1 for cancer
  all.test.response <- c(all.test.response, ifelse(testing$dx == "cancer", 1, 0))
  # Save the test set predicted probabilities of highest class in all.test.predictor
  all.test.predictor <- c(all.test.predictor, probs[[2]])
}
stopCluster(cl)
# Get the ROC curve and AUC values for all the iterations 
test_roc <- roc(all.test.response, all.test.predictor, auc=TRUE, ci=TRUE)
```

```{r LogitROCplot, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

pdf("../results/figures/LogReg_inR.pdf")
par(mar=c(4,4,1,1))
# Plot random line on ROC curve
plot(c(1,0),c(0,1),
     type='l',
     lty=3,
     xlim=c(1.01,0), ylim=c(-0.01,1.01),
     xaxs='i', yaxs='i',
     ylab='', xlab='')
# Plot Test ROC in red line
plot(test_roc,
     col='black',
     lwd=2,
     add=T,
     lty=1)
#Calculate Confidence Interval sensitivities at given specificities
sens.ci <- ci.se(test_roc)
# Compute the CI of the AUC
auc.ci <- ci.auc(test_roc)
# Plot CV ROC in blue line
#plot(cv_roc,
#     col='blue',
#     lwd=2,
#     add=T,
#     lty=1)
# Label the axes
mtext(side=2,
      text="Sensitivity",
      line=2.5,
      cex=1.5)
mtext(side=1,
      text="Specificity",
      line=2.5,
      cex=1.5)
# Add legends for both lines
legend(x=0.7,y=0.2,
       legend=(sprintf('Test AUC: %.3g, CI: %.3g', test_roc$auc, auc.ci)),
       bty='n',
       xjust=0,
       lty=c(1,1),
       col='black',
       text.col='black')

plot(sens.ci, type="shape", col="gray88")

# Save the figure
dev.off()
```

\vspace{35mm}

Running title: INSERT RUNNING TITLE HERE

\vspace{35mm}

Begüm D. Topçuoğlu^1, Jenna Wiens^2, Patrick D. Schloss^1$\dagger$^

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109

\newpage
\linenumbers

## Abstract

\newpage

## Introduction

As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning  methods to analyze microbiome sequence read count data and to determine association with a continuous or categorical phenotype of interest.

Colorectal cancer is one of the leading cause of death among cancers in the United States. Early diagnosis increases the chance of survival. However the current diagnostic methods are expensive and invasive. As a less invasive tool, numerous studies use relative abundances of the gut bacteria populations to predict disease progression. Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small. It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features.

ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc. There is a need for guidance on how to properly implement these different methods. We need to emphasize good machine learning practices and pipelines and discuss the reproducibility, robustness and actionability of models.

We established a non-leaky pipeline. We performed L1 and L2-regularized logistic regression, Linear SVM, Non-Linear SVM, Decision tree, Random forest, XGBoost and Feed Forward Neural Net classification models. We evaluated the classification performance of different machine learning methods. We also want to discuss the reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting of each method.

Generalisation Perfomance of each model. 
Is there a maximum threshold of prediction with all these methods?
Does an increase in model complexity improve predictibility?
Synthesis statement regarding modeling 16S microbiome data

## Results and Discussion

## Conclusions

## Materials and Methods

\newpage

Insert figure legends with the first sentence in bold, for example:

**Figure 1. Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B). Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).

\newpage

## References

### 2018-10-18 submission/manuscript.tex
initial commit

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={NAME OF THIS STUDY},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{\textbf{NAME OF THIS STUDY}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\begin{document}
\maketitle

\begin{verbatim}
## Loading required package: foreach
\end{verbatim}

\begin{verbatim}
## Loading required package: iterators
\end{verbatim}

\begin{verbatim}
## Loading required package: parallel
\end{verbatim}

\begin{verbatim}
## Type 'citation("pROC")' for a citation.
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'pROC'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:stats':
## 
##     cov, smooth, var
\end{verbatim}

\begin{verbatim}
## Loading required package: lattice
\end{verbatim}

\begin{verbatim}
## Loading required package: ggplot2
\end{verbatim}

\begin{verbatim}
## Loading required package: permute
\end{verbatim}

\begin{verbatim}
## This is vegan 2.5-2
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'vegan'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:caret':
## 
##     tolerance
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'gtools'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:permute':
## 
##     permute
\end{verbatim}

\begin{verbatim}
## -- Attaching packages ---------------------------------- tidyverse 1.2.1 --
\end{verbatim}

\begin{verbatim}
## v tibble  1.4.2     v purrr   0.2.5
## v tidyr   0.8.1     v dplyr   0.7.6
## v readr   1.1.1     v stringr 1.3.1
## v tibble  1.4.2     v forcats 0.3.0
\end{verbatim}

\begin{verbatim}
## -- Conflicts ------------------------------------- tidyverse_conflicts() --
## x purrr::accumulate() masks foreach::accumulate()
## x dplyr::filter()     masks stats::filter()
## x dplyr::lag()        masks stats::lag()
## x purrr::lift()       masks caret::lift()
## x purrr::when()       masks foreach::when()
\end{verbatim}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\vspace{35mm}

Running title: INSERT RUNNING TITLE HERE

\vspace{35mm}

Begüm D. Topçuoğlu\^{}1, Jenna Wiens\^{}2, Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

2. Department of Computer Science and Engineering, University or
Michigan, Ann Arbor, MI 49109

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

\newpage

\subsection{Introduction}\label{introduction}

As gut microbiome field continues to grow, there will be an
ever-increasing demand for reproducible machine learning methods to
analyze microbiome sequence read count data and to determine association
with a continuous or categorical phenotype of interest.

Colorectal cancer is one of the leading cause of death among cancers in
the United States. Early diagnosis increases the chance of survival.
However the current diagnostic methods are expensive and invasive. As a
less invasive tool, numerous studies use relative abundances of the gut
bacteria populations to predict disease progression. Most microbial
communities are pretty patchy and the likelihood of a single feature
that explains the differences in health is pretty small. It is likely
that many biomarkers are needed to account for the patchiness as well as
the context dependency of the features.

ML use in microbiome literature is a bit like the wild west with lack of
clarity over methods, testing, validation, etc. There is a need for
guidance on how to properly implement these different methods. We need
to emphasize good machine learning practices and pipelines and discuss
the reproducibility, robustness and actionability of models.

We established a non-leaky pipeline. We performed L1 and L2-regularized
logistic regression, Linear SVM, Non-Linear SVM, Decision tree, Random
forest, XGBoost and Feed Forward Neural Net classification models. We
evaluated the classification performance of different machine learning
methods. We also want to discuss the reproducibility, robustness,
actionability, interpretibility and susceptibility to overfitting of
each method.

Generalisation Perfomance of each model. Is there a maximum threshold of
prediction with all these methods? Does an increase in model complexity
improve predictibility? Synthesis statement regarding modeling 16S
microbiome data

\subsection{Results and Discussion}\label{results-and-discussion}

\subsection{Conclusions}\label{conclusions}

\subsection{Materials and Methods}\label{materials-and-methods}

\newpage

Insert figure legends with the first sentence in bold, for example:

\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
16S rRNA gene sequences for different OTU definitions and level of
sequencing effort.} Rarefaction curves for different OTU definitions of
Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
environments in Table 1 for Bacteria (C) and Archaea (D).

\newpage

\subsection{References}\label{references}

\end{document}

### 2018-10-22 submission/manuscript.Rmd
working on ML manuscript

@@
 data$dx <- factor(data$dx, labels=c("normal", "cancer"))
+# calculate the pre-process parameters from the dataset
+# in python we use StandardScaler, here we center and scale to get the same effect
+preprocessParams <- preProcess(data[,-1], method=c("center", "scale"))
+# transform the dataset using the parameters
+data[,-1] <- predict(preprocessParams, data[,-1])
 ```
@@
 ```{r BinaryClassification, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
-cl <- makePSOCKcluster(4)
-registerDoParallel(cl)
-# Create
+# Create lists to store the predicted labels and actuals labels at each iteration 
 all.test.response <- all.test.predictor <- test_aucs <- c()
+
+## We will split the dataset 80%-20% and tune hyper-parameter on the 80% training.
+This tuning step on training data will be done 50 times with 5 folds on the training set and an optimal hyper-parameter/optimal model will be chosen.
+
+## The chosen best model and hyper-parameter will be tested on the %20 test set that was not seen before during training. This will give a TEST AUC.
+
+## We will do the previous steps 50 iterations.
+Each iteration will have a unique hyper-parameter and optimal model.
+For each of these models we will have a TEST AUC.
+We will report the TEST AUC with 95 %CI.
+
+# For each epoch, we can also keep track of AUC values with CI for each cross-validation during training.But these code lines are not included in this Rmd file.
+I will have those commented out in the main.R code file.
 for (i in 1:50) {
+  # Partition the data 80-20. This function in caret keeps it stratified if I remember correctly.
   inTraining <- createDataPartition(data$dx, p = .80, list = FALSE)
@@
   # L2 regularization is used for Logistic regression
-  # cost is the inverse of penalty and we provide a grid
+  # Cost is our regularization parameter which here is the inverse of penalty and we provide a grid to tune the best one. 
   grid <-  expand.grid(cost = c(0.0000001, 0.000001, 0.00001, 0.0001),
@@
                        epsilon = 0.1)
+
   cv <- trainControl(method="repeatedcv",
@@
                      indexFinal=NULL,
-                     preProc = "scale",
                      savePredictions = TRUE)
 
-  L2Logit <- train(x_train, y_train,
+  L2Logit <- train(x_train, 
+                   y_train,
                    method = "regLogistic",
@@
                    tuneGrid = grid,
-                               preProcess = c("center", "scale"), 
                    family = "binomial")
@@
 }
-stopCluster(cl)
 # Get the ROC curve and AUC values for all the iterations
+# auc= TRUE and ci=TRUE will keep track of the auc and 95 CI for all the sensitivities and specificities. 
 test_roc <- roc(all.test.response, all.test.predictor, auc=TRUE, ci=TRUE)
@@
 
-```{r LogitROCplot, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+```{r LogitROCplot, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
+# Here we plot the test AUC and 95%CI for logistic regression.
 
@@
      ylab='', xlab='')
-# Plot Test ROC in red line
+# Plot Test ROC in black line
 plot(test_roc,
@@
 sens.ci <- ci.se(test_roc)
-# Compute the CI of the AUC
-auc.ci <- ci.auc(test_roc)
+# Calculate CI for AUC values
+auc_ci <- test_roc$ci[3]-test_roc$ci[2]
+
 # Plot CV ROC in blue line
@@
 legend(x=0.7,y=0.2,
-       legend=(sprintf('Test AUC: %.3g, CI: %.3g', test_roc$auc, auc.ci)),
+       legend=(sprintf('Test AUC: %.3g, %s CI: %0.3g', test_roc$auc, '%95', auc_ci)),
        bty='n',

### 2018-10-22 submission/manuscript.tex
working on ML manuscript

@@
 \begin{verbatim}
-## Loading required package: foreach
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: iterators
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: parallel
-\end{verbatim}
-
-\begin{verbatim}
-## Type 'citation("pROC")' for a citation.
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'pROC'
-\end{verbatim}
-
-\begin{verbatim}
-## The following objects are masked from 'package:stats':
-## 
-##     cov, smooth, var
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: lattice
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: ggplot2
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: permute
-\end{verbatim}
-
-\begin{verbatim}
-## This is vegan 2.5-2
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'vegan'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:caret':
-## 
-##     tolerance
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'gtools'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:permute':
-## 
-##     permute
-\end{verbatim}
-
-\begin{verbatim}
-## -- Attaching packages ---------------------------------- tidyverse 1.2.1 --
-\end{verbatim}
-
-\begin{verbatim}
-## v tibble  1.4.2     v purrr   0.2.5
-## v tidyr   0.8.1     v dplyr   0.7.6
-## v readr   1.1.1     v stringr 1.3.1
-## v tibble  1.4.2     v forcats 0.3.0
-\end{verbatim}
-
-\begin{verbatim}
-## -- Conflicts ------------------------------------- tidyverse_conflicts() --
-## x purrr::accumulate() masks foreach::accumulate()
-## x dplyr::filter()     masks stats::filter()
-## x dplyr::lag()        masks stats::lag()
-## x purrr::lift()       masks caret::lift()
-## x purrr::when()       masks foreach::when()
-\end{verbatim}
-
-\begin{verbatim}
 ## pdf 

### 2018-10-22 code/learning/LogisticRegression.R

@@ -111,6 +111,10 @@ plot(test_roc,
      lwd=2,
      add=T,
      lty=1)
+#Calculate Confidence Interval sensitivities at given specificities
+sens.ci <- ci.se(test_roc)
+# Compute the CI of the AUC
+auc.ci <- ci.auc(test_roc)
 # Plot CV ROC in blue line
 #plot(cv_roc,
 #     col='blue',
@@ -128,14 +132,13 @@ mtext(side=1,
       cex=1.5)
 # Add legends for both lines
 legend(x=0.7,y=0.2,
-       legend=(sprintf('Test - AUC: %.3g', test_roc$auc)),
+       legend=(sprintf('Test - AUC: %.3g, CI: %.3g', test_roc$auc, sens.ci)),
        bty='n',
        xjust=0,
        lty=c(1,1),
        col='black',
        text.col='black')
 
-sens.ci <- ci.se(test_roc)
 plot(sens.ci, type="shape", col="gray88")
 
 

### 2018-10-23 submission/manuscript.Rmd
trouble with confidence interval plot

@@
   # Save the test set labels in all.test.response. Labels converted to 0 for normal and 1 for cancer
-  all.test.response <- c(all.test.response, ifelse(testing$dx == "cancer", 1, 0))
+  all.test.response <- c(all.test.response, 
+                         ifelse(testing$dx == "cancer", 1, 0))
   # Save the test set predicted probabilities of highest class in all.test.predictor
-  all.test.predictor <- c(all.test.predictor, probs[[2]])
+  all.test.predictor <- c(all.test.predictor, 
+                          probs[[2]])
 }
@@
 # auc= TRUE and ci=TRUE will keep track of the auc and 95 CI for all the sensitivities and specificities. 
-test_roc <- roc(all.test.response, all.test.predictor, auc=TRUE, ci=TRUE)
+test_roc <- roc(all.test.response, 
+                all.test.predictor, 
+                auc=TRUE, 
+                ci=TRUE)
 ```
@@
 #Calculate Confidence Interval sensitivities at given specificities
-sens.ci <- ci.se(test_roc)
+sens.ci <- ci.se(test_roc, 
+          specificities=seq(0, 100, 5)) # over a select set of specificities
+
 # Calculate CI for AUC values
@@
 legend(x=0.7,y=0.2,
-       legend=(sprintf('Test AUC: %.3g, %s CI: %0.3g', test_roc$auc, '%95', auc_ci)),
+       legend=(sprintf('Test AUC: %.3g, %s CI: %0.3g', 
+                       test_roc$auc, 
+                       '%95', 
+                       auc_ci)),
        bty='n',

### 2018-11-06 submission/manuscript.Rmd

--- 2018-10-18 submission/manuscript.Rmd
@@
    select(-label, -numOtus)
+
+# Read in AUCs table generated from l2 logistic regression model for all samples
+  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+  # FIT is a feature
+  # The scaler is MinMax scaler
+logit <- read.delim('../data/process/L2_Logistic_Regression.tsv', header=T, sep='\t') %>%
+  select(level_1, AUC) %>% 
+  rename(Performance = level_1) %>% 
+  mutate(model="L2-Logistic Regression")
+
+# Read in AUCs table generated from l1 SVM linear kernel for all samples
+  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+  # FIT is a feature
+  # The scaler is StandardScaler
+l1svm <- read.delim('../data/process/L1_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
+  select(level_1, AUC) %>% 
+  rename(Performance = level_1) %>% 
+  mutate(model="L1-SVM Linear")
+
+# Read in AUCs table generated from l2 SVM linear kernel for all samples
+  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+  # FIT is a feature
+  # The scaler is StandardScaler
+l2svm <- read.delim('../data/process/L2_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
+  select(level_1, AUC) %>% 
+  rename(Performance = level_1) %>% 
+  mutate(model="L2-SVM Linear")
+
+# Read in AUCs table generated from  SVM RBF kernel for all samples
+  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+  # FIT is a feature
+  # The scaler is StandardScaler
+svmRBF <- read.delim('../data/process/SVM_RBF.tsv', header=T, sep='\t') %>%
+  select(level_1, AUC) %>% 
+  rename(Performance = level_1) %>% 
+  mutate(model="SVM RBF")
+
+# Read in AUCs table generated from xgboost for all samples
+  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+  # FIT is a feature
+  # The scaler is MinMaxScaler
+xgboost <- read.delim('../data/process/XGBoost.tsv', header=T, sep='\t') %>%
+  select(level_1, AUC) %>% 
+  rename(Performance = level_1) %>% 
+  mutate(model="XGBoost")
+
+# Read in AUCs table generated from random forest for all samples
+  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+  # FIT is a feature
+  # The scaler is MinMaxScaler
+rf <- read.delim('../data/process/Random_Forest.tsv', header=T, sep='\t') %>%
+  select(level_1, AUC) %>% 
+  rename(Performance = level_1) %>% 
+  mutate(model="Random Forest ")
+
+# Read in AUCs table generated from decision tree for all samples
+  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+  # FIT is a feature
+  # The scaler is MinMaxScaler
+dt <- read.delim('../data/process/Decision_Tree.tsv', header=T, sep='\t') %>%
+  select(level_1, AUC) %>% 
+  rename(Performance = level_1) %>% 
+  mutate(model="Decision Tree")
+
+# Put  all the AUCs for all the models together and group by model
+all <- bind_rows(logit, l1svm, l2svm, svmRBF, xgboost, rf, dt) %>% 
+  group_by(model)
 ```
 
-```{r BinaryPreProcess, eval=TRUE, echo=FALSE, cache=FALSE}
-# Merge metadata and OTU table and remove all the samples that are diagnosed with adenomas. Keep only cancer and normal.
-# Then remove the sample ID column
-data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
-  filter(dx != 'adenoma') %>%
-  select(-sample)
-
-# We want the diagnosis column to a factor
-data$dx <- factor(data$dx, labels=c("normal", "cancer"))
+```{r model_parameters, echo=FALSE, results='asis'}
+model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t')
+kable(model_param, caption = "Table 1 Parameters and software implementation of the classification algorithms")
 ```
-
-```{r BinaryClassification, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
-cl <- makePSOCKcluster(4)
-registerDoParallel(cl)
-# Create
-all.test.response <- all.test.predictor <- test_aucs <- c()
-for (i in 1:50) {
-  inTraining <- createDataPartition(data$dx, p = .80, list = FALSE)
-  training <- data[ inTraining,]
-  testing  <- data[-inTraining,]
-  x_train <- training %>% select(-dx)
-  y_train <- training$dx
-  y_train <- as.factor(y_train)
-  # L2 regularization is used for Logistic regression
-  # cost is the inverse of penalty and we provide a grid
-  grid <-  expand.grid(cost = c(0.0000001, 0.000001, 0.00001, 0.0001),
-                       loss = "L2_dual",
-                       epsilon = 0.1)
-  cv <- trainControl(method="repeatedcv",
-                     repeats = 50,
-                     number=5,
-                     returnResamp="final",
-                     classProbs=TRUE,
-                     summaryFunction=twoClassSummary,
-                     indexFinal=NULL,
-                     preProc = "scale",
-                     savePredictions = TRUE)
-
-  L2Logit <- train(x_train, y_train,
-                               method = "regLogistic",
-                               trControl = cv,
-                               metric = "ROC",
-                               tuneGrid = grid,
-                               preProcess = c("center", "scale"), 
-                               family = "binomial")
-  # Mean AUC value of the best lambda parameter training over repeats
-  cv_auc <- getTrainPerf(L2Logit)$TrainROC
-  # Predict on the test set and get predicted probabilities
-  probs <- predict(L2Logit, testing, type="prob")
-  # Test AUC calculation
-  test_roc <- roc(ifelse(testing$dx == "cancer", 1, 0), probs[[2]])
-  test_auc <- test_roc$auc
-  # Save all the test AUCs over iterations in test_aucs
-  test_aucs <- c(test_aucs, test_auc)
-  # Save the test set labels in all.test.response. Labels converted to 0 for normal and 1 for cancer
-  all.test.response <- c(all.test.response, ifelse(testing$dx == "cancer", 1, 0))
-  # Save the test set predicted probabilities of highest class in all.test.predictor
-  all.test.predictor <- c(all.test.predictor, probs[[2]])
-}
-stopCluster(cl)
-# Get the ROC curve and AUC values for all the iterations 
-test_roc <- roc(all.test.response, all.test.predictor, auc=TRUE, ci=TRUE)
-```
-
-```{r LogitROCplot, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
-
-pdf("../results/figures/LogReg_inR.pdf")
-par(mar=c(4,4,1,1))
-# Plot random line on ROC curve
-plot(c(1,0),c(0,1),
-     type='l',
-     lty=3,
-     xlim=c(1.01,0), ylim=c(-0.01,1.01),
-     xaxs='i', yaxs='i',
-     ylab='', xlab='')
-# Plot Test ROC in red line
-plot(test_roc,
-     col='black',
-     lwd=2,
-     add=T,
-     lty=1)
-#Calculate Confidence Interval sensitivities at given specificities
-sens.ci <- ci.se(test_roc)
-# Compute the CI of the AUC
-auc.ci <- ci.auc(test_roc)
-# Plot CV ROC in blue line
-#plot(cv_roc,
-#     col='blue',
-#     lwd=2,
-#     add=T,
-#     lty=1)
-# Label the axes
-mtext(side=2,
-      text="Sensitivity",
-      line=2.5,
-      cex=1.5)
-mtext(side=1,
-      text="Specificity",
-      line=2.5,
-      cex=1.5)
-# Add legends for both lines
-legend(x=0.7,y=0.2,
-       legend=(sprintf('Test AUC: %.3g, CI: %.3g', test_roc$auc, auc.ci)),
-       bty='n',
-       xjust=0,
-       lty=c(1,1),
-       col='black',
-       text.col='black')
-
-plot(sens.ci, type="shape", col="gray88")
-
-# Save the figure
-dev.off()
-```
-
 \vspace{35mm}

### 2018-11-06 submission/manuscript.Rmd
playing with tables

@@
 # Read in necessary libraries
-deps = c("doParallel","pROC", "caret","knitr","rmarkdown","vegan","gtools", "tidyverse");
+deps = c("kableExtra", "kable","doParallel","pROC", "caret","knitr","rmarkdown","vegan","gtools", "tidyverse");
 for (dep in deps){
@@
 model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t')
-kable(model_param, caption = "Table 1 Parameters and software implementation of the classification algorithms")
+kable(model_param, caption = "Table 1 Parameters and software implementation of the classification algorithms") %>%
+  kable_styling("striped", full_width = F)
 ```

### 2018-11-06 submission/manuscript.tex
playing with tables

@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
 
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-
-\vspace{35mm}
+\begin{longtable}[]{@{}lllllrl@{}}
+\caption{Table 1 Parameters and software implementation of the
+classification algorithms}\tabularnewline
+\toprule
+Method & Parameter & Value & Cross.Validation & Scaler & Epochs &
+Sofware.Implementation\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Method & Parameter & Value & Cross.Validation & Scaler & Epochs &
+Sofware.Implementation\tabularnewline
+\midrule
+\endhead
+Logistic Regression & C & {[}0.01, 0.1, 1, 10{]} & 5-fold 100 repeats &
+MinMaxScaler & 100 & Sklearn-LogisticRegression\tabularnewline
+L1 SVM Linear Kernel & C & {[}0.001, 0.01, 0.1 ,1{]} & 5-fold 100
+repeats & StandardScaler & 100 & Sklearn-LinearSVC\tabularnewline
+L2 SVM Linear Kernel & C & {[}0.01, 0.1, 1{]} & 5-fold 100 repeats &
+StandardScaler & 100 & Sklearn-LinearSVC\tabularnewline
+SVM RBF Kernel & C, gamma & {[}0.000001, 0.00001, 0.0001, 0.001,
+0.01{]}, {[}0.000000001, 0.00000001, 0.0000001{]} & 5-fold 100 repeats &
+StandardScaler & 100 & Sklearn-SVC\tabularnewline
+Decision Tree & max\_depth, min\_samples\_split & {[}6, 8, 10, 50{]},
+{[}10, 25, 50{]} & 5-fold 100 repeats & MinMaxScaler & 100 &
+DecisionTreeClassifier\tabularnewline
+Random Forest & n\_estimators, max\_features & 1000, {[}10, 80, 500,
+1000, 1500{]} & 5-fold 100 repeats & MinMaxScaler & 100 &
+RandomForestClassifier\tabularnewline
+XGBoost & n\_estimators, colsample\_bytree, learning\_rate, subsample,
+max\_depth, min\_child\_weight & 100, 0.8, {[}0.01, 0.1, 1{]},
+{[}0.7,0.8,0.9{]}, {[}6,7,8{]}, {[}1,2,3{]} & 5-fold 100 repeats &
+MinMaxScaler & 100 & XGBClassifier\tabularnewline
+\begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
+\vspace{35mm}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
+\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
+\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
+\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
+\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedleft\strut
+\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
 

### 2018-11-06 submission/manuscript.Rmd
styling

@@
 ```{r model_parameters, echo=FALSE, results='asis'}
-model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t')
+model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t') %>% 
+  select(-Value)
 kable(model_param, caption = "Table 1 Parameters and software implementation of the classification algorithms") %>%
-  kable_styling("striped", full_width = F)
+  kable_styling("striped", full_width = F, font_size=10)
 ```

### 2018-11-07 submission/manuscript.Rmd
table for model parameters added

--- 2018-11-06 submission/manuscript.Rmd
@@
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
-# Load useful functions
-source('../code/learning/functions.R')
 # Read in necessary libraries
@@
 knitr::knit_hooks$set(inline=inline_hook)
+options(kableExtra.latex.load_packages = FALSE)
+options(kableExtra.auto_format = FALSE)
+options(knitr.table.format = "latex")
+library(kableExtra)
 ```
@@
 ```{r model_parameters, echo=FALSE, results='asis'}
-model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t')
-kable(model_param, caption = "Table 1 Parameters and software implementation of the classification algorithms")
+model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t') %>% 
+   select(Method, Parameter, Cross.Validation, Epochs, Scaler, Sklearn.Function) %>% 
+  rename("Cross Validation"="Cross.Validation") %>%
+  rename("Epoch"="Epochs") %>%
+  rename("Sklearn Function"="Sklearn.Function")
+
+Table1 <- kable(model_param, "latex", caption = "Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms", booktabs = T) %>%
+kable_styling(latex_options = c("striped", "scale_down"), position = "left") %>%  
+ column_spec(1, width="10em") %>%
+  column_spec(2, width="8em") %>% 
+  column_spec(3, width="9em") %>%
+  column_spec(5, width="5em") %>% 
+  row_spec(1:7, align="l") %>% 
+  row_spec(0,bold=TRUE)
+
+param_range <- read.table('../data/process/model_parameters.txt', header=T, sep='\t', na.strings="", stringsAsFactors = FALSE) %>% 
+  replace(is.na(.), "-") %>%  
+  select(-Value, -Parameter, -Cross.Validation, -Epochs, -Scaler, -Sklearn.Function)
+
+Table2 <- kable(param_range, "latex", caption = "The range of optimized hyper-parameters for each of the classification algorithms", booktabs = T) %>%
+kable_styling(latex_options = c("striped", "scale_down"), position = "left")%>%
+ column_spec(2:3, width="5em")
+
 ```
@@
 
-**Figure 1.
-Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B).
-Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).
+**Figure 1. Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances.
+```{r}
+```
 
+```{r}
+Table1
+```
 \newpage

### 2018-11-07 submission/manuscript.tex
table for model parameters added

--- 2018-10-22 submission/manuscript.tex
@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage[table]{xcolor}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+
 \usepackage{helvet} % Helvetica font
@@
 \maketitle
-
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
 
@@
 
-\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
-16S rRNA gene sequences for different OTU definitions and level of
-sequencing effort.} Rarefaction curves for different OTU definitions of
-Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
-environments in Table 1 for Bacteria (C) and Archaea (D).
+\textbf{Figure 1. Generalization and classification performance of
+modeling methods } AUC values of all cross validation and testing
+performances.
+\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf}
 
-\newpage
+\rowcolors{2}{gray!6}{white}
+
+\begin{table}
+
+\caption{\label{tab:model_parameters}Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{8em}>{\raggedright\arraybackslash}p{9em}r>{\raggedright\arraybackslash}p{5em}l}
+\hiderowcolors
+\toprule
+\textbf{Method} & \textbf{Parameter} & \textbf{Cross Validation} & \textbf{Epoch} & \textbf{Scaler} & \textbf{Sklearn Function}\\
+\midrule
+\showrowcolors
+\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Logistic Regression} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{LogisticRegression}\\
+\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{L1 SVM Linear Kernel} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{Standard} & \multicolumn{1}{l}{LinearSVC}\\
+\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{L2 SVM Linear Kernel} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{Standard} & \multicolumn{1}{l}{LinearSVC}\\
+\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{SVM RBF Kernel} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C, gamma} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{Standard} & \multicolumn{1}{l}{SVC}\\
+\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Decision Tree} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{max\_depth, min\_samples\_split} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{DecisionTreeClassifier}\\
+\addlinespace
+\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Random Forest} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{n\_estimators, max\_features} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{RandomForestClassifier}\\
+\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{XGBoost} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{n\_estimators, colsample\_bytree, learning\_rate, subsample, max\_depth, min\_child\_weight} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{XGBClassifier}\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\rowcolors{2}{white}{white} \newpage
 

### 2018-11-08 submission/manuscript.Rmd
tables added

@@
 
-```{r model_parameters, echo=FALSE, results='asis'}
+```{r table_1_models, echo=FALSE, results='asis'}
+# We load a model parameters file to make our Table 1.
+Model parameters file has the information of optimized hyper-parameters, pre-processing and cross-validation methods and software implementation for all the classification algorithms.
+
+# This file is hand-made. Therefore needs to be changed manually if new information wants to be added or information needs to be changed.
+
 model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t') %>% 
@@
 
-Table1 <- kable(model_param, "latex", caption = "Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms", booktabs = T) %>%
+             ### Table 1 generation ###
+
+Table1 <- kable(model_param, "latex", caption = "Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.", booktabs = T, linesep = "") %>%
 kable_styling(latex_options = c("striped", "scale_down"), position = "left") %>%  
@@
   row_spec(0,bold=TRUE)
+```
 
-param_range <- read.table('../data/process/model_parameters.txt', header=T, sep='\t', na.strings="", stringsAsFactors = FALSE) %>% 
+```{r table_2_3_params echo=FALSE, results='asis'}
+# We load a param_grid.scv file to make our Table 2 and 3. This file has the information of the ranges of the optimized hyper-parameters for each model.
+
+# This file is generated from model_selection.py that is used to define model parameters for main.py. Therefore any changes to those files would allow us to represent them in the tables automatically.
+
+################ Non-tree based model parameters ##############
+linear_param_range <- read.table('../data/process/param_grid.csv', header=T, sep=',', na.strings="", stringsAsFactors = FALSE) %>% 
   replace(is.na(.), "-")  %>% 
-  select(-Value, -Parameter, -Cross.Validation, -Epochs, -Scaler, -Sklearn.Function)
+  select(1:17) %>% 
+   slice(1:2)
 
-Table2 <- kable(param_range, "latex", caption = "The range of optimized hyper-parameters for each of the classification algorithms", booktabs = T) %>%
-kable_styling(latex_options = c("striped", "scale_down"), position = "left")%>%
- column_spec(2:3, width="5em")
+              ## Table 2 generation ##
+Table2 <- kable(linear_param_range, "latex", caption = "The range of optimized hyper-parameters for logistic regression and support vector machines.", booktabs = T, col.names = NULL, linesep = "") %>%
+kable_styling(latex_options = c("striped", "scale_down"), position = "left") %>% 
+  add_header_above(c("Parameter", "L2 Logistic" = 4, "L1 SVM Linear" = 4, "L2 SVM Linear " = 3, "SVM RBF"=5), bold=TRUE) %>% 
+  column_spec(2, border_left = T) %>%
+  column_spec(6, border_left = T) %>% 
+  column_spec(10, border_left = T) %>% 
+  column_spec(13, border_left = T)
+################ Tree based model parameters ##############
+tree_param_range <- read.table('../data/process/param_grid.csv', header=T, sep=',', na.strings="", stringsAsFactors = FALSE) %>% 
+  replace(is.na(.), "-")  %>% 
+  select(1, 18:29) %>% 
+  slice(3:9)
 
+                  ## Table 3 generation ##
+Table3 <- kable(tree_param_range, "latex", caption = "The range of optimized hyper-parameters for tree based classification algorithms.", booktabs = T, col.names = NULL, linesep = "") %>%
+kable_styling(latex_options = c("striped", "scale_down"), position = "left", font_size=6) %>% 
+  add_header_above(c("Parameter", "Random Forest"=5, "Decision Tree"= 4, "XGBoost"=3 ), bold=TRUE) %>% 
+  column_spec(2, border_left = T) %>% 
+  column_spec(7, border_left = T)%>% 
+  column_spec(11, border_left = T)
 ```
@@
 
-Insert figure legends with the first sentence in bold, for example:
-
 **Figure 1. Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances.
-```{r}
-```
+\begin{center}
+\includegraphics[width=7in]{../results/figures/AUC_comparison.pdf}
+\end{center}
 
@@
 Table1
+Table2
+Table3
 ```
-\newpage
-
-## References

### 2018-11-08 submission/manuscript.tex
tables added

@@
 
-Insert figure legends with the first sentence in bold, for example:
-
 \textbf{Figure 1. Generalization and classification performance of
@@
 performances.
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf}
+
+\begin{center}
+\includegraphics[width=7in]{../results/figures/AUC_comparison.pdf}
+\end{center}
 
@@
 
-\caption{\label{tab:model_parameters}Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms}
+\caption{\label{tab:model_parameters}Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.}
 \resizebox{\linewidth}{!}{
@@
 \multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Decision Tree} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{max\_depth, min\_samples\_split} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{DecisionTreeClassifier}\\
-\addlinespace
 \multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Random Forest} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{n\_estimators, max\_features} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{RandomForestClassifier}\\
@@
 
-\rowcolors{2}{white}{white} \newpage
+\rowcolors{2}{white}{white}
 
-\subsection{References}\label{references}
+\rowcolors{1}{white}{gray!6}
+
+\begin{table}
+
+\caption{\label{tab:model_parameters}The range of optimized hyper-parameters for logistic regression and support vector machines.}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{l|>{}llll|>{}llll|>{}lll|>{}lllll}
+\hiderowcolors
+\toprule
+\multicolumn{1}{c}{\bfseries Parameter} & \multicolumn{4}{c}{\bfseries L2 Logistic} & \multicolumn{4}{c}{\bfseries L1 SVM Linear} & \multicolumn{3}{c}{\bfseries L2 SVM Linear } & \multicolumn{5}{c}{\bfseries SVM RBF} \\
+\cmidrule(l{2pt}r{2pt}){1-1} \cmidrule(l{2pt}r{2pt}){2-5} \cmidrule(l{2pt}r{2pt}){6-9} \cmidrule(l{2pt}r{2pt}){10-12} \cmidrule(l{2pt}r{2pt}){13-17}
+\showrowcolors
+C & 0.01 & 0.1 & 1 & 10 & 0.001 & 0.01 & 0.1 & 1 & 0.01 & 0.1 & 1 & 1e-06 & 1e-05 & 1e-04 & 0.001 & 0.01\\
+gamma & - & - & - & - & - & - & - & - & - & - & - & 1e-09 & 1e-08 & 1e-07 & - & -\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\rowcolors{2}{white}{white}
+
+\rowcolors{1}{white}{gray!6}
+
+\begin{table}
+
+\caption{\label{tab:model_parameters}The range of optimized hyper-parameters for tree based classification algorithms.}
+\resizebox{\linewidth}{!}{
+\fontsize{6}{8}\selectfont
+\begin{tabular}[t]{l|>{}lllll|>{}llll|>{}lll}
+\hiderowcolors
+\toprule
+\multicolumn{1}{c}{\bfseries Parameter} & \multicolumn{5}{c}{\bfseries Random Forest} & \multicolumn{4}{c}{\bfseries Decision Tree} & \multicolumn{3}{c}{\bfseries XGBoost} \\
+\cmidrule(l{2pt}r{2pt}){1-1} \cmidrule(l{2pt}r{2pt}){2-6} \cmidrule(l{2pt}r{2pt}){7-10} \cmidrule(l{2pt}r{2pt}){11-13}
+\showrowcolors
+learning\_rate & - & - & - & - & - & - & - & - & - & 0.01 & 0.1 & 1\\
+max\_depth & - & - & - & - & - & 6 & 8 & 10 & 50 & 6 & 7 & 8\\
+max\_features & 10 & 80 & 500 & 1000 & 1500 & - & - & - & - & - & - & -\\
+min\_child\_weight & - & - & - & - & - & - & - & - & - & 1 & 2 & 3\\
+min\_samples\_split & - & - & - & - & - & 10 & 25 & 50 & - & - & - & -\\
+n\_estimators & 1000 & - & - & - & - & - & - & - & - & 100 & - & -\\
+subsample & - & - & - & - & - & - & - & - & - & 0.7 & 0.8 & 0.9\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\rowcolors{2}{white}{white}
 

### 2018-11-08 submission/manuscript.Rmd
legend size bigger

@@
 
-```{r table_2_3_params echo=FALSE, results='asis'}
+```{r table_2_3_params, echo=FALSE, results='asis'}
 # We load a param_grid.scv file to make our Table 2 and 3. This file has the information of the ranges of the optimized hyper-parameters for each model.
@@
   replace(is.na(.), "-")  %>% 
+  ## only select Logistic Regression and SVM model columns
   select(1:17) %>% 
+  ## only select C (penalty) and gamma rows as parameters
    slice(1:2)
@@
 Table2 <- kable(linear_param_range, "latex", caption = "The range of optimized hyper-parameters for logistic regression and support vector machines.", booktabs = T, col.names = NULL, linesep = "") %>%
+  # latex options make table to be condensed
+  # table itseld if left aligned
   kable_styling(latex_options = c("striped", "scale_down"), position = "left") %>% 
+  # the column names are removed when the file is read so I put headers above the grouped/replicated modeling method for
   add_header_above(c("Parameter", "L2 Logistic" = 4, "L1 SVM Linear" = 4, "L2 SVM Linear " = 3, "SVM RBF"=5), bold=TRUE) %>% 
+  # Put a left border to seperate above header lines
   column_spec(2, border_left = T) %>%
@@
   column_spec(13, border_left = T)
+
 ################ Tree based model parameters ##############
@@
   replace(is.na(.), "-")  %>% 
+  ## only select Logistic Regression and SVM model columns
   select(1, 18:29) %>% 
+  ## select all rows except C (penalty) and gamma for parameters
   slice(3:9)

### 2018-11-08 submission/manuscript.tex
legend size bigger

@@
 
-\caption{\label{tab:model_parameters}Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.}
+\caption{\label{tab:table_1_models}Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.}
 \resizebox{\linewidth}{!}{
@@
 
-\caption{\label{tab:model_parameters}The range of optimized hyper-parameters for logistic regression and support vector machines.}
+\caption{\label{tab:table_2_3_params}The range of optimized hyper-parameters for logistic regression and support vector machines.}
 \resizebox{\linewidth}{!}{
@@
 
-\caption{\label{tab:model_parameters}The range of optimized hyper-parameters for tree based classification algorithms.}
+\caption{\label{tab:table_2_3_params}The range of optimized hyper-parameters for tree based classification algorithms.}
 \resizebox{\linewidth}{!}{

### 2018-11-08 code/learning/compareAUC.R

@@ -89,7 +89,9 @@ ggplot(all, aes(x = fct_reorder(model, AUC, fun = median, .asc =TRUE), y = AUC, 
                      limits=c(0.3, 1), expand=c(0,0)) +
   scale_x_discrete(name = "") +
   theme_bw() +
-  theme(panel.grid.major = element_blank(), 
+  theme(legend.text=element_text(size=18),
+        legend.title=element_text(size=22),
+        panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         text = element_text(size = 12),

### 2018-11-08 submission/manuscript.Rmd
comments added

@@
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
-# Read in necessary libraries
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
 deps = c("doParallel","pROC", "caret","knitr","rmarkdown","vegan","gtools", "tidyverse");
@@
 }
+######################################################################
+#-------------- Define the chunk options ----------------#
+######################################################################
 opts_chunk$set("tidy" = TRUE)
@@
 }
+######################################################################
+#-------------- Define knitr options to allow kable ----------------#
+######################################################################
 knitr::knit_hooks$set(inline=inline_hook)
@@
 ```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
+######################################################################
+#----------------- Load OTU table and MetaData -----------------#
+######################################################################
+
 # Read in metadata and select only sample Id and diagnosis columns
@@
 
-# Read in AUCs table generated from l2 logistic regression model for all samples
-  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+######################################################################
+#------------ Load .tsv data generated in Python -----------------#
+######################################################################
+
+# Read in AUCs table generated from L2 logistic regression model for all samples:
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
 #       FIT is a feature
@@
 # Read in AUCs table generated from l1 SVM linear kernel for all samples
-  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
 #       FIT is a feature
-  # The scaler is StandardScaler
+#       The scaler is Standard scaler
 l1svm <- read.delim('../data/process/L1_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
@@
 # Read in AUCs table generated from l2 SVM linear kernel for all samples
-  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
 #       FIT is a feature
-  # The scaler is StandardScaler
+#       The scaler is Standard scaler
 l2svm <- read.delim('../data/process/L2_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
@@
 # Read in AUCs table generated from  SVM RBF kernel for all samples
-  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
 #       FIT is a feature
-  # The scaler is StandardScaler
+#       The scaler is Standard scaler
 svmRBF <- read.delim('../data/process/SVM_RBF.tsv', header=T, sep='\t') %>%
@@
 # Read in AUCs table generated from xgboost for all samples
-  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
 #       FIT is a feature
-  # The scaler is MinMaxScaler
+#       The scaler is MinMax (0-1) scaler
 xgboost <- read.delim('../data/process/XGBoost.tsv', header=T, sep='\t') %>%
@@
 # Read in AUCs table generated from random forest for all samples
-  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
 #       FIT is a feature
-  # The scaler is MinMaxScaler
+#       The scaler is MinMax (0-1) scaler
 rf <- read.delim('../data/process/Random_Forest.tsv', header=T, sep='\t') %>%
@@
 # Read in AUCs table generated from decision tree for all samples
-  # Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
 #       FIT is a feature
-  # The scaler is MinMaxScaler
+#       The scaler is MinMax (0-1) scaler
 dt <- read.delim('../data/process/Decision_Tree.tsv', header=T, sep='\t') %>%
@@
 
-# Put  all the AUCs for all the models together and group by model
+######################################################################
+#------------ Put all the loaded AUC tables together -----------------#
+######################################################################
 all <- bind_rows(logit, l1svm, l2svm, svmRBF, xgboost, rf, dt) %>% 
@@
 ```{r table_1_models, echo=FALSE, results='asis'}
+######################################################################
 # We load a model parameters file to make our Table 1.
@@
 # This file is hand-made. Therefore needs to be changed manually if new information wants to be added or information needs to be changed. 
-
+######################################################################
 model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t') %>% 
@@
 
-             ### Table 1 generation ###
+#------------------------Table 1 generation------------------------#
 
@@
 ```{r table_2_3_params, echo=FALSE, results='asis'}
+######################################################################
 # We load a param_grid.scv file to make our Table 2 and 3. This file has the information of the ranges of the optimized hyper-parameters for each model.
@@
 # This file is generated from model_selection.py that is used to define model parameters for main.py. Therefore any changes to those files would allow us to represent them in the tables automatically.
+
+# 1) First we will generate a Table for non-tree based models that shows C(penalty) and gamma hyperparameters.
+
+# 2) Second we will generate a Table for tree based models that show all the hyperparameters used for it. 
+######################################################################
 
@@
 
-              ## Table 2 generation ##
+#------------------------Table 2 generation------------------------#
 Table2 <- kable(linear_param_range, "latex", caption = "The range of optimized hyper-parameters for logistic regression and support vector machines.", booktabs = T, col.names = NULL, linesep = "") %>%
@@
 
-                  ## Table 3 generation ##
+#------------------------Table 3 generation------------------------#
 Table3 <- kable(tree_param_range, "latex", caption = "The range of optimized hyper-parameters for tree based classification algorithms.", booktabs = T, col.names = NULL, linesep = "") %>%

### 2018-11-08 code/learning/compareAUC.R

@@ -1,11 +1,15 @@
 # Author: Begum Topcuoglu
 # Date: 2018-11-06
 #
-# 
+######################################################################
 # 1. This script loads the .tsv files generated from python code main.py for all the models. The .tsv files have AUC values for cross validation (for each repeat and n-fold) and for testing for each iteration over different splits of the training-testing data. 
 # 2. This script then binds all the models together and makes a boxplot for AUC for all the models. 
+######################################################################
 
 
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
 deps = c("ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
 for (dep in deps){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
@@ -14,74 +18,84 @@ for (dep in deps){
   library(dep, verbose=FALSE, character.only=TRUE)
 }
 
-# Read in AUCs table generated from l2 logistic regression model for all samples
-# Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
-# FIT is a feature
-# The scaler is MinMax scaler
+######################################################################
+#------------ Load .tsv data generated in Python -----------------#
+######################################################################
+
+# Read in AUCs table generated from L2 logistic regression model for all samples:
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
+#       FIT is a feature
+#       The scaler is MinMax scaler
 logit <- read.delim('data/process/L2_Logistic_Regression.tsv', header=T, sep='\t') %>%
   select(level_1, AUC) %>% 
   rename(Performance = level_1) %>% 
   mutate(model="L2-Logistic Regression")
 
 # Read in AUCs table generated from l1 SVM linear kernel for all samples
-# Carcinomas + Adenomas are 1 and Normal is 0 for binary predictive model
-# FIT is a feature
-# The scaler is StandardScaler
+#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
+#       FIT is a feature
+#       The scaler is Standard scaler
 l1svm <- read.delim('data/process/L1_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
   select(level_1, AUC) %>% 
[…]

### 2018-11-08 submission/manuscript.Rmd
shape the code better
title added

@@
 ---
-title: "**NAME OF THIS STUDY**"
+title: "**Evalution of binary classification pipelines and methods for 16S rRNA gene data**"
 bibliography: references.bib
@@
 
-Table1 <- kable(model_param, "latex", caption = "Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.", booktabs = T, linesep = "") %>%
-kable_styling(latex_options = c("striped", "scale_down"), position = "left") %>%  
+Table1 <- kable(model_param, 
+                "latex", 
+                caption = "Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.", 
+                booktabs = T, 
+                linesep = "") %>%
+kable_styling(latex_options = c("striped", "scale_down"), 
+              position = "left") %>%  
  column_spec(1, width="10em") %>%
@@
 #------------------------Table 2 generation------------------------#
-Table2 <- kable(linear_param_range, "latex", caption = "The range of optimized hyper-parameters for logistic regression and support vector machines.", booktabs = T, col.names = NULL, linesep = "") %>%
+Table2 <- kable(linear_param_range, 
+                "latex", 
+                caption = "The range of optimized hyper-parameters for logistic regression and support vector machines.", 
+                booktabs = T, 
+                col.names = NULL, 
+                linesep = "") %>%
   # latex options make table to be condensed
   # table itseld if left aligned
-  kable_styling(latex_options = c("striped", "scale_down"), position = "left") %>% 
+  kable_styling(latex_options = c("striped", "scale_down"), 
+                position = "left") %>% 
   # the column names are removed when the file is read so I put headers above the grouped/replicated modeling method for
@@
 #------------------------Table 3 generation------------------------#
-Table3 <- kable(tree_param_range, "latex", caption = "The range of optimized hyper-parameters for tree based classification algorithms.", booktabs = T, col.names = NULL, linesep = "") %>%
-kable_styling(latex_options = c("striped", "scale_down"), position = "left", font_size=6) %>% 
+Table3 <- kable(tree_param_range, 
+                "latex", 
+                caption = "The range of optimized hyper-parameters for tree based classification algorithms.", 
+                booktabs = T, 
+                col.names = NULL, 
+                linesep = "") %>%
+kable_styling(latex_options = c("striped", "scale_down"), 
+              position = "left", font_size=6) %>% 
   add_header_above(c("Parameter", "Random Forest"=5, "Decision Tree"= 4, "XGBoost"=3 ), bold=TRUE) %>% 

### 2018-11-09 results/tables/Table1.Rmd
tables generated as seperate pdf

---
title: ""
output:
  pdf_document:
    keep_tex: false
    includes:
      in_header: header.tex
geometry: margin=1.0in
---
```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
######################################################################
#----------------- Read in necessary libraries -------------------#
######################################################################
deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
for (dep in deps){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}

######################################################################
#-------------- Define knitr options to allow kable ----------------#
######################################################################
options(kableExtra.latex.load_packages = FALSE)
options(kableExtra.auto_format = FALSE)
options(knitr.table.format = "latex")
library(kableExtra)
```

```{r table_1_models, echo=FALSE, results='asis'}
######################################################################
# We load a model parameters file to make our Table 1. Model parameters file has the information of optimized hyper-parameters, pre-processing and cross-validation methods and software implementation for all the classification algorithms.

# This file is hand-made. Therefore needs to be changed manually if new information wants to be added or information needs to be changed. 
######################################################################
model_param <- read.table('../../data/process/model_parameters.txt', header=T, sep='\t') %>% 
   select(Method, Parameter, Cross.Validation, Epochs, Scaler, Sklearn.Function) %>% 
  rename("Inner CV"="Cross.Validation") %>%
  rename("Outer CV"="Epochs") %>%
  rename("Function"="Sklearn.Function") %>% 
  rename("Scaling"="Scaler") %>% 
  rename("Hyper-Parameter"="Parameter")

#------------------------Table 1 generation------------------------#

Table1 <- kable(model_param, 
                "latex", 
                caption = "Table 1. Optimized hyper-parameters, pre-processing and inner and outer cross-validation (CV) methods and software implementation of the classification algorithms.", 
                booktabs = T, 
                linesep = "") %>%
kable_styling(latex_options = c("striped", "scale_down", "hold_position"), 
              position = "left") %>%  
 column_spec(1, width="10em") %>%
  column_spec(2, width="8em") %>% 
  column_spec(3, width="9em") %>%
  column_spec(5, width="5em") %>% 
  row_spec(1:7, align="l") %>% 
  row_spec(0,bold=TRUE)

Table1
```

### 2018-11-09 results/tables/Table2.Rmd
tables generated as seperate pdf

---
title: ""
output:
  pdf_document:
    keep_tex: false
    includes:
      in_header: header.tex
geometry: margin=1.0in
---
```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
######################################################################
#----------------- Read in necessary libraries -------------------#
######################################################################
deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
for (dep in deps){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}

######################################################################
#-------------- Define knitr options to allow kable ----------------#
######################################################################
options(kableExtra.latex.load_packages = FALSE)
options(kableExtra.auto_format = FALSE)
options(knitr.table.format = "latex")
library(kableExtra)
```

```{r table_2, echo=FALSE, results='asis'}
######################################################################
# We load a param_grid.scv file to make our Table 2 and 3. This file has the information of the ranges of the optimized hyper-parameters for each model.

# This file is generated from model_selection.py that is used to define model parameters for main.py. Therefore any changes to those files would allow us to represent them in the tables automatically.

# We will generate a Table for non-tree based models that shows C(penalty) and gamma hyperparameters.
######################################################################

################ Non-tree based model parameters ##############
linear_param_range <- read.table('../../data/process/param_grid.csv', header=T, sep=',', na.strings="", stringsAsFactors = FALSE) %>% 
  replace(is.na(.), "-")  %>% 
  ## only select Logistic Regression and SVM model columns
  select(1:17) %>% 
  ## only select C (penalty) and gamma rows as parameters
   slice(1:2)

#------------------------Table 2 generation------------------------#
Table2 <- kable(linear_param_range, 
                "latex", 
                caption = "Table 2. The range of optimized hyper-parameters for logistic regression and support vector machines.", 
                booktabs = T, 
                col.names = NULL, 
                linesep = "") %>%
  # latex options make table to be condensed
  # table itseld if left aligned
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"), 
                position = "left") %>% 
  # the column names are removed when the file is read so I put headers above the grouped/replicated modeling method for
  add_header_above(c("Parameter", "L2 Logistic" = 4, "L1 SVM Linear" = 4, "L2 SVM Linear " = 3, "SVM RBF"=5), bold=TRUE) %>% 
  # Put a left border to seperate above header lines
  column_spec(2, border_left = T) %>%
  column_spec(6, border_left = T) %>% 
  column_spec(10, border_left = T) %>% 
  column_spec(13, border_left = T)

Table2
```

### 2018-11-09 results/tables/Table3.Rmd
tables generated as seperate pdf

--- 2018-11-09 results/tables/Table2.Rmd
@@
 
-```{r table_2, echo=FALSE, results='asis'}
+```{r table_3, echo=FALSE, results='asis'}
 ######################################################################
@@
 
-# We will generate a Table for non-tree based models that shows C(penalty) and gamma hyperparameters.
+# We will generate a Table for tree based models that show all the hyperparameters used for it. 
 ######################################################################
 
-################ Non-tree based model parameters ##############
-linear_param_range <- read.table('../../data/process/param_grid.csv', header=T, sep=',', na.strings="", stringsAsFactors = FALSE) %>% 
+################ Tree based model parameters ##############
+tree_param_range <- read.table('../../data/process/param_grid.csv', header=T, sep=',', na.strings="", stringsAsFactors = FALSE) %>% 
   replace(is.na(.), "-")  %>% 
   ## only select Logistic Regression and SVM model columns
-  select(1:17) %>% 
-  ## only select C (penalty) and gamma rows as parameters
-   slice(1:2)
+  select(1, 18:29) %>% 
+  ## select all rows except C (penalty) and gamma for parameters
+  slice(3:9)
 
-#------------------------Table 2 generation------------------------#
-Table2 <- kable(linear_param_range, 
+#------------------------Table 3 generation------------------------#
+Table3 <- kable(tree_param_range, 
                 "latex", 
-                caption = "Table 2. The range of optimized hyper-parameters for logistic regression and support vector machines.", 
+                caption = "Table 3. The range of optimized hyper-parameters for tree based classification algorithms.", 
                 booktabs = T, 
@@
                 linesep = "") %>%
-  # latex options make table to be condensed
-  # table itseld if left aligned
 kable_styling(latex_options = c("striped", "scale_down", "hold_position"), 
-                position = "left") %>% 
-  # the column names are removed when the file is read so I put headers above the grouped/replicated modeling method for
-  add_header_above(c("Parameter", "L2 Logistic" = 4, "L1 SVM Linear" = 4, "L2 SVM Linear " = 3, "SVM RBF"=5), bold=TRUE) %>% 
-  # Put a left border to seperate above header lines
+              position = "left", font_size=6) %>% 
+  add_header_above(c("Parameter", "Random Forest"=5, "Decision Tree"= 4, "XGBoost"=3 ), bold=TRUE) %>% 
   column_spec(2, border_left = T) %>% 
-  column_spec(6, border_left = T) %>% 
-  column_spec(10, border_left = T) %>% 
-  column_spec(13, border_left = T)
+  column_spec(7, border_left = T)%>% 
+  column_spec(11, border_left = T)
 
-Table2
+Table3
 ```

### 2018-11-09 submission/manuscript.Rmd
running title
tables generated as seperate pdf

@@
 
-```{r table_1_models, echo=FALSE, results='asis'}
-######################################################################
-# We load a model parameters file to make our Table 1.
-Model parameters file has the information of optimized hyper-parameters, pre-processing and cross-validation methods and software implementation for all the classification algorithms.
-
-# This file is hand-made. Therefore needs to be changed manually if new information wants to be added or information needs to be changed. 
-######################################################################
-model_param <- read.table('../data/process/model_parameters.txt', header=T, sep='\t') %>% 
-   select(Method, Parameter, Cross.Validation, Epochs, Scaler, Sklearn.Function) %>% 
-  rename("Cross Validation"="Cross.Validation") %>%
-  rename("Epoch"="Epochs") %>%
-  rename("Sklearn Function"="Sklearn.Function")
-
-#------------------------Table 1 generation------------------------#
-
-Table1 <- kable(model_param, 
-                "latex", 
-                caption = "Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.", 
-                booktabs = T, 
-                linesep = "") %>%
-kable_styling(latex_options = c("striped", "scale_down"), 
-              position = "left") %>%  
- column_spec(1, width="10em") %>%
-  column_spec(2, width="8em") %>% 
-  column_spec(3, width="9em") %>%
-  column_spec(5, width="5em") %>% 
-  row_spec(1:7, align="l") %>% 
-  row_spec(0,bold=TRUE)
-```
-
-```{r table_2_3_params, echo=FALSE, results='asis'}
-######################################################################
-# We load a param_grid.scv file to make our Table 2 and 3. This file has the information of the ranges of the optimized hyper-parameters for each model.
-
-# This file is generated from model_selection.py that is used to define model parameters for main.py. Therefore any changes to those files would allow us to represent them in the tables automatically.
-
-# 1) First we will generate a Table for non-tree based models that shows C(penalty) and gamma hyperparameters.
-
-# 2) Second we will generate a Table for tree based models that show all the hyperparameters used for it. 
-######################################################################
-
-################ Non-tree based model parameters ##############
-linear_param_range <- read.table('../data/process/param_grid.csv', header=T, sep=',', na.strings="", stringsAsFactors = FALSE) %>% 
-  replace(is.na(.), "-")  %>% 
-  ## only select Logistic Regression and SVM model columns
-  select(1:17) %>% 
-  ## only select C (penalty) and gamma rows as parameters
-   slice(1:2)
-
-#------------------------Table 2 generation------------------------#
-Table2 <- kable(linear_param_range, 
-                "latex", 
-                caption = "The range of optimized hyper-parameters for logistic regression and support vector machines.", 
-                booktabs = T, 
-                col.names = NULL, 
-                linesep = "") %>%
-  # latex options make table to be condensed
-  # table itseld if left aligned
-  kable_styling(latex_options = c("striped", "scale_down"), 
-                position = "left") %>% 
-  # the column names are removed when the file is read so I put headers above the grouped/replicated modeling method for
-  add_header_above(c("Parameter", "L2 Logistic" = 4, "L1 SVM Linear" = 4, "L2 SVM Linear " = 3, "SVM RBF"=5), bold=TRUE) %>% 
-  # Put a left border to seperate above header lines
-  column_spec(2, border_left = T) %>%
-  column_spec(6, border_left = T) %>% 
-  column_spec(10, border_left = T) %>% 
-  column_spec(13, border_left = T)
-
-################ Tree based model parameters ##############
-tree_param_range <- read.table('../data/process/param_grid.csv', header=T, sep=',', na.strings="", stringsAsFactors = FALSE) %>% 
-  replace(is.na(.), "-")  %>% 
-  ## only select Logistic Regression and SVM model columns
-  select(1, 18:29) %>% 
-  ## select all rows except C (penalty) and gamma for parameters
-  slice(3:9)
-
-#------------------------Table 3 generation------------------------#
-Table3 <- kable(tree_param_range, 
-                "latex", 
-                caption = "The range of optimized hyper-parameters for tree based classification algorithms.", 
-                booktabs = T, 
-                col.names = NULL, 
-                linesep = "") %>%
-kable_styling(latex_options = c("striped", "scale_down"), 
-              position = "left", font_size=6) %>% 
-  add_header_above(c("Parameter", "Random Forest"=5, "Decision Tree"= 4, "XGBoost"=3 ), bold=TRUE) %>% 
-  column_spec(2, border_left = T) %>% 
-  column_spec(7, border_left = T)%>% 
-  column_spec(11, border_left = T)
-```
 \vspace{35mm}
 
-Running title: INSERT RUNNING TITLE HERE
+Running title: Machine learning methods in microbiome studies
 

### 2018-11-09 submission/manuscript.tex
tables generated as seperate pdf

@@
 \hypersetup{unicode=true,
-            pdftitle={NAME OF THIS STUDY},
+            pdftitle={Evalution of binary classification pipelines and methods for 16S rRNA gene data},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{NAME OF THIS STUDY}}
+  \title{\textbf{Evalution of binary classification pipelines and methods for 16S
+rRNA gene data}}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-Running title: INSERT RUNNING TITLE HERE
+Running title: Machine learning methods in microbiome studies
 

### 2018-11-09 results/tables/Table1.Rmd
comment fix

@@
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
-######################################################################
-#----------------- Read in necessary libraries -------------------#
-######################################################################
-deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
+####################################################################################################
+#------------------------------- Read in necessary libraries --------------------------------------#
+####################################################################################################deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
 for (dep in deps){
@@
 
-######################################################################
-#-------------- Define knitr options to allow kable ----------------#
-######################################################################
-options(kableExtra.latex.load_packages = FALSE)
+#####################################################################################################------------------------------- Define knitr options to allow kable ------------------------------#
+####################################################################################################options(kableExtra.latex.load_packages = FALSE)
 options(kableExtra.auto_format = FALSE)
@@
 ```{r table_1_models, echo=FALSE, results='asis'}
-######################################################################
-# We load a model parameters file to make our Table 1.
+##################################################################################################### We load a model parameters file to make our Table 1.
 Model parameters file has the information of optimized hyper-parameters, pre-processing and cross-validation methods and software implementation for all the classification algorithms.
@@
 # This file is hand-made. Therefore needs to be changed manually if new information wants to be added or information needs to be changed.
-######################################################################
-model_param <- read.table('../../data/process/model_parameters.txt', header=T, sep='\t') %>% 
+####################################################################################################model_param <- read.table('../../data/process/model_parameters.txt', header=T, sep='\t') %>%
    select(Method, Parameter, Cross.Validation, Epochs, Scaler, Sklearn.Function) %>%
@@
 
-#------------------------Table 1 generation------------------------#
+#--------------------------------------- Table 1 generation ---------------------------------------#
 

### 2018-11-09 results/tables/Table1.Rmd
fix comment

@@
 ---
+
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
+
 ####################################################################################################
@@
 ```{r table_1_models, echo=FALSE, results='asis'}
+
 ##################################################################################################### We load a model parameters file to make our Table 1.
@@
 # This file is hand-made. Therefore needs to be changed manually if new information wants to be added or information needs to be changed.
-####################################################################################################model_param <- read.table('../../data/process/model_parameters.txt', header=T, sep='\t') %>%
+##################################################################################################
+
+model_param <- read.table('../../data/process/model_parameters.txt', header=T, sep='\t') %>%
   select(Method, Parameter, Cross.Validation, Epochs, Scaler, Sklearn.Function) %>%

### 2018-11-14 results/tables/Table1.Rmd
fix commented code

@@
 #------------------------------- Read in necessary libraries --------------------------------------#
-####################################################################################################deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
+####################################################################################################
+deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
 for (dep in deps){
@@
 
-#####################################################################################################------------------------------- Define knitr options to allow kable ------------------------------#
-####################################################################################################options(kableExtra.latex.load_packages = FALSE)
+#####################################################################################################
+#------------------------------- Define knitr options to allow kable ------------------------------#
+####################################################################################################
+options(kableExtra.latex.load_packages = FALSE)
 options(kableExtra.auto_format = FALSE)
@@
 
-##################################################################################################### We load a model parameters file to make our Table 1.
+#####################################################################################################
+
+# We load a model parameters file to make our Table 1.
 Model parameters file has the information of optimized hyper-parameters, pre-processing and cross-validation methods and software implementation for all the classification algorithms.

### 2018-11-14 code/learning/compareHP.R
initial commit

@@ -1,0 +1,133 @@
+# Author: Begum Topcuoglu
+# Date: 2018-11-14
+#
+######################################################################
+# 1. This script loads the *_parameters.tsv files generated from python code main.py for all the models. The .tsv files have mean AUC values of all the hyper-parameters tested on the training data for the 100 times that outer-validation runs. 
+# 2. It also plots the hyper-parameter range and performance during cross validation. 
+######################################################################
+
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
+deps = c("cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+######################################################################
+#------------ Load .tsv data generated in Python -----------------#
+######################################################################
+
+# Read in hyper-parameter AUCs generated from L2 logistic regression model for all samples:
+logit <- read.delim('data/process/L2_Logistic_Regression_parameters.tsv', header=T, sep='\t')%>%
+  select(-X) %>% 
+  mutate(meanAUC=rowMeans(.[, 2:101])) %>% 
+  select(C, meanAUC) 
+
+logit_plot <- ggplot(logit,aes(x=C,y=meanAUC)) + 
+  geom_line() + 
+  geom_point() +
+  scale_y_continuous(name="Logistic Regression cvAUC", 
+                     limits = c(0.50, 0.80), 
+                     breaks = seq(0.5, 0.8, 0.05)) + 
+  theme_bw() +
+  theme(legend.text=element_text(size=18),
+        legend.title=element_text(size=22),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        panel.background = element_blank(),
+        text = element_text(size = 12),
+        axis.text.x=element_text(size = 12, colour='black'),
+        axis.text.y=element_text(size = 12, colour='black'),
[…]

### 2018-11-14 submission/manuscript.Rmd
explain bxplot
clean up

@@
 ######################################################################
-deps = c("doParallel","pROC", "caret","knitr","rmarkdown","vegan","gtools", "tidyverse");
+deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
 for (dep in deps){
@@
 }
-######################################################################
-#-------------- Define knitr options to allow kable ----------------#
-######################################################################
 knitr::knit_hooks$set(inline=inline_hook)
-options(kableExtra.latex.load_packages = FALSE)
-options(kableExtra.auto_format = FALSE)
-options(knitr.table.format = "latex")
-library(kableExtra)
 ```
@@
 
-**Figure 1. Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances.
-\begin{center}
-\includegraphics[width=7in]{../results/figures/AUC_comparison.pdf}
-\end{center}
+**Figure 1.
+Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances.
+The boxplot shows quartiles at the box ends and the statistical median as athe horizontal line in the box.
+The whiskers show the farthest points that are not outliers.
+Outliers are data points that are not within 3/2 times the interquartile ranges.
 

### 2018-11-14 submission/manuscript.Rmd
add cvAUCs for HP

@@
 ######################################################################
-#------------ Load .tsv data generated in Python -----------------#
+#---Load .tsv data with CV and test AUC data generated in Python ----#
 ######################################################################
@@
   group_by(model)
+
+######################################################################
+#---Load .tsv data with cvAUCs for hyper-parameters  ----#
+######################################################################
+# Read in hyper-parameter AUCs generated from L2 logistic regression model for all samples:
+logit <- read.delim('../data/process/L2_Logistic_Regression_parameters.tsv', header=T, sep='\t')%>%
+  select(-X) %>% 
+  mutate(meanAUC=rowMeans(.[, 2:101])) %>% 
+  select(C, meanAUC) 
+# Read in hyper-parameter AUCs generated from L1 Linear SVM model for all samples:
+l1svm <- read.delim('data/process/L1_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
+  select(-X) %>% 
+  mutate(meanAUC=rowMeans(.[, 2:101])) %>% 
+  select(C, meanAUC) 
+# Read in hyper-parameter AUCs generated from L2 Linear SVM model for all samples:
+l2svm <- read.delim('data/process/L2_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
+  select(-X) %>% 
+  mutate(meanAUC=rowMeans(.[, 2:101])) %>% 
+  select(C, meanAUC)
+
 ```

### 2018-11-16 submission/manuscript.Rmd
update introduction

@@
 meta <- read.delim('../data/metadata.tsv', header=T, sep='\t') %>%
-  select(sample, dx)
+  select(sample, dx, Dx_Bin, fit_result)
 
@@
 
-As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning  methods to analyze microbiome sequence read count data and to determine association with a continuous or categorical phenotype of interest.
+As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning  methods to analyze 16S rRNA gene sequence data and to determine association of the microbiome with a continuous or categorical phenotype of interest.
+The use of machine learning in microbiome literature lack clarity over the learning pipeline which spans the problem formulation, feature selection, feature pre-processing, model learning, and output.
+There is a need for guidance on how to properly implement good machine learning practices to generate reproducible, robust and actionable models.
+We also need to generate interpretable models for biomedical researchers to adopt and use regularly.
+In this study we chose to focus on a binary classification model that predicts if an individual has screen-relevant tumors in their colon or not, to establish a model pipeline.
 
 Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Early diagnosis increases the chance of survival.
-However the current diagnostic methods are expensive and invasive.
-As a less invasive tool, numerous studies use relative abundances of the gut bacteria populations to predict disease progression.
-Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small.
-It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features.
-
-ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc.
-There is a need for guidance on how to properly implement these different methods.
-We need to emphasize good machine learning practices and pipelines and discuss the reproducibility, robustness and actionability of models.
-
-We established a non-leaky pipeline.
-We performed L1 and L2-regularized logistic regression, Linear SVM, Non-Linear SVM, Decision tree, Random forest, XGBoost and Feed Forward Neural Net classification models.
-We evaluated the classification performance of different machine learning methods.
-We also want to discuss the reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting of each method.
+Each person in the industrialized world has on average a one-in-twenty chance of developing colorectal cancer (CRC) in their lifetime and once diagnosed, more than one-third will not survive 5 years.
+Colonoscopy as a screening tool is very effective, however it is very invasive, expensive and have a low rate of patient adherence.
+Therefore, there is a need for improved non-invasive methods to screen individuals.
+One proposed non-invasive screening tool is using gut microbiome-based biomarkers.
+Patients with colorectal cancer have different stool community of microbes compared to adults with normal colons.
+This difference however cannot be explained by a single or a handful of features in the gut microbiome but by many of them in relation to one another.
+Therefore, machine learning is a great tool to investigate the differences between the gut microbiomes of CRC patients and healthy individuals.
+Previous studies have shown that human hemoglobin levels and bacterial population abundances in the stool help us predict screen relevant growth in the colon, however the literature for the problem of classification colorectal cancer diagnosis vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.8.
+The variation in classification performance is based in part on differences in the task definition, in
+part on differences in the study populations, and in part on the evaluation method. The highest
+reported AUCs were from studies of ...
+Additionally, some of the reported results were not obtained from testing on held-out sets.
+In this study, we have defined classification pipelines with L2-regularized logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM , decision tree, random forest and XGBoost.
+We evaluated the generalization and prediction performance of these methods.
+We also compared each method based on their reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting.
 
@@
 ## Materials and Methods
+
+#### Data collection
+We obtained stool OTU abundance data and metadata from the Sze et al.
+(1).
+The stool OTU abundance data and metadata were generated from the Great Lakes-New England Early Detection Research Network study which collected stool samples from recruited participants who were undergoing scheduled screening or surveillance colonoscpy.
+Colonoscopies were performed and fecal samples were collected from adult participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
+Patients' colonic disease status was defined by colonoscopy with adequate preparation and tissue histopathology of all resected lesions.
+Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma.
+Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies , `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas.
+Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas.
+Fecal material provided by the patients was used for Fecal Immunological Tests (FIT) which measure human hemoglobin concentrations and for 16S rRNA gene sequencing to measure bacterial population abundances.
+The OTU abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm.
+
+#### Data definitions and pre-processing
+We use FIT results and OTU abundances as features. However these are in different scales. We use MinMaxScaler to transforms features by scaling each feature to a [0-1] range.
 

### 2018-11-16 submission/manuscript.tex
update introduction

@@
 ever-increasing demand for reproducible machine learning methods to
-analyze microbiome sequence read count data and to determine association
-with a continuous or categorical phenotype of interest.
+analyze 16S rRNA gene sequence data and to determine association of the
+microbiome with a continuous or categorical phenotype of interest. The
+use of machine learning in microbiome literature lack clarity over the
+learning pipeline which spans the problem formulation, feature
+selection, feature pre-processing, model learning, and output. There is
+a need for guidance on how to properly implement good machine learning
+practices to generate reproducible, robust and actionable models. We
+also need to generate interpretable models for biomedical researchers to
+adopt and use regularly. In this study we chose to focus on a binary
+classification model that predicts if an individual has screen-relevant
+tumors in their colon or not, to establish a model pipeline.
 
 Colorectal cancer is one of the leading cause of death among cancers in
-the United States. Early diagnosis increases the chance of survival.
-However the current diagnostic methods are expensive and invasive. As a
-less invasive tool, numerous studies use relative abundances of the gut
-bacteria populations to predict disease progression. Most microbial
-communities are pretty patchy and the likelihood of a single feature
-that explains the differences in health is pretty small. It is likely
-that many biomarkers are needed to account for the patchiness as well as
-the context dependency of the features.
-
-ML use in microbiome literature is a bit like the wild west with lack of
-clarity over methods, testing, validation, etc. There is a need for
-guidance on how to properly implement these different methods. We need
-to emphasize good machine learning practices and pipelines and discuss
-the reproducibility, robustness and actionability of models.
-
-We established a non-leaky pipeline. We performed L1 and L2-regularized
-logistic regression, Linear SVM, Non-Linear SVM, Decision tree, Random
-forest, XGBoost and Feed Forward Neural Net classification models. We
-evaluated the classification performance of different machine learning
-methods. We also want to discuss the reproducibility, robustness,
-actionability, interpretibility and susceptibility to overfitting of
-each method.
+the United States. Each person in the industrialized world has on
+average a one-in-twenty chance of developing colorectal cancer (CRC) in
+their lifetime and once diagnosed, more than one-third will not survive
+5 years. Colonoscopy as a screening tool is very effective, however it
+is very invasive, expensive and have a low rate of patient adherence.
+Therefore, there is a need for improved non-invasive methods to screen
+individuals. One proposed non-invasive screening tool is using gut
+microbiome-based biomarkers. Patients with colorectal cancer have
+different stool community of microbes compared to adults with normal
+colons. This difference however cannot be explained by a single or a
+handful of features in the gut microbiome but by many of them in
+relation to one another. Therefore, machine learning is a great tool to
+investigate the differences between the gut microbiomes of CRC patients
+and healthy individuals. Previous studies have shown that human
+hemoglobin levels and bacterial population abundances in the stool help
+us predict screen relevant growth in the colon, however the literature
+for the problem of classification colorectal cancer diagnosis vary
+greatly, with areas under the receiver operating characteristic curve
+(AUC) of 0.7-0.8. The variation in classification performance is based
+in part on differences in the task definition, in part on differences in
+the study populations, and in part on the evaluation method. The highest
+reported AUCs were from studies of \ldots{} Additionally, some of the
+reported results were not obtained from testing on held-out sets. In
+this study, we have defined classification pipelines with L2-regularized
+logistic regression, L1 and L2 linear suppor vector machines (SVM),
+radial basis function SVM , decision tree, random forest and XGBoost. We
+evaluated the generalization and prediction performance of these
+methods. We also compared each method based on their reproducibility,
+robustness, actionability, interpretibility and susceptibility to
+overfitting.
 
@@
 
+\paragraph{The data}\label{the-data}
+
+We obtained stool OTU abundance data and metadata from the Sze et al.
+(1). The stool OTU abundance data and metadata comes from the Great
+Lakes- New England Early Detection Research Network study which
+collected stool samples from eligible individuals. Briefly, eligible
+patients for this study were aged at least 18 years and willing to
+collect stool samples. Colonoscopies were performed and fecal samples
+were collected from participants in four locations: Toronto (ON,
+Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
+Patient diagnoses were determined by colonoscopic examination and
+histopathological review of any biopsies taken. Patients with an adenoma
+greater than 1 cm, more than three adenomas of any size, or an adenoma
+with villous histology were classified as advanced adenoma. Fecal
+material was used for Fecal Immunological Tests and these tests were
+used to measure human hemoglobin concentrations. This study had 172
+control, 198 adenomas and 120 carcinomas. Of the 198 adenomas, 109 were
+advanced adenomas.
+
 \newpage
@@
 modeling methods } AUC values of all cross validation and testing
-performances.
-
-\begin{center}
-\includegraphics[width=7in]{../results/figures/AUC_comparison.pdf}
-\end{center}
-
-\rowcolors{2}{gray!6}{white}
-
-\begin{table}
-
-\caption{\label{tab:table_1_models}Optimized hyper-parameters, pre-processing and cross-validation methods and software implementation of the classification algorithms.}
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{8em}>{\raggedright\arraybackslash}p{9em}r>{\raggedright\arraybackslash}p{5em}l}
-\hiderowcolors
-\toprule
-\textbf{Method} & \textbf{Parameter} & \textbf{Cross Validation} & \textbf{Epoch} & \textbf{Scaler} & \textbf{Sklearn Function}\\
-\midrule
-\showrowcolors
-\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Logistic Regression} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{LogisticRegression}\\
-\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{L1 SVM Linear Kernel} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{Standard} & \multicolumn{1}{l}{LinearSVC}\\
-\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{L2 SVM Linear Kernel} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{Standard} & \multicolumn{1}{l}{LinearSVC}\\
-\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{SVM RBF Kernel} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{C, gamma} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{Standard} & \multicolumn{1}{l}{SVC}\\
-\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Decision Tree} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{max\_depth, min\_samples\_split} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{DecisionTreeClassifier}\\
-\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{Random Forest} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{n\_estimators, max\_features} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{RandomForestClassifier}\\
-\multicolumn{1}{>{\raggedright\arraybackslash}p{10em}}{XGBoost} & \multicolumn{1}{>{\raggedright\arraybackslash}p{8em}}{n\_estimators, colsample\_bytree, learning\_rate, subsample, max\_depth, min\_child\_weight} & \multicolumn{1}{>{\raggedright\arraybackslash}p{9em}}{5-fold, 100-repeats} & \multicolumn{1}{l}{100} & \multicolumn{1}{>{\raggedright\arraybackslash}p{5em}}{MinMax} & \multicolumn{1}{l}{XGBClassifier}\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\rowcolors{2}{white}{white}
-
-\rowcolors{1}{white}{gray!6}
-
-\begin{table}
-
-\caption{\label{tab:table_2_3_params}The range of optimized hyper-parameters for logistic regression and support vector machines.}
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{l|>{}llll|>{}llll|>{}lll|>{}lllll}
-\hiderowcolors
-\toprule
-\multicolumn{1}{c}{\bfseries Parameter} & \multicolumn{4}{c}{\bfseries L2 Logistic} & \multicolumn{4}{c}{\bfseries L1 SVM Linear} & \multicolumn{3}{c}{\bfseries L2 SVM Linear } & \multicolumn{5}{c}{\bfseries SVM RBF} \\
-\cmidrule(l{2pt}r{2pt}){1-1} \cmidrule(l{2pt}r{2pt}){2-5} \cmidrule(l{2pt}r{2pt}){6-9} \cmidrule(l{2pt}r{2pt}){10-12} \cmidrule(l{2pt}r{2pt}){13-17}
-\showrowcolors
-C & 0.01 & 0.1 & 1 & 10 & 0.001 & 0.01 & 0.1 & 1 & 0.01 & 0.1 & 1 & 1e-06 & 1e-05 & 1e-04 & 0.001 & 0.01\\
-gamma & - & - & - & - & - & - & - & - & - & - & - & 1e-09 & 1e-08 & 1e-07 & - & -\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\rowcolors{2}{white}{white}
-
-\rowcolors{1}{white}{gray!6}
-
-\begin{table}
-
-\caption{\label{tab:table_2_3_params}The range of optimized hyper-parameters for tree based classification algorithms.}
-\resizebox{\linewidth}{!}{
-\fontsize{6}{8}\selectfont
-\begin{tabular}[t]{l|>{}lllll|>{}llll|>{}lll}
-\hiderowcolors
-\toprule
-\multicolumn{1}{c}{\bfseries Parameter} & \multicolumn{5}{c}{\bfseries Random Forest} & \multicolumn{4}{c}{\bfseries Decision Tree} & \multicolumn{3}{c}{\bfseries XGBoost} \\
-\cmidrule(l{2pt}r{2pt}){1-1} \cmidrule(l{2pt}r{2pt}){2-6} \cmidrule(l{2pt}r{2pt}){7-10} \cmidrule(l{2pt}r{2pt}){11-13}
-\showrowcolors
-learning\_rate & - & - & - & - & - & - & - & - & - & 0.01 & 0.1 & 1\\
-max\_depth & - & - & - & - & - & 6 & 8 & 10 & 50 & 6 & 7 & 8\\
-max\_features & 10 & 80 & 500 & 1000 & 1500 & - & - & - & - & - & - & -\\
-min\_child\_weight & - & - & - & - & - & - & - & - & - & 1 & 2 & 3\\
-min\_samples\_split & - & - & - & - & - & 10 & 25 & 50 & - & - & - & -\\
-n\_estimators & 1000 & - & - & - & - & - & - & - & - & 100 & - & -\\
-subsample & - & - & - & - & - & - & - & - & - & 0.7 & 0.8 & 0.9\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\rowcolors{2}{white}{white}
+performances. The boxplot shows quartiles at the box ends and the
+statistical median as athe horizontal line in the box. The whiskers show
+the farthest points that are not outliers. Outliers are data points that
+are not within 3/2 times the interquartile ranges.
 

### 2018-11-19 submission/manuscript.Rmd
add sklearn version
versions

@@
 knitr::knit_hooks$set(inline=inline_hook)
+library(reticulate)
+use_python("/Library/Frameworks/Python.framework/Versions/3.6/bin/python3", required = TRUE)
 ```
@@
 # Read in hyper-parameter AUCs generated from L1 Linear SVM model for all samples:
-l1svm <- read.delim('data/process/L1_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
+l1svm <- read.delim('../data/process/L1_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
   select(-X) %>% 
@@
 # Read in hyper-parameter AUCs generated from L2 Linear SVM model for all samples:
-l2svm <- read.delim('data/process/L2_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
+l2svm <- read.delim('../data/process/L2_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
   select(-X) %>% 
@@
 
+```
+
+```{python, cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
+import platform
+python_version = platform.python_version()
+import sklearn
+sklearn_module = sklearn.__version__
 ```
@@
 reported AUCs were from studies of ...
-Additionally, some of the reported results were not obtained from testing on held-out sets.
 In this study, we have defined classification pipelines with L2-regularized logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM , decision tree, random forest and XGBoost.
@@
 #### Data definitions and pre-processing
-We use FIT results and OTU abundances as features. However these are in different scales. We use MinMaxScaler to transforms features by scaling each feature to a [0-1] range.
+We use FIT results and OTU abundances as features.
+However these are in different scales.
+We use python v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` function MinMaxScaler() to transform features by scaling each feature to a [0-1] range.
 
@@
 Outliers are data points that are not within 3/2 times the interquartile ranges.
-
-```{r}
-Table1
-Table2
-Table3
-```

### 2018-11-20 submission/manuscript.Rmd
update

@@
 ---
-title: "**Evalution of binary classification pipelines and methods for 16S rRNA gene data**"
+title: "**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**"
 bibliography: references.bib
@@
 
-Begüm D. Topçuoğlu^1, Jenna Wiens^2, Patrick D. Schloss^1$\dagger$^
+Begüm D. Topçuoğlu${^1}$, Jenna Wiens${^2}$, Patrick D. Schloss^1$\dagger$^
 
@@
 There is a need for guidance on how to properly implement good machine learning practices to generate reproducible, robust and actionable models.
-We also need to generate interpretable models for biomedical researchers to adopt and use regularly.
-In this study we chose to focus on a binary classification model that predicts if an individual has screen-relevant tumors in their colon or not, to establish a model pipeline.
+There is also a clinical need to generate interpretable models for biomedical researchers and clinicans to adopt and use regularly [@wiens_editorial:_2016].
 
+Recently, there is an interest on using machine learning to predict colorectal cancer progression with microbiota-assiciated biomarkers.
 Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Each person in the industrialized world has on average a one-in-twenty chance of developing colorectal cancer (CRC) in their lifetime and once diagnosed, more than one-third will not survive 5 years.
+Each person in the industrialized world has on average a one-in-twenty chance of developing colorectal cancer (CRC) in their lifetime and once diagnosed, more than one-third will not survive 5 years [@seer_2016; @street_colorectal_nodate; @weir_past_2015].
 Colonoscopy as a screening tool is very effective, however it is very invasive, expensive and have a low rate of patient adherence.
 Therefore, there is a need for improved non-invasive methods to screen individuals.
-One proposed non-invasive screening tool is using gut microbiome-based biomarkers.
+Gut microbiome-based biomarkers can be used as a non-invasive screening method.
+
 Patients with colorectal cancer have different stool community of microbes compared to adults with normal colons.
-This difference however cannot be explained by a single or a handful of features in the gut microbiome but by many of them in relation to one another.
-Therefore, machine learning is a great tool to investigate the differences between the gut microbiomes of CRC patients and healthy individuals.
-Previous studies have shown that human hemoglobin levels and bacterial population abundances in the stool help us predict screen relevant growth in the colon, however the literature for the problem of classification colorectal cancer diagnosis vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.8.
-The variation in classification performance is based in part on differences in the task definition, in
-part on differences in the study populations, and in part on the evaluation method. The highest
-reported AUCs were from studies of ...
-In this study, we have defined classification pipelines with L2-regularized logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM , decision tree, random forest and XGBoost.
-We evaluated the generalization and prediction performance of these methods.
-We also compared each method based on their reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting.
+This difference however cannot be explained by a single or a handful of candidate taxa in the gut microbiome but by many of them in relation to one another.
+Therefore, machine learning emerges as a tool to detect the difference between the gut microbiomes of CRC patients and healthy individuals.
+Previous studies have shown that human hemoglobin levels and bacterial population abundances in the stool help us predict screen relevant growth in the colon, however the literature for the problem of classifying colorectal disease status vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014].
+The variation in classification performance is based in part on differences in the task definition, in part on differences in the study populations, and in part on the learning pipeline.
+In this study, classification pipelines with L2-regularized logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM , decision tree, random forest and XGBoost classifiers are established.
+The generalization and prediction performance of these classifiers are evaluated and each classifier is examined for its reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting.
 
-Generalisation Perfomance of each model. 
-Is there a maximum threshold of prediction with all these methods?
-Does an increase in model complexity improve predictibility?
-Synthesis statement regarding modeling 16S microbiome data
+Here, colonic disease status is defined as Normal or Screen Relevant Neoplasias (SRN).
+Stool bacterial population abundances and stool hemoglobin levels of `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` Normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples were used to learn binary classifiers and evaluate their performances.
+The results show that ......
+(Is there a maximum threshold of prediction with all these methods?
+Does an increase in model complexity improve predictibility?)
 
 ## Results and Discussion
+
+##### Results of modeling in text, tables and figures
+
+##### Comparisons among modeling approaches
+
+#### Interpretation of modeling results in terms of reproducibility, robustness, actionability, interpretibility and susceptibility
+
+#### Consideration of possible weaknesses for each model
+
+The interactions between the biomarkers may be nonlinear.
+Obviously, the linear models will not incorporate this because they are linear.
+Tools like linear models (e.g.
+metastats, lefse, wilcoxon, etc) are likely worthless.
+
+#### Consideration of possible weaknesses for our approach and chosen dataset
+
+#### Relationship of results to previous literature and broader implications of this work
+
+#### Prospects of future progress
 
@@
 #### Data collection
-We obtained stool OTU abundance data and metadata from the Sze et al.
-(1).
-The stool OTU abundance data and metadata were generated from the Great Lakes-New England Early Detection Research Network study which collected stool samples from recruited participants who were undergoing scheduled screening or surveillance colonoscpy.
-Colonoscopies were performed and fecal samples were collected from adult participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
+  The data used for this analysis are stool bacterial abundances, stool hemoglobin levels and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study.
+These data were obtained from Sze et al [@sze_leveraging_2018].
+  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscpy.
+Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
 Patients' colonic disease status was defined by colonoscopy with adequate preparation and tissue histopathology of all resected lesions.
 Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma.
-Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies , `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas.
+Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies, `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas.
 Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas.
-Fecal material provided by the patients was used for Fecal Immunological Tests (FIT) which measure human hemoglobin concentrations and for 16S rRNA gene sequencing to measure bacterial population abundances.
-The OTU abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm.
+Stool provided by the patients was used for Fecal Immunological Tests (FIT) which measure human hemoglobin concentrations and for 16S rRNA gene sequencing to measure bacterial population abundances.
+The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].
 
 #### Data definitions and pre-processing
-We use FIT results and OTU abundances as features.
-However these are in different scales.
-We use python v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` function MinMaxScaler() to transform features by scaling each feature to a [0-1] range.
+
+  The colonic disease status is re-defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs).
+Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas.
+Colonic disease status is the label predicted with each classifier.
+The bacterial abundances and FIT results are the features used to predict colonic disease status.
+Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
+There are 6920 OTUs for each sample.
+FIT levels are continuous data present for each sample.
+Because the data are in different scales, Python programming language v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` is used to transform features by scaling each feature to a [0-1] range (Table 1) [@scikit-learn].
+
+#### Learning the Classifier
+
+  To train and validate our model, labeled data is randomly split 80/20 into a training set and testing
+set.
+Then, seven binary class classifiers, L2 logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM, decision tree, random forest and XGBoost, are learned.
+The training set is used for training purposes and validation of hyperparameter selection, and the test set is used for evaluation purposes.
+Hyperparameters are selected using 5-fold cross-validation with 100-repeats on the training set.
+Since the colonic disease status are not uniformly represented in the data, 5-fold splits are stratified to maintain the overall label distribution on the training set.
+Python programming langugae v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` functions are used to learn the seven classifiers (Table 1).
+
+#### Classifier Performance
+
+  The classification performance of learned classifier is evaluated on the labeled held-out testing set.
+The optimal  classifier with optimal hyperparameters selected in the cross-validation step is used to produce a prediction for the testing set.
+The performance of this prediction is measured in terms of the sensitivity and specificity, in addition to Area Under the Curve (AUC) metrics.
+This process of splitting the data, learning a classifier with cross-validation, and testing the classifier is repeated on 100 different splits.
+In the end cross-validation AUC and testing AUC averaged over the 100 different training/test splits are reported.
+Hyperparameter budget and performance for each split is also reported.
 
@@
 Outliers are data points that are not within 3/2 times the interquartile ranges.
+
+\newpage
+
+## References
+
+<div id="refs"></div>

### 2018-11-28 submission/manuscript.tex
update manuscript

@@
 \hypersetup{unicode=true,
-            pdftitle={Evalution of binary classification pipelines and methods for 16S rRNA gene data},
+            pdftitle={Evaluation of binary classification pipelines and methods for 16S rRNA gene data},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Evalution of binary classification pipelines and methods for 16S
-rRNA gene data}}
+  \title{\textbf{Evaluation of binary classification pipelines and methods for
+16S rRNA gene data}}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-Begüm D. Topçuoğlu\^{}1, Jenna Wiens\^{}2, Patrick D.
+Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Patrick D.
 Schloss\textsuperscript{1\(\dagger\)}
@@
 a need for guidance on how to properly implement good machine learning
-practices to generate reproducible, robust and actionable models. We
-also need to generate interpretable models for biomedical researchers to
-adopt and use regularly. In this study we chose to focus on a binary
-classification model that predicts if an individual has screen-relevant
-tumors in their colon or not, to establish a model pipeline.
+practices to generate reproducible, robust and actionable models. There
+is also a clinical need to generate interpretable models for biomedical
+researchers and clinicans to adopt and use regularly (1).
 
+Recently, there is an interest on using machine learning to predict
+colorectal cancer progression with microbiota-assiciated biomarkers.
 Colorectal cancer is one of the leading cause of death among cancers in
@@
 their lifetime and once diagnosed, more than one-third will not survive
-5 years. Colonoscopy as a screening tool is very effective, however it
-is very invasive, expensive and have a low rate of patient adherence.
-Therefore, there is a need for improved non-invasive methods to screen
-individuals. One proposed non-invasive screening tool is using gut
-microbiome-based biomarkers. Patients with colorectal cancer have
-different stool community of microbes compared to adults with normal
-colons. This difference however cannot be explained by a single or a
-handful of features in the gut microbiome but by many of them in
-relation to one another. Therefore, machine learning is a great tool to
-investigate the differences between the gut microbiomes of CRC patients
-and healthy individuals. Previous studies have shown that human
-hemoglobin levels and bacterial population abundances in the stool help
-us predict screen relevant growth in the colon, however the literature
-for the problem of classification colorectal cancer diagnosis vary
-greatly, with areas under the receiver operating characteristic curve
-(AUC) of 0.7-0.8. The variation in classification performance is based
-in part on differences in the task definition, in part on differences in
-the study populations, and in part on the evaluation method. The highest
-reported AUCs were from studies of \ldots{} Additionally, some of the
-reported results were not obtained from testing on held-out sets. In
-this study, we have defined classification pipelines with L2-regularized
-logistic regression, L1 and L2 linear suppor vector machines (SVM),
-radial basis function SVM , decision tree, random forest and XGBoost. We
-evaluated the generalization and prediction performance of these
-methods. We also compared each method based on their reproducibility,
+5 years (2--4). Colonoscopy as a screening tool is very effective,
+however it is very invasive, expensive and have a low rate of patient
+adherence. Therefore, there is a need for improved non-invasive methods
+to screen individuals. Gut microbiome-based biomarkers can be used as a
+non-invasive screening method.
+
+Patients with colorectal cancer have different stool community of
+microbes compared to adults with normal colons. This difference however
+cannot be explained by a single or a handful of candidate taxa in the
+gut microbiome but by many of them in relation to one another.
+Therefore, machine learning emerges as a tool to detect the difference
+between the gut microbiomes of CRC patients and healthy individuals.
+Previous studies have shown that human hemoglobin levels and bacterial
+population abundances in the stool help us predict screen relevant
+growth in the colon, however the literature for the problem of
+classifying colorectal disease status vary greatly, with areas under the
+receiver operating characteristic curve (AUC) of 0.7-0.9 (5--8). The
+variation in classification performance is based in part on differences
+in the task definition, in part on differences in the study populations,
+and in part on the learning pipeline. In this study, classification
+pipelines with L2-regularized logistic regression, L1 and L2 linear
+suppor vector machines (SVM), radial basis function SVM , decision tree,
+random forest and XGBoost classifiers are established. The
+generalization and prediction performance of these classifiers are
+evaluated and each classifier is examined for its reproducibility,
 robustness, actionability, interpretibility and susceptibility to
@@
 
-Generalisation Perfomance of each model. Is there a maximum threshold of
-prediction with all these methods? Does an increase in model complexity
-improve predictibility? Synthesis statement regarding modeling 16S
-microbiome data
+Here, colonic disease status is defined as Normal or Screen Relevant
+Neoplasias (SRN). Stool bacterial population abundances and stool
+hemoglobin levels of 261 Normal and 229 SRN samples were used to learn
+binary classifiers and evaluate their performances. The results show
+that \ldots{}\ldots{} (Is there a maximum threshold of prediction with
+all these methods? Does an increase in model complexity improve
+predictibility?)
 
 \subsection{Results and Discussion}\label{results-and-discussion}
+
+\subparagraph{Results of modeling in text, tables and
+figures}\label{results-of-modeling-in-text-tables-and-figures}
+
+\subparagraph{Comparisons among modeling
+approaches}\label{comparisons-among-modeling-approaches}
+
+\paragraph{Interpretation of modeling results in terms of
+reproducibility, robustness, actionability, interpretibility and
+susceptibility}\label{interpretation-of-modeling-results-in-terms-of-reproducibility-robustness-actionability-interpretibility-and-susceptibility}
+
+\paragraph{Consideration of possible weaknesses for each
+model}\label{consideration-of-possible-weaknesses-for-each-model}
+
+The interactions between the biomarkers may be nonlinear. Obviously, the
+linear models will not incorporate this because they are linear. Tools
+like linear models (e.g.~metastats, lefse, wilcoxon, etc) are likely
+worthless.
+
+\paragraph{Consideration of possible weaknesses for our approach and
+chosen
+dataset}\label{consideration-of-possible-weaknesses-for-our-approach-and-chosen-dataset}
+
+\paragraph{Relationship of results to previous literature and broader
+implications of this
+work}\label{relationship-of-results-to-previous-literature-and-broader-implications-of-this-work}
+
+\paragraph{Prospects of future
+progress}\label{prospects-of-future-progress}
 
@@
 
-\paragraph{The data}\label{the-data}
+\paragraph{Data collection}\label{data-collection}
 
-We obtained stool OTU abundance data and metadata from the Sze et al.
-(1). The stool OTU abundance data and metadata comes from the Great
-Lakes- New England Early Detection Research Network study which
-collected stool samples from eligible individuals. Briefly, eligible
-patients for this study were aged at least 18 years and willing to
-collect stool samples. Colonoscopies were performed and fecal samples
-were collected from participants in four locations: Toronto (ON,
+The data used for this analysis are stool bacterial abundances, stool
+hemoglobin levels and clinical information of the patients recruited by
+Great Lakes-New England Early Detection Research Network study. These
+data were obtained from Sze et al (5). The stool samples were provided
+by recruited adult participants who were undergoing scheduled screening
+or surveillance colonoscpy. Colonoscopies were performed and fecal
+samples were collected from participants in four locations: Toronto (ON,
 Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
-Patient diagnoses were determined by colonoscopic examination and
-histopathological review of any biopsies taken. Patients with an adenoma
-greater than 1 cm, more than three adenomas of any size, or an adenoma
-with villous histology were classified as advanced adenoma. Fecal
-material was used for Fecal Immunological Tests and these tests were
-used to measure human hemoglobin concentrations. This study had 172
-control, 198 adenomas and 120 carcinomas. Of the 198 adenomas, 109 were
-advanced adenomas.
+Patients' colonic disease status was defined by colonoscopy with
+adequate preparation and tissue histopathology of all resected lesions.
+Patients with an adenoma greater than 1 cm, more than three adenomas of
+any size, or an adenoma with villous histology were classified as
+advanced adenoma. Study had 172 patients with normal colonoscopies, 198
+with adenomas and 120 with carcinomas. Of the 198 adenomas, 109 were
+identified as advanced adenomas. Stool provided by the patients was used
+for Fecal Immunological Tests (FIT) which measure human hemoglobin
+concentrations and for 16S rRNA gene sequencing to measure bacterial
+population abundances. The bacterial abundance data was generated by Sze
+et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the
+default quality filtering methods, identifying and removing chimeric
+sequences using VSEARCH and assigning to OTUs at 97\% similarity using
+the OptiClust algorithm (9--11).
+
+\paragraph{Data definitions and
+pre-processing}\label{data-definitions-and-pre-processing}
+
+The colonic disease status is re-defined as two encompassing classes;
+Normal or Screen Relevant Neoplasias (SRNs). Normal class includes
+patients with non-advanced adenomas or normal colons whereas SRN class
+includes patients with advanced adenomas or carcinomas. Colonic disease
+status is the label predicted with each classifier. The bacterial
+abundances and FIT results are the features used to predict colonic
+disease status. Bacterial abundances are discrete data in the form of
+Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
+sample. FIT levels are continuous data present for each sample. These 2
+different data are in different scales. Python programming language
+v3.6.6, module scikit-learn v0.19.2 is used to transform features by
+scaling each feature to a {[}0-1{]} range (Table 1) (12).
+
+\paragraph{Learning the Classifier}\label{learning-the-classifier}
+
+To train and validate our model, labeled data is randomly split 80/20
+into a training set and testing set. Then, seven binary class
+classifiers, L2 logistic regression, L1 and L2 linear suppor vector
+machines (SVM), radial basis function SVM, decision tree, random forest
+and XGBoost, are learned. The training set is used for training purposes
+and validation of hyperparameter selection, and the test set is used for
+evaluation purposes. Hyperparameters are selected using 5-fold
+cross-validation with 100-repeats on the training set. Since the colonic
+disease status are not uniformly represented in the data, 5-fold splits
+are stratified to maintain the overall label distribution on the
+training set. v3.6.6, module scikit-learn v0.19.2 functions are used to
+learn the seven classifiers (Table 1).
+
+\paragraph{Classifier Performance}\label{classifier-performance}
+
+The classification performance of learned classifier is evaluated on the
+labeled held-out testing set. The optimal classifier with optimal
+hyperparameters selected in the cross-validation step is used to produce
+a prediction for the testing set. The performance of this prediction is
+measured in terms of the sensitivity and specificity, in addition to
+Area Under the Curve (AUC) metrics. This process of splitting the data,
+learning a classifier with cross-validation, and testing the classifier
+is repeated on 100 different splits. In the end cross-validation AUC and
+testing AUC averaged over the 100 different training/test splits are
+reported. Hyperparameter budget and performance for each split is also
+reported.
 
@@
 
+\newpage
+
+\subsection{References}\label{references}
+
+\hypertarget{refs}{}
+\hypertarget{ref-wiens_editorial:_2016}{}
+1. \textbf{Wiens J}, \textbf{Wallace BC}. 2016. Editorial: Special issue
+on machine learning for health and medicine. Machine Learning
+\textbf{102}:305--307.
+doi:\href{https://doi.org/10.1007/s10994-015-5533-9}{10.1007/s10994-015-5533-9}.
+
+\hypertarget{ref-seer_2016}{}
+2. \textbf{Howlader N KM \textnormal{Noone AM}}. SEER cancer statistics
+review, 1975-2013, (national cancer institute. bethesda, md).
+
+\hypertarget{ref-street_colorectal_nodate}{}
+3. \textbf{Street W}. Colorectal cancer facts \& figures 2017-2019 40.
+
+\hypertarget{ref-weir_past_2015}{}
+4. \textbf{Weir HK}, \textbf{Thompson TD}, \textbf{Soman A},
+\textbf{MÃžller B}, \textbf{Leadbetter S}. 2015. The past, present, and
+future of cancer incidence in the united states: 1975 through 2020.
+Cancer \textbf{121}:1827--1837.
+doi:\href{https://doi.org/10.1002/cncr.29258}{10.1002/cncr.29258}.
+
+\hypertarget{ref-sze_leveraging_2018}{}
+5. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+rRNA gene surveys to identify reproducible biomarkers in individuals
+with colorectal tumors. mBio \textbf{9}:e00630--18.
+doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
+
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+6. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\hypertarget{ref-baxter_dna_2016}{}
+7. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
+\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
+immunochemical test can replace stool for detection of colonic lesions
+using a microbiota-based model. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
+
+\hypertarget{ref-zackular_human_2014}{}
+8. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
+
+\hypertarget{ref-schloss_introducing_2009}{}
+9. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-Source, Platform-Independent,
+Community-Supported Software for Describing and Comparing Microbial
+Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.
+
+\hypertarget{ref-westcott_opticlust_2017}{}
+10. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+Improved Method for Assigning Amplicon-Based Sequence Data to
+Operational Taxonomic Units. mSphere \textbf{2}.
+doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
+
+\hypertarget{ref-rognes_vsearch_2016}{}
+11. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
+source tool for metagenomics. PeerJ \textbf{4}:e2584.
+doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+
+\hypertarget{ref-scikit-learn}{}
+12. \textbf{Pedregosa F}, \textbf{Varoquaux G}, \textbf{Gramfort A},
+\textbf{Michel V}, \textbf{Thirion B}, \textbf{Grisel O},
+\textbf{Blondel M}, \textbf{Prettenhofer P}, \textbf{Weiss R},
+\textbf{Dubourg V}, \textbf{Vanderplas J}, \textbf{Passos A},
+\textbf{Cournapeau D}, \textbf{Brucher M}, \textbf{Perrot M},
+\textbf{Duchesnay E}. 2011. Scikit-learn: Machine learning in Python.
+Journal of Machine Learning Research \textbf{12}:2825--2830.
+
 \end{document}

### 2018-11-28 submission/Microbe_Abstract2019.Rmd
microbe conference submission

--- 2018-11-19 submission/manuscript.Rmd
@@
 ---
-title: "**Evalution of binary classification pipelines and methods for 16S rRNA gene data**"
 bibliography: references.bib
-output:
-  pdf_document:
-    keep_tex: true
-    includes:
-      in_header: header.tex
+output: word_document
 csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
@@
 
+logit_summary <- logit %>%  
+  group_by(Performance) %>% 
+  summarise(meanAUC=mean(AUC), std=sd(AUC))
+
+cv_meanAUC <- logit_summary[1,2]
+cv_sdAUC <- logit_summary[1,3]
+test_meanAUC <- logit_summary[2,2]
+test_sdAUC <- logit_summary[2,3]
+
 # Read in AUCs table generated from l1 SVM linear kernel for all samples
@@
   mutate(model="L1-SVM Linear")
+
+l1svm_summary <- l1svm %>%  
+  group_by(Performance) %>% 
+  summarise(meanAUC=mean(AUC), std=sd(AUC))
 
@@
 
+l2svm_summary <- l2svm %>%  
+  group_by(Performance) %>% 
+  summarise(meanAUC=mean(AUC), std=sd(AUC))
+
 # Read in AUCs table generated from  SVM RBF kernel for all samples
@@
   mutate(model="SVM RBF")
+
+svmRBF_summary <- svmRBF %>%  
+  group_by(Performance) %>% 
+  summarise(meanAUC=mean(AUC), std=sd(AUC))
 
@@
 
+xgboost_summary <- xgboost %>%  
+  group_by(Performance) %>% 
+  summarise(meanAUC=mean(AUC), std=sd(AUC))
+
 # Read in AUCs table generated from random forest for all samples
@@
 
+rf_summary <- rf %>%  
+  group_by(Performance) %>% 
+  summarise(meanAUC=mean(AUC), std=sd(AUC))
 # Read in AUCs table generated from decision tree for all samples
@@
 
-######################################################################
-#------------ Put all the loaded AUC tables together -----------------#
-######################################################################
-all <- bind_rows(logit, l1svm, l2svm, svmRBF, xgboost, rf, dt) %>% 
-  group_by(model)
-
-######################################################################
-#---Load .tsv data with cvAUCs for hyper-parameters  ----#
-######################################################################
-# Read in hyper-parameter AUCs generated from L2 logistic regression model for all samples:
-logit <- read.delim('../data/process/L2_Logistic_Regression_parameters.tsv', header=T, sep='\t')%>%
-  select(-X) %>% 
-  mutate(meanAUC=rowMeans(.[, 2:101])) %>% 
-  select(C, meanAUC) 
-# Read in hyper-parameter AUCs generated from L1 Linear SVM model for all samples:
-l1svm <- read.delim('../data/process/L1_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
-  select(-X) %>% 
-  mutate(meanAUC=rowMeans(.[, 2:101])) %>% 
-  select(C, meanAUC) 
-# Read in hyper-parameter AUCs generated from L2 Linear SVM model for all samples:
-l2svm <- read.delim('../data/process/L2_SVM_Linear_Kernel_parameters.tsv', header=T, sep='\t')%>%
-  select(-X) %>% 
-  mutate(meanAUC=rowMeans(.[, 2:101])) %>% 
-  select(C, meanAUC)
-
+dt_summary <- dt %>%  
+  group_by(Performance) %>% 
+  summarise(meanAUC=mean(AUC), std=sd(AUC))
 ```
@@
 ```
+\linenumbers
+__Evaluation of machine learning methods that predict colorectal cancer progression with microbiota-associated biomarkers__
 
-\vspace{35mm}
+Begüm D. Topçuoğlu, Jenna Wiens, Mack Ruffin, Patrick D. Schloss
 
-Running title: Machine learning methods in microbiome studies
+As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning methods to determine association of the microbiome with a continuous or categorical phenotype of interest.
+Currently, the use of machine learning in microbiome research lack clarity and consistency over the training, validation and testing of models.
+There is a need for guidance on how to properly implement good machine learning practices to generate reproducible and robust models.
 
-\vspace{35mm}
+Recently, there has been growing interest in using machine learning to predict colorectal cancer progression with microbiota-associated biomarkers.
+Colorectal cancer is one of the leading cause of death among cancers in the United States.
+Colonoscopy as a screening tool is effective, however it is invasive, expensive and have a low rate of patient adherence.
+Previous studies have shown that bacterial population abundances in the stool can predict screen relevant growth in the colon and can be used as a non-invasive screening tool.
+However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014].
+The variation in prediction performance is based in part on differences in the study populations, and in part on the differences in training, validation and testing pipelines.
 
-Begüm D. Topçuoğlu^1, Jenna Wiens^2, Patrick D. Schloss^1$\dagger$^
-
-\vspace{40mm}
-
-$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
-
-1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
-
-2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109
-
-\newpage
-\linenumbers
-
-## Abstract
+In this study, hemoglobin levels and bacterial population abundances in the stool were used to predict colorectal disease status of `r paste(nrow(meta))` patients.
+The colorectal disease status was defined as showing screen-relevant colonic growth or not.
+Training, validation and testing pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function (RBF) SVM, Decision Tree, Random Forest and XGBoost binary classifiers.
+L2-regularized Logistic Regression, L1 Linear SVM, L2 Linear SVM, and RBF SVM had mean AUCs of `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, and `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, respectively.
+Decision Tree, Random Forest and XGBoost had mean AUCs of `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively.
+Tree-based models were less susceptible to overfitting and in general had higher sensitivity and specificity for colonic screen-relevant growth.
+Aside from evaluating generalization and classification performance of each classifier, this study established standards for training, validation and testing of the microbiome-associated machine learning models.
 
@@
 
-## Introduction
+__References__
 
-As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning  methods to analyze 16S rRNA gene sequence data and to determine association of the microbiome with a continuous or categorical phenotype of interest.
-The use of machine learning in microbiome literature lack clarity over the learning pipeline which spans the problem formulation, feature selection, feature pre-processing, model learning, and output.
-There is a need for guidance on how to properly implement good machine learning practices to generate reproducible, robust and actionable models.
-We also need to generate interpretable models for biomedical researchers to adopt and use regularly.
-In this study we chose to focus on a binary classification model that predicts if an individual has screen-relevant tumors in their colon or not, to establish a model pipeline.
-
-Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Each person in the industrialized world has on average a one-in-twenty chance of developing colorectal cancer (CRC) in their lifetime and once diagnosed, more than one-third will not survive 5 years.
-Colonoscopy as a screening tool is very effective, however it is very invasive, expensive and have a low rate of patient adherence.
-Therefore, there is a need for improved non-invasive methods to screen individuals.
-One proposed non-invasive screening tool is using gut microbiome-based biomarkers.
-Patients with colorectal cancer have different stool community of microbes compared to adults with normal colons.
-This difference however cannot be explained by a single or a handful of features in the gut microbiome but by many of them in relation to one another.
-Therefore, machine learning is a great tool to investigate the differences between the gut microbiomes of CRC patients and healthy individuals.
-Previous studies have shown that human hemoglobin levels and bacterial population abundances in the stool help us predict screen relevant growth in the colon, however the literature for the problem of classification colorectal cancer diagnosis vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.8.
-The variation in classification performance is based in part on differences in the task definition, in
-part on differences in the study populations, and in part on the evaluation method. The highest
-reported AUCs were from studies of ...
-In this study, we have defined classification pipelines with L2-regularized logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM , decision tree, random forest and XGBoost.
-We evaluated the generalization and prediction performance of these methods.
-We also compared each method based on their reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting.
-
-Generalisation Perfomance of each model. 
-Is there a maximum threshold of prediction with all these methods?
-Does an increase in model complexity improve predictibility?
-Synthesis statement regarding modeling 16S microbiome data
-
-## Results and Discussion
-
-## Conclusions
-
-## Materials and Methods
-
-#### Data collection
-We obtained stool OTU abundance data and metadata from the Sze et al.
-(1).
-The stool OTU abundance data and metadata were generated from the Great Lakes-New England Early Detection Research Network study which collected stool samples from recruited participants who were undergoing scheduled screening or surveillance colonoscpy.
-Colonoscopies were performed and fecal samples were collected from adult participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
-Patients' colonic disease status was defined by colonoscopy with adequate preparation and tissue histopathology of all resected lesions.
-Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma.
-Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies , `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas.
-Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas.
-Fecal material provided by the patients was used for Fecal Immunological Tests (FIT) which measure human hemoglobin concentrations and for 16S rRNA gene sequencing to measure bacterial population abundances.
-The OTU abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm.
-
-#### Data definitions and pre-processing
-We use FIT results and OTU abundances as features.
-However these are in different scales.
-We use python v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` function MinMaxScaler() to transform features by scaling each feature to a [0-1] range.
-
-\newpage
-
-**Figure 1.
-Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances.
-The boxplot shows quartiles at the box ends and the statistical median as athe horizontal line in the box.
-The whiskers show the farthest points that are not outliers.
-Outliers are data points that are not within 3/2 times the interquartile ranges.
+<div id="refs"></div>

### 2018-11-28 submission/Microbe_Abstract2019.tex
microbe conference submission

--- 2018-11-16 submission/manuscript.tex
@@
 \hypersetup{unicode=true,
-            pdftitle={Evalution of binary classification pipelines and methods for 16S rRNA gene data},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Evalution of binary classification pipelines and methods for 16S
-rRNA gene data}}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
+  \title{}
+    \pretitle{\vspace{\droptitle}}
+  \posttitle{}
     \author{}
@@
 \begin{document}
-\maketitle
 
-\vspace{35mm}
+\linenumbers
+\textbf{Evaluation of classification pipelines that predict colorectal
+cancer progression with microbiota-associated biomarkers}
 
-Running title: Machine learning methods in microbiome studies
-
-\vspace{35mm}
-
-Begüm D. Topçuoğlu\^{}1, Jenna Wiens\^{}2, Patrick D.
+Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Patrick D.
 Schloss\textsuperscript{1\(\dagger\)}
 
-\vspace{40mm}
+As gut microbiome field continues to grow, there will be an
+ever-increasing demand for reproducible machine learning methods to
+determine association of the microbiome with a continuous or categorical
+phenotype of interest. Currently, the use of machine learning in
+microbiome literature lack clarity over the training, validation and
+testing of the models used. There is a need to properly implement good
+machine learning practices to generate reproducible and robust models.
 
-\(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+Recently, there is an interest in using machine learning to predict
+colorectal cancer progression with microbiota-associated biomarkers.
+Colorectal cancer is one of the leading cause of death among cancers in
+the United States. Each person in the industrialized world has on
+average a one-in-twenty chance of developing colorectal cancer (CRC)
+(1--3). Colonoscopy as a screening tool is effective, however it is
+invasive, expensive and have a low rate of patient adherence. Therefore,
+gut microbiome-based biomarkers emerged as a non-invasive screening
+method.
 
-1. Department of Microbiology and Immunology, University of Michigan,
-Ann Arbor, MI 48109
-
-2. Department of Computer Science and Engineering, University or
-Michigan, Ann Arbor, MI 49109
+In this study, classification models that use human hemoglobin levels
+and bacterial population abundances in the stool were used to predict
+colorectal disease status as screen-relevant colonic growth or not.
+Training, validation and testing pipelines were established for
+L2-regularized Logistic Regression, L1 and L2 Linear Suppor Vector
+Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest
+and XGBoost classifiers. The generalization and prediction performance
+of these classifiers were evaluated and each classifier was examined for
+its reproducibility, robustness and susceptibility to overfitting.
+L2-regularized Logistic Regression had a mean AUC of 0.68 +/- 0.04, L1
+Linear SVM had a mean AUC of 0.76 +/- 0.05, L2 Linear SVM had a mean AUC
+of 0.68 +/- 0.05 and Radial Basis Function SVM had a mean AUC of 0.69
++/- 0.05. Decision Tree had a mean AUC of 0.7 +/- 0.05, Random Forest
+had a mean AUC of 0.76 +/- 0.06 and XGBoost had a mean AUC of 0.76 +/-
+0.04. Tree-based models were less susceptible to overfiting and in
+general had higher sensitivity and specificity for colonic
+screen-relevant growth.
 
@@
 
-\linenumbers
+\subsection{References}\label{references}
 
-\subsection{Abstract}\label{abstract}
+\hypertarget{refs}{}
+\hypertarget{ref-seer_2016}{}
+1. \textbf{Howlader N KM \textnormal{Noone AM}}. SEER cancer statistics
+review, 1975-2013, (national cancer institute. bethesda, md).
 
-\newpage
+\hypertarget{ref-street_colorectal_nodate}{}
+2. \textbf{Street W}. Colorectal cancer facts \& figures 2017-2019 40.
 
-\subsection{Introduction}\label{introduction}
-
-As gut microbiome field continues to grow, there will be an
-ever-increasing demand for reproducible machine learning methods to
-analyze 16S rRNA gene sequence data and to determine association of the
-microbiome with a continuous or categorical phenotype of interest. The
-use of machine learning in microbiome literature lack clarity over the
-learning pipeline which spans the problem formulation, feature
-selection, feature pre-processing, model learning, and output. There is
-a need for guidance on how to properly implement good machine learning
-practices to generate reproducible, robust and actionable models. We
-also need to generate interpretable models for biomedical researchers to
-adopt and use regularly. In this study we chose to focus on a binary
-classification model that predicts if an individual has screen-relevant
-tumors in their colon or not, to establish a model pipeline.
-
-Colorectal cancer is one of the leading cause of death among cancers in
-the United States. Each person in the industrialized world has on
-average a one-in-twenty chance of developing colorectal cancer (CRC) in
-their lifetime and once diagnosed, more than one-third will not survive
-5 years. Colonoscopy as a screening tool is very effective, however it
-is very invasive, expensive and have a low rate of patient adherence.
-Therefore, there is a need for improved non-invasive methods to screen
-individuals. One proposed non-invasive screening tool is using gut
-microbiome-based biomarkers. Patients with colorectal cancer have
-different stool community of microbes compared to adults with normal
-colons. This difference however cannot be explained by a single or a
-handful of features in the gut microbiome but by many of them in
-relation to one another. Therefore, machine learning is a great tool to
-investigate the differences between the gut microbiomes of CRC patients
-and healthy individuals. Previous studies have shown that human
-hemoglobin levels and bacterial population abundances in the stool help
-us predict screen relevant growth in the colon, however the literature
-for the problem of classification colorectal cancer diagnosis vary
-greatly, with areas under the receiver operating characteristic curve
-(AUC) of 0.7-0.8. The variation in classification performance is based
-in part on differences in the task definition, in part on differences in
-the study populations, and in part on the evaluation method. The highest
-reported AUCs were from studies of \ldots{} Additionally, some of the
-reported results were not obtained from testing on held-out sets. In
-this study, we have defined classification pipelines with L2-regularized
-logistic regression, L1 and L2 linear suppor vector machines (SVM),
-radial basis function SVM , decision tree, random forest and XGBoost. We
-evaluated the generalization and prediction performance of these
-methods. We also compared each method based on their reproducibility,
-robustness, actionability, interpretibility and susceptibility to
-overfitting.
-
-Generalisation Perfomance of each model. Is there a maximum threshold of
-prediction with all these methods? Does an increase in model complexity
-improve predictibility? Synthesis statement regarding modeling 16S
-microbiome data
-
-\subsection{Results and Discussion}\label{results-and-discussion}
-
-\subsection{Conclusions}\label{conclusions}
-
-\subsection{Materials and Methods}\label{materials-and-methods}
-
-\paragraph{The data}\label{the-data}
-
-We obtained stool OTU abundance data and metadata from the Sze et al.
-(1). The stool OTU abundance data and metadata comes from the Great
-Lakes- New England Early Detection Research Network study which
-collected stool samples from eligible individuals. Briefly, eligible
-patients for this study were aged at least 18 years and willing to
-collect stool samples. Colonoscopies were performed and fecal samples
-were collected from participants in four locations: Toronto (ON,
-Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
-Patient diagnoses were determined by colonoscopic examination and
-histopathological review of any biopsies taken. Patients with an adenoma
-greater than 1 cm, more than three adenomas of any size, or an adenoma
-with villous histology were classified as advanced adenoma. Fecal
-material was used for Fecal Immunological Tests and these tests were
-used to measure human hemoglobin concentrations. This study had 172
-control, 198 adenomas and 120 carcinomas. Of the 198 adenomas, 109 were
-advanced adenomas.
-
-\newpage
-
-\textbf{Figure 1. Generalization and classification performance of
-modeling methods } AUC values of all cross validation and testing
-performances. The boxplot shows quartiles at the box ends and the
-statistical median as athe horizontal line in the box. The whiskers show
-the farthest points that are not outliers. Outliers are data points that
-are not within 3/2 times the interquartile ranges.
+\hypertarget{ref-weir_past_2015}{}
+3. \textbf{Weir HK}, \textbf{Thompson TD}, \textbf{Soman A},
+\textbf{MÃžller B}, \textbf{Leadbetter S}. 2015. The past, present, and
+future of cancer incidence in the united states: 1975 through 2020.
+Cancer \textbf{121}:1827--1837.
+doi:\href{https://doi.org/10.1002/cncr.29258}{10.1002/cncr.29258}.
 

### 2018-11-29 submission/Microbe_Abstract2019.Rmd
final abstract

@@
 bibliography: references.bib
-output: word_document
-csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
+csl: mbio.csl
 fontsize: 11pt
 geometry: margin=1.0in
+output:
+  word_document: default
+  pdf_document: 
+    includes:
+      in_header: header.tex
+    keep_tex: yes
 ---
@@
 \linenumbers
-__Evaluation of machine learning methods that predict colorectal cancer progression with microbiota-associated biomarkers__
+__Evaluation of machine learning methods that identify colorectal tumors with microbiota-associated biomarkers__
 
@@
 
-As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning methods to determine association of the microbiome with a continuous or categorical phenotype of interest.
-Currently, the use of machine learning in microbiome research lack clarity and consistency over the training, validation and testing of models.
-There is a need for guidance on how to properly implement good machine learning practices to generate reproducible and robust models.
+As gut microbiome field continues to grow, there is an ever-increasing demand for reproducible machine learning methods to determine associations between the microbiome and a phenotype of interest.
+Currently, the use of machine learning in microbiome research lacks clarity and consistency over the modeling pipeline (training, validation and testing steps).
+There is a need for guidance on how to implement good machine learning practices to generate reproducible and robust models.
 
-Recently, there has been growing interest in using machine learning to predict colorectal cancer progression with microbiota-associated biomarkers.
+Recently, there has been growing interest in using machine learning to identify colorectal tumors with microbiota-associated biomarkers.
 Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Colonoscopy as a screening tool is effective, however it is invasive, expensive and have a low rate of patient adherence.
-Previous studies have shown that bacterial population abundances in the stool can predict screen relevant growth in the colon and can be used as a non-invasive screening tool.
+Colonoscopy as a screening tool is effective, however it is invasive and have a low rate of patient adherence.
+Previous studies have shown that bacterial abundances in the stool can predict colorectal tumors in the colon and can be used as a non-invasive screening tool.
 However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014].
-The variation in prediction performance is based in part on differences in the study populations, and in part on the differences in training, validation and testing pipelines.
+The variation in prediction performance is based in part on differences in the study populations, and in part on the differences in modeling pipelines.
 
-In this study, hemoglobin levels and bacterial population abundances in the stool were used to predict colorectal disease status of `r paste(nrow(meta))` patients.
-The colorectal disease status was defined as showing screen-relevant colonic growth or not.
-Training, validation and testing pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function (RBF) SVM, Decision Tree, Random Forest and XGBoost binary classifiers.
-L2-regularized Logistic Regression, L1 Linear SVM, L2 Linear SVM, and RBF SVM had mean AUCs of `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, and `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, respectively.
-Decision Tree, Random Forest and XGBoost had mean AUCs of `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively.
-Tree-based models were less susceptible to overfitting and in general had higher sensitivity and specificity for colonic screen-relevant growth.
-Aside from evaluating generalization and classification performance of each classifier, this study established standards for training, validation and testing of the microbiome-associated machine learning models.
+In this study, hemoglobin counts and 16S gene sequences in the stool were used to identify colorectal tumors of `r paste(nrow(meta))` patients as advanced tumors or not.
+Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models.
+The mean AUCs of these models were `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, and `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively.
+Tree-based methods, namely Decision Tree, Random Forest and XGBoost were less susceptible to overfitting and in general had higher sensitivity and specificity for advanced tumors.
+Aside from evaluating generalization and classification performance of each model, this study established standards for modeling pipelines of microbiome-associated machine learning models.
 

### 2018-12-03 submission/manuscript.Rmd
typo
update

@@
 output:
+  word_document: default
   pdf_document:
@@
 (Is there a maximum threshold of prediction with all these methods?
-Does an increase in model complexity improve predictibility?)
+Does an increase in model complexity improve predictability?)
 
@@
 These data were obtained from Sze et al [@sze_leveraging_2018].
-  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscpy.
+  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscopy.
 Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
@@
 Since the colonic disease status are not uniformly represented in the data, 5-fold splits are stratified to maintain the overall label distribution on the training set.
-Python programming langugae v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` functions are used to learn the seven classifiers (Table 1).
+Python programming language v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` functions are used to learn the seven classifiers (Table 1).
 
@@
 Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances.
-The boxplot shows quartiles at the box ends and the statistical median as athe horizontal line in the box.
+The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
 The whiskers show the farthest points that are not outliers.

### 2018-12-07 code/learning/individual_models/L1LinearSVM.R
updates

@@ -7,7 +7,7 @@
 #### To be able to run this script we need to be in our project directory.
 
 #### The dependinces for this script are consolidated in the first part
-deps = c("doParallel","pROC", "caret","knitr","rmarkdown","vegan","gtools", "tidyverse");
+deps = c("LiblineaR", "doParallel","pROC", "caret","knitr","rmarkdown","vegan","gtools", "tidyverse");
 for (dep in deps){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), quiet=TRUE);
@@ -32,71 +32,61 @@ data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
 # We want the diagnosis column to a factor
 data$dx <- factor(data$dx, labels=c("normal", "cancer"))
 
-# Create
-all.test.response <- all.test.predictor <- test_aucs <- c()
-#all.cv.response <- all.cv.predictor <- cv_aucs <- c()
+all.test.response <- all.test.predictor <- c()
+all.cv.response <- all.cv.predictor <- c()
 cl <- makePSOCKcluster(4)
 registerDoParallel(cl)
-for (i in 1:50) {
+for (i in 1:5) {
   inTraining <- createDataPartition(data$dx, p = .80, list = FALSE)
   training <- data[ inTraining,]
   testing  <- data[-inTraining,]
+  preProcValues <- preProcess(training, method = "range")
+  training <- predict(preProcValues, training)
+  testing <- predict(preProcValues, testing)
   x_train <- training %>% select(-dx)
-  y_train <- training$dx
+  y_train <- ifelse(training$dx == "cancer", 1, 0)
   y_train <- as.factor(y_train)
-  grid <-  expand.grid(cost = c(0.0000001, 0.000001, 0.00001, 0.0001),
-                       loss = "L2_dual",
-                       epsilon = 0.1)
-  cv <- trainControl(method="repeatedcv",
-                     repeats = 50,
-                     number=5,
-                     returnResamp="final",
-                     classProbs=TRUE,
-                     summaryFunction=twoClassSummary,
-                     indexFinal=NULL,
-                     preProc = "scale",
-                     savePredictions = TRUE)
+  x_test <- testing %>% select(-dx)
+  y_test <- testing$dx
[…]

### 2018-12-07 code/learning/individual_models/RBF_SVM.R

@@ -34,8 +34,8 @@ data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
     Dx_Bin== "High Risk Normal" ~ "normal",
     Dx_Bin== "adv Adenoma" ~ "cancer",
     Dx_Bin== "Cancer" ~ "cancer"
-  )) %>% 
-  select(-sample, -Dx_Bin) %>% 
+  )) %>%
+  select(-sample, -Dx_Bin) %>%
   drop_na()
 
 # We want the diagnosis column to a factor
@@ -56,8 +56,8 @@ for (i in 1:100) {
   preProcValues <- preProcess(training, method = "range")
   trainTransformed <- predict(preProcValues, training)
   testTransformed <- predict(preProcValues, testing)
-  grid <-  expand.grid(sigma = c(0.001, 0.005, 0.01, 0.05),
-                       C = c(0.005, 0.01, 0.05, 0.1))
+  grid <-  expand.grid(sigma = c(0.00001, 0.0001, 0.001, 0.005),
+                       C = c(0.00001,0.0001,0.001, 0.01))
   cv <- trainControl(method="repeatedcv",
                      repeats = 10,
                      number=5,
@@ -66,7 +66,7 @@ for (i in 1:100) {
                      summaryFunction=twoClassSummary,
                      indexFinal=NULL,
                      savePredictions = TRUE)
-  
+
   rbf_SVM <- train(dx ~ .,
                    data=trainTransformed,
                    method = "svmRadial",
@@ -80,7 +80,7 @@ for (i in 1:100) {
   # Save the best C parameter
   best.tunes.sigma <- c(best.tunes.sigma, best.tune.sigma)
   best.tunes.cost <- c(best.tunes.cost, best.tune.cost)
-  # Print the best C parameter 
+  # Print the best C parameter
   print(rbf_SVM$bestTune)
   # Mean AUC value over repeats of the best cost parameter during training
   cv_auc <- getTrainPerf(rbf_SVM)$TrainROC
@@ -161,4 +161,3 @@ legend(x=0.7,y=0.2,
 
 # Save the figure
 dev.off()
-

### 2018-12-07 submission/HHMI_Plot.Rmd
HHMI plot initial commit

--- 2018-11-29 submission/Microbe_Abstract2019.Rmd
@@
 ---
-bibliography: references.bib
-csl: mbio.csl
+title: ""
+output:
+  pdf_document:
+    keep_tex: true
+    includes:
+      in_header: header.tex
+csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
 fontsize: 11pt
 geometry: margin=1.0in
-output:
-  word_document: default
-  pdf_document: 
-    includes:
-      in_header: header.tex
-    keep_tex: yes
 ---
@@
 
-```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
-######################################################################
-#----------------- Load OTU table and MetaData -----------------#
-######################################################################
-
-# Read in metadata and select only sample Id and diagnosis columns
-meta <- read.delim('../data/metadata.tsv', header=T, sep='\t') %>%
-  select(sample, dx, Dx_Bin, fit_result)
-
-# Read in OTU table and remove label and numOtus columns
-shared <- read.delim('../data/baxter.0.03.subsample.shared', header=T, sep='\t') %>%
-   select(-label, -numOtus)
-
-######################################################################
-#---Load .tsv data with CV and test AUC data generated in Python ----#
-######################################################################
-
-# Read in AUCs table generated from L2 logistic regression model for all samples:
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax scaler
-logit <- read.delim('../data/process/L2_Logistic_Regression.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="L2-Logistic Regression")
-
-logit_summary <- logit %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-cv_meanAUC <- logit_summary[1,2]
-cv_sdAUC <- logit_summary[1,3]
-test_meanAUC <- logit_summary[2,2]
-test_sdAUC <- logit_summary[2,3]
-
-# Read in AUCs table generated from l1 SVM linear kernel for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is Standard scaler
-l1svm <- read.delim('../data/process/L1_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="L1-SVM Linear")
-
-l1svm_summary <- l1svm %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from l2 SVM linear kernel for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is Standard scaler
-l2svm <- read.delim('../data/process/L2_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="L2-SVM Linear")
-
-l2svm_summary <- l2svm %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from  SVM RBF kernel for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is Standard scaler
-svmRBF <- read.delim('../data/process/SVM_RBF.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="SVM RBF")
-
-svmRBF_summary <- svmRBF %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from xgboost for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax (0-1) scaler
-xgboost <- read.delim('../data/process/XGBoost.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="XGBoost")
-
-xgboost_summary <- xgboost %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from random forest for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax (0-1) scaler
-rf <- read.delim('../data/process/Random_Forest.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="Random Forest ")
-
-rf_summary <- rf %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-# Read in AUCs table generated from decision tree for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax (0-1) scaler
-dt <- read.delim('../data/process/Decision_Tree.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="Decision Tree")
-
-dt_summary <- dt %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-```
-
-```{python, cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
-import platform
-python_version = platform.python_version()
-import sklearn
-sklearn_module = sklearn.__version__
-```
-\linenumbers
-__Evaluation of machine learning methods that identify colorectal tumors with microbiota-associated biomarkers__
-
-Begüm D. Topçuoğlu, Jenna Wiens, Mack Ruffin, Patrick D. Schloss
-
-As gut microbiome field continues to grow, there is an ever-increasing demand for reproducible machine learning methods to determine associations between the microbiome and a phenotype of interest.
-Currently, the use of machine learning in microbiome research lacks clarity and consistency over the modeling pipeline (training, validation and testing steps).
-There is a need for guidance on how to implement good machine learning practices to generate reproducible and robust models.
-
-Recently, there has been growing interest in using machine learning to identify colorectal tumors with microbiota-associated biomarkers.
-Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Colonoscopy as a screening tool is effective, however it is invasive and have a low rate of patient adherence.
-Previous studies have shown that bacterial abundances in the stool can predict colorectal tumors in the colon and can be used as a non-invasive screening tool.
-However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014].
-The variation in prediction performance is based in part on differences in the study populations, and in part on the differences in modeling pipelines.
-
-In this study, hemoglobin counts and 16S gene sequences in the stool were used to identify colorectal tumors of `r paste(nrow(meta))` patients as advanced tumors or not.
-Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models.
-The mean AUCs of these models were `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, and `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively.
-Tree-based methods, namely Decision Tree, Random Forest and XGBoost were less susceptible to overfitting and in general had higher sensitivity and specificity for advanced tumors.
-Aside from evaluating generalization and classification performance of each model, this study established standards for modeling pipelines of microbiome-associated machine learning models.
-
-\newpage
-
-__References__
-
-<div id="refs"></div>
+![Generalization and classification performance of modeling methods.
+Tukey box-plots show the cross validation and testing AUC values of the L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models over the 100 different training/test splits.](../results/figures/AUC_comparison.png)

### 2018-12-15 submission/manuscript.Rmd

@@
 
+As the microbiome field continues to grow, there is an ever-increasing demand for reproducible machine learning methods for identifying associations between members of the microbiome and phenotypes.
+Currently, the field’s use of machine learning lacks clarity and consistency.
+There is a need to implement good machine learning practices to generate reproducible and robust models.
+One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
+Colorectal cancer is one of the leading cause of death among cancers in the United States.
+Colonoscopy as a screening tool is effective, however it is invasive and have a low rate of patient adherence.
+Previous studies have shown that bacterial abundances in the stool can predict colorectal tumors in the colon and can be used as a non-invasive screening tool.
+However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014].
+We used the fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples to classify `r paste(nrow(meta))` patients as having advanced tumors or not.
+Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and Extreme Gradient Boosting (XGBoost) binary classification models.
+Tree-based methods, namely Decision Tree, Random Forest and XGBoost were less susceptible to overfitting and in general had higher sensitivity and specificity for advanced tumors.
+Aside from evaluating generalization and classification performance of each model, this study established standards for modeling pipelines of microbiome-associated machine learning models.
+
 \newpage
@@
 
-As gut microbiome field continues to grow, there will be an ever-increasing demand for reproducible machine learning  methods to analyze 16S rRNA gene sequence data and to determine association of the microbiome with a continuous or categorical phenotype of interest.
-The use of machine learning in microbiome literature lack clarity over the learning pipeline which spans the problem formulation, feature selection, feature pre-processing, model learning, and output.
-There is a need for guidance on how to properly implement good machine learning practices to generate reproducible, robust and actionable models.
-There is also a clinical need to generate interpretable models for biomedical researchers and clinicans to adopt and use regularly [@wiens_editorial:_2016].
+Colorectal cancer is the second most commonly occurring cancer in men and third in women, with 1.8 million new cases in 2018 (8).
+Early detection of screen-relevant neoplasias (SRN) which are cancerous or advanced precancerous lesions, is crucial for treatment of colorectal cancer.
+Patients whose lesions are detected at an early stage have a greater than 90 % chance of survival, yet a significant proportion of the US population remain unscreened (9).
+The high cost and invasiveness of screening methods such as colonoscopy and sigmoidoscopy lower patient adherence.
+There is a need for non-invasive screening tests that have high sensitivity and specificity for SRN.
 
-Recently, there is an interest on using machine learning to predict colorectal cancer progression with microbiota-assiciated biomarkers.
-Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Each person in the industrialized world has on average a one-in-twenty chance of developing colorectal cancer (CRC) in their lifetime and once diagnosed, more than one-third will not survive 5 years [@seer_2016; @street_colorectal_nodate; @weir_past_2015].
-Colonoscopy as a screening tool is very effective, however it is very invasive, expensive and have a low rate of patient adherence.
-Therefore, there is a need for improved non-invasive methods to screen individuals.
-Gut microbiome-based biomarkers can be used as a non-invasive screening method.
-
-Patients with colorectal cancer have different stool community of microbes compared to adults with normal colons.
-This difference however cannot be explained by a single or a handful of candidate taxa in the gut microbiome but by many of them in relation to one another.
-Therefore, machine learning emerges as a tool to detect the difference between the gut microbiomes of CRC patients and healthy individuals.
-Previous studies have shown that human hemoglobin levels and bacterial population abundances in the stool help us predict screen relevant growth in the colon, however the literature for the problem of classifying colorectal disease status vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014].
-The variation in classification performance is based in part on differences in the task definition, in part on differences in the study populations, and in part on the learning pipeline.
-In this study, classification pipelines with L2-regularized logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM , decision tree, random forest and XGBoost classifiers are established.
-The generalization and prediction performance of these classifiers are evaluated and each classifier is examined for its reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting.
+Patients with colorectal cancer have different stool community of microbes than adults with normal colons (10–15).
+This difference however cannot be explained by a single species in the gut but by many of them in relation to one another.
+Therefore, machine learning emerged as a tool to identify the microbial biomarkers of CRC.
+Previous studies that used cohorts of 90 (30 SRN, 30 adenomas, 30 normal colons) (7) and 490 (229 SRN, 89 adenoma, 172 normal colons) patients (5) suggest that machine learning models can predict SRN.
+However, the prediction performance of previous models varies greatly.
+There is a need for implementing consistent and transparent machine learning practices to generate reproducible and replicable CRC biomarker models.
+This study aims to determine whether a reproducible and replicable classification model that incorporate the relative abundance of fecal bacterial populations can be developed to discriminate between individuals with and without, SRN.
 
@@
 Stool bacterial population abundances and stool hemoglobin levels of `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` Normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples were used to learn binary classifiers and evaluate their performances.
-The results show that ......
-(Is there a maximum threshold of prediction with all these methods?
-Does an increase in model complexity improve predictability?)
+Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models.
+The mean AUCs of these models were `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, and `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively.
 

### 2019-01-09 submission/ML_outline.Rmd
for Pat

---
title: "Machine Learning Manuscript Outline"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

##### General Context of the work
As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and phenotypes. Currently, the field’s use of machine learning lacks clarity and consistency.

##### Narrower research area and statement of its importance

* One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers. Colorectal cancer is one of the leading cause of death among cancers in the United States. Early diagnosis increases the chance of survival. However the current diagnostic methods are expensive and invasive. As a less invasive tool,  studies use relative abundances of the gut bacteria populations to predict disease progression.

* Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small. It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features.

##### Identification of a gap or other need for research
* ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc. There is a need for guidance on how to properly implement these different methods.
* The prediction performance of previous CRC microbiome models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9.

##### Summary of appraoch to answer the research question
* We establish a non-leaky pipeline. 
* We perform L2-regularized logistic regression, Linear L1 and L2 SVM, Non-Linear SVM, Decision tree, Random forest, XGBoost. 
* We evaluate the classification success of different machine learning methods and discuss the reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting of each method.

##### Announcement of principal findings
* Prediction perfomance of each model. 
* Is there a maximum threshold of prediction with all these methods?
* Does an increase in model complexity improve predictibility?
* Synthesis statement regarding modeling 16S microbiome data

## Methods

##### Brief explanation of study design/patient sampling and 16S rRNA gene sequencing/curation

##### Analysis of data
* What are the features and what are the labels?
We used the fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples to classify 490 patients as having advanced tumors (advanced adenoma or carinoma) or not (non-advanced adenoma or normal colon). 
* What is the data (temporal or not)? Assumptions we make when we use a dataset. What will the future data look like?
* Pre-proccessing of the data
* Machine Learning pipeline backbone. How do I split, train, validate and test?
A diagram to explain the modeling pipeline.
* Which methods are linear/non-linear? Talk about interpretibility.
* Cross-validation and hyper-parameter tuning methods for each modeling method
* Programming languages and packages/modules utilized
* Statistical methods for comparison of model performance

## Results

##### AUC results of 7 modeling approaches. (FIT + OTUs)

##### Comparisons among the 7 modeling approaches. (FIT + OTUs)

* Compare prediction performance, generalization performance and susceptibility to overfitting.

##### Hyper-parameter tuning budgets and corresponding AUC values.

* This will show that we have used the right budgets to allow the model to pick the right hyper-parameter.

##### AUC results of models with just (FIT)

* This could be in supplemental. We want to make sure that using just FIT as a feature does worse than FIT+OTUs.

## Discussion

##### Interpretation of modeling results in terms of reproducibility, robustness, actionability, interpretibility and susceptibility

* What are the metrics to talk about these concepts?

##### Consideration of possible weaknesses for each model

* The interactions between the biomarkers may be nonlinear. Obviously, the linear models will not incorporate this because they are linear. Tools like linear models (e.g. metastats, lefse, wilcoxon, etc) are likely worthless.

##### Consideration of possible weaknesses for our approach and chosen dataset

##### Relationship of results to previous literature and broader implications of this work

##### Prospects of future progress

### 2019-01-17 submission/ML_outline.Rmd
for Pat and Jenna

@@
 title: "Machine Learning Manuscript Outline"
-output: word_document
+bibliography: references.bib
+output:
+  pdf_document:
+    keep_tex: true
+    includes:
+      in_header: header.tex
+csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
+fontsize: 11pt
+geometry: margin=1.0in
 ---
@@
 ##### General Context of the work
-As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and phenotypes.
-Currently, the field’s use of machine learning lacks clarity and consistency.
+
+* As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
+
+* Most microbial communities are pretty patchy and the likelihood of a single species that explains the differences in health is pretty small.
+It is likely that subsets of those communities in relation to one another are needed to account for the patchiness as well as the context dependency of the features.
+
+* Thus, researchers have started to explore the utility of machine learning (ML) techniques.
+
+* Currently, the field’s use of machine learning lacks clarity and consistency on
+    1. Which methods are used?
+    2. Why those methods are used? Can these methods explain if there are linear or non-linear relationships between phenotype and features?
+    3. How those methods are used?
+    4. Reproducibilty
+
+* ML methods have a trade-off for interpretibility, susceptibility to overfitting and actionability versus prediction performance. It is important to prioritize and define task accordingly.
 
@@
 
-* One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
-Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Early diagnosis increases the chance of survival.
-However the current diagnostic methods are expensive and invasive.
-As a less invasive tool,  studies use relative abundances of the gut bacteria populations to predict disease progression.
+* Among cancers, colorectal cancer is one of the leading causes of death in the US.
+While colonoscopy is an effective screening tool, it is invasive and as a result has a low rate of patient adherence.
+In contrast, microbiome analysis of stool is non-invasive.
+Machine learning (ML) techniques are used to identify patients with colorectal tumors based on microbiota-associated biomarkers.
 
-* Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small.
-It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features.
+* However, the discriminative performance of these models varies greatly, with areas under the receiver operating characteristic curve (AUROC) of 0.7-0.9.
 
-##### Identification of a gap or other need for research
-* ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc.
-There is a need for guidance on how to properly implement these different methods.
-* The prediction performance of previous CRC microbiome models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9.
+* Previously:
+    - Differences in task definition (what is it that we want to predict?)
+    - Use of logistic regression without discussion of how it explains non-linearity of microbiome & CRC.
+    - Use of random forest only without discussion of interpretibility.
+    - Use of random forest without proper pipeline.
 
-##### Summary of appraoch to answer the research question
-* We establish a non-leaky pipeline. 
-* We perform L2-regularized logistic regression, Linear L1 and L2 SVM, Non-Linear SVM, Decision tree, Random forest, XGBoost. 
-* We evaluate the classification success of different machine learning methods and discuss the reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting of each method.
+* To shed light on how much differences in modeling can affect the results, we performed an empirical analysis comparing several different modeling pipelines.
 
-##### Announcement of principal findings
-* Prediction perfomance of each model. 
-* Is there a maximum threshold of prediction with all these methods?
-* Does an increase in model complexity improve predictibility?
-* Synthesis statement regarding modeling 16S microbiome data
+##### Summary of appraoch and findings
 
-## Methods
+* Modeling pipelines were established for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
 
-##### Brief explanation of study design/patient sampling and 16S rRNA gene sequencing/curation
+* These methods increase in complexity while decrease in interpretibility.
 
-##### Analysis of data
-* What are the features and what are the labels?
-We used the fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples to classify 490 patients as having advanced tumors (advanced adenoma or carinoma) or not (non-advanced adenoma or normal colon).
-* What is the data (temporal or not)? Assumptions we make when we use a dataset. What will the future data look like?
-* Pre-proccessing of the data
-* Machine Learning pipeline backbone. How do I split, train, validate and test?
-A diagram to explain the modeling pipeline.
-* Which methods are linear/non-linear? Talk about interpretibility.
-* Cross-validation and hyper-parameter tuning methods for each modeling method
-* Programming languages and packages/modules utilized
-* Statistical methods for comparison of model performance
+* We established ML pipeline with held-out test data and performed 100 data-splits to evaluate generalization and predicton performance of the ML method.
+
+* Applied to held-out test data, the mean AUROC varied from 0.68 (std ± 0.04) to 0.82 (std ± 0.04).
+
+* Random Forest had the highest mean AUROC for detecting SRN and was less susceptible to overfitting compared to other methods.
+
+* Despite the lower mean AUROC value, the L1-regularized linear kernel SVM offered the greatest interpretability and stability.
+
+* In terms of computational efficiency, x trained the fastest, while y took the longest.
+
+* We found that cross-validation and testing AUROC could vary by as much as 0.06, highlighting the importance of a separate held-out test set for evaluation.
+
+* Aside from evaluating generalization and classification performance for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
 
@@
 
-##### AUC results of 7 modeling approaches. (FIT + OTUs)
+##### AUROC results of 7 modeling approaches. (FIT + OTUs)
 
@@
 
-##### Hyper-parameter tuning budgets and corresponding AUC values.
+##### Hyper-parameter tuning budgets and corresponding AUROC values.
 
@@
 
+* Emphasize trade-off results we observe in this study.
+
 ##### Consideration of possible weaknesses for each model
@@
 metastats, lefse, wilcoxon, etc) are likely worthless.
+How many OTUs would you find with these methods?
 
 ##### Consideration of possible weaknesses for our approach and chosen dataset
+
+* Why did we use only one dataset? Becuase we are comparing several methods on 1 dataset.
 
@@
 ##### Prospects of future progress
+
+## Methods
+
+##### Brief explanation of study design/patient sampling and 16S rRNA gene sequencing/curation
+
+* Baxter et al, 2016
+
+##### Analysis of data
+
+* What are the features and what are the labels?
+
+    1.  Features: Fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples
+
+    2.  Labels:  490 patients as having advanced tumors (advanced adenoma or carinoma) or not (non-advanced adenoma or normal colon).
+
+* What is the data (temporal or not)? Assumptions we make when we use a dataset. What will the future data look like?
+
+* Pre-proccessing of the data
+
+* Machine Learning pipeline backbone. How do I split, train, validate and test?
+
+      - A diagram to explain the modeling pipeline.
+
+* Which methods are linear/non-linear? Talk about interpretibility.
+
+* Cross-validation and hyper-parameter tuning methods for each modeling method
+
+* Programming languages and packages/modules utilized
+
+* Statistical methods for comparison of model performance
+
+* Code/data availability

### 2019-01-17 submission/ML_outline.tex
for Pat and Jenna

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Machine Learning Manuscript Outline},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Machine Learning Manuscript Outline}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\floatplacement{figure}{H} % Keep the figure up top of the page

\begin{document}
\maketitle

\subsection{Introduction}\label{introduction}

\subparagraph{General Context of the
work}\label{general-context-of-the-work}

\begin{itemize}
\item
  As the microbiome field continues to grow, there is an ever-increasing
  demand for reproducible methods for identifying associations between
  members of the microbiome and human health.
\item
  Most microbial communities are pretty patchy and the likelihood of a
  single species that explains the differences in health is pretty
  small. It is likely that subsets of those communities in relation to
  one another are needed to account for the patchiness as well as the
  context dependency of the features.
\item
  Thus, researchers have started to explore the utility of machine
  learning (ML) techniques.
\item
  Currently, the field's use of machine learning lacks clarity and
  consistency on

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Which methods are used?
  \item
    Why those methods are used? Can these methods explain if there are
    linear or non-linear relationships between phenotype and features?
  \item
    How those methods are used?
  \item
    Reproducibilty
  \end{enumerate}
\item
  ML methods have a trade-off for interpretibility, susceptibility to
  overfitting and actionability versus prediction performance. It is
  important to prioritize and define task accordingly.
\end{itemize}

\subparagraph{Narrower research area and statement of its
importance}\label{narrower-research-area-and-statement-of-its-importance}

\begin{itemize}
\item
  Among cancers, colorectal cancer is one of the leading causes of death
  in the US. While colonoscopy is an effective screening tool, it is
  invasive and as a result has a low rate of patient adherence. In
  contrast, microbiome analysis of stool is non-invasive. Machine
  learning (ML) techniques are used to identify patients with colorectal
  tumors based on microbiota-associated biomarkers.
\item
  However, the discriminative performance of these models varies
  greatly, with areas under the receiver operating characteristic curve
  (AUROC) of 0.7-0.9.
\item
  Previously:

  \begin{itemize}
  \tightlist
  \item
    Differences in task definition (what is it that we want to predict?)
  \item
    Use of logistic regression without discussion of how it explains
    non-linearity of microbiome \& CRC.
  \item
    Use of random forest only without discussion of interpretibility.
  \item
    Use of random forest without proper pipeline.
  \end{itemize}
\item
  To shed light on how much differences in modeling can affect the
  results, we performed an empirical analysis comparing several
  different modeling pipelines.
\end{itemize}

\subparagraph{Summary of appraoch and
findings}\label{summary-of-appraoch-and-findings}

\begin{itemize}
\item
  Modeling pipelines were established for L2-regularized logistic
  regression, L1 and L2 support vector machines (SVM) with linear and
  radial basis function kernels, a decision tree, random forest and
  XGBoost.
\item
  These methods increase in complexity while decrease in
  interpretibility.
\item
  We established ML pipeline with held-out test data and performed 100
  data-splits to evaluate generalization and predicton performance of
  the ML method.
\item
  Applied to held-out test data, the mean AUROC varied from 0.68 (std ±
  0.04) to 0.82 (std ± 0.04).
\item
  Random Forest had the highest mean AUROC for detecting SRN and was
  less susceptible to overfitting compared to other methods.
\item
  Despite the lower mean AUROC value, the L1-regularized linear kernel
  SVM offered the greatest interpretability and stability.
\item
  In terms of computational efficiency, x trained the fastest, while y
  took the longest.
\item
  We found that cross-validation and testing AUROC could vary by as much
  as 0.06, highlighting the importance of a separate held-out test set
  for evaluation.
\item
  Aside from evaluating generalization and classification performance
  for each of these models, this study established standards for
  modeling pipelines of microbiome-associated machine learning models.
\end{itemize}

\subsection{Results}\label{results}

\subparagraph{AUROC results of 7 modeling approaches. (FIT +
OTUs)}\label{auroc-results-of-7-modeling-approaches.-fit-otus}

\subparagraph{Comparisons among the 7 modeling approaches. (FIT +
OTUs)}\label{comparisons-among-the-7-modeling-approaches.-fit-otus}

\begin{itemize}
\tightlist
\item
  Compare prediction performance, generalization performance and
  susceptibility to overfitting.
\end{itemize}

\subparagraph{Hyper-parameter tuning budgets and corresponding AUROC
values.}\label{hyper-parameter-tuning-budgets-and-corresponding-auroc-values.}

\begin{itemize}
\tightlist
\item
  This will show that we have used the right budgets to allow the model
  to pick the right hyper-parameter.
\end{itemize}

\subparagraph{AUC results of models with just
(FIT)}\label{auc-results-of-models-with-just-fit}

\begin{itemize}
\tightlist
\item
  This could be in supplemental. We want to make sure that using just
  FIT as a feature does worse than FIT+OTUs.
\end{itemize}

\subsection{Discussion}\label{discussion}

\subparagraph{Interpretation of modeling results in terms of
reproducibility, robustness, actionability, interpretibility and
susceptibility}\label{interpretation-of-modeling-results-in-terms-of-reproducibility-robustness-actionability-interpretibility-and-susceptibility}

\begin{itemize}
\item
  What are the metrics to talk about these concepts?
\item
  Emphasize trade-off results we observe in this study.
\end{itemize}

\subparagraph{Consideration of possible weaknesses for each
model}\label{consideration-of-possible-weaknesses-for-each-model}

\begin{itemize}
\tightlist
\item
  The interactions between the biomarkers may be nonlinear. Obviously,
  the linear models will not incorporate this because they are linear.
  Tools like linear models (e.g.~metastats, lefse, wilcoxon, etc) are
  likely worthless. How many OTUs would you find with these methods?
\end{itemize}

\subparagraph{Consideration of possible weaknesses for our approach and
chosen
dataset}\label{consideration-of-possible-weaknesses-for-our-approach-and-chosen-dataset}

\begin{itemize}
\tightlist
\item
  Why did we use only one dataset? Becuase we are comparing several
  methods on 1 dataset.
\end{itemize}

\subparagraph{Relationship of results to previous literature and broader
implications of this
work}\label{relationship-of-results-to-previous-literature-and-broader-implications-of-this-work}

\subparagraph{Prospects of future
progress}\label{prospects-of-future-progress}

\subsection{Methods}\label{methods}

\subparagraph{Brief explanation of study design/patient sampling and 16S
rRNA gene
sequencing/curation}\label{brief-explanation-of-study-designpatient-sampling-and-16s-rrna-gene-sequencingcuration}

\begin{itemize}
\tightlist
\item
  Baxter et al, 2016
\end{itemize}

\subparagraph{Analysis of data}\label{analysis-of-data}

\begin{itemize}
\item
  What are the features and what are the labels?

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \item
    Features: Fecal hemoglobin concentration and 16S rRNA gene sequences
    from stool samples
  \item
    Labels: 490 patients as having advanced tumors (advanced adenoma or
    carinoma) or not (non-advanced adenoma or normal colon).
  \end{enumerate}
\item
  What is the data (temporal or not)? Assumptions we make when we use a
  dataset. What will the future data look like?
\item
  Pre-proccessing of the data
\item
  Machine Learning pipeline backbone. How do I split, train, validate
  and test?

  \begin{itemize}
  \tightlist
  \item
    A diagram to explain the modeling pipeline.
  \end{itemize}
\item
  Which methods are linear/non-linear? Talk about interpretibility.
\item
  Cross-validation and hyper-parameter tuning methods for each modeling
  method
\item
  Programming languages and packages/modules utilized
\item
  Statistical methods for comparison of model performance
\item
  Code/data availability
\end{itemize}

\end{document}

### 2019-01-17 submission/manuscript.Rmd
intro update

@@
 
-######################################################################
-#---Load .tsv data with CV and test AUC data generated in Python ----#
-######################################################################
-
-# Read in AUCs table generated from L2 logistic regression model for all samples:
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax scaler
-logit <- read.delim('../data/process/L2_Logistic_Regression.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="L2-Logistic Regression")
-
-logit_summary <- logit %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-cv_meanAUC <- logit_summary[1,2]
-cv_sdAUC <- logit_summary[1,3]
-test_meanAUC <- logit_summary[2,2]
-test_sdAUC <- logit_summary[2,3]
-
-# Read in AUCs table generated from l1 SVM linear kernel for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is Standard scaler
-l1svm <- read.delim('../data/process/L1_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="L1-SVM Linear")
-
-l1svm_summary <- l1svm %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from l2 SVM linear kernel for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is Standard scaler
-l2svm <- read.delim('../data/process/L2_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="L2-SVM Linear")
-
-l2svm_summary <- l2svm %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from  SVM RBF kernel for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is Standard scaler
-svmRBF <- read.delim('../data/process/SVM_RBF.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="SVM RBF")
-
-svmRBF_summary <- svmRBF %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from xgboost for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax (0-1) scaler
-xgboost <- read.delim('../data/process/XGBoost.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="XGBoost")
-
-xgboost_summary <- xgboost %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-
-# Read in AUCs table generated from random forest for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax (0-1) scaler
-rf <- read.delim('../data/process/Random_Forest.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="Random Forest ")
-
-rf_summary <- rf %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
-# Read in AUCs table generated from decision tree for all samples
-#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
-#       FIT is a feature
-#       The scaler is MinMax (0-1) scaler
-dt <- read.delim('../data/process/Decision_Tree.tsv', header=T, sep='\t') %>%
-  select(level_1, AUC) %>% 
-  rename(Performance = level_1) %>% 
-  mutate(model="Decision Tree")
-
-dt_summary <- dt %>%  
-  group_by(Performance) %>% 
-  summarise(meanAUC=mean(AUC), std=sd(AUC))
 ```
@@
 
-As the microbiome field continues to grow, there is an ever-increasing demand for reproducible machine learning methods for identifying associations between members of the microbiome and phenotypes.
-Currently, the field’s use of machine learning lacks clarity and consistency.
-There is a need to implement good machine learning practices to generate reproducible and robust models.
-One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
-Colorectal cancer is one of the leading cause of death among cancers in the United States.
-Colonoscopy as a screening tool is effective, however it is invasive and have a low rate of patient adherence.
-Previous studies have shown that bacterial abundances in the stool can predict colorectal tumors in the colon and can be used as a non-invasive screening tool.
-However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014].
-We used the fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples to classify `r paste(nrow(meta))` patients as having advanced tumors or not.
-Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and Extreme Gradient Boosting (XGBoost) binary classification models.
-Tree-based methods, namely Decision Tree, Random Forest and XGBoost were less susceptible to overfitting and in general had higher sensitivity and specificity for advanced tumors.
-Aside from evaluating generalization and classification performance of each model, this study established standards for modeling pipelines of microbiome-associated machine learning models.
-
 \newpage
@@
 
-Colorectal cancer is the second most commonly occurring cancer in men and third in women, with 1.8 million new cases in 2018 (8).
-Early detection of screen-relevant neoplasias (SRN) which are cancerous or advanced precancerous lesions, is crucial for treatment of colorectal cancer.
-Patients whose lesions are detected at an early stage have a greater than 90 % chance of survival, yet a significant proportion of the US population remain unscreened (9).
-The high cost and invasiveness of screening methods such as colonoscopy and sigmoidoscopy lower patient adherence.
-There is a need for non-invasive screening tests that have high sensitivity and specificity for SRN.
+Advances in sequencing technology and decreasing costs of generating multitudes of 16S rRNA gene sequences allowed rapid exploration and taxonomic characterization of the human associated microbiome.
+Currently, the microbiome field is growing at an unprecedented rate and with it, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
+Human associated microbial communities are remarkably uneven and it is unlikely that a single species can explain a disease state universally.
+It is more likely that subsets of those communities, in relation to one another, reliably account for the unevenness and be used as biomarkers of a specific disease state.
+Thus, researchers have started to explore the utility of machine learning (ML) techniques to identify the biomarkers associated with healty and diseased individuals.
+However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are reproducible.
+Moreover, there is a lack of consideration on why a particular method is utilized.
 
@@
 ## Results and Discussion
-
-##### Results of modeling in text, tables and figures
-
-##### Comparisons among modeling approaches
-
-#### Interpretation of modeling results in terms of reproducibility, robustness, actionability, interpretibility and susceptibility
-
-#### Consideration of possible weaknesses for each model
-
-The interactions between the biomarkers may be nonlinear.
-Obviously, the linear models will not incorporate this because they are linear.
-Tools like linear models (e.g.
-metastats, lefse, wilcoxon, etc) are likely worthless.
-
-#### Consideration of possible weaknesses for our approach and chosen dataset
-
-#### Relationship of results to previous literature and broader implications of this work
-
-#### Prospects of future progress
 

### 2019-01-23 submission/manuscript.Rmd
add 2nd par

@@
 
-Advances in sequencing technology and decreasing costs of generating multitudes of 16S rRNA gene sequences allowed rapid exploration and taxonomic characterization of the human associated microbiome.
-Currently, the microbiome field is growing at an unprecedented rate and with it, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
-Human associated microbial communities are remarkably uneven and it is unlikely that a single species can explain a disease state universally.
-It is more likely that subsets of those communities, in relation to one another, reliably account for the unevenness and be used as biomarkers of a specific disease state.
-Thus, researchers have started to explore the utility of machine learning (ML) techniques to identify the biomarkers associated with healty and diseased individuals.
+Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences allowed rapid exploration and taxonomic characterization of the human associated microbiome.
+Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
+Human associated microbial communities are remarkably uneven and it is unlikely that a single species can explain a disease state comprehensively.
+It is more likely that subsets of those communities, in relation to one another and to their host, can reliably account for the unevenness and be used as biomarkers of a specific disease state.
+Thus, researchers have started to explore the utility of machine learning (ML) techniques to identify the biomarkers associated with healthy and diseased individuals.
 However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are reproducible.
 Moreover, there is a lack of consideration on why a particular method is utilized.
+Often, there is a trade-off between the interpretibility and accuracy of ML methods.
+As researchers, we have to find the right balance to accomplish complex knowledge tasks such as identifying a disease state based on microbiome-associated biomarkers while also understanding how these tasks are accomplished.
+
+One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
+Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a microbiome assay as a non-invasive screening tool would be a useful innovation.
+However, the efforts to combine ML methods with relative abundances of bacterial poplations in stool to predict CRC tumors have been afflicted by the aforementioned flawed ML practices.
+As a result, the predictive performance of these models varies greatly with areas under the receiver operating characteristic curve (AUROC) of 0.7-0.9.
 

### 2019-02-01 submission/ML_outline.Rmd
jenna notes

@@
 
-* As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
-
-* Most microbial communities are pretty patchy and the likelihood of a single species that explains the differences in health is pretty small.
-It is likely that subsets of those communities in relation to one another are needed to account for the patchiness as well as the context dependency of the features.
-
-* Thus, researchers have started to explore the utility of machine learning (ML) techniques.
-
-* Currently, the field’s use of machine learning lacks clarity and consistency on
-    1. Which methods are used?
-    2. Why those methods are used? Can these methods explain if there are linear or non-linear relationships between phenotype and features?
-    3. How those methods are used?
-    4. Reproducibilty
-
-* ML methods have a trade-off for interpretibility, susceptibility to overfitting and actionability versus prediction performance. It is important to prioritize and define task accordingly.
+Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences allowed rapid exploration and taxonomic characterization of the human associated microbiome.
+Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
+Human associated microbial communities are remarkably uneven and it is unlikely that a single species can explain a disease state comprehensively.
+It is more likely that subsets of those communities, in relation to one another and to their host, account for the unevenness and can be used as biomarkers of a specific disease state.
+Thus, researchers have started to explore the utility of machine learning (ML) techniques to identify the biomarkers associated with healthy and diseased individuals.
+However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are reproducible.
+Moreover, there is a lack of consideration on why a particular method is utilized.
+Often, there is a trade-off between the interpretibility and accuracy of ML methods.
+As researchers, we have to find the right balance to accomplish complex knowledge tasks such as identifying a disease state based on microbiome-associated biomarkers while also understanding how these tasks are accomplished.
 
@@
 
-* Among cancers, colorectal cancer is one of the leading causes of death in the US.
-While colonoscopy is an effective screening tool, it is invasive and as a result has a low rate of patient adherence.
-In contrast, microbiome analysis of stool is non-invasive.
-Machine learning (ML) techniques are used to identify patients with colorectal tumors based on microbiota-associated biomarkers.
+One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
+However, the efforts to combine ML methods with relative abundances of bacterial poplations in stool to predict CRC tumors have been negatively effected by the aforementioned flawed ML practices.
+As a result, the predictive performance of these models varies greatly with areas under the receiver operating characteristic curve (AUROC) of 0.7-0.9.
 
-* However, the discriminative performance of these models varies greatly, with areas under the receiver operating characteristic curve (AUROC) of 0.7-0.9.
-
-* Previously:
+* What we see in previous studies:
     - Differences in task definition (what is it that we want to predict?)
-    - Use of logistic regression without discussion of how it explains non-linearity of microbiome & CRC.
     - Use of random forest only without discussion of interpretibility.
@@
 
-* To shed light on how much differences in modeling can affect the results, we performed an empirical analysis comparing several different modeling pipelines.
+* What we do here: 
+    - To shed light on how much differences in modeling can affect the results, we performed an empirical analysis comparing several different modeling pipelines.
 

### 2019-02-01 submission/ML_outline.tex
jenna notes

@@
 
-\begin{itemize}
-\item
-  As the microbiome field continues to grow, there is an ever-increasing
-  demand for reproducible methods for identifying associations between
-  members of the microbiome and human health.
-\item
-  Most microbial communities are pretty patchy and the likelihood of a
-  single species that explains the differences in health is pretty
-  small. It is likely that subsets of those communities in relation to
-  one another are needed to account for the patchiness as well as the
-  context dependency of the features.
-\item
-  Thus, researchers have started to explore the utility of machine
-  learning (ML) techniques.
-\item
-  Currently, the field's use of machine learning lacks clarity and
-  consistency on
-
-  \begin{enumerate}
-  \def\labelenumi{\arabic{enumi}.}
-  \tightlist
-  \item
-    Which methods are used?
-  \item
-    Why those methods are used? Can these methods explain if there are
-    linear or non-linear relationships between phenotype and features?
-  \item
-    How those methods are used?
-  \item
-    Reproducibilty
-  \end{enumerate}
-\item
-  ML methods have a trade-off for interpretibility, susceptibility to
-  overfitting and actionability versus prediction performance. It is
-  important to prioritize and define task accordingly.
-\end{itemize}
+Advances in sequencing technology and decreasing costs of generating 16S
+rRNA gene sequences allowed rapid exploration and taxonomic
+characterization of the human associated microbiome. Currently, the
+microbiome field is growing at an unprecedented rate and as a result,
+there is an ever-increasing demand for reproducible methods for
+identifying associations between members of the microbiome and human
+health. Human associated microbial communities are remarkably uneven and
+it is unlikely that a single species can explain a disease state
+comprehensively. It is more likely that subsets of those communities, in
+relation to one another and to their host, account for the unevenness
+and can be used as biomarkers of a specific disease state. Thus,
+researchers have started to explore the utility of machine learning (ML)
+techniques to identify the biomarkers associated with healthy and
+diseased individuals. However, currently the field's use of ML lacks
+clarity and consistency on which methods are used, how these methods are
+implemented and if these methods are reproducible. Moreover, there is a
+lack of consideration on why a particular method is utilized. Often,
+there is a trade-off between the interpretibility and accuracy of ML
+methods. As researchers, we have to find the right balance to accomplish
+complex knowledge tasks such as identifying a disease state based on
+microbiome-associated biomarkers while also understanding how these
+tasks are accomplished.
 
@@
 
+One application of machine learning to microbiome data has been to
+classify patients as having colorectal tumors based on
+microbiota-associated biomarkers. However, the efforts to combine ML
+methods with relative abundances of bacterial poplations in stool to
+predict CRC tumors have been negatively effected by the aforementioned
+flawed ML practices. As a result, the predictive performance of these
+models varies greatly with areas under the receiver operating
+characteristic curve (AUROC) of 0.7-0.9.
+
 \begin{itemize}
+\tightlist
 \item
-  Among cancers, colorectal cancer is one of the leading causes of death
-  in the US. While colonoscopy is an effective screening tool, it is
-  invasive and as a result has a low rate of patient adherence. In
-  contrast, microbiome analysis of stool is non-invasive. Machine
-  learning (ML) techniques are used to identify patients with colorectal
-  tumors based on microbiota-associated biomarkers.
-\item
-  However, the discriminative performance of these models varies
-  greatly, with areas under the receiver operating characteristic curve
-  (AUROC) of 0.7-0.9.
-\item
-  Previously:
+  What we see in previous studies:
 
@@
   \item
-    Use of logistic regression without discussion of how it explains
-    non-linearity of microbiome \& CRC.
-  \item
     Use of random forest only without discussion of interpretibility.
@@
 \item
+  What we do here:
+
+  \begin{itemize}
+  \tightlist
+  \item
     To shed light on how much differences in modeling can affect the
@@
     different modeling pipelines.
+  \end{itemize}
 \end{itemize}

### 2019-02-28 submission/manuscript.Rmd
compare time

--- 2019-01-17 submission/manuscript.Rmd
@@
 output:
-  word_document: default
   pdf_document: 
@@
 ######################################################################
-deps = c("knitr","rmarkdown","vegan","gtools", "tidyverse");
+deps = c("reshape2","knitr","rmarkdown","vegan","gtools", "tidyverse");
 for (dep in deps){
@@
 knitr::knit_hooks$set(inline=inline_hook)
-library(reticulate)
-use_python("/Library/Frameworks/Python.framework/Versions/3.6/bin/python3", required = TRUE)
 ```
@@
 ```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
+######################################################################
+# Load in needed functions and libraries
+melt_data <-  function(data) {
+  data_melt <- data %>%
+    melt(measure.vars=c('cv_aucs', 'test_aucs')) %>%
+    rename(AUC=value) %>%
+    mutate(Performance = case_when(variable == "cv_aucs" ~ 'cross-validation', variable == "test_aucs" ~ 'testing')) %>%
+    group_by(Performance)
+  return(data_melt)
+}
+
+# Read in files as delim that are saved in a list with a pattern
+read_files <- function(filenames){
+  for(file in filenames){
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',')
+  }
+  return(data)
+}
+
+# Get model name with sub from file name
+get_model_name <- function(files){
+  pat1 <- "../data/process/walltime_"
+  name_files <- sub(pat1, "", files)
+  pat2 <- ".csv"
+  names <- sub(pat2, "", name_files)
+  return(names)
+}
+######################################################################
+
 ######################################################################
@@
 
-```
+######################################################################
+#--------- Load .csv datato get mean test AUC for 7 models--------#
+######################################################################
 
-```{python, cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
-import platform
-python_version = platform.python_version()
-import sklearn
-sklearn_module = sklearn.__version__
+# Read in the cvAUCs, testAUCs for 100 splits.
+best_files <- list.files(path= '../data/process', pattern='combined_best.*', full.names = TRUE)
+logit <- read_files(best_files[4])
+l2svm <- read_files(best_files[3])
+l1svm <- read_files(best_files[2])
+rbf <- read_files(best_files[6])
+rf <- read_files(best_files[5])
+dt <- read_files(best_files[1])
+xgboost <- read_files(best_files[7])
+
+# Bind them together and summarise mean testAUC by model
+test_all <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
+  melt_data() %>% 
+  select(-variable) %>% 
+  group_by(model, Performance) %>% 
+  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
+  filter(Performance=="testing")
+
+# Get the order index from small to large meanAUC
+performance_index <- order(test_all$mean_AUC)
+
+# Bind them together and summarise mean cvAUC by model
+cv_all <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
+  melt_data() %>% 
+  select(-variable) %>% 
+  group_by(model, Performance) %>% 
+  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
+  filter(Performance=="cross-validation")
+
+# Get the difference between mean cvAUC and testAUC for each model
+difference <- cv_all$mean_AUC - test_all$mean_AUC[match(cv_all$model, test_all$model)]
+difference_model <- data.frame(difference, test_all$model)
+# Get the order index from small to large of the differences
+difference_index <- order(difference_model$difference)
+
+######################################################################
+#---------- Load .csv datato get mean walltime for 7 models-------#
+######################################################################
+
+# Read in the walltime for each split.
+walltime_files <- list.files(path= '../data/process', pattern='walltime*', full.names = TRUE)
+
+# ------- 1. In a loop read the files as delim
+# --------2. get the model name for each with get_model_name() 
+#---------3. Add a column with model name to each delim
+result <- list()
+for(file in walltime_files){
+  model_walltime <- read_files(file) %>%  
+  #summarise_walltime() %>% 
+  mutate(model=get_model_name(file))
+  result[[length(result)+1]] <- model_walltime
+}
+
+# ------- 1. If models are L1 or L2 Linear SVM or L2 Logit then the walltime is in minutes
+# --------2. If models are Random Forest or XGBoost then the walltime is in days
+#---------3. If other models, the walltime is in hours.
+#---------4. This loop converts all to hours.
+min_fixed_result <- list()
+for(i in result){
+  if((sum(i$model=="L1_Linear_SVM")==100 || sum(i$model=="L2_Linear_SVM")==100 || sum(i$model=="L2_Logistic_Regression")==100)){
+    i$x <- i$x/60 # The walltimes were saved as minutes # We convert these to hours by diving with 60
+  }
+  if((sum(i$model=="Random_Forest")==100 || sum(i$model=="XGBoost")==100)){
+    i$x <- i$x*24 # The walltimes were saved as minutes # We convert these to hours by diving with 60
+  }
+  else{
+  }
+  min_fixed_result[[length(min_fixed_result)+1]] <- i
+}
+
+# Bind all delims with model name and walltime together
+# Summarise mean walltime
+walltime_df <- bind_rows(min_fixed_result) %>% 
+  group_by(model) %>% 
+  summarise(mean_walltime=mean(x), sd_walltime=sd(x))
+
+# Get the sorted mean walltime for each model small to large
+walltime_index <- order(walltime_df$mean_walltime)
+
 ```
@@
 
-Advances in sequencing technology and decreasing costs of generating multitudes of 16S rRNA gene sequences allowed rapid exploration and taxonomic characterization of the human associated microbiome.
-Currently, the microbiome field is growing at an unprecedented rate and with it, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
-Human associated microbial communities are remarkably uneven and it is unlikely that a single species can explain a disease state universally.
-It is more likely that subsets of those communities, in relation to one another, reliably account for the unevenness and be used as biomarkers of a specific disease state.
-Thus, researchers have started to explore the utility of machine learning (ML) techniques to identify the biomarkers associated with healty and diseased individuals.
-However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are reproducible.
-Moreover, there is a lack of consideration on why a particular method is utilized.
+Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome.
+Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
+However, this is an undertaking as human associated microbial communities are remarkably uneven.
+It is unlikely that a single species can explain a disease state comprehensively.
+It is more likely that subsets of those communities, in relation to one another and to their host, account for different disease states.
+Thus, researchers have started to explore the utility of machine learning (ML) techniques to develop predictive models that use microbiota associated biomarkers to identify healthy and diseased states.
+However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are robust and reproducible.
+Moreover, there is a lack of consideration on why a particular ML model is utilized.
+There is a wide range of complexity and interpretibility in models used in microbiome field [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @pasolli_machine_2016]; recently with many that are black boxes which cannot be interpreted by humans [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017].
+ML model choice should reflect the goal of the study; whether it is to develop the best predictive model or to understand the ecology of microbiota associated diseases.
 
-Patients with colorectal cancer have different stool community of microbes than adults with normal colons (10–15).
-This difference however cannot be explained by a single species in the gut but by many of them in relation to one another.
-Therefore, machine learning emerged as a tool to identify the microbial biomarkers of CRC.
-Previous studies that used cohorts of 90 (30 SRN, 30 adenomas, 30 normal colons) (7) and 490 (229 SRN, 89 adenoma, 172 normal colons) patients (5) suggest that machine learning models can predict SRN.
-However, the prediction performance of previous models varies greatly.
+One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
+Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a predictive model from a stool microbiome assay as a non-invasive screening tool is a remarkable innovation.
+However, the efforts to combine ML methods with relative abundances of bacterial poplations in stool to predict CRC tumors have been afflicted by flawed ML methods and by the lack of discussion on modeling choices.
+In previous studies, ...
 There is a need for implementing consistent and transparent machine learning practices to generate reproducible and replicable CRC biomarker models.
-This study aims to determine whether a reproducible and replicable classification model that incorporate the relative abundance of fecal bacterial populations can be developed to discriminate between individuals with and without, SRN.
+This study aims to shed light on how much task definition and model choices can affect the results.
+Here we performed an empirical analysis comparing several different modeling pipelines using the same dataset.
 
-Here, colonic disease status is defined as Normal or Screen Relevant Neoplasias (SRN).
-Stool bacterial population abundances and stool hemoglobin levels of `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` Normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples were used to learn binary classifiers and evaluate their performances.
-Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models.
-The mean AUCs of these models were `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, and `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively.
+As our dataset, we used a previously published CRC study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and immunochemical test (FIT) results from `r paste(nrow(shared))` patients.
+We built ML models using 16S abundances and FIT to predict patients with normal or screen relevant neoplasias (SRN) disease status.
+The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
+We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
+These models increase in complexity while they decrease in interpretability.
+Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model.
+The mean AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`).
+Random Forest and XGBoost had the highest mean AUROC for detecting SRN.
+In terms of computational efficiency, L2 Logistic Regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
+We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation.
+Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
 
 ## Results and Discussion
+
+__The prediction and generalization performance of classifiers during cross-validation and when applied to the held-out test data.__
+
+We evaluated the prediction performance of seven binary classification models when applied to held-out test data using AUROC as our metric.
+Random Forest and XGBoost had the highest mean AUROC for detecting SRN, `r paste(round(mean(rf$test_aucs), 3))` (std ± `r paste(round(sd(rf$test_aucs), 3))`) and `r paste(round(mean(xgboost$test_aucs), 3))` (std ± `r paste(round(sd(xgboost$test_aucs), 3))`) respectively [Figure 2].
+L1 linear SVM and decision tree had significantly lower AUROC values, `r paste(round(mean(l1svm$test_aucs), 3))` (std ± `r paste(round(sd(l1svm$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`)  [Figure 2].
+However, they had significanly higher performances than L2 linear SVM, RBF SVM and L2 logistic regression which had mean AUROC values of `r paste(round(mean(l2svm$test_aucs), 3))` (std ± `r paste(round(sd(l2svm$test_aucs), 3))`),  `r paste(round(mean(rbf$test_aucs), 3))` (std ± `r paste(round(sd(rbf$test_aucs), 3))`) and `r paste(round(mean(logit$test_aucs), 3))` (std ± `r paste(round(sd(logit$test_aucs), 3))`) respectively [Figure 2].
+We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC.
+We found a statistically significant difference for classifiers L2 support vector machines (SVM) with linear and radial basis function kernels and L2 logistic regression.
+These differences were `r paste(round(difference_model[3,1], 3))`, `r paste(round(difference_model[6,1], 3))` and `r paste(round(difference_model[4,1], 3))`, respectively [Figure 2] .
+
+__Hyperparameter optimization for each classifier.__
+
+__The interpretation of classifiers.__
 
@@
 FIT levels are continuous data present for each sample.
-Because the data are in different scales, Python programming language v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` is used to transform features by scaling each feature to a [0-1] range (Table 1) [@scikit-learn].
+Because the data are in different scales, Python programming language v, module scikit-learn v is used to transform features by scaling each feature to a [0-1] range (Table 1) [@scikit-learn].
 
@@
 Since the colonic disease status are not uniformly represented in the data, 5-fold splits are stratified to maintain the overall label distribution on the training set.
-Python programming language v`r paste(py$python_version)`, module scikit-learn v`r paste(py$sklearn_module)` functions are used to learn the seven classifiers (Table 1).
 

### 2019-02-28 submission/manuscript.tex
compare time

@@
 \usepackage{pdfpages}
+\floatplacement{figure}{H} % Keep the figure up top of the page
 
@@
 
-As gut microbiome field continues to grow, there will be an
-ever-increasing demand for reproducible machine learning methods to
-analyze 16S rRNA gene sequence data and to determine association of the
-microbiome with a continuous or categorical phenotype of interest. The
-use of machine learning in microbiome literature lack clarity over the
-learning pipeline which spans the problem formulation, feature
-selection, feature pre-processing, model learning, and output. There is
-a need for guidance on how to properly implement good machine learning
-practices to generate reproducible, robust and actionable models. There
-is also a clinical need to generate interpretable models for biomedical
-researchers and clinicans to adopt and use regularly (1).
+Advances in sequencing technology and decreasing costs of generating 16S
+rRNA gene sequences have allowed rapid exploration and taxonomic
+characterization of the human associated microbiome. Currently, the
+microbiome field is growing at an unprecedented rate and as a result,
+there is an ever-increasing demand for reproducible methods for
+identifying associations between members of the microbiome and human
+health. However, this is an undertaking as human associated microbial
+communities are remarkably uneven. It is unlikely that a single species
+can explain a disease state comprehensively. It is more likely that
+subsets of those communities, in relation to one another and to their
+host, account for different disease states. Thus, researchers have
+started to explore the utility of machine learning (ML) techniques to
+develop predictive models that use microbiota associated biomarkers to
+identify healthy and diseased states. However, currently the field's use
+of ML lacks clarity and consistency on which methods are used, how these
+methods are implemented and if these methods are robust and
+reproducible. Moreover, there is a lack of consideration on why a
+particular ML model is utilized. There is a wide range of complexity and
+interpretibility in models used in microbiome field (1--5); recently
+with many that are black boxes which cannot be interpreted by humans
+(6--8). ML model choice should reflect the goal of the study; whether it
+is to develop the best predictive model or to understand the ecology of
+microbiota associated diseases.
 
-Recently, there is an interest on using machine learning to predict
-colorectal cancer progression with microbiota-assiciated biomarkers.
-Colorectal cancer is one of the leading cause of death among cancers in
-the United States. Each person in the industrialized world has on
-average a one-in-twenty chance of developing colorectal cancer (CRC) in
-their lifetime and once diagnosed, more than one-third will not survive
-5 years (2--4). Colonoscopy as a screening tool is very effective,
-however it is very invasive, expensive and have a low rate of patient
-adherence. Therefore, there is a need for improved non-invasive methods
-to screen individuals. Gut microbiome-based biomarkers can be used as a
-non-invasive screening method.
+One application of machine learning to microbiome data has been to
+classify patients as having colorectal tumors based on
+microbiota-associated biomarkers. Colorectal cancer (CRC) is one of the
+leading causes of death in the US, therefore developing a predictive
+model from a stool microbiome assay as a non-invasive screening tool is
+a remarkable innovation. However, the efforts to combine ML methods with
+relative abundances of bacterial poplations in stool to predict CRC
+tumors have been afflicted by flawed ML methods and by the lack of
+discussion on modeling choices. In previous studies, \ldots{} There is a
+need for implementing consistent and transparent machine learning
+practices to generate reproducible and replicable CRC biomarker models.
+This study aims to shed light on how much task definition and model
+choices can affect the results. Here we performed an empirical analysis
+comparing several different modeling pipelines using the same dataset.
 
-Patients with colorectal cancer have different stool community of
-microbes compared to adults with normal colons. This difference however
-cannot be explained by a single or a handful of candidate taxa in the
-gut microbiome but by many of them in relation to one another.
-Therefore, machine learning emerges as a tool to detect the difference
-between the gut microbiomes of CRC patients and healthy individuals.
-Previous studies have shown that human hemoglobin levels and bacterial
-population abundances in the stool help us predict screen relevant
-growth in the colon, however the literature for the problem of
-classifying colorectal disease status vary greatly, with areas under the
-receiver operating characteristic curve (AUC) of 0.7-0.9 (5--8). The
-variation in classification performance is based in part on differences
-in the task definition, in part on differences in the study populations,
-and in part on the learning pipeline. In this study, classification
-pipelines with L2-regularized logistic regression, L1 and L2 linear
-suppor vector machines (SVM), radial basis function SVM , decision tree,
-random forest and XGBoost classifiers are established. The
-generalization and prediction performance of these classifiers are
-evaluated and each classifier is examined for its reproducibility,
-robustness, actionability, interpretibility and susceptibility to
-overfitting.
-
-Here, colonic disease status is defined as Normal or Screen Relevant
-Neoplasias (SRN). Stool bacterial population abundances and stool
-hemoglobin levels of 261 Normal and 229 SRN samples were used to learn
-binary classifiers and evaluate their performances. The results show
-that \ldots{}\ldots{} (Is there a maximum threshold of prediction with
-all these methods? Does an increase in model complexity improve
-predictibility?)
+As our dataset, we used a previously published CRC study (3) which had
+fecal 16S rRNA gene sequences and immunochemical test (FIT) results from
+490 patients. We built ML models using 16S abundances and FIT to predict
+patients with normal or screen relevant neoplasias (SRN) disease status.
+The study had 261 normal and 229 SRN samples. We established modeling
+pipelines for L2-regularized logistic regression, L1 and L2 support
+vector machines (SVM) with linear and radial basis function kernels, a
+decision tree, random forest and XGBoost. These models increase in
+complexity while they decrease in interpretability. Our ML pipeline
+performed 100 data-splits and utilized held-out test data to evaluate
+generalization and predicton performance of each ML model. The mean
+AUROC varied from 0.67 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
+and XGBoost had the highest mean AUROC for detecting SRN. In terms of
+computational efficiency, L2 Logistic Regression trained the fastest
+(0.098 hours, std ± 0.008), while XGBoost took the longest (3.601 hours,
+std ± 0.524). We found that mean cross-validation and testing AUROC
+could vary by as much as 0.062, which highlights the importance of a
+seperate held-out test set for evaluation. Aside from evaluating
+generalization and classification performances for each of these models,
+this study established standards for modeling pipelines of
+microbiome-associated machine learning models.
 
@@
 
-\subparagraph{Results of modeling in text, tables and
-figures}\label{results-of-modeling-in-text-tables-and-figures}
+\textbf{The prediction and generalization performance of classifiers
+during cross-validation and when applied to the held-out test data.}
 
-\subparagraph{Comparisons among modeling
-approaches}\label{comparisons-among-modeling-approaches}
+We evaluated the prediction performance of seven binary classification
+models when applied to held-out test data using AUROC as our metric.
+Random Forest and XGBoost had the highest mean AUROC for detecting SRN,
+0.818 (std ± 0.036) and 0.814 (std ± 0.038) respectively {[}Figure 2{]}.
+L1 linear SVM and decision tree had significantly lower AUROC values,
+0.748 (std ± 0.043) and 0.741 (std ± 0.038) {[}Figure 2{]}. However,
+they had significanly higher performances than L2 linear SVM, RBF SVM
+and L2 logistic regression which had mean AUROC values of 0.67 (std ±
+0.054), 0.683 (std ± 0.048) and 0.677 (std ± 0.053)
+respectively{[}Figure 2{]}. We also evaluated the generalization
+performance of each classifier by comparing their mean cross-validation
+AUROC and mean testing AUROC. We found a significant difference for
+classifiers L2 support vector machines (SVM) with linear and radial
+basis function kernels and L2 logistic regression. These differences
+were 0.062, 0.047 and 0.061, respectively {[}Figure 2{]} .
 
-\paragraph{Interpretation of modeling results in terms of
-reproducibility, robustness, actionability, interpretibility and
-susceptibility}\label{interpretation-of-modeling-results-in-terms-of-reproducibility-robustness-actionability-interpretibility-and-susceptibility}
+\textbf{The prediction performance of classifiers for different
+hyperparameter settings.}
 
-\paragraph{Consideration of possible weaknesses for each
-model}\label{consideration-of-possible-weaknesses-for-each-model}
-
-The interactions between the biomarkers may be nonlinear. Obviously, the
-linear models will not incorporate this because they are linear. Tools
-like linear models (e.g.~metastats, lefse, wilcoxon, etc) are likely
-worthless.
-
-\paragraph{Consideration of possible weaknesses for our approach and
-chosen
-dataset}\label{consideration-of-possible-weaknesses-for-our-approach-and-chosen-dataset}
-
-\paragraph{Relationship of results to previous literature and broader
-implications of this
-work}\label{relationship-of-results-to-previous-literature-and-broader-implications-of-this-work}
-
-\paragraph{Prospects of future
-progress}\label{prospects-of-future-progress}
+\textbf{The interpretation of classifiers.}
 
@@
 Great Lakes-New England Early Detection Research Network study. These
-data were obtained from Sze et al (5). The stool samples were provided
+data were obtained from Sze et al (9). The stool samples were provided
 by recruited adult participants who were undergoing scheduled screening
-or surveillance colonoscpy. Colonoscopies were performed and fecal
+or surveillance colonoscopy. Colonoscopies were performed and fecal
 samples were collected from participants in four locations: Toronto (ON,
@@
 sequences using VSEARCH and assigning to OTUs at 97\% similarity using
-the OptiClust algorithm (9--11).
+the OptiClust algorithm (10--12).
 
@@
 Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
-sample. FIT levels are continuous data present for each sample. These 2
-different data are in different scales. Python programming language
-v3.6.6, module scikit-learn v0.19.2 is used to transform features by
-scaling each feature to a {[}0-1{]} range (Table 1) (12).
+sample. FIT levels are continuous data present for each sample. Because
+the data are in different scales, Python programming language v, module
+scikit-learn v is used to transform features by scaling each feature to
+a {[}0-1{]} range (Table 1) (13).
 
@@
 are stratified to maintain the overall label distribution on the
-training set. v3.6.6, module scikit-learn v0.19.2 functions are used to
-learn the seven classifiers (Table 1).
+training set.
 
@@
 performances. The boxplot shows quartiles at the box ends and the
-statistical median as athe horizontal line in the box. The whiskers show
+statistical median as the horizontal line in the box. The whiskers show
 the farthest points that are not outliers. Outliers are data points that
@@
 \hypertarget{refs}{}
-\hypertarget{ref-wiens_editorial:_2016}{}
-1. \textbf{Wiens J}, \textbf{Wallace BC}. 2016. Editorial: Special issue
-on machine learning for health and medicine. Machine Learning
-\textbf{102}:305--307.
-doi:\href{https://doi.org/10.1007/s10994-015-5533-9}{10.1007/s10994-015-5533-9}.
+\hypertarget{ref-zeller_potential_2014}{}
+1. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
+\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
+\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
+\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
+\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
+J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
+\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
+\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
+for early-stage detection of colorectal cancer. Mol Syst Biol
+\textbf{10}.
+doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
 
-\hypertarget{ref-seer_2016}{}
-2. \textbf{Howlader N KM \textnormal{Noone AM}}. SEER cancer statistics
-review, 1975-2013, (national cancer institute. bethesda, md).
+\hypertarget{ref-zackular_human_2014}{}
+2. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
 
-\hypertarget{ref-street_colorectal_nodate}{}
-3. \textbf{Street W}. Colorectal cancer facts \& figures 2017-2019 40.
-
-\hypertarget{ref-weir_past_2015}{}
-4. \textbf{Weir HK}, \textbf{Thompson TD}, \textbf{Soman A},
-\textbf{MÃžller B}, \textbf{Leadbetter S}. 2015. The past, present, and
-future of cancer incidence in the united states: 1975 through 2020.
-Cancer \textbf{121}:1827--1837.
-doi:\href{https://doi.org/10.1002/cncr.29258}{10.1002/cncr.29258}.
-
-\hypertarget{ref-sze_leveraging_2018}{}
-5. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
-rRNA gene surveys to identify reproducible biomarkers in individuals
-with colorectal tumors. mBio \textbf{9}:e00630--18.
-doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
+\hypertarget{ref-baxter_dna_2016}{}
+3. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
+\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
+immunochemical test can replace stool for detection of colonic lesions
+using a microbiota-based model. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
 
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-6. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+4. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 
-\hypertarget{ref-baxter_dna_2016}{}
-7. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
-\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
-immunochemical test can replace stool for detection of colonic lesions
-using a microbiota-based model. Microbiome \textbf{4}.
-doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
+\hypertarget{ref-pasolli_machine_2016}{}
+5. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
+\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
+meta-analysis of large metagenomic datasets: Tools and biological
+insights. PLoS Comput Biol \textbf{12}.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.
 
-\hypertarget{ref-zackular_human_2014}{}
-8. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
-\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
-for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
-doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
+\hypertarget{ref-galkin_human_2018}{}
+6. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
+2018. Human microbiome aging clocks based on deep learning and tandem of
+permutation feature importance and accumulated local effects. bioRxiv.
+doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
+
+\hypertarget{ref-reiman_using_2017}{}
+7. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+convolutional neural networks to explore the microbiome, pp. 4269--4272.
+\emph{In} 2017 39th annual international conference of the IEEE
+engineering in medicine and biology society (EMBC).
+
+\hypertarget{ref-fioravanti_phylogenetic_2017}{}
+8. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
+\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
+in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
+
+\hypertarget{ref-sze_leveraging_2018}{}
+9. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+rRNA gene surveys to identify reproducible biomarkers in individuals
+with colorectal tumors. mBio \textbf{9}:e00630--18.
+doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
 
 \hypertarget{ref-schloss_introducing_2009}{}
-9. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+10. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-10. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+11. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-11. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+12. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-scikit-learn}{}
-12. \textbf{Pedregosa F}, \textbf{Varoquaux G}, \textbf{Gramfort A},
+13. \textbf{Pedregosa F}, \textbf{Varoquaux G}, \textbf{Gramfort A},
 \textbf{Michel V}, \textbf{Thirion B}, \textbf{Grisel O},

### 2019-02-28 code/learning/Figure1.R

@@ -15,7 +15,7 @@ source('code/learning/functions.R')
 
 
 ######################################################################
-# Load .tsv data generated with modeling pipeline 
+# Load .csv data generated with modeling pipeline 
 ######################################################################
 
 # Read in the cvAUCs, testAUCs for 100 splits.

### 2019-02-28 code/learning/Figure2.R

@@ -88,6 +88,29 @@ l2svm_plot <- base_plot(l2svm, l2svm$cost, l2svm$mean_Acc) +
                      breaks = seq(0.5, 1, 0.05)) +
   geom_errorbar(aes(ymin=mean_Acc-sd_Acc, ymax=mean_Acc+sd_Acc), width=.001)
 
+logit_plot <- logit_all %>%
+  group_by(cost, loss, epsilon) %>%
+  summarise(mean_AUC = mean(ROC), sd_AUC = sd(ROC)) %>%
+  ggplot(aes(x=cost,y=mean_AUC)) +
+  geom_line() +
+  geom_point() +
+  scale_x_continuous(name="C (penalty)") +
+  scale_y_continuous(name="L2 Logistic Regression mean cvAUC",
+                     limits = c(0.50, 1),
+                     breaks = seq(0.5, 1, 0.05)) +
+  geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.01) +
+  theme_bw() +
+  theme(legend.text=element_text(size=18),
+        legend.title=element_text(size=22),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        panel.background = element_blank(),
+        text = element_text(size = 12),
+        axis.text.x=element_text(size = 12, colour='black'),
+        axis.text.y=element_text(size = 12, colour='black'),
+        axis.title.y=element_text(size = 13),
+        axis.title.x=element_text(size = 13))
+
 dt <- dt_all %>%
   group_by(maxdepth) %>%
   summarise(mean_AUC = mean(ROC), sd_AUC = sd(ROC))
@@ -112,103 +135,24 @@ scale_x_continuous(name="mtry",
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=1)
 
 # Start plotting models with 2 hyper-parameters individually
+
 rbf_plot <- rbf_all %>%
   group_by(sigma, C) %>%
   summarise(mean_AUC = mean(ROC), sd_AUC = sd(ROC)) %>%
-  ggplot(aes(x=sigma, y=C) ) +
-  stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
-  scale_fill_distiller(palette="Greys", direction=1) +
-  scale_x_continuous(expand = c(0, 0)) +
[…]

### 2019-02-28 code/learning/compareInterp.R

@@ -36,22 +36,43 @@ read_files <- function(filenames){
   return(data)
 }
 
-get_interp_info <- function(model, model_name){ 
+get_interp_info <- function(data, model_name){ 
   if(model_name=="L2_Logistic_Regression" || 
      model_name=="L1_Linear_SVM" || 
      model_name=="L2_Linear_SVM")
       {
-      ranked_imp <- model %>% 
-      select(-normal) %>% 
-      group_by(names) %>% 
-      summarise(mean_imp = mean(cancer), sd_imp = sd(cancer), n = n()) %>% 
-      arrange(-n) %>% 
-      head(n=10)
+    # If the models are linear, we saved the weights of every OTU in the linear model for each datasplit
+      # 1. Get the mean of weigths for each OTU (OTUs are columns)
+      weights <- data %>% 
+        colMeans()
+      # 2. Create a dataframe where column 1 shows the OTU names and column 2 weights
+      means <- as.data.frame(weights) 
+      means$names <- rownames(means) 
+      # 3. We now want to save to a new column the sign of the weights
+      full <- means %>% 
+        mutate(sign = case_when(weights<0 ~ "negative",
+                                weights>0 ~ "positive")) 
+      # 4. We change all the weights to their absolute value
+      full$weights <- abs(full$weights)
+      # 5.  a) Order the dataframe from largest weights to smallest.
+      #     b) Select the largest 10 
+      #     c) Put the signs back to weights
+      imp_means <- full %>% 
+        arrange(desc(weights)) %>% 
+        head(n=10) %>% 
+        mutate(values = case_when(sign=="negative" ~ weights*-1,
+                                  sign=="positive"~ weights))
       }
   else{
-      ranked_imp <- model %>% 
+    # If the models are not linear, we saved variable importance of the top 10 variables per each datasplit
+      # We will group by the OTU names 
+      ranked_imp <- data %>% 
       group_by(names) %>% 
+      # We then get the mean of importance of each OTU and how many times that OTU was saved 
+        # How many data-splits actually chose that OTU as important
       summarise(mean_imp = mean(Overall), sd_imp = sd(Overall), n = n()) %>% 
+      # Order the dataframe by how many times the OTU was observed 
+        # Choose the top 10
       arrange(-n) %>% 
       head(n=10)
       }

### 2019-02-28 code/learning/model_pipeline.R

@@ -144,8 +144,13 @@ pipeline <- function(dataset, model){
         results_individual <- trained_model$results
         results_total <- rbind(results_total, results_individual)
     }
-  # Here we look at the top 10 important features
-  feature_importance <- model_interpret(trained_model)
+    # Here we look at the top 10 important features
+    if(model=="L1_Linear_SVM" || model=="L2_Linear_SVM" || model=="L2_Logistic_Regression"){
+      feature_importance <- trained_model$finalModel$W
+    }
+    else{
+      feature_importance <- model_interpret(trained_model)
+    }
   # Return all the metrics
   results <- list(cv_aucs, test_aucs, results_total, feature_importance, trained_model)
   return(results)

### 2019-02-28 code/learning/model_selection.R

@@ -109,7 +109,7 @@ tuning_grid <- function(model){
                          max_depth=8,
                          colsample_bytree= 0.8,
                          min_child_weight=1,
-                         subsample=c(0.5, 0.6, 0.7))
+                         subsample=c(0.4, 0.5, 0.6, 0.7))
     method <- "xgbTree"
   }
   else {

### 2019-03-01 submission/manuscript.Rmd

@@
 ---
-title: "**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**"
+title: "**Evaluation of machine learning methods for 16S rRNA gene data**"
 bibliography: references.bib
@@
 Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome.
-Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.
+Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for reproducible methods to identify associations between members of the microbiome and human host.
 However, this is an undertaking as human associated microbial communities are remarkably uneven.
-It is unlikely that a single species can explain a disease state comprehensively.
-It is more likely that subsets of those communities, in relation to one another and to their host, account for different disease states.
-Thus, researchers have started to explore the utility of machine learning (ML) techniques to develop predictive models that use microbiota associated biomarkers to identify healthy and diseased states.
-However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are robust and reproducible.
+It is unlikely that a single species can explain a phenotype.
+Instead, subsets of those communities, in relation to one another and to their host, account for the differences.
+Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict a phenotype and to understand the microbial ecology of the host.
+However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented.
 Moreover, there is a lack of consideration on why a particular ML model is utilized.
-There is a wide range of complexity and interpretibility in models used in microbiome field [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @pasolli_machine_2016]; recently with many that are black boxes which cannot be interpreted by humans [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017].
-ML model choice should reflect the goal of the study; whether it is to develop the best predictive model or to understand the ecology of microbiota associated diseases.
+Recently, there is a trend towards using more complex ML models [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017] without a discussion on if interpretibility is necessary for the study.
+If we want to generate reproducible and robust microbiota-associated biomarker models we need to (1) implement consistent, accessible and transparent machine learning practices; (2) use ML models that reflect the goal of the study as it will inform/define the expectations of model accuracy, complexity, interpretibility, computational efficiency and scalability.
 
-One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
-Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a predictive model from a stool microbiome assay as a non-invasive screening tool is a remarkable innovation.
-However, the efforts to combine ML methods with relative abundances of bacterial poplations in stool to predict CRC tumors have been afflicted by flawed ML methods and by the lack of discussion on modeling choices.
-In previous studies, ...
-There is a need for implementing consistent and transparent machine learning practices to generate reproducible and replicable CRC biomarker models.
-This study aims to shed light on how much task definition and model choices can affect the results.
+One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers in the stool [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017].
+Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a non-invasive screening tool with ML is a remarkable innovation.
+However, afforementioned standards have not been established in these ML models.
 Here we performed an empirical analysis comparing several different modeling pipelines using the same dataset.
-
-As our dataset, we used a previously published CRC study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and immunochemical test (FIT) results from `r paste(nrow(shared))` patients.
+We used a previously published CRC study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and immunochemical test (FIT) results from `r paste(nrow(shared))` patients.
 We built ML models using 16S abundances and FIT to predict patients with normal or screen relevant neoplasias (SRN) disease status.

### 2019-03-01 submission/manuscript.tex

@@
 \hypersetup{unicode=true,
-            pdftitle={Evaluation of binary classification pipelines and methods for 16S rRNA gene data},
+            pdftitle={Evaluation of machine learning methods for 16S rRNA gene data},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Evaluation of binary classification pipelines and methods for
-16S rRNA gene data}}
+  \title{\textbf{Evaluation of machine learning methods for 16S rRNA gene data}}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 microbiome field is growing at an unprecedented rate and as a result,
-there is an ever-increasing demand for reproducible methods for
-identifying associations between members of the microbiome and human
-health. However, this is an undertaking as human associated microbial
-communities are remarkably uneven. It is unlikely that a single species
-can explain a disease state comprehensively. It is more likely that
-subsets of those communities, in relation to one another and to their
-host, account for different disease states. Thus, researchers have
-started to explore the utility of machine learning (ML) techniques to
-develop predictive models that use microbiota associated biomarkers to
-identify healthy and diseased states. However, currently the field's use
-of ML lacks clarity and consistency on which methods are used, how these
-methods are implemented and if these methods are robust and
-reproducible. Moreover, there is a lack of consideration on why a
-particular ML model is utilized. There is a wide range of complexity and
-interpretibility in models used in microbiome field (1--5); recently
-with many that are black boxes which cannot be interpreted by humans
-(6--8). ML model choice should reflect the goal of the study; whether it
-is to develop the best predictive model or to understand the ecology of
-microbiota associated diseases.
+there is an increasing demand for reproducible methods to identify
+associations between members of the microbiome and human host. However,
+this is an undertaking as human associated microbial communities are
+remarkably uneven. It is unlikely that a single species can explain a
+phenotype. Instead, subsets of those communities, in relation to one
+another and to their host, account for the differences. Therefore,
+researchers have started to explore the utility of machine learning (ML)
+models that use microbiota associated biomarkers to predict a phenotype
+and to understand the microbial ecology of the host. However, currently
+the field's use of ML lacks clarity and consistency on which methods are
+used and how these methods are implemented. Moreover, there is a lack of
+consideration on why a particular ML model is utilized. Recently, there
+is a trend towards using more complex ML models (1--3) without a
+discussion on if interpretibility is important for the study. If we want
+to generate reproducible and robust microbiota-associated biomarker
+models we need to (1) implement consistent, accessible and transparent
+machine learning practices; (2) use ML models that reflect the goal of
+the study as it will inform/define the expectations of model accuracy,
+complexity, interpretibility, computational efficiency and scalability.
 
@@
 classify patients as having colorectal tumors based on
-microbiota-associated biomarkers. Colorectal cancer (CRC) is one of the
-leading causes of death in the US, therefore developing a predictive
-model from a stool microbiome assay as a non-invasive screening tool is
-a remarkable innovation. However, the efforts to combine ML methods with
-relative abundances of bacterial poplations in stool to predict CRC
-tumors have been afflicted by flawed ML methods and by the lack of
-discussion on modeling choices. In previous studies, \ldots{} There is a
-need for implementing consistent and transparent machine learning
-practices to generate reproducible and replicable CRC biomarker models.
-This study aims to shed light on how much task definition and model
-choices can affect the results. Here we performed an empirical analysis
-comparing several different modeling pipelines using the same dataset.
-
-As our dataset, we used a previously published CRC study (3) which had
-fecal 16S rRNA gene sequences and immunochemical test (FIT) results from
-490 patients. We built ML models using 16S abundances and FIT to predict
-patients with normal or screen relevant neoplasias (SRN) disease status.
-The study had 261 normal and 229 SRN samples. We established modeling
-pipelines for L2-regularized logistic regression, L1 and L2 support
-vector machines (SVM) with linear and radial basis function kernels, a
-decision tree, random forest and XGBoost. These models increase in
-complexity while they decrease in interpretability. Our ML pipeline
-performed 100 data-splits and utilized held-out test data to evaluate
-generalization and predicton performance of each ML model. The mean
-AUROC varied from 0.67 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
-and XGBoost had the highest mean AUROC for detecting SRN. In terms of
-computational efficiency, L2 Logistic Regression trained the fastest
-(0.098 hours, std ± 0.008), while XGBoost took the longest (3.601 hours,
-std ± 0.524). We found that mean cross-validation and testing AUROC
-could vary by as much as 0.062, which highlights the importance of a
-seperate held-out test set for evaluation. Aside from evaluating
-generalization and classification performances for each of these models,
-this study established standards for modeling pipelines of
-microbiome-associated machine learning models.
+microbiota-associated biomarkers in the stool (4--8). Colorectal cancer
+(CRC) is one of the leading causes of death in the US, therefore
+developing a non-invasive screening tool with ML is a remarkable
+innovation. However, afforementioned standards have not been established
+in these ML models. Here we performed an empirical analysis comparing
+several different modeling pipelines using the same dataset. We used a
+previously published CRC study (6) which had fecal 16S rRNA gene
+sequences and immunochemical test (FIT) results from 490 patients. We
+built ML models using 16S abundances and FIT to predict patients with
+normal or screen relevant neoplasias (SRN) disease status. The study had
+261 normal and 229 SRN samples. We established modeling pipelines for
+L2-regularized logistic regression, L1 and L2 support vector machines
+(SVM) with linear and radial basis function kernels, a decision tree,
+random forest and XGBoost. These models increase in complexity while
+they decrease in interpretability. Our ML pipeline performed 100
+data-splits and utilized held-out test data to evaluate generalization
+and predicton performance of each ML model. The mean AUROC varied from
+0.67 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest and XGBoost had
+the highest mean AUROC for detecting SRN. In terms of computational
+efficiency, L2 Logistic Regression trained the fastest (0.098 hours, std
+± 0.008), while XGBoost took the longest (82.47 hours, std ± 17.045). We
+found that mean cross-validation and testing AUROC could vary by as much
+as 0.062, which highlights the importance of a seperate held-out test
+set for evaluation. Aside from evaluating generalization and
+classification performances for each of these models, this study
+established standards for modeling pipelines of microbiome-associated
+machine learning models.
 
@@
 Random Forest and XGBoost had the highest mean AUROC for detecting SRN,
-0.818 (std ± 0.036) and 0.814 (std ± 0.038) respectively {[}Figure 2{]}.
+0.816 (std ± 0.039) and 0.814 (std ± 0.038) respectively {[}Figure 2{]}.
 L1 linear SVM and decision tree had significantly lower AUROC values,
@@
 and L2 logistic regression which had mean AUROC values of 0.67 (std ±
-0.054), 0.683 (std ± 0.048) and 0.677 (std ± 0.053)
-respectively{[}Figure 2{]}. We also evaluated the generalization
-performance of each classifier by comparing their mean cross-validation
-AUROC and mean testing AUROC. We found a significant difference for
+0.054), 0.684 (std ± 0.048) and 0.677 (std ± 0.053) respectively
+{[}Figure 2{]}. We also evaluated the generalization performance of each
+classifier by comparing their mean cross-validation AUROC and mean
+testing AUROC. We found a statistically significant difference for
 classifiers L2 support vector machines (SVM) with linear and radial
 basis function kernels and L2 logistic regression. These differences
-were 0.062, 0.047 and 0.061, respectively {[}Figure 2{]} .
+were 0.062, 0.046 and 0.061, respectively {[}Figure 2{]} .
 
-\textbf{The prediction performance of classifiers for different
-hyperparameter settings.}
+\textbf{Hyperparameter optimization for each classifier.}
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-galkin_human_2018}{}
+1. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
+2018. Human microbiome aging clocks based on deep learning and tandem of
+permutation feature importance and accumulated local effects. bioRxiv.
+doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
+
+\hypertarget{ref-reiman_using_2017}{}
+2. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+convolutional neural networks to explore the microbiome, pp. 4269--4272.
+\emph{In} 2017 39th annual international conference of the IEEE
+engineering in medicine and biology society (EMBC).
+
+\hypertarget{ref-fioravanti_phylogenetic_2017}{}
+3. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
+\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
+in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
+
 \hypertarget{ref-zeller_potential_2014}{}
-1. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+4. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
 \textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
@@
 \hypertarget{ref-zackular_human_2014}{}
-2. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+5. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
 \textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
@@
 \hypertarget{ref-baxter_dna_2016}{}
-3. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
+6. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
@@
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-4. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+7. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 
-\hypertarget{ref-pasolli_machine_2016}{}
-5. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
-\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
-meta-analysis of large metagenomic datasets: Tools and biological
-insights. PLoS Comput Biol \textbf{12}.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.
-
-\hypertarget{ref-galkin_human_2018}{}
-6. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
-\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
-2018. Human microbiome aging clocks based on deep learning and tandem of
-permutation feature importance and accumulated local effects. bioRxiv.
-doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
-
-\hypertarget{ref-reiman_using_2017}{}
-7. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
-convolutional neural networks to explore the microbiome, pp. 4269--4272.
-\emph{In} 2017 39th annual international conference of the IEEE
-engineering in medicine and biology society (EMBC).
-
-\hypertarget{ref-fioravanti_phylogenetic_2017}{}
-8. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
-\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
-\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
-in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
+\hypertarget{ref-hale_shifts_2017}{}
+8. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
+\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
+\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
+\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
+\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
+microbiota associated with adenomatous polyps. Cancer Epidemiol
+Biomarkers Prev \textbf{26}:85--94.
+doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.
 

### 2019-03-04 submission/manuscript.Rmd

@@
 Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome.
-Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for reproducible methods to identify associations between members of the microbiome and human host.
+Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for reproducible methods to identify associations between members of the microbiome and human health.
 However, this is an undertaking as human associated microbial communities are remarkably uneven.
-It is unlikely that a single species can explain a phenotype.
-Instead, subsets of those communities, in relation to one another and to their host, account for the differences.
-Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict a phenotype and to understand the microbial ecology of the host.
+It is unlikely that a single species can explain a disease.
+Instead, subsets of those communities, in relation to one another and to their host, account for the differences in the health outcomes.
+Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of the human diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented.
-Moreover, there is a lack of consideration on why a particular ML model is utilized.
-Recently, there is a trend towards using more complex ML models [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017] without a discussion on if interpretibility is necessary for the study.
-If we want to generate reproducible and robust microbiota-associated biomarker models we need to (1) implement consistent, accessible and transparent machine learning practices; (2) use ML models that reflect the goal of the study as it will inform/define the expectations of model accuracy, complexity, interpretibility, computational efficiency and scalability.
+Moreover, there is a lack of deliberation on why a particular ML model is utilized.
+Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
+If we want to generate reproducible and robust microbiota-associated biomarker models we need to (1) implement consistent, accessible and transparent machine learning practices; (2) use ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility, computational efficiency and scalability.
 
-One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers in the stool [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017].
-Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a non-invasive screening tool with ML is a remarkable innovation.
-However, afforementioned standards have not been established in these ML models.
-Here we performed an empirical analysis comparing several different modeling pipelines using the same dataset.
-We used a previously published CRC study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and immunochemical test (FIT) results from `r paste(nrow(shared))` patients.
-We built ML models using 16S abundances and FIT to predict patients with normal or screen relevant neoplasias (SRN) disease status.
+To shed light on how much task definition and modeling choices can affect the outcomes of ML studies, we performed an empirical analysis comparing several different modeling pipelines using the same dataset.
+We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients.
+We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal or screen relevant neoplasias (SRN) disease status.
 The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
@@
 Random Forest and XGBoost had the highest mean AUROC for detecting SRN.
+Despite the simplicity, the L1-regularized linear kernel SVM performed on par with many of the non-linear models.
 In terms of computational efficiency, L2 Logistic Regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
@@
 FIT levels are continuous data present for each sample.
-Because the data are in different scales, Python programming language v, module scikit-learn v is used to transform features by scaling each feature to a [0-1] range (Table 1) [@scikit-learn].
+Because the data are in different scales, features are transformd by scaling each feature to a [0-1] range (Table 1).
 

### 2019-03-04 submission/manuscript.tex

@@
 there is an increasing demand for reproducible methods to identify
-associations between members of the microbiome and human host. However,
-this is an undertaking as human associated microbial communities are
-remarkably uneven. It is unlikely that a single species can explain a
-phenotype. Instead, subsets of those communities, in relation to one
-another and to their host, account for the differences. Therefore,
-researchers have started to explore the utility of machine learning (ML)
-models that use microbiota associated biomarkers to predict a phenotype
-and to understand the microbial ecology of the host. However, currently
-the field's use of ML lacks clarity and consistency on which methods are
-used and how these methods are implemented. Moreover, there is a lack of
-consideration on why a particular ML model is utilized. Recently, there
-is a trend towards using more complex ML models (1--3) without a
-discussion on if interpretibility is important for the study. If we want
-to generate reproducible and robust microbiota-associated biomarker
-models we need to (1) implement consistent, accessible and transparent
-machine learning practices; (2) use ML models that reflect the goal of
-the study as it will inform/define the expectations of model accuracy,
-complexity, interpretibility, computational efficiency and scalability.
+associations between members of the microbiome and human health.
+However, this is an undertaking as human associated microbial
+communities are remarkably uneven. It is unlikely that a single species
+can explain a disease. Instead, subsets of those communities, in
+relation to one another and to their host, account for the differences
+in the health outcomes. Therefore, researchers have started to explore
+the utility of machine learning (ML) models that use microbiota
+associated biomarkers to predict human health and to understand the
+microbial ecology of the human diseases such as liver cirrhosis,
+colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type
+2 diabetes (1--11). However, currently the field's use of ML lacks
+clarity and consistency on which methods are used and how these methods
+are implemented. Moreover, there is a lack of deliberation on why a
+particular ML model is utilized. Recently, there is a trend towards
+using more complex ML models such as random forest, extreme gradient
+boosting and neural networks (11--14) without a discussion on if and how
+much model interpretibility is necessary for the study. If we want to
+generate reproducible and robust microbiota-associated biomarker models
+we need to (1) implement consistent, accessible and transparent machine
+learning practices; (2) use ML models that reflect the goal of the study
+as it will inform the expectations of model accuracy, complexity,
+interpretibility, computational efficiency and scalability.
 
-One application of machine learning to microbiome data has been to
-classify patients as having colorectal tumors based on
-microbiota-associated biomarkers in the stool (4--8). Colorectal cancer
-(CRC) is one of the leading causes of death in the US, therefore
-developing a non-invasive screening tool with ML is a remarkable
-innovation. However, afforementioned standards have not been established
-in these ML models. Here we performed an empirical analysis comparing
-several different modeling pipelines using the same dataset. We used a
-previously published CRC study (6) which had fecal 16S rRNA gene
-sequences and immunochemical test (FIT) results from 490 patients. We
-built ML models using 16S abundances and FIT to predict patients with
+To shed light on how much task definition and modeling choices can
+affect the outcomes of ML studies, we performed an empirical analysis
+comparing several different modeling pipelines using the same dataset.
+We used a previously published colorectal cancer (CRC) study (3) which
+had fecal 16S rRNA gene sequences and human hemoglobin concentrations
+from 490 patients. We built ML models using fecal 16S rRNA gene
+sequences and human hemoglobin concentrations to predict patients with
 normal or screen relevant neoplasias (SRN) disease status. The study had
@@
 0.67 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest and XGBoost had
-the highest mean AUROC for detecting SRN. In terms of computational
-efficiency, L2 Logistic Regression trained the fastest (0.098 hours, std
-± 0.008), while XGBoost took the longest (82.47 hours, std ± 17.045). We
-found that mean cross-validation and testing AUROC could vary by as much
-as 0.062, which highlights the importance of a seperate held-out test
-set for evaluation. Aside from evaluating generalization and
-classification performances for each of these models, this study
-established standards for modeling pipelines of microbiome-associated
-machine learning models.
+the highest mean AUROC for detecting SRN. Despite the simplicity, the
+L1-regularized linear kernel SVM performed on par with many of the
+non-linear models. In terms of computational efficiency, L2 Logistic
+Regression trained the fastest (0.105 hours, std ± 0.013), while XGBoost
+took the longest (82.47 hours, std ± 17.045). We found that mean
+cross-validation and testing AUROC could vary by as much as 0.062, which
+highlights the importance of a seperate held-out test set for
+evaluation. Aside from evaluating generalization and classification
+performances for each of these models, this study established standards
+for modeling pipelines of microbiome-associated machine learning models.
 
@@
 and L2 logistic regression which had mean AUROC values of 0.67 (std ±
-0.054), 0.684 (std ± 0.048) and 0.677 (std ± 0.053) respectively
+0.054), 0.684 (std ± 0.049) and 0.677 (std ± 0.053) respectively
 {[}Figure 2{]}. We also evaluated the generalization performance of each
@@
 Great Lakes-New England Early Detection Research Network study. These
-data were obtained from Sze et al (9). The stool samples were provided
+data were obtained from Sze et al (15). The stool samples were provided
 by recruited adult participants who were undergoing scheduled screening
@@
 sequences using VSEARCH and assigning to OTUs at 97\% similarity using
-the OptiClust algorithm (10--12).
+the OptiClust algorithm (16--18).
 
@@
 sample. FIT levels are continuous data present for each sample. Because
-the data are in different scales, Python programming language v, module
-scikit-learn v is used to transform features by scaling each feature to
-a {[}0-1{]} range (Table 1) (13).
+the data are in different scales, features are transformd by scaling
+each feature to a {[}0-1{]} range (Table 1).
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-galkin_human_2018}{}
-1. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
-\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
-2018. Human microbiome aging clocks based on deep learning and tandem of
-permutation feature importance and accumulated local effects. bioRxiv.
-doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
-
-\hypertarget{ref-reiman_using_2017}{}
-2. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
-convolutional neural networks to explore the microbiome, pp. 4269--4272.
-\emph{In} 2017 39th annual international conference of the IEEE
-engineering in medicine and biology society (EMBC).
-
-\hypertarget{ref-fioravanti_phylogenetic_2017}{}
-3. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
-\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
-\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
-in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
-
 \hypertarget{ref-zeller_potential_2014}{}
-4. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+1. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
 \textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
@@
 \hypertarget{ref-zackular_human_2014}{}
-5. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+2. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
 \textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
@@
 \hypertarget{ref-baxter_dna_2016}{}
-6. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
+3. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
@@
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-7. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+4. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-hale_shifts_2017}{}
-8. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
+5. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
 \textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
@@
 
+\hypertarget{ref-pasolli_machine_2016}{}
+6. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
+\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
+meta-analysis of large metagenomic datasets: Tools and biological
+insights. PLoS Comput Biol \textbf{12}.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.
+
+\hypertarget{ref-sze_looking_2016}{}
+7. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
+the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
+doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.
+
+\hypertarget{ref-walters_meta-analyses_2014}{}
+8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
+Meta-analyses of human gut microbes associated with obesity and IBD.
+FEBS Lett \textbf{588}:4223--4233.
+doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
+
+\hypertarget{ref-vazquez-baeza_guiding_2018}{}
+9. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
+\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
+\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
+Gut \textbf{67}:1743--1745.
+doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.
+
+\hypertarget{ref-qin_alterations_2014}{}
+10. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
+\textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
+E}, \textbf{Yao J}, \textbf{Wu L}, \textbf{Zhou J}, \textbf{Ni S},
+\textbf{Liu L}, \textbf{Pons N}, \textbf{Batto JM}, \textbf{Kennedy SP},
+\textbf{Leonard P}, \textbf{Yuan C}, \textbf{Ding W}, \textbf{Chen Y},
+\textbf{Hu X}, \textbf{Zheng B}, \textbf{Qian G}, \textbf{Xu W},
+\textbf{Ehrlich SD}, \textbf{Zheng S}, \textbf{Li L}. 2014. Alterations
+of the human gut microbiome in liver cirrhosis. Nature
+\textbf{513}:59--64.
+doi:\href{https://doi.org/10.1038/nature13568}{10.1038/nature13568}.
+
+\hypertarget{ref-geman_deep_2018}{}
+11. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
+\textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
+learning tools for human microbiome big data, pp. 265--275. \emph{In}
+Balas, VE, Jain, LC, Balas, MM (eds.), Soft computing applications.
+Springer International Publishing.
+
+\hypertarget{ref-galkin_human_2018}{}
+12. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
+2018. Human microbiome aging clocks based on deep learning and tandem of
+permutation feature importance and accumulated local effects. bioRxiv.
+doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
+
+\hypertarget{ref-reiman_using_2017}{}
+13. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+convolutional neural networks to explore the microbiome, pp. 4269--4272.
+\emph{In} 2017 39th annual international conference of the IEEE
+engineering in medicine and biology society (EMBC).
+
+\hypertarget{ref-fioravanti_phylogenetic_2017}{}
+14. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
+\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
+in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
+
 \hypertarget{ref-sze_leveraging_2018}{}
-9. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+15. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-10. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+16. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-11. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+17. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-12. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+18. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 
-\hypertarget{ref-scikit-learn}{}
-13. \textbf{Pedregosa F}, \textbf{Varoquaux G}, \textbf{Gramfort A},
-\textbf{Michel V}, \textbf{Thirion B}, \textbf{Grisel O},
-\textbf{Blondel M}, \textbf{Prettenhofer P}, \textbf{Weiss R},
-\textbf{Dubourg V}, \textbf{Vanderplas J}, \textbf{Passos A},
-\textbf{Cournapeau D}, \textbf{Brucher M}, \textbf{Perrot M},
-\textbf{Duchesnay E}. 2011. Scikit-learn: Machine learning in Python.
-Journal of Machine Learning Research \textbf{12}:2825--2830.
-
 \end{document}

### 2019-03-06 submission/manuscript.Rmd
update

@@
 
+# Get the The unpaired two-samples Wilcoxon test to see if models differ from one another signigicantly
+
+rf_xgboost <- wilcox.test(rf$test_aucs, xgboost$test_aucs)
+rf_l1svm <- wilcox.test(rf$test_aucs, l1svm$test_aucs)
+l1svm_dt <- wilcox.test(l1svm$test_aucs, dt$test_aucs)
+l1svm_rbf <- wilcox.test(l1svm$test_aucs, rbf$test_aucs)
+l1svm_l2svm <- wilcox.test(l1svm$test_aucs, l2svm$test_aucs)
+l1svm_logit <- wilcox.test(l1svm$test_aucs, logit$test_aucs)
+
+# Get the unpaired two-samples Wilcoxon test to see if cv and test differ significantly
+logit_diff <- wilcox.test(logit$test_aucs, logit$cv_aucs)
+l2svm_diff <- wilcox.test(l2svm$test_aucs, l2svm$cv_aucs)
+rbf_diff <- wilcox.test(rbf$test_aucs, rbf$cv_aucs)
+
 # Bind them together and summarise mean testAUC by model
@@
 
-Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome.
-Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for reproducible methods to identify associations between members of the microbiome and human health.
-However, this is an undertaking as human associated microbial communities are remarkably uneven.
+Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications.
+Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health.
+However, this is an undertaking as human associated microbial communities are remarkably complex and uneven.
 It is unlikely that a single species can explain a disease.
 Instead, subsets of those communities, in relation to one another and to their host, account for the differences in the health outcomes.
-Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of the human diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+ML methods are effective at recognizing and highlighting patterns in complex microbial datasets.
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented.
@@
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-If we want to generate reproducible and robust microbiota-associated biomarker models we need to (1) implement consistent, accessible and transparent machine learning practices; (2) use ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility, computational efficiency and scalability.
+The lack of transparency on methodology and modeling decisions can have negative consequences on model reproducibility and reliability which we need avoid by (1) implementing consistent, accessible and transparent machine learning practices; (2) selecting ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility and computational efficiency.
 
-To shed light on how much task definition and modeling choices can affect the outcomes of ML studies, we performed an empirical analysis comparing several different modeling pipelines using the same dataset.
+To showcase transparent ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset.
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients.
@@
 The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
-We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
-These models increase in complexity while they decrease in interpretability.
+We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which increase in complexity and decrease in interpretability respectively.
 Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model.
@@
 Random Forest and XGBoost had the highest mean AUROC for detecting SRN.
-Despite the simplicity, the L1-regularized linear kernel SVM performed on par with many of the non-linear models.
-In terms of computational efficiency, L2 Logistic Regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
+Despite the simplicity, the L1-regularized linear kernel SVM followed Random Forest and XGBoost in performance.
+In terms of computational efficiency, L2 SVM with linear kernel trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
 We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation.
@@
 
-## Results and Discussion
+## Results
+
+__Model selection and construction__
+
+We established modeling pipelines for binary classification with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility, computational efficiency and scalability due to  model selection.
+We randomly split the data into stratified training and test sets 100 times with a 80/20 split [Figure 1].
+The training set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
+The training data was used for training purposes and validation of parameter selection, and the test set was used for evaluation purposes.
+Validation of parameter selection was repeated 100 times to get a robust parameter selection for each model [Figure 1].
 
@@
 
-We evaluated the prediction performance of seven binary classification models when applied to held-out test data using AUROC as our metric.
+We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using AUROC as classification performance metric.
 Random Forest and XGBoost had the highest mean AUROC for detecting SRN, `r paste(round(mean(rf$test_aucs), 3))` (std ± `r paste(round(sd(rf$test_aucs), 3))`) and `r paste(round(mean(xgboost$test_aucs), 3))` (std ± `r paste(round(sd(xgboost$test_aucs), 3))`) respectively [Figure 2].
@@
 We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC.
-We found a statistically significant difference for classifiers L2 support vector machines (SVM) with linear and radial basis function kernels and L2 logistic regression.
+We found a statistically significant difference for classifiers L2 support vector machines (SVM) with linear and radial basis function kernels and L2 logistic regression (p-values = `r paste(format(l2svm_diff$p.value, digits=2))`, `r paste(format(rbf_diff$p.value, digits=2))` and `r paste(format(logit_diff$p.value, digits=2))` respectively).
 These differences were `r paste(round(difference_model[3,1], 3))`, `r paste(round(difference_model[6,1], 3))` and `r paste(round(difference_model[4,1], 3))`, respectively [Figure 2] .
 
-__Hyperparameter optimization for each classifier.__
+__The complexity and interpretibility of each classifier.__
 
-__The interpretation of classifiers.__
+We interpreted the feature weights of L1 and L2 SVM with linear kernel and log odds ratios of L2 logistic regression using the training data.
+To get a sense of the importance of each feature we randomly subsampled 80% of the data 100 times and trained 100 models; this resulted in 100 different sets of feature weights.
+We reported the mean weights of the 10 most important features for[Figure 4].
+FIT levels were
+
+__The computational efficiency of each classifier.__
+
+Linear models trained faster than non-linear models.
+L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(60*(round(walltime_df[walltime_index[1],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[1],3], 2)))`), `r paste(60*(round(walltime_df[walltime_index[2],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[2],3], 2)))`) and `r paste(60*(round(walltime_df[walltime_index[3],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[3],3], 2)))`) respectively.
+Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[walltime_index[4],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[4],3], 2))`), `r paste((round(walltime_df[walltime_index[5],2], 2)))` hrs, (std ± `r paste((round(walltime_df[walltime_index[5],3], 2)))`), `r paste(round(walltime_df[walltime_index[6],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[6],3], 2))`) and `r paste(round(walltime_df[walltime_index[7],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[7],3], 2))`), respectively.
 

### 2019-03-06 submission/manuscript.tex
update

@@
 Advances in sequencing technology and decreasing costs of generating 16S
-rRNA gene sequences have allowed rapid exploration and taxonomic
-characterization of the human associated microbiome. Currently, the
-microbiome field is growing at an unprecedented rate and as a result,
-there is an increasing demand for reproducible methods to identify
-associations between members of the microbiome and human health.
-However, this is an undertaking as human associated microbial
-communities are remarkably uneven. It is unlikely that a single species
-can explain a disease. Instead, subsets of those communities, in
-relation to one another and to their host, account for the differences
-in the health outcomes. Therefore, researchers have started to explore
-the utility of machine learning (ML) models that use microbiota
-associated biomarkers to predict human health and to understand the
-microbial ecology of the human diseases such as liver cirrhosis,
-colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type
-2 diabetes (1--11). However, currently the field's use of ML lacks
-clarity and consistency on which methods are used and how these methods
-are implemented. Moreover, there is a lack of deliberation on why a
-particular ML model is utilized. Recently, there is a trend towards
-using more complex ML models such as random forest, extreme gradient
-boosting and neural networks (11--14) without a discussion on if and how
-much model interpretibility is necessary for the study. If we want to
-generate reproducible and robust microbiota-associated biomarker models
-we need to (1) implement consistent, accessible and transparent machine
-learning practices; (2) use ML models that reflect the goal of the study
-as it will inform the expectations of model accuracy, complexity,
-interpretibility, computational efficiency and scalability.
+rRNA gene sequences have allowed rapid exploration of human associated
+microbiome and its health implications. Currently, the human microbiome
+field is growing at an unprecedented rate and as a result, there is an
+increasing demand for methods that identify associations between members
+of the microbiome and human health. However, this is an undertaking as
+human associated microbial communities are remarkably complex and
+uneven. It is unlikely that a single species can explain a disease.
+Instead, subsets of those communities, in relation to one another and to
+their host, account for the differences in the health outcomes.
+Therefore, researchers have started to explore the utility of machine
+learning (ML) models that use microbiota associated biomarkers to
+predict human health and to understand the microbial ecology of diseases
+such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases
+(IBD), obesity, and type 2 diabetes (1--11). ML methods are effective at
+recognizing and highlighting patterns in complex microbial datasets.
+However, currently the field's use of ML lacks clarity and consistency
+on which methods are used and how these methods are implemented.
+Moreover, there is a lack of deliberation on why a particular ML model
+is utilized. Recently, there is a trend towards using more complex ML
+models such as random forest, extreme gradient boosting and neural
+networks without a discussion on if and how much model interpretibility
+is necessary for the study (11--14). The lack of transparency on
+methodology and modeling decisions can have negative consequences on
+model reproducibility and reliability which we need avoid by (1)
+implementing consistent, accessible and transparent machine learning
+practices; (2) selecting ML models that reflect the goal of the study as
+it will inform the expectations of model accuracy, complexity,
+interpretibility and computational efficiency.
 
-To shed light on how much task definition and modeling choices can
-affect the outcomes of ML studies, we performed an empirical analysis
-comparing several different modeling pipelines using the same dataset.
-We used a previously published colorectal cancer (CRC) study (3) which
-had fecal 16S rRNA gene sequences and human hemoglobin concentrations
-from 490 patients. We built ML models using fecal 16S rRNA gene
-sequences and human hemoglobin concentrations to predict patients with
-normal or screen relevant neoplasias (SRN) disease status. The study had
-261 normal and 229 SRN samples. We established modeling pipelines for
-L2-regularized logistic regression, L1 and L2 support vector machines
-(SVM) with linear and radial basis function kernels, a decision tree,
-random forest and XGBoost. These models increase in complexity while
-they decrease in interpretability. Our ML pipeline performed 100
-data-splits and utilized held-out test data to evaluate generalization
-and predicton performance of each ML model. The mean AUROC varied from
-0.67 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest and XGBoost had
-the highest mean AUROC for detecting SRN. Despite the simplicity, the
-L1-regularized linear kernel SVM performed on par with many of the
-non-linear models. In terms of computational efficiency, L2 Logistic
-Regression trained the fastest (0.105 hours, std ± 0.013), while XGBoost
-took the longest (82.47 hours, std ± 17.045). We found that mean
-cross-validation and testing AUROC could vary by as much as 0.062, which
+To showcase transparent ML methodologies and to shed light on how much
+ML model selection can affect modeling results, we performed an
+empirical analysis comparing several different ML models using the same
+dataset. We used a previously published colorectal cancer (CRC) study
+(3) which had fecal 16S rRNA gene sequences and human hemoglobin
+concentrations from 490 patients. We built ML models using fecal 16S
+rRNA gene sequences and human hemoglobin concentrations to predict
+patients with normal or screen relevant neoplasias (SRN) disease status.
+The study had 261 normal and 229 SRN samples. We established modeling
+pipelines for L2-regularized logistic regression, L1 and L2 support
+vector machines (SVM) with linear and radial basis function kernels, a
+decision tree, random forest and XGBoost which increase in complexity
+and decrease in interpretability respectively. Our ML pipeline performed
+100 data-splits and utilized held-out test data to evaluate
+generalization and predicton performance of each ML model. The mean
+AUROC varied from 0.68 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
+and XGBoost had the highest mean AUROC for detecting SRN. Despite the
+simplicity, the L1-regularized linear kernel SVM followed Random Forest
+and XGBoost in performance. In terms of computational efficiency, L2 SVM
+with linear kernel trained the fastest (0.127 hours, std ± 0.01), while
+XGBoost took the longest (82.47 hours, std ± 17.045). We found that mean
+cross-validation and testing AUROC could vary by as much as 0.061, which
 highlights the importance of a seperate held-out test set for
@@
 
-\subsection{Results and Discussion}\label{results-and-discussion}
+\subsection{Results}\label{results}
+
+\textbf{Model selection and construction}
+
+We established modeling pipelines for binary classification with
+L2-regularized logistic regression, L1 and L2 support vector machines
+(SVM) with linear and radial basis function kernels, a decision tree,
+random forest and XGBoost to emphasize the differences in model
+accuracy, complexity, interpretibility, computational efficiency and
+scalability due to model selection. We randomly split the data into
+stratified training and test sets 100 times with a 80/20 split {[}Figure
+1{]}. The training set consisted of 393 patients (209 SRN), while the
+test set was composed of 97 patients (52 SRN). The training data was
+used for training purposes and validation of parameter selection, and
+the test set was used for evaluation purposes. Validation of parameter
+selection was repeated 100 times to get a robust parameter selection for
+each model {[}Figure 1{]}.
 
@@
 We evaluated the prediction performance of seven binary classification
-models when applied to held-out test data using AUROC as our metric.
-Random Forest and XGBoost had the highest mean AUROC for detecting SRN,
-0.816 (std ± 0.039) and 0.814 (std ± 0.038) respectively {[}Figure 2{]}.
-L1 linear SVM and decision tree had significantly lower AUROC values,
-0.748 (std ± 0.043) and 0.741 (std ± 0.038) {[}Figure 2{]}. However,
-they had significanly higher performances than L2 linear SVM, RBF SVM
-and L2 logistic regression which had mean AUROC values of 0.67 (std ±
-0.054), 0.684 (std ± 0.049) and 0.677 (std ± 0.053) respectively
-{[}Figure 2{]}. We also evaluated the generalization performance of each
-classifier by comparing their mean cross-validation AUROC and mean
-testing AUROC. We found a statistically significant difference for
-classifiers L2 support vector machines (SVM) with linear and radial
-basis function kernels and L2 logistic regression. These differences
-were 0.062, 0.046 and 0.061, respectively {[}Figure 2{]} .
+models when applied to held-out test data over 100 data-splits using
+AUROC as classification performance metric. Random Forest and XGBoost
+had the highest mean AUROC for detecting SRN, 0.816 (std ± 0.039) and
+0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM and
+decision tree had significantly lower AUROC values, 0.75 (std ± 0.045)
+and 0.741 (std ± 0.038) {[}Figure 2{]}. However, they had significanly
+higher performances than L2 linear SVM, RBF SVM and L2 logistic
+regression which had mean AUROC values of 0.68 (std ± 0.052), 0.684 (std
+± 0.049) and 0.677 (std ± 0.053) respectively {[}Figure 2{]}. We also
+evaluated the generalization performance of each classifier by comparing
+their mean cross-validation AUROC and mean testing AUROC. We found a
+statistically significant difference for classifiers L2 support vector
+machines (SVM) with linear and radial basis function kernels and L2
+logistic regression (p-values = 2.4e-16, 3.5e-13 and 9.4e-20
+respectively). These differences were 0.052, 0.047 and 0.061,
+respectively {[}Figure 2{]} .
 
-\textbf{Hyperparameter optimization for each classifier.}
+\textbf{The complexity and interpretibility of each classifier.}
 
-\textbf{The interpretation of classifiers.}
+\textbf{The computational efficiency of each classifier.}
+
+Linear models trained faster than non-linear models. L2 logistic
+Rregression and L1 and L2 SVM with linear kernel had training times of
+7.8 min, (std ± 0.6), 9.6 min, (std ± 0.6) and 13.2 min, (std ± 1.2)
+respectively. Whereas, SVM with radial basis function kernel, decision
+tree, random forest and xgboost had training times of 2.21 hrs, (std ±
+0.25), 4.2 hrs, (std ± 0.49), 82.47 hrs, (std ± 17.05) and 116.83 hrs,
+(std ± 17.24), respectively.
 

### 2019-03-06 submission/manuscript.Rmd
figures and text changes

@@
 
-We interpreted the feature weights of L1 and L2 SVM with linear kernel and log odds ratios of L2 logistic regression using the training data.
+We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data.
 To get a sense of the importance of each feature we randomly subsampled 80% of the data 100 times and trained 100 models; this resulted in 100 different sets of feature weights.
-We reported the mean weights of the 10 most important features for[Figure 4].
+We reported the mean weights of the 10 most important features [Figure 4].
 FIT levels were

### 2019-03-06 code/learning/Figure2.R

@@ -71,7 +71,7 @@ l1svm <- l1svm_all %>%
 
 l1svm_plot <- base_plot(l1svm, l1svm$cost, l1svm$mean_Acc) +
   scale_x_continuous(name="C (penalty)") +
-  scale_y_continuous(name="L1 Linear Kernel SVM mean cvAUC",
+  scale_y_continuous(name="L1 SVM with linear kernel mean cvAUC",
                      limits = c(0.30, 1),
                      breaks = seq(0.3, 1, 0.05)) +
   geom_errorbar(aes(ymin=mean_Acc-sd_Acc, ymax=mean_Acc+sd_Acc), width=.001)
@@ -83,7 +83,7 @@ l2svm <- l2svm_all %>%
 
 l2svm_plot <- base_plot(l2svm, l2svm$cost, l2svm$mean_Acc) +
   scale_x_continuous(name="C (penalty)") +
-  scale_y_continuous(name="L2 Linear Kernel SVM mean cvAUC",
+  scale_y_continuous(name="L2 SVM with linear kernel mean cvAUC",
                      limits = c(0.30, 1),
                      breaks = seq(0.3, 1, 0.05)) +
   geom_errorbar(aes(ymin=mean_Acc-sd_Acc, ymax=mean_Acc+sd_Acc), width=.001)
@@ -95,7 +95,7 @@ logit_plot <- logit_all %>%
   geom_line() +
   geom_point() +
   scale_x_continuous(name="C (penalty)") +
-  scale_y_continuous(name="L2 Logistic Regression mean cvAUC",
+  scale_y_continuous(name="L2 logistic regression mean cvAUC",
                      limits = c(0.30, 1),
                      breaks = seq(0.3, 1, 0.05)) +
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.01) +
@@ -117,7 +117,7 @@ dt <- dt_all %>%
 
 dt_plot <- base_plot(dt, dt$maxdepth, dt$mean_AUC) +
 scale_x_continuous(name="max depth") +
-  scale_y_continuous(name="Decision Tree mean cvAUC",
+  scale_y_continuous(name="Decision tree mean cvAUC",
                      limits = c(0.30, 1),
                      breaks = seq(0.3, 1, 0.05)) +
[…]

### 2019-03-06 code/learning/Figure3.R

@@ -54,7 +54,7 @@ walltime_df <- bind_rows(min_fixed_result)
 
 walltime_plot <- ggplot(walltime_df, aes(x = fct_reorder(model, x, fun = median, .asc =TRUE), y = x)) +
   geom_boxplot(alpha=0.7) +
-  scale_y_continuous(name = "Walltime") +
+  scale_y_continuous(name = "Walltime (hours)") +
   scale_x_discrete(name = "") +
   theme_bw() +
   theme(legend.title=element_text(size=22),

### 2019-03-07 submission/manuscript.Rmd
update text

@@
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-The lack of transparency on methodology and modeling decisions can have negative consequences on model reproducibility and reliability which we need avoid by (1) implementing consistent, accessible and transparent machine learning practices; (2) selecting ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility and computational efficiency.
+The lack of transparency on methodology and modeling decisions can have negative consequences on model reproducibility and reliability which we need to avoid by (1) implementing consistent, accessible and transparent machine learning practices; (2) selecting ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility and computational efficiency.
 
@@
 We reported the mean weights of the 10 most important features [Figure 4].
-FIT levels were
 

### 2019-03-07 submission/manuscript.tex
update text

@@
 generalization and predicton performance of each ML model. The mean
-AUROC varied from 0.68 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
+AUROC varied from 0.5 (std ± 0.01) to 0.82 (std ± 0.04). Random Forest
 and XGBoost had the highest mean AUROC for detecting SRN. Despite the
@@
 and XGBoost in performance. In terms of computational efficiency, L2 SVM
-with linear kernel trained the fastest (0.127 hours, std ± 0.01), while
-XGBoost took the longest (82.47 hours, std ± 17.045). We found that mean
-cross-validation and testing AUROC could vary by as much as 0.061, which
-highlights the importance of a seperate held-out test set for
+with linear kernel trained the fastest (0.123 hours, std ± 0.014), while
+XGBoost took the longest (106.568 hours, std ± 6.839). We found that
+mean cross-validation and testing AUROC could vary by as much as 0.238,
+which highlights the importance of a seperate held-out test set for
 evaluation. Aside from evaluating generalization and classification
@@
 0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM and
-decision tree had significantly lower AUROC values, 0.75 (std ± 0.045)
+decision tree had significantly lower AUROC values, 0.503 (std ± 0.027)
 and 0.741 (std ± 0.038) {[}Figure 2{]}. However, they had significanly
 higher performances than L2 linear SVM, RBF SVM and L2 logistic
-regression which had mean AUROC values of 0.68 (std ± 0.052), 0.684 (std
-± 0.049) and 0.677 (std ± 0.053) respectively {[}Figure 2{]}. We also
+regression which had mean AUROC values of 0.501 (std ± 0.013), 0.68 (std
+± 0.05) and 0.672 (std ± 0.05) respectively {[}Figure 2{]}. We also
 evaluated the generalization performance of each classifier by comparing
@@
 machines (SVM) with linear and radial basis function kernels and L2
-logistic regression (p-values = 2.4e-16, 3.5e-13 and 9.4e-20
-respectively). These differences were 0.052, 0.047 and 0.061,
+logistic regression (p-values = 2.4e-37, 1.3e-14 and 2.6e-21
+respectively). These differences were 0.222, 0.05 and 0.064,
 respectively {[}Figure 2{]} .
@@
 \textbf{The complexity and interpretibility of each classifier.}
+
+We interpreted the feature weights of L1 and L2 SVM with linear kernel
+and regression coefficients of L2 logistic regression using the training
+data. To get a sense of the importance of each feature we randomly
+subsampled 80\% of the data 100 times and trained 100 models; this
+resulted in 100 different sets of feature weights. We reported the mean
+weights of the 10 most important features {[}Figure 4{]}.
 
@@
 Rregression and L1 and L2 SVM with linear kernel had training times of
-7.8 min, (std ± 0.6), 9.6 min, (std ± 0.6) and 13.2 min, (std ± 1.2)
+7.2 min, (std ± 0.6), 10.2 min, (std ± 0.6) and 13.2 min, (std ± 1.8)
 respectively. Whereas, SVM with radial basis function kernel, decision
-tree, random forest and xgboost had training times of 2.21 hrs, (std ±
-0.25), 4.2 hrs, (std ± 0.49), 82.47 hrs, (std ± 17.05) and 116.83 hrs,
+tree, random forest and xgboost had training times of 2.01 hrs, (std ±
+0.28), 4.2 hrs, (std ± 0.49), 106.57 hrs, (std ± 6.84) and 116.83 hrs,
 (std ± 17.24), respectively.

### 2019-03-13 submission/manuscript.Rmd
update

@@
 
+# Merge metadata and OTU table.
+# Group advanced adenomas and cancers together as cancer and normal, high risk normal and non-advanced adenomas as normal
+# Then remove the sample ID column
+data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
+  mutate(dx = case_when(
+    Dx_Bin== "Adenoma" ~ "normal",
+    Dx_Bin== "Normal" ~ "normal",
+    Dx_Bin== "High Risk Normal" ~ "normal",
+    Dx_Bin== "adv Adenoma" ~ "cancer",
+    Dx_Bin== "Cancer" ~ "cancer"
+  )) %>%
+  select(-sample, -Dx_Bin) %>%
+  drop_na()
+# We want the diagnosis column to be a factor
+data$dx <- factor(data$dx)
+
 ######################################################################
@@
 result <- list()
+
 for(file in walltime_files){
@@
 for(i in result){
-  if((sum(i$model=="L1_Linear_SVM")==100 || sum(i$model=="L2_Linear_SVM")==100 || sum(i$model=="L2_Logistic_Regression")==100)){
-    i$x <- i$x/60 # The walltimes were saved as minutes # We convert these to hours by diving with 60
-  }
-  if((sum(i$model=="Random_Forest")==100 || sum(i$model=="XGBoost")==100)){
-    i$x <- i$x*24 # The walltimes were saved as minutes # We convert these to hours by diving with 60
-  }
-  else{
-  }
+  i$x <- i$x/3600 # The walltimes were saved as seconds we covert to hours
   min_fixed_result[[length(min_fixed_result)+1]] <- i
 }
-
 # Bind all delims with model name and walltime together
@@
 It is unlikely that a single species can explain a disease.
-Instead, subsets of those communities, in relation to one another and to their host, account for the differences in the health outcomes.
-Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-ML methods are effective at recognizing and highlighting patterns in complex microbial datasets.
+Instead, subsets of those communities, in relation to one another and to their host, account for the differences in health outcomes.
+Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
+Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented.
@@
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-The lack of transparency on methodology and modeling decisions can have negative consequences on model reproducibility and reliability which we need to avoid by (1) implementing consistent, accessible and transparent machine learning practices; (2) selecting ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility and computational efficiency.
+The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability.
+We need to strive toward better machine learning practices by (1) implementing consistent, accessible and transparent machine learning methods; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
 
@@
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients.
-We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal or screen relevant neoplasias (SRN) disease status.
+We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN).
 The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
-We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which increase in complexity and decrease in interpretability respectively.
-Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model.
-The mean AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`).
+We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which are ML models that increase in complexity respectively.
+Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model.
+The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`).
 Random Forest and XGBoost had the highest mean AUROC for detecting SRN.
 Despite the simplicity, the L1-regularized linear kernel SVM followed Random Forest and XGBoost in performance.
-In terms of computational efficiency, L2 SVM with linear kernel trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
-We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation.
+In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
+We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing/splitting data for evaluation.
 Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
@@
 
-We established modeling pipelines for binary classification with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility, computational efficiency and scalability due to  model selection.
-We randomly split the data into stratified training and test sets 100 times with a 80/20 split [Figure 1].
-The training set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
-The training data was used for training purposes and validation of parameter selection, and the test set was used for evaluation purposes.
-Validation of parameter selection was repeated 100 times to get a robust parameter selection for each model [Figure 1].
+The data has `r paste(ncol(data)-1)` features (`r paste(ncol(data)-2)` OTUs and FIT) and a two-class label that defines the colonic health of the patients (SRN or normal).
+We established modeling pipelines for binary classification with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
+We randomly split the data into stratified training/validation and test sets 100 times with a 80/20 proportion [Figure 1].
+For each data-split, training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
+The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
+Validation of hyperparameter selection was performed by randomly splitting the training/validation set into stratified training and validation sets 100 times with a 80/20 proportion [Figure 1].
+For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
+For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
+For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has.
+For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
+For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
+We validated the prediction performances of each hyperparameter setting (Figure S1].
 
@@
 Random Forest and XGBoost had the highest mean AUROC for detecting SRN, `r paste(round(mean(rf$test_aucs), 3))` (std ± `r paste(round(sd(rf$test_aucs), 3))`) and `r paste(round(mean(xgboost$test_aucs), 3))` (std ± `r paste(round(sd(xgboost$test_aucs), 3))`) respectively [Figure 2].
-L1 linear SVM and decision tree had significantly lower AUROC values, `r paste(round(mean(l1svm$test_aucs), 3))` (std ± `r paste(round(sd(l1svm$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`)  [Figure 2].
-However, they had significanly higher performances than L2 linear SVM, RBF SVM and L2 logistic regression which had mean AUROC values of `r paste(round(mean(l2svm$test_aucs), 3))` (std ± `r paste(round(sd(l2svm$test_aucs), 3))`),  `r paste(round(mean(rbf$test_aucs), 3))` (std ± `r paste(round(sd(rbf$test_aucs), 3))`) and `r paste(round(mean(logit$test_aucs), 3))` (std ± `r paste(round(sd(logit$test_aucs), 3))`) respectively [Figure 2].
+L1 linear SVM had significantly lower AUROC values, `r paste(round(mean(l1svm$test_aucs), 3))` (std ± `r paste(round(sd(l1svm$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`)  [Figure 2].
+However, it also had significanly higher performances than L2 linear SVM, RBF SVM, L2 logistic regression and decision tree which had mean AUROC values of `r paste(round(mean(l2svm$test_aucs), 3))` (std ± `r paste(round(sd(l2svm$test_aucs), 3))`),  `r paste(round(mean(rbf$test_aucs), 3))` (std ± `r paste(round(sd(rbf$test_aucs), 3))`), `r paste(round(mean(logit$test_aucs), 3))` (std ± `r paste(round(sd(logit$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`), respectively [Figure 2].
 We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC.
-We found a statistically significant difference for classifiers L2 support vector machines (SVM) with linear and radial basis function kernels and L2 logistic regression (p-values = `r paste(format(l2svm_diff$p.value, digits=2))`, `r paste(format(rbf_diff$p.value, digits=2))` and `r paste(format(logit_diff$p.value, digits=2))` respectively).
-These differences were `r paste(round(difference_model[3,1], 3))`, `r paste(round(difference_model[6,1], 3))` and `r paste(round(difference_model[4,1], 3))`, respectively [Figure 2] .
+We found that mean cross-validation and testing AUROC difference for each model was below 0.01.
 
@@
 We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data.
-To get a sense of the importance of each feature we randomly subsampled 80% of the data 100 times and trained 100 models; this resulted in 100 different sets of feature weights.
-We reported the mean weights of the 10 most important features [Figure 4].
+To get a sense of the most important features in linear models, we calculated the mean weights of all the features over the 100 data-splits for each model.
+[Figure 4].
+In the three linear models, FIT was the feature that drove the detection of SRNs, followed by eight OTUs which belong to family *Lachnospiraceae* and *Ruminococcaceae* (four with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four with negative signs; Otu00015, Otu000150, Otu00609, Otu00629).
+To get the importance of features in a decision tree, we used gini importance which allows us to use the improvement in the split-criterion is the importance measure attributed to the splitting variable
 
@@
 Linear models trained faster than non-linear models.
-L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(60*(round(walltime_df[walltime_index[1],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[1],3], 2)))`), `r paste(60*(round(walltime_df[walltime_index[2],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[2],3], 2)))`) and `r paste(60*(round(walltime_df[walltime_index[3],2], 2)))` min, (std ± `r paste(60*(round(walltime_df[walltime_index[3],3], 2)))`) respectively.
-Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[walltime_index[4],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[4],3], 2))`), `r paste((round(walltime_df[walltime_index[5],2], 2)))` hrs, (std ± `r paste((round(walltime_df[walltime_index[5],3], 2)))`), `r paste(round(walltime_df[walltime_index[6],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[6],3], 2))`) and `r paste(round(walltime_df[walltime_index[7],2], 2))` hrs, (std ± `r paste(round(walltime_df[walltime_index[7],3], 2))`), respectively.
+L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(60*(round(walltime_df[4,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[4,3], 2)))`), `r paste(60*(round(walltime_df[2,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[2,3], 2)))`) and `r paste(60*(round(walltime_df[3,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[3,3], 2)))`) respectively.
+Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[6,2], 2))` hrs, (std ± `r paste(round(walltime_df[6,3], 2))`), `r paste((round(walltime_df[1,2], 2)))` hrs, (std ± `r paste((round(walltime_df[1,3], 2)))`), `r paste(round(walltime_df[5,2], 2))` hrs, (std ± `r paste(round(walltime_df[5,3], 2))`) and `r paste(round(walltime_df[7,2], 2))` hrs, (std ± `r paste(round(walltime_df[7,3], 2))`), respectively [Figure 4].
 

### 2019-03-13 submission/manuscript.tex
update

--- 2019-03-06 submission/manuscript.tex
@@
 Instead, subsets of those communities, in relation to one another and to
-their host, account for the differences in the health outcomes.
-Therefore, researchers have started to explore the utility of machine
-learning (ML) models that use microbiota associated biomarkers to
-predict human health and to understand the microbial ecology of diseases
-such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases
-(IBD), obesity, and type 2 diabetes (1--11). ML methods are effective at
-recognizing and highlighting patterns in complex microbial datasets.
-However, currently the field's use of ML lacks clarity and consistency
-on which methods are used and how these methods are implemented.
-Moreover, there is a lack of deliberation on why a particular ML model
-is utilized. Recently, there is a trend towards using more complex ML
-models such as random forest, extreme gradient boosting and neural
-networks without a discussion on if and how much model interpretibility
-is necessary for the study (11--14). The lack of transparency on
-methodology and modeling decisions can have negative consequences on
-model reproducibility and reliability which we need avoid by (1)
-implementing consistent, accessible and transparent machine learning
-practices; (2) selecting ML models that reflect the goal of the study as
-it will inform the expectations of model accuracy, complexity,
-interpretibility and computational efficiency.
+their host, account for the differences in health outcomes. Machine
+learning (ML) methods are effective at recognizing and highlighting
+patterns in complex microbial datasets. Therefore, researchers have
+started to explore the utility of ML models that use microbiota
+associated biomarkers to predict human health and to understand the
+microbial ecology of diseases such as liver cirrhosis, colorectal
+cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes
+(1--11). However, currently the field's use of ML lacks clarity and
+consistency on which methods are used and how these methods are
+implemented. Moreover, there is a lack of deliberation on why a
+particular ML model is utilized. Recently, there is a trend towards
+using more complex ML models such as random forest, extreme gradient
+boosting and neural networks without a discussion on if and how much
+model interpretibility is necessary for the study (11--14). The lack of
+transparency on modeling methodology and model selection negatively
+impact model reproducibility and reliability. We need to strive toward
+better machine learning practices by (1) implementing consistent,
+accessible and transparent machine learning methods; (2) selecting ML
+models that reflect the goal of the study as it will inform our
+expectations of model accuracy, complexity, interpretibility and
+computational efficiency.
 
@@
 rRNA gene sequences and human hemoglobin concentrations to predict
-patients with normal or screen relevant neoplasias (SRN) disease status.
-The study had 261 normal and 229 SRN samples. We established modeling
-pipelines for L2-regularized logistic regression, L1 and L2 support
-vector machines (SVM) with linear and radial basis function kernels, a
-decision tree, random forest and XGBoost which increase in complexity
-and decrease in interpretability respectively. Our ML pipeline performed
-100 data-splits and utilized held-out test data to evaluate
-generalization and predicton performance of each ML model. The mean
-AUROC varied from 0.68 (std ± 0.05) to 0.82 (std ± 0.04). Random Forest
-and XGBoost had the highest mean AUROC for detecting SRN. Despite the
-simplicity, the L1-regularized linear kernel SVM followed Random Forest
-and XGBoost in performance. In terms of computational efficiency, L2 SVM
-with linear kernel trained the fastest (0.127 hours, std ± 0.01), while
-XGBoost took the longest (82.47 hours, std ± 17.045). We found that mean
-cross-validation and testing AUROC could vary by as much as 0.061, which
-highlights the importance of a seperate held-out test set for
-evaluation. Aside from evaluating generalization and classification
-performances for each of these models, this study established standards
-for modeling pipelines of microbiome-associated machine learning models.
+patients with normal colons or patients with colonic tumors which are
+called screen relevant neoplasias (SRN). The study had 261 normal and
+229 SRN samples. We established modeling pipelines for L2-regularized
+logistic regression, L1 and L2 support vector machines (SVM) with linear
+and radial basis function kernels, a decision tree, random forest and
+XGBoost which are ML models that increase in complexity respectively.
+Our ML pipeline utilized held-out test data to evaluate generalization
+and prediction performance of each ML model. The mean test AUROC varied
+from 0.73 (std ± 0.04) to 0.82 (std ± 0.03). Random Forest and XGBoost
+had the highest mean AUROC for detecting SRN. Despite the simplicity,
+the L1-regularized linear kernel SVM followed Random Forest and XGBoost
+in performance. In terms of computational efficiency, L2 logistic
+regression trained the fastest (0.001 hours, std ± 0.001), while XGBoost
+took the longest (0.231 hours, std ± 0). We found that mean
+cross-validation and testing AUROC varied only 0.007, which highlights
+the importance of a seperate held-out test set and consistent
+preprocessing/splitting data for evaluation. Aside from evaluating
+generalization and classification performances for each of these models,
+this study established standards for modeling pipelines of
+microbiome-associated machine learning models.
 
@@
 
-We established modeling pipelines for binary classification with
+The data has 6921 features (6920 OTUs and FIT) and a two-class label
+that defines the colonic health of the patients (SRN or normal). We
+established modeling pipelines for binary classification with
 L2-regularized logistic regression, L1 and L2 support vector machines
@@
 random forest and XGBoost to emphasize the differences in model
-accuracy, complexity, interpretibility, computational efficiency and
-scalability due to model selection. We randomly split the data into
-stratified training and test sets 100 times with a 80/20 split {[}Figure
-1{]}. The training set consisted of 393 patients (209 SRN), while the
-test set was composed of 97 patients (52 SRN). The training data was
-used for training purposes and validation of parameter selection, and
-the test set was used for evaluation purposes. Validation of parameter
-selection was repeated 100 times to get a robust parameter selection for
-each model {[}Figure 1{]}.
+accuracy, complexity, interpretibility and computational efficiency due
+to model selection. We randomly split the data into stratified
+training/validation and test sets 100 times with a 80/20 proportion
+{[}Figure 1{]}. For each data-split, training/validation set consisted
+of 393 patients (209 SRN), while the test set was composed of 97
+patients (52 SRN). The training/validation data was used for training
+purposes and validation of hyperparameter selection, and the test set
+was used for evaluation purposes. Validation of hyperparameter selection
+was performed by randomly splitting the training/validation set into
+stratified training and validation sets 100 times with a 80/20
+proportion {[}Figure 1{]}. For L2-regularized logistic regression, L1
+and L2 support vector machines (SVM) with linear and radial basis
+function kernels we tuned the \textbf{cost} hyperparameter which
+determines the regularization strength where smaller values specify
+stronger regularization. For SVM with radial basis function kernel we
+also tuned \textbf{sigma} hyperparameter which determines the reach of a
+single training instance where for a high value of sigma, the SVM
+decision boundary will be dependent on the points that are closest to
+the decision boundary. For the decision tree model, we tuned the
+\textbf{depth of the tree} where deeper the tree, the more splits it
+has. For random forest, we tuned the \textbf{number of features} to
+consider when looking for the best tree split. For xgboost, we tuned for
+\textbf{learning rate} and the \textbf{fraction of samples} to be used
+for fitting the individual base learners. We validated the prediction
+performances of each hyperparameter setting (Figure S1{]}.
 
@@
 AUROC as classification performance metric. Random Forest and XGBoost
-had the highest mean AUROC for detecting SRN, 0.816 (std ± 0.039) and
-0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM and
-decision tree had significantly lower AUROC values, 0.75 (std ± 0.045)
-and 0.741 (std ± 0.038) {[}Figure 2{]}. However, they had significanly
-higher performances than L2 linear SVM, RBF SVM and L2 logistic
-regression which had mean AUROC values of 0.68 (std ± 0.052), 0.684 (std
-± 0.049) and 0.677 (std ± 0.053) respectively {[}Figure 2{]}. We also
-evaluated the generalization performance of each classifier by comparing
-their mean cross-validation AUROC and mean testing AUROC. We found a
-statistically significant difference for classifiers L2 support vector
-machines (SVM) with linear and radial basis function kernels and L2
-logistic regression (p-values = 2.4e-16, 3.5e-13 and 9.4e-20
-respectively). These differences were 0.052, 0.047 and 0.061,
-respectively {[}Figure 2{]} .
+had the highest mean AUROC for detecting SRN, 0.82 (std ± 0.033) and
+0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM had
+significantly lower AUROC values, 0.76 (std ± 0.054) and 0.742 (std ±
+0.04) {[}Figure 2{]}. However, it also had significanly higher
+performances than L2 linear SVM, RBF SVM, L2 logistic regression and
+decision tree which had mean AUROC values of 0.74 (std ± 0.045), 0.73
+(std ± 0.041), 0.745 (std ± 0.039) and 0.742 (std ± 0.04), respectively
+{[}Figure 2{]}. We also evaluated the generalization performance of each
+classifier by comparing their mean cross-validation AUROC and mean
+testing AUROC. We found that mean cross-validation and testing AUROC
+difference for each model was below 0.01.
 
 \textbf{The complexity and interpretibility of each classifier.}
+
+We interpreted the feature weights of L1 and L2 SVM with linear kernel
+and regression coefficients of L2 logistic regression using the training
+data. To get a sense of the most important features in linear models, we
+calculated the mean weights of all the features over the 100 data-splits
+for each model. {[}Figure 4{]}. In the three linear models, FIT was the
+feature that drove the detection of SRNs, followed by eight OTUs which
+belong to family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (four
+with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four
+with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get
+the importance of features in non-linear models, we
 
@@
 Linear models trained faster than non-linear models. L2 logistic
-Rregression and L1 and L2 SVM with linear kernel had training times of
-7.8 min, (std ± 0.6), 9.6 min, (std ± 0.6) and 13.2 min, (std ± 1.2)
-respectively. Whereas, SVM with radial basis function kernel, decision
-tree, random forest and xgboost had training times of 2.21 hrs, (std ±
-0.25), 4.2 hrs, (std ± 0.49), 82.47 hrs, (std ± 17.05) and 116.83 hrs,
-(std ± 17.24), respectively.
+Rregression and L1 and L2 SVM with linear kernel had training times of 0
+min, (std ± 0), 13.8 min, (std ± 1.2) and 0 min, (std ± 0) respectively.
+Whereas, SVM with radial basis function kernel, decision tree, random
+forest and xgboost had training times of 0 hrs, (std ± 0), 0 hrs, (std ±
+0), 0 hrs, (std ± 0) and 0 hrs, (std ± 0), respectively {[}Figure 4{]}.
 

### 2019-03-18 submission/manuscript.Rmd
permutation

@@
 
+The classification models were built to classify normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations.
+The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.
+The featThe bacterial abundances are the features used to predict colonic disease status.
+Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
+There are 6920 OTUs for each sample.
+FIT levels are continuous data present for each sample.
+Because the data are in different scales, features are transformd by scaling each feature to a [0-1] range
+
+To evaluate the model performances, we randomly split the data into training/validation and test sets 100 times with a 80/20 proportion.
+Data-splits were stratified for diagnosis or SCFA concentration levels in classification models.
+For each data-split, training/validation set consisted of x patients (x carcinoma, y adenoma, z normal), while the test set was composed of x patients (x carcinoma, y adenoma, z normal).
+The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
+Validation of hyperparameter selection was performed by randomly splitting the training/validation set into stratified training and validation sets 100 times with a 80/20 proportion [Figure 1].
+
 #### Data collection

### 2019-03-18 code/learning/main.R

@@ -79,6 +79,8 @@ data$dx <- factor(data$dx)
 # We want the first sample to be a cancer so we shuffle the dataset with a specific seed to get cancer as the first sample
 set.seed(0)
 data <- data[sample(1:nrow(data)), ]
+
+
 ###################################################################
 
 ######################## RUN PIPELINE #############################

### 2019-03-18 code/learning/model_pipeline.R

@@ -141,7 +141,7 @@ pipeline <- function(dataset, model){
     feature_importance <- trained_model$finalModel$W
   }
   else{
-    feature_importance <- roc_results[3]
+    feature_importance <- roc_results[2]
   }
   # ---------------------------------------------------------------------------------->  
   

### 2019-03-18 code/learning/permutation_importance.R

@@ -7,9 +7,9 @@
 ######################################################################
 # Description: 
 
-# This script will read in data from Baxter et al. 2016
-#     - 0.03 subsampled OTU dataset
-#     - CRC metadata: SRN information
+# This script will read in:
+#     - Trained model
+#     - Pre-processed hel-out test data
 
 
 
@@ -48,29 +48,37 @@ for (dep in deps){
 ######################################################################
 
 
+
 ####################### DEFINE FUNCTION  #############################
 permutation_importance <- function(model, full){
-  # Calculate the test-auc for the 
+  # Calculate the test-auc for the actual pre-processed held-out data
   rpartProbs <- predict(model, full, type="prob")
   base_roc <- roc(ifelse(full$dx == "cancer", 1, 0), rpartProbs[[1]])
   base_auc <- base_roc$auc
   
-  imp <- c()
+  # Start the timer
   library(tictoc)
   tic("perm")
-  for (i in 1:6921){
+  # Permutate each feature in a 6921 dimensional feature vector
+  imp <- do.call('rbind', lapply(1:5, function(i){    
     full_permuted <- full 
     full_permuted[,i] <- sample(full[,i])
+    # Predict the diagnosis outcome with the one-feature-permuted test dataset
     rpartProbs_permuted <- predict(model, full_permuted, type="prob")
-    new_roc <- roc(ifelse(full_permuted$dx == "cancer", 1, 0), rpartProbs_permuted[[1]])
-    new_auc <- new_roc$auc
-    percent_increase_error=((base_auc-new_auc)/base_auc)*100
-    imp <- c(imp, percent_increase_error)
-  }
+    # Calculate the new auc
+    new_auc <- roc(ifelse(full_permuted$dx == "cancer", 1, 0), rpartProbs_permuted[[1]])$auc
+    # Return how does this feature being permuted effect the auc
+    return(((base_auc-new_auc)/base_auc)*100)
+  }))
+  # stop timer
   secs <- toc()
   walltime <- secs$toc-secs$tic
   print(walltime)
-  roc_results <- c(base_auc, new_auc, imp)
+  # save results
+  imp <- as.data.frame(imp) %>% 
+    mutate(names=colnames(full[,1:5])) %>% 
+    rename(percent_auc_change=V1)
+  roc_results <- list(base_auc, imp)
   return(roc_results)
 }
 

### 2019-05-15 submission/manuscript.Rmd
add changes par lab meeting
modulate and change

@@
 ######################################################################
-# Load in needed functions and libraries
-melt_data <-  function(data) {
-  data_melt <- data %>%
-    melt(measure.vars=c('cv_aucs', 'test_aucs')) %>%
-    rename(AUC=value) %>%
-    mutate(Performance = case_when(variable == "cv_aucs" ~ 'cross-validation', variable == "test_aucs" ~ 'testing')) %>%
-    group_by(Performance)
-  return(data_melt)
-}
+#----------------- Read in important functions -----------------#
+######################################################################
+source("../code/learning/functions.R")
 
-# Read in files as delim that are saved in a list with a pattern
-read_files <- function(filenames){
-  for(file in filenames){
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',')
-  }
-  return(data)
-}
-
-# Get model name with sub from file name
-get_model_name <- function(files){
-  pat1 <- "../data/process/walltime_"
-  name_files <- sub(pat1, "", files)
-  pat2 <- ".csv"
-  names <- sub(pat2, "", name_files)
-  return(names)
-}
-######################################################################
+#####################################################################
 
@@
 ######################################################################
-
-# Read in metadata and select only sample Id and diagnosis columns
-meta <- read.delim('../data/metadata.tsv', header=T, sep='\t') %>%
-  select(sample, dx, Dx_Bin, fit_result)
-
-# Read in OTU table and remove label and numOtus columns
-shared <- read.delim('../data/baxter.0.03.subsample.shared', header=T, sep='\t') %>%
-   select(-label, -numOtus)
-
-# Merge metadata and OTU table.
-# Group advanced adenomas and cancers together as cancer and normal, high risk normal and non-advanced adenomas as normal
-# Then remove the sample ID column
-data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
-  mutate(dx = case_when(
-    Dx_Bin== "Adenoma" ~ "normal",
-    Dx_Bin== "Normal" ~ "normal",
-    Dx_Bin== "High Risk Normal" ~ "normal",
-    Dx_Bin== "adv Adenoma" ~ "cancer",
-    Dx_Bin== "Cancer" ~ "cancer"
-  )) %>%
-  select(-sample, -Dx_Bin) %>%
-  drop_na()
-# We want the diagnosis column to be a factor
-data$dx <- factor(data$dx)
+source("../code/learning/load_data.R")
 
 ######################################################################
-#--------- Load .csv datato get mean test AUC for 7 models--------#
+#--------- Load .csv data to get mean test AUC for 7 models--------#
 ######################################################################
@@
 best_files <- list.files(path= '../data/process', pattern='combined_best.*', full.names = TRUE)
-logit <- read_files(best_files[4])
-l2svm <- read_files(best_files[3])
-l1svm <- read_files(best_files[2])
-rbf <- read_files(best_files[6])
-rf <- read_files(best_files[5])
-dt <- read_files(best_files[1])
-xgboost <- read_files(best_files[7])
+all <- map_df(best_files, read_files)
 
@@
 
-rf_xgboost <- wilcox.test(rf$test_aucs, xgboost$test_aucs)
-rf_l1svm <- wilcox.test(rf$test_aucs, l1svm$test_aucs)
-l1svm_dt <- wilcox.test(l1svm$test_aucs, dt$test_aucs)
-l1svm_rbf <- wilcox.test(l1svm$test_aucs, rbf$test_aucs)
-l1svm_l2svm <- wilcox.test(l1svm$test_aucs, l2svm$test_aucs)
-l1svm_logit <- wilcox.test(l1svm$test_aucs, logit$test_aucs)
-
-# Get the unpaired two-samples Wilcoxon test to see if cv and test differ significantly
-logit_diff <- wilcox.test(logit$test_aucs, logit$cv_aucs)
-l2svm_diff <- wilcox.test(l2svm$test_aucs, l2svm$cv_aucs)
-rbf_diff <- wilcox.test(rbf$test_aucs, rbf$cv_aucs)
+rf_xgboost <- wilcoxon_test(all, "Random_Forest", "XGBoost")
+rf_logit <- wilcoxon_test(all, "Random_Forest", "L2_Logistic_Regression")
+rf_L2svm <- wilcoxon_test(all, "Random_Forest", "L2_Linear_SVM")
+xgboost_L2svm <- wilcoxon_test(all, "XGBoost", "L2_Linear_SVM")
+logit_L2svm <- wilcoxon_test(all, "L2_Logistic_Regression", "L2_Linear_SVM")
 
 # Bind them together and summarise mean testAUC by model
-test_all <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
+test_all <- all %>%
   melt_data() %>% 
@@
 # Bind them together and summarise mean cvAUC by model
-cv_all <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
+cv_all <- all %>%
   melt_data() %>% 
@@
 # Get the order index from small to large of the differences
-difference_index <- order(difference_model$difference)
+difference_index <- order(abs(difference_model$difference))
 
@@
 Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health.
-However, this is an undertaking as human associated microbial communities are remarkably complex and uneven.
+However, this is difficult as human associated microbial communities are remarkably complex and uneven.
 It is unlikely that a single species can explain a disease.
@@
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented.
-Moreover, there is a lack of deliberation on why a particular ML model is utilized.
+Most notably, there are misleading practices of using "leaky" ML pipelines where models are tested on training data or reporting the best outcome of different iterations of cross-validation.
+Moreover, there is a lack of discussion on why a particular ML model is utilized.
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
 The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability.
-We need to strive toward better machine learning practices by (1) implementing consistent, accessible and transparent machine learning methods; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
+We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
 
-To showcase transparent ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset.
-We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients.
-We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN).
+To showcase reliable ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and a robust ML pipeline.
+We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients.
+We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN).
 The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
@@
 Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model.
-The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`).
-Random Forest and XGBoost had the highest mean AUROC for detecting SRN.
-Despite the simplicity, the L1-regularized linear kernel SVM followed Random Forest and XGBoost in performance.
+The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))` (std ± `r paste(round(test_all[performance_index[1],4], 3))`) to `r paste(round(test_all[performance_index[7],3], 3))` (std ± `r paste(round(test_all[performance_index[7],4], 3))`).
+Random forest had the highest mean AUROC for detecting SRN.
+Despite the simplicity, the L2-regularized logistic regression followed random forest in performance.
 In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
-We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing/splitting data for evaluation.
+We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
 Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
@@
 
-The data has `r paste(ncol(data)-1)` features (`r paste(ncol(data)-2)` OTUs and FIT) and a two-class label that defines the colonic health of the patients (SRN or normal).
-We established modeling pipelines for binary classification with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
+The data has `r paste(ncol(data)-1)` features (`r paste(ncol(data)-1)` OTUs) and a two-class label that defines the colonic health of the patients (SRN or normal).
+All the cases were independently labeled through colonoscopies.
+We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
 We randomly split the data into stratified training/validation and test sets 100 times with a 80/20 proportion [Figure 1].
@@
 For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
-We validated the prediction performances of each hyperparameter setting (Figure S1].
+We validated the cross-validation performances of each hyperparameter setting [Figures S1 and S2].
 
@@
 
-We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using AUROC as classification performance metric.
-Random Forest and XGBoost had the highest mean AUROC for detecting SRN, `r paste(round(mean(rf$test_aucs), 3))` (std ± `r paste(round(sd(rf$test_aucs), 3))`) and `r paste(round(mean(xgboost$test_aucs), 3))` (std ± `r paste(round(sd(xgboost$test_aucs), 3))`) respectively [Figure 2].
-L1 linear SVM had significantly lower AUROC values, `r paste(round(mean(l1svm$test_aucs), 3))` (std ± `r paste(round(sd(l1svm$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`)  [Figure 2].
-However, it also had significanly higher performances than L2 linear SVM, RBF SVM, L2 logistic regression and decision tree which had mean AUROC values of `r paste(round(mean(l2svm$test_aucs), 3))` (std ± `r paste(round(sd(l2svm$test_aucs), 3))`),  `r paste(round(mean(rbf$test_aucs), 3))` (std ± `r paste(round(sd(rbf$test_aucs), 3))`), `r paste(round(mean(logit$test_aucs), 3))` (std ± `r paste(round(sd(logit$test_aucs), 3))`) and `r paste(round(mean(dt$test_aucs), 3))` (std ± `r paste(round(sd(dt$test_aucs), 3))`), respectively [Figure 2].
+We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits in terms of the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric.
+Random Forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`) [Figure 2].
+L2 logistic regression and XGBoost had significantly lower AUROC values, `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) [Figure 2].
+Random forest also had significanly higher performances than L2 linear SVM, RBF SVM and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
 We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC.
@@
 We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data.
-To get a sense of the most important features in linear models, we calculated the mean weights of all the features over the 100 data-splits for each model.
-[Figure 4].
-In the three linear models, FIT was the feature that drove the detection of SRNs, followed by eight OTUs which belong to family *Lachnospiraceae* and *Ruminococcaceae* (four with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four with negative signs; Otu00015, Otu000150, Otu00609, Otu00629).
-To get the importance of features in a decision tree, we used gini importance which allows us to use the improvement in the split-criterion is the importance measure attributed to the splitting variable
+We calculated the mean weights of all the features over training the each model 100 times [Figure 3a].
+In the three linear models, OTUs that drove the detection of SRNs belong to family *Lachnospiraceae* and *Ruminococcaceae* (OTUs..).
+To explain the importance of features in non-linear models, we used permutation importance on the held-out test data [Figure 3b].
 

### 2019-05-15 submission/manuscript.tex
modulate and change

@@
 increasing demand for methods that identify associations between members
-of the microbiome and human health. However, this is an undertaking as
-human associated microbial communities are remarkably complex and
-uneven. It is unlikely that a single species can explain a disease.
-Instead, subsets of those communities, in relation to one another and to
-their host, account for the differences in health outcomes. Machine
-learning (ML) methods are effective at recognizing and highlighting
-patterns in complex microbial datasets. Therefore, researchers have
-started to explore the utility of ML models that use microbiota
-associated biomarkers to predict human health and to understand the
-microbial ecology of diseases such as liver cirrhosis, colorectal
-cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes
-(1--11). However, currently the field's use of ML lacks clarity and
-consistency on which methods are used and how these methods are
-implemented. Moreover, there is a lack of deliberation on why a
-particular ML model is utilized. Recently, there is a trend towards
-using more complex ML models such as random forest, extreme gradient
-boosting and neural networks without a discussion on if and how much
-model interpretibility is necessary for the study (11--14). The lack of
-transparency on modeling methodology and model selection negatively
-impact model reproducibility and reliability. We need to strive toward
-better machine learning practices by (1) implementing consistent,
-accessible and transparent machine learning methods; (2) selecting ML
+of the microbiome and human health. However, this is difficult as human
+associated microbial communities are remarkably complex and uneven. It
+is unlikely that a single species can explain a disease. Instead,
+subsets of those communities, in relation to one another and to their
+host, account for the differences in health outcomes. Machine learning
+(ML) methods are effective at recognizing and highlighting patterns in
+complex microbial datasets. Therefore, researchers have started to
+explore the utility of ML models that use microbiota associated
+biomarkers to predict human health and to understand the microbial
+ecology of diseases such as liver cirrhosis, colorectal cancer,
+inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
+However, currently the field's use of ML lacks clarity and consistency
+on which methods are used and how these methods are implemented. Most
+notably, there are misleading practices of using ``leaky'' ML pipelines
+where models are tested on training data or reporting the best outcome
+of different iterations of cross-validation. Moreover, there is a lack
+of discussion on why a particular ML model is utilized. Recently, there
+is a trend towards using more complex ML models such as random forest,
+extreme gradient boosting and neural networks without a discussion on if
+and how much model interpretibility is necessary for the study (11--14).
+The lack of transparency on modeling methodology and model selection
+negatively impact model reproducibility and reliability. We need to
+strive toward better machine learning practices by (1) implementing
+consistent and correct machine learning pipelines; (2) selecting ML
 models that reflect the goal of the study as it will inform our
@@
 
-To showcase transparent ML methodologies and to shed light on how much
-ML model selection can affect modeling results, we performed an
-empirical analysis comparing several different ML models using the same
-dataset. We used a previously published colorectal cancer (CRC) study
-(3) which had fecal 16S rRNA gene sequences and human hemoglobin
-concentrations from 490 patients. We built ML models using fecal 16S
-rRNA gene sequences and human hemoglobin concentrations to predict
-patients with normal colons or patients with colonic tumors which are
-called screen relevant neoplasias (SRN). The study had 261 normal and
-229 SRN samples. We established modeling pipelines for L2-regularized
-logistic regression, L1 and L2 support vector machines (SVM) with linear
-and radial basis function kernels, a decision tree, random forest and
-XGBoost which are ML models that increase in complexity respectively.
-Our ML pipeline utilized held-out test data to evaluate generalization
-and prediction performance of each ML model. The mean test AUROC varied
-from 0.73 (std ± 0.04) to 0.82 (std ± 0.03). Random Forest and XGBoost
-had the highest mean AUROC for detecting SRN. Despite the simplicity,
-the L1-regularized linear kernel SVM followed Random Forest and XGBoost
-in performance. In terms of computational efficiency, L2 logistic
-regression trained the fastest (0.001 hours, std ± 0.001), while XGBoost
-took the longest (0.231 hours, std ± 0). We found that mean
-cross-validation and testing AUROC varied only 0.007, which highlights
+To showcase reliable ML methodologies and to shed light on how much ML
+model selection can affect modeling results, we performed an empirical
+analysis comparing several different ML models using the same dataset
+and a robust ML pipeline. We used a previously published colorectal
+cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
+patients. We built ML models using fecal 16S rRNA gene sequences to
+predict patients with normal colons or patients with colonic tumors
+which are called screen relevant neoplasias (SRN). The study had 261
+normal and 229 SRN samples. We established modeling pipelines for
+L2-regularized logistic regression, L1 and L2 support vector machines
+(SVM) with linear and radial basis function kernels, a decision tree,
+random forest and XGBoost which are ML models that increase in
+complexity respectively. Our ML pipeline utilized held-out test data to
+evaluate generalization and prediction performance of each ML model. The
+mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ± 0.037).
+Random forest had the highest mean AUROC for detecting SRN. Despite the
+simplicity, the L2-regularized logistic regression followed random
+forest in performance. In terms of computational efficiency, L2 logistic
+regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
+took the longest (162.843 hours, std ± 10.018). We found that mean
+cross-validation and testing AUROC varied only 0.01, which highlights
 the importance of a seperate held-out test set and consistent
-preprocessing/splitting data for evaluation. Aside from evaluating
+preprocessing of the data prior to evaluation. Aside from evaluating
 generalization and classification performances for each of these models,
@@
 
-The data has 6921 features (6920 OTUs and FIT) and a two-class label
-that defines the colonic health of the patients (SRN or normal). We
+The data has 6920 features (6920 OTUs) and a two-class label that
+defines the colonic health of the patients (SRN or normal). We
 established modeling pipelines for binary classification with
@@
 models when applied to held-out test data over 100 data-splits using
-AUROC as classification performance metric. Random Forest and XGBoost
-had the highest mean AUROC for detecting SRN, 0.82 (std ± 0.033) and
-0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM had
-significantly lower AUROC values, 0.76 (std ± 0.054) and 0.742 (std ±
-0.04) {[}Figure 2{]}. However, it also had significanly higher
-performances than L2 linear SVM, RBF SVM, L2 logistic regression and
-decision tree which had mean AUROC values of 0.74 (std ± 0.045), 0.73
-(std ± 0.041), 0.745 (std ± 0.039) and 0.742 (std ± 0.04), respectively
+AUROC as classification performance metric. Random Forest had the
+highest mean AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure
+2{]}. L2 logistic regression and XGBoost had significantly lower AUROC
+values, 0.676 (std ± 0.042) and 0.678 (std ± 0.04) {[}Figure 2{]}.
+Random forest also had significanly higher performances than L2 linear
+SVM, RBF SVM and decision tree which had mean AUROC values of 0.671 (std
+± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044), respectively
 {[}Figure 2{]}. We also evaluated the generalization performance of each
@@
 belong to family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (four
-with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four
+with positive signs; Otu01239, Otu00742, Otu00008, Otu00659, and four
 with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get
-the importance of features in non-linear models, we
+the importance of features in a decision tree, we used gini importance
+which allows us to use the improvement in the split-criterion is the
+importance measure attributed to the splitting variable
 
@@
 Linear models trained faster than non-linear models. L2 logistic
-Rregression and L1 and L2 SVM with linear kernel had training times of 0
-min, (std ± 0), 13.8 min, (std ± 1.2) and 0 min, (std ± 0) respectively.
-Whereas, SVM with radial basis function kernel, decision tree, random
-forest and xgboost had training times of 0 hrs, (std ± 0), 0 hrs, (std ±
-0), 0 hrs, (std ± 0) and 0 hrs, (std ± 0), respectively {[}Figure 4{]}.
+Rregression and L1 and L2 SVM with linear kernel had training times of
+13.2 min, (std ± 1.8), 563.4 min, (std ± 46.8) and 12.6 min, (std ± 1.8)
+respectively. Whereas, SVM with radial basis function kernel, decision
+tree, random forest and xgboost had training times of 101.32 hrs, (std ±
+10.02), 162.84 hrs, (std ± 3.99), 0.2 hrs, (std ± 0.02) and 64.71 hrs,
+(std ± 9.85), respectively {[}Figure 4{]}.
 
@@
 \subsection{Materials and Methods}\label{materials-and-methods}
+
+The classification models were built to classify normal versus adenoma,
+normal versus carcinoma, and high versus low SCFA concentrations. The
+regression models were built to classify the SCFA concentrations of
+acetate, butyrate, and propionate regardless of disease status. The
+featThe bacterial abundances are the features used to predict colonic
+disease status. Bacterial abundances are discrete data in the form of
+Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
+sample. FIT levels are continuous data present for each sample. Because
+the data are in different scales, features are transformd by scaling
+each feature to a {[}0-1{]} range
+
+To evaluate the model performances, we randomly split the data into
+training/validation and test sets 100 times with a 80/20 proportion.
+Data-splits were stratified for diagnosis or SCFA concentration levels
+in classification models. For each data-split, training/validation set
+consisted of x patients (x carcinoma, y adenoma, z normal), while the
+test set was composed of x patients (x carcinoma, y adenoma, z normal).
+The training/validation data was used for training purposes and
+validation of hyperparameter selection, and the test set was used for
+evaluation purposes. Validation of hyperparameter selection was
+performed by randomly splitting the training/validation set into
+stratified training and validation sets 100 times with a 80/20
+proportion {[}Figure 1{]}.
 

### 2019-05-15 submission/manuscript.Rmd
manuscript update with fig 3

@@
 min_fixed_result <- list()
+
 for(i in result){
@@
 
-To showcase reliable ML methodologies and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and a robust ML pipeline.
+To showcase reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline.
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients.
@@
 The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
-We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost which are ML models that increase in complexity respectively.
+We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
 Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model.
@@
 Despite the simplicity, the L2-regularized logistic regression followed random forest in performance.
-In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`).
+In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`).
 We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
@@
 We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
-We randomly split the data into stratified training/validation and test sets 100 times with a 80/20 proportion [Figure 1].
+We randomly split the data into training/validation and test sets 100 times with a 80/20 proportion [Figure 1].
+Since the colonic health is not uniformly represented in the data, data-splits are stratified to maintain the overall label distribution on the training set.
 For each data-split, training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
 The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
-Validation of hyperparameter selection was performed by randomly splitting the training/validation set into stratified training and validation sets 100 times with a 80/20 proportion [Figure 1].
-For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
-For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
-For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has.
-For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
-For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
-We validated the cross-validation performances of each hyperparameter setting [Figures S1 and S2].
+Validation of hyperparameter selection was performed using 100 randomizations of five-fold cross-validation on the training/validation set [Figure 1].
+Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets.
+We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting to train the full training/validation dataset [Figures S1 and S2].
+We then used the held-out test set to evaluate the prediction performance of each ML model.
 
@@
 
-We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits in terms of the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric.
+We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric.
 Random Forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`) [Figure 2].
@@
 We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data.
-We calculated the mean weights of all the features over training the each model 100 times [Figure 3a].
-In the three linear models, OTUs that drove the detection of SRNs belong to family *Lachnospiraceae* and *Ruminococcaceae* (OTUs..).
+We calculated the mean weights of all the features fromt the 100 randomizations of five-fold cross-validation [Figure 3a].
+In the three linear models, OTUs that drove the detection of SRNs belong to family *Lachnospiraceae* and *Ruminococcaceae* (OTUs x, y, z).
 To explain the importance of features in non-linear models, we used permutation importance on the held-out test data [Figure 3b].
+Perfectly correlated OTUs were grouped and permuted together to get a better representation of their effect in the prediction performance of the model.
+Non-correlated OTUs were permuted individually.
 
@@
 Linear models trained faster than non-linear models.
-L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(60*(round(walltime_df[4,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[4,3], 2)))`), `r paste(60*(round(walltime_df[2,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[2,3], 2)))`) and `r paste(60*(round(walltime_df[3,2], 2)))` min, (std ± `r paste(60*(round(walltime_df[3,3], 2)))`) respectively.
-Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[6,2], 2))` hrs, (std ± `r paste(round(walltime_df[6,3], 2))`), `r paste((round(walltime_df[1,2], 2)))` hrs, (std ± `r paste((round(walltime_df[1,3], 2)))`), `r paste(round(walltime_df[5,2], 2))` hrs, (std ± `r paste(round(walltime_df[5,3], 2))`) and `r paste(round(walltime_df[7,2], 2))` hrs, (std ± `r paste(round(walltime_df[7,3], 2))`), respectively [Figure 4].
+L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(walltime_df[walltime_index[1],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[1],3], 2))`), `r paste(round(walltime_df[walltime_index[2],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[2],3],2))`), and `r paste(round(walltime_df[walltime_index[3],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[3],3], 2))`), respectively.
+Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[walltime_index[4],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[4],3], 1))`), `r paste(round(walltime_df[walltime_index[5],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[5],3], 1))`), `r paste(round(walltime_df[walltime_index[6],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[6],3], 1))`) and `r paste(round(walltime_df[walltime_index[7],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[7],3], 1))`), respectively [Figure 4].
 
@@
 
-The classification models were built to classify normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations.
-The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.
-The featThe bacterial abundances are the features used to predict colonic disease status.
-Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
-There are 6920 OTUs for each sample.
-FIT levels are continuous data present for each sample.
-Because the data are in different scales, features are transformd by scaling each feature to a [0-1] range
-
-To evaluate the model performances, we randomly split the data into training/validation and test sets 100 times with a 80/20 proportion.
-Data-splits were stratified for diagnosis or SCFA concentration levels in classification models.
-For each data-split, training/validation set consisted of x patients (x carcinoma, y adenoma, z normal), while the test set was composed of x patients (x carcinoma, y adenoma, z normal).
-The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
-Validation of hyperparameter selection was performed by randomly splitting the training/validation set into stratified training and validation sets 100 times with a 80/20 proportion [Figure 1].
-
-#### Data collection
-  The data used for this analysis are stool bacterial abundances, stool hemoglobin levels and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study.
+#### Data collection and study population
+  The data used for this analysis are stool bacterial abundances and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study.
 These data were obtained from Sze et al [@sze_leveraging_2018].
@@
 Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas.
-Stool provided by the patients was used for Fecal Immunological Tests (FIT) which measure human hemoglobin concentrations and for 16S rRNA gene sequencing to measure bacterial population abundances.
+Stool provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances.
 The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].
@@
 
-  The colonic disease status is re-defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs).
+  The colonic health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs).
 Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas.
-Colonic disease status is the label predicted with each classifier.
-The bacterial abundances and FIT results are the features used to predict colonic disease status.
+The bacterial abundances are the features used to predict colonic health of the patients.
 Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
-There are 6920 OTUs for each sample.
-FIT levels are continuous data present for each sample.
-Because the data are in different scales, features are transformd by scaling each feature to a [0-1] range (Table 1).
+OTU counts were set to the size of our smallest sample and were subsampled at the same distances.
+They were then transformd by scaling to a [0-1] range.
 
-#### Learning the Classifier
+#### Model training and evaluation
 
-  To train and validate our model, labeled data is randomly split 80/20 into a training set and testing
-set.
-Then, seven binary class classifiers, L2 logistic regression, L1 and L2 linear suppor vector machines (SVM), radial basis function SVM, decision tree, random forest and XGBoost, are learned.
-The training set is used for training purposes and validation of hyperparameter selection, and the test set is used for evaluation purposes.
-Hyperparameters are selected using 5-fold cross-validation with 100-repeats on the training set.
-Since the colonic disease status are not uniformly represented in the data, 5-fold splits are stratified to maintain the overall label distribution on the training set.
+For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
+For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
+For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has.
+For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
+For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
+Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).
 
-#### Classifier Performance
+**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1).
+We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best discriminative performance.
 
-  The classification performance of learned classifier is evaluated on the labeled held-out testing set.
-The optimal  classifier with optimal hyperparameters selected in the cross-validation step is used to produce a prediction for the testing set.
-The performance of this prediction is measured in terms of the sensitivity and specificity, in addition to Area Under the Curve (AUC) metrics.
-This process of splitting the data, learning a classifier with cross-validation, and testing the classifier is repeated on 100 different splits.
-In the end cross-validation AUC and testing AUC averaged over the 100 different training/test splits are reported.
-Hyperparameter budget and performance for each split is also reported.
+**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.
 

### 2019-05-15 submission/manuscript.tex
manuscript update with fig 3

@@
 strive toward better machine learning practices by (1) implementing
-consistent and correct machine learning pipelines; (2) selecting ML
+consistent and reliable machine learning pipelines; (2) selecting ML
 models that reflect the goal of the study as it will inform our
@@
 
-To showcase reliable ML methodologies and to shed light on how much ML
-model selection can affect modeling results, we performed an empirical
+To showcase reliable ML pipeline and to shed light on how much ML model
+selection can affect modeling results, we performed an empirical
 analysis comparing several different ML models using the same dataset
-and a robust ML pipeline. We used a previously published colorectal
+and with a robust ML pipeline. We used a previously published colorectal
 cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
@@
 (SVM) with linear and radial basis function kernels, a decision tree,
-random forest and XGBoost which are ML models that increase in
-complexity respectively. Our ML pipeline utilized held-out test data to
-evaluate generalization and prediction performance of each ML model. The
-mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ± 0.037).
-Random forest had the highest mean AUROC for detecting SRN. Despite the
-simplicity, the L2-regularized logistic regression followed random
-forest in performance. In terms of computational efficiency, L2 logistic
-regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
-took the longest (162.843 hours, std ± 10.018). We found that mean
-cross-validation and testing AUROC varied only 0.01, which highlights
-the importance of a seperate held-out test set and consistent
+random forest and XGBoost. Our ML pipeline utilized held-out test data
+to evaluate generalization and prediction performance of each ML model.
+The mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ±
+0.037). Random forest had the highest mean AUROC for detecting SRN.
+Despite the simplicity, the L2-regularized logistic regression followed
+random forest in performance. In terms of computational efficiency, L2
+logistic regression trained the fastest (0.202 hours, std ± 0.019),
+while XGBoost took the longest (162.843 hours, std ± 3.986). We found
+that mean cross-validation and testing AUROC varied only 0.01, which
+highlights the importance of a seperate held-out test set and consistent
 preprocessing of the data prior to evaluation. Aside from evaluating
@@
 The data has 6920 features (6920 OTUs) and a two-class label that
-defines the colonic health of the patients (SRN or normal). We
-established modeling pipelines for binary classification with
-L2-regularized logistic regression, L1 and L2 support vector machines
-(SVM) with linear and radial basis function kernels, a decision tree,
-random forest and XGBoost to emphasize the differences in model
-accuracy, complexity, interpretibility and computational efficiency due
-to model selection. We randomly split the data into stratified
-training/validation and test sets 100 times with a 80/20 proportion
-{[}Figure 1{]}. For each data-split, training/validation set consisted
-of 393 patients (209 SRN), while the test set was composed of 97
-patients (52 SRN). The training/validation data was used for training
-purposes and validation of hyperparameter selection, and the test set
-was used for evaluation purposes. Validation of hyperparameter selection
-was performed by randomly splitting the training/validation set into
-stratified training and validation sets 100 times with a 80/20
-proportion {[}Figure 1{]}. For L2-regularized logistic regression, L1
-and L2 support vector machines (SVM) with linear and radial basis
-function kernels we tuned the \textbf{cost} hyperparameter which
-determines the regularization strength where smaller values specify
-stronger regularization. For SVM with radial basis function kernel we
-also tuned \textbf{sigma} hyperparameter which determines the reach of a
-single training instance where for a high value of sigma, the SVM
-decision boundary will be dependent on the points that are closest to
-the decision boundary. For the decision tree model, we tuned the
-\textbf{depth of the tree} where deeper the tree, the more splits it
-has. For random forest, we tuned the \textbf{number of features} to
-consider when looking for the best tree split. For xgboost, we tuned for
-\textbf{learning rate} and the \textbf{fraction of samples} to be used
-for fitting the individual base learners. We validated the prediction
-performances of each hyperparameter setting (Figure S1{]}.
+defines the colonic health of the patients (SRN or normal). All the
+cases were independently labeled through colonoscopies. We established
+modeling pipelines for a binary prediction task with L2-regularized
+logistic regression, L1 and L2 support vector machines (SVM) with linear
+and radial basis function kernels, a decision tree, random forest and
+XGBoost to emphasize the differences in model accuracy, complexity,
+interpretibility and computational efficiency due to model selection. We
+randomly split the data into training/validation and test sets 100 times
+with a 80/20 proportion {[}Figure 1{]}. Since the colonic health is not
+uniformly represented in the data, data-splits are stratified to
+maintain the overall label distribution on the training set. For each
+data-split, training/validation set consisted of 393 patients (209 SRN),
+while the test set was composed of 97 patients (52 SRN). The
+training/validation data was used for training purposes and validation
+of hyperparameter selection, and the test set was used for evaluation
+purposes. Validation of hyperparameter selection was performed using 100
+randomizations of five-fold cross-validation on the training/validation
+set {[}Figure 1{]}. Similar to the initial data-split, five-fold
+cross-validation was also stratified to maintain the overall label
+distribution on the training and validation sets. We validated the
+cross-validation performances of each hyperparameter setting over the
+100 randomizations and selected the best performing hyper-parameter
+setting to train the full training/validation dataset {[}Figures S1 and
+S2{]}. We then used the held-out test set to evaluate the prediction
+performance of each ML model.
 
@@
 We evaluated the prediction performance of seven binary classification
-models when applied to held-out test data over 100 data-splits using
-AUROC as classification performance metric. Random Forest had the
-highest mean AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure
-2{]}. L2 logistic regression and XGBoost had significantly lower AUROC
-values, 0.676 (std ± 0.042) and 0.678 (std ± 0.04) {[}Figure 2{]}.
-Random forest also had significanly higher performances than L2 linear
-SVM, RBF SVM and decision tree which had mean AUROC values of 0.671 (std
-± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044), respectively
-{[}Figure 2{]}. We also evaluated the generalization performance of each
-classifier by comparing their mean cross-validation AUROC and mean
-testing AUROC. We found that mean cross-validation and testing AUROC
-difference for each model was below 0.01.
+models when applied to held-out test data over 100 data-splits using the
+area under the receiver operating characteristic curve (AUROC) as the
+discriminative performance metric. Random Forest had the highest mean
+AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure 2{]}. L2 logistic
+regression and XGBoost had significantly lower AUROC values, 0.676 (std
+± 0.042) and 0.678 (std ± 0.04) {[}Figure 2{]}. Random forest also had
+significanly higher performances than L2 linear SVM, RBF SVM and
+decision tree which had mean AUROC values of 0.671 (std ± 0.045), 0.663
+(std ± 0.044), and 0.598 (std ± 0.044), respectively {[}Figure 2{]}. We
+also evaluated the generalization performance of each classifier by
+comparing their mean cross-validation AUROC and mean testing AUROC. We
+found that mean cross-validation and testing AUROC difference for each
+model was below 0.01.
 
@@
 and regression coefficients of L2 logistic regression using the training
-data. To get a sense of the most important features in linear models, we
-calculated the mean weights of all the features over the 100 data-splits
-for each model. {[}Figure 4{]}. In the three linear models, FIT was the
-feature that drove the detection of SRNs, followed by eight OTUs which
-belong to family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (four
-with positive signs; Otu01239, Otu00742, Otu00008, Otu00659, and four
-with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get
-the importance of features in a decision tree, we used gini importance
-which allows us to use the improvement in the split-criterion is the
-importance measure attributed to the splitting variable
+data. We calculated the mean weights of all the features fromt the 100
+randomizations of five-fold cross-validation {[}Figure 3a{]}. In the
+three linear models, OTUs that drove the detection of SRNs belong to
+family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (OTUs x, y, z).
+To explain the importance of features in non-linear models, we used
+permutation importance on the held-out test data {[}Figure 3b{]}.
+Perfectly correlated OTUs were grouped and permuted together to get a
+better representation of their effect in the prediction performance of
+the model. Non-correlated OTUs were permuted individually.
 
@@
 Rregression and L1 and L2 SVM with linear kernel had training times of
-13.2 min, (std ± 1.8), 563.4 min, (std ± 46.8) and 12.6 min, (std ± 1.8)
-respectively. Whereas, SVM with radial basis function kernel, decision
-tree, random forest and xgboost had training times of 101.32 hrs, (std ±
-10.02), 162.84 hrs, (std ± 3.99), 0.2 hrs, (std ± 0.02) and 64.71 hrs,
-(std ± 9.85), respectively {[}Figure 4{]}.
+0.2 hours, (std ± 0.02), 0.2 hours, (std ± 0.03), and 0.2 hours, (std ±
+0.03), respectively. Whereas, SVM with radial basis function kernel,
+decision tree, random forest and xgboost had training times of 9.4
+hours, (std ± 0.8), 64.7 hours, (std ± 9.9), 101.3 hours, (std ± 10) and
+162.8 hours, (std ± 4), respectively {[}Figure 4{]}.
 
@@
 
-The classification models were built to classify normal versus adenoma,
-normal versus carcinoma, and high versus low SCFA concentrations. The
-regression models were built to classify the SCFA concentrations of
-acetate, butyrate, and propionate regardless of disease status. The
-featThe bacterial abundances are the features used to predict colonic
-disease status. Bacterial abundances are discrete data in the form of
-Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
-sample. FIT levels are continuous data present for each sample. Because
-the data are in different scales, features are transformd by scaling
-each feature to a {[}0-1{]} range
+\paragraph{Data collection and study
+population}\label{data-collection-and-study-population}
 
-To evaluate the model performances, we randomly split the data into
-training/validation and test sets 100 times with a 80/20 proportion.
-Data-splits were stratified for diagnosis or SCFA concentration levels
-in classification models. For each data-split, training/validation set
-consisted of x patients (x carcinoma, y adenoma, z normal), while the
-test set was composed of x patients (x carcinoma, y adenoma, z normal).
-The training/validation data was used for training purposes and
-validation of hyperparameter selection, and the test set was used for
-evaluation purposes. Validation of hyperparameter selection was
-performed by randomly splitting the training/validation set into
-stratified training and validation sets 100 times with a 80/20
-proportion {[}Figure 1{]}.
-
-\paragraph{Data collection}\label{data-collection}
-
-The data used for this analysis are stool bacterial abundances, stool
-hemoglobin levels and clinical information of the patients recruited by
-Great Lakes-New England Early Detection Research Network study. These
-data were obtained from Sze et al (15). The stool samples were provided
-by recruited adult participants who were undergoing scheduled screening
-or surveillance colonoscopy. Colonoscopies were performed and fecal
-samples were collected from participants in four locations: Toronto (ON,
-Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
-Patients' colonic disease status was defined by colonoscopy with
-adequate preparation and tissue histopathology of all resected lesions.
-Patients with an adenoma greater than 1 cm, more than three adenomas of
-any size, or an adenoma with villous histology were classified as
-advanced adenoma. Study had 172 patients with normal colonoscopies, 198
-with adenomas and 120 with carcinomas. Of the 198 adenomas, 109 were
+The data used for this analysis are stool bacterial abundances and
+clinical information of the patients recruited by Great Lakes-New
+England Early Detection Research Network study. These data were obtained
+from Sze et al (15). The stool samples were provided by recruited adult
+participants who were undergoing scheduled screening or surveillance
+colonoscopy. Colonoscopies were performed and fecal samples were
+collected from participants in four locations: Toronto (ON, Canada),
+Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
+colonic disease status was defined by colonoscopy with adequate
+preparation and tissue histopathology of all resected lesions. Patients
+with an adenoma greater than 1 cm, more than three adenomas of any size,
+or an adenoma with villous histology were classified as advanced
+adenoma. Study had 172 patients with normal colonoscopies, 198 with
+adenomas and 120 with carcinomas. Of the 198 adenomas, 109 were
 identified as advanced adenomas. Stool provided by the patients was used
-for Fecal Immunological Tests (FIT) which measure human hemoglobin
-concentrations and for 16S rRNA gene sequencing to measure bacterial
-population abundances. The bacterial abundance data was generated by Sze
-et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the
-default quality filtering methods, identifying and removing chimeric
-sequences using VSEARCH and assigning to OTUs at 97\% similarity using
-the OptiClust algorithm (16--18).
+for 16S rRNA gene sequencing to measure bacterial population abundances.
+The bacterial abundance data was generated by Sze et al, by processing
+16S rRNA sequences in Mothur (v1.39.3) using the default quality
+filtering methods, identifying and removing chimeric sequences using
+VSEARCH and assigning to OTUs at 97\% similarity using the OptiClust
+algorithm (16--18).
 
@@
 
-The colonic disease status is re-defined as two encompassing classes;
-Normal or Screen Relevant Neoplasias (SRNs). Normal class includes
-patients with non-advanced adenomas or normal colons whereas SRN class
-includes patients with advanced adenomas or carcinomas. Colonic disease
-status is the label predicted with each classifier. The bacterial
-abundances and FIT results are the features used to predict colonic
-disease status. Bacterial abundances are discrete data in the form of
-Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
-sample. FIT levels are continuous data present for each sample. Because
-the data are in different scales, features are transformd by scaling
-each feature to a {[}0-1{]} range (Table 1).
+The colonic health of the patient was defined as two encompassing
+classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
+includes patients with non-advanced adenomas or normal colons whereas
+SRN class includes patients with advanced adenomas or carcinomas. The
+bacterial abundances are the features used to predict colonic health of
+the patients. Bacterial abundances are discrete data in the form of
+Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size
+of our smallest sample and were subsampled at the same distances. They
+were then transformd by scaling to a {[}0-1{]} range.
 
-\paragraph{Learning the Classifier}\label{learning-the-classifier}
+\paragraph{Model training and
+evaluation}\label{model-training-and-evaluation}
 
-To train and validate our model, labeled data is randomly split 80/20
-into a training set and testing set. Then, seven binary class
-classifiers, L2 logistic regression, L1 and L2 linear suppor vector
-machines (SVM), radial basis function SVM, decision tree, random forest
-and XGBoost, are learned. The training set is used for training purposes
-and validation of hyperparameter selection, and the test set is used for
-evaluation purposes. Hyperparameters are selected using 5-fold
-cross-validation with 100-repeats on the training set. Since the colonic
-disease status are not uniformly represented in the data, 5-fold splits
-are stratified to maintain the overall label distribution on the
-training set.
+For L2-regularized logistic regression, L1 and L2 support vector
+machines (SVM) with linear and radial basis function kernels we tuned
+the \textbf{cost} hyperparameter which determines the regularization
+strength where smaller values specify stronger regularization. For SVM
+with radial basis function kernel we also tuned \textbf{sigma}
+hyperparameter which determines the reach of a single training instance
+where for a high value of sigma, the SVM decision boundary will be
+dependent on the points that are closest to the decision boundary. For
+the decision tree model, we tuned the \textbf{depth of the tree} where
+deeper the tree, the more splits it has. For random forest, we tuned the
+\textbf{number of features} to consider when looking for the best tree
+split. For xgboost, we tuned for \textbf{learning rate} and the
+\textbf{fraction of samples} to be used for fitting the individual base
+learners. Models were trained using the machine learning wrapper caret
+package (v.6.0.81) in R (v.3.5.0).
 
-\paragraph{Classifier Performance}\label{classifier-performance}
+\textbf{Statistical analysis workflow.} Data summaries, statistical
+analysis, and data visualizations were performed using R (v.3.5.0) with
+the tidyverse package (v.1.2.1). We compared the AUROC values of the
+seven ML models by Wilcoxon rank sum tests to determine the best
+discriminative performance.
 
-The classification performance of learned classifier is evaluated on the
-labeled held-out testing set. The optimal classifier with optimal
-hyperparameters selected in the cross-validation step is used to produce
-a prediction for the testing set. The performance of this prediction is
-measured in terms of the sensitivity and specificity, in addition to
-Area Under the Curve (AUC) metrics. This process of splitting the data,
-learning a classifier with cross-validation, and testing the classifier
-is repeated on 100 different splits. In the end cross-validation AUC and
-testing AUC averaged over the 100 different training/test splits are
-reported. Hyperparameter budget and performance for each split is also
-reported.
+\textbf{Code availability.} The code for all sequence curation and
+analysis steps including an Rmarkdown version of this manuscript is
+available at
+\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/}.
 

### 2019-05-15 code/learning/Figure3a_3b.R

@@ -255,19 +255,24 @@ base_nonlin_plot <-  function(data, name){
   data_full <- read_files(paste0("data/process/combined_all_imp_features_non_cor_results_", name, ".csv")) %>%
     # Only keep the OTUs and their AUCs for the ones that are in the top 5 changed (decreased the most) ones
     filter(names %in% data_first_ten$names) %>% 
-    # Add the base test auc to the dataframe
-    bind_rows(data_base) %>% 
     group_by(names)
   # Keep the 5 OTU names as a list to use in the plot as factors
   otus <- data_first_ten$names %>% 
     droplevels() %>% 
     as.character()
   # Base auc at the top, then followed by the most changed OTU, followed by others ordered by median
-  data_full$names <- factor(data_full$names,levels = c(otus[5], otus[4], otus[3], otus[2], otus[1], "base_auc"))
+  data_full$names <- factor(data_full$names,levels = c(otus[5], otus[4], otus[3], otus[2], otus[1]))
   # Plot boxplot
-  plot <- ggplot(data_full, aes(x=names, y=new_auc)) +
-    geom_boxplot(alpha=0.7, fill="#E69F00") +
+  lowerq = quantile(data_base$new_auc)[2]
+  upperq = quantile(data_base$new_auc)[4]
+
+  plot <- ggplot() +
+    geom_boxplot(data=data_full, aes(x=names, y=new_auc), fill="#E69F00") +
+    geom_rect(aes(ymin=lowerq, ymax=upperq, xmin=0, xmax=Inf), fill="grey") +
+    geom_boxplot(data=data_full, aes(x=names, y=new_auc), fill="#E69F00") +
     geom_hline(yintercept = data_base_means$imp , linetype="dashed") +
+    #geom_hline(yintercept = upperq, alpha=0.5) +
+    #geom_hline(yintercept = lowerq, alpha=0.5) +
     coord_flip() +
     theme_classic() +
     scale_y_continuous(name = " AUROC with the OTU permuted randomly", 

### 2019-05-16 submission/manuscript.Rmd
prep data and figures in the submission folder

@@
 ######################################################################
-#--------- Load .csv data to get mean test AUC for 7 models--------#
+#----------------- Load feature importance data -----------------#
+# Median of base testing AUC for each model
+# 100 datasplit permutation importances new testing AUC for top 5 features
 ######################################################################
 
-# Read in the cvAUCs, testAUCs for 100 splits.
-best_files <- list.files(path= '../data/process', pattern='combined_best.*', full.names = TRUE)
-all <- map_df(best_files, read_files)
-
-# Get the The unpaired two-samples Wilcoxon test to see if models differ from one another signigicantly
-
-rf_xgboost <- wilcoxon_test(all, "Random_Forest", "XGBoost")
-rf_logit <- wilcoxon_test(all, "Random_Forest", "L2_Logistic_Regression")
-rf_L2svm <- wilcoxon_test(all, "Random_Forest", "L2_Linear_SVM")
-xgboost_L2svm <- wilcoxon_test(all, "XGBoost", "L2_Linear_SVM")
-logit_L2svm <- wilcoxon_test(all, "L2_Logistic_Regression", "L2_Linear_SVM")
-
-# Bind them together and summarise mean testAUC by model
-test_all <- all %>%
-  melt_data() %>% 
-  select(-variable) %>% 
-  group_by(model, Performance) %>% 
-  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
-  filter(Performance=="testing")
-
-# Get the order index from small to large meanAUC
-performance_index <- order(test_all$mean_AUC)
-
-# Bind them together and summarise mean cvAUC by model
-cv_all <- all %>%
-  melt_data() %>% 
-  select(-variable) %>% 
-  group_by(model, Performance) %>% 
-  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
-  filter(Performance=="cross-validation")
-
-# Get the difference between mean cvAUC and testAUC for each model
-difference <- cv_all$mean_AUC - test_all$mean_AUC[match(cv_all$model, test_all$model)]
-difference_model <- data.frame(difference, test_all$model)
-# Get the order index from small to large of the differences
-difference_index <- order(abs(difference_model$difference))
+source("../code/learning/load_imp_data.R")
 
 ######################################################################
-#---------- Load .csv datato get mean walltime for 7 models-------#
+#-------------------------- Load walltime data -----------------#
 ######################################################################
+source("../code/learning/load_walltime.R")
 
-# Read in the walltime for each split.
-walltime_files <- list.files(path= '../data/process', pattern='walltime*', full.names = TRUE)
-
-# ------- 1. In a loop read the files as delim
-# --------2. get the model name for each with get_model_name() 
-#---------3. Add a column with model name to each delim
-result <- list()
-
-for(file in walltime_files){
-  model_walltime <- read_files(file) %>%  
-  #summarise_walltime() %>% 
-  mutate(model=get_model_name(file))
-  result[[length(result)+1]] <- model_walltime
-}
-
-# ------- 1. If models are L1 or L2 Linear SVM or L2 Logit then the walltime is in minutes
-# --------2. If models are Random Forest or XGBoost then the walltime is in days
-#---------3. If other models, the walltime is in hours.
-#---------4. This loop converts all to hours.
-min_fixed_result <- list()
-
-for(i in result){
-  i$x <- i$x/3600 # The walltimes were saved as seconds we covert to hours
-  min_fixed_result[[length(min_fixed_result)+1]] <- i
-}
-# Bind all delims with model name and walltime together
-# Summarise mean walltime
-walltime_df <- bind_rows(min_fixed_result) %>% 
-  group_by(model) %>% 
-  summarise(mean_walltime=mean(x), sd_walltime=sd(x))
-
-# Get the sorted mean walltime for each model small to large
-walltime_index <- order(walltime_df$mean_walltime)
-
+######################################################################
+#---------------------- Load AUROC performance data -----------------#
+######################################################################
+source("../code/learning/load_AUROC_data.R")
 ```
-
 \vspace{35mm}
@@
 
-Begüm D. Topçuoğlu${^1}$, Jenna Wiens${^2}$, Patrick D. Schloss^1$\dagger$^
+Begüm D. Topçuoğlu${^1}$, Mack Ruffin${^2}$, Jenna Wiens${^3}$, Patrick D. Schloss^1$\dagger$^
 
@@
 
-2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109
+2\. Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
+
+3\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109
 
@@
 
-__Model selection and construction__
+__Model selection and pipeline construction__
 
-The data has `r paste(ncol(data)-1)` features (`r paste(ncol(data)-1)` OTUs) and a two-class label that defines the colonic health of the patients (SRN or normal).
+We used a cohort of 490 patients with 261 cases of SRN.
+For each patient fecal sample, we had `r paste(ncol(data)-1)` features (`r paste(ncol(data)-1)` OTUs) and a two-class label that defines the colonic health of the patients (SRN or normal).
 All the cases were independently labeled through colonoscopies.
 We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
-We randomly split the data into training/validation and test sets 100 times with a 80/20 proportion [Figure 1].
-Since the colonic health is not uniformly represented in the data, data-splits are stratified to maintain the overall label distribution on the training set.
-For each data-split, training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
+We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
+Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution on the training set.
+Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
 The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
@@
 We then used the held-out test set to evaluate the prediction performance of each ML model.
+The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance [Figure 1].
 
@@
 Random Forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`) [Figure 2].
-L2 logistic regression and XGBoost had significantly lower AUROC values, `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) [Figure 2].
+L2 logistic regression and XGBoost followed random forest however they had significantly lower AUROC values, `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) [Figure 2].
 Random forest also had significanly higher performances than L2 linear SVM, RBF SVM and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
@@
 
-We interpreted the feature weights of L1 and L2 SVM with linear kernel and regression coefficients of L2 logistic regression using the training data.
-We calculated the mean weights of all the features fromt the 100 randomizations of five-fold cross-validation [Figure 3a].
-In the three linear models, OTUs that drove the detection of SRNs belong to family *Lachnospiraceae* and *Ruminococcaceae* (OTUs x, y, z).
-To explain the importance of features in non-linear models, we used permutation importance on the held-out test data [Figure 3b].
-Perfectly correlated OTUs were grouped and permuted together to get a better representation of their effect in the prediction performance of the model.
-Non-correlated OTUs were permuted individually.
+The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost increase in complexity as they decrase in interpretibility, respectively.
+We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models.
+We calculated the mean weights and coefficients of all the features over the 100 data-splits.
+In the three linear models, OTUs that had the largest mean weights and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629) and genera *Gamella*  (OTU00426) [Figure 3].
+We explained the feature importances in non-linear models using permutation importance on the held-out test data.
+For the tree-based models, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most.
+Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
+The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].
+Other OTUs had none to minimal effect on the predictive performance.
 
@@
 
-## Conclusions
+## Discussion
 
@@
 \newpage
-
+\includegraphics{Figure_1}
 **Figure 1.
-Generalization and classification performance of modeling methods ** AUC values of all cross validation and testing performances.
+Machine learning pipeline showing predictive model training and evaluation flowchart.
+**  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data.
+The model was evaluated on a held-out set of data (not used in selecting the model).
+Abbreviations: AUROC, area under the receiver operating characteristic curve
+\newpage
+\includegraphics{Figure_2.png}
+**Figure 2.
+Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.
+**   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
+Discriminative performance of random forest model was higher than other ML models.
 The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
@@
 Outliers are data points that are not within 3/2 times the interquartile ranges.
-
+Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
+\newpage
+\includegraphics{Figure_3.png}
+**Figure 3.
+Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights.
+The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits.
+Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel.
+Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
+\newpage
+\includegraphics{Figure_4.png}
+**Figure 4.
+Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set.
+For all the tree-based models, a **Peptostreptococcus** species (OTU00367) had the largest impact on predictive performance of the model.
+Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+\newpage
+\includegraphics{Figure_5.png}
+**Figure 5.
+Computational efficiency of seven ML models.** The walltimes for training and testing of each data-split showed the differences in computational efficieny of the seven models.
+The median walltime in hours was the highest for XGBoost and shortest for L2 logistic regression.
+The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
+The whiskers show the farthest points that are not outliers.
+Outliers are data points that are not within 3/2 times the interquartile ranges.
+Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
 \newpage

### 2019-05-16 submission/manuscript.tex
prep data and figures in the submission folder

@@
 
-Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Patrick D.
-Schloss\textsuperscript{1\(\dagger\)}
+Begüm D. Topçuoğlu\({^1}\), Mack Ruffin\({^2}\), Jenna Wiens\({^3}\),
+Patrick D. Schloss\textsuperscript{1\(\dagger\)}
 
@@
 
-2. Department of Computer Science and Engineering, University or
+2. Department of Family Medicine and Community Medicine, Penn State
+Hershey Medical Center, Hershey, PA
+
+3. Department of Computer Science and Engineering, University or
 Michigan, Ann Arbor, MI 49109
@@
 
-\textbf{Model selection and construction}
+\textbf{Model selection and pipeline construction}
 
-The data has 6920 features (6920 OTUs) and a two-class label that
-defines the colonic health of the patients (SRN or normal). All the
+We used a cohort of 490 patients with 261 cases of SRN. For each patient
+fecal sample, we had 6920 features (6920 OTUs) and a two-class label
+that defines the colonic health of the patients (SRN or normal). All the
 cases were independently labeled through colonoscopies. We established
@@
 interpretibility and computational efficiency due to model selection. We
-randomly split the data into training/validation and test sets 100 times
-with a 80/20 proportion {[}Figure 1{]}. Since the colonic health is not
-uniformly represented in the data, data-splits are stratified to
-maintain the overall label distribution on the training set. For each
-data-split, training/validation set consisted of 393 patients (209 SRN),
-while the test set was composed of 97 patients (52 SRN). The
+randomly split the data into training/validation and test sets so that
+the training/validation set consisted of 80\% of the full dataset while
+the test set was composed of the remaining data {[}Figure 1{]}. Since
+the cases are not uniformly represented in the data, the initial
+data-split was stratified to maintain the overall label distribution on
+the training set. Training/validation set consisted of 393 patients (209
+SRN), while the test set was composed of 97 patients (52 SRN). The
 training/validation data was used for training purposes and validation
@@
 S2{]}. We then used the held-out test set to evaluate the prediction
-performance of each ML model.
+performance of each ML model. The data-split, hyperparameter selection,
+training and testing steps were repeated 100 times to get a reliable and
+robust reading of model prediction performance {[}Figure 1{]}.
 
@@
 AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure 2{]}. L2 logistic
-regression and XGBoost had significantly lower AUROC values, 0.676 (std
-± 0.042) and 0.678 (std ± 0.04) {[}Figure 2{]}. Random forest also had
-significanly higher performances than L2 linear SVM, RBF SVM and
-decision tree which had mean AUROC values of 0.671 (std ± 0.045), 0.663
-(std ± 0.044), and 0.598 (std ± 0.044), respectively {[}Figure 2{]}. We
-also evaluated the generalization performance of each classifier by
-comparing their mean cross-validation AUROC and mean testing AUROC. We
-found that mean cross-validation and testing AUROC difference for each
-model was below 0.01.
+regression and XGBoost followed random forest however they had
+significantly lower AUROC values, 0.676 (std ± 0.042) and 0.678 (std ±
+0.04) {[}Figure 2{]}. Random forest also had significanly higher
+performances than L2 linear SVM, RBF SVM and decision tree which had
+mean AUROC values of 0.671 (std ± 0.045), 0.663 (std ± 0.044), and 0.598
+(std ± 0.044), respectively {[}Figure 2{]}. We also evaluated the
+generalization performance of each classifier by comparing their mean
+cross-validation AUROC and mean testing AUROC. We found that mean
+cross-validation and testing AUROC difference for each model was below
+0.01.
 
@@
 
-We interpreted the feature weights of L1 and L2 SVM with linear kernel
-and regression coefficients of L2 logistic regression using the training
-data. We calculated the mean weights of all the features fromt the 100
-randomizations of five-fold cross-validation {[}Figure 3a{]}. In the
-three linear models, OTUs that drove the detection of SRNs belong to
-family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (OTUs x, y, z).
-To explain the importance of features in non-linear models, we used
-permutation importance on the held-out test data {[}Figure 3b{]}.
-Perfectly correlated OTUs were grouped and permuted together to get a
-better representation of their effect in the prediction performance of
-the model. Non-correlated OTUs were permuted individually.
+The ML models we built using L2-regularized logistic regression, L1 and
+L2 support vector machines (SVM) with linear and radial basis function
+kernels, a decision tree, random forest and XGBoost increase in
+complexity as they decrase in interpretibility, respectively. We
+interpreted L1 and L2 SVM with linear kernel using the feature weights
+and L2 logistic regression using regression coefficients of the trained
+models. We calculated the mean weights and coefficients of all the
+features over the 100 data-splits. In the three linear models, OTUs that
+had the largest mean weights and drove the detection of SRNs belong to
+family \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
+OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822,
+OTU00609, OTU01212, OTU00629) and genera \emph{Gamella} (OTU00426)
+{[}Figure 3{]}. We explained the feature importances in non-linear
+models using permutation importance on the held-out test data. For the
+tree-based models, permuting \emph{Peptostreptococcus} (OTU00367)
+abundances randomly, dropped the predictive performances the most.
+Decision tree, random forest and XGBoost models' predictive performance
+dropped from 0.6 base testing AUROC median to 0.52, from 0.69 to 0.68
+and from 0.68 to 0.65, respectively {[}Figure 4{]}. The negative
+predictive impact of \emph{Peptostreptococcus} in the decision tree
+model was followed by a \emph{Lachnospiraceae} species (OTU00058) (0.6
+base testing AUROC median to 0.58) {[}Figure 4B{]}. Other OTUs had none
+to minimal effect on the predictive performance.
 
@@
 
-\subsection{Conclusions}\label{conclusions}
+\subsection{Discussion}\label{discussion}
 
@@
 
-\textbf{Figure 1. Generalization and classification performance of
-modeling methods } AUC values of all cross validation and testing
-performances. The boxplot shows quartiles at the box ends and the
+\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
+showing predictive model training and evaluation flowchart. } We split
+the data 80\%/20\% stratified to maintain the overall label
+distribution, performed five-fold cross-validation on the training data
+to select the best hyperparameter setting and then using these
+hyperparameters to train all of the training data. The model was
+evaluated on a held-out set of data (not used in selecting the model).
+Abbreviations: AUROC, area under the receiver operating characteristic
+curve \newpage
+\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
+classification performance of ML models using AUROC values of all cross
+validation and testing performances. } The median AUROC for diagnosing
+individuals with SRN using bacterial abundances was higher than chance
+(depicted by horizontal line at 0.50) for all the ML models.
+Discriminative performance of random forest model was higher than other
+ML models. The boxplot shows quartiles at the box ends and the
 statistical median as the horizontal line in the box. The whiskers show
 the farthest points that are not outliers. Outliers are data points that
-are not within 3/2 times the interquartile ranges.
-
+are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
+screen-relevant neoplasias; AUROC, area under the receiver operating
+characteristic curve; SVM, support vector machine; XGBoost, extreme
+gradient boosting \newpage
+\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
+linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
+with linear kernel feature weights (C) L2 SVM with linear kernel feature
+weights. The means weights and coefficients of the most important five
+OTUs for each model are shown here with the standard deviation over 100
+data-splits. Similar OTUs had the largest impact on the predictive
+performance of L2 logistic regression and L2 SVM with linear kernel.
+Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
+Unit. \newpage
+\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
+non-linear ML models.} (A) SVM with radial basis kernel (B) decision
+tree (C) random forest (D) XGboost feaure importances were explanied
+using permutation importance using held-out test set. For all the
+tree-based models, a \textbf{Peptostreptococcus} species (OTU00367) had
+the largest impact on predictive performance of the model.
+Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
+Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+\newpage
+\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
+efficiency of seven ML models.} The walltimes for training and testing
+of each data-split showed the differences in computational efficieny of
+the seven models. The median walltime in hours was the highest for
+XGBoost and shortest for L2 logistic regression. The boxplot shows
+quartiles at the box ends and the statistical median as the horizontal
+line in the box. The whiskers show the farthest points that are not
+outliers. Outliers are data points that are not within 3/2 times the
+interquartile ranges. Abbreviations: AUROC, area under the receiver
+operating characteristic curve; SVM, support vector machine; XGBoost,
+extreme gradient boosting.\\
 \newpage

### 2019-05-16 submission/manuscript.Rmd
update'

@@
 We used a cohort of 490 patients with 261 cases of SRN.
-For each patient fecal sample, we had `r paste(ncol(data)-1)` features (`r paste(ncol(data)-1)` OTUs) and a two-class label that defines the colonic health of the patients (SRN or normal).
+For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colonic health (SRN or normal).
 All the cases were independently labeled through colonoscopies.
@@
 We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
-Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution on the training set.
+Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution on both the training/validation and test sets.
 Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
@@
 
-__The prediction and generalization performance of classifiers during cross-validation and when applied to the held-out test data.__
+__The prediction and generalization performance of seven ML models during cross-validation and when applied to the held-out test data.__
 
 We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric.
-Random Forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`) [Figure 2].
-L2 logistic regression and XGBoost followed random forest however they had significantly lower AUROC values, `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) [Figure 2].
-Random forest also had significanly higher performances than L2 linear SVM, RBF SVM and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
-We also evaluated the generalization performance of each classifier by comparing their mean cross-validation AUROC and mean testing AUROC.
+Random forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`), followed by L2 logistic regression and XGBoost with mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) respectively [Figure 2].
+Random forest had significanly higher predictive performances than L2 SVM with linear and radial basis function kernels and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
+We also evaluated the generalization performance of each ML model by comparing their mean cross-validation AUROC and mean testing AUROC.
 We found that mean cross-validation and testing AUROC difference for each model was below 0.01.
 
-__The complexity and interpretibility of each classifier.__
+__The complexity and interpretibility of each ML model.__
 
-The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost increase in complexity as they decrase in interpretibility, respectively.
+The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost increase in complexity as they decrase in interpretibility.
 We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models.
@@
 
-__The computational efficiency of each classifier.__
+__The computational efficiency of each ML model.__
 
@@
 Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
-Patients' colonic disease status was defined by colonoscopy with adequate preparation and tissue histopathology of all resected lesions.
+Patients' colonic health was labeled by colonoscopy with adequate preparation and tissue histopathology of all resected lesions.
 Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma.
@@
 Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set.
-For all the tree-based models, a **Peptostreptococcus** species (OTU00367) had the largest impact on predictive performance of the model.
+For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model.
 Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.

### 2019-05-16 submission/manuscript.tex
update'

@@
 
-We used a cohort of 490 patients with 261 cases of SRN. For each patient
-fecal sample, we had 6920 features (6920 OTUs) and a two-class label
-that defines the colonic health of the patients (SRN or normal). All the
-cases were independently labeled through colonoscopies. We established
-modeling pipelines for a binary prediction task with L2-regularized
-logistic regression, L1 and L2 support vector machines (SVM) with linear
-and radial basis function kernels, a decision tree, random forest and
-XGBoost to emphasize the differences in model accuracy, complexity,
-interpretibility and computational efficiency due to model selection. We
-randomly split the data into training/validation and test sets so that
-the training/validation set consisted of 80\% of the full dataset while
-the test set was composed of the remaining data {[}Figure 1{]}. Since
-the cases are not uniformly represented in the data, the initial
-data-split was stratified to maintain the overall label distribution on
-the training set. Training/validation set consisted of 393 patients (209
-SRN), while the test set was composed of 97 patients (52 SRN). The
-training/validation data was used for training purposes and validation
-of hyperparameter selection, and the test set was used for evaluation
-purposes. Validation of hyperparameter selection was performed using 100
-randomizations of five-fold cross-validation on the training/validation
-set {[}Figure 1{]}. Similar to the initial data-split, five-fold
-cross-validation was also stratified to maintain the overall label
-distribution on the training and validation sets. We validated the
-cross-validation performances of each hyperparameter setting over the
-100 randomizations and selected the best performing hyper-parameter
-setting to train the full training/validation dataset {[}Figures S1 and
-S2{]}. We then used the held-out test set to evaluate the prediction
-performance of each ML model. The data-split, hyperparameter selection,
-training and testing steps were repeated 100 times to get a reliable and
-robust reading of model prediction performance {[}Figure 1{]}.
+We used a cohort of 490 patients with 261 cases of SRN. For each
+patient, we had 6920 features (fecal bacterial abundances) and a
+two-class label that defines their colonic health (SRN or normal). All
+the cases were independently labeled through colonoscopies. We
+established modeling pipelines for a binary prediction task with
+L2-regularized logistic regression, L1 and L2 support vector machines
+(SVM) with linear and radial basis function kernels, a decision tree,
+random forest and XGBoost to emphasize the differences in model
+accuracy, complexity, interpretibility and computational efficiency due
+to model selection. We randomly split the data into training/validation
+and test sets so that the training/validation set consisted of 80\% of
+the full dataset while the test set was composed of the remaining data
+{[}Figure 1{]}. Since the cases are not uniformly represented in the
+data, the initial data-split was stratified to maintain the overall
+label distribution on both the training/validation and test sets.
+Training/validation set consisted of 393 patients (209 SRN), while the
+test set was composed of 97 patients (52 SRN). The training/validation
+data was used for training purposes and validation of hyperparameter
+selection, and the test set was used for evaluation purposes. Validation
+of hyperparameter selection was performed using 100 randomizations of
+five-fold cross-validation on the training/validation set {[}Figure
+1{]}. Similar to the initial data-split, five-fold cross-validation was
+also stratified to maintain the overall label distribution on the
+training and validation sets. We validated the cross-validation
+performances of each hyperparameter setting over the 100 randomizations
+and selected the best performing hyper-parameter setting to train the
+full training/validation dataset {[}Figures S1 and S2{]}. We then used
+the held-out test set to evaluate the prediction performance of each ML
+model. The data-split, hyperparameter selection, training and testing
+steps were repeated 100 times to get a reliable and robust reading of
+model prediction performance {[}Figure 1{]}.
 
-\textbf{The prediction and generalization performance of classifiers
+\textbf{The prediction and generalization performance of seven ML models
 during cross-validation and when applied to the held-out test data.}
@@
 area under the receiver operating characteristic curve (AUROC) as the
-discriminative performance metric. Random Forest had the highest mean
-AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure 2{]}. L2 logistic
-regression and XGBoost followed random forest however they had
-significantly lower AUROC values, 0.676 (std ± 0.042) and 0.678 (std ±
-0.04) {[}Figure 2{]}. Random forest also had significanly higher
-performances than L2 linear SVM, RBF SVM and decision tree which had
-mean AUROC values of 0.671 (std ± 0.045), 0.663 (std ± 0.044), and 0.598
-(std ± 0.044), respectively {[}Figure 2{]}. We also evaluated the
-generalization performance of each classifier by comparing their mean
-cross-validation AUROC and mean testing AUROC. We found that mean
-cross-validation and testing AUROC difference for each model was below
-0.01.
+discriminative performance metric. Random forest had the highest mean
+AUROC for detecting SRN, 0.697 (std ± 0.037) followed by L2 logistic
+regression and XGBoost with mean AUROC values of 0.676 (std ± 0.042) and
+0.678 (std ± 0.04) respectively {[}Figure 2{]}. Random forest had
+significanly higher predictive performances than those as well as L2
+linear SVM, RBF SVM and decision tree which had mean AUROC values of
+0.671 (std ± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044),
+respectively {[}Figure 2{]}. We also evaluated the generalization
+performance of each ML model by comparing their mean cross-validation
+AUROC and mean testing AUROC. We found that mean cross-validation and
+testing AUROC difference for each model was below 0.01.
 
-\textbf{The complexity and interpretibility of each classifier.}
+\textbf{The complexity and interpretibility of each ML model.}
 
@@
 
-\textbf{The computational efficiency of each classifier.}
+\textbf{The computational efficiency of each ML model.}
 
@@
 using permutation importance using held-out test set. For all the
-tree-based models, a \textbf{Peptostreptococcus} species (OTU00367) had
+tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
 the largest impact on predictive performance of the model.

### 2019-05-16 submission/manuscript.Rmd
update

@@
 
-Begüm D. Topçuoğlu${^1}$, Mack Ruffin${^2}$, Jenna Wiens${^3}$, Patrick D. Schloss^1$\dagger$^
+Begüm D. Topçuoğlu${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^
 
@@
 
-2\. Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
+2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109
 
-3\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109
+3\. Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
 
@@
 
-To showcase reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline.
+To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline.
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients.
@@
 Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model.
-The mean test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))` (std ± `r paste(round(test_all[performance_index[1],4], 3))`) to `r paste(round(test_all[performance_index[7],3], 3))` (std ± `r paste(round(test_all[performance_index[7],4], 3))`).
-Random forest had the highest mean AUROC for detecting SRN.
+The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`.
+Random forest had the highest median AUROC for detecting SRN.
 Despite the simplicity, the L2-regularized logistic regression followed random forest in performance.
 In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`).
-We found that mean cross-validation and testing AUROC varied only `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
+We found that median cross-validation and testing AUROC varied only `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
 Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
@@
 
-__The prediction and generalization performance of seven ML models during cross-validation and when applied to the held-out test data.__
+__Random forest has the highest prediction performance when applied to the held-out test data compared the other six ML models .__
 
-We evaluated the prediction performance of seven binary classification models when applied to held-out test data over 100 data-splits using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric.
-Random forest had the highest mean AUROC for detecting SRN, `r paste(round(mean((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`), followed by L2 logistic regression and XGBoost with mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`) and `r paste(round(mean((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="XGBoost"))$test_aucs), 3))`) respectively [Figure 2].
-Random forest had significanly higher predictive performances than L2 SVM with linear and radial basis function kernels and decision tree which had mean AUROC values of `r paste(round(mean((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`),  `r paste(round(mean((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), and `r paste(round(mean((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (std ± `r paste(round(sd((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
-We also evaluated the generalization performance of each ML model by comparing their mean cross-validation AUROC and mean testing AUROC.
-We found that mean cross-validation and testing AUROC difference for each model was below 0.01.
+We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric.
+Random forest had significantly higher test AUROC values than the other models for detecting SRNs.
+The median AUROC of random forest was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
+L2 logistic regression, XGBoost, L2 SVM with linear and radial basis function kernel AUROC values were not significantly different from one another.
+They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively.
+L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
+
+__Predictive performances of the seven ML models during cross-validation and when applied to the held-out test data do not show large differences.__
+
+We evaluated the generalization performance of each ML model by comparing their median cross-validation AUROC and median testing AUROC.
+The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively.
 
@@
 Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set.
+The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
 For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model.

### 2019-05-16 submission/manuscript.tex
update

@@
 
-Begüm D. Topçuoğlu\({^1}\), Mack Ruffin\({^2}\), Jenna Wiens\({^3}\),
+Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Mack Ruffin\({^3}\),
 Patrick D. Schloss\textsuperscript{1\(\dagger\)}
@@
 
-2. Department of Family Medicine and Community Medicine, Penn State
+2. Department of Computer Science and Engineering, University or
+Michigan, Ann Arbor, MI 49109
+
+3. Department of Family Medicine and Community Medicine, Penn State
 Hershey Medical Center, Hershey, PA
-
-3. Department of Computer Science and Engineering, University or
-Michigan, Ann Arbor, MI 49109
 
@@
 to evaluate generalization and prediction performance of each ML model.
-The mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ±
-0.037). Random forest had the highest mean AUROC for detecting SRN.
-Despite the simplicity, the L2-regularized logistic regression followed
-random forest in performance. In terms of computational efficiency, L2
-logistic regression trained the fastest (0.202 hours, std ± 0.019),
-while XGBoost took the longest (162.843 hours, std ± 3.986). We found
-that mean cross-validation and testing AUROC varied only 0.01, which
-highlights the importance of a seperate held-out test set and consistent
+The median test AUROC varied from 0.601 to 0.695. Random forest had the
+highest median AUROC for detecting SRN. Despite the simplicity, the
+L2-regularized logistic regression followed random forest in
+performance. In terms of computational efficiency, L2 logistic
+regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
+took the longest (162.843 hours, std ± 3.986). We found that median
+cross-validation and testing AUROC varied only 0.021, which highlights
+the importance of a seperate held-out test set and consistent
 preprocessing of the data prior to evaluation. Aside from evaluating
@@
 
-\textbf{The prediction and generalization performance of seven ML models
-during cross-validation and when applied to the held-out test data.}
+\textbf{Random forest has the highest prediction performance when
+applied to the held-out test data compared the other six ML models .}
 
-We evaluated the prediction performance of seven binary classification
-models when applied to held-out test data over 100 data-splits using the
-area under the receiver operating characteristic curve (AUROC) as the
-discriminative performance metric. Random forest had the highest mean
-AUROC for detecting SRN, 0.697 (std ± 0.037) followed by L2 logistic
-regression and XGBoost with mean AUROC values of 0.676 (std ± 0.042) and
-0.678 (std ± 0.04) respectively {[}Figure 2{]}. Random forest had
-significanly higher predictive performances than those as well as L2
-linear SVM, RBF SVM and decision tree which had mean AUROC values of
-0.671 (std ± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044),
-respectively {[}Figure 2{]}. We also evaluated the generalization
-performance of each ML model by comparing their mean cross-validation
-AUROC and mean testing AUROC. We found that mean cross-validation and
-testing AUROC difference for each model was below 0.01.
+We evaluated the prediction performances of seven binary classification
+models when applied to held-out test data using the area under the
+receiver operating characteristic curve (AUROC) as the discriminative
+performance metric. Random forest had significantly higher test AUROC
+values than the other models for detecting SRNs. The median AUROC of
+random forest was 0.695 (IQR 0.044). L2 logistic regression, XGBoost, L2
+SVM with linear and radial basis function kernel AUROC values were not
+significantly different from one another. They had median AUROC values
+of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
+0.056) respectively. L1 SVM with linear kernel and decision tree had
+significantly lower AUROC values than the other ML models with median
+AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
+2{]}.
+
+\textbf{Predictive performances of the seven ML models during
+cross-validation and when applied to the held-out test data do not show
+large differences.}
+
+We evaluated the generalization performance of each ML model by
+comparing their median cross-validation AUROC and median testing AUROC.
+The largest difference between the two was 0.021 in L1 SVM with linear
+kernel followed by SVM with radial basis function kernel and decision
+tree with a difference of 0.007 and 0.006, respectively.
 
@@
 kernels, a decision tree, random forest and XGBoost increase in
-complexity as they decrase in interpretibility, respectively. We
-interpreted L1 and L2 SVM with linear kernel using the feature weights
-and L2 logistic regression using regression coefficients of the trained
-models. We calculated the mean weights and coefficients of all the
-features over the 100 data-splits. In the three linear models, OTUs that
-had the largest mean weights and drove the detection of SRNs belong to
-family \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
-OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822,
-OTU00609, OTU01212, OTU00629) and genera \emph{Gamella} (OTU00426)
-{[}Figure 3{]}. We explained the feature importances in non-linear
-models using permutation importance on the held-out test data. For the
-tree-based models, permuting \emph{Peptostreptococcus} (OTU00367)
-abundances randomly, dropped the predictive performances the most.
-Decision tree, random forest and XGBoost models' predictive performance
-dropped from 0.6 base testing AUROC median to 0.52, from 0.69 to 0.68
-and from 0.68 to 0.65, respectively {[}Figure 4{]}. The negative
-predictive impact of \emph{Peptostreptococcus} in the decision tree
-model was followed by a \emph{Lachnospiraceae} species (OTU00058) (0.6
-base testing AUROC median to 0.58) {[}Figure 4B{]}. Other OTUs had none
-to minimal effect on the predictive performance.
+complexity as they decrase in interpretibility. We interpreted L1 and L2
+SVM with linear kernel using the feature weights and L2 logistic
+regression using regression coefficients of the trained models. We
+calculated the mean weights and coefficients of all the features over
+the 100 data-splits. In the three linear models, OTUs that had the
+largest mean weights and drove the detection of SRNs belong to family
+\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239, OTU00659,
+OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609,
+OTU01212, OTU00629) and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}.
+We explained the feature importances in non-linear models using
+permutation importance on the held-out test data. For the tree-based
+models, permuting \emph{Peptostreptococcus} (OTU00367) abundances
+randomly, dropped the predictive performances the most. Decision tree,
+random forest and XGBoost models' predictive performance dropped from
+0.6 base testing AUROC median to 0.52, from 0.69 to 0.68 and from 0.68
+to 0.65, respectively {[}Figure 4{]}. The negative predictive impact of
+\emph{Peptostreptococcus} in the decision tree model was followed by a
+\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
+to 0.58) {[}Figure 4B{]}. Other OTUs had none to minimal effect on the
+predictive performance.
 
@@
 Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
-colonic disease status was defined by colonoscopy with adequate
-preparation and tissue histopathology of all resected lesions. Patients
-with an adenoma greater than 1 cm, more than three adenomas of any size,
-or an adenoma with villous histology were classified as advanced
-adenoma. Study had 172 patients with normal colonoscopies, 198 with
-adenomas and 120 with carcinomas. Of the 198 adenomas, 109 were
-identified as advanced adenomas. Stool provided by the patients was used
-for 16S rRNA gene sequencing to measure bacterial population abundances.
-The bacterial abundance data was generated by Sze et al, by processing
-16S rRNA sequences in Mothur (v1.39.3) using the default quality
-filtering methods, identifying and removing chimeric sequences using
-VSEARCH and assigning to OTUs at 97\% similarity using the OptiClust
-algorithm (16--18).
+colonic health was labeled by colonoscopy with adequate preparation and
+tissue histopathology of all resected lesions. Patients with an adenoma
+greater than 1 cm, more than three adenomas of any size, or an adenoma
+with villous histology were classified as advanced adenoma. Study had
+172 patients with normal colonoscopies, 198 with adenomas and 120 with
+carcinomas. Of the 198 adenomas, 109 were identified as advanced
+adenomas. Stool provided by the patients was used for 16S rRNA gene
+sequencing to measure bacterial population abundances. The bacterial
+abundance data was generated by Sze et al, by processing 16S rRNA
+sequences in Mothur (v1.39.3) using the default quality filtering
+methods, identifying and removing chimeric sequences using VSEARCH and
+assigning to OTUs at 97\% similarity using the OptiClust algorithm
+(16--18).
 
@@
 tree (C) random forest (D) XGboost feaure importances were explanied
-using permutation importance using held-out test set. For all the
-tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
-the largest impact on predictive performance of the model.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
-Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
-\newpage
+using permutation importance using held-out test set. The gray rectangle
+and the dashed line show the IQR range and median of the base testing
+AUROC without any permutation performed. For all the tree-based models,
+a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
+predictive performance of the model. Abbreviations: SVM, support vector
+machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
+Operational Taxonomic Unit. \newpage
 \includegraphics{Figure_5.png} \textbf{Figure 5. Computational

### 2019-05-16 code/learning/Figure2.R

@@ -41,7 +41,7 @@ best_performance <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
 
 
 performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC, fill = Performance)) +
-  geom_boxplot(alpha=0.7) +
+  geom_boxplot() +
   geom_hline(yintercept = 0.5, linetype="dashed") +
   scale_fill_manual(values=c("#999999", "#E69F00")) +
   scale_y_continuous(name = "AUROC",

### 2019-05-16 code/learning/load_AUROC_data.R

@@ -1,3 +1,4 @@
+
 ######################################################################
 #--------- Load .tsv data to get mean test AUC for 7 models--------#
 ######################################################################
@@ -6,35 +7,39 @@
 best_files <- list.files(path= '../data/process', pattern='combined_best.*', full.names = TRUE)
 all <- map_df(best_files, read_files) 
 
-# Get the The unpaired two-samples Wilcoxon test to see if models differ from one another signigicantly
+# Get the The unpaired two-samples Wilcoxon test to see if test_aucs of models differ from one another signigicantly
 
 rf_xgboost <- wilcoxon_test(all, "Random_Forest", "XGBoost")
 rf_logit <- wilcoxon_test(all, "Random_Forest", "L2_Logistic_Regression")
 rf_L2svm <- wilcoxon_test(all, "Random_Forest", "L2_Linear_SVM")
+rf_rbf <- wilcoxon_test(all, "Random_Forest", "RBF_SVM")
 xgboost_L2svm <- wilcoxon_test(all, "XGBoost", "L2_Linear_SVM")
 logit_L2svm <- wilcoxon_test(all, "L2_Logistic_Regression", "L2_Linear_SVM")
+rbf_L2svm <- wilcoxon_test(all, "RBF_SVM", "L2_Linear_SVM")
+l1svm_L2svm <- wilcoxon_test(all, "L1_Linear_SVM", "L2_Linear_SVM")
+dt_L2svm <- wilcoxon_test(all, "Decision_Tree", "L2_Linear_SVM")
 
 # Bind them together and summarise mean testAUC by model
 test_all <- all %>%
   melt_data() %>% 
   select(-variable) %>% 
   group_by(model, Performance) %>% 
-  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
+  summarise(median_AUC = median(AUC), iqr_AUC = IQR(AUC)) %>% 
   filter(Performance=="testing") 
 
-# Get the order index from small to large meanAUC
-performance_index <- order(test_all$mean_AUC)
+# Get the order index from small to large medianAUC
+performance_index <- order(test_all$median_AUC)
 
-# Bind them together and summarise mean cvAUC by model
+# Bind them together and summarise median cvAUC by model
 cv_all <- all %>%
   melt_data() %>% 
   select(-variable) %>% 
   group_by(model, Performance) %>% 
-  summarise(mean_AUC = mean(AUC), sd_AUC = sd(AUC)) %>% 
[…]

### 2019-05-17 submission/manuscript.Rmd
update bibs

@@
 
-Begüm D. Topçuoğlu${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^
+Begüm D. Topçuoğlu${^1}$, Nick Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^
 
@@
 Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
-Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented.
-Most notably, there are misleading practices of using "leaky" ML pipelines where models are tested on training data or reporting the best outcome of different iterations of cross-validation.
+Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016;  .
+More notably, there are flawed ML practices like using ML pipelines where there is no seperate held-out test dataset to evaulate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
 Moreover, there is a lack of discussion on why a particular ML model is utilized.
@@
 ## Discussion
+
+#### Interpretation of results
+ Our results show that 
+ Suprisingly, l2 logistic..
+ computational efficiency and interp of logistic vs random forest
+ decision tree and l1 why so low?
+
+#### Consideration of possible weaknesses
+ What happens with imbalanced data
+ What happens with smaller datasets
+ Does the code work on any microbiome data?
+  Why did we use only one dataset? Becuase we are comparing several methods on 1 dataset.
+  What will the future data look like?
+
+#### Relationship of results to previous literature and broader implications of having answered research question
+
+#### Prospects for future progress
+  Subsample dataset and refit see how method behaves for 7 methods. (500->250->50)
 

### 2019-05-17 submission/manuscript.tex
update bibs

@@
 
-Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Mack Ruffin\({^3}\),
-Patrick D. Schloss\textsuperscript{1\(\dagger\)}
+Begüm D. Topçuoğlu\({^1}\), Nick Lesniak\({^1}\), Jenna Wiens\({^2}\),
+Mack Ruffin\({^3}\), Patrick D. Schloss\textsuperscript{1\(\dagger\)}
 
@@
 ecology of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
-However, currently the field's use of ML lacks clarity and consistency
-on which methods are used and how these methods are implemented. Most
-notably, there are misleading practices of using ``leaky'' ML pipelines
-where models are tested on training data or reporting the best outcome
-of different iterations of cross-validation. Moreover, there is a lack
+inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
+(1--11). However, currently the field's use of ML lacks clarity and
+consistency on which methods are used and how these methods are
+implemented {[}(12); . Most notably, there are misleading practices of
+using ``leaky'' ML pipelines where models are tested on training data,
+reporting few or only the best outcomes of different randomizations of
+cross-validation and showing a disregard for large differences between
+cross-validation and testing performances as well as large confidence
+intervals of testing performances (4, 13--18). Moreover, there is a lack
 of discussion on why a particular ML model is utilized. Recently, there
@@
 extreme gradient boosting and neural networks without a discussion on if
-and how much model interpretibility is necessary for the study (11--14).
-The lack of transparency on modeling methodology and model selection
-negatively impact model reproducibility and reliability. We need to
-strive toward better machine learning practices by (1) implementing
-consistent and reliable machine learning pipelines; (2) selecting ML
-models that reflect the goal of the study as it will inform our
-expectations of model accuracy, complexity, interpretibility and
+and how much model interpretibility is necessary for the study (11,
+19--21). The lack of transparency on modeling methodology and model
+selection negatively impact model reproducibility and reliability. We
+need to strive toward better machine learning practices by (1)
+implementing consistent and reliable machine learning pipelines; (2)
+selecting ML models that reflect the goal of the study as it will inform
+our expectations of model accuracy, complexity, interpretibility and
 computational efficiency.
 
-To showcase reliable ML pipeline and to shed light on how much ML model
-selection can affect modeling results, we performed an empirical
+To showcase a reliable ML pipeline and to shed light on how much ML
+model selection can affect modeling results, we performed an empirical
 analysis comparing several different ML models using the same dataset
@@
 
+\paragraph{Interpretation of results}\label{interpretation-of-results}
+
+Our results show that Suprisingly, l2 logistic.. computational
+efficiency and interp of logistic vs random forest decision tree and l1
+why so low?
+
+\paragraph{Consideration of possible
+weaknesses}\label{consideration-of-possible-weaknesses}
+
+What happens with imbalanced data What happens with smaller datasets
+Does the code work on any microbiome data? Why did we use only one
+dataset? Becuase we are comparing several methods on 1 dataset. What
+will the future data look like?
+
+\paragraph{Relationship of results to previous literature and broader
+implications of having answered research
+question}\label{relationship-of-results-to-previous-literature-and-broader-implications-of-having-answered-research-question}
+
+\paragraph{Prospects for future
+progress}\label{prospects-for-future-progress}
+
+Subsample dataset and refit see how method behaves for 7 methods.
+(500-\textgreater{}250-\textgreater{}50)
+
 \subsection{Materials and Methods}\label{materials-and-methods}
@@
 England Early Detection Research Network study. These data were obtained
-from Sze et al (15). The stool samples were provided by recruited adult
+from Sze et al (22). The stool samples were provided by recruited adult
 participants who were undergoing scheduled screening or surveillance
@@
 assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(16--18).
+(23--25).
 
@@
 
+\hypertarget{ref-thaiss_persistent_2016}{}
+12. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
+\textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
+\textbf{Dohnalová L}, \textbf{Braverman S}, \textbf{Rozin S},
+\textbf{Malitsky S}, \textbf{Dori-Bachash M}, \textbf{Kuperman Y},
+\textbf{Biton I}, \textbf{Gertler A}, \textbf{Harmelin A},
+\textbf{Shapiro H}, \textbf{Halpern Z}, \textbf{Aharoni A},
+\textbf{Segal E}, \textbf{Elinav E}. 2016. Persistent microbiome
+alterations modulate the rate of post-dieting weight regain. Nature
+\textbf{540}:544--551.
+doi:\href{https://doi.org/10.1038/nature20796}{10.1038/nature20796}.
+
+\hypertarget{ref-flemer_oral_2018}{}
+13. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
+\textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
+E}, \textbf{O`Riordain M}, \textbf{Shanahan F}, \textbf{O`Toole PW}.
+2018. The oral microbiota in colorectal cancer is distinctive and
+predictive. Gut \textbf{67}:1454--1463.
+doi:\href{https://doi.org/10.1136/gutjnl-2017-314814}{10.1136/gutjnl-2017-314814}.
+
+\hypertarget{ref-dai_multi-cohort_2018}{}
+14. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
+WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
+\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
+J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
+identified altered bacteria across populations and universal bacterial
+markers. Microbiome \textbf{6}:70.
+doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.
+
+\hypertarget{ref-montassier_pretreatment_2016}{}
+15. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
+\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
+P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
+2016. Pretreatment gut microbiome predicts chemotherapy-related
+bloodstream infection. Genome Medicine \textbf{8}:49.
+doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.
+
+\hypertarget{ref-papa_non-invasive_2012}{}
+16. \textbf{Papa E}, \textbf{Docktor M}, \textbf{Smillie C},
+\textbf{Weber S}, \textbf{Preheim SP}, \textbf{Gevers D},
+\textbf{Giannoukos G}, \textbf{Ciulla D}, \textbf{Tabbaa D},
+\textbf{Ingram J}, \textbf{Schauer DB}, \textbf{Ward DV},
+\textbf{Korzenik JR}, \textbf{Xavier RJ}, \textbf{Bousvaros A},
+\textbf{Alm EJ}. 2012. Non-invasive mapping of the gastrointestinal
+microbiota identifies children with inflammatory bowel disease. PLOS ONE
+\textbf{7}:e39242.
+doi:\href{https://doi.org/10.1371/journal.pone.0039242}{10.1371/journal.pone.0039242}.
+
+\hypertarget{ref-mossotto_classification_2017}{}
+17. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
+\textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
+Classification of paediatric inflammatory bowel disease using machine
+learning. Scientific Reports \textbf{7}.
+doi:\href{https://doi.org/10.1038/s41598-017-02606-2}{10.1038/s41598-017-02606-2}.
+
+\hypertarget{ref-ai_systematic_2017}{}
+18. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
+\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
+supervised classifiers for fecal microbiota-based prediction of
+colorectal cancer. Oncotarget \textbf{8}:9546--9556.
+doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.
+
 \hypertarget{ref-galkin_human_2018}{}
-12. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+19. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
 \textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
@@
 \hypertarget{ref-reiman_using_2017}{}
-13. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+20. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
 convolutional neural networks to explore the microbiome, pp. 4269--4272.
@@
 \hypertarget{ref-fioravanti_phylogenetic_2017}{}
-14. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+21. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
 \textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
@@
 \hypertarget{ref-sze_leveraging_2018}{}
-15. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+22. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-16. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+23. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-17. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+24. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-18. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+25. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open

### 2019-05-22 submission/manuscript.Rmd
update to figures and paper and abstract:

@@
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016;  .
-More notably, there are flawed ML practices like using ML pipelines where there is no seperate held-out test dataset to evaulate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+More notably, flawed ML practices are prevalent such as using ML pipelines where there is no seperate held-out test dataset to evaulate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
 Moreover, there is a lack of discussion on why a particular ML model is utilized.
@@
 All the cases were independently labeled through colonoscopies.
-We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
+We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost) to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
+
+For regularized logistic regression and SVM with linear kernel we used L2 regularization to keep all potentially important features.
+For comparison, we also trained an L1 regularized SVM model with linear kernel.
+L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero), removing features that could be correlated with other important features, making it more difficult include the relationships between the microbes and possibly dropping features that are important in detecting SRNs.
+Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.
+
+We established a robust and reliable ML pipeline where we train and validate each of the seven models [Figure 1].
 We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
@@
 
-__Random forest has the highest prediction performance when applied to the held-out test data compared the other six ML models .__
+__Model performance and generalizability.__
 
-We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric.
+We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric [Figure 1].
 Random forest had significantly higher test AUROC values than the other models for detecting SRNs.
@@
 
-__Predictive performances of the seven ML models during cross-validation and when applied to the held-out test data do not show large differences.__
+We evaluated the generalization performance of each ML model by comparing their median cross-validation AUROC and median testing AUROC.
+This difference should be low to suggest the model is not overfitting despite the large number of features.
+The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
 
-We evaluated the generalization performance of each ML model by comparing their median cross-validation AUROC and median testing AUROC.
-The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively.
-
-__The complexity and interpretibility of each ML model.__
+__Interpretation of each ML model.__
 
@@
 We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models.
-We calculated the mean weights and coefficients of all the features over the 100 data-splits.
+We examined the 5 predictive (highest weight) and protective (lowest weight) features identified by the model.
+We calculated the mean weights and coefficients of these features over the 100 data-splits.
 In the three linear models, OTUs that had the largest mean weights and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629) and genera *Gamella*  (OTU00426) [Figure 3].
@@
 #### Interpretation of results
- Our results show that 
- Suprisingly, l2 logistic..
+ In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+We showed the importance of held-out test set and reporting the results of many iterations of cross-validation and testing.
+Each data-split has a weight that comes from random splitting.
+Seeing that testing and cross-validation median differences being low show that the models are generalizable and the pipeline works.
+Our results also show that we can choose different models for different reasons.
+Getting feature importances as weights and regression coefficients is the fastest and easiest interpratation.
+Similarly, linear models require less computational burden, fastest to train.
+THinking about the applicability, and actionability of a model in clinical setting, we might want faster results.
+
+ Suprisingly, l2 logistic regression does really well. 
  computational efficiency and interp of logistic vs random forest
@@
  Does the code work on any microbiome data?
-  Why did we use only one dataset? Becuase we are comparing several methods on 1 dataset.
+  Why did we use only one dataset? Why not validate on other data instead of held-out testing set? Becuase we are comparing several methods on 1 dataset.
   What will the future data look like?
@@
 \newpage
+\includegraphics{Figure_S1.png}
+**Figure S1. Hyperparameter setting performances for linear models.** 
+Abbreviations: 
+\newpage
+\includegraphics{Figure_S2.png}
+**Figure S2. Hyperparameter setting performances for non-linear models.** 
+Abbreviations: 
+\newpage
 

### 2019-05-22 submission/manuscript.tex
update to figures and paper and abstract:

@@
 consistency on which methods are used and how these methods are
-implemented {[}(12); . Most notably, there are misleading practices of
-using ``leaky'' ML pipelines where models are tested on training data,
-reporting few or only the best outcomes of different randomizations of
-cross-validation and showing a disregard for large differences between
-cross-validation and testing performances as well as large confidence
-intervals of testing performances (4, 13--18). Moreover, there is a lack
-of discussion on why a particular ML model is utilized. Recently, there
-is a trend towards using more complex ML models such as random forest,
-extreme gradient boosting and neural networks without a discussion on if
-and how much model interpretibility is necessary for the study (11,
-19--21). The lack of transparency on modeling methodology and model
-selection negatively impact model reproducibility and reliability. We
-need to strive toward better machine learning practices by (1)
-implementing consistent and reliable machine learning pipelines; (2)
-selecting ML models that reflect the goal of the study as it will inform
-our expectations of model accuracy, complexity, interpretibility and
-computational efficiency.
+implemented {[}(12); . More notably, flawed ML practices are prevalent
+such as using ML pipelines where there is no seperate held-out test
+dataset to evaulate model performance, reporting few or only the best
+outcomes of different randomizations of cross-validation and showing a
+disregard for large differences between cross-validation and testing
+performances as well as large confidence intervals of testing
+performances (4, 13--19). Moreover, there is a lack of discussion on why
+a particular ML model is utilized. Recently, there is a trend towards
+using more complex ML models such as random forest, extreme gradient
+boosting and neural networks without a discussion on if and how much
+model interpretibility is necessary for the study (11, 20--22). The lack
+of transparency on modeling methodology and model selection negatively
+impact model reproducibility and reliability. We need to strive toward
+better machine learning practices by (1) implementing consistent and
+reliable machine learning pipelines; (2) selecting ML models that
+reflect the goal of the study as it will inform our expectations of
+model accuracy, complexity, interpretibility and computational
+efficiency.
 
@@
 (SVM) with linear and radial basis function kernels, a decision tree,
-random forest and XGBoost to emphasize the differences in model
-accuracy, complexity, interpretibility and computational efficiency due
-to model selection. We randomly split the data into training/validation
-and test sets so that the training/validation set consisted of 80\% of
-the full dataset while the test set was composed of the remaining data
-{[}Figure 1{]}. Since the cases are not uniformly represented in the
-data, the initial data-split was stratified to maintain the overall
-label distribution on both the training/validation and test sets.
-Training/validation set consisted of 393 patients (209 SRN), while the
-test set was composed of 97 patients (52 SRN). The training/validation
-data was used for training purposes and validation of hyperparameter
-selection, and the test set was used for evaluation purposes. Validation
-of hyperparameter selection was performed using 100 randomizations of
-five-fold cross-validation on the training/validation set {[}Figure
-1{]}. Similar to the initial data-split, five-fold cross-validation was
-also stratified to maintain the overall label distribution on the
-training and validation sets. We validated the cross-validation
-performances of each hyperparameter setting over the 100 randomizations
-and selected the best performing hyper-parameter setting to train the
-full training/validation dataset {[}Figures S1 and S2{]}. We then used
-the held-out test set to evaluate the prediction performance of each ML
-model. The data-split, hyperparameter selection, training and testing
-steps were repeated 100 times to get a reliable and robust reading of
-model prediction performance {[}Figure 1{]}.
+random forest and extreme gradient boosted decision tree (XGBoost) to
+emphasize the differences in model accuracy, complexity,
+interpretibility and computational efficiency due to model selection.
 
-\textbf{Random forest has the highest prediction performance when
-applied to the held-out test data compared the other six ML models .}
+For regularized logistic regression and SVM with linear kernel we used
+L2 regularization to keep all potentially important features. For
+comparison, we also trained an L1 regularized SVM model with linear
+kernel. L1-regularization on microbiome data lead to a sparser solution
+(i.e., force many coefficients to zero), removing features that could be
+correlated with other important features, making it more difficult
+include the relationships between the microbes and possibly dropping
+features that are important in detecting SRNs. Finally, to explore the
+potential for non-linear relationships among features and the outcome of
+interest, we trained tree based models, decision tree, random forest and
+XGboost, as well as an SVM with non-linear kernel.
+
+We established a robust and reliable ML pipeline where we train and
+validate each of the seven models {[}Figure 1{]}. We randomly split the
+data into training/validation and test sets so that the
+training/validation set consisted of 80\% of the full dataset while the
+test set was composed of the remaining data {[}Figure 1{]}. Since the
+cases are not uniformly represented in the data, the initial data-split
+was stratified to maintain the overall label distribution on both the
+training/validation and test sets. Training/validation set consisted of
+393 patients (209 SRN), while the test set was composed of 97 patients
+(52 SRN). The training/validation data was used for training purposes
+and validation of hyperparameter selection, and the test set was used
+for evaluation purposes. Validation of hyperparameter selection was
+performed using 100 randomizations of five-fold cross-validation on the
+training/validation set {[}Figure 1{]}. Similar to the initial
+data-split, five-fold cross-validation was also stratified to maintain
+the overall label distribution on the training and validation sets. We
+validated the cross-validation performances of each hyperparameter
+setting over the 100 randomizations and selected the best performing
+hyper-parameter setting to train the full training/validation dataset
+{[}Figures S1 and S2{]}. We then used the held-out test set to evaluate
+the prediction performance of each ML model. The data-split,
+hyperparameter selection, training and testing steps were repeated 100
+times to get a reliable and robust reading of model prediction
+performance {[}Figure 1{]}.
+
+\textbf{Model performance and generalizability.}
 
@@
 
-\textbf{Predictive performances of the seven ML models during
-cross-validation and when applied to the held-out test data do not show
-large differences.}
-
 We evaluated the generalization performance of each ML model by
 comparing their median cross-validation AUROC and median testing AUROC.
-The largest difference between the two was 0.021 in L1 SVM with linear
-kernel followed by SVM with radial basis function kernel and decision
-tree with a difference of 0.007 and 0.006, respectively.
+This difference should be low to suggest the model is not overfitting
+despite the large number of features. The largest difference between the
+two was 0.021 in L1 SVM with linear kernel followed by SVM with radial
+basis function kernel and decision tree with a difference of 0.007 and
+0.006, respectively {[}Figure 2{]}.
 
-\textbf{The complexity and interpretibility of each ML model.}
+\textbf{Interpretation of each ML model.}
 
@@
 regression using regression coefficients of the trained models. We
-calculated the mean weights and coefficients of all the features over
-the 100 data-splits. In the three linear models, OTUs that had the
-largest mean weights and drove the detection of SRNs belong to family
-\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239, OTU00659,
-OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609,
-OTU01212, OTU00629) and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}.
-We explained the feature importances in non-linear models using
-permutation importance on the held-out test data. For the tree-based
-models, permuting \emph{Peptostreptococcus} (OTU00367) abundances
-randomly, dropped the predictive performances the most. Decision tree,
-random forest and XGBoost models' predictive performance dropped from
-0.6 base testing AUROC median to 0.52, from 0.69 to 0.68 and from 0.68
-to 0.65, respectively {[}Figure 4{]}. The negative predictive impact of
+examined the 5 predictive (highest weight) and protective (lowest
+weight) features identified by the model. We calculated the mean weights
+and coefficients of these features over the 100 data-splits. In the
+three linear models, OTUs that had the largest mean weights and drove
+the detection of SRNs belong to family \emph{Lachnospiraceae}, and
+\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
+OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629)
+and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}. We explained the
+feature importances in non-linear models using permutation importance on
+the held-out test data. For the tree-based models, permuting
+\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
+predictive performances the most. Decision tree, random forest and
+XGBoost models' predictive performance dropped from 0.6 base testing
+AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
+respectively {[}Figure 4{]}. The negative predictive impact of
 \emph{Peptostreptococcus} in the decision tree model was followed by a
@@
 
-Our results show that Suprisingly, l2 logistic.. computational
+In this study we established a robust ML pipeline to use 16S rRNA
+sequence counts to predict a binary health outcome. We showed the
+importance of held-out test set and reporting the results of many
+iterations of cross-validation and testing. Each data-split has a weight
+that comes from random splitting. Seeing that testing and
+cross-validation median differences being low show that the models are
+generalizable and the pipeline works. Our results also show that we can
+choose different models for different reasons. Getting feature
+importances as weights and regression coefficients is the fastest and
+easiest interpratation. Similarly, linear models require less
+computational burden, fastest to train. THinking about the
+applicability, and actionability of a model in clinical setting, we
+might want faster results.
+
+Suprisingly, l2 logistic regression does really well. computational
 efficiency and interp of logistic vs random forest decision tree and l1
@@
 Does the code work on any microbiome data? Why did we use only one
-dataset? Becuase we are comparing several methods on 1 dataset. What
-will the future data look like?
+dataset? Why not validate on other data instead of held-out testing set?
+Becuase we are comparing several methods on 1 dataset. What will the
+future data look like?
 
@@
 England Early Detection Research Network study. These data were obtained
-from Sze et al (22). The stool samples were provided by recruited adult
+from Sze et al (23). The stool samples were provided by recruited adult
 participants who were undergoing scheduled screening or surveillance
@@
 assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(23--25).
+(24--26).
 
@@
 \newpage
+\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
+setting performances for linear models.} Abbreviations: \newpage
+\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
+setting performances for non-linear models.} Abbreviations: \newpage
 
@@
 
+\hypertarget{ref-wong_quantitation_2017}{}
+19. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
+AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
+\textbf{Zhang L}, \textbf{Wu JCY}, \textbf{Chan FKL}, \textbf{Ng SSM},
+\textbf{Wong MCS}, \textbf{Ng SC}, \textbf{Wu WKK}, \textbf{Yu J},
+\textbf{Sung JJY}. 2017. Quantitation of faecal fusobacterium improves
+faecal immunochemical test in detecting advanced colorectal neoplasia.
+Gut \textbf{66}:1441--1448.
+doi:\href{https://doi.org/10.1136/gutjnl-2016-312766}{10.1136/gutjnl-2016-312766}.
+
 \hypertarget{ref-galkin_human_2018}{}
-19. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+20. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
 \textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
@@
 \hypertarget{ref-reiman_using_2017}{}
-20. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+21. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
 convolutional neural networks to explore the microbiome, pp. 4269--4272.
@@
 \hypertarget{ref-fioravanti_phylogenetic_2017}{}
-21. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+22. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
 \textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
@@
 \hypertarget{ref-sze_leveraging_2018}{}
-22. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+23. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-23. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-24. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+25. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-25. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+26. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open

### 2019-05-22 code/learning/Figure2.R

@@ -41,36 +41,39 @@ best_performance <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
 
 
 performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC, fill = Performance)) +
-  geom_boxplot() +
+  geom_boxplot(alpha=0.5, fatten = 4) +
   geom_hline(yintercept = 0.5, linetype="dashed") +
-  scale_fill_manual(values=c("#999999", "#E69F00")) +
+  scale_fill_manual(values=c("blue4", "springgreen4")) +
+  coord_flip() +
   scale_y_continuous(name = "AUROC",
-                     breaks = seq(0.3, 1, 0.02),
-                     limits=c(0.3, 1),
+                     breaks = seq(0.4, 1, 0.1),
+                     limits=c(0.4, 1),
                      expand=c(0,0)) +
   scale_x_discrete(name = "",
-                   labels=c("Decision Tree",
-                           "L1 Linear SVM",
-                            "RBF SVM",
-                            "L2 Linear SVM",
+                   labels=c("Decision tree",
+                           "L1 linear SVM",
+                            "SVM RBF",
+                            "L2 linear SVM",
                             "XGBoost",
-                            "L2 Logistic Regression",
-                            "Random Forest")) +
+                            "L2 logistic regression",
+                            "Random forest")) +
   theme_bw() +
-  theme(legend.title = element_blank(),
+  theme(plot.margin=unit(c(1.1,1.1,1.1,1.1),"cm"),
+        legend.justification=c(1,0),
+        legend.position=c(1,0),
+        #legend.position="bottom",
+        legend.title = element_blank(),
         legend.background = element_rect(size=0.5, linetype="solid", color="black"),
-        legend.justification=c(0,1),
-        legend.position=c(0,1),
-        legend.box.margin=margin(c(10,10,10,10)),
-        legend.text=element_text(size=20),
+        legend.box.margin=margin(c(12,12,12, 12)),
+        legend.text=element_text(size=12),
         #legend.title=element_text(size=22),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         text = element_text(size = 12),
         axis.text.x=element_text(size = 15, colour='black'),
-        axis.text.y=element_text(size = 12, colour='black'),
-        axis.title.y=element_text(size = 30),
[…]

### 2019-05-22 code/learning/Figure3_4.R

@@ -174,7 +174,8 @@ xgboost <- read_files(best_files[7])
 # Define the base plot for the linear modeling methods
 base_plot <-  function(data, x_axis, y_axis){
   plot <- ggplot(data, aes(fct_reorder(x_axis, -abs(y_axis)), y_axis)) +
-    geom_point(colour = "brown2", size = 3) +
+    geom_point(colour = "#D55E00", size = 3) +
+    coord_flip() +
     theme_classic() +
     scale_x_discrete(name = "") +
     theme(legend.text=element_text(size=18),
@@ -199,9 +200,9 @@ base_plot <-  function(data, x_axis, y_axis){
 l1svm <- read.delim("data/process/L1_Linear_SVM_non_cor_importance.tsv", header=T, sep='\t') 
 
 l1svm_plot <- base_plot(l1svm, x=l1svm$key,y=l1svm$mean_weights) +
-  scale_y_continuous(name="L1 linear kernel SVM feature weights",
+  scale_y_continuous(name="L1 linear SVM feature weights",
                     limits = c(-2, 2),
-                    breaks = seq(-2, 2, 0.5)) +
+                    breaks = seq(-2, 2, 1)) +
   geom_errorbar(aes(ymin=l1svm$mean_weights-l1svm$sd_weights, 
                     ymax=l1svm$mean_weights+l1svm$sd_weights), 
                 width=.01) 
@@ -210,9 +211,9 @@ l1svm_plot <- base_plot(l1svm, x=l1svm$key,y=l1svm$mean_weights) +
 # ------------------L2 SVM with linear kernel---------------------------->
 l2svm <- read.delim("data/process/L2_Linear_SVM_non_cor_importance.tsv", header=T, sep='\t') 
 l2svm_plot <- base_plot(l2svm, x=l2svm$key,y=l2svm$mean_weights) +
-  scale_y_continuous(name="L2 linear kernel SVM feature weights",
+  scale_y_continuous(name="L2 linear SVM feature weights",
                      limits = c(-2, 2),
-                     breaks = seq(-2, 2, 0.5)) +    
+                     breaks = seq(-2, 2, 1)) +    
   geom_errorbar(aes(ymin=l2svm$mean_weights-l2svm$sd_weights, 
                     ymax=l2svm$mean_weights+l2svm$sd_weights), 
                 width=.01) 
[…]

### 2019-05-22 code/learning/Figure5.R

@@ -43,33 +43,34 @@ walltime_df <- bind_rows(min_fixed_result)
 
 
 walltime_plot <- ggplot(walltime_df, aes(x = fct_reorder(model, x), y = x)) +
-  geom_boxplot(fill="darkgoldenrod1") +
+  geom_boxplot(fill="#0072B2", alpha=0.5, fatten = 1.5) +
+  coord_flip() +
   scale_y_continuous(name = "Walltime (hours)") +
   scale_x_discrete(name = "",
-                   labels=c("L2 Logistic Regression",
-                     "L1 Linear SVM",
-                     "L2 Linear SVM",
-                     "Decision Tree",
+                   labels=c("L2 logistic regression",
+                     "L1 linear SVM",
+                     "L2 linear SVM",
+                     "Decision tree",
                      "RBF SVM",
-                      "Random Forest", 
+                      "Random forest", 
                      "XGBoost")) +
   theme_bw() +
-  theme(legend.title=element_text(size=22),
+  theme(legend.title=element_text(size=15),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
-        text = element_text(size = 12),
-        axis.text.x=element_text(size = 16, colour='black'),
-        axis.text.y=element_text(size = 16, colour='black'),
-        axis.title.y=element_text(size = 20),
-        axis.title.x=element_text(size = 20), 
+        text = element_text(size = 10),
+        axis.text.x=element_text(size = 8, colour='black'),
+        axis.text.y=element_text(size = 8, colour='black'),
+        axis.title.y=element_text(size = 10),
+        axis.title.x=element_text(size = 10), 
         panel.border = element_rect(colour = "black", fill=NA, size=1)) 
 
 ######################################################################
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 
-ggsave("Figure_5.png", plot = walltime_plot, device = 'png', path = 'submission', width = 15, height = 7)
+ggsave("Figure_5.png", plot = walltime_plot, device = 'png', path = 'submission', width = 3, height = 2)
 
 
             

### 2019-05-22 code/learning/FigureS1_S2.R

@@ -51,6 +51,7 @@ base_plot <-  function(data, x_axis, y_axis){
   geom_line() +
   geom_point() +
   theme_bw() +
+  geom_hline(yintercept = 0.5, linetype="dashed") +
   theme(legend.text=element_text(size=10),
         legend.title=element_text(size=10),
         panel.grid.major = element_blank(),
@@ -75,7 +76,7 @@ l1svm_plot <- base_plot(l1svm, l1svm$cost, l1svm$mean_AUC) +
   scale_y_continuous(name="L1 linear SVM
 mean cvAUROC",
                      limits = c(0.30, 1),
-                     breaks = seq(0.3, 1, 0.05)) +
+                     breaks = seq(0.3, 1, 0.1)) +
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.001)
 
 
@@ -89,7 +90,7 @@ l2svm_plot <- base_plot(l2svm, l2svm$cost, l2svm$mean_AUC) +
   scale_y_continuous(name="L2 linear SVM
 mean cvAUROC",
                      limits = c(0.30, 1),
-                     breaks = seq(0.3, 1, 0.05)) +
+                     breaks = seq(0.3, 1, 0.1)) +
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.001)
 
 logit <- logit_all %>%
@@ -102,7 +103,7 @@ logit_plot <- base_plot(logit, logit$cost, logit$mean_AUC) +
   scale_y_continuous(name="L2 logistic regression
 mean cvAUROC",
                      limits = c(0.30, 1),
-                     breaks = seq(0.3, 1, 0.05)) +
+                     breaks = seq(0.3, 1, 0.1)) +
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.001)
 
 
@@ -115,7 +116,7 @@ scale_x_continuous(name="max depth") +
   scale_y_continuous(name="Decision tree
 mean cvAUROC",
                      limits = c(0.30, 1),
-                     breaks = seq(0.3, 1, 0.05)) +
+                     breaks = seq(0.3, 1, 0.1)) +
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.001)
 
[…]

### 2019-05-24 submission/manuscript.Rmd
update manuscript

@@
 ## Abstract
+
+Members of the human microbiota have been implicated in a wide range of chronic diseases and can be used as biomarkers to diagnose them.
+However, identifying a set of biomarkers and making generalizable disease diagnoses is challenging because human-associated microbiome is high-dimensional and uneven within and between individuals with the same disease.
+Machine learning (ML) modeling of human-associated microbial communities is used to overcome this challenge.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable and reproducible ML models are scarce due to the flawed modeling methods and a lack of transparency in application and selection of ML models.
+We established a standard for ML modeling pipelines that use 16 rRNA gene sequences to predict diseases.
+We used fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions in patients (N=490) from a previously published CRC study.
+We developed seven ML models with different classification algorithms (L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost)) to show the effect of model selection on model accuracy, interpretability and computational efficiency.
+Random forest model was best at detecting CRC lesions but it was slow to train and hard to interpret.
+Despite the simplicity, the L2-regularized logistic regression followed random forest in predictive performance and it trained the fastest.
+Our results in comparing seven ML models with one dataset, showed the importance of model selection in ML model development.
+It also showcased a robust ML pipeline that creates generalizable and reproducible ML models.
+
+## Importance
 

### 2019-05-24 submission/manuscript.tex
update manuscript

@@
 
+Members of the human microbiota have been implicated in a wide range of
+chronic diseases and can be used as biomarkers to diagnose them.
+However, identifying a set of biomarkers and making generalizable
+disease diagnoses is challenging because human-associated microbiome is
+high-dimensional and uneven within and between individuals with the same
+disease. Machine learning (ML) modeling of human-associated microbial
+communities is used to overcome this challenge. Progress has been made
+towards developing ML models that predict health outcomes from bacterial
+abundances, but reliable and reproducible ML models are scarce due to
+the flawed modeling methods and a lack of transparency in application
+and selection of ML models. We established a standard for ML modeling
+pipelines that use 16 rRNA gene sequences to predict diseases. We used
+fecal 16S rRNA sequences to predict the presence/absence of colorectal
+cancer (CRC) lesions in patients (N=490) from a previously published CRC
+study. We developed seven ML models with different classification
+algorithms (L2-regularized logistic regression, L1 and L2 support vector
+machines (SVM) with linear and radial basis function kernels, a decision
+tree, random forest and extreme gradient boosted decision tree
+(XGBoost)) to show the effect of model selection on model accuracy,
+interpretability and computational efficiency. Random forest model was
+best at detecting CRC lesions but it was slow to train and hard to
+interpret. Despite the simplicity, the L2-regularized logistic
+regression followed random forest in predictive performance and it
+trained the fastest. Our results in comparing seven ML models with one
+dataset, showed the importance of model selection in ML model
+development. It also showcased a robust ML pipeline that creates
+generalizable and reproducible ML models.
+
+\subsection{Importance}\label{importance}
+
 \newpage
@@
 receiver operating characteristic curve (AUROC) as the discriminative
-performance metric. Random forest had significantly higher test AUROC
-values than the other models for detecting SRNs. The median AUROC of
-random forest was 0.695 (IQR 0.044). L2 logistic regression, XGBoost, L2
-SVM with linear and radial basis function kernel AUROC values were not
-significantly different from one another. They had median AUROC values
-of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
-0.056) respectively. L1 SVM with linear kernel and decision tree had
-significantly lower AUROC values than the other ML models with median
-AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
-2{]}.
+performance metric {[}Figure 1{]}. Random forest had significantly
+higher test AUROC values than the other models for detecting SRNs. The
+median AUROC of random forest was 0.695 (IQR 0.044). L2 logistic
+regression, XGBoost, L2 SVM with linear and radial basis function kernel
+AUROC values were not significantly different from one another. They had
+median AUROC values of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR
+0.056) and 0.668 (IQR 0.056) respectively. L1 SVM with linear kernel and
+decision tree had significantly lower AUROC values than the other ML
+models with median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059),
+respectively {[}Figure 2{]}.
 

### 2019-05-30 submission/manuscript.Rmd
edit for Jennas comment

--- 2019-05-22 submission/manuscript.Rmd
@@
 
+Machine learning (ML) modeling of the human microbiome is used to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable ML models are scarce in part to the flawed modeling methods.
+Furthermore, using black box ML models has hindered the validation of microbial biomarkers.
+To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
+To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosting (XGBoost).
+The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret.
+Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min).
+This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources.
+It also established standards for modeling pipelines of microbiome-associated ML models.
+
+## Importance
+
 \newpage
@@
 Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health.
-However, this is difficult as human associated microbial communities are remarkably complex and uneven.
+However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease.
 It is unlikely that a single species can explain a disease.
-Instead, subsets of those communities, in relation to one another and to their host, account for the differences in health outcomes.
+Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.
 Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
 Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016;  .
-More notably, flawed ML practices are prevalent such as using ML pipelines where there is no seperate held-out test dataset to evaulate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019].
+More notably, flawed ML practices are prevalent such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
 Moreover, there is a lack of discussion on why a particular ML model is utilized.
@@
 The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability.
-We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines; (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
+We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
 
@@
 We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
-Our ML pipeline utilized held-out test data to evaluate generalization and prediction performance of each ML model.
+Our ML pipeline utilized held-out test data to evaluate generalization performance of each ML model.
 The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`.
@@
 For comparison, we also trained an L1 regularized SVM model with linear kernel.
-L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero), removing features that could be correlated with other important features, making it more difficult include the relationships between the microbes and possibly dropping features that are important in detecting SRNs.
+L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero), removing features that could be correlated with other important features.
 Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.
@@
 We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
-Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution on both the training/validation and test sets.
+Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets.
 Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
-The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
-Validation of hyperparameter selection was performed using 100 randomizations of five-fold cross-validation on the training/validation set [Figure 1].
+The training/validation data was used for training purposes and validation of hyperparameter selection (i.e.
+model selection), and the test set was used for evaluation purposes.
+Validation of hyperparameter selection was performed using repeated five-fold cross-validation on the training/validation set [Figure 1].
 Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets.
-We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting to train the full training/validation dataset [Figures S1 and S2].
+We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting in terms of AUROC to train the full training/validation dataset [Figures S1 and S2].
 We then used the held-out test set to evaluate the prediction performance of each ML model.
@@
 We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric [Figure 1].
-Random forest had significantly higher test AUROC values than the other models for detecting SRNs.
+Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`).
 The median AUROC of random forest was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
@@
 
-We evaluated the generalization performance of each ML model by comparing their median cross-validation AUROC and median testing AUROC.
+We compared the median cross-validation AUROC and median testing AUROC for each model.
 This difference should be low to suggest the model is not overfitting despite the large number of features.
@@
 
+We reported the testing AUROC values over 100 random splits of the full dataset [Figure 1].
+The testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
+For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.
+
 __Interpretation of each ML model.__
 
-The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost increase in complexity as they decrase in interpretibility.
+The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity.
 We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models.
@@
 In the three linear models, OTUs that had the largest mean weights and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629) and genera *Gamella*  (OTU00426) [Figure 3].
-We explained the feature importances in non-linear models using permutation importance on the held-out test data.
+We explained the feature importances in non-linear models using permutation importance on the held-out test data where we permuted groups of correlated features (see methods).
 For the tree-based models, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most.
@@
 #### Interpretation of results
- In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
-We showed the importance of held-out test set and reporting the results of many iterations of cross-validation and testing.
+ - In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+
+ - We showed the importance of held-out test set and reporting the results of many iterations of cross-validation and testing.
 Each data-split has a weight that comes from random splitting.
 Seeing that testing and cross-validation median differences being low show that the models are generalizable and the pipeline works.
-Our results also show that we can choose different models for different reasons.
+
+ - Our results also show that we can choose different models for different reasons.
 Getting feature importances as weights and regression coefficients is the fastest and easiest interpratation.
 Similarly, linear models require less computational burden, fastest to train.
-THinking about the applicability, and actionability of a model in clinical setting, we might want faster results.
+Thinking about the applicability, and actionability of a model in clinical setting, we might want faster results.
 
- Suprisingly, l2 logistic regression does really well. 
- computational efficiency and interp of logistic vs random forest
- decision tree and l1 why so low?
+ -This is a huge difference which shows that you can get "lucky" or "unlucky" depending on the random split.
+It is crucial to report all the findings, many randomizations of these splits to get reliable performance metrics.
+
+ - Suprisingly, l2 logistic regression does really well.
+
+ - Computational efficiency and interp of logistic vs random forest
+
+ - decision tree and l1 why so low?
 
 #### Consideration of possible weaknesses
- What happens with imbalanced data
- What happens with smaller datasets
- Does the code work on any microbiome data?
-  Why did we use only one dataset? Why not validate on other data instead of held-out testing set? Becuase we are comparing several methods on 1 dataset.
-  What will the future data look like?
+ - What happens with imbalanced data
+
+ - What happens with smaller datasets
+
+ - Does the code work on any microbiome data?
+
+  - Why did we use only one dataset?
+
+  - Why not validate on other data instead of held-out testing set? Becuase we are comparing several methods on 1 dataset.
+
+  - What will the future data look like?
 
@@
 
-#### Data collection and study population
+**Data collection and study population.**
   The data used for this analysis are stool bacterial abundances and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study.
@@
 
-#### Data definitions and pre-processing
+**Data definitions and pre-processing.**
 
@@
 
-#### Model training and evaluation
+**Model training and evaluation.**
 
@@
 Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).
+
+**Permutation importance workflow.**
 

### 2019-05-30 submission/manuscript.tex
edit for Jennas comment

--- 2019-05-22 submission/manuscript.tex
@@
 
+Machine learning (ML) modeling of the human microbiome is used to
+identify the microbial biomarkers and aid in diagnosis of many chronic
+diseases such as inflammatory bowel disease, diabetes and colorectal
+cancer. Progress has been made towards developing ML models that predict
+health outcomes from bacterial abundances, but reliable ML models are
+scarce in part to the flawed modeling methods. Furthermore, using black
+box ML models has hindered the validation of microbial biomarkers. To
+overcome these challenges, we benchmarked seven different ML models that
+use fecal 16S rRNA sequences to predict the presence/absence of
+colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
+cases). To show the effect of model selection, we assessed the accuracy,
+interpretability, and computational efficiency of the following models:
+L2-regularized logistic regression, L1 and L2 support vector machines
+(SVM) with linear and radial basis function kernels, a decision tree,
+random forest and extreme gradient boosting (XGBoost). The random forest
+model was best at detecting CRC lesions with an AUROC of 0.695 but it
+was slow to train (101.3 h) and hard to interpret. Despite its
+simplicity, L2-regularized logistic regression followed random forest in
+predictive performance with an AUROC of 0.680, and it trained the
+fastest (12 min). This study showed that we should choose ML models
+based on our expectations of accuracy, interpretability and
+computational efficiency. It also established standards for modeling
+pipelines of microbiome-associated ML models.
+
+\subsection{Importance}\label{importance}
+
 \newpage
@@
 of the microbiome and human health. However, this is difficult as human
-associated microbial communities are remarkably complex and uneven. It
-is unlikely that a single species can explain a disease. Instead,
-subsets of those communities, in relation to one another and to their
-host, account for the differences in health outcomes. Machine learning
+associated microbial communities are remarkably complex,
+high-dimensional and uneven within and between individuals with the same
+disease. It is unlikely that a single species can explain a disease.
+Instead, subsets of those communities, in relation to one another and to
+their host, account for differences in health outcomes. Machine learning
 (ML) methods are effective at recognizing and highlighting patterns in
@@
 consistency on which methods are used and how these methods are
-implemented {[}(12); . More notably, flawed ML practices are prevalent
-such as using ML pipelines where there is no seperate held-out test
-dataset to evaulate model performance, reporting few or only the best
-outcomes of different randomizations of cross-validation and showing a
-disregard for large differences between cross-validation and testing
-performances as well as large confidence intervals of testing
-performances (4, 13--19). Moreover, there is a lack of discussion on why
-a particular ML model is utilized. Recently, there is a trend towards
-using more complex ML models such as random forest, extreme gradient
-boosting and neural networks without a discussion on if and how much
-model interpretibility is necessary for the study (11, 20--22). The lack
-of transparency on modeling methodology and model selection negatively
-impact model reproducibility and reliability. We need to strive toward
-better machine learning practices by (1) implementing consistent and
-reliable machine learning pipelines; (2) selecting ML models that
-reflect the goal of the study as it will inform our expectations of
-model accuracy, complexity, interpretibility and computational
-efficiency.
+implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, flawed ML
+practices are prevalent such as using ML pipelines where there is no
+seperate held-out test dataset to evaluate model performance, reporting
+few or only the best outcomes of different randomizations of
+cross-validation and showing a disregard for large differences between
+cross-validation and testing performances as well as large confidence
+intervals of testing performances (4, 13--19). Moreover, there is a lack
+of discussion on why a particular ML model is utilized. Recently, there
+is a trend towards using more complex ML models such as random forest,
+extreme gradient boosting and neural networks without a discussion on if
+and how much model interpretibility is necessary for the study (11,
+20--22). The lack of transparency on modeling methodology and model
+selection negatively impact model reproducibility and reliability. We
+need to strive toward better machine learning practices by (1)
+implementing consistent and reliable machine learning pipelines and (2)
+selecting ML models that reflect the goal of the study as it will inform
+our expectations of model accuracy, complexity, interpretibility and
+computational efficiency.
 
@@
 random forest and XGBoost. Our ML pipeline utilized held-out test data
-to evaluate generalization and prediction performance of each ML model.
-The median test AUROC varied from 0.601 to 0.695. Random forest had the
-highest median AUROC for detecting SRN. Despite the simplicity, the
-L2-regularized logistic regression followed random forest in
-performance. In terms of computational efficiency, L2 logistic
-regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
-took the longest (162.843 hours, std ± 3.986). We found that median
-cross-validation and testing AUROC varied only 0.021, which highlights
-the importance of a seperate held-out test set and consistent
-preprocessing of the data prior to evaluation. Aside from evaluating
-generalization and classification performances for each of these models,
-this study established standards for modeling pipelines of
-microbiome-associated machine learning models.
+to evaluate generalization performance of each ML model. The median test
+AUROC varied from 0.601 to 0.695. Random forest had the highest median
+AUROC for detecting SRN. Despite the simplicity, the L2-regularized
+logistic regression followed random forest in performance. In terms of
+computational efficiency, L2 logistic regression trained the fastest
+(0.202 hours, std ± 0.019), while XGBoost took the longest (162.843
+hours, std ± 3.986). We found that median cross-validation and testing
+AUROC varied only 0.021, which highlights the importance of a seperate
+held-out test set and consistent preprocessing of the data prior to
+evaluation. Aside from evaluating generalization and classification
+performances for each of these models, this study established standards
+for modeling pipelines of microbiome-associated machine learning models.
 
@@
 (i.e., force many coefficients to zero), removing features that could be
-correlated with other important features, making it more difficult
-include the relationships between the microbes and possibly dropping
-features that are important in detecting SRNs. Finally, to explore the
+correlated with other important features. Finally, to explore the
 potential for non-linear relationships among features and the outcome of
@@
 cases are not uniformly represented in the data, the initial data-split
-was stratified to maintain the overall label distribution on both the
+was stratified to maintain the overall label distribution in both the
 training/validation and test sets. Training/validation set consisted of
@@
 (52 SRN). The training/validation data was used for training purposes
-and validation of hyperparameter selection, and the test set was used
-for evaluation purposes. Validation of hyperparameter selection was
-performed using 100 randomizations of five-fold cross-validation on the
-training/validation set {[}Figure 1{]}. Similar to the initial
-data-split, five-fold cross-validation was also stratified to maintain
-the overall label distribution on the training and validation sets. We
-validated the cross-validation performances of each hyperparameter
-setting over the 100 randomizations and selected the best performing
-hyper-parameter setting to train the full training/validation dataset
-{[}Figures S1 and S2{]}. We then used the held-out test set to evaluate
-the prediction performance of each ML model. The data-split,
-hyperparameter selection, training and testing steps were repeated 100
-times to get a reliable and robust reading of model prediction
-performance {[}Figure 1{]}.
+and validation of hyperparameter selection (i.e.~model selection), and
+the test set was used for evaluation purposes. Validation of
+hyperparameter selection was performed using repeated five-fold
+cross-validation on the training/validation set {[}Figure 1{]}. Similar
+to the initial data-split, five-fold cross-validation was also
+stratified to maintain the overall label distribution on the training
+and validation sets. We validated the cross-validation performances of
+each hyperparameter setting over the 100 randomizations and selected the
+best performing hyper-parameter setting in terms of AUROC to train the
+full training/validation dataset {[}Figures S1 and S2{]}. We then used
+the held-out test set to evaluate the prediction performance of each ML
+model. The data-split, hyperparameter selection, training and testing
+steps were repeated 100 times to get a reliable and robust reading of
+model prediction performance {[}Figure 1{]}.
 
@@
 receiver operating characteristic curve (AUROC) as the discriminative
-performance metric. Random forest had significantly higher test AUROC
-values than the other models for detecting SRNs. The median AUROC of
-random forest was 0.695 (IQR 0.044). L2 logistic regression, XGBoost, L2
-SVM with linear and radial basis function kernel AUROC values were not
-significantly different from one another. They had median AUROC values
-of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
-0.056) respectively. L1 SVM with linear kernel and decision tree had
-significantly lower AUROC values than the other ML models with median
-AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
-2{]}.
+performance metric {[}Figure 1{]}. Random forest had significantly
+higher test AUROC values than the other models for detecting SRNs when
+AUROC values were compared to the other six by Wilcoxon rank sum test (p
+= 0.0011). The median AUROC of random forest was 0.695 (IQR 0.044). L2
+logistic regression, XGBoost, L2 SVM with linear and radial basis
+function kernel AUROC values were not significantly different from one
+another. They had median AUROC values of 0.68 (IQR 0.055), 0.679 (IQR
+0.053), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1 SVM
+with linear kernel and decision tree had significantly lower AUROC
+values than the other ML models with median AUROC of 0.65 (IQR 0.066)
+and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.
 
-We evaluated the generalization performance of each ML model by
-comparing their median cross-validation AUROC and median testing AUROC.
-This difference should be low to suggest the model is not overfitting
-despite the large number of features. The largest difference between the
-two was 0.021 in L1 SVM with linear kernel followed by SVM with radial
-basis function kernel and decision tree with a difference of 0.007 and
-0.006, respectively {[}Figure 2{]}.
+We compared the median cross-validation AUROC and median testing AUROC
+for each model. This difference should be low to suggest the model is
+not overfitting despite the large number of features. The largest
+difference between the two was 0.021 in L1 SVM with linear kernel
+followed by SVM with radial basis function kernel and decision tree with
+a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
+
+We reported the testing AUROC values over 100 random splits of the full
+dataset {[}Figure 1{]}. The testing AUROC values varied 0.23 on average,
+depending on the data-split. For instance, the lowest AUROC value of the
+random forest model was 0.59 whereas the highest was 0.81.
 
@@
 L2 support vector machines (SVM) with linear and radial basis function
-kernels, a decision tree, random forest and XGBoost increase in
-complexity as they decrase in interpretibility. We interpreted L1 and L2
-SVM with linear kernel using the feature weights and L2 logistic
+kernels, a decision tree, random forest and XGBoost decrease in
+interpretibility as they increase in complexity. We interpreted L1 and
+L2 SVM with linear kernel using the feature weights and L2 logistic
 regression using regression coefficients of the trained models. We
@@
 feature importances in non-linear models using permutation importance on
-the held-out test data. For the tree-based models, permuting
+the held-out test data where we permuted groups of correlated features
+(see methods). For the tree-based models, permuting
 \emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
@@
 
+\begin{itemize}
+\item
   In this study we established a robust ML pipeline to use 16S rRNA
-sequence counts to predict a binary health outcome. We showed the
-importance of held-out test set and reporting the results of many
-iterations of cross-validation and testing. Each data-split has a weight
-that comes from random splitting. Seeing that testing and
-cross-validation median differences being low show that the models are
-generalizable and the pipeline works. Our results also show that we can
-choose different models for different reasons. Getting feature
-importances as weights and regression coefficients is the fastest and
-easiest interpratation. Similarly, linear models require less
-computational burden, fastest to train. THinking about the
-applicability, and actionability of a model in clinical setting, we
-might want faster results.
+  sequence counts to predict a binary health outcome.
+\item
+  We showed the importance of held-out test set and reporting the
+  results of many iterations of cross-validation and testing. Each
+  data-split has a weight that comes from random splitting. Seeing that
+  testing and cross-validation median differences being low show that
+  the models are generalizable and the pipeline works.
+\item
+  Our results also show that we can choose different models for
+  different reasons. Getting feature importances as weights and
+  regression coefficients is the fastest and easiest interpratation.
+  Similarly, linear models require less computational burden, fastest to
+  train. Thinking about the applicability, and actionability of a model
+  in clinical setting, we might want faster results.
+\end{itemize}
 
-Suprisingly, l2 logistic regression does really well. computational
-efficiency and interp of logistic vs random forest decision tree and l1
-why so low?
+-This is a huge difference which shows that you can get ``lucky'' or
+``unlucky'' depending on the random split. It is crucial to report all
+the findings, many randomizations of these splits to get reliable
+performance metrics.
+
+\begin{itemize}
+\item
+  Suprisingly, l2 logistic regression does really well.
+\item
+  Computational efficiency and interp of logistic vs random forest
+\item
+  decision tree and l1 why so low?
+\end{itemize}
 
@@
 
-What happens with imbalanced data What happens with smaller datasets
-Does the code work on any microbiome data? Why did we use only one
-dataset? Why not validate on other data instead of held-out testing set?
-Becuase we are comparing several methods on 1 dataset. What will the
-future data look like?
+\begin{itemize}
+\item
+  What happens with imbalanced data
+\item
+  What happens with smaller datasets
+\item
+  Does the code work on any microbiome data?
+\item
+  Why did we use only one dataset?
+\item
+  Why not validate on other data instead of held-out testing set?
+  Becuase we are comparing several methods on 1 dataset.
+\item
+  What will the future data look like?
+\end{itemize}
 
@@
 
-\paragraph{Data collection and study
-population}\label{data-collection-and-study-population}
-
-The data used for this analysis are stool bacterial abundances and
-clinical information of the patients recruited by Great Lakes-New
-England Early Detection Research Network study. These data were obtained
-from Sze et al (23). The stool samples were provided by recruited adult
-participants who were undergoing scheduled screening or surveillance
-colonoscopy. Colonoscopies were performed and fecal samples were
-collected from participants in four locations: Toronto (ON, Canada),
-Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients'
-colonic health was labeled by colonoscopy with adequate preparation and
-tissue histopathology of all resected lesions. Patients with an adenoma
-greater than 1 cm, more than three adenomas of any size, or an adenoma
-with villous histology were classified as advanced adenoma. Study had
-172 patients with normal colonoscopies, 198 with adenomas and 120 with
+\textbf{Data collection and study population.} The data used for this
+analysis are stool bacterial abundances and clinical information of the
+patients recruited by Great Lakes-New England Early Detection Research
+Network study. These data were obtained from Sze et al (23). The stool
+samples were provided by recruited adult participants who were
+undergoing scheduled screening or surveillance colonoscopy.
+Colonoscopies were performed and fecal samples were collected from
+participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
+Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
+labeled by colonoscopy with adequate preparation and tissue
+histopathology of all resected lesions. Patients with an adenoma greater
+than 1 cm, more than three adenomas of any size, or an adenoma with
+villous histology were classified as advanced adenoma. Study had 172
+patients with normal colonoscopies, 198 with adenomas and 120 with
 carcinomas. Of the 198 adenomas, 109 were identified as advanced
@@
 
-\paragraph{Data definitions and
-pre-processing}\label{data-definitions-and-pre-processing}
+\textbf{Data definitions and pre-processing.}
 
@@
 
-\paragraph{Model training and
-evaluation}\label{model-training-and-evaluation}
+\textbf{Model training and evaluation.}
 
@@
 package (v.6.0.81) in R (v.3.5.0).
+
+\textbf{Permutation importance workflow.}
 

### 2019-05-31 submission/manuscript.Rmd
update manuscript after Jennas edits?

@@
 ---
-title: "**Evaluation of machine learning methods for 16S rRNA gene data**"
-bibliography: references.bib
+title: '**Evaluation of machine learning methods for 16S rRNA gene data**'
+csl: mbio.csl
+fontsize: 11pt
 output:
   pdf_document:
-    keep_tex: true
     includes:
       in_header: header.tex
-csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
-fontsize: 11pt
+    keep_tex: yes
+  word_document: default
 geometry: margin=1.0in
+bibliography: references.bib
 ---
@@
 Machine learning (ML) modeling of the human microbiome is used to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable ML models are scarce in part to the flawed modeling methods.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable ML models are scarce in part due to the flawed modeling methods.
 Furthermore, using black box ML models has hindered the validation of microbial biomarkers.
 To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
-To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosting (XGBoost).
+To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
 The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret.
@@
 
-## Introduction
+## Background
 
@@
 Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.
+
 Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
@@
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
+
 The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability.
@@
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients.
+CRC is the third most common form of cancer worldwide in which the human-associated human microbiome is hypothesized to directly contribute to its development.
 We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN).
@@
 Random forest had the highest median AUROC for detecting SRN.
-Despite the simplicity, the L2-regularized logistic regression followed random forest in performance.
+Despite its simplicity, the L2-regularized logistic regression followed random forest in performance.
 In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`).
-We found that median cross-validation and testing AUROC varied only `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
+We found that median cross-validation and testing AUROC varied `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
 Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
@@
 
-__Model performance and generalizability.__
+__Discriminative performance of the seven models.__
 
@@
 Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`).
-The median AUROC of random forest was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
-L2 logistic regression, XGBoost, L2 SVM with linear and radial basis function kernel AUROC values were not significantly different from one another.
+The median AUROC of therandom forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another.
 They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively.
 L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
+
+__Generalizability of the seven models.__
 
@@
 #### Interpretation of results
+
 - In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
+When the differences between cross-validation and test performance is low, this suggets the models will perform similar with similar data.
+In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data.
 
- - We showed the importance of held-out test set and reporting the results of many iterations of cross-validation and testing.
-Each data-split has a weight that comes from random splitting.
-Seeing that testing and cross-validation median differences being low show that the models are generalizable and the pipeline works.
+- We also reported the results of 100 random dataset splits because there are differences in AUROC values of each of the seven models between different random data-splits.
+Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
+This result shows that you can get "lucky" in terms of prediction performance depending on the random split.
+Therefore, it is crucial to report all the findings over many random data-splits to get reliable performance metrics.
 
- - Our results also show that we can choose different models for different reasons.
+ - Our results also showed that we can choose different models for different reasons.
 Getting feature importances as weights and regression coefficients is the fastest and easiest interpratation.
@@
 
- -This is a huge difference which shows that you can get "lucky" or "unlucky" depending on the random split.
-It is crucial to report all the findings, many randomizations of these splits to get reliable performance metrics.
+ -
 

### 2019-05-31 submission/manuscript.tex
update manuscript after Jennas edits?

@@
 health outcomes from bacterial abundances, but reliable ML models are
-scarce in part to the flawed modeling methods. Furthermore, using black
-box ML models has hindered the validation of microbial biomarkers. To
-overcome these challenges, we benchmarked seven different ML models that
-use fecal 16S rRNA sequences to predict the presence/absence of
+scarce in part due to the flawed modeling methods. Furthermore, using
+black box ML models has hindered the validation of microbial biomarkers.
+To overcome these challenges, we benchmarked seven different ML models
+that use fecal 16S rRNA sequences to predict the presence/absence of
 colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
@@
 (SVM) with linear and radial basis function kernels, a decision tree,
-random forest and extreme gradient boosting (XGBoost). The random forest
-model was best at detecting CRC lesions with an AUROC of 0.695 but it
-was slow to train (101.3 h) and hard to interpret. Despite its
+random forest, and extreme gradient boosting (XGBoost). The random
+forest model was best at detecting CRC lesions with an AUROC of 0.695
+but it was slow to train (101.3 h) and hard to interpret. Despite its
 simplicity, L2-regularized logistic regression followed random forest in
@@
 fastest (12 min). This study showed that we should choose ML models
-based on our expectations of accuracy, interpretability and
-computational efficiency. It also established standards for modeling
+based on our expectations of performance and interpretability as well as
+our computational resources. It also established standards for modeling
 pipelines of microbiome-associated ML models.
@@
 
-\subsection{Introduction}\label{introduction}
+\subsection{Background}\label{background}
 
@@
 Instead, subsets of those communities, in relation to one another and to
-their host, account for differences in health outcomes. Machine learning
-(ML) methods are effective at recognizing and highlighting patterns in
-complex microbial datasets. Therefore, researchers have started to
-explore the utility of ML models that use microbiota associated
-biomarkers to predict human health and to understand the microbial
-ecology of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
-(1--11). However, currently the field's use of ML lacks clarity and
-consistency on which methods are used and how these methods are
-implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, flawed ML
-practices are prevalent such as using ML pipelines where there is no
-seperate held-out test dataset to evaluate model performance, reporting
-few or only the best outcomes of different randomizations of
+their host, account for differences in health outcomes.
+
+Machine learning (ML) methods are effective at recognizing and
+highlighting patterns in complex microbial datasets. Therefore,
+researchers have started to explore the utility of ML models that use
+microbiota associated biomarkers to predict human health and to
+understand the microbial ecology of diseases such as liver cirrhosis,
+colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
+diabetes and others (1--11). However, currently the field's use of ML
+lacks clarity and consistency on which methods are used and how these
+methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably,
+flawed ML practices are prevalent such as using ML pipelines where there
+is no seperate held-out test dataset to evaluate model performance,
+reporting few or only the best outcomes of different randomizations of
 cross-validation and showing a disregard for large differences between
@@
 and how much model interpretibility is necessary for the study (11,
-20--22). The lack of transparency on modeling methodology and model
-selection negatively impact model reproducibility and reliability. We
-need to strive toward better machine learning practices by (1)
-implementing consistent and reliable machine learning pipelines and (2)
-selecting ML models that reflect the goal of the study as it will inform
-our expectations of model accuracy, complexity, interpretibility and
+20--22).
+
+The lack of transparency on modeling methodology and model selection
+negatively impact model reproducibility and reliability. We need to
+strive toward better machine learning practices by (1) implementing
+consistent and reliable machine learning pipelines and (2) selecting ML
+models that reflect the goal of the study as it will inform our
+expectations of model accuracy, complexity, interpretibility and
 computational efficiency.
@@
 cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
-patients. We built ML models using fecal 16S rRNA gene sequences to
-predict patients with normal colons or patients with colonic tumors
-which are called screen relevant neoplasias (SRN). The study had 261
-normal and 229 SRN samples. We established modeling pipelines for
-L2-regularized logistic regression, L1 and L2 support vector machines
-(SVM) with linear and radial basis function kernels, a decision tree,
-random forest and XGBoost. Our ML pipeline utilized held-out test data
-to evaluate generalization performance of each ML model. The median test
-AUROC varied from 0.601 to 0.695. Random forest had the highest median
-AUROC for detecting SRN. Despite the simplicity, the L2-regularized
-logistic regression followed random forest in performance. In terms of
-computational efficiency, L2 logistic regression trained the fastest
-(0.202 hours, std ± 0.019), while XGBoost took the longest (162.843
-hours, std ± 3.986). We found that median cross-validation and testing
-AUROC varied only 0.021, which highlights the importance of a seperate
-held-out test set and consistent preprocessing of the data prior to
-evaluation. Aside from evaluating generalization and classification
-performances for each of these models, this study established standards
-for modeling pipelines of microbiome-associated machine learning models.
+patients. CRC is the third most common form of cancer worldwide in which
+the human-associated human microbiome is hypothesized to directly
+contribute to its development. We built ML models using fecal 16S rRNA
+gene sequences to predict patients with normal colons or patients with
+colonic tumors which are called screen relevant neoplasias (SRN). The
+study had 261 normal and 229 SRN samples. We established modeling
+pipelines for L2-regularized logistic regression, L1 and L2 support
+vector machines (SVM) with linear and radial basis function kernels, a
+decision tree, random forest and XGBoost. Our ML pipeline utilized
+held-out test data to evaluate generalization performance of each ML
+model. The median test AUROC varied from 0.601 to 0.695. Random forest
+had the highest median AUROC for detecting SRN. Despite its simplicity,
+the L2-regularized logistic regression followed random forest in
+performance. In terms of computational efficiency, L2 logistic
+regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
+took the longest (162.843 hours, std ± 3.986). We found that median
+cross-validation and testing AUROC varied 0.021, which highlights the
+importance of a seperate held-out test set and consistent preprocessing
+of the data prior to evaluation. Aside from evaluating generalization
+and classification performances for each of these models, this study
+established standards for modeling pipelines of microbiome-associated
+machine learning models.
 
@@
 
-\textbf{Model performance and generalizability.}
+\textbf{Discriminative performance of the seven models.}
 
@@
 AUROC values were compared to the other six by Wilcoxon rank sum test (p
-= 0.0011). The median AUROC of random forest was 0.695 (IQR 0.044). L2
-logistic regression, XGBoost, L2 SVM with linear and radial basis
-function kernel AUROC values were not significantly different from one
-another. They had median AUROC values of 0.68 (IQR 0.055), 0.679 (IQR
-0.053), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1 SVM
-with linear kernel and decision tree had significantly lower AUROC
-values than the other ML models with median AUROC of 0.65 (IQR 0.066)
-and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.
+= 0.0011). The median AUROC of therandom forest model was 0.695 (IQR
+0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
+with linear and radial basis function kernel AUROC values were not
+significantly different from one another. They had median AUROC values
+of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
+0.056) respectively. L1 SVM with linear kernel and decision tree had
+significantly lower AUROC values than the other ML models with median
+AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
+2{]}.
+
+\textbf{Generalizability of the seven models.}
 
@@
   In this study we established a robust ML pipeline to use 16S rRNA
-  sequence counts to predict a binary health outcome.
+  sequence counts to predict a binary health outcome. We used a held-out
+  test set to illustrate the difference between cross-validation and
+  testing AUROC values. When the differences between cross-validation
+  and test performance is low, this suggets the models will perform
+  similar with similar data. In all seven models, the differences in
+  cross-validation and testing AUROC values did not exceed 0.021 which
+  suggests that these models will be able to test similar new data.
 \item
-  We showed the importance of held-out test set and reporting the
-  results of many iterations of cross-validation and testing. Each
-  data-split has a weight that comes from random splitting. Seeing that
-  testing and cross-validation median differences being low show that
-  the models are generalizable and the pipeline works.
+  We also reported the results of 100 random dataset splits because
+  there are differences in AUROC values of each of the seven models
+  between different random data-splits. Our results showed that the
+  testing AUROC values varied 0.23 on average, depending on the
+  data-split. This result shows that you can get ``lucky'' in terms of
+  prediction performance depending on the random split. Therefore, it is
+  crucial to report all the findings over many random data-splits to get
+  reliable performance metrics.
 \item
-  Our results also show that we can choose different models for
+  Our results also showed that we can choose different models for
   different reasons. Getting feature importances as weights and
@@
   in clinical setting, we might want faster results.
-\end{itemize}
-
--This is a huge difference which shows that you can get ``lucky'' or
-``unlucky'' depending on the random split. It is crucial to report all
-the findings, many randomizations of these splits to get reliable
-performance metrics.
-
-\begin{itemize}
+\item
 \item

### 2019-05-31 code/learning/Figure2.R

@@ -43,6 +43,10 @@ best_performance <- bind_rows(logit, l2svm, rbf, dt, xgboost, rf, l1svm)%>%
 performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC, fill = Performance)) +
   geom_boxplot(alpha=0.5, fatten = 4) +
   geom_hline(yintercept = 0.5, linetype="dashed") +
+  #geom_hline(yintercept = 0.6, linetype="dashed") +
+  #geom_hline(yintercept = 0.7, linetype="dashed") +
+  #geom_hline(yintercept = 0.8, linetype="dashed") +
+  #geom_hline(yintercept = 0.9, linetype="dashed") +
   scale_fill_manual(values=c("blue4", "springgreen4")) +
   coord_flip() +
   scale_y_continuous(name = "AUROC",
@@ -67,7 +71,8 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
         legend.box.margin=margin(c(12,12,12, 12)),
         legend.text=element_text(size=12),
         #legend.title=element_text(size=22),
-        panel.grid.major = element_blank(),
+        panel.grid.major.y = element_blank(),
+        panel.grid.major.x = element_line( size=0.6),
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         text = element_text(size = 12),

### 2019-06-05 submission/manuscript.Rmd
add some comments from meeting
intergrate jennas changes

@@
 
-Machine learning (ML) modeling of the human microbiome is used to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable ML models are scarce in part due to the flawed modeling methods.
+Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed (suboptimal) modeling methods that call into question the validity of developed ML models.
 Furthermore, using black box ML models has hindered the validation of microbial biomarkers.
@@
 ## Importance
+
+- Without rigorous pipelines, we can't believe the machine learning models.
+
+- To speed up progress, let's slow down and make it right.
+
+- Let's define and start using good ML practices.
 
@@
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019].
-More notably, flawed ML practices are prevalent such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+More notably, ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance and reporting few or only the best outcomes of different randomizations of cross-validation are commonly used.
+When there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
 Moreover, there is a lack of discussion on why a particular ML model is utilized.
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-
 The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability.
-We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
+We need to strive toward better machine learning practices by (1) implementing consistent and rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
 
@@
 
+- decision tree has no rgularization.
+
 __The computational efficiency of each ML model.__
@@
 Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[walltime_index[4],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[4],3], 1))`), `r paste(round(walltime_df[walltime_index[5],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[5],3], 1))`), `r paste(round(walltime_df[walltime_index[6],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[6],3], 1))`) and `r paste(round(walltime_df[walltime_index[7],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[7],3], 1))`), respectively [Figure 4].
+
+- Testing with each model will take the same amount of time but if you are training a model and optimize hyper-parameters, then you need to consider the computational burden.
+The computational efficiency of different models have been investigated before.
+Explain for each model.
+In tree based models, depth of tree, number of features, ..
+matter.
+L1 linear is very fast.
+
+- But we can reduce time paralellizing different components of the pipeline.
+Here I have parallelized each datasplit but we can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.
 
@@
 
- -
+ - There is variability if we don't do a full grid search for the hyper-parameters.
+We can give each model 15 minutes to train and see how well they do.
+So an L1 linear model will give you terible results with one and not the other C value as seen in figure S1.
+We have done a granular search first and then limited the possible settings to a range that made the more sense.
 
@@
  - What happens with imbalanced data
+ The metric you use as the discriminative performance becomes important. Don't use accuracy.
+ AUROC is good and random at 0.5 always.
 
  - What happens with smaller datasets
+ Generalization performance depends on size. 
+ The more complex the model, more data you will need.
+ Use linear models if you have little data.
 
  - Does the code work on any microbiome data?
+ So if yu have a new sample, do you have the same OTUs? NO
 
@@
 For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
+
+Hyperparameter selection is very important.
+We start with a granular grid search by then narrow and refine.
+Range of the grid depends on the problem.
+We can use hyper-band to help us with our hyperparameter selection.
+If not a full grid search is done, there could be a lot of variability in testing performance.
 Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).

### 2019-06-07 submission/manuscript.Rmd
redo interpret figures based on Jenna

@@
 **Permutation importance workflow.**
+We created a Spearman's rank-order correlation matrix, corrected for multiple pairwise comparisons.
+We then defined correlated OTUs as having perfect correlation (correlation coef=1 and p<0.01).Other OTUs were permuted individually to get permutation importance but the correlated ones are grouped together and permuted at the same time.
+The reported OTUs that have impact on testing AUROC are reported in Figures xxx.
+If we want we can decrease the correlation coefficient to consider OTUs that are correlated with ecological consequences but this was out of the scope of this study but will be followed up in further analyses.
 

### 2019-06-07 code/learning/Figure4.R

@@ -78,6 +78,44 @@ dt <- read_files(best_files[1])
 xgboost <- read_files(best_files[7])
 
 
+# Define the base plot 
+top_five_otus <-  function(name){
+  
+  data_first_five <- read.delim(paste0("data/process/", name, "_non_cor_importance.tsv"), header=T, sep='\t') %>%
+    arrange(imp) %>% 
+    head(5)
+  
+  return(data_first_five)
+}
+
+# --------------------------- RBF_SVM ----------------------------->
+# Plot most important 5 features effect on AUROC
+rbf_top <- top_five_otus("RBF_SVM")
+
+# --------------------------- Decision Tree ----------------------------->
+# Plot most important 5 features effect on AUROC
+dt_top <- top_five_otus("Decision_Tree")
+# ----------------------------------------------------------------------->
+
+# --------------------------- Random Forest ----------------------------->
+# Plot most important 5 features effect on AUROC
+rf_top <- top_five_otus("Random_Forest")
+# ----------------------------------------------------------------------->
+
+# --------------------------- XGBoost ----------------------------->
+# Plot most important 5 features effect on AUROC
+xgboost_top <- top_five_otus("XGBoost")
+
+intersect_three <- Reduce(intersect, list(dt_top$names,rf_top$names,xgboost_top$names))
+intersect_rf_dt <- setdiff(intersect(dt_top$names,rf_top$names), intersect_three)
+intersect_xgboost_dt <- setdiff(intersect(dt_top$names,xgboost_top$names), intersect_three)
+intersect_xgboost_rf <- setdiff(intersect(rf_top$names,xgboost_top$names), intersect_three)
+intersect_rbf_dt <- setdiff(intersect(rbf_top$names,dt_top$names), intersect_three)
+intersect_rbf_xgboost <- setdiff(intersect(xgboost_top$names,rbf_top$names), intersect_three)
+
+list1 <- do.call(c, list(intersect_three, intersect_rf_dt, intersect_xgboost_dt, intersect_xgboost_rf, intersect_rbf_xgboost, intersect_rbf_dt))
+
+
 
 
 ######################################################################
@@ -85,7 +123,7 @@ xgboost <- read_files(best_files[7])
 ######################################################################
 # -----------------------Base plot function -------------------------->
 # Define the base plot 
[…]

### 2019-06-07 code/learning/FigureS3.R

@@ -37,14 +37,14 @@ grab_importance <-  function(name){
   return(data_first_10)
 }
 
-x <- list("Random_Forest", "XGBoost")
+x <- list("Decision_Tree", "Random_Forest", "XGBoost")
 
 total <- map_df(x, grab_importance) %>% group_by(model)
 
 # Grouped
 ggplot(total, aes(y=rank, x=names, fill=model)) +
-  geom_bar(stat="identity", width=0.2, position = position_dodge2(preserve = "single")) +
-  geom_point(position = position_dodge(width=0.15), size=1, aes(color=model)) +
+  geom_bar(stat="identity", width=0.35, position = position_dodge2(preserve = "single")) +
+  geom_point(position = position_dodge(width=0.25), size=1, aes(color=model)) +
   scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9", "black")) +
   scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9", "black")) +
   coord_flip() +

### 2019-06-07 code/learning/permutation_importance.R

@@ -5,7 +5,7 @@
 ######################################################################
 
 ######################################################################
-# Description: 
+# Description:
 
 # This script will read in:
 #     - Trained model
@@ -17,12 +17,12 @@
 #     - Predict the held-out test data for one data-split
 #     - Permutate each feature in the test data randomly
 #     - Predict transformed test data for each permutation
-#     - Substract transformed prediction auc from original prediction auc 
+#     - Substract transformed prediction auc from original prediction auc
 #     - Determine which feature makes the biggest difference in auc
 ######################################################################
 
 ######################################################################
-# Dependencies and Outputs: 
+# Dependencies and Outputs:
 
 #     - The funtion needs transformed test set
 #     - Trained model for one data-split
@@ -30,7 +30,7 @@
 # Be in the project directory.
 
 # The outputs are:
-#   (1) AUC difference for each feature transformation 
+#   (1) AUC difference for each feature transformation
 ######################################################################
 
 
@@ -39,8 +39,8 @@
 deps = c("tictoc", "caret", "pROC", "tidyverse");
 for (dep in deps){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), 
-                     quiet=TRUE, 
+    install.packages(as.character(dep),
+                     quiet=TRUE,
                      repos = "http://cran.us.r-project.org", dependencies=TRUE);
   }
   library(dep, verbose=FALSE, character.only=TRUE)
@@ -51,52 +51,52 @@ for (dep in deps){
 
 ####################### DEFINE FUNCTION  #############################
 permutation_importance <- function(model, full){
-  
-  # -----------Get the original testAUC from held-out test data--------->  
+
+  # -----------Get the original testAUC from held-out test data--------->
   # Calculate the test-auc for the actual pre-processed held-out data
   rpartProbs <- predict(model, full, type="prob")
   base_roc <- roc(ifelse(full$dx == "cancer", 1, 0), rpartProbs[[1]])
   base_auc <- base_roc$auc
-  # -------------------------------------------------------------------->  
-  
-  # ----------- Read in the correlation matrix of full dataset---------->  
+  # -------------------------------------------------------------------->
+
[…]

### 2019-06-10 submission/TableS1.Rmd
table S1

---
title: ""
output:
  pdf_document:
    keep_tex: false
    includes:
      in_header: header.tex
geometry: margin=1.0in
fontsize: 11pt
---

\begin{center}
\captionof{table}{Model complexity: $n$ is the number of training samples, $f$ the number of features, $n_{trees}$ is the number of trees.}
\begin{tabular}{|l|l|}
\hline
\textbf{Machine learning model} & \textbf{Training complexity} \\ \hline
Logistic regression & $O(nf)$ \\ \hline
SVM with linear kernel & $O(nf)$  \\ \hline
SVM with radial basis kernel & $O(n^2f)$  \\ \hline
Decision tree & $O(n^2f)$  \\ \hline
Random forest & $O(n^2fn_{trees})$  \\ \hline
XGBoost & $O(nfn_{trees})?$  \\ \hline
\end{tabular}
\end{center}

### 2019-06-10 submission/manuscript.Rmd
table S1

@@
 Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed (suboptimal) modeling methods that call into question the validity of developed ML models.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed modeling methods that call into question the validity of developed ML models.
 Furthermore, using black box ML models has hindered the validation of microbial biomarkers.
@@
 
-- Without rigorous pipelines, we can't believe the machine learning models.
-
-- To speed up progress, let's slow down and make it right.
-
-- Let's define and start using good ML practices.
+Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies.
+However, the results so far are overoptimistic both in terms of prediction performance and validity of developed ML models.
+Without rigorous ML pipelines, we cannot trust ML models.
+Before we can speed up progress, we need to slow down, define and start using good ML practices.
 
@@
 For comparison, we also trained an L1 regularized SVM model with linear kernel.
-L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero), removing features that could be correlated with other important features.
+L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero).
 Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.
@@
 Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`).
-The median AUROC of therandom forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
+The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
 L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another.
@@
 The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity.
-We interpreted L1 and L2 SVM with linear kernel using the feature weights and L2 logistic regression using regression coefficients of the trained models.
-We examined the 5 predictive (highest weight) and protective (lowest weight) features identified by the model.
-We calculated the mean weights and coefficients of these features over the 100 data-splits.
-In the three linear models, OTUs that had the largest mean weights and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629) and genera *Gamella*  (OTU00426) [Figure 3].
-We explained the feature importances in non-linear models using permutation importance on the held-out test data where we permuted groups of correlated features (see methods).
-For the tree-based models, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most.
+We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models.
+We ranked the absolute weights of all the OTUs for each data-split.
+We calculated the median ranks of these features over the 100 data-splits.
+In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609) and genera *Gamella*  (OTU00426) [Figure 3].
+Some of the OTUs with the highest ranks were shared among the linear models.
+We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect on testing AUROC.
+The top 5 OTUs with the largest negative impact on testing AUROC were overl in tree-based models [Figure 4].
+Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods.
 Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
@@
 Other OTUs had none to minimal effect on the predictive performance.
-
-- decision tree has no rgularization.
 
@@
 
-- Testing with each model will take the same amount of time but if you are training a model and optimize hyper-parameters, then you need to consider the computational burden.
-The computational efficiency of different models have been investigated before.
-Explain for each model.
-In tree based models, depth of tree, number of features, ..
-matter.
-L1 linear is very fast.
-
-- But we can reduce time paralellizing different components of the pipeline.
-Here I have parallelized each datasplit but we can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.
+As the complexity of a ML model increase, its training times will increase (Table S1).
+When regularization and hyperparameter optimization (i.e.
+model selection) steps are added to the model pipeline, training times will increase further.
+The more hyperparameters a model has, the longer it will take to train.
+The computational burden that comes from model complexity, can be reduced by parallelizing different segments of the ML pipeline.
+In this study we have parallelized each datasplit which allowed 100 atasplits to be run at the same time for each model.
+We can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.
 
@@
 
- - decision tree and l1 why so low?
+ - decision tree and l1 why so low?  Decision tree model has no regularization which is why the AUROC values were low.
 

### 2019-06-11 submission/manuscript.Rmd
changes on train time

@@
 ######################################################################
-#-------------------------- Load walltime data -----------------#
+#-------------------------- Load traintime data -----------------#
 ######################################################################
-source("../code/learning/load_walltime.R")
+source("../code/learning/load_traintime.R")
 
 ######################################################################
-#---------------------- Load AUROC performance data -----------------#
+#-------- Load AUROC performance data and Wilcoxon test ------------#
+
 ######################################################################
@@
 The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret.
-Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min).
+Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
 This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources.
@@
 Despite its simplicity, the L2-regularized logistic regression followed random forest in performance.
-In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`).
+In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`).
 We found that median cross-validation and testing AUROC varied `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
@@
 Linear models trained faster than non-linear models.
-L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(walltime_df[walltime_index[1],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[1],3], 2))`), `r paste(round(walltime_df[walltime_index[2],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[2],3],2))`), and `r paste(round(walltime_df[walltime_index[3],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[3],3], 2))`), respectively.
-Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[walltime_index[4],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[4],3], 1))`), `r paste(round(walltime_df[walltime_index[5],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[5],3], 1))`), `r paste(round(walltime_df[walltime_index[6],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[6],3], 1))`) and `r paste(round(walltime_df[walltime_index[7],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[7],3], 1))`), respectively [Figure 4].
+L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
+Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 4].
 
@@
 **Figure 5.
-Computational efficiency of seven ML models.** The walltimes for training and testing of each data-split showed the differences in computational efficieny of the seven models.
-The median walltime in hours was the highest for XGBoost and shortest for L2 logistic regression.
+Computational efficiency of seven ML models.** The traintimes for training and testing of each data-split showed the differences in computational efficieny of the seven models.
+The median traintime in hours was the highest for XGBoost and shortest for L2 logistic regression.
 The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.

### 2019-06-12 submission/manuscript.Rmd
figure axis labels bigger

@@
 
-We compared the median cross-validation AUROC and median testing AUROC for each model.
+We compared the median cross-validation AUROC and median testing AUROC of each model to one another.
 This difference should be low to suggest the model is not overfitting despite the large number of features.
@@
 
-We reported the testing AUROC values over 100 random splits of the full dataset [Figure 1].
-The testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
+We reported the testing AUROC values over 100 random splits of the full dataset which showed that depending on the data-split, the testing AUROC value showed great variability [Figure 1].
+The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
 For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.
@@
 We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect on testing AUROC.
-The top 5 OTUs with the largest negative impact on testing AUROC were overl in tree-based models [Figure 4].
+The top 5 OTUs with the largest negative impact on testing AUROC were overlapped in tree-based models [Figure 4].
 Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods.
@@
 The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].
-Other OTUs had none to minimal effect on the predictive performance.
 
@@
 
-As the complexity of a ML model increase, its training times will increase (Table S1).
+As the complexity of a ML model increase, its training times will increase as well (Table S1).
 When regularization and hyperparameter optimization (i.e.
@@
 The more hyperparameters a model has, the longer it will take to train.
-The computational burden that comes from model complexity, can be reduced by parallelizing different segments of the ML pipeline.
-In this study we have parallelized each datasplit which allowed 100 atasplits to be run at the same time for each model.
-We can also parallelize cross-validation hyerp-parameter search for each hyper-parameter setting.
+The computational burden that comes from model complexity can be reduced by parallelizing different segments of the ML pipeline.
+In this study we have parallelized each data-split which allowed 100 data-splits to be run at the same time for each model.
+We can further parallelize the cross-validation step for each hyper-parameter setting.
 

### 2019-06-12 code/learning/Figure2.R

@@ -55,7 +55,7 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
         legend.position=c(1,0),
         #legend.position="bottom",
         legend.title = element_blank(),
-        legend.background = element_rect(size=0.5, linetype="solid", color="black"),
+        legend.background = element_rect(linetype="solid", color="black", size=0.5),
         legend.box.margin=margin(c(12,12,12, 12)),
         legend.text=element_text(size=12),
         #legend.title=element_text(size=22),
@@ -64,11 +64,11 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         text = element_text(size = 12),
-        axis.text.x=element_text(size = 15, colour='black'),
-        axis.text.y=element_text(size = 15, colour='black'),
-        axis.title.y=element_text(size = 20),
-        axis.title.x=element_text(size = 20),
-        panel.border = element_rect(colour = "black", fill=NA, size=1))
+        axis.text.x=element_text(size = 20, colour='black'),
+        axis.text.y=element_text(size = 20, colour='black'),
+        axis.title.y=element_text(size = 24),
+        axis.title.x=element_text(size = 24),
+        panel.border = element_rect(linetype="solid", colour = "black", fill=NA, size=1.5))
 
 ######################################################################
 #-----------------------Save figure as .pdf ------------------------ #

### 2019-06-13 submission/manuscript.Rmd
discussion additions

@@
 Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed modeling methods that call into question the validity of developed ML models.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call into question the validity of developed ML models.
 Furthermore, using black box ML models has hindered the validation of microbial biomarkers.
@@
 The computational burden that comes from model complexity can be reduced by parallelizing different segments of the ML pipeline.
-In this study we have parallelized each data-split which allowed 100 data-splits to be run at the same time for each model.
+In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
 We can further parallelize the cross-validation step for each hyper-parameter setting.
@@
 
-- In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
 We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
@@
 
-- We also reported the results of 100 random dataset splits because there are differences in AUROC values of each of the seven models between different random data-splits.
+In our ML pipeline we performed the 80%-20% random datasplit 100 times.
+The repetition of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop generalizable ML models and to report reliable performance metrics for a ML model.
+Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model.
+In our study, we showed that there are differences in AUROC values between different random data-splits in each of the models we tested.
 Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
-This result shows that you can get "lucky" in terms of prediction performance depending on the random split.
-Therefore, it is crucial to report all the findings over many random data-splits to get reliable performance metrics.
 
- - Our results also showed that we can choose different models for different reasons.
-Getting feature importances as weights and regression coefficients is the fastest and easiest interpratation.
-Similarly, linear models require less computational burden, fastest to train.
-Thinking about the applicability, and actionability of a model in clinical setting, we might want faster results.
+Our results also showed that we should choose to use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden.
+In terms of predictive performance, all models except a decision tree and L1-regularized SVM with linear kernel behaved similarly.
+Random forest model had testing AUROC values statistically significantly larger than others.
+However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015.
+The decision tree model performed poorly because it had no regularization.
+L1-regularized SVM with linear kernel also performed poorly possibly because it was more sensitive to hyper-parameter tuning.
+In terms of interpretation, random forest is a more complex ML model and it can only be explained using methods such as permutation importance.
+On the other hand, L2-regularized logistic regression is easy to interpret (i.e.
+regression coefficients of the trained model).
+If the goal of the study is to learn the ecology behind a disease and to identify biomarkers, then easily interpretable models could be a better fit.
+It is also important to consider the computational burden of developing ML models.
+Random forest model was slow to train with upto `r` hours whereas L2-regularized logistic regression trained in `r` minutes.
 
- - There is variability if we don't do a full grid search for the hyper-parameters.
-We can give each model 15 minutes to train and see how well they do.
-So an L1 linear model will give you terible results with one and not the other C value as seen in figure S1.
-We have done a granular search first and then limited the possible settings to a range that made the more sense.
-
- - Suprisingly, l2 logistic regression does really well.
-
- - Computational efficiency and interp of logistic vs random forest
-
- - decision tree and l1 why so low?  Decision tree model has no regularization which is why the AUROC values were low.
+Our study underscored the importance of performing a full grid search for hyperparameter settings when building a ML model.
+We first performed a granular grid search and then a full grid search on the limited range of possible hyperparameter settings.
+In the example of L1-regularized SVM with linear kernel, the model showed variability between different regularization coefficients (C).
 

### 2019-06-17 submission/manuscript.Rmd
make a supplemental dotplot for permutation

@@
 To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
-To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
+To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
 The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret.
 Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
-This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources.
+This study showed that we should choose ML models based on our expectations of predictive performance and interpretability as well as our  computational resources.
 It also established standards for modeling pipelines of microbiome-associated ML models.
@@
 Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies.
-However, the results so far are overoptimistic both in terms of prediction performance and validity of developed ML models.
+However, the developed ML so far are overoptimistic both in terms of predictive performance and validity.
 Without rigorous ML pipelines, we cannot trust ML models.
@@
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019].
-More notably, ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance and reporting few or only the best outcomes of different randomizations of cross-validation are commonly used.
+More notably, we commonly see ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation.
 When there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
@@
 We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
-Our ML pipeline utilized held-out test data to evaluate generalization performance of each ML model.
+Our ML pipeline utilized held-out test data to evaluate predictive performance and generalizability of each ML model.
+We used the area under the receiver operating characteristic curve (AUROC) as the predictive performance metric.
 The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`.
 Random forest had the highest median AUROC for detecting SRN.
-Despite its simplicity, the L2-regularized logistic regression followed random forest in performance.
+Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.
 In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`).
-We found that median cross-validation and testing AUROC varied `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation.
-Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
+We also found that depending on how the data is split to create a held-out test set, the AUROC values a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs.
+Aside from evaluating generalization and predictive performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
 
@@
 
-We established a robust and reliable ML pipeline where we train and validate each of the seven models [Figure 1].
+We established a robust ML pipeline where we train and validate each of the seven models [Figure 1].
 We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
@@
 
-__Discriminative performance of the seven models.__
+__Discriminative performance and generalizability of the seven models.__
 
-We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric [Figure 1].
-Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`).
+We evaluated the prediction performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 1].
+Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01).
 The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
@@
 
-__Generalizability of the seven models.__
-
-We compared the median cross-validation AUROC and median testing AUROC of each model to one another.
-This difference should be low to suggest the model is not overfitting despite the large number of features.
-The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
+For each model, we compared the median cross-validation AUROC to the median testing AUROC.
+The difference between the two should be low to suggest the model is not overfitting despite the large number of features.
+The largest difference between the two was in L1 SVM with linear kernel, `r paste(round(abs(difference_model[difference_index[7],1]), 3))`,  followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
 
@@
 We calculated the median ranks of these features over the 100 data-splits.
-In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609) and genera *Gamella*  (OTU00426) [Figure 3].
+In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3].
 Some of the OTUs with the highest ranks were shared among the linear models.
-We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect on testing AUROC.
-The top 5 OTUs with the largest negative impact on testing AUROC were overlapped in tree-based models [Figure 4].
+We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC.
+The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
 Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods.
@@
 The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].
+Permuting other OTUs did not have an effect on the decision tree.
 
@@
 ## Discussion
-
-#### Interpretation of results
 

### 2019-06-17 code/learning/Figure3a_3b.R

@@ -1,352 +1,0 @@
-# Author: Begum Topcuoglu
-# Date: 2018-02-13
-#
-######################################################################
-# This script looks at the model interpretation
-######################################################################
-
-
-######################################################################
-#----------------- Read in necessary libraries -------------------#
-######################################################################
-deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
-for (dep in deps){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
-  }
-  library(dep, verbose=FALSE, character.only=TRUE)
-}
-######################################################################
-#----------------- Define the functions we will use -----------------#
-######################################################################
-
-
-# -------------------- Read files ------------------------------------>
-# This function:
-#     1. takes a list of files(with their path)
-#     2. reads them as delim files with comma seperator
-#     3. returns the dataframe
-read_files <- function(filenames){
-  for(file in filenames){
-    # Read the files generated in main.R 
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',')
-  }
-  return(data)
-}
-# -------------------------------------------------------------------->
-
-
-# ------------------- Re-organize feature importance  ----------------->
-# This function:
-#     1. Takes in a dataframe (different data for each model) and the model name
-#     2. If the models are linear, returns the mean and sd weights of highest weight 10 features
-#     3. If the models are not linear, returns the permutation importance results for:
-#         - Correlated and non-correlated OTUs:
-#         - Top 10 features or feature groups will be listed
-#         - Mean percent AUROC change from original AUROC after permutation
-get_interp_info <- function(data, model_name){ 
-  if(model_name=="L2_Logistic_Regression" || 
-     model_name=="L1_Linear_SVM" || 
-     model_name=="L2_Linear_SVM"){
-    # If the models are linear, we saved the weights of every OTU for each datasplit
-    # 1. Get dataframe transformed into long form
-    #         The OTU names are in 1 column(repeated for 100 datasplits)
-    #         The weight value are in 1 column(for each of the datasplits)
[…]

### 2019-06-17 code/learning/Figure4.R

@@ -212,17 +212,6 @@ base_nonlin_plot <-  function(data, name, list){
 }
 # ----------------------------------------------------------------------->
 
-# --------------------- Linear models ----------------------------------->
-logit_plot <- base_nonlin_plot(logit, "L2_Logistic_Regression", list2) +
-  scale_x_discrete(name = "L2 Logistic Regression ") 
-
-l1_plot <- base_nonlin_plot(l1svm, "L1_Linear_SVM", list2) +
-  scale_x_discrete(name = "L1 Linear SVM") 
-
-l2_plot <- base_nonlin_plot(l2svm, "L2_Linear_SVM", list2) +
-  scale_x_discrete(name = "L2 Linear SVM ") 
-
-# ----------------------------------------------------------------------->
 
 # ----------------- SVM with radial basis function------------------------>
 # Plot most important 5 features effect on AUROC
@@ -252,11 +241,6 @@ xgboost_plot <- base_nonlin_plot(xgboost, "XGBoost", list1)+
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 #combine with cowplot
-perm_non_tree_based <- plot_grid(logit_plot, l1_plot, l2_plot, rbf_plot, labels = c("A", "B", "C", "D"))
-ggdraw(add_sub(perm_non_tree_based, "AUROC with the OTU permuted randomly", vpadding=grid::unit(0,"lines"), y=5, x=0.5, vjust=4.5))
-
-ggsave("Figure_S4.png", plot = last_plot(), device = 'png', path = 'submission', width = 8, height = 6)
-
 perm_tree_based <- plot_grid(rbf_plot, dt_plot, rf_plot, xgboost_plot, labels = c("A", "B", "C", "D"), cols=2)
 ggdraw(add_sub(perm_tree_based, "AUROC with the OTU permuted randomly", size=20, vpadding=grid::unit(0,"lines"), y=5, x=0.5, vjust=4.75))
 

### 2019-06-17 code/learning/FigureS3.R

@@ -2,7 +2,7 @@
 # Date: 2018-06-05
 #
 ######################################################################
-# This script plots permutation importances as ranks and fingerprints over 7 models
+# This script plots permutation importances of 7 models
 ######################################################################
 
 
@@ -16,6 +16,7 @@ for (dep in deps){
   }
   library(dep, verbose=FALSE, character.only=TRUE)
 }
+library(ggrepel)
 ######################################################################
 #----------------- Call the functions we will use -----------------#
 ######################################################################
@@ -27,31 +28,89 @@ source("code/learning/functions.R")
 ######################################################################
 
 grab_importance <-  function(name){
+  
   # Grab the names of the OTUs that have the lowest median AUC when they are permuted
-  data_first_10 <- read.delim(paste0("data/process/", name, "_non_cor_importance.tsv"), header=T, sep='\t') %>%
+  data_all <- read.delim(paste0("data/process/", name, "_non_cor_importance.tsv"), header=T, sep='\t') %>%
     arrange(imp) %>% 
-    head(5) %>%
-    mutate(rank = 1:nrow(.)) %>% 
-    mutate(model=name)
-    
-  return(data_first_10)
+    head(50) %>% 
+    mutate(model=name) %>% 
+    mutate(new_auc = imp) %>%
+    select(-sd_imp, -imp) 
+  return(data_all)
 }
 
-x <- list("Decision_Tree", "Random_Forest", "XGBoost")
+grab_base_auc <-  function(name){
+    # Grab the base test auc values for 100 datasplits
+  data_base <- read.delim(paste0("data/process/combined_best_hp_results_", name, ".csv"), header=T, sep=',') %>%
+    select(-cv_aucs) %>% 
+    group_by(model) %>% 
+    summarise(median_base=median(test_aucs))
+  return(data_base)
+}
 
-total <- map_df(x, grab_importance) %>% group_by(model)
+get_delta_auc <- function(name){
+  new_total <- total %>% 
+    filter(model==name) %>% 
+    mutate(delta= (base_aucs %>% filter(model==name))$median_base - (total %>% filter(model == name))$new_auc)
+  return(new_total)
+}
+  
 
-# Grouped
-ggplot(total, aes(y=rank, x=names, fill=model)) +
-  geom_bar(stat="identity", width=0.35, position = position_dodge2(preserve = "single")) +
[…]

### 2019-06-17 code/learning/FigureS4.R

@@ -78,8 +78,6 @@ dt <- read_files(best_files[1])
 xgboost <- read_files(best_files[7])
 
 
-
-
 ######################################################################
 #-------------- Plot the permutation importance of all models ----------#
 ######################################################################
@@ -162,7 +160,8 @@ base_nonlin_plot <-  function(data, name){
 
 # --------------------- Linear models ----------------------------------->
 logit_plot <- base_nonlin_plot(logit, "L2_Logistic_Regression") +
-  scale_x_discrete(name = "L2 Logistic Regression ") 
+  scale_x_discrete(name = "L2 Logistic 
+Regression ") 
 
 l1_plot <- base_nonlin_plot(l1svm, "L1_Linear_SVM") +
   scale_x_discrete(name = "L1 Linear SVM") 
@@ -172,42 +171,15 @@ l2_plot <- base_nonlin_plot(l2svm, "L2_Linear_SVM") +
 
 # ----------------------------------------------------------------------->
 
-# ----------------- SVM with radial basis function------------------------>
-# Plot most important 5 features effect on AUROC
-rbf_plot <- base_nonlin_plot(rbf, "RBF_SVM") +
-  scale_x_discrete(name = "RBF SVM ") 
-# ----------------------------------------------------------------------->
-
-# --------------------------- Decision Tree ----------------------------->
-# Plot most important 5 features effect on AUROC
-dt_plot <- base_nonlin_plot(dt, "Decision_Tree") +
-  scale_x_discrete(name = "Decision tree ") 
-# ----------------------------------------------------------------------->
-
-# --------------------------- Random Forest ----------------------------->
-# Plot most important 5 features effect on AUROC
-rf_plot <- base_nonlin_plot(rf, "Random_Forest") +
-  scale_x_discrete(name = "Random forest ") 
-# ----------------------------------------------------------------------->
-
-# --------------------------- XGBoost ----------------------------->
-# Plot most important 5 features effect on AUROC
-xgboost_plot <- base_nonlin_plot(xgboost, "XGBoost")+
-  scale_x_discrete(name = "XGBoost ") 
-# ----------------------------------------------------------------------->
-
 ######################################################################
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
[…]

### 2019-06-17 code/learning/Figure4.R
color changes

@@ -163,12 +163,12 @@ base_nonlin_plot <-  function(data, name, list){
   median <-  median(data_base$new_auc) %>% 
     data.frame()
   
-  cols <- c("same1" = "orange", "same2" = "lightblue", "same3" = "darkseagreen3", "same4"="red", "same5" = "lightsalmon", "same6" = "hotpink" , "diff"="white")
+  cols <- c("same1" = "darkgoldenrod1", "same2" = "darkgoldenrod1", "same3" = "darkgoldenrod1", "same4"="darkgoldenrod1", "same5" = "darkgoldenrod1", "same6" = "darkgoldenrod1" , "diff"="darkgoldenrod1")
 
   plot <- ggplot() +
-    geom_boxplot(data=data_full, aes(x=names, y=new_auc, fill=factor(common_otus)), alpha=0.8) +
+    geom_boxplot(data=data_full, aes(x=names, y=new_auc, fill=factor(common_otus)), alpha=0.7) +
     geom_rect(aes(ymin=lowerq, ymax=upperq, xmin=0, xmax=Inf), fill="grey") +
-    geom_boxplot(data=data_full, aes(x=names, y=new_auc, fill=factor(common_otus)), alpha=0.8) +
+    geom_boxplot(data=data_full, aes(x=names, y=new_auc, fill=factor(common_otus)), alpha=0.7) +
     scale_fill_manual(values=cols) +
     geom_hline(yintercept = data_base_means$imp , linetype="dashed") +
     #geom_hline(yintercept = upperq, alpha=0.5) +

### 2019-06-17 submission/manuscript.Rmd
edit some

@@
 The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
-Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods.
+Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods [Figure S3].
 Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
-The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].
-Permuting other OTUs did not have an effect on the decision tree.
+Decision tree was more susceptible to The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].
 

### 2019-06-26 submission/TableS1.Rmd
update

@@
 \captionof{table}{Model complexity: $n$ is the number of training samples, $f$ the number of features, $n_{trees}$ is the number of trees.}
-\begin{tabular}{|l|l|}
+\begin{tabular}{|l|l|l|}
 \hline
-\textbf{Machine learning model} & \textbf{Training complexity} \\ \hline
-Logistic regression & $O(nf)$ \\ \hline
-SVM with linear kernel & $O(nf)$  \\ \hline
-SVM with radial basis kernel & $O(n^2f)$  \\ \hline
-Decision tree & $O(n^2f)$  \\ \hline
-Random forest & $O(n^2fn_{trees})$  \\ \hline
+\textbf{Machine learning model} & \textbf{Training complexity} & \textbf{Hyperparameters} \\ \hline
+Logistic regression & $O(nf)$ & $C$ \\ \hline
+SVM with linear kernel & $O(nf)$ & $C$ \\ \hline
+SVM with radial basis kernel & $O(n^2f)$ & $C$, $\sigma$ \\ \hline
+Decision tree & $O(n^2f)$ & tree depth \\ \hline
+Random forest & $O(n^2fn_{trees})$ & number of features \\ \hline
 XGBoost & $O(nfn_{trees})?$  \\ \hline

### 2019-06-26 submission/manuscript.Rmd
update

@@
 ```
+
 \vspace{35mm}
@@
 Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call into question the validity of developed ML models.
-Furthermore, using black box ML models has hindered the validation of microbial biomarkers.
+Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers.
 To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
@@
 Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
-This study showed that we should choose ML models based on our expectations of predictive performance and interpretability as well as our  computational resources.
+This study showed that we should choose ML models based on our expectations of predictive performance, interpretability and our computational resources.
 It also established standards for modeling pipelines of microbiome-associated ML models.
@@
 Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies.
-However, the developed ML so far are overoptimistic both in terms of predictive performance and validity.
+However, the developed ML so far are overoptimistic in terms of predictive performance and validity.
 Without rigorous ML pipelines, we cannot trust ML models.
@@
 More notably, we commonly see ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation.
-When there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+Even when there are seperate testing sets to evaluate model performance, we observe large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
 Moreover, there is a lack of discussion on why a particular ML model is utilized.
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability.
-We need to strive toward better machine learning practices by (1) implementing consistent and rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
+Black box machine learning models require posthoc explanations to determine the feature importances in making a prediction.
+These explanations can be misleading and at times unreliable when making high-stake decisions about someone's health [@rudin_please_2018].
+The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliyu.
+We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
 
-To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline.
+To showcase a rigorous ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients.
-CRC is the third most common form of cancer worldwide in which the human-associated human microbiome is hypothesized to directly contribute to its development.
-We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN).
+CRC is a type of cancer which the human-associated human microbiome is hypothesized to directly contribute to its development therefore fecal 16S rRNA gene sequences can be used to detect CRC.
+We built ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
 The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
@@
 In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`).
-We also found that depending on how the data is split to create a held-out test set, the AUROC values a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs.
-Aside from evaluating generalization and predictive performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.
+We also found that depending on how the data is split to create a held-out test set, the AUROC values of a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs.
+This study established standards for microbiome-associated ML models and underscored the importance of model selection.
 
@@
 
-We established a robust ML pipeline where we train and validate each of the seven models [Figure 1].
+We established a ML pipeline where we train and validate each of the seven models [Figure 1].
 We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
@@
 We then used the held-out test set to evaluate the prediction performance of each ML model.
-The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance [Figure 1].
+The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].
 
@@
 
-We reported the testing AUROC values over 100 random splits of the full dataset which showed that depending on the data-split, the testing AUROC value showed great variability [Figure 1].
+We reported the testing AUROC values over 100 randomizations of the initial data-split.
+The results showed that depending on the data-split, the testing AUROC value showed great variability [Figure 2].
 The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
@@
 
+As the complexity of a ML model increased [Table S1], its training times increased as well [Figure 5].
+Additionally, the more hyperparameters a model had [Table S1], the longer it took to train [Figure 5].
 Linear models trained faster than non-linear models.
 L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
-Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 4].
-
-As the complexity of a ML model increase, its training times will increase as well (Table S1).
-When regularization and hyperparameter optimization (i.e.
-model selection) steps are added to the model pipeline, training times will increase further.
-The more hyperparameters a model has, the longer it will take to train.
-The computational burden that comes from model complexity can be reduced by parallelizing different segments of the ML pipeline.
-In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
-We can further parallelize the cross-validation step for each hyper-parameter setting.
+Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
 
@@
 
+Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).
+Within the caret package, we have made modifications to L2-regularized SVM with linear kernel function __svmLinear3__ and developed a L1-regularized SVM with linear kernel function __svmLinear4__ to calculate decision values instead of predicted probabilities.
+These changes are available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.
+
 For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
@@
 For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
-For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
+For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.For hyperparameter selection, we started with a granular grid search.
+Then we narrowed and fine-tuned the range of each hyperparameter.
+The range of the grid depends on the ML task and ML model.
+A full grid search needs to be performed to avoid variability in testing performance.
+We can use hyper-band to help us with our hyperparameter selection [@li_hyperband:_2016].
 
-Hyperparameter selection is very important.
-We start with a granular grid search by then narrow and refine.
-Range of the grid depends on the problem.
-We can use hyper-band to help us with our hyperparameter selection.
-If not a full grid search is done, there could be a lot of variability in testing performance.
-Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).
+The computational burden during model training due to model complexity was reduced by parallelizing segments of the ML pipeline.
+In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
+We can further parallelize the cross-validation step for each hyper-parameter setting.
 

### 2019-06-26 submission/manuscript.tex
update

@@
 \usepackage{makecell}
+\usepackage{caption}
 
@@
 
-Machine learning (ML) modeling of the human microbiome is used to
-identify the microbial biomarkers and aid in diagnosis of many chronic
-diseases such as inflammatory bowel disease, diabetes and colorectal
-cancer. Progress has been made towards developing ML models that predict
-health outcomes from bacterial abundances, but reliable ML models are
-scarce in part due to the flawed modeling methods. Furthermore, using
+Machine learning (ML) modeling of the human microbiome has the potential
+to identify the microbial biomarkers and aid in diagnosis of many
+chronic diseases such as inflammatory bowel disease, diabetes and
+colorectal cancer. Progress has been made towards developing ML models
+that predict health outcomes from bacterial abundances, but rigourous ML
+models are scarce due to the flawed modeling methods that call into
+question the validity of developed ML models. Furthermore, the use of
 black box ML models has hindered the validation of microbial biomarkers.
@@
 colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
-cases). To show the effect of model selection, we assessed the accuracy,
-interpretability, and computational efficiency of the following models:
-L2-regularized logistic regression, L1 and L2 support vector machines
-(SVM) with linear and radial basis function kernels, a decision tree,
-random forest, and extreme gradient boosting (XGBoost). The random
-forest model was best at detecting CRC lesions with an AUROC of 0.695
-but it was slow to train (101.3 h) and hard to interpret. Despite its
-simplicity, L2-regularized logistic regression followed random forest in
-predictive performance with an AUROC of 0.680, and it trained the
-fastest (12 min). This study showed that we should choose ML models
-based on our expectations of performance and interpretability as well as
-our computational resources. It also established standards for modeling
-pipelines of microbiome-associated ML models.
+cases). To show the effect of model selection, we assessed the
+predictive performance, interpretability, and computational efficiency
+of the following models: L2-regularized logistic regression, L1 and L2
+support vector machines (SVM) with linear and radial basis function
+kernels, a decision tree, random forest, and extreme gradient boosting
+(XGBoost). The random forest model was best at detecting CRC lesions
+with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to
+interpret. Despite its simplicity, L2-regularized logistic regression
+followed random forest in predictive performance with an AUROC of 0.680,
+and it trained much faster (12 min). This study showed that we should
+choose ML models based on our expectations of predictive performance,
+interpretability and our computational resources. It also established
+standards for modeling pipelines of microbiome-associated ML models.
 
 \subsection{Importance}\label{importance}
+
+Prediction of health outcomes using ML is rapidly being adopted by
+human-associated microbiome studies. However, the developed ML so far
+are overoptimistic in terms of predictive performance and validity.
+Without rigorous ML pipelines, we cannot trust ML models. Before we can
+speed up progress, we need to slow down, define and start using good ML
+practices.
 
@@
 lacks clarity and consistency on which methods are used and how these
-methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably,
-flawed ML practices are prevalent such as using ML pipelines where there
-is no seperate held-out test dataset to evaluate model performance,
-reporting few or only the best outcomes of different randomizations of
-cross-validation and showing a disregard for large differences between
-cross-validation and testing performances as well as large confidence
-intervals of testing performances (4, 13--19). Moreover, there is a lack
-of discussion on why a particular ML model is utilized. Recently, there
-is a trend towards using more complex ML models such as random forest,
-extreme gradient boosting and neural networks without a discussion on if
-and how much model interpretibility is necessary for the study (11,
-20--22).
+methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
+commonly see ML practices such as using ML pipelines where there is no
+seperate held-out test dataset to evaluate model performance or
+reporting few or only the best outcomes of cross-validation. Even when
+there are seperate testing sets to evaluate model performance, we
+observe large differences between cross-validation and testing
+performances that indicate overfitting as well as large confidence
+intervals for testing performances (4, 13--19). Moreover, there is a
+lack of discussion on why a particular ML model is utilized. Recently,
+there is a trend towards using more complex ML models such as random
+forest, extreme gradient boosting and neural networks without a
+discussion on if and how much model interpretibility is necessary for
+the study (11, 20--22). Black box machine learning models require
+posthoc explanations to determine the importances of features in making
+a prediction. These explanations can be misleading and at times
+unreliable when making high-stake decisions about someone's health (23).
+The lack of transparency on modeling methodology, model interpretation
+and model selection negatively impact model reproducibility and
+validity. We need to strive toward better machine learning practices by
+(1) implementing consistent and rigourous machine learning pipelines and
+(2) selecting ML models that reflect the goal of the study as it will
+inform our expectations of model accuracy, complexity, interpretibility
+and computational efficiency.
 
-The lack of transparency on modeling methodology and model selection
-negatively impact model reproducibility and reliability. We need to
-strive toward better machine learning practices by (1) implementing
-consistent and reliable machine learning pipelines and (2) selecting ML
-models that reflect the goal of the study as it will inform our
-expectations of model accuracy, complexity, interpretibility and
-computational efficiency.
-
-To showcase a reliable ML pipeline and to shed light on how much ML
+To showcase a rigorous ML pipeline and to shed light on how much ML
 model selection can affect modeling results, we performed an empirical
 analysis comparing several different ML models using the same dataset
-and with a robust ML pipeline. We used a previously published colorectal
+and the same ML pipeline. We used a previously published colorectal
 cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
-patients. CRC is the third most common form of cancer worldwide in which
-the human-associated human microbiome is hypothesized to directly
-contribute to its development. We built ML models using fecal 16S rRNA
-gene sequences to predict patients with normal colons or patients with
-colonic tumors which are called screen relevant neoplasias (SRN). The
-study had 261 normal and 229 SRN samples. We established modeling
-pipelines for L2-regularized logistic regression, L1 and L2 support
-vector machines (SVM) with linear and radial basis function kernels, a
-decision tree, random forest and XGBoost. Our ML pipeline utilized
-held-out test data to evaluate generalization performance of each ML
-model. The median test AUROC varied from 0.601 to 0.695. Random forest
-had the highest median AUROC for detecting SRN. Despite its simplicity,
-the L2-regularized logistic regression followed random forest in
-performance. In terms of computational efficiency, L2 logistic
-regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
-took the longest (162.843 hours, std ± 3.986). We found that median
-cross-validation and testing AUROC varied 0.021, which highlights the
-importance of a seperate held-out test set and consistent preprocessing
-of the data prior to evaluation. Aside from evaluating generalization
-and classification performances for each of these models, this study
-established standards for modeling pipelines of microbiome-associated
-machine learning models.
+patients. CRC is a type of cancer which the human-associated human
+microbiome is hypothesized to directly contribute to its development
+therefore fecal 16S rRNA gene sequences can be used to detect CRC. We
+built ML models using fecal 16S rRNA gene sequences to predict healthy
+patients versus patients with colorectal lesions that were identified by
+colonoscopy as screen relevant neoplasias (SRN). The study had 261
+normal and 229 SRN samples. We established modeling pipelines for
+L2-regularized logistic regression, L1 and L2 support vector machines
+(SVM) with linear and radial basis function kernels, a decision tree,
+random forest and XGBoost. Our ML pipeline utilized held-out test data
+to evaluate predictive performance and generalizability of each ML
+model. We used the area under the receiver operating characteristic
+curve (AUROC) as the predictive performance metric. The median test
+AUROC varied from 0.601 to 0.695. Random forest had the highest median
+AUROC for detecting SRN. Despite its simplicity, the L2-regularized
+logistic regression was second best in predictive performance. In terms
+of computational efficiency, L1 SVM with linear kernel trained the
+fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
+(155.104 hours, std ± 0.959). We also found that depending on how the
+data is split to create a held-out test set, the AUROC values of a ML
+model can vary up to 0.295 which highlighted the importance of
+performing many randomized modeling runs. This study established
+standards for microbiome-associated ML models and underscored the
+importance of model selection.
 
@@
 kernel. L1-regularization on microbiome data lead to a sparser solution
-(i.e., force many coefficients to zero), removing features that could be
-correlated with other important features. Finally, to explore the
+(i.e., force many coefficients to zero). Finally, to explore the
 potential for non-linear relationships among features and the outcome of
@@
 
-We established a robust and reliable ML pipeline where we train and
-validate each of the seven models {[}Figure 1{]}. We randomly split the
-data into training/validation and test sets so that the
-training/validation set consisted of 80\% of the full dataset while the
-test set was composed of the remaining data {[}Figure 1{]}. Since the
-cases are not uniformly represented in the data, the initial data-split
-was stratified to maintain the overall label distribution in both the
-training/validation and test sets. Training/validation set consisted of
-393 patients (209 SRN), while the test set was composed of 97 patients
-(52 SRN). The training/validation data was used for training purposes
-and validation of hyperparameter selection (i.e.~model selection), and
-the test set was used for evaluation purposes. Validation of
-hyperparameter selection was performed using repeated five-fold
-cross-validation on the training/validation set {[}Figure 1{]}. Similar
-to the initial data-split, five-fold cross-validation was also
-stratified to maintain the overall label distribution on the training
-and validation sets. We validated the cross-validation performances of
-each hyperparameter setting over the 100 randomizations and selected the
-best performing hyper-parameter setting in terms of AUROC to train the
-full training/validation dataset {[}Figures S1 and S2{]}. We then used
-the held-out test set to evaluate the prediction performance of each ML
+We established a robust ML pipeline where we train and validate each of
+the seven models {[}Figure 1{]}. We randomly split the data into
+training/validation and test sets so that the training/validation set
+consisted of 80\% of the full dataset while the test set was composed of
+the remaining data {[}Figure 1{]}. Since the cases are not uniformly
+represented in the data, the initial data-split was stratified to
+maintain the overall label distribution in both the training/validation
+and test sets. Training/validation set consisted of 393 patients (209
+SRN), while the test set was composed of 97 patients (52 SRN). The
+training/validation data was used for training purposes and validation
+of hyperparameter selection (i.e.~model selection), and the test set was
+used for evaluation purposes. Validation of hyperparameter selection was
+performed using repeated five-fold cross-validation on the
+training/validation set {[}Figure 1{]}. Similar to the initial
+data-split, five-fold cross-validation was also stratified to maintain
+the overall label distribution on the training and validation sets. We
+validated the cross-validation performances of each hyperparameter
+setting over the 100 randomizations and selected the best performing
+hyper-parameter setting in terms of AUROC to train the full
+training/validation dataset {[}Figures S1 and S2{]}. We then used the
+held-out test set to evaluate the prediction performance of each ML
 model. The data-split, hyperparameter selection, training and testing
@@
 
-\textbf{Discriminative performance of the seven models.}
+\textbf{Discriminative performance and generalizability of the seven
+models.}
 
 We evaluated the prediction performances of seven binary classification
-models when applied to held-out test data using the area under the
-receiver operating characteristic curve (AUROC) as the discriminative
-performance metric {[}Figure 1{]}. Random forest had significantly
-higher test AUROC values than the other models for detecting SRNs when
-AUROC values were compared to the other six by Wilcoxon rank sum test (p
-= 0.0011). The median AUROC of therandom forest model was 0.695 (IQR
-0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
-with linear and radial basis function kernel AUROC values were not
-significantly different from one another. They had median AUROC values
-of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
-0.056) respectively. L1 SVM with linear kernel and decision tree had
-significantly lower AUROC values than the other ML models with median
-AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
-2{]}.
+models when applied to held-out test data using AUROC metric {[}Figure
+1{]}. Random forest had significantly higher test AUROC values than the
+other models for detecting SRNs when AUROC values were compared to the
+other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
+AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
+logistic regression, XGBoost, L2-regularized SVM with linear and radial
+basis function kernel AUROC values were not significantly different from
+one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
+(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
+SVM with linear kernel and decision tree had significantly lower AUROC
+values than the other ML models with median AUROC of 0.65 (IQR 0.066)
+and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.
 
-\textbf{Generalizability of the seven models.}
-
-We compared the median cross-validation AUROC and median testing AUROC
-for each model. This difference should be low to suggest the model is
-not overfitting despite the large number of features. The largest
-difference between the two was 0.021 in L1 SVM with linear kernel
-followed by SVM with radial basis function kernel and decision tree with
-a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
+For each model, we compared the median cross-validation AUROC to the
+median testing AUROC. The difference between the two should be low to
+suggest the model is not overfitting despite the large number of
+features. The largest difference between the two was in L1 SVM with
+linear kernel, 0.021, followed by SVM with radial basis function kernel
+and decision tree with a difference of 0.007 and 0.006, respectively
+{[}Figure 2{]}.
 
 We reported the testing AUROC values over 100 random splits of the full
-dataset {[}Figure 1{]}. The testing AUROC values varied 0.23 on average,
-depending on the data-split. For instance, the lowest AUROC value of the
-random forest model was 0.59 whereas the highest was 0.81.
+dataset which showed that depending on the data-split, the testing AUROC
+value showed great variability {[}Figure 2{]}. The testing AUROC values
+within each model varied 0.23 on average across the seven models. For
+instance, the lowest AUROC value of the random forest model was 0.59
+whereas the highest was 0.81.
 
@@
 interpretibility as they increase in complexity. We interpreted L1 and
-L2 SVM with linear kernel using the feature weights and L2 logistic
-regression using regression coefficients of the trained models. We
-examined the 5 predictive (highest weight) and protective (lowest
-weight) features identified by the model. We calculated the mean weights
-and coefficients of these features over the 100 data-splits. In the
-three linear models, OTUs that had the largest mean weights and drove
-the detection of SRNs belong to family \emph{Lachnospiraceae}, and
-\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
-OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629)
-and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}. We explained the
-feature importances in non-linear models using permutation importance on
-the held-out test data where we permuted groups of correlated features
-(see methods). For the tree-based models, permuting
+L2 SVM with linear kernel and L2 logistic regression using the feature
+weights of the trained models. We ranked the absolute weights of all the
+OTUs for each data-split. We calculated the median ranks of these
+features over the 100 data-splits. In the three linear models, OTUs that
+had the largest median ranks and drove the detection of SRNs belong to
+family \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
+OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609),
+genera \emph{Gamella} (OTU00426) and genera \emph{Peptostreptococcus}
+(OTU00367) {[}Figure 3{]}. Some of the OTUs with the highest ranks were
+shared among the linear models. We explained the feature importances in
+non-linear models using permutation importance on the held-out test data
+where we randomly permuted non-correlated features individually and
+groups of correlated features together (see methods) to calculate the
+effect of permuted OTUs or group of OTUs on testing AUROC. The top 5
+OTUs with the largest negative impact on testing AUROC overlapped in
+tree-based models {[}Figure 4{]}. Specifically, permuting
 \emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
-predictive performances the most. Decision tree, random forest and
-XGBoost models' predictive performance dropped from 0.6 base testing
-AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
-respectively {[}Figure 4{]}. The negative predictive impact of
+predictive performances the most for all tree-based methods {[}Figure
+S3{]}. Decision tree, random forest and XGBoost models' predictive
+performance dropped from 0.6 base testing AUROC median to 0.52, from
+0.69 to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.
+Decision tree was more susceptible to The negative predictive impact of
 \emph{Peptostreptococcus} in the decision tree model was followed by a
 \emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
-to 0.58) {[}Figure 4B{]}. Other OTUs had none to minimal effect on the
-predictive performance.
+to 0.58) {[}Figure 4B{]}.
 
@@
 
-Linear models trained faster than non-linear models. L2 logistic
-Rregression and L1 and L2 SVM with linear kernel had training times of
-0.2 hours, (std ± 0.02), 0.2 hours, (std ± 0.03), and 0.2 hours, (std ±
-0.03), respectively. Whereas, SVM with radial basis function kernel,
-decision tree, random forest and xgboost had training times of 9.4
-hours, (std ± 0.8), 64.7 hours, (std ± 9.9), 101.3 hours, (std ± 10) and
-162.8 hours, (std ± 4), respectively {[}Figure 4{]}.
+As the complexity of a ML model increased {[}Table S1{]}, its training
+times increased as well {[}Figure 5{]}. Additionally, the more
+hyperparameters a model had {[}Table S1{]}, the longer it took to train
+{[}Figure 5{]}. Linear models trained faster than non-linear models. L2
+logistic Rregression and L1 and L2 SVM with linear kernel had training
+times of 0.2 hours, (std ± 0.03), 0.2 hours, (std ± 0.02), and 0.2
+hours, (std ± 0.02), respectively. Whereas, SVM with radial basis
+function kernel, decision tree, random forest and xgboost had training
+times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ± 8.8), 83.2 hours,
+(std ± 11.3) and 155.1 hours, (std ± 1), respectively {[}Figure 5{]}.
 
@@
 
-\paragraph{Interpretation of results}\label{interpretation-of-results}
-
-\begin{itemize}
-\item
 In this study we established a robust ML pipeline to use 16S rRNA
@@
 test set to illustrate the difference between cross-validation and
-  testing AUROC values. When the differences between cross-validation
-  and test performance is low, this suggets the models will perform
-  similar with similar data. In all seven models, the differences in
+testing AUROC values. When the differences between cross-validation and
+test performance is low, this suggets the models will perform similar
+with similar data. In all seven models, the differences in
 cross-validation and testing AUROC values did not exceed 0.021 which
 suggests that these models will be able to test similar new data.
-\item
-  We also reported the results of 100 random dataset splits because
-  there are differences in AUROC values of each of the seven models
-  between different random data-splits. Our results showed that the
+
+In our ML pipeline we performed the 80\%-20\% random datasplit 100
+times. The repetition of the initial data-split to create a held-out
+test set is a crucial step in the ML pipeline to develop generalizable
+ML models and to report reliable performance metrics for a ML model.
+Depending on the data-split, there is the chance of being overoptimistic
+about the predictive performance of a model. In our study, we showed
+that there are differences in AUROC values between different random
+data-splits in each of the models we tested. Our results showed that the
 testing AUROC values varied 0.23 on average, depending on the
-  data-split. This result shows that you can get ``lucky'' in terms of
-  prediction performance depending on the random split. Therefore, it is
-  crucial to report all the findings over many random data-splits to get
-  reliable performance metrics.
-\item
-  Our results also showed that we can choose different models for
-  different reasons. Getting feature importances as weights and
-  regression coefficients is the fastest and easiest interpratation.
-  Similarly, linear models require less computational burden, fastest to
-  train. Thinking about the applicability, and actionability of a model
-  in clinical setting, we might want faster results.
-\item
-\item
-  Suprisingly, l2 logistic regression does really well.
-\item
-  Computational efficiency and interp of logistic vs random forest
-\item
-  decision tree and l1 why so low?
-\end{itemize}
+data-split.
+
+Our results also showed that we should choose to use ML models based on
+the goal of the study and our expectations of predictive performance,
+interpretibility and computational burden. In terms of predictive
+performance, all models except a decision tree and L1-regularized SVM
+with linear kernel behaved similarly. Random forest model had testing
+AUROC values statistically significantly larger than others. However the
+second best model was L2-regularized logistic regression following
+random forest with a median AUROC difference of only 0.015. The decision
+tree model performed poorly because it had no regularization.
+L1-regularized SVM with linear kernel also performed poorly possibly
+because it was more sensitive to hyper-parameter tuning. In terms of
+interpretation, random forest is a more complex ML model and it can only
+be explained using methods such as permutation importance. On the other
+hand, L2-regularized logistic regression is easy to interpret
+(i.e.~regression coefficients of the trained model). If the goal of the
+study is to learn the ecology behind a disease and to identify
+biomarkers, then easily interpretable models could be a better fit. It
+is also important to consider the computational burden of developing ML
+models. Random forest model was slow to train with upto \texttt{r} hours
+whereas L2-regularized logistic regression trained in \texttt{r}
+minutes.
+
+Our study underscored the importance of performing a full grid search
+for hyperparameter settings when building a ML model. We first performed
+a granular grid search and then a full grid search on the limited range
+of possible hyperparameter settings. In the example of L1-regularized
+SVM with linear kernel, the model showed variability between different
+regularization coefficients (C).
 
@@
 \item
-  What happens with imbalanced data
+  What happens with imbalanced data The metric you use as the
+  discriminative performance becomes important. Don't use accuracy.
+  AUROC is good and random at 0.5 always.
 \item
-  What happens with smaller datasets
+  What happens with smaller datasets Generalization performance depends
+  on size. The more complex the model, more data you will need. Use
+  linear models if you have little data.
 \item
-  Does the code work on any microbiome data?
+  Does the code work on any microbiome data? So if yu have a new sample,
+  do you have the same OTUs? NO
 \item
@@
 patients recruited by Great Lakes-New England Early Detection Research
-Network study. These data were obtained from Sze et al (23). The stool
+Network study. These data were obtained from Sze et al (24). The stool
 samples were provided by recruited adult participants who were
@@
 assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(24--26).
+(25--27).
 
@@
 
+Models were trained using the machine learning wrapper caret package
+(v.6.0.81) in R (v.3.5.0). Within the caret package, we have made
+modifications to L2-regularized SVM with linear kernel function
+\textbf{svmLinear3} and developed a L1-regularized SVM with linear
+kernel function \textbf{svmLinear4} to calculate decision values instead
+of predicted probabilities. These changes are available at
+\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/}.
+
 For L2-regularized logistic regression, L1 and L2 support vector
@@
 \textbf{number of features} to consider when looking for the best tree
-split. For xgboost, we tuned for \textbf{learning rate} and the
+split. For XGBoost, we tuned for \textbf{learning rate} and the
 \textbf{fraction of samples} to be used for fitting the individual base
-learners. Models were trained using the machine learning wrapper caret
-package (v.6.0.81) in R (v.3.5.0).
+learners.For hyperparameter selection, we started with a granular grid
+search. Then we narrowed and fine-tuned the range of each
+hyperparameter. The range of the grid depends on the ML task and ML
+model. A full grid search needs to be performed to avoid variability in
+testing performance. We can use hyper-band to help us with our
+hyperparameter selection (28).
 
-\textbf{Permutation importance workflow.}
+The computational burden during model training due to model complexity
+was reduced by parallelizing segments of the ML pipeline. In this study
+we have parallelized each data-split which allowed 100 data-splits to be
+processed through the ML pipeline at the same time for each model. We
+can further parallelize the cross-validation step for each
+hyper-parameter setting.
+
+\textbf{Permutation importance workflow.} We created a Spearman's
+rank-order correlation matrix, corrected for multiple pairwise
+comparisons. We then defined correlated OTUs as having perfect
+correlation (correlation coef=1 and p\textless{}0.01).Other OTUs were
+permuted individually to get permutation importance but the correlated
+ones are grouped together and permuted at the same time. The reported
+OTUs that have impact on testing AUROC are reported in Figures xxx. If
+we want we can decrease the correlation coefficient to consider OTUs
+that are correlated with ecological consequences but this was out of the
+scope of this study but will be followed up in further analyses.
 
@@
 \includegraphics{Figure_5.png} \textbf{Figure 5. Computational
-efficiency of seven ML models.} The walltimes for training and testing
+efficiency of seven ML models.} The traintimes for training and testing
 of each data-split showed the differences in computational efficieny of
-the seven models. The median walltime in hours was the highest for
+the seven models. The median traintime in hours was the highest for
 XGBoost and shortest for L2 logistic regression. The boxplot shows
@@
 
+\hypertarget{ref-rudin_please_2018}{}
+23. \textbf{Rudin C}. 2018. Please stop explaining black box models for
+high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
+
 \hypertarget{ref-sze_leveraging_2018}{}
-23. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+24. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+25. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-25. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+26. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-26. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+27. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 
+\hypertarget{ref-li_hyperband:_2016}{}
+28. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
+bandit-based approach to hyperparameter optimization. arXiv:160306560
+{[}cs, stat{]}.
+
 \end{document}

### 2019-06-27 submission/manuscript.Rmd
add

@@
 Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
-Decision tree was more susceptible to The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].
 
@@
 
-As the complexity of a ML model increased [Table S1], its training times increased as well [Figure 5].
-Additionally, the more hyperparameters a model had [Table S1], the longer it took to train [Figure 5].
+As the complexity of a ML model and the number of tuned hyperparameter settings increased [Table S1], its training times increased as well [Figure 5].
 Linear models trained faster than non-linear models.
-L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
-Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
+L1 and L2 SVM with linear kernel and L2 logistic regression had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
+Whereas, a decision tree, SVM with radial basis function kernel, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
 

### 2019-06-27 submission/manuscript.tex
add

@@
 the study (11, 20--22). Black box machine learning models require
-posthoc explanations to determine the importances of features in making
-a prediction. These explanations can be misleading and at times
-unreliable when making high-stake decisions about someone's health (23).
-The lack of transparency on modeling methodology, model interpretation
-and model selection negatively impact model reproducibility and
-validity. We need to strive toward better machine learning practices by
-(1) implementing consistent and rigourous machine learning pipelines and
-(2) selecting ML models that reflect the goal of the study as it will
-inform our expectations of model accuracy, complexity, interpretibility
-and computational efficiency.
+posthoc explanations to determine the feature importances in making a
+prediction. These explanations can be misleading and at times unreliable
+when making high-stake decisions about someone's health (23). The lack
+of transparency on model selection and interpretation as well as flawed
+modeling methods negatively impact model validity and reproducibiliyu.
+We need to strive toward better machine learning practices by (1)
+implementing rigourous machine learning pipelines and (2) selecting ML
+models that reflect the goal of the study as it will inform our
+expectations of model accuracy, complexity, interpretibility and
+computational efficiency.
 
@@
 
-We established a robust ML pipeline where we train and validate each of
-the seven models {[}Figure 1{]}. We randomly split the data into
+We established a ML pipeline where we train and validate each of the
+seven models {[}Figure 1{]}. We randomly split the data into
 training/validation and test sets so that the training/validation set
@@
 steps were repeated 100 times to get a reliable and robust reading of
-model prediction performance {[}Figure 1{]}.
+model performance {[}Figure 1{]}.
 
@@
 
-We reported the testing AUROC values over 100 random splits of the full
-dataset which showed that depending on the data-split, the testing AUROC
-value showed great variability {[}Figure 2{]}. The testing AUROC values
-within each model varied 0.23 on average across the seven models. For
-instance, the lowest AUROC value of the random forest model was 0.59
-whereas the highest was 0.81.
+We reported the testing AUROC values over 100 randomizations of the
+initial data-split. The results showed that depending on the data-split,
+the testing AUROC value showed great variability {[}Figure 2{]}. The
+testing AUROC values within each model varied 0.23 on average across the
+seven models. For instance, the lowest AUROC value of the random forest
+model was 0.59 whereas the highest was 0.81.
 

### 2019-07-03 submission/manuscript.Rmd
update for meeting

@@
 Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call into question the validity of developed ML models.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call the validity of developed ML models into question.
 Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers.
@@
 To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
-The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret.
+The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
 Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
@@
 More notably, we commonly see ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation.
-Even when there are seperate testing sets to evaluate model performance, we observe large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+Even when there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
 Moreover, there is a lack of discussion on why a particular ML model is utilized.
@@
 These explanations can be misleading and at times unreliable when making high-stake decisions about someone's health [@rudin_please_2018].
-The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliyu.
+The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliy.
 We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
@@
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients.
-CRC is a type of cancer which the human-associated human microbiome is hypothesized to directly contribute to its development therefore fecal 16S rRNA gene sequences can be used to detect CRC.
+CRC is a type of cancer which the human-associated human microbiome is hypothesized to directly contribute to its development and fecal 16S rRNA gene sequences have been used to detect CRC.
 We built ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
@@
 We used a cohort of 490 patients with 261 cases of SRN.
-For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colonic health (SRN or normal).
+For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colonic health (having colorectal lesions that are identified as SRN or normal).
 All the cases were independently labeled through colonoscopies.
@@
 Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets.
-We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting in terms of AUROC to train the full training/validation dataset [Figures S1 and S2].
+We validated the cross-validation performances of each hyperparameter setting over 100 randomizations and selected the best performing hyper-parameter setting in terms of AUROC metric and trained the full training/validation dataset [Figures S1 and S2].
 We then used the held-out test set to evaluate the prediction performance of each ML model.
@@
 We calculated the median ranks of these features over the 100 data-splits.
-In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3].
+In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to families *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3].
 Some of the OTUs with the highest ranks were shared among the linear models.
@@
 The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
-Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods [Figure S3].
+Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure S3].
 Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
@@
 
-In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+We set-up standards for developing and evaluating ML models for microbiome data.
 We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
-When the differences between cross-validation and test performance is low, this suggets the models will perform similar with similar data.
+When the differences between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data.
 In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data.
-
-In our ML pipeline we performed the 80%-20% random datasplit 100 times.
+We performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
 The repetition of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop generalizable ML models and to report reliable performance metrics for a ML model.
 Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model.
-In our study, we showed that there are differences in AUROC values between different random data-splits in each of the models we tested.
+In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
 Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
+We used AUROC metric instead of accuracy in our study to evaluate predictive performance of the ML models.
+AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
+We performed a full grid search for hyperparameter settings when building a ML model.
+It is not enough to use the default hyperparameter settings when using previously developed ML packages in programming languages like R, Python and Matlab.
+In the example of L1-regularized SVM with linear kernel (Figure S1), the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
 
-Our results also showed that we should choose to use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden.
-In terms of predictive performance, all models except a decision tree and L1-regularized SVM with linear kernel behaved similarly.
-Random forest model had testing AUROC values statistically significantly larger than others.
+Our results showed that we should choose to use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden.
+In terms of predictive performance, random forest model had testing AUROC values statistically significantly larger than others.
 However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015.
-The decision tree model performed poorly because it had no regularization.
-L1-regularized SVM with linear kernel also performed poorly possibly because it was more sensitive to hyper-parameter tuning.
-In terms of interpretation, random forest is a more complex ML model and it can only be explained using methods such as permutation importance.
-On the other hand, L2-regularized logistic regression is easy to interpret (i.e.
+In terms of interpretation, random forest was a more complex ML model and it could only be explained using methods such as permutation importance.
+On the other hand, L2-regularized logistic regression was easy to interpret (i.e.
 regression coefficients of the trained model).
-If the goal of the study is to learn the ecology behind a disease and to identify biomarkers, then easily interpretable models could be a better fit.
+If the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers of a disease, then easily interpretable models would be better suited.
 It is also important to consider the computational burden of developing ML models.
-Random forest model was slow to train with upto `r` hours whereas L2-regularized logistic regression trained in `r` minutes.
-
-Our study underscored the importance of performing a full grid search for hyperparameter settings when building a ML model.
-We first performed a granular grid search and then a full grid search on the limited range of possible hyperparameter settings.
-In the example of L1-regularized SVM with linear kernel, the model showed variability between different regularization coefficients (C).
-
-#### Consideration of possible weaknesses
- - What happens with imbalanced data
- The metric you use as the discriminative performance becomes important. Don't use accuracy.
- AUROC is good and random at 0.5 always.
-
- - What happens with smaller datasets
- Generalization performance depends on size. 
- The more complex the model, more data you will need.
- Use linear models if you have little data.
-
- - Does the code work on any microbiome data?
- So if yu have a new sample, do you have the same OTUs? NO
-
-  - Why did we use only one dataset?
-
-  - Why not validate on other data instead of held-out testing set? Becuase we are comparing several methods on 1 dataset.
-
-  - What will the future data look like?
-
-#### Relationship of results to previous literature and broader implications of having answered research question
-
-#### Prospects for future progress
-  Subsample dataset and refit see how method behaves for 7 methods. (500->250->50)
+Random forest model was slow to train 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
+The dataset we used for our study had 490 samples.
+Another criteria when choosing a ML model is the generalizability of it.
+The generalization performance of ML models depends on size.
+The more complex the model, the more data it will need.
+Microbiome studies that have small sample sizes should use less complex models such as L2-regularized logistic regression.
 
@@
 **Figure 3.
-Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights.
-The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits.
+Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature ranks.
+The feature ranks of the most important five OTUs in each data-split are shown here.
 Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel.
@@
 **Figure 5.
-Computational efficiency of seven ML models.** The traintimes for training and testing of each data-split showed the differences in computational efficieny of the seven models.
+Computational efficiency of seven ML models.** The training times for training and testing of each data-split showed the differences in computational efficieny of the seven models.
 The median traintime in hours was the highest for XGBoost and shortest for L2 logistic regression.

### 2019-07-03 submission/manuscript.tex
update for meeting

@@
 that predict health outcomes from bacterial abundances, but rigourous ML
-models are scarce due to the flawed modeling methods that call into
-question the validity of developed ML models. Furthermore, the use of
+models are scarce due to the flawed modeling methods that call the
+validity of developed ML models into question. Furthermore, the use of
 black box ML models has hindered the validation of microbial biomarkers.
@@
 (XGBoost). The random forest model was best at detecting CRC lesions
-with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to
+with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to
 interpret. Despite its simplicity, L2-regularized logistic regression
@@
 reporting few or only the best outcomes of cross-validation. Even when
-there are seperate testing sets to evaluate model performance, we
-observe large differences between cross-validation and testing
-performances that indicate overfitting as well as large confidence
-intervals for testing performances (4, 13--19). Moreover, there is a
-lack of discussion on why a particular ML model is utilized. Recently,
-there is a trend towards using more complex ML models such as random
-forest, extreme gradient boosting and neural networks without a
-discussion on if and how much model interpretibility is necessary for
-the study (11, 20--22). Black box machine learning models require
-posthoc explanations to determine the feature importances in making a
-prediction. These explanations can be misleading and at times unreliable
-when making high-stake decisions about someone's health (23). The lack
-of transparency on model selection and interpretation as well as flawed
-modeling methods negatively impact model validity and reproducibiliyu.
-We need to strive toward better machine learning practices by (1)
-implementing rigourous machine learning pipelines and (2) selecting ML
-models that reflect the goal of the study as it will inform our
-expectations of model accuracy, complexity, interpretibility and
-computational efficiency.
+there are seperate testing sets to evaluate model performance, there are
+large differences between cross-validation and testing performances that
+indicate overfitting as well as large confidence intervals for testing
+performances (4, 13--19). Moreover, there is a lack of discussion on why
+a particular ML model is utilized. Recently, there is a trend towards
+using more complex ML models such as random forest, extreme gradient
+boosting and neural networks without a discussion on if and how much
+model interpretibility is necessary for the study (11, 20--22). Black
+box machine learning models require posthoc explanations to determine
+the feature importances in making a prediction. These explanations can
+be misleading and at times unreliable when making high-stake decisions
+about someone's health (23). The lack of transparency on model selection
+and interpretation as well as flawed modeling methods negatively impact
+model validity and reproducibiliy. We need to strive toward better
+machine learning practices by (1) implementing rigourous machine
+learning pipelines and (2) selecting ML models that reflect the goal of
+the study as it will inform our expectations of model accuracy,
+complexity, interpretibility and computational efficiency.
 
@@
 patients. CRC is a type of cancer which the human-associated human
-microbiome is hypothesized to directly contribute to its development
-therefore fecal 16S rRNA gene sequences can be used to detect CRC. We
-built ML models using fecal 16S rRNA gene sequences to predict healthy
-patients versus patients with colorectal lesions that were identified by
+microbiome is hypothesized to directly contribute to its development and
+fecal 16S rRNA gene sequences have been used to detect CRC. We built ML
+models using fecal 16S rRNA gene sequences to predict healthy patients
+versus patients with colorectal lesions that were identified by
 colonoscopy as screen relevant neoplasias (SRN). The study had 261
@@
 patient, we had 6920 features (fecal bacterial abundances) and a
-two-class label that defines their colonic health (SRN or normal). All
-the cases were independently labeled through colonoscopies. We
-established modeling pipelines for a binary prediction task with
-L2-regularized logistic regression, L1 and L2 support vector machines
-(SVM) with linear and radial basis function kernels, a decision tree,
-random forest and extreme gradient boosted decision tree (XGBoost) to
-emphasize the differences in model accuracy, complexity,
-interpretibility and computational efficiency due to model selection.
+two-class label that defines their colonic health (having colorectal
+lesions that are identified as SRN or normal). All the cases were
+independently labeled through colonoscopies. We established modeling
+pipelines for a binary prediction task with L2-regularized logistic
+regression, L1 and L2 support vector machines (SVM) with linear and
+radial basis function kernels, a decision tree, random forest and
+extreme gradient boosted decision tree (XGBoost) to emphasize the
+differences in model accuracy, complexity, interpretibility and
+computational efficiency due to model selection.
 
@@
 validated the cross-validation performances of each hyperparameter
-setting over the 100 randomizations and selected the best performing
-hyper-parameter setting in terms of AUROC to train the full
+setting over 100 randomizations and selected the best performing
+hyper-parameter setting in terms of AUROC metric and trained the full
 training/validation dataset {[}Figures S1 and S2{]}. We then used the
@@
 had the largest median ranks and drove the detection of SRNs belong to
-family \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
+families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
 OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609),
@@
 \emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
-predictive performances the most for all tree-based methods {[}Figure
+predictive performances the most in all tree-based methods {[}Figure
 S3{]}. Decision tree, random forest and XGBoost models' predictive
@@
 0.69 to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.
-Decision tree was more susceptible to The negative predictive impact of
-\emph{Peptostreptococcus} in the decision tree model was followed by a
-\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
-to 0.58) {[}Figure 4B{]}.
 
@@
 
-As the complexity of a ML model increased {[}Table S1{]}, its training
-times increased as well {[}Figure 5{]}. Additionally, the more
-hyperparameters a model had {[}Table S1{]}, the longer it took to train
-{[}Figure 5{]}. Linear models trained faster than non-linear models. L2
-logistic Rregression and L1 and L2 SVM with linear kernel had training
+As the complexity of a ML model and the number of tuned hyperparameter
+settings increased {[}Table S1{]}, its training times increased as well
+{[}Figure 5{]}. Linear models trained faster than non-linear models. L1
+and L2 SVM with linear kernel and L2 logistic regression had training
 times of 0.2 hours, (std ± 0.03), 0.2 hours, (std ± 0.02), and 0.2
-hours, (std ± 0.02), respectively. Whereas, SVM with radial basis
-function kernel, decision tree, random forest and xgboost had training
+hours, (std ± 0.02), respectively. Whereas, a decision tree, SVM with
+radial basis function kernel, random forest and xgboost had training
 times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ± 8.8), 83.2 hours,
@@
 
-In this study we established a robust ML pipeline to use 16S rRNA
-sequence counts to predict a binary health outcome. We used a held-out
-test set to illustrate the difference between cross-validation and
-testing AUROC values. When the differences between cross-validation and
-test performance is low, this suggets the models will perform similar
-with similar data. In all seven models, the differences in
-cross-validation and testing AUROC values did not exceed 0.021 which
-suggests that these models will be able to test similar new data.
+In this study we established a rigorous ML pipeline to use 16S rRNA
+sequence counts to predict a binary health outcome. We set-up standards
+for developing and evaluating ML models for microbiome data. We used a
+held-out test set to illustrate the difference between cross-validation
+and testing AUROC values. When the differences between cross-validation
+and test performance is low, this suggets the models are not overfit and
+that they will perform similar with similar data. In all seven models,
+the differences in cross-validation and testing AUROC values did not
+exceed 0.021 which suggests that these models will be able to test
+similar new data. We performed the initial 80\%-20\% random datasplit
+100 times in our ML pipeline. The repetition of the initial data-split
+to create a held-out test set is a crucial step in the ML pipeline to
+develop generalizable ML models and to report reliable performance
+metrics for a ML model. Depending on the data-split, there is the chance
+of being overoptimistic about the predictive performance of a model. In
+our study, we showed that there was variability in AUROC values between
+different random data-splits in each of the models we tested. Our
+results showed that the testing AUROC values varied 0.23 on average,
+depending on the data-split. We used AUROC metric instead of accuracy in
+our study to evaluate predictive performance of the ML models. AUROC is
+always random at the value 0.5 and is a robust metric when a dataset is
+imbalanced. We performed a full grid search for hyperparameter settings
+when building a ML model. It is not enough to use the default
+hyperparameter settings when using previously developed ML packages in
+programming languages like R, Python and Matlab. In the example of
+L1-regularized SVM with linear kernel (Figure S1), the model showed
+large variability between different regularization coefficients (C) and
+was susceptible to performing poorly if the wrong regularization
+coefficient was assigned to the model by default.
 
-In our ML pipeline we performed the 80\%-20\% random datasplit 100
-times. The repetition of the initial data-split to create a held-out
-test set is a crucial step in the ML pipeline to develop generalizable
-ML models and to report reliable performance metrics for a ML model.
-Depending on the data-split, there is the chance of being overoptimistic
-about the predictive performance of a model. In our study, we showed
-that there are differences in AUROC values between different random
-data-splits in each of the models we tested. Our results showed that the
-testing AUROC values varied 0.23 on average, depending on the
-data-split.
-
-Our results also showed that we should choose to use ML models based on
-the goal of the study and our expectations of predictive performance,
+Our results showed that we should choose to use ML models based on the
+goal of the study and our expectations of predictive performance,
 interpretibility and computational burden. In terms of predictive
-performance, all models except a decision tree and L1-regularized SVM
-with linear kernel behaved similarly. Random forest model had testing
-AUROC values statistically significantly larger than others. However the
-second best model was L2-regularized logistic regression following
-random forest with a median AUROC difference of only 0.015. The decision
-tree model performed poorly because it had no regularization.
-L1-regularized SVM with linear kernel also performed poorly possibly
-because it was more sensitive to hyper-parameter tuning. In terms of
-interpretation, random forest is a more complex ML model and it can only
-be explained using methods such as permutation importance. On the other
-hand, L2-regularized logistic regression is easy to interpret
-(i.e.~regression coefficients of the trained model). If the goal of the
-study is to learn the ecology behind a disease and to identify
-biomarkers, then easily interpretable models could be a better fit. It
-is also important to consider the computational burden of developing ML
-models. Random forest model was slow to train with upto \texttt{r} hours
-whereas L2-regularized logistic regression trained in \texttt{r}
-minutes.
-
-Our study underscored the importance of performing a full grid search
-for hyperparameter settings when building a ML model. We first performed
-a granular grid search and then a full grid search on the limited range
-of possible hyperparameter settings. In the example of L1-regularized
-SVM with linear kernel, the model showed variability between different
-regularization coefficients (C).
-
-\paragraph{Consideration of possible
-weaknesses}\label{consideration-of-possible-weaknesses}
-
-\begin{itemize}
-\item
-  What happens with imbalanced data The metric you use as the
-  discriminative performance becomes important. Don't use accuracy.
-  AUROC is good and random at 0.5 always.
-\item
-  What happens with smaller datasets Generalization performance depends
-  on size. The more complex the model, more data you will need. Use
-  linear models if you have little data.
-\item
-  Does the code work on any microbiome data? So if yu have a new sample,
-  do you have the same OTUs? NO
-\item
-  Why did we use only one dataset?
-\item
-  Why not validate on other data instead of held-out testing set?
-  Becuase we are comparing several methods on 1 dataset.
-\item
-  What will the future data look like?
-\end{itemize}
-
-\paragraph{Relationship of results to previous literature and broader
-implications of having answered research
-question}\label{relationship-of-results-to-previous-literature-and-broader-implications-of-having-answered-research-question}
-
-\paragraph{Prospects for future
-progress}\label{prospects-for-future-progress}
-
-Subsample dataset and refit see how method behaves for 7 methods.
-(500-\textgreater{}250-\textgreater{}50)
+performance, random forest model had testing AUROC values statistically
+significantly larger than others. However the second best model was
+L2-regularized logistic regression following random forest with a median
+AUROC difference of only 0.015. In terms of interpretation, random
+forest was a more complex ML model and it could only be explained using
+methods such as permutation importance. On the other hand,
+L2-regularized logistic regression was easy to interpret
+(i.e.~regression coefficients of the trained model). If the goal of a
+study is to learn the ecology behind a disease and to identify microbial
+biomarkers of a disease, then easily interpretable models would be
+better suited. It is also important to consider the computational burden
+of developing ML models. Random forest model was slow to train 83.2
+hours whereas L2-regularized logistic regression trained in 12 minutes.
+The dataset we used for our study had 490 samples. Another criteria when
+choosing a ML model is the generalizability of it. The generalization
+performance of ML models depends on size. The more complex the model,
+the more data it will need. Microbiome studies that have small sample
+sizes should use less complex models such as L2-regularized logistic
+regression.
 
@@
 with linear kernel feature weights (C) L2 SVM with linear kernel feature
-weights. The means weights and coefficients of the most important five
-OTUs for each model are shown here with the standard deviation over 100
-data-splits. Similar OTUs had the largest impact on the predictive
-performance of L2 logistic regression and L2 SVM with linear kernel.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
-Unit. \newpage
+ranks. The feature ranks of the most important five OTUs in each
+data-split are shown here. Similar OTUs had the largest impact on the
+predictive performance of L2 logistic regression and L2 SVM with linear
+kernel. Abbreviations: SVM, support vector machine; OTU, Operational
+Taxonomic Unit. \newpage
 \includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
@@
 \includegraphics{Figure_5.png} \textbf{Figure 5. Computational
-efficiency of seven ML models.} The traintimes for training and testing
-of each data-split showed the differences in computational efficieny of
-the seven models. The median traintime in hours was the highest for
-XGBoost and shortest for L2 logistic regression. The boxplot shows
-quartiles at the box ends and the statistical median as the horizontal
-line in the box. The whiskers show the farthest points that are not
-outliers. Outliers are data points that are not within 3/2 times the
-interquartile ranges. Abbreviations: AUROC, area under the receiver
+efficiency of seven ML models.} The training times for training and
+testing of each data-split showed the differences in computational
+efficieny of the seven models. The median traintime in hours was the
+highest for XGBoost and shortest for L2 logistic regression. The boxplot
+shows quartiles at the box ends and the statistical median as the
+horizontal line in the box. The whiskers show the farthest points that
+are not outliers. Outliers are data points that are not within 3/2 times
+the interquartile ranges. Abbreviations: AUROC, area under the receiver
 operating characteristic curve; SVM, support vector machine; XGBoost,

### 2019-07-03 code/learning/FigureS1.R
initial commit

@@ -1,0 +1,111 @@
+# Author: Begum Topcuoglu
+# Date: 2018-02-12
+#
+######################################################################
+# This script plots Figure 2:
+#   1. Y axis: mean cvAUC of 100 datasplits
+#   2. X axis: different hyper-parameters tested in cv(hp)
+######################################################################
+
+
+######################################################################
+# Load in needed functions and libraries
+source('code/learning/functions.R')
+# detach("package:randomForest", unload=TRUE) to run
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
+deps = c("scales","cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+######################################################################
+
+
+######################################################################
+# Load .tsv data generated with modeling pipeline for Logistic Regression
+######################################################################
+
+# Read in the results of trained model of 100 data-splits
+
+all_files <- list.files(path= 'data/process', pattern='combined_all_hp.*', full.names = TRUE)
+
+logit_all <- read_files(all_files[4])
+l2svm_all <- read_files(all_files[3])
+l1svm_all <- read_files(all_files[2])
+
+######################################################################
+#Plot the mean AUC values for hyper parameters tested #
+######################################################################
+
+# Define the base plot for all the modeling methods
+base_plot <-  function(data, x_axis, y_axis){
+  plot <- ggplot(data, aes(x_axis, y_axis)) +
+    geom_line() +
+    geom_point() +
+    theme_bw() +
+    geom_hline(yintercept = 0.5, linetype="dashed") +
+    theme(legend.text=element_text(size=10),
+          legend.title=element_text(size=10),
+          panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank(),
+          panel.background = element_blank(),
[…]

### 2019-07-05 submission/manuscript.Rmd
sent to Joshua:

@@
 
+\newpage
 ## Importance
@@
 Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies.
-However, the developed ML so far are overoptimistic in terms of predictive performance and validity.
+However, the developed ML models so far are overoptimistic in terms of validity and predictive performance.
 Without rigorous ML pipelines, we cannot trust ML models.
 Before we can speed up progress, we need to slow down, define and start using good ML practices.
-
 \newpage
@@
 Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019].
-More notably, we commonly see ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation.
+However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019].
+More notably, we commonly see flawed ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation.
 Even when there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
@@
 CRC is a type of cancer which the human-associated human microbiome is hypothesized to directly contribute to its development and fecal 16S rRNA gene sequences have been used to detect CRC.
-We built ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
+We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
 The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
@@
 We used a cohort of 490 patients with 261 cases of SRN.
-For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colonic health (having colorectal lesions that are identified as SRN or normal).
+For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (having colorectal lesions that are identified as SRN or normal).
 All the cases were independently labeled through colonoscopies.
@@
 Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets.
-We validated the cross-validation performances of each hyperparameter setting over 100 randomizations and selected the best performing hyper-parameter setting in terms of AUROC metric and trained the full training/validation dataset [Figures S1 and S2].
+We validated the cross-validation performances of each hyperparameter setting over 100 randomizations and selected the best performing hyperparameter setting in terms of AUROC metric and trained the full training/validation dataset [Figures S1 and S2].
 We then used the held-out test set to evaluate the prediction performance of each ML model.
@@
 
-We evaluated the prediction performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 1].
+We evaluated the prediction performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 2].
 Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01).
@@
 The difference between the two should be low to suggest the model is not overfitting despite the large number of features.
-The largest difference between the two was in L1 SVM with linear kernel, `r paste(round(abs(difference_model[difference_index[7],1]), 3))`,  followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
+The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel,  followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
 
@@
 We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models.
-We ranked the absolute weights of all the OTUs for each data-split.
+We ranked the absolute weights of all the OTUs for each data-split [Figure 3].
 We calculated the median ranks of these features over the 100 data-splits.
-In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to families *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3].
+In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belonged to families *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3].
 Some of the OTUs with the highest ranks were shared among the linear models.
@@
 We set-up standards for developing and evaluating ML models for microbiome data.
-We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
+First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
 When the differences between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data.
 In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data.
-We performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
-The repetition of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop generalizable ML models and to report reliable performance metrics for a ML model.
+Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
+The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop generalizable ML models and to report reliable performance metrics for a ML model.
 Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model.
 In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
-Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
-We used AUROC metric instead of accuracy in our study to evaluate predictive performance of the ML models.
+Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.Third,  we used AUROC metric instead of accuracy in our study to evaluate predictive performance of the ML models.
 AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
-We performed a full grid search for hyperparameter settings when building a ML model.
-It is not enough to use the default hyperparameter settings when using previously developed ML packages in programming languages like R, Python and Matlab.
+We also performed a full grid search for hyperparameter settings when building a ML model.
+It is not enough to use the default hyperparameter settings when using previously developed ML packages in programming languages such as R, Python and Matlab.
 In the example of L1-regularized SVM with linear kernel (Figure S1), the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
@@
 Random forest model was slow to train 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
-The dataset we used for our study had 490 samples.
-Another criteria when choosing a ML model is the generalizability of it.
+Another criteria when choosing a ML model is generalizability.
 The generalization performance of ML models depends on size.
 The more complex the model, the more data it will need.
-Microbiome studies that have small sample sizes should use less complex models such as L2-regularized logistic regression.
+The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.
 
@@
 
-  The colonic health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs).
+  The colorectal health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs).
 Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas.
@@
 In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
-We can further parallelize the cross-validation step for each hyper-parameter setting.
+We can further parallelize the cross-validation step for each hyperparameter setting.
 
@@
 \includegraphics{Figure_S1.png}
-**Figure S1. Hyperparameter setting performances for linear models.** 
-Abbreviations: 
+**Figure S1.
+Hyperparameter setting performances for linear models.** (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters are used in training the model.
+The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
+
 \newpage
 \includegraphics{Figure_S2.png}
-**Figure S2. Hyperparameter setting performances for non-linear models.** 
-Abbreviations: 
+**Figure S2.
+Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGoost mean cross-validation AUROC values when different hyperparameters are used in training the model.
+The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
+\newpage
+
+\includegraphics{TableS1.pdf}
+**Table S1. ML model complexity and hyperparameter range.** 
 \newpage

### 2019-07-05 submission/manuscript.tex
sent to Joshua:

@@
 
+\newpage
+
 \subsection{Importance}\label{importance}
@@
 Prediction of health outcomes using ML is rapidly being adopted by
-human-associated microbiome studies. However, the developed ML so far
-are overoptimistic in terms of predictive performance and validity.
+human-associated microbiome studies. However, the developed ML models so
+far are overoptimistic in terms of validity and predictive performance.
 Without rigorous ML pipelines, we cannot trust ML models. Before we can
 speed up progress, we need to slow down, define and start using good ML
-practices.
-
-\newpage
+practices. \newpage
 
@@
 lacks clarity and consistency on which methods are used and how these
-methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
-commonly see ML practices such as using ML pipelines where there is no
-seperate held-out test dataset to evaluate model performance or
-reporting few or only the best outcomes of cross-validation. Even when
-there are seperate testing sets to evaluate model performance, there are
-large differences between cross-validation and testing performances that
-indicate overfitting as well as large confidence intervals for testing
-performances (4, 13--19). Moreover, there is a lack of discussion on why
+methods are implemented (12, 13). More notably, we commonly see flawed
+ML practices such as using ML pipelines where there is no seperate
+held-out test dataset to evaluate model performance or reporting few or
+only the best outcomes of cross-validation. Even when there are seperate
+testing sets to evaluate model performance, there are large differences
+between cross-validation and testing performances that indicate
+overfitting as well as large confidence intervals for testing
+performances (4, 14--20). Moreover, there is a lack of discussion on why
 a particular ML model is utilized. Recently, there is a trend towards
@@
 boosting and neural networks without a discussion on if and how much
-model interpretibility is necessary for the study (11, 20--22). Black
+model interpretibility is necessary for the study (11, 21--23). Black
 box machine learning models require posthoc explanations to determine
@@
 be misleading and at times unreliable when making high-stake decisions
-about someone's health (23). The lack of transparency on model selection
+about someone's health (24). The lack of transparency on model selection
 and interpretation as well as flawed modeling methods negatively impact
@@
 microbiome is hypothesized to directly contribute to its development and
-fecal 16S rRNA gene sequences have been used to detect CRC. We built ML
-models using fecal 16S rRNA gene sequences to predict healthy patients
-versus patients with colorectal lesions that were identified by
+fecal 16S rRNA gene sequences have been used to detect CRC. We built
+seven ML models using fecal 16S rRNA gene sequences to predict healthy
+patients versus patients with colorectal lesions that were identified by
 colonoscopy as screen relevant neoplasias (SRN). The study had 261
@@
 patient, we had 6920 features (fecal bacterial abundances) and a
-two-class label that defines their colonic health (having colorectal
+two-class label that defines their colorectal health (having colorectal
 lesions that are identified as SRN or normal). All the cases were
@@
 setting over 100 randomizations and selected the best performing
-hyper-parameter setting in terms of AUROC metric and trained the full
+hyperparameter setting in terms of AUROC metric and trained the full
 training/validation dataset {[}Figures S1 and S2{]}. We then used the
@@
 models when applied to held-out test data using AUROC metric {[}Figure
-1{]}. Random forest had significantly higher test AUROC values than the
+2{]}. Random forest had significantly higher test AUROC values than the
 other models for detecting SRNs when AUROC values were compared to the
@@
 suggest the model is not overfitting despite the large number of
-features. The largest difference between the two was in L1 SVM with
-linear kernel, 0.021, followed by SVM with radial basis function kernel
+features. The largest difference between the two was 0.021 in L1 SVM
+with linear kernel, followed by SVM with radial basis function kernel
 and decision tree with a difference of 0.007 and 0.006, respectively
@@
 weights of the trained models. We ranked the absolute weights of all the
-OTUs for each data-split. We calculated the median ranks of these
-features over the 100 data-splits. In the three linear models, OTUs that
-had the largest median ranks and drove the detection of SRNs belong to
-families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
-OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609),
-genera \emph{Gamella} (OTU00426) and genera \emph{Peptostreptococcus}
-(OTU00367) {[}Figure 3{]}. Some of the OTUs with the highest ranks were
-shared among the linear models. We explained the feature importances in
-non-linear models using permutation importance on the held-out test data
-where we randomly permuted non-correlated features individually and
-groups of correlated features together (see methods) to calculate the
-effect of permuted OTUs or group of OTUs on testing AUROC. The top 5
-OTUs with the largest negative impact on testing AUROC overlapped in
-tree-based models {[}Figure 4{]}. Specifically, permuting
-\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
-predictive performances the most in all tree-based methods {[}Figure
-S3{]}. Decision tree, random forest and XGBoost models' predictive
-performance dropped from 0.6 base testing AUROC median to 0.52, from
-0.69 to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.
+OTUs for each data-split {[}Figure 3{]}. We calculated the median ranks
+of these features over the 100 data-splits. In the three linear models,
+OTUs that had the largest median ranks and drove the detection of SRNs
+belonged to families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
+(OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822,
+OTU00609), genera \emph{Gamella} (OTU00426) and genera
+\emph{Peptostreptococcus} (OTU00367) {[}Figure 3{]}. Some of the OTUs
+with the highest ranks were shared among the linear models. We explained
+the feature importances in non-linear models using permutation
+importance on the held-out test data where we randomly permuted
+non-correlated features individually and groups of correlated features
+together (see methods) to calculate the effect of permuted OTUs or group
+of OTUs on testing AUROC. The top 5 OTUs with the largest negative
+impact on testing AUROC overlapped in tree-based models {[}Figure 4{]}.
+Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
+randomly, dropped the predictive performances the most in all tree-based
+methods {[}Figure S3{]}. Decision tree, random forest and XGBoost
+models' predictive performance dropped from 0.6 base testing AUROC
+median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65, respectively
+{[}Figure 4{]}.
 
@@
 sequence counts to predict a binary health outcome. We set-up standards
-for developing and evaluating ML models for microbiome data. We used a
-held-out test set to illustrate the difference between cross-validation
-and testing AUROC values. When the differences between cross-validation
-and test performance is low, this suggets the models are not overfit and
-that they will perform similar with similar data. In all seven models,
-the differences in cross-validation and testing AUROC values did not
-exceed 0.021 which suggests that these models will be able to test
-similar new data. We performed the initial 80\%-20\% random datasplit
-100 times in our ML pipeline. The repetition of the initial data-split
-to create a held-out test set is a crucial step in the ML pipeline to
-develop generalizable ML models and to report reliable performance
-metrics for a ML model. Depending on the data-split, there is the chance
-of being overoptimistic about the predictive performance of a model. In
-our study, we showed that there was variability in AUROC values between
-different random data-splits in each of the models we tested. Our
-results showed that the testing AUROC values varied 0.23 on average,
-depending on the data-split. We used AUROC metric instead of accuracy in
-our study to evaluate predictive performance of the ML models. AUROC is
-always random at the value 0.5 and is a robust metric when a dataset is
-imbalanced. We performed a full grid search for hyperparameter settings
-when building a ML model. It is not enough to use the default
-hyperparameter settings when using previously developed ML packages in
-programming languages like R, Python and Matlab. In the example of
-L1-regularized SVM with linear kernel (Figure S1), the model showed
-large variability between different regularization coefficients (C) and
-was susceptible to performing poorly if the wrong regularization
-coefficient was assigned to the model by default.
+for developing and evaluating ML models for microbiome data. First, we
+used a held-out test set to illustrate the difference between
+cross-validation and testing AUROC values. When the differences between
+cross-validation and test performance is low, this suggets the models
+are not overfit and that they will perform similar with similar data. In
+all seven models, the differences in cross-validation and testing AUROC
+values did not exceed 0.021 which suggests that these models will be
+able to test similar new data. Second, we performed the initial
+80\%-20\% random datasplit 100 times in our ML pipeline. The
+randomization of the initial data-split to create a held-out test set is
+a crucial step in the ML pipeline to develop generalizable ML models and
+to report reliable performance metrics for a ML model. Depending on the
+data-split, there is the chance of being overoptimistic about the
+predictive performance of a model. In our study, we showed that there
+was variability in AUROC values between different random data-splits in
+each of the models we tested. Our results showed that the testing AUROC
+values varied 0.23 on average, depending on the data-split.Third, we
+used AUROC metric instead of accuracy in our study to evaluate
+predictive performance of the ML models. AUROC is always random at the
+value 0.5 and is a robust metric when a dataset is imbalanced. We also
+performed a full grid search for hyperparameter settings when building a
+ML model. It is not enough to use the default hyperparameter settings
+when using previously developed ML packages in programming languages
+such as R, Python and Matlab. In the example of L1-regularized SVM with
+linear kernel (Figure S1), the model showed large variability between
+different regularization coefficients (C) and was susceptible to
+performing poorly if the wrong regularization coefficient was assigned
+to the model by default.
 
@@
 hours whereas L2-regularized logistic regression trained in 12 minutes.
-The dataset we used for our study had 490 samples. Another criteria when
-choosing a ML model is the generalizability of it. The generalization
-performance of ML models depends on size. The more complex the model,
-the more data it will need. Microbiome studies that have small sample
-sizes should use less complex models such as L2-regularized logistic
-regression.
+Another criteria when choosing a ML model is generalizability. The
+generalization performance of ML models depends on size. The more
+complex the model, the more data it will need. The dataset we used for
+our study had 490 samples, however microbiome studies that have smaller
+sample sizes would benefit from using less complex models such as
+L2-regularized logistic regression.
 
@@
 patients recruited by Great Lakes-New England Early Detection Research
-Network study. These data were obtained from Sze et al (24). The stool
+Network study. These data were obtained from Sze et al (25). The stool
 samples were provided by recruited adult participants who were
@@
 assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(25--27).
+(26--28).
 
@@
 
-The colonic health of the patient was defined as two encompassing
+The colorectal health of the patient was defined as two encompassing
 classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
@@
 testing performance. We can use hyper-band to help us with our
-hyperparameter selection (28).
+hyperparameter selection (29).
 
@@
 can further parallelize the cross-validation step for each
-hyper-parameter setting.
+hyperparameter setting.
 
@@
 \includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
-setting performances for linear models.} Abbreviations: \newpage
+setting performances for linear models.} (A) L2 logistic regression (B)
+L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
+cross-validation AUROC values when different hyperparameters are used in
+training the model. The differences in AUROC values when hyperparameters
+change show that hyperparameter tuning is a crucial step in building a
+ML model.
+
+\newpage
+
 \includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
-setting performances for non-linear models.} Abbreviations: \newpage
+setting performances for non-linear models.} (A) Decision tree (B)
+Random forest (C) SVM with radial basis kernel (D) XGoost mean
+cross-validation AUROC values when different hyperparameters are used in
+training the model. The differences in AUROC values when hyperparameters
+change show that hyperparameter tuning is a crucial step in building a
+ML model. \newpage
+
+\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
+hyperparameter range.} \newpage
 
@@
 
+\hypertarget{ref-dadkhah_gut_2019}{}
+13. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
+\textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
+G}, \textbf{Kuehn T}, \textbf{Gillevet PM}. 2019. Gut microbiome
+identifies risk for colorectal polyps. BMJ Open Gastroenterology
+\textbf{6}:e000297.
+doi:\href{https://doi.org/10.1136/bmjgast-2019-000297}{10.1136/bmjgast-2019-000297}.
+
 \hypertarget{ref-flemer_oral_2018}{}
-13. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
+14. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
 \textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
@@
 \hypertarget{ref-dai_multi-cohort_2018}{}
-14. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
+15. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
 WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
@@
 \hypertarget{ref-montassier_pretreatment_2016}{}
-15. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
+16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
 \textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
@@
 \hypertarget{ref-papa_non-invasive_2012}{}
-16. \textbf{Papa E}, \textbf{Docktor M}, \textbf{Smillie C},
+17. \textbf{Papa E}, \textbf{Docktor M}, \textbf{Smillie C},
 \textbf{Weber S}, \textbf{Preheim SP}, \textbf{Gevers D},
@@
 \hypertarget{ref-mossotto_classification_2017}{}
-17. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
+18. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
 \textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
@@
 \hypertarget{ref-ai_systematic_2017}{}
-18. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
+19. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
 \textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
@@
 \hypertarget{ref-wong_quantitation_2017}{}
-19. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
+20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
 AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
@@
 \hypertarget{ref-galkin_human_2018}{}
-20. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+21. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
 \textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
@@
 \hypertarget{ref-reiman_using_2017}{}
-21. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+22. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
 convolutional neural networks to explore the microbiome, pp. 4269--4272.
@@
 \hypertarget{ref-fioravanti_phylogenetic_2017}{}
-22. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+23. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
 \textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
@@
 \hypertarget{ref-rudin_please_2018}{}
-23. \textbf{Rudin C}. 2018. Please stop explaining black box models for
+24. \textbf{Rudin C}. 2018. Please stop explaining black box models for
 high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
@@
 \hypertarget{ref-sze_leveraging_2018}{}
-24. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+25. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-25. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+26. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-26. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+27. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-27. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+28. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-28. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+29. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-07-05 submission/manuscript.Rmd
collinearity:

@@
 
+We also used permutation importance to interpret linear models to show the differences between the two interpretation methods [Figure S3].
+L1-regularized SVM with linear kernel picked out the same OTUs as the most important features in methods that use feature rankings [Figure 3] as well as permutation importance [Figure S3].
+However, L2-regularized models had different feature importances in the two methods of interpretation due to collinearity in microbial communities.
+
 __The computational efficiency of each ML model.__
@@
 The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.
+
+ML model interpretation methods in the previous studies microbiome studies lacked clarity and transparency.
+The studies that made the model interpreatation code publicly available, used default functions in previously built packages in R, python and matlab.
+In this study we used two methods to interpret our models.
+Feature weights of trained models ad permutation importance.
+the differences between the two interpretation methods
 
@@
 **Figure 3.
-Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature ranks.
-The feature ranks of the most important five OTUs in each data-split are shown here.
+Interpretation of the linear ML models.** The absolute feature weights of (A) L2 logistic regression coefficients (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel were ranked from highest rank 1 to 100 for each data-split.
+The feature ranks of the highest ranked five OTUs based on their median ranks are shown here.
 Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel.
@@
 The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
+The colors of the boxplots stand for the unique OTUs that are shared among the different models; pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for OTU00882.
 For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model.
@@
 **Figure 5.
-Computational efficiency of seven ML models.** The training times for training and testing of each data-split showed the differences in computational efficieny of the seven models.
-The median traintime in hours was the highest for XGBoost and shortest for L2 logistic regression.
+Computational efficiency of seven ML models.** The training times for of each data-split showed the differences in computational efficieny of the seven models.
+The median training time in hours was the highest for XGBoost and shortest for L1-regularized SVM with linear kernel.
 The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
@@
 \newpage
-
+\includegraphics{Figure_S4.png}
+**Figure S3.
+Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel (B) L2-regularized SVM with linear kernel and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
+The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
+Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
 \includegraphics{TableS1.pdf}

### 2019-07-05 submission/manuscript.tex
collinearity:

@@
 
+We also used permutation importance to interpret linear models to show
+the differences between the two interpretation methods {[}Figure S3{]}.
+L1-regularized SVM with linear kernel picked out the same OTUs as the
+most important features in methods that use feature rankings {[}Figure
+3{]} as well as permutation importance {[}Figure S3{]}. However,
+L2-regularized models had different feature importances in the two
+methods of interpretation due to collinearity in microbial communities.
+
 \textbf{The computational efficiency of each ML model.}
@@
 L2-regularized logistic regression.
+
+ML model interpretation methods in the previous studies microbiome
+studies lacked clarity and transparency. The studies that made the model
+interpreatation code publicly available, used default functions in
+previously built packages in R, python and matlab. In this study we used
+two methods to interpret our models. Feature weights of trained models
+ad permutation importance. the differences between the two
+interpretation methods
 
@@
 \includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
-linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
-with linear kernel feature weights (C) L2 SVM with linear kernel feature
-ranks. The feature ranks of the most important five OTUs in each
-data-split are shown here. Similar OTUs had the largest impact on the
-predictive performance of L2 logistic regression and L2 SVM with linear
-kernel. Abbreviations: SVM, support vector machine; OTU, Operational
-Taxonomic Unit. \newpage
+linear ML models.} The absolute feature weights of (A) L2 logistic
+regression coefficients (B) L1 SVM with linear kernel (C) L2 SVM with
+linear kernel were ranked from highest rank 1 to 100 for each
+data-split. The feature ranks of the highest ranked five OTUs based on
+their median ranks are shown here. Similar OTUs had the largest impact
+on the predictive performance of L2 logistic regression and L2 SVM with
+linear kernel. Abbreviations: SVM, support vector machine; OTU,
+Operational Taxonomic Unit. \newpage
 \includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
@@
 and the dashed line show the IQR range and median of the base testing
-AUROC without any permutation performed. For all the tree-based models,
-a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
-predictive performance of the model. Abbreviations: SVM, support vector
-machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
-Operational Taxonomic Unit. \newpage
+AUROC without any permutation performed. The colors of the boxplots
+stand for the unique OTUs that are shared among the different models;
+pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for
+OTU00110, green for OTU00361 and red for OTU00882. For all the
+tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
+the largest impact on predictive performance of the model.
+Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
+Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+\newpage
 \includegraphics{Figure_5.png} \textbf{Figure 5. Computational
-efficiency of seven ML models.} The training times for training and
-testing of each data-split showed the differences in computational
-efficieny of the seven models. The median traintime in hours was the
-highest for XGBoost and shortest for L2 logistic regression. The boxplot
-shows quartiles at the box ends and the statistical median as the
-horizontal line in the box. The whiskers show the farthest points that
-are not outliers. Outliers are data points that are not within 3/2 times
-the interquartile ranges. Abbreviations: AUROC, area under the receiver
-operating characteristic curve; SVM, support vector machine; XGBoost,
-extreme gradient boosting.\\
+efficiency of seven ML models.} The training times for of each
+data-split showed the differences in computational efficieny of the
+seven models. The median training time in hours was the highest for
+XGBoost and shortest for L1-regularized SVM with linear kernel. The
+boxplot shows quartiles at the box ends and the statistical median as
+the horizontal line in the box. The whiskers show the farthest points
+that are not outliers. Outliers are data points that are not within 3/2
+times the interquartile ranges. Abbreviations: AUROC, area under the
+receiver operating characteristic curve; SVM, support vector machine;
+XGBoost, extreme gradient boosting.\\
 \newpage
@@
 ML model. \newpage
-
-\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
-hyperparameter range.} \newpage
+\includegraphics{Figure_S4.png} \textbf{Figure S3. Interpretation of the
+linear ML models with permutation importance.} (A) L1-regularized SVM
+with linear kernel (B) L2-regularized SVM with linear kernel and (C)
+L2-regularized logistic regression were interpreted using permutation
+importance using held-out test set. The gray rectangle and the dashed
+line show the IQR range and median of the base testing AUROC without any
+permutation performed. Abbreviations: SVM, support vector machine; OTU,
+Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational
+Taxonomic Unit. \includegraphics{TableS1.pdf} \textbf{Table S1. ML model
+complexity and hyperparameter range.} \newpage
 

### 2019-07-08 submission/manuscript.Rmd
plot sizes and cowplot adjustments

@@
 Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
-Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+ML models learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction.
+Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
 However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019].
@@
 Even when there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+
 Moreover, there is a lack of discussion on why a particular ML model is utilized.
@@
 To showcase a rigorous ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
-We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients.
+We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of `r paste(nrow(shared))` patients.
 CRC is a type of cancer which the human-associated human microbiome is hypothesized to directly contribute to its development and fecal 16S rRNA gene sequences have been used to detect CRC.
@@
 
-We evaluated the prediction performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 2].
+We evaluated the predictive performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 2].
 Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01).
@@
 The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity.
-We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models.
+We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the absolute feature weights of the trained models.
 We ranked the absolute weights of all the OTUs for each data-split [Figure 3].
@@
 Some of the OTUs with the highest ranks were shared among the linear models.
-We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC.
+
+We explained the feature importances in non-linear models, SVM with radial basis kernel, decision tree, random forest and XGBoost, using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC.
 The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
@@
 
-We also used permutation importance to interpret linear models to show the differences between the two interpretation methods [Figure S3].
-L1-regularized SVM with linear kernel picked out the same OTUs as the most important features in methods that use feature rankings [Figure 3] as well as permutation importance [Figure S3].
-However, L2-regularized models had different feature importances in the two methods of interpretation due to collinearity in microbial communities.
+We used permutation importance to interpret linear models as well to highlight the differences between the two interpretation methods [Figure S3].
+The rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods.
+L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings [Figure 3] and permutation importance methods [Figure S3].
+However, in general L2-regularized models had different OTUs picked out as important in the two methods of interpretation due to collinearity in microbial communities.
 
@@
 
-ML model interpretation methods in the previous studies microbiome studies lacked clarity and transparency.
-The studies that made the model interpreatation code publicly available, used default functions in previously built packages in R, python and matlab.
+Microbiome studies use ML models with a classification task to learn the training data to assign labels, healthy or not, to new data (e.g.
+new samples), but also to learn which features are important enough to discriminate between labels.
+This is a difficult task because we need to correctly interpret our ML models.
 In this study we used two methods to interpret our models.
-Feature weights of trained models ad permutation importance.
-the differences between the two interpretation methods
+Absolute feature weights of trained models and permutation importance.
+When interpreting linear models with the differences we observed between the two interpretation methods is possibly due to collinearity in the dataset.
+Collinearity is that one OTU is highly correlated with other OTUs which means that the regression coefficients/weights are not unique and that they are influenced by each other.
+If they are influenced and not unique then it becomes harder to interpret the model.
+You may want to use different models for prediction and one for interpretation/inference, in one you remove highly correlated OTUs to interpret and in other keep them all to predict.
 
@@
 \includegraphics{Figure_2.png}
+
 **Figure 2.
@@
 \newpage
-\includegraphics{Figure_3.png}
+\includegraphics[height=15cm, width=10cm]{Figure_3.png}
+
 **Figure 3.

### 2019-07-08 submission/manuscript.tex
plot sizes and cowplot adjustments

--- 2019-07-05 submission/manuscript.tex
@@
 Machine learning (ML) methods are effective at recognizing and
-highlighting patterns in complex microbial datasets. Therefore,
-researchers have started to explore the utility of ML models that use
-microbiota associated biomarkers to predict human health and to
-understand the microbial ecology of diseases such as liver cirrhosis,
-colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
-diabetes and others (1--11). However, currently the field's use of ML
-lacks clarity and consistency on which methods are used and how these
-methods are implemented (12, 13). More notably, we commonly see flawed
-ML practices such as using ML pipelines where there is no seperate
-held-out test dataset to evaluate model performance or reporting few or
-only the best outcomes of cross-validation. Even when there are seperate
-testing sets to evaluate model performance, there are large differences
-between cross-validation and testing performances that indicate
-overfitting as well as large confidence intervals for testing
-performances (4, 14--20). Moreover, there is a lack of discussion on why
-a particular ML model is utilized. Recently, there is a trend towards
-using more complex ML models such as random forest, extreme gradient
-boosting and neural networks without a discussion on if and how much
-model interpretibility is necessary for the study (11, 21--23). Black
-box machine learning models require posthoc explanations to determine
-the feature importances in making a prediction. These explanations can
-be misleading and at times unreliable when making high-stake decisions
-about someone's health (24). The lack of transparency on model selection
-and interpretation as well as flawed modeling methods negatively impact
-model validity and reproducibiliy. We need to strive toward better
-machine learning practices by (1) implementing rigourous machine
-learning pipelines and (2) selecting ML models that reflect the goal of
-the study as it will inform our expectations of model accuracy,
-complexity, interpretibility and computational efficiency.
+highlighting patterns in complex microbial datasets. ML models learn
+from existing data to predict the outcomes of new data and allow us to
+infer on the reasons underlying that prediction. Therefore, researchers
+have started to explore the utility of ML models that use microbiota
+associated biomarkers to predict human health and to understand the
+ecological basis of diseases such as liver cirrhosis, colorectal cancer,
+inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
+(1--11). However, currently the field's use of ML lacks clarity and
+consistency on which methods are used and how these methods are
+implemented (12, 13). More notably, we commonly see flawed ML practices
+such as using ML pipelines where there is no seperate held-out test
+dataset to evaluate model performance or reporting few or only the best
+outcomes of cross-validation. Even when there are seperate testing sets
+to evaluate model performance, there are large differences between
+cross-validation and testing performances that indicate overfitting as
+well as large confidence intervals for testing performances (4, 14--20).
+
+Moreover, there is a lack of discussion on why a particular ML model is
+utilized. Recently, there is a trend towards using more complex ML
+models such as random forest, extreme gradient boosting and neural
+networks without a discussion on if and how much model interpretibility
+is necessary for the study (11, 21--23). Black box machine learning
+models require posthoc explanations to determine the feature importances
+in making a prediction. These explanations can be misleading and at
+times unreliable when making high-stake decisions about someone's health
+(24). The lack of transparency on model selection and interpretation as
+well as flawed modeling methods negatively impact model validity and
+reproducibiliy. We need to strive toward better machine learning
+practices by (1) implementing rigourous machine learning pipelines and
+(2) selecting ML models that reflect the goal of the study as it will
+inform our expectations of model accuracy, complexity, interpretibility
+and computational efficiency.
 
@@
 and the same ML pipeline. We used a previously published colorectal
-cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
+cancer (CRC) study (3) which had fecal 16S rRNA gene sequences of 490
 patients. CRC is a type of cancer which the human-associated human
@@
 
-We evaluated the prediction performances of seven binary classification
+We evaluated the predictive performances of seven binary classification
 models when applied to held-out test data using AUROC metric {[}Figure
@@
 interpretibility as they increase in complexity. We interpreted L1 and
-L2 SVM with linear kernel and L2 logistic regression using the feature
-weights of the trained models. We ranked the absolute weights of all the
-OTUs for each data-split {[}Figure 3{]}. We calculated the median ranks
-of these features over the 100 data-splits. In the three linear models,
-OTUs that had the largest median ranks and drove the detection of SRNs
-belonged to families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
-(OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822,
-OTU00609), genera \emph{Gamella} (OTU00426) and genera
-\emph{Peptostreptococcus} (OTU00367) {[}Figure 3{]}. Some of the OTUs
-with the highest ranks were shared among the linear models. We explained
-the feature importances in non-linear models using permutation
-importance on the held-out test data where we randomly permuted
-non-correlated features individually and groups of correlated features
-together (see methods) to calculate the effect of permuted OTUs or group
-of OTUs on testing AUROC. The top 5 OTUs with the largest negative
-impact on testing AUROC overlapped in tree-based models {[}Figure 4{]}.
-Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
-randomly, dropped the predictive performances the most in all tree-based
-methods {[}Figure S3{]}. Decision tree, random forest and XGBoost
-models' predictive performance dropped from 0.6 base testing AUROC
-median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65, respectively
-{[}Figure 4{]}.
+L2 SVM with linear kernel and L2 logistic regression using the absolute
+feature weights of the trained models. We ranked the absolute weights of
+all the OTUs for each data-split {[}Figure 3{]}. We calculated the
+median ranks of these features over the 100 data-splits. In the three
+linear models, OTUs that had the largest median ranks and drove the
+detection of SRNs belonged to families \emph{Lachnospiraceae}, and
+\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
+OTU00015, OTU00768, OTU00822, OTU00609), genera \emph{Gamella}
+(OTU00426) and genera \emph{Peptostreptococcus} (OTU00367) {[}Figure
+3{]}. Some of the OTUs with the highest ranks were shared among the
+linear models.
+
+We explained the feature importances in non-linear models, SVM with
+radial basis kernel, decision tree, random forest and XGBoost, using
+permutation importance on the held-out test data where we randomly
+permuted non-correlated features individually and groups of correlated
+features together (see methods) to calculate the effect of permuted OTUs
+or group of OTUs on testing AUROC. The top 5 OTUs with the largest
+negative impact on testing AUROC overlapped in tree-based models
+{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
+(OTU00367) abundances randomly, dropped the predictive performances the
+most in all tree-based methods {[}Figure S3{]}. Decision tree, random
+forest and XGBoost models' predictive performance dropped from 0.6 base
+testing AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
+respectively {[}Figure 4{]}.
+
+We used permutation importance to interpret linear models as well to
+highlight the differences between the two interpretation methods
+{[}Figure S3{]}. The rankings of OTUs that discriminate between having
+an SRN or not, changed between different interpretation methods.
+L1-regularized SVM with linear kernel picked out some of the same OTUs
+(OTU00822, OTU01239, OTU00609) as important in feature rankings
+{[}Figure 3{]} and permutation importance methods {[}Figure S3{]}.
+However, in general L2-regularized models had different OTUs picked out
+as important in the two methods of interpretation due to collinearity in
+microbial communities.
 
@@
 L2-regularized logistic regression.
+
+Microbiome studies use ML models with a classification task to learn the
+training data to assign labels, healthy or not, to new data (e.g.~new
+samples), but also to learn which features are important enough to
+discriminate between labels. This is a difficult task because we need to
+correctly interpret our ML models. In this study we used two methods to
+interpret our models. Absolute feature weights of trained models and
+permutation importance. When interpreting linear models with the
+differences we observed between the two interpretation methods is
+possibly due to collinearity in the dataset. Collinearity is that one
+OTU is highly correlated with other OTUs which means that the regression
+coefficients/weights are not unique and that they are influenced by each
+other. If they are influenced and not unique then it becomes harder to
+interpret the model. You may want to use different models for prediction
+and one for interpretation/inference, in one you remove highly
+correlated OTUs to interpret and in other keep them all to predict.
 
@@
 curve \newpage
-\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
-classification performance of ML models using AUROC values of all cross
-validation and testing performances. } The median AUROC for diagnosing
-individuals with SRN using bacterial abundances was higher than chance
-(depicted by horizontal line at 0.50) for all the ML models.
-Discriminative performance of random forest model was higher than other
-ML models. The boxplot shows quartiles at the box ends and the
-statistical median as the horizontal line in the box. The whiskers show
-the farthest points that are not outliers. Outliers are data points that
-are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
-screen-relevant neoplasias; AUROC, area under the receiver operating
-characteristic curve; SVM, support vector machine; XGBoost, extreme
-gradient boosting \newpage
-\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
-linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
-with linear kernel feature weights (C) L2 SVM with linear kernel feature
-ranks. The feature ranks of the most important five OTUs in each
-data-split are shown here. Similar OTUs had the largest impact on the
-predictive performance of L2 logistic regression and L2 SVM with linear
-kernel. Abbreviations: SVM, support vector machine; OTU, Operational
-Taxonomic Unit. \newpage
+\includegraphics{Figure_2.png}
+
+\textbf{Figure 2. Generalization and classification performance of ML
+models using AUROC values of all cross validation and testing
+performances. } The median AUROC for diagnosing individuals with SRN
+using bacterial abundances was higher than chance (depicted by
+horizontal line at 0.50) for all the ML models. Discriminative
+performance of random forest model was higher than other ML models. The
+boxplot shows quartiles at the box ends and the statistical median as
+the horizontal line in the box. The whiskers show the farthest points
+that are not outliers. Outliers are data points that are not within 3/2
+times the interquartile ranges. Abbreviations: SRN, screen-relevant
+neoplasias; AUROC, area under the receiver operating characteristic
+curve; SVM, support vector machine; XGBoost, extreme gradient boosting
+\newpage
+\includegraphics[height=15cm, width=10cm]{Figure_3.png}
+
+\textbf{Figure 3. Interpretation of the linear ML models.} The absolute
+feature weights of (A) L2 logistic regression coefficients (B) L1 SVM
+with linear kernel (C) L2 SVM with linear kernel were ranked from
+highest rank 1 to 100 for each data-split. The feature ranks of the
+highest ranked five OTUs based on their median ranks are shown here.
+Similar OTUs had the largest impact on the predictive performance of L2
+logistic regression and L2 SVM with linear kernel. Abbreviations: SVM,
+support vector machine; OTU, Operational Taxonomic Unit. \newpage
 \includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the

### 2019-07-08 code/learning/Figure3.R

@@ -70,9 +70,9 @@ plot_feature_ranks <- function(data){
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           panel.background = element_blank(),
-          axis.text.x=element_text(size = 8, colour='black'),
-          axis.text.y=element_text(size = 8, colour='black'),
-          axis.title.x=element_blank())
+          axis.title.x=element_blank(),
+          axis.text.x=element_blank(),
+          axis.text.y=element_text(size = 8, colour='black'))
     return(plot)
 }
 
@@ -83,17 +83,20 @@ plot_feature_ranks <- function(data){
 
 L1_SVM_imp <- get_feature_ranked_files("data/process/combined_L1_Linear_SVM_feature_ranking.tsv", "L1_Linear_SVM")
 l1_svm_graph <- plot_feature_ranks(L1_SVM_imp)  +
-  scale_x_discrete(name = "L1 Linear SVM")
+  scale_x_discrete(name = "L1 Linear 
+SVM")
 
 
 L2_SVM_imp <- get_feature_ranked_files("data/process/combined_L2_Linear_SVM_feature_ranking.tsv", "L2_Linear_SVM")
 l2_svm_graph <- plot_feature_ranks(L2_SVM_imp)+
-  scale_x_discrete(name = "L2 Linear SVM ")
+  scale_x_discrete(name = "L2 Linear 
+SVM ")
 
 logit_imp <- get_feature_ranked_files("data/process/combined_L2_Logistic_Regression_feature_ranking.tsv", "L2_Logistic_Regression")
 logit_graph <- plot_feature_ranks(logit_imp) +
   scale_x_discrete(name = "L2 Logistic 
-Regression ")
+Regression ") +
+  theme(axis.text.x=element_text(size = 8, colour='black'))
 # -------------------------------------------------------------------->
 
 
@@ -102,8 +105,8 @@ Regression ")
 ######################################################################
 #combine with cowplot
 
-linear <- plot_grid(logit_graph, l1_svm_graph, l2_svm_graph, labels = c("A", "B", "C"), align = 'v', ncol = 1)
+linear <- plot_grid(l1_svm_graph, l2_svm_graph, logit_graph, labels = c("A", "B", "C"), align = 'v', ncol = 1)
 
 ggdraw(add_sub(linear, "Feature Ranks", vpadding=grid::unit(0,"lines"), size=10, x=0.6))
 
[…]

### 2019-07-08 code/learning/Figure4.R

@@ -186,9 +186,10 @@ base_nonlin_plot <-  function(data, name, list){
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           panel.background = element_blank(),
-          axis.text.x=element_text(size = 16, colour='black'),
           axis.text.y=element_text(size = 16, colour='black'), 
-          axis.title.x=element_blank()) 
+          axis.title.x=element_blank(),
+          axis.text.x=element_blank(), 
+          axis.ticks = element_line(colour = "black", size = 1.1)) 
   
   #-----------------------Save median info ------------------------ #
   #-----------------------Save top 5 features ------------------------ #
@@ -228,20 +229,22 @@ dt_plot <- base_nonlin_plot(dt, "Decision_Tree", list1) +
 # --------------------------- Random Forest ----------------------------->
 # Plot most important 5 features effect on AUROC
 rf_plot <- base_nonlin_plot(rf, "Random_Forest", list1) +
-  scale_x_discrete(name = "Random forest ") 
+  scale_x_discrete(name = "Random forest ") +
+  theme(axis.text.x=element_text(size = 16, colour='black'))
 # ----------------------------------------------------------------------->
 
 # --------------------------- XGBoost ----------------------------->
 # Plot most important 5 features effect on AUROC
 xgboost_plot <- base_nonlin_plot(xgboost, "XGBoost", list1)+
-  scale_x_discrete(name = "XGBoost ") 
+  scale_x_discrete(name = "XGBoost ") +
+  theme(axis.text.x=element_text(size = 16, colour='black'))
 # ----------------------------------------------------------------------->
 
 ######################################################################
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 #combine with cowplot
-perm_tree_based <- plot_grid(rbf_plot, dt_plot, rf_plot, xgboost_plot, labels = c("A", "B", "C", "D"), cols=2)
+perm_tree_based <- plot_grid(rbf_plot, dt_plot, rf_plot, xgboost_plot, labels = c("A", "B", "C", "D"), cols=2, scale = 0.97)
 ggdraw(add_sub(perm_tree_based, "AUROC with the OTU permuted randomly", size=20, vpadding=grid::unit(0,"lines"), y=5, x=0.5, vjust=4.75))
 
 ggsave("Figure_4.png", plot = last_plot(), device = 'png', path = 'submission', width = 10, height = 8)

### 2019-07-09 submission/manuscript.Rmd
prep for sarah and add interp paragraph

@@
 Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure S3].
-Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
+Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
 
-We used permutation importance to interpret linear models as well to highlight the differences between the two interpretation methods [Figure S3].
+We used permutation importance to interpret linear models as well, to highlight the differences between the two interpretation methods [Figure S3].
 The rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods.
-L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings [Figure 3] and permutation importance methods [Figure S3].
-However, in general L2-regularized models had different OTUs picked out as important in the two methods of interpretation due to collinearity in microbial communities.
+L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings [Figure 3] and permutation importance methods [Figure S3] but with different ranks.
+The difference was more pronounced in L2-regularized models as they had mostly different OTUs picked out as important due to collinearity in microbial communities.
 
@@
 
-As the complexity of a ML model and the number of tuned hyperparameter settings increased [Table S1], its training times increased as well [Figure 5].
+As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5].
 Linear models trained faster than non-linear models.
@@
 First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
-When the differences between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data.
-In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data.
+When the difference between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data.
+In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data.
 Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
-The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop generalizable ML models and to report reliable performance metrics for a ML model.
+The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics for a ML model.
 Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model.
 In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
-Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.Third,  we used AUROC metric instead of accuracy in our study to evaluate predictive performance of the ML models.
+Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
+Third, we used the AUROC metric instead of accuracy in our study to evaluate the predictive performance of the ML models.
 AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
@@
 It is not enough to use the default hyperparameter settings when using previously developed ML packages in programming languages such as R, Python and Matlab.
-In the example of L1-regularized SVM with linear kernel (Figure S1), the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
+In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
 
-Our results showed that we should choose to use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden.
+In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden.
+Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not, to new data, but also to learn which features are important to discriminate between labels.
+If the goal of a study is not just to predict a health outcome but also to learn the ecology behind a disease and to identify microbial biomarkers of a disease, then the ML model has to be interpretable.
 In terms of predictive performance, random forest model had testing AUROC values statistically significantly larger than others.
 However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015.
-In terms of interpretation, random forest was a more complex ML model and it could only be explained using methods such as permutation importance.
-On the other hand, L2-regularized logistic regression was easy to interpret (i.e.
-regression coefficients of the trained model).
-If the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers of a disease, then easily interpretable models would be better suited.
-It is also important to consider the computational burden of developing ML models.
-Random forest model was slow to train 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
-Another criteria when choosing a ML model is generalizability.
-The generalization performance of ML models depends on size.
+In terms of interpretability, random forest was a complex ML model and it could only be explained using methods such as permutation importance.
+On the other hand, L2-regularized logistic regression was easier to interpret (i.e.
+ranking absolute regression coefficients of the trained model).
+
+Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models.
+In this study we used two methods to interpret our linear models; absolute feature weights of trained models and permutation importance.
+We observed differences in outcome between the two interpretation methods due to collinearity in the dataset.
+Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes regression coefficients to not be unique.
+The regression coefficients of correlated OTUs are influenced by one another which makes it difficult to interpret a ML model.
+There are ways to make it easier to interpret models.; (1) use different models for prediction and interpretation in which highly correlated OTUs are removed, (2) get the highly ranked OTUs and check what othe OTUs they are correlated with.
+The interpretation however should not be a final answer but should be used to generate hypothesis that are tested with followup experiments.
+
+There are other criteria when choosing ML models such as the computational burden of developing it, sample size and balance of the dataset.
+In terms of computation burden, random forest model was slow to train 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
+The generalization performance of ML models depends on sample size.
 The more complex the model, the more data it will need.
 The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.
-
-Microbiome studies use ML models with a classification task to learn the training data to assign labels, healthy or not, to new data (e.g.
-new samples), but also to learn which features are important enough to discriminate between labels.
-This is a difficult task because we need to correctly interpret our ML models.
-In this study we used two methods to interpret our models.
-Absolute feature weights of trained models and permutation importance.
-When interpreting linear models with the differences we observed between the two interpretation methods is possibly due to collinearity in the dataset.
-Collinearity is that one OTU is highly correlated with other OTUs which means that the regression coefficients/weights are not unique and that they are influenced by each other.
-If they are influenced and not unique then it becomes harder to interpret the model.
-You may want to use different models for prediction and one for interpretation/inference, in one you remove highly correlated OTUs to interpret and in other keep them all to predict.
 

### 2019-07-09 submission/manuscript.tex
prep for sarah and add interp paragraph

@@
 (OTU00822, OTU01239, OTU00609) as important in feature rankings
-{[}Figure 3{]} and permutation importance methods {[}Figure S3{]}.
-However, in general L2-regularized models had different OTUs picked out
-as important in the two methods of interpretation due to collinearity in
-microbial communities.
+{[}Figure 3{]} and permutation importance methods {[}Figure S3{]} but
+with different ranks. The difference was more pronounced in
+L2-regularized models as they had mostly different OTUs picked out as
+due to collinearity in microbial communities.
 
@@
 all seven models, the differences in cross-validation and testing AUROC
-values did not exceed 0.021 which suggests that these models will be
-able to test similar new data. Second, we performed the initial
-80\%-20\% random datasplit 100 times in our ML pipeline. The
-randomization of the initial data-split to create a held-out test set is
-a crucial step in the ML pipeline to develop generalizable ML models and
-to report reliable performance metrics for a ML model. Depending on the
-data-split, there is the chance of being overoptimistic about the
-predictive performance of a model. In our study, we showed that there
-was variability in AUROC values between different random data-splits in
-each of the models we tested. Our results showed that the testing AUROC
-values varied 0.23 on average, depending on the data-split.Third, we
-used AUROC metric instead of accuracy in our study to evaluate
-predictive performance of the ML models. AUROC is always random at the
-value 0.5 and is a robust metric when a dataset is imbalanced. We also
-performed a full grid search for hyperparameter settings when building a
-ML model. It is not enough to use the default hyperparameter settings
-when using previously developed ML packages in programming languages
-such as R, Python and Matlab. In the example of L1-regularized SVM with
-linear kernel (Figure S1), the model showed large variability between
-different regularization coefficients (C) and was susceptible to
-performing poorly if the wrong regularization coefficient was assigned
-to the model by default.
+values did not exceed 0.021 which suggests that these models are
+generalizable and can be used to test similar new data. Second, we
+performed the initial 80\%-20\% random datasplit 100 times in our ML
+pipeline. The randomization of the initial data-split to create a
+held-out test set is a crucial step in the ML pipeline to develop robust
+ML models and to report reliable performance metrics for a ML model.
+Depending on the data-split, there is the chance of being overoptimistic
+about the predictive performance of a model. In our study, we showed
+that there was variability in AUROC values between different random
+data-splits in each of the models we tested. Our results showed that the
+testing AUROC values varied 0.23 on average, depending on the
+data-split. Third, we used AUROC metric instead of accuracy in our study
+to evaluate predictive performance of the ML models. AUROC is always
+random at the value 0.5 and is a robust metric when a dataset is
+imbalanced. We also performed a full grid search for hyperparameter
+settings when building a ML model. It is not enough to use the default
+hyperparameter settings when using previously developed ML packages in
+programming languages such as R, Python and Matlab. In the example of
+L1-regularized SVM with linear kernel (Figure S1), the model showed
+large variability between different regularization coefficients (C) and
+was susceptible to performing poorly if the wrong regularization
+coefficient was assigned to the model by default.
 
-Our results showed that we should choose to use ML models based on the
-goal of the study and our expectations of predictive performance,
-interpretibility and computational burden. In terms of predictive
-performance, random forest model had testing AUROC values statistically
-significantly larger than others. However the second best model was
-L2-regularized logistic regression following random forest with a median
-AUROC difference of only 0.015. In terms of interpretation, random
-forest was a more complex ML model and it could only be explained using
-methods such as permutation importance. On the other hand,
-L2-regularized logistic regression was easy to interpret
-(i.e.~regression coefficients of the trained model). If the goal of a
-study is to learn the ecology behind a disease and to identify microbial
-biomarkers of a disease, then easily interpretable models would be
-better suited. It is also important to consider the computational burden
-of developing ML models. Random forest model was slow to train 83.2
-hours whereas L2-regularized logistic regression trained in 12 minutes.
-Another criteria when choosing a ML model is generalizability. The
-generalization performance of ML models depends on size. The more
-complex the model, the more data it will need. The dataset we used for
-our study had 490 samples, however microbiome studies that have smaller
-sample sizes would benefit from using less complex models such as
-L2-regularized logistic regression.
+In this study, we benchmarked seven ML models with different
+classification algorithms to show that we should use ML models based on
+the goal of the study and our expectations of predictive performance,
+interpretibility and computational burden. Microbiome studies use ML
+models with the classification task to learn the training data to assign
+labels, such as healthy or not, to new data (e.g.~new samples), but also
+to learn which features are important enough to discriminate between
+labels. If the goal of a study is not just to predict a health outcome
+but also to learn the ecology behind a disease and to identify microbial
+biomarkers of a disease, then interpretibility of a ML model is crucial.
+In terms of predictive performance, random forest model had testing
+AUROC values statistically significantly larger than others. However the
+second best model was L2-regularized logistic regression following
+random forest with a median AUROC difference of only 0.015. In terms of
+interpretability, random forest was a more complex ML model and it could
+only be explained using methods such as permutation importance. On the
+other hand, L2-regularized logistic regression was easier to interpret
+(i.e.~ranking absolute regression coefficients of the trained model).
 
-Microbiome studies use ML models with a classification task to learn the
-training data to assign labels, healthy or not, to new data (e.g.~new
-samples), but also to learn which features are important enough to
-discriminate between labels. This is a difficult task because we need to
-correctly interpret our ML models. In this study we used two methods to
-interpret our models. Absolute feature weights of trained models and
-permutation importance. When interpreting linear models with the
-differences we observed between the two interpretation methods is
-possibly due to collinearity in the dataset. Collinearity is that one
-OTU is highly correlated with other OTUs which means that the regression
-coefficients/weights are not unique and that they are influenced by each
-other. If they are influenced and not unique then it becomes harder to
-interpret the model. You may want to use different models for prediction
-and one for interpretation/inference, in one you remove highly
-correlated OTUs to interpret and in other keep them all to predict.
+Even with interpretable models, we need to be careful with how we treat
+the information we get from the models. In this study we used two
+methods to interpret our linear models; absolute feature weights of
+trained models and permutation importance. We observed differences in
+outcome between the two interpretation methods due to collinearity in
+the dataset. Collinearity in a microbial dataset is when one OTU is
+highly correlated with another OTU which causes regression coefficients
+to not be unique. The regression coefficients of correlated OTUs are
+influenced by one another which makes it difficult to interpret a ML
+model. There are ways to make it easier to interpret models.; (1) use
+different models for prediction and interpretation in which highly
+correlated OTUs are removed, (2) get the highly ranked OTUs and check
+what othe OTUs they are correlated with. The interpretation however
+should not be a final answer but should be used to generate hypothesis
+that are tested with followup experiments.
+
+There are other criteria when choosing ML models such as the
+computational burden of developing it, sample size and balance of the
+dataset. In terms of computation burden, random forest model was slow to
+train 83.2 hours whereas L2-regularized logistic regression trained in
+12 minutes. The generalization performance of ML models depends on
+sample size. The more complex the model, the more data it will need. The
+dataset we used for our study had 490 samples, however microbiome
+studies that have smaller sample sizes would benefit from using less
+complex models such as L2-regularized logistic regression.
 

### 2019-07-10 submission/manuscript.Rmd
update FigS4 and add more to manuscript to send Sarah

@@
 Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call the validity of developed ML models into question.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed ML models into question.
 Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers.
@@
 Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
-This study showed that we should choose ML models based on our expectations of predictive performance, interpretability and our computational resources.
-It also established standards for modeling pipelines of microbiome-associated ML models.
+In this study, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.
+Additionally, we established standards for the development of modeling pipelines for microbiome-associated ML models
 
 \newpage
-## Importance
+## Importance (needs work)
 
@@
 Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
-ML models learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction.
+They learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction.
 Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019].
-More notably, we commonly see flawed ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation.
+However, the field's use of ML lacks clarity and consistency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019].
+More notably, we commonly see flawed ML practices such as (1) using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or (2) reporting few or only the best outcomes of cross-validation.
 Even when there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
@@
 Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-Black box machine learning models require posthoc explanations to determine the feature importances in making a prediction.
-These explanations can be misleading and at times unreliable when making high-stake decisions about someone's health [@rudin_please_2018].
+Black box machine learning models are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction.
+These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018].
+The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].
+
 The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliy.
-We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.
-
-To showcase a rigorous ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
+We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
 We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of `r paste(nrow(shared))` patients.
-CRC is a type of cancer which the human-associated human microbiome is hypothesized to directly contribute to its development and fecal 16S rRNA gene sequences have been used to detect CRC.
+The human-associated human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
 We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
@@
 We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
-Our ML pipeline utilized held-out test data to evaluate predictive performance and generalizability of each ML model.
-We used the area under the receiver operating characteristic curve (AUROC) as the predictive performance metric.
+We used the area under the receiver operating characteristic curve (AUROC) as a metric for determining the predictibve performance of each model.
+The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance.
 The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`.
@@
 
-We explained the feature importances in non-linear models, SVM with radial basis kernel, decision tree, random forest and XGBoost, using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC.
+We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a metod called permutation importance on the held-out test data.
+Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together and calculate how much the predictive performance of the model (i.e AUROC values) decrease if those specific OTUs or group of OTUs are permuted randomly.
+We ranked the OTUs based on how much they decreased the test AUROC values if they were randomly permuted, with the largest decrease ranking highest.
 The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
@@
 
-We used permutation importance to interpret linear models as well, to highlight the differences between the two interpretation methods [Figure S3].
-The rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods.
-L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings [Figure 3] and permutation importance methods [Figure S3] but with different ranks.
-The difference was more pronounced in L2-regularized models as they had mostly different OTUs picked out as important due to collinearity in microbial communities.
+To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3].
+The importance rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods.
+L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in making a prediction in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3].
+Similarly,  L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation models, OTU00659 and OTU00012, respectively.
+However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.
 
@@
 The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics for a ML model.
-Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model.
+Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model.
 In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
-Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.
+Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits.
 Third, we used the AUROC metric instead of accuracy in our study to evaluate the predictive performance of the ML models.
@@
 We also performed a full grid search for hyperparameter settings when building a ML model.
-It is not enough to use the default hyperparameter settings when using previously developed ML packages in programming languages such as R, Python and Matlab.
+Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model.
 In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
@@
 Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models.
-In this study we used two methods to interpret our linear models; absolute feature weights of trained models and permutation importance.
-We observed differences in outcome between the two interpretation methods due to collinearity in the dataset.
-Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes regression coefficients to not be unique.
-The regression coefficients of correlated OTUs are influenced by one another which makes it difficult to interpret a ML model.
-There are ways to make it easier to interpret models.; (1) use different models for prediction and interpretation in which highly correlated OTUs are removed, (2) get the highly ranked OTUs and check what othe OTUs they are correlated with.
-The interpretation however should not be a final answer but should be used to generate hypothesis that are tested with followup experiments.
+In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
+We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset.
+Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique.
+The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model.
+To avoid misinterpreting the models we can; (1) use ML models that has highly correlated OTUs removed, (2) use the highly ranked OTUs that are correlated with other OTUs to generate hypotheses about the ecology of the disease and test them with followup experiments.
 
-There are other criteria when choosing ML models such as the computational burden of developing it, sample size and balance of the dataset.
-In terms of computation burden, random forest model was slow to train 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
+There are other criteria when choosing ML models such as the computational burden of developing it and sample size.
+In terms of computational burden, random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
 The generalization performance of ML models depends on sample size.
@@
 The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.
+
+This study highlights the need to make educated choices at every step of developing a ML model with microbiome data.
+Model selection, setting up a rigorous ML pipeline and choosing the right interpretation method contribute to achieving the level of validity and accountability we want from models built for patient health.
 
@@
 \includegraphics{Figure_4.png}
+
 **Figure 4.
@@
 \includegraphics{Figure_5.png}
+
 **Figure 5.
@@
 \includegraphics{Figure_S1.png}
+
 **Figure S1.
@@
 \includegraphics{Figure_S2.png}
+
 **Figure S2.
@@
 \newpage
-\includegraphics{Figure_S4.png}
+\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
+
 **Figure S3.
@@
 Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+
+\newpage
 \includegraphics{TableS1.pdf}

### 2019-07-10 submission/manuscript.tex
update FigS4 and add more to manuscript to send Sarah

@@
 that predict health outcomes from bacterial abundances, but rigourous ML
-models are scarce due to the flawed modeling methods that call the
-validity of developed ML models into question. Furthermore, the use of
-black box ML models has hindered the validation of microbial biomarkers.
-To overcome these challenges, we benchmarked seven different ML models
-that use fecal 16S rRNA sequences to predict the presence/absence of
-colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
-cases). To show the effect of model selection, we assessed the
-predictive performance, interpretability, and computational efficiency
-of the following models: L2-regularized logistic regression, L1 and L2
-support vector machines (SVM) with linear and radial basis function
-kernels, a decision tree, random forest, and extreme gradient boosting
-(XGBoost). The random forest model was best at detecting CRC lesions
-with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to
-interpret. Despite its simplicity, L2-regularized logistic regression
-followed random forest in predictive performance with an AUROC of 0.680,
-and it trained much faster (12 min). This study showed that we should
-choose ML models based on our expectations of predictive performance,
-interpretability and our computational resources. It also established
-standards for modeling pipelines of microbiome-associated ML models.
+models are scarce due to the flawed methods that call the validity of
+developed ML models into question. Furthermore, the use of black box ML
+models has hindered the validation of microbial biomarkers. To overcome
+these challenges, we benchmarked seven different ML models that use
+fecal 16S rRNA sequences to predict the presence/absence of colorectal
+cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To
+show the effect of model selection, we assessed the predictive
+performance, interpretability, and computational efficiency of the
+following models: L2-regularized logistic regression, L1 and L2 support
+vector machines (SVM) with linear and radial basis function kernels, a
+decision tree, random forest, and extreme gradient boosting (XGBoost).
+The random forest model was best at detecting CRC lesions with an AUROC
+of 0.695 but it was slow to train (83.2 h) and hard to interpret.
+Despite its simplicity, L2-regularized logistic regression followed
+random forest in predictive performance with an AUROC of 0.680, and it
+trained much faster (12 min). In this study, we showed that ML models
+should be chosen based on expectations of predictive performance,
+interpretability and available computational resources. Additionally, we
+established standards for the development of modeling pipelines for
+microbiome-associated ML models
 
@@
 Machine learning (ML) methods are effective at recognizing and
-highlighting patterns in complex microbial datasets. ML models learn
-from existing data to predict the outcomes of new data and allow us to
-infer on the reasons underlying that prediction. Therefore, researchers
-have started to explore the utility of ML models that use microbiota
+highlighting patterns in complex microbial datasets. They learn from
+existing data to predict the outcomes of new data and allow us to infer
+on the reasons underlying that prediction. Therefore, researchers have
+started to explore the utility of ML models that use microbiota
 associated biomarkers to predict human health and to understand the
@@
 inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
-(1--11). However, currently the field's use of ML lacks clarity and
-consistency on which methods are used and how these methods are
-implemented (12, 13). More notably, we commonly see flawed ML practices
-such as using ML pipelines where there is no seperate held-out test
-dataset to evaluate model performance or reporting few or only the best
-outcomes of cross-validation. Even when there are seperate testing sets
-to evaluate model performance, there are large differences between
-cross-validation and testing performances that indicate overfitting as
-well as large confidence intervals for testing performances (4, 14--20).
+(1--11). However, the field's use of ML lacks clarity and consistency in
+which methods are used and how these methods are implemented (12, 13).
+More notably, we commonly see flawed ML practices such as (1) using ML
+pipelines where there is no seperate held-out test dataset to evaluate
+model performance or (2) reporting few or only the best outcomes of
+cross-validation. Even when there are seperate testing sets to evaluate
+model performance, there are large differences between cross-validation
+and testing performances that indicate overfitting as well as large
+confidence intervals for testing performances (4, 14--20).
 
@@
 is necessary for the study (11, 21--23). Black box machine learning
-models require posthoc explanations to determine the feature importances
-in making a prediction. These explanations can be misleading and at
-times unreliable when making high-stake decisions about someone's health
-(24). The lack of transparency on model selection and interpretation as
-well as flawed modeling methods negatively impact model validity and
+models are not inherently interpretable and require posthoc explanations
+to determine the feature importances in making a prediction. These
+explanations can be misleading and at times unreliable when making
+high-stake decisions about patient health (24). The models we develop
+for healthcare, both to predict disease and to understand underlying
+reasons behind that prediction, should be transparent and accountable
+(25).
+
+The lack of transparency on model selection and interpretation as well
+as flawed modeling methods negatively impact model validity and
 reproducibiliy. We need to strive toward better machine learning
@@
 inform our expectations of model accuracy, complexity, interpretibility
-and computational efficiency.
-
-To showcase a rigorous ML pipeline and to shed light on how much ML
-model selection can affect modeling results, we performed an empirical
-analysis comparing several different ML models using the same dataset
-and the same ML pipeline. We used a previously published colorectal
-cancer (CRC) study (3) which had fecal 16S rRNA gene sequences of 490
-patients. CRC is a type of cancer which the human-associated human
-microbiome is hypothesized to directly contribute to its development and
-fecal 16S rRNA gene sequences have been used to detect CRC. We built
-seven ML models using fecal 16S rRNA gene sequences to predict healthy
-patients versus patients with colorectal lesions that were identified by
-colonoscopy as screen relevant neoplasias (SRN). The study had 261
-normal and 229 SRN samples. We established modeling pipelines for
-L2-regularized logistic regression, L1 and L2 support vector machines
-(SVM) with linear and radial basis function kernels, a decision tree,
-random forest and XGBoost. Our ML pipeline utilized held-out test data
-to evaluate predictive performance and generalizability of each ML
-model. We used the area under the receiver operating characteristic
-curve (AUROC) as the predictive performance metric. The median test
-AUROC varied from 0.601 to 0.695. Random forest had the highest median
-AUROC for detecting SRN. Despite its simplicity, the L2-regularized
-logistic regression was second best in predictive performance. In terms
-of computational efficiency, L1 SVM with linear kernel trained the
-fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
-(155.104 hours, std ± 0.959). We also found that depending on how the
-data is split to create a held-out test set, the AUROC values of a ML
-model can vary up to 0.295 which highlighted the importance of
-performing many randomized modeling runs. This study established
-standards for microbiome-associated ML models and underscored the
-importance of model selection.
+and computational efficiency.To showcase a rigorous ML pipeline and to
+shed light on how ML model selection can affect modeling results, we
+performed an empirical analysis comparing several different ML models
+using the same dataset and the same ML pipeline. We used a previously
+published colorectal cancer (CRC) study (3) which had fecal 16S rRNA
+gene sequences of 490 patients. The human-associated human microbiome is
+hypothesized to directly contribute to the development of CRC and fecal
+16S rRNA gene sequences have been used to detect it (1, 3, 4, 26). We
+built seven ML models using fecal 16S rRNA gene sequences to predict
+healthy patients versus patients with colorectal lesions that were
+identified by colonoscopy as screen relevant neoplasias (SRN). The study
+had 261 normal and 229 SRN samples. We established modeling pipelines
+for L2-regularized logistic regression, L1 and L2 support vector
+machines (SVM) with linear and radial basis function kernels, a decision
+tree, random forest and XGBoost. We used the area under the receiver
+operating characteristic curve (AUROC) as a metric for determining the
+predictibve performance of each model. The AUROC ranges from 1.0, where
+the model perfectly distinguishes between cases and controls, to 0.50,
+where the model's predictions are no different from random chance. The
+median test AUROC varied from 0.601 to 0.695. Random forest had the
+highest median AUROC for detecting SRN. Despite its simplicity, the
+L2-regularized logistic regression was second best in predictive
+performance. In terms of computational efficiency, L1 SVM with linear
+kernel trained the fastest (0.202 hours, std ± 0.028), while XGBoost
+took the longest (155.104 hours, std ± 0.959). We also found that
+depending on how the data is split to create a held-out test set, the
+AUROC values of a ML model can vary up to 0.295 which highlighted the
+importance of performing many randomized modeling runs. This study
+established standards for microbiome-associated ML models and
+underscored the importance of model selection.
 
@@
 
-We explained the feature importances in non-linear models, SVM with
-radial basis kernel, decision tree, random forest and XGBoost, using
-permutation importance on the held-out test data where we randomly
-permuted non-correlated features individually and groups of correlated
-features together (see methods) to calculate the effect of permuted OTUs
-or group of OTUs on testing AUROC. The top 5 OTUs with the largest
+We explained the feature importances in non-linear models; SVM with
+radial basis kernel, decision tree, random forest and XGBoost, using a
+metod called permutation importance on the held-out test data.
+Permutation importance analysis is where we randomly permute
+non-correlated features individually and groups of highly correlated
+features together and calculate how much the predictive performance of
+the model (i.e AUROC values) decrease if those specific OTUs or group of
+OTUs are permuted randomly. We ranked the OTUs based on how much they
+decreased the test AUROC values if they were randomly permuted, with the
+largest decrease ranking highest. The top 5 OTUs with the largest
 negative impact on testing AUROC overlapped in tree-based models
@@
 forest and XGBoost models' predictive performance dropped from 0.6 base
-testing AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
+testing median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
 respectively {[}Figure 4{]}.
 
-We used permutation importance to interpret linear models as well to
-highlight the differences between the two interpretation methods
-{[}Figure S3{]}. The rankings of OTUs that discriminate between having
+To highlight the differences between the two interpretation methods, we
+used permutation importance to interpret linear models as well {[}Figure
+S3{]}. The importance rankings of OTUs that discriminate between having
 an SRN or not, changed between different interpretation methods.
 L1-regularized SVM with linear kernel picked out some of the same OTUs
-(OTU00822, OTU01239, OTU00609) as important in feature rankings
-{[}Figure 3{]} and permutation importance methods {[}Figure S3{]} but
-with different ranks. The difference was more pronounced in
-L2-regularized models as they had mostly different OTUs picked out as
-due to collinearity in microbial communities.
+(OTU00822, OTU01239, OTU00609) as important in making a prediction in
+the two interpretation methods; feature rankings based on weights
+{[}Figure 3{]} and permutation importance {[}Figure S3{]}. Similarly,
+L2-regularized SVM and L2-regularized logistic regression picked out
+some of the same OTUs in both interpretation models, OTU00659 and
+OTU00012, respectively. However, for all the linear models, the rankings
+of these features were different due to the collinearity in microbial
+communities.
 
@@
 
-As the complexity of a ML model and the number of tuned hyperparameter
-settings increased {[}Table S1{]}, its training times increased as well
-{[}Figure 5{]}. Linear models trained faster than non-linear models. L1
-and L2 SVM with linear kernel and L2 logistic regression had training
-times of 0.2 hours, (std ± 0.03), 0.2 hours, (std ± 0.02), and 0.2
-hours, (std ± 0.02), respectively. Whereas, a decision tree, SVM with
-radial basis function kernel, random forest and xgboost had training
-times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ± 8.8), 83.2 hours,
-(std ± 11.3) and 155.1 hours, (std ± 1), respectively {[}Figure 5{]}.
+As the complexity of a ML model {[}Table S1{]} and the number of tuned
+hyperparameter settings increased {[}Figures S1-S2{]}, its training
+times increased as well {[}Figure 5{]}. Linear models trained faster
+than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic
+regression had training times of 0.2 hours, (std ± 0.03), 0.2 hours,
+(std ± 0.02), and 0.2 hours, (std ± 0.02), respectively. Whereas, a
+decision tree, SVM with radial basis function kernel, random forest and
+xgboost had training times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ±
+8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1), respectively
+{[}Figure 5{]}.
 
@@
 used a held-out test set to illustrate the difference between
-cross-validation and testing AUROC values. When the differences between
+cross-validation and testing AUROC values. When the difference between
 cross-validation and test performance is low, this suggets the models
 are not overfit and that they will perform similar with similar data. In
-all seven models, the differences in cross-validation and testing AUROC
-values did not exceed 0.021 which suggests that these models are
-generalizable and can be used to test similar new data. Second, we
-performed the initial 80\%-20\% random datasplit 100 times in our ML
-pipeline. The randomization of the initial data-split to create a
+all seven models, the difference between median cross-validation and
+testing AUROC values did not exceed 0.021 which suggests that these
+models are generalizable and can be used to test similar new data.
+Second, we performed the initial 80\%-20\% random datasplit 100 times in
+our ML pipeline. The randomization of the initial data-split to create a
 held-out test set is a crucial step in the ML pipeline to develop robust
 ML models and to report reliable performance metrics for a ML model.
-Depending on the data-split, there is the chance of being overoptimistic
-about the predictive performance of a model. In our study, we showed
-that there was variability in AUROC values between different random
-data-splits in each of the models we tested. Our results showed that the
-testing AUROC values varied 0.23 on average, depending on the
-data-split. Third, we used AUROC metric instead of accuracy in our study
-to evaluate predictive performance of the ML models. AUROC is always
-random at the value 0.5 and is a robust metric when a dataset is
-imbalanced. We also performed a full grid search for hyperparameter
-settings when building a ML model. It is not enough to use the default
-hyperparameter settings when using previously developed ML packages in
-programming languages such as R, Python and Matlab. In the example of
-L1-regularized SVM with linear kernel (Figure S1), the model showed
+Depending on how the data is split, there is the chance of being
+overoptimistic about the predictive performance of a model. In our
+study, we showed that there was variability in AUROC values between
+different random data-splits in each of the models we tested. Our
+results showed that the testing AUROC values varied 0.23 on average
+between different data-splits. Third, we used the AUROC metric instead
+of accuracy in our study to evaluate the predictive performance of the
+ML models. AUROC is always random at the value 0.5 and is a robust
+metric when a dataset is imbalanced. We also performed a full grid
+search for hyperparameter settings when building a ML model. Default
+hyperparameter settings in previously developed ML packages in R,
+Python, and Matlab programming languages are inadequate for effective
+application of classification algorithms and need to be optimized for
+each new dataset used to generate a model. In the example of
+L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
 large variability between different regularization coefficients (C) and
@@
 interpretibility and computational burden. Microbiome studies use ML
-models with the classification task to learn the training data to assign
-labels, such as healthy or not, to new data (e.g.~new samples), but also
-to learn which features are important enough to discriminate between
-labels. If the goal of a study is not just to predict a health outcome
-but also to learn the ecology behind a disease and to identify microbial
-biomarkers of a disease, then interpretibility of a ML model is crucial.
-In terms of predictive performance, random forest model had testing
-AUROC values statistically significantly larger than others. However the
-second best model was L2-regularized logistic regression following
-random forest with a median AUROC difference of only 0.015. In terms of
-interpretability, random forest was a more complex ML model and it could
-only be explained using methods such as permutation importance. On the
-other hand, L2-regularized logistic regression was easier to interpret
+models with a classification task to learn the training data to assign
+labels, such as healthy or not, to new data, but also to learn which
+features are important to discriminate between labels. If the goal of a
+study is not just to predict a health outcome but also to learn the
+ecology behind a disease and to identify microbial biomarkers of a
+disease, then the ML model has to be interpretable. In terms of
+predictive performance, random forest model had testing AUROC values
+statistically significantly larger than others. However the second best
+model was L2-regularized logistic regression following random forest
+with a median AUROC difference of only 0.015. In terms of
+interpretability, random forest was a complex ML model and it could only
+be explained using methods such as permutation importance. On the other
+hand, L2-regularized logistic regression was easier to interpret
 (i.e.~ranking absolute regression coefficients of the trained model).
 
-Even with interpretable models, we need to be careful with how we treat
-the information we get from the models. In this study we used two
-methods to interpret our linear models; absolute feature weights of
-trained models and permutation importance. We observed differences in
-outcome between the two interpretation methods due to collinearity in
-the dataset. Collinearity in a microbial dataset is when one OTU is
-highly correlated with another OTU which causes regression coefficients
-to not be unique. The regression coefficients of correlated OTUs are
-influenced by one another which makes it difficult to interpret a ML
-model. There are ways to make it easier to interpret models.; (1) use
-different models for prediction and interpretation in which highly
-correlated OTUs are removed, (2) get the highly ranked OTUs and check
-what othe OTUs they are correlated with. The interpretation however
-should not be a final answer but should be used to generate hypothesis
-that are tested with followup experiments.
+Even with interpretable models such as L2-regularized logistic
+regression, we need to be careful with how we treat the information we
+get from the models. In this study we used two different methods to
+interpret our linear models; ranking each OTU by (1) their absolute
+weights in the trained models and (2) their impact on the predictive
+performance based on permutation importance. We observed differences in
+the OTU rankings between the two interpretation methods due to
+collinearity in the dataset. Collinearity in a microbial dataset is when
+one OTU is highly correlated with another OTU which causes feature
+weights to not be unique. The feature weights of correlated OTUs are
+influenced by one another which makes it difficult to interpret the ML
+model. To avoid misinterpreting the models we can; (1) use ML models
+that has highly correlated OTUs removed, (2) use the highly ranked OTUs
+that are correlated with other OTUs to generate hypotheses about the
+ecology of the disease and test them with followup experiments.
 
 There are other criteria when choosing ML models such as the
-computational burden of developing it, sample size and balance of the
-dataset. In terms of computation burden, random forest model was slow to
-train 83.2 hours whereas L2-regularized logistic regression trained in
-12 minutes. The generalization performance of ML models depends on
-sample size. The more complex the model, the more data it will need. The
+computational burden of developing it and sample size. In terms of
+computational burden, random forest model trained each data-split in
+83.2 hours whereas L2-regularized logistic regression trained in 12
+minutes. The generalization performance of ML models depends on sample
+size. The more complex the model, the more data it will need. The
 dataset we used for our study had 490 samples, however microbiome
@@
 complex models such as L2-regularized logistic regression.
+
+This study highlights the need to make educated choices at every step of
+developing a ML model with microbiome data. Model selection, setting up
+a rigorous ML pipeline and choosing the right interpretation method
+contribute to achieving the level of validity and accountability we want
+from models built for patient health.
 
@@
 patients recruited by Great Lakes-New England Early Detection Research
-Network study. These data were obtained from Sze et al (25). The stool
+Network study. These data were obtained from Sze et al (27). The stool
 samples were provided by recruited adult participants who were
@@
 assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(26--28).
+(28--30).
 
@@
 testing performance. We can use hyper-band to help us with our
-hyperparameter selection (29).
+hyperparameter selection (31).
 
@@
 support vector machine; OTU, Operational Taxonomic Unit. \newpage
-\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
-non-linear ML models.} (A) SVM with radial basis kernel (B) decision
-tree (C) random forest (D) XGboost feaure importances were explanied
-using permutation importance using held-out test set. The gray rectangle
-and the dashed line show the IQR range and median of the base testing
-AUROC without any permutation performed. The colors of the boxplots
-stand for the unique OTUs that are shared among the different models;
-pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for
-OTU00110, green for OTU00361 and red for OTU00882. For all the
-tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
-the largest impact on predictive performance of the model.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
-Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+\includegraphics{Figure_4.png}
+
+\textbf{Figure 4. Explanation of the non-linear ML models.} (A) SVM with
+radial basis kernel (B) decision tree (C) random forest (D) XGboost
+feaure importances were explanied using permutation importance using
+held-out test set. The gray rectangle and the dashed line show the IQR
+range and median of the base testing AUROC without any permutation
+performed. The colors of the boxplots stand for the unique OTUs that are
+shared among the different models; pink for OTU0008, salmon for OTU0050,
+yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for
+OTU00882. For all the tree-based models, a \emph{Peptostreptococcus}
+species (OTU00367) had the largest impact on predictive performance of
+the model. Abbreviations: SVM, support vector machine; OTU, Operational
+Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic
+Unit. \newpage
+\includegraphics{Figure_5.png}
+
+\textbf{Figure 5. Computational efficiency of seven ML models.} The
+training times for of each data-split showed the differences in
+computational efficieny of the seven models. The median training time in
+hours was the highest for XGBoost and shortest for L1-regularized SVM
+with linear kernel. The boxplot shows quartiles at the box ends and the
+statistical median as the horizontal line in the box. The whiskers show
+the farthest points that are not outliers. Outliers are data points that
+are not within 3/2 times the interquartile ranges. Abbreviations: AUROC,
+area under the receiver operating characteristic curve; SVM, support
+vector machine; XGBoost, extreme gradient boosting.\\
 \newpage
-\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
-efficiency of seven ML models.} The training times for of each
-data-split showed the differences in computational efficieny of the
-seven models. The median training time in hours was the highest for
-XGBoost and shortest for L1-regularized SVM with linear kernel. The
-boxplot shows quartiles at the box ends and the statistical median as
-the horizontal line in the box. The whiskers show the farthest points
-that are not outliers. Outliers are data points that are not within 3/2
-times the interquartile ranges. Abbreviations: AUROC, area under the
-receiver operating characteristic curve; SVM, support vector machine;
-XGBoost, extreme gradient boosting.\\
-\newpage
-\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
-setting performances for linear models.} (A) L2 logistic regression (B)
-L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
-cross-validation AUROC values when different hyperparameters are used in
-training the model. The differences in AUROC values when hyperparameters
-change show that hyperparameter tuning is a crucial step in building a
-ML model.
+\includegraphics{Figure_S1.png}
+
+\textbf{Figure S1. Hyperparameter setting performances for linear
+models.} (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2
+SVM with linear kernel mean cross-validation AUROC values when different
+hyperparameters are used in training the model. The differences in AUROC
+values when hyperparameters change show that hyperparameter tuning is a
+crucial step in building a ML model.
 
@@
 
-\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
-setting performances for non-linear models.} (A) Decision tree (B)
-Random forest (C) SVM with radial basis kernel (D) XGoost mean
-cross-validation AUROC values when different hyperparameters are used in
-training the model. The differences in AUROC values when hyperparameters
-change show that hyperparameter tuning is a crucial step in building a
-ML model. \newpage
-\includegraphics{Figure_S4.png} \textbf{Figure S3. Interpretation of the
-linear ML models with permutation importance.} (A) L1-regularized SVM
-with linear kernel (B) L2-regularized SVM with linear kernel and (C)
-L2-regularized logistic regression were interpreted using permutation
-importance using held-out test set. The gray rectangle and the dashed
-line show the IQR range and median of the base testing AUROC without any
-permutation performed. Abbreviations: SVM, support vector machine; OTU,
-Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational
-Taxonomic Unit. \includegraphics{TableS1.pdf} \textbf{Table S1. ML model
-complexity and hyperparameter range.} \newpage
+\includegraphics{Figure_S2.png}
+
+\textbf{Figure S2. Hyperparameter setting performances for non-linear
+models.} (A) Decision tree (B) Random forest (C) SVM with radial basis
+kernel (D) XGoost mean cross-validation AUROC values when different
+hyperparameters are used in training the model. The differences in AUROC
+values when hyperparameters change show that hyperparameter tuning is a
+crucial step in building a ML model. \newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
+
+\textbf{Figure S3. Interpretation of the linear ML models with
+permutation importance.} (A) L1-regularized SVM with linear kernel (B)
+L2-regularized SVM with linear kernel and (C) L2-regularized logistic
+regression were interpreted using permutation importance using held-out
+test set. The gray rectangle and the dashed line show the IQR range and
+median of the base testing AUROC without any permutation performed.
+Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
+Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+
+\newpage
+
+\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
+hyperparameter range.} \newpage
 
@@
 
+\hypertarget{ref-rudin_optimized_2018}{}
+25. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
+Toward trust in machine learning for healthcare and criminal justice.
+Interfaces \textbf{48}:449--466.
+doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.
+
+\hypertarget{ref-knights_human-associated_2011}{}
+26. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
+\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
+signatures: Examining their predictive value. Cell Host Microbe
+\textbf{10}:292--296.
+doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.
+
 \hypertarget{ref-sze_leveraging_2018}{}
-25. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+27. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-26. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-27. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+29. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-28. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+30. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-29. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+31. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-07-10 code/learning/FigureS4.R

@@ -160,8 +160,7 @@ base_nonlin_plot <-  function(data, name){
 
 # --------------------- Linear models ----------------------------------->
 logit_plot <- base_nonlin_plot(logit, "L2_Logistic_Regression") +
-  scale_x_discrete(name = "L2 Logistic 
-Regression ") 
+  scale_x_discrete(name = "L2 Logistic Regression ") 
 
 l1_plot <- base_nonlin_plot(l1svm, "L1_Linear_SVM") +
   scale_x_discrete(name = "L1 Linear SVM") 

### 2019-07-11 code/learning/Figure3.R
fix makefile

@@ -154,4 +154,4 @@ linear <- plot_grid(l1_svm_graph, l2_svm_graph, logit_graph, labels = c("A", "B"
 
 ggdraw(add_sub(linear, "Feature Ranks", vpadding=grid::unit(0,"lines"), y=5, x=0.7, vjust=4.75, size=15))
 
-ggsave("Figure_3.png", plot = last_plot(), device = 'png', path = 'submission', width = 6, height = 8.5)
+ggsave("Figure_3.png", plot = last_plot(), device = 'png', path = 'submission', width = 6, height = 9.2)

### 2019-07-12 submission/manuscript.Rmd
fix makefile
update

@@
 We reported the testing AUROC values over 100 randomizations of the initial data-split.
-The results showed that depending on the data-split, the testing AUROC value showed great variability [Figure 2].
+The results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
 The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
@@
 
-We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a metod called permutation importance on the held-out test data.
-Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together and calculate how much the predictive performance of the model (i.e AUROC values) decrease if those specific OTUs or group of OTUs are permuted randomly.
-We ranked the OTUs based on how much they decreased the test AUROC values if they were randomly permuted, with the largest decrease ranking highest.
+We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.
+Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together.
+We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly.
+We ranked the OTUs based on how much they decreased the median testing AUROC; the OTU with the largest decrease ranking highest.
 The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
-Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure S3].
-Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
+Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure 4].
+Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
 
 To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3].
-The importance rankings of OTUs that discriminate between having an SRN or not, changed between different interpretation methods.
-L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in making a prediction in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3].
-Similarly,  L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation models, OTU00659 and OTU00012, respectively.
+L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3].
+Similarly,  L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively.
 However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.
@@
 \newpage
-\includegraphics[height=15cm, width=10cm]{Figure_3.png}
+\includegraphics[height=17cm, width=12cm]{Figure_3.png}
 
@@
 \newpage
-\includegraphics{Figure_S2.png}
+\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
 
@@
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
+\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
 

### 2019-07-12 submission/manuscript.tex
add comment on importance
fix makefile
update

@@
 
-\subsection{Importance}\label{importance}
+\subsection{Importance (needs work)}\label{importance-needs-work}
 
@@
 initial data-split. The results showed that depending on the data-split,
-the testing AUROC value showed great variability {[}Figure 2{]}. The
+the testing AUROC values showed great variability {[}Figure 2{]}. The
 testing AUROC values within each model varied 0.23 on average across the
@@
 radial basis kernel, decision tree, random forest and XGBoost, using a
-metod called permutation importance on the held-out test data.
+method called permutation importance on the held-out test data.
 Permutation importance analysis is where we randomly permute
 non-correlated features individually and groups of highly correlated
-features together and calculate how much the predictive performance of
-the model (i.e AUROC values) decrease if those specific OTUs or group of
-OTUs are permuted randomly. We ranked the OTUs based on how much they
-decreased the test AUROC values if they were randomly permuted, with the
-largest decrease ranking highest. The top 5 OTUs with the largest
-negative impact on testing AUROC overlapped in tree-based models
-{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
-(OTU00367) abundances randomly, dropped the predictive performances the
-most in all tree-based methods {[}Figure S3{]}. Decision tree, random
-forest and XGBoost models' predictive performance dropped from 0.6 base
-testing median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
-respectively {[}Figure 4{]}.
+features together. We then calculate how much the predictive performance
+of the model (i.e AUROC values) decrease when each OTU or group of OTUs
+is permuted randomly. We ranked the OTUs based on how much they
+decreased the median testing AUROC; the OTU with the largest decrease
+ranking highest. The top 5 OTUs with the largest negative impact on
+testing AUROC overlapped in tree-based models {[}Figure 4{]}.
+Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
+randomly, dropped the predictive performances the most in all tree-based
+methods {[}Figure 4{]}. Decision tree, random forest and XGBoost models'
+predictive performance dropped from 0.6 median AUROC to 0.52, from 0.69
+to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.
 
@@
 used permutation importance to interpret linear models as well {[}Figure
-S3{]}. The importance rankings of OTUs that discriminate between having
-an SRN or not, changed between different interpretation methods.
-L1-regularized SVM with linear kernel picked out some of the same OTUs
-(OTU00822, OTU01239, OTU00609) as important in making a prediction in
-the two interpretation methods; feature rankings based on weights
-{[}Figure 3{]} and permutation importance {[}Figure S3{]}. Similarly,
-L2-regularized SVM and L2-regularized logistic regression picked out
-some of the same OTUs in both interpretation models, OTU00659 and
-OTU00012, respectively. However, for all the linear models, the rankings
-of these features were different due to the collinearity in microbial
+S3{]}. L1-regularized SVM with linear kernel picked out some of the same
+OTUs (OTU00822, OTU01239, OTU00609) as important in the two
+interpretation methods; feature rankings based on weights {[}Figure 3{]}
+and permutation importance {[}Figure S3{]}. Similarly, L2-regularized
+SVM and L2-regularized logistic regression picked out some of the same
+OTUs in both interpretation methods, OTU00659 and OTU00012,
+respectively. However, for all the linear models, the rankings of these
+features were different due to the collinearity in microbial
 communities.
@@
 
-\includegraphics{Figure_S2.png}
+\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
 
@@
 crucial step in building a ML model. \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
+\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
 

### 2019-07-15 submission/manuscript.Rmd
Sent to Pat

@@
 ---
-title: '**Evaluation of machine learning methods for 16S rRNA gene data**'
+title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
 csl: mbio.csl
@@
 
-Begüm D. Topçuoğlu${^1}$, Nick Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^
+Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^
 
@@
 Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers.
-To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
-To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
+To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
+To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
 The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
 Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
-In this study, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.
-Additionally, we established standards for the development of modeling pipelines for microbiome-associated ML models
+In this study, we established standards for the development of modeling pipelines for microbiome-associated ML models.
+Additionally, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.
 
@@
 
-Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies.
+Prediction of health outcomes using ML is rapidly being adopted by human microbiome studies.
 However, the developed ML models so far are overoptimistic in terms of validity and predictive performance.
 Without rigorous ML pipelines, we cannot trust ML models.
-Before we can speed up progress, we need to slow down, define and start using good ML practices.
+Before we can speed up progress, we need to slow down, define and implement good ML practices.
 \newpage
@@
 
-Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications.
-Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health.
+Advances in sequencing technology and decreases in the costs of generating 16S rRNA gene sequences have allowed rapid exploration of the human microbiome and its health implications.
+Currently, the human microbiome field is growing at an unprecedented rate and there is an increasing demand for methods that identify associations between microbiome members and human health.
 However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease.
@@
 They learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction.
-Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+The high-stake task of predicting the correct diagnosis with high confidence relies on a ML model that is built with rigorous methods.
 However, the field's use of ML lacks clarity and consistency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019].
-More notably, we commonly see flawed ML practices such as (1) using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or (2) reporting few or only the best outcomes of cross-validation.
-Even when there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+More notably, we commonly see flawed ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate predictive performance or reporting few or only the best outcomes of cross-validation.
+Even when there are seperate testing sets to evaluate predictive performance, there are large differences between cross-validation and testing performances and large confidence intervals in testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
+These practices prevent the development of generalizable models, where the model makes accurate predictions with newly acquired data as well as it does with the training data.
 
-Moreover, there is a lack of discussion on why a particular ML model is utilized.
-Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-Black box machine learning models are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction.
+Moreover, there is a lack of discussion on why a particular ML model is utilized and a lack of emphasis on the strengths and weaknesses of it.
+Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
+These models are also called black box ML models because they are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction.
 These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018].
@@
 
-The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliy.
-We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
-We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of `r paste(nrow(shared))` patients.
-The human-associated human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
+The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibility.
+We need to strive toward better machine learning practices by (1) implementing rigorous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of predictive performance, complexity, interpretability and computational efficiency.
+To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
+We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of CRC patients.
+The human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
 We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
-The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
-We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
-We used the area under the receiver operating characteristic curve (AUROC) as a metric for determining the predictibve performance of each model.
-The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance.
-The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`.
-Random forest had the highest median AUROC for detecting SRN.
-Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.
-In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`).
-We also found that depending on how the data is split to create a held-out test set, the AUROC values of a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs.
-This study established standards for microbiome-associated ML models and underscored the importance of model selection.
+We established modeling pipelines for three linear models with different forms of regularization; L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear kernel.
+We also developed four non-linear models; SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
+We compared the predictive performance, interpretability and computational efficiency of the seven models to highlight the importance of model selection.
+We established standards for ML pipeline construction, predictive performance evaluation and model interpretation for microbiome-associated ML models that have a binary prediction task.
 
@@
 
-We used a cohort of 490 patients with 261 cases of SRN.
-For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (having colorectal lesions that are identified as SRN or normal).
-All the cases were independently labeled through colonoscopies.
-We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost) to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
+We first determined the dataset and the ML models to use in our study.
+We used a previously published study on a CRC cohort of 490 patients with 261 cases of SRN.
+For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (normal or SRN colorectal lesions as defined by colonoscopies).
+We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost).
 
-For regularized logistic regression and SVM with linear kernel we used L2 regularization to keep all potentially important features.
+We used ML models with different classification algorithms and regularization parameters.
+For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features.
 For comparison, we also trained an L1 regularized SVM model with linear kernel.
 L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero).
-Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.
+Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGBoost, as well as an SVM with non-linear kernel.
 
@@
 Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
-The training/validation data was used for training purposes and validation of hyperparameter selection (i.e.
-model selection), and the test set was used for evaluation purposes.
-Validation of hyperparameter selection was performed using repeated five-fold cross-validation on the training/validation set [Figure 1].
-Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets.
-We validated the cross-validation performances of each hyperparameter setting over 100 randomizations and selected the best performing hyperparameter setting in terms of AUROC metric and trained the full training/validation dataset [Figures S1 and S2].
-We then used the held-out test set to evaluate the prediction performance of each ML model.
+The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
+
+Hyperparameters are the rules that are learned from the training data in a classification algorithm.
+When they are tuned over a full grid search and selected by validation, they make the ML model better are predicting.
+We selected hyperparameter settings using repeated five-fold cross-validation on the training/validation set [Figure 1].
+We chose the best hyperparameter setting for each model based on its predictive performance on the validation set using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2].
+The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance.
+Similar to the initial data-split, five-fold cross-validation for hyperparameter selection was also stratified to maintain the overall label distribution on the training and validation sets.
+The cross-validation of each hyperparameter setting was repeated over 100 randomizations to get a robust reading of predictive performance.
+
+We then trained the full training/validation dataset with the selected hyperparameters.
+We used the held-out test set to evaluate the predictive performance of each ML model.
 The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].
 
-__Discriminative performance and generalizability of the seven models.__
+__Predictive performance and generalizability of the seven models.__
 
-We evaluated the predictive performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 2].
+We evaluated the predictive performances of seven binary classification models when applied to held-out test data using the AUROC metric [Figure 2].
 Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01).
@@
 They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively.
-L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
+L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
+Random forest had the highest median AUROC for detecting SRN.
+Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.
 
-For each model, we compared the median cross-validation AUROC to the median testing AUROC.
+To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC.
 The difference between the two should be low to suggest the model is not overfitting despite the large number of features.
-The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel,  followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
-
-We reported the testing AUROC values over 100 randomizations of the initial data-split.
-The results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
+The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
+We also reported the testing AUROC values over 100 randomizations of the initial data-split.
 The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
 For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.
+These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
 
@@
 
-The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity.
-We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the absolute feature weights of the trained models.
+Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017].
+Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies.
+The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretability as they increase in complexity.
+In this study we highlighted two methods to interpret models with varying complexity.
+
+We interpreted linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the absolute feature weights of the trained models.
 We ranked the absolute weights of all the OTUs for each data-split [Figure 3].
@@
 We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.
-Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together.
+Permutation importance analysis is a posthoc explanation of the model where we randomly permute non-correlated features individually and groups of highly correlated features together.
 We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly.
@@
 To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3].
-L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3].
+L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings based on weights [Figure 3] and permutation importance [Figure S3].
 Similarly, L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively.
@@
 
+We compared the training times of the seven ML models.
 As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5].
 Linear models trained faster than non-linear models.
-L1 and L2 SVM with linear kernel and L2 logistic regression had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
-Whereas, a decision tree, SVM with radial basis function kernel, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
+L1 and L2 SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
+Whereas, a decision tree, SVM with radial basis function kernel, random forest and XGBoost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
 
@@
 In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
-We set-up standards for developing and evaluating ML models for microbiome data.
+We built on others' work in the microbiome field to set-up standards for developing and evaluating ML models for microbiome data [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019].
 First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
-When the difference between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data.
+When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data.
 In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data.
 Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
-The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics for a ML model.
+The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics.
 Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model.
@@
 Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits.
-Third, we used the AUROC metric instead of accuracy in our study to evaluate the predictive performance of the ML models.
+Third, we used the AUROC metric in our study to evaluate the predictive performance of the ML models.
 AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
@@
 
-In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden.
-Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not, to new data, but also to learn which features are important to discriminate between labels.
-If the goal of a study is not just to predict a health outcome but also to learn the ecology behind a disease and to identify microbial biomarkers of a disease, then the ML model has to be interpretable.
-In terms of predictive performance, random forest model had testing AUROC values statistically significantly larger than others.
-However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015.
-In terms of interpretability, random forest was a complex ML model and it could only be explained using methods such as permutation importance.
+In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretability and computational burden.
+Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not to new data, but also to learn which features are important to discriminate between labels [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+Our results show that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we could create ML models that are inherently interpretable without losing predictive power.
+In terms of predictive performance, random forest model had the best testing AUROC values compared to the other 6 models.
+However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
+In terms of interpretability, random forest was a complex ML model and could only be explained using methods such as permutation importance.
 On the other hand, L2-regularized logistic regression was easier to interpret (i.e.
@@
 The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model.
-To avoid misinterpreting the models we can; (1) use ML models that has highly correlated OTUs removed, (2) use the highly ranked OTUs that are correlated with other OTUs to generate hypotheses about the ecology of the disease and test them with followup experiments.
+To avoid misinterpreting the models we should use the highly ranked correlated OTUs to generate hypotheses about the ecology of the disease and test them with follow-up experiments.
 
-There are other criteria when choosing ML models such as the computational burden of developing it and sample size.
+There are other criteria when choosing ML models such as the computational burden of developing it and the sample size.
 In terms of computational burden, random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
@@
 This study highlights the need to make educated choices at every step of developing a ML model with microbiome data.
-Model selection, setting up a rigorous ML pipeline and choosing the right interpretation method contribute to achieving the level of validity and accountability we want from models built for patient health.
+Model selection should be done with a solid understanding of model complexity and interpretability, rigorous ML pipelines should be built with cross-validation for hyperparameter tuning and with a held-out test set for evaluating predictive performance and models should be interpreted while considering collinearity in datasets.
+The right methods will help us achieve the level of validity and accountability we want from models built for patient health.
 
@@
 Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas.
+The study had 261 normal and 229 SRN samples.
 The bacterial abundances are the features used to predict colonic health of the patients.
@@
 OTU counts were set to the size of our smallest sample and were subsampled at the same distances.
-They were then transformd by scaling to a [0-1] range.
+They were then transformed by scaling to a [0-1] range.
 
@@
 We created a Spearman's rank-order correlation matrix, corrected for multiple pairwise comparisons.
-We then defined correlated OTUs as having perfect correlation (correlation coef=1 and p<0.01).Other OTUs were permuted individually to get permutation importance but the correlated ones are grouped together and permuted at the same time.
-The reported OTUs that have impact on testing AUROC are reported in Figures xxx.
-If we want we can decrease the correlation coefficient to consider OTUs that are correlated with ecological consequences but this was out of the scope of this study but will be followed up in further analyses.
+We then defined correlated OTUs as having perfect correlation (correlation coef=1 and p<0.01).
+Non-correlated OTUs were permuted individually whereas correlated ones were grouped together and permuted at the same time.
 
 **Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1).
-We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best discriminative performance.
+We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best predictive performance.
 
@@
 The model was evaluated on a held-out set of data (not used in selecting the model).
-Abbreviations: AUROC, area under the receiver operating characteristic curve
+Abbreviations: cvAUROC, cross-validation area under the receiver operating characteristic curve
 \newpage
@@
 **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-Discriminative performance of random forest model was higher than other ML models.
+Predictive performance of random forest model was higher than other ML models.
 The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
@@
 **Figure 4.
-Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set.
+Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGBoost feature importances were explained using permutation importance using held-out test set.
 The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
-The colors of the boxplots stand for the unique OTUs that are shared among the different models; pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for OTU00882.
+The colors of the box plots stand for the unique OTUs that are shared among the different models; pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for OTU00882.
 For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model.
@@
 **Figure 5.
-Computational efficiency of seven ML models.** The training times for of each data-split showed the differences in computational efficieny of the seven models.
+Computational efficiency of seven ML models.** The training times for of each data-split showed the differences in computational efficiency of the seven models.
 The median training time in hours was the highest for XGBoost and shortest for L1-regularized SVM with linear kernel.
@@
 **Figure S2.
-Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGoost mean cross-validation AUROC values when different hyperparameters are used in training the model.
+Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGBoost mean cross-validation AUROC values when different hyperparameters are used in training the model.
 The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.

### 2019-07-15 submission/manuscript.tex
Sent to Pat

@@
 \hypersetup{unicode=true,
-            pdftitle={Evaluation of machine learning methods for 16S rRNA gene data},
+            pdftitle={Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Evaluation of machine learning methods for 16S rRNA gene data}}
+  \title{\textbf{Best practices for applying machine learning to bacterial 16S
+rRNA gene sequencing data}}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-Begüm D. Topçuoğlu\({^1}\), Nick Lesniak\({^1}\), Jenna Wiens\({^2}\),
-Mack Ruffin\({^3}\), Patrick D. Schloss\textsuperscript{1\(\dagger\)}
+Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Jenna
+Wiens\({^2}\), Mack Ruffin\({^3}\), Patrick D.
+Schloss\textsuperscript{1\(\dagger\)}
 
@@
 these challenges, we benchmarked seven different ML models that use
-fecal 16S rRNA sequences to predict the presence/absence of colorectal
-cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To
-show the effect of model selection, we assessed the predictive
-performance, interpretability, and computational efficiency of the
-following models: L2-regularized logistic regression, L1 and L2 support
-vector machines (SVM) with linear and radial basis function kernels, a
-decision tree, random forest, and extreme gradient boosting (XGBoost).
-The random forest model was best at detecting CRC lesions with an AUROC
-of 0.695 but it was slow to train (83.2 h) and hard to interpret.
-Despite its simplicity, L2-regularized logistic regression followed
-random forest in predictive performance with an AUROC of 0.680, and it
-trained much faster (12 min). In this study, we showed that ML models
-should be chosen based on expectations of predictive performance,
-interpretability and available computational resources. Additionally, we
-established standards for the development of modeling pipelines for
-microbiome-associated ML models
+fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions
+(n=490 patients, 261 controls and 229 cases). To show the effect of
+model selection, we assessed the predictive performance,
+interpretability, and computational efficiency of the following models:
+L2-regularized logistic regression, L1 and L2-regularized support vector
+machines (SVM) with linear and radial basis function kernels, a decision
+tree, random forest, and extreme gradient boosting (XGBoost). The random
+forest model was best at detecting CRC lesions with an AUROC of 0.695
+but it was slow to train (83.2 h) and hard to interpret. Despite its
+simplicity, L2-regularized logistic regression followed random forest in
+predictive performance with an AUROC of 0.680, and it trained much
+faster (12 min). In this study, we established standards for the
+development of modeling pipelines for microbiome-associated ML models.
+Additionally, we showed that ML models should be chosen based on
+expectations of predictive performance, interpretability and available
+computational resources.
 
@@
 
-Prediction of health outcomes using ML is rapidly being adopted by
-human-associated microbiome studies. However, the developed ML models so
-far are overoptimistic in terms of validity and predictive performance.
-Without rigorous ML pipelines, we cannot trust ML models. Before we can
-speed up progress, we need to slow down, define and start using good ML
-practices. \newpage
+Prediction of health outcomes using ML is rapidly being adopted by human
+microbiome studies. However, the developed ML models so far are
+overoptimistic in terms of validity and predictive performance. Without
+rigorous ML pipelines, we cannot trust ML models. Before we can speed up
+progress, we need to slow down, define and implement good ML practices.
+\newpage
 
@@
 
-Advances in sequencing technology and decreasing costs of generating 16S
-rRNA gene sequences have allowed rapid exploration of human associated
-microbiome and its health implications. Currently, the human microbiome
-field is growing at an unprecedented rate and as a result, there is an
-increasing demand for methods that identify associations between members
-of the microbiome and human health. However, this is difficult as human
+Advances in sequencing technology and decreases in the costs of
+generating 16S rRNA gene sequences have allowed rapid exploration of the
+human microbiome and its health implications. Currently, the human
+microbiome field is growing at an unprecedented rate and there is an
+increasing demand for methods that identify associations between
+microbiome members and human health. However, this is difficult as human
 associated microbial communities are remarkably complex,
@@
 ecological basis of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
-(1--11). However, the field's use of ML lacks clarity and consistency in
-which methods are used and how these methods are implemented (12, 13).
-More notably, we commonly see flawed ML practices such as (1) using ML
-pipelines where there is no seperate held-out test dataset to evaluate
-model performance or (2) reporting few or only the best outcomes of
+inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
+The high-stake task of predicting the correct diagnosis with high
+confidence relies on a ML model that is built with rigorous methods.
+However, the field's use of ML lacks clarity and consistency in which
+methods are used and how these methods are implemented (12, 13). More
+notably, we commonly see flawed ML practices such as using ML pipelines
+where there is no seperate held-out test dataset to evaluate predictive
+performance or reporting few or only the best outcomes of
 cross-validation. Even when there are seperate testing sets to evaluate
-model performance, there are large differences between cross-validation
-and testing performances that indicate overfitting as well as large
-confidence intervals for testing performances (4, 14--20).
+predictive performance, there are large differences between
+cross-validation and testing performances and large confidence intervals
+in testing performances (4, 14--20). These practices prevent the
+development of generalizable models, where the model makes accurate
+predictions with newly acquired data as well as it does with the
+training data.
 
 Moreover, there is a lack of discussion on why a particular ML model is
-utilized. Recently, there is a trend towards using more complex ML
-models such as random forest, extreme gradient boosting and neural
-networks without a discussion on if and how much model interpretibility
-is necessary for the study (11, 21--23). Black box machine learning
-models are not inherently interpretable and require posthoc explanations
-to determine the feature importances in making a prediction. These
-explanations can be misleading and at times unreliable when making
-high-stake decisions about patient health (24). The models we develop
-for healthcare, both to predict disease and to understand underlying
-reasons behind that prediction, should be transparent and accountable
-(25).
+utilized and a lack of emphasis on the strengths and weaknesses of it.
+Recently, there is a trend towards using more complex ML models such as
+random forest, extreme gradient boosting (XGBoost) and neural networks
+without a discussion on if and how much model interpretibility is
+necessary for the study (11, 21--23). These models are also called black
+box ML models because they are not inherently interpretable and require
+posthoc explanations to determine the feature importances in making a
+prediction. These explanations can be misleading and at times unreliable
+when making high-stake decisions about patient health (24). The models
+we develop for healthcare, both to predict disease and to understand
+underlying reasons behind that prediction, should be transparent and
+accountable (25).
 
@@
 as flawed modeling methods negatively impact model validity and
-reproducibiliy. We need to strive toward better machine learning
-practices by (1) implementing rigourous machine learning pipelines and
+reproducibility. We need to strive toward better machine learning
+practices by (1) implementing rigorous machine learning pipelines and
 (2) selecting ML models that reflect the goal of the study as it will
-inform our expectations of model accuracy, complexity, interpretibility
-and computational efficiency.To showcase a rigorous ML pipeline and to
-shed light on how ML model selection can affect modeling results, we
-performed an empirical analysis comparing several different ML models
-using the same dataset and the same ML pipeline. We used a previously
-published colorectal cancer (CRC) study (3) which had fecal 16S rRNA
-gene sequences of 490 patients. The human-associated human microbiome is
+inform our expectations of predictive performance, complexity,
+interpretability and computational efficiency. To showcase a rigorous ML
+pipeline and to shed light on how ML model selection can affect modeling
+results, we performed an empirical analysis comparing several different
+ML models using the same dataset and the same ML pipeline. We used a
+previously published colorectal cancer (CRC) study (3) which had fecal
+16S rRNA gene sequences of CRC patients. The human microbiome is
 hypothesized to directly contribute to the development of CRC and fecal
@@
 healthy patients versus patients with colorectal lesions that were
-identified by colonoscopy as screen relevant neoplasias (SRN). The study
-had 261 normal and 229 SRN samples. We established modeling pipelines
-for L2-regularized logistic regression, L1 and L2 support vector
-machines (SVM) with linear and radial basis function kernels, a decision
-tree, random forest and XGBoost. We used the area under the receiver
-operating characteristic curve (AUROC) as a metric for determining the
-predictibve performance of each model. The AUROC ranges from 1.0, where
-the model perfectly distinguishes between cases and controls, to 0.50,
-where the model's predictions are no different from random chance. The
-median test AUROC varied from 0.601 to 0.695. Random forest had the
-highest median AUROC for detecting SRN. Despite its simplicity, the
-L2-regularized logistic regression was second best in predictive
-performance. In terms of computational efficiency, L1 SVM with linear
-kernel trained the fastest (0.202 hours, std ± 0.028), while XGBoost
-took the longest (155.104 hours, std ± 0.959). We also found that
-depending on how the data is split to create a held-out test set, the
-AUROC values of a ML model can vary up to 0.295 which highlighted the
-importance of performing many randomized modeling runs. This study
-established standards for microbiome-associated ML models and
-underscored the importance of model selection.
+identified by colonoscopy as screen relevant neoplasias (SRN). We
+established modeling pipelines for three linear models with different
+forms of regularization; L2-regularized logistic regression, L1 and
+L2-regularized support vector machines (SVM) with linear kernel. We also
+developed four non-linear models; SVM with radial basis function kernel,
+a decision tree, random forest and XGBoost. We compared the predictive
+performance, interpretability and computational efficiency of the seven
+models to highlight the importance of model selection. We established
+standards for ML pipeline construction, predictive performance
+evaluation and model interpretation for microbiome-associated ML models
+that have a binary prediction task.
 
@@
 
-We used a cohort of 490 patients with 261 cases of SRN. For each
-patient, we had 6920 features (fecal bacterial abundances) and a
-two-class label that defines their colorectal health (having colorectal
-lesions that are identified as SRN or normal). All the cases were
-independently labeled through colonoscopies. We established modeling
-pipelines for a binary prediction task with L2-regularized logistic
-regression, L1 and L2 support vector machines (SVM) with linear and
-radial basis function kernels, a decision tree, random forest and
-extreme gradient boosted decision tree (XGBoost) to emphasize the
-differences in model accuracy, complexity, interpretibility and
-computational efficiency due to model selection.
+We first determined the dataset and the ML models to use in our study.
+We used a previously published study on a CRC cohort of 490 patients
+with 261 cases of SRN. For each patient, we had 6920 features (fecal
+bacterial abundances) and a two-class label that defines their
+colorectal health (normal or SRN colorectal lesions as defined by
+colonoscopies). We established modeling pipelines for a binary
+prediction task with L2-regularized logistic regression, L1 and L2
+support vector machines (SVM) with linear and radial basis function
+kernels, a decision tree, random forest and extreme gradient boosted
+decision tree (XGBoost).
 
-For regularized logistic regression and SVM with linear kernel we used
-L2 regularization to keep all potentially important features. For
-comparison, we also trained an L1 regularized SVM model with linear
-kernel. L1-regularization on microbiome data lead to a sparser solution
-(i.e., force many coefficients to zero). Finally, to explore the
-potential for non-linear relationships among features and the outcome of
-interest, we trained tree based models, decision tree, random forest and
-XGboost, as well as an SVM with non-linear kernel.
+We used ML models with different classification algorithms and
+regularization parameters. For regularized logistic regression and SVM
+with linear kernel, we used L2 regularization to keep all potentially
+important features. For comparison, we also trained an L1 regularized
+SVM model with linear kernel. L1-regularization on microbiome data lead
+to a sparser solution (i.e., force many coefficients to zero). Finally,
+to explore the potential for non-linear relationships among features and
+the outcome of interest, we trained tree based models, decision tree,
+random forest and XGBoost, as well as an SVM with non-linear kernel.
 
@@
 training/validation data was used for training purposes and validation
-of hyperparameter selection (i.e.~model selection), and the test set was
-used for evaluation purposes. Validation of hyperparameter selection was
-performed using repeated five-fold cross-validation on the
-training/validation set {[}Figure 1{]}. Similar to the initial
-data-split, five-fold cross-validation was also stratified to maintain
-the overall label distribution on the training and validation sets. We
-validated the cross-validation performances of each hyperparameter
-setting over 100 randomizations and selected the best performing
-hyperparameter setting in terms of AUROC metric and trained the full
-training/validation dataset {[}Figures S1 and S2{]}. We then used the
-held-out test set to evaluate the prediction performance of each ML
-model. The data-split, hyperparameter selection, training and testing
-steps were repeated 100 times to get a reliable and robust reading of
-model performance {[}Figure 1{]}.
+of hyperparameter selection, and the test set was used for evaluation
+purposes.
 
-\textbf{Discriminative performance and generalizability of the seven
+Hyperparameters are the rules that are learned from the training data in
+a classification algorithm. When they are tuned over a full grid search
+and selected by validation, they make the ML model better are
+predicting. We selected hyperparameter settings using repeated five-fold
+cross-validation on the training/validation set {[}Figure 1{]}. We chose
+the best hyperparameter setting for each model based on its predictive
+performance on the validation set using the area under the receiver
+operating characteristic curve (AUROC) metric {[}Figure S1 and S2{]}.
+The AUROC ranges from 1.0, where the model perfectly distinguishes
+between cases and controls, to 0.50, where the model's predictions are
+no different from random chance. Similar to the initial data-split,
+five-fold cross-validation for hyperparameter selection was also
+stratified to maintain the overall label distribution on the training
+and validation sets. The cross-validation of each hyperparameter setting
+was repeated over 100 randomizations to get a robust reading of
+predictive performance.
+
+We then trained the full training/validation dataset with the selected
+hyperparameters. We used the held-out test set to evaluate the
+predictive performance of each ML model. The data-split, hyperparameter
+selection, training and testing steps were repeated 100 times to get a
+reliable and robust reading of model performance {[}Figure 1{]}.
+
+\textbf{Predictive performance and generalizability of the seven
 models.}
@@
 We evaluated the predictive performances of seven binary classification
-models when applied to held-out test data using AUROC metric {[}Figure
-2{]}. Random forest had significantly higher test AUROC values than the
-other models for detecting SRNs when AUROC values were compared to the
-other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
-AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
-logistic regression, XGBoost, L2-regularized SVM with linear and radial
-basis function kernel AUROC values were not significantly different from
-one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
-(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
-SVM with linear kernel and decision tree had significantly lower AUROC
-values than the other ML models with median AUROC of 0.65 (IQR 0.066)
-and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.
+models when applied to held-out test data using the AUROC metric
+{[}Figure 2{]}. Random forest had significantly higher test AUROC values
+than the other models for detecting SRNs when AUROC values were compared
+to the other six by Wilcoxon rank sum test (p \textless{} 0.01). The
+median AUROC of the random forest model was 0.695 (IQR 0.044).
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with
+linear and radial basis function kernel AUROC values were not
+significantly different from one another. They had median AUROC values
+of 0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
+0.056) respectively. L1-regularized SVM with linear kernel and decision
+tree had significantly lower AUROC values than the other ML models with
+median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
+{[}Figure 2{]}. Random forest had the highest median AUROC for detecting
+SRN. Despite its simplicity, the L2-regularized logistic regression was
+second best in predictive performance.
 
-For each model, we compared the median cross-validation AUROC to the
-median testing AUROC. The difference between the two should be low to
-suggest the model is not overfitting despite the large number of
-features. The largest difference between the two was 0.021 in L1 SVM
-with linear kernel, followed by SVM with radial basis function kernel
-and decision tree with a difference of 0.007 and 0.006, respectively
-{[}Figure 2{]}.
-
-We reported the testing AUROC values over 100 randomizations of the
-initial data-split. The results showed that depending on the data-split,
-the testing AUROC values showed great variability {[}Figure 2{]}. The
-testing AUROC values within each model varied 0.23 on average across the
-seven models. For instance, the lowest AUROC value of the random forest
-model was 0.59 whereas the highest was 0.81.
+To evaluate the generalizability of each model, we compared the median
+cross-validation AUROC to the median testing AUROC. The difference
+between the two should be low to suggest the model is not overfitting
+despite the large number of features. The largest difference between the
+two was 0.021 in L1-regularized SVM with linear kernel, followed by SVM
+with radial basis function kernel and decision tree with a difference of
+0.007 and 0.006, respectively {[}Figure 2{]}. We also reported the
+testing AUROC values over 100 randomizations of the initial data-split.
+The testing AUROC values within each model varied 0.23 on average across
+the seven models. For instance, the lowest AUROC value of the random
+forest model was 0.59 whereas the highest was 0.81. These results showed
+that depending on the data-split, the testing AUROC values showed great
+variability {[}Figure 2{]}.
 
@@
 
-The ML models we built using L2-regularized logistic regression, L1 and
-L2 support vector machines (SVM) with linear and radial basis function
-kernels, a decision tree, random forest and XGBoost decrease in
-interpretibility as they increase in complexity. We interpreted L1 and
-L2 SVM with linear kernel and L2 logistic regression using the absolute
+Interpretability is the degree to which humans can understand the
+reasons behind a model prediction (27). Because we often use ML models
+not just to predict a health outcome but also to learn the ecology
+behind a disease, model interpretation becomes crucial for microbiome
+studies. The ML models we built using L2-regularized logistic
+regression, L1 and L2 support vector machines (SVM) with linear and
+radial basis function kernels, a decision tree, random forest and
+XGBoost decrease in interpretability as they increase in complexity. In
+this study we highlighted two methods to interpret models with varying
+complexity.
+
+We interpreted linear models (L1 and L2-regularized SVM with linear
+kernel and L2-regularized logistic regression) using the absolute
 feature weights of the trained models. We ranked the absolute weights of
@@
 method called permutation importance on the held-out test data.
-Permutation importance analysis is where we randomly permute
-non-correlated features individually and groups of highly correlated
-features together. We then calculate how much the predictive performance
-of the model (i.e AUROC values) decrease when each OTU or group of OTUs
-is permuted randomly. We ranked the OTUs based on how much they
-decreased the median testing AUROC; the OTU with the largest decrease
-ranking highest. The top 5 OTUs with the largest negative impact on
-testing AUROC overlapped in tree-based models {[}Figure 4{]}.
-Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
-randomly, dropped the predictive performances the most in all tree-based
-methods {[}Figure 4{]}. Decision tree, random forest and XGBoost models'
-predictive performance dropped from 0.6 median AUROC to 0.52, from 0.69
-to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.
+Permutation importance analysis is a posthoc explanation of the model
+where we randomly permute non-correlated features individually and
+groups of highly correlated features together. We then calculate how
+much the predictive performance of the model (i.e AUROC values) decrease
+when each OTU or group of OTUs is permuted randomly. We ranked the OTUs
+based on how much they decreased the median testing AUROC; the OTU with
+the largest decrease ranking highest. The top 5 OTUs with the largest
+negative impact on testing AUROC overlapped in tree-based models
+{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
+(OTU00367) abundances randomly, dropped the predictive performances the
+most in all tree-based methods {[}Figure 4{]}. Decision tree, random
+forest and XGBoost models' predictive performance dropped from 0.6
+median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
+respectively {[}Figure 4{]}.
 
@@
 S3{]}. L1-regularized SVM with linear kernel picked out some of the same
-OTUs (OTU00822, OTU01239, OTU00609) as important in the two
-interpretation methods; feature rankings based on weights {[}Figure 3{]}
-and permutation importance {[}Figure S3{]}. Similarly, L2-regularized
-SVM and L2-regularized logistic regression picked out some of the same
-OTUs in both interpretation methods, OTU00659 and OTU00012,
-respectively. However, for all the linear models, the rankings of these
-features were different due to the collinearity in microbial
-communities.
+OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings
+based on weights {[}Figure 3{]} and permutation importance {[}Figure
+S3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
+regression picked out some of the same OTUs in both interpretation
+methods, OTU00659 and OTU00012, respectively. However, for all the
+linear models, the rankings of these features were different due to the
+collinearity in microbial communities.
 
@@
 
-As the complexity of a ML model {[}Table S1{]} and the number of tuned
-hyperparameter settings increased {[}Figures S1-S2{]}, its training
-times increased as well {[}Figure 5{]}. Linear models trained faster
-than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic
-regression had training times of 0.2 hours, (std ± 0.03), 0.2 hours,
-(std ± 0.02), and 0.2 hours, (std ± 0.02), respectively. Whereas, a
-decision tree, SVM with radial basis function kernel, random forest and
-xgboost had training times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ±
-8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1), respectively
-{[}Figure 5{]}.
+We compared the training times of the seven ML models. As the complexity
+of a ML model {[}Table S1{]} and the number of tuned hyperparameter
+settings increased {[}Figures S1-S2{]}, its training times increased as
+well {[}Figure 5{]}. Linear models trained faster than non-linear
+models. L1 and L2 SVM with linear kernel and L2-regularized logistic
+regression had the shortest training times with 0.2 hours, (std ± 0.03),
+0.2 hours, (std ± 0.02), and 0.2 hours, (std ± 0.02), respectively.
+Whereas, a decision tree, SVM with radial basis function kernel, random
+forest and XGBoost had training times of 4.4 hours, (std ± 0.3), 59.6
+hours, (std ± 8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1),
+respectively {[}Figure 5{]}.
 
@@
 In this study we established a rigorous ML pipeline to use 16S rRNA
-sequence counts to predict a binary health outcome. We set-up standards
-for developing and evaluating ML models for microbiome data. First, we
-used a held-out test set to illustrate the difference between
-cross-validation and testing AUROC values. When the difference between
-cross-validation and test performance is low, this suggets the models
-are not overfit and that they will perform similar with similar data. In
-all seven models, the difference between median cross-validation and
-testing AUROC values did not exceed 0.021 which suggests that these
-models are generalizable and can be used to test similar new data.
-Second, we performed the initial 80\%-20\% random datasplit 100 times in
-our ML pipeline. The randomization of the initial data-split to create a
-held-out test set is a crucial step in the ML pipeline to develop robust
-ML models and to report reliable performance metrics for a ML model.
-Depending on how the data is split, there is the chance of being
-overoptimistic about the predictive performance of a model. In our
-study, we showed that there was variability in AUROC values between
-different random data-splits in each of the models we tested. Our
-results showed that the testing AUROC values varied 0.23 on average
-between different data-splits. Third, we used the AUROC metric instead
-of accuracy in our study to evaluate the predictive performance of the
-ML models. AUROC is always random at the value 0.5 and is a robust
-metric when a dataset is imbalanced. We also performed a full grid
-search for hyperparameter settings when building a ML model. Default
-hyperparameter settings in previously developed ML packages in R,
-Python, and Matlab programming languages are inadequate for effective
+sequence counts to predict a binary health outcome. We built on others'
+work in the microbiome field to set-up standards for developing and
+evaluating ML models for microbiome data (28--30). First, we used a
+held-out test set to illustrate the difference between cross-validation
+and testing AUROC values. When the difference between cross-validation
+and test performance is low, this suggest the models are not overfit and
+that they will perform similar with similar data. In all seven models,
+the difference between median cross-validation and testing AUROC values
+did not exceed 0.021 which suggests that these models are generalizable
+and can be used to test similar new data. Second, we performed the
+initial 80\%-20\% random datasplit 100 times in our ML pipeline. The
+randomization of the initial data-split to create a held-out test set is
+a crucial step in the ML pipeline to develop robust ML models and to
+report reliable performance metrics. Depending on how the data is split,
+there is the chance of being overoptimistic about the predictive
+performance of a model. In our study, we showed that there was
+variability in AUROC values between different random data-splits in each
+of the models we tested. Our results showed that the testing AUROC
+values varied 0.23 on average between different data-splits. Third, we
+used the AUROC metric in our study to evaluate the predictive
+performance of the ML models. AUROC is always random at the value 0.5
+and is a robust metric when a dataset is imbalanced. We also performed a
+full grid search for hyperparameter settings when building a ML model.
+Default hyperparameter settings in previously developed ML packages in
+R, Python, and Matlab programming languages are inadequate for effective
 application of classification algorithms and need to be optimized for
@@
 the goal of the study and our expectations of predictive performance,
-interpretibility and computational burden. Microbiome studies use ML
+interpretability and computational burden. Microbiome studies use ML
 models with a classification task to learn the training data to assign
-labels, such as healthy or not, to new data, but also to learn which
-features are important to discriminate between labels. If the goal of a
-study is not just to predict a health outcome but also to learn the
-ecology behind a disease and to identify microbial biomarkers of a
-disease, then the ML model has to be interpretable. In terms of
-predictive performance, random forest model had testing AUROC values
-statistically significantly larger than others. However the second best
-model was L2-regularized logistic regression following random forest
-with a median AUROC difference of only 0.015. In terms of
-interpretability, random forest was a complex ML model and it could only
-be explained using methods such as permutation importance. On the other
+labels, such as healthy or not to new data, but also to learn which
+features are important to discriminate between labels (2--11). Our
+results show that if the goal of a study is to learn the ecology behind
+a disease and to identify microbial biomarkers, we could create ML
+models that are inherently interpretable without losing predictive
+power. In terms of predictive performance, random forest model had the
+best testing AUROC values compared to the other 6 models. However, the
+second best model was L2-regularized logistic regression with a median
+AUROC difference of only 0.015 compared to random forest. In terms of
+interpretability, random forest was a complex ML model and could only be
+explained using methods such as permutation importance. On the other
 hand, L2-regularized logistic regression was easier to interpret
@@
 influenced by one another which makes it difficult to interpret the ML
-model. To avoid misinterpreting the models we can; (1) use ML models
-that has highly correlated OTUs removed, (2) use the highly ranked OTUs
-that are correlated with other OTUs to generate hypotheses about the
-ecology of the disease and test them with followup experiments.
+model. To avoid misinterpreting the models we should use the highly
+ranked correlated OTUs to generate hypotheses about the ecology of the
+disease and test them with follow-up experiments.
 
 There are other criteria when choosing ML models such as the
-computational burden of developing it and sample size. In terms of
+computational burden of developing it and the sample size. In terms of
 computational burden, random forest model trained each data-split in
@@
 This study highlights the need to make educated choices at every step of
-developing a ML model with microbiome data. Model selection, setting up
-a rigorous ML pipeline and choosing the right interpretation method
-contribute to achieving the level of validity and accountability we want
-from models built for patient health.
+developing a ML model with microbiome data. Model selection should be
+done with a solid understanding of model complexity and
+interpretability, rigorous ML pipelines should be built with
+cross-validation for hyperparameter tuning and with a held-out test set
+for evaluating predictive performance and models should be interpreted
+while considering collinearity in datasets. The right methods will help
+us achieve the level of validity and accountability we want from models
+built for patient health.
 
@@
 patients recruited by Great Lakes-New England Early Detection Research
-Network study. These data were obtained from Sze et al (27). The stool
+Network study. These data were obtained from Sze et al (31). The stool
 samples were provided by recruited adult participants who were
@@
 assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(28--30).
+(32--34).
 
@@
 SRN class includes patients with advanced adenomas or carcinomas. The
-bacterial abundances are the features used to predict colonic health of
-the patients. Bacterial abundances are discrete data in the form of
-Operational Taxonomic Unit (OTU) counts. OTU counts were set to the size
-of our smallest sample and were subsampled at the same distances. They
-were then transformd by scaling to a {[}0-1{]} range.
+study had 261 normal and 229 SRN samples. The bacterial abundances are
+the features used to predict colonic health of the patients. Bacterial
+abundances are discrete data in the form of Operational Taxonomic Unit
+(OTU) counts. OTU counts were set to the size of our smallest sample and
+were subsampled at the same distances. They were then transformed by
+scaling to a {[}0-1{]} range.
 
@@
 testing performance. We can use hyper-band to help us with our
-hyperparameter selection (31).
+hyperparameter selection (35).
 
@@
 comparisons. We then defined correlated OTUs as having perfect
-correlation (correlation coef=1 and p\textless{}0.01).Other OTUs were
-permuted individually to get permutation importance but the correlated
-ones are grouped together and permuted at the same time. The reported
-OTUs that have impact on testing AUROC are reported in Figures xxx. If
-we want we can decrease the correlation coefficient to consider OTUs
-that are correlated with ecological consequences but this was out of the
-scope of this study but will be followed up in further analyses.
+correlation (correlation coef=1 and p\textless{}0.01). Non-correlated
+OTUs were permuted individually whereas correlated ones were grouped
+together and permuted at the same time.
 
@@
 seven ML models by Wilcoxon rank sum tests to determine the best
-discriminative performance.
+predictive performance.
 
@@
 evaluated on a held-out set of data (not used in selecting the model).
-Abbreviations: AUROC, area under the receiver operating characteristic
-curve \newpage
+Abbreviations: cvAUROC, cross-validation area under the receiver
+operating characteristic curve \newpage
 \includegraphics{Figure_2.png}
@@
 using bacterial abundances was higher than chance (depicted by
-horizontal line at 0.50) for all the ML models. Discriminative
-performance of random forest model was higher than other ML models. The
-boxplot shows quartiles at the box ends and the statistical median as
-the horizontal line in the box. The whiskers show the farthest points
-that are not outliers. Outliers are data points that are not within 3/2
-times the interquartile ranges. Abbreviations: SRN, screen-relevant
+horizontal line at 0.50) for all the ML models. Predictive performance
+of random forest model was higher than other ML models. The boxplot
+shows quartiles at the box ends and the statistical median as the
+horizontal line in the box. The whiskers show the farthest points that
+are not outliers. Outliers are data points that are not within 3/2 times
+the interquartile ranges. Abbreviations: SRN, screen-relevant
 neoplasias; AUROC, area under the receiver operating characteristic
@@
 
-\textbf{Figure 4. Explanation of the non-linear ML models.} (A) SVM with
-radial basis kernel (B) decision tree (C) random forest (D) XGboost
-feaure importances were explanied using permutation importance using
+\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
+with radial basis kernel (B) decision tree (C) random forest (D) XGBoost
+feature importances were explained using permutation importance using
 held-out test set. The gray rectangle and the dashed line show the IQR
 range and median of the base testing AUROC without any permutation
-performed. The colors of the boxplots stand for the unique OTUs that are
-shared among the different models; pink for OTU0008, salmon for OTU0050,
-yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for
-OTU00882. For all the tree-based models, a \emph{Peptostreptococcus}
-species (OTU00367) had the largest impact on predictive performance of
-the model. Abbreviations: SVM, support vector machine; OTU, Operational
-Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic
-Unit. \newpage
+performed. The colors of the box plots stand for the unique OTUs that
+are shared among the different models; pink for OTU0008, salmon for
+OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and
+red for OTU00882. For all the tree-based models, a
+\emph{Peptostreptococcus} species (OTU00367) had the largest impact on
+predictive performance of the model. Abbreviations: SVM, support vector
+machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
+Operational Taxonomic Unit. \newpage
 \includegraphics{Figure_5.png}
@@
 training times for of each data-split showed the differences in
-computational efficieny of the seven models. The median training time in
-hours was the highest for XGBoost and shortest for L1-regularized SVM
+computational efficiency of the seven models. The median training time
+in hours was the highest for XGBoost and shortest for L1-regularized SVM
 with linear kernel. The boxplot shows quartiles at the box ends and the
@@
 models.} (A) Decision tree (B) Random forest (C) SVM with radial basis
-kernel (D) XGoost mean cross-validation AUROC values when different
+kernel (D) XGBoost mean cross-validation AUROC values when different
 hyperparameters are used in training the model. The differences in AUROC
@@
 
+\hypertarget{ref-miller_explanation_2017}{}
+27. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
+Insights from the social sciences. arXiv:170607269 {[}cs{]}.
+
+\hypertarget{ref-statnikov_comprehensive_2013}{}
+28. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
+\textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
+\textbf{Blaser MJ}, \textbf{Aliferis CF}, \textbf{Alekseyenko AV}. 2013.
+A comprehensive evaluation of multicategory classification methods for
+microbiomic data. Microbiome \textbf{1}:11.
+doi:\href{https://doi.org/10.1186/2049-2618-1-11}{10.1186/2049-2618-1-11}.
+
+\hypertarget{ref-knights_supervised_2011}{}
+29. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
+Supervised classification of human microbiota. FEMS Microbiology Reviews
+\textbf{35}:343--359.
+doi:\href{https://doi.org/10.1111/j.1574-6976.2010.00251.x}{10.1111/j.1574-6976.2010.00251.x}.
+
+\hypertarget{ref-wirbel_meta-analysis_2019}{}
+30. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
+K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
+\textbf{Voigt AY}, \textbf{Palleja A}, \textbf{Ponnudurai R},
+\textbf{Sunagawa S}, \textbf{Coelho LP}, \textbf{Schrotz-King P},
+\textbf{Vogtmann E}, \textbf{Habermann N}, \textbf{Niméus E},
+\textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Gandini S},
+\textbf{Serrano D}, \textbf{Mizutani S}, \textbf{Shiroma H},
+\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
+T}, \textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N},
+\textbf{Sinha R}, \textbf{Ulrich CM}, \textbf{Brenner H},
+\textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller G}. 2019.
+Meta-analysis of fecal metagenomes reveals global microbial signatures
+that are specific for colorectal cancer. Nature Medicine
+\textbf{25}:679.
+doi:\href{https://doi.org/10.1038/s41591-019-0406-6}{10.1038/s41591-019-0406-6}.
+
 \hypertarget{ref-sze_leveraging_2018}{}
-27. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+31. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+32. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-29. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-30. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+34. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-31. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+35. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-07-18 submission/manuscript.Rmd
add a Table 1

@@
 \newpage
-\includegraphics{TableS1.pdf}
-**Table S1. ML model complexity and hyperparameter range.** 
+\captionof{table}{Characteristics of the machine learning models in our comparative study.}
+\small
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+\rowcolor{lightgray}
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} & \textbf{Refs.} \\ \hline
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable & $^{36}$\\ \hline
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\seperating hyperplane that discriminates between labels.} & Linear & Interpratable & $^{37}$\\ \hline
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal gaussian \\seperating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
+Decision tree & \makecell[l]{A classifier that sorts samples down from the 
+root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline
+Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ & $^{40-41}$  \\ \hline
+XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ & $^{42-43}$ \\ \hline
+\end{tabular}
+\begin{tablenotes}\footnotesize
+\item[1]{$^*$Explainable models are not inherently interpretable but can be explained with post-hoc analyses.}
+\end{tablenotes}
 \newpage

### 2019-07-18 submission/manuscript.tex
add a Table 1

@@
 \usepackage{caption}
+\usepackage{makecell}
+\usepackage{threeparttable}
 
@@
 \usepackage[T1]{fontenc}
+\usepackage[labelfont=bf]{caption}
 
@@
 
-\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
-hyperparameter range.} \newpage
+\captionof{table}{Characteristics of the machine learning models in our comparative study.}
+\small
+
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+\rowcolor{lightgray}
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} & \textbf{Refs.} \\ \hline
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable & $^{36}$\\ \hline
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\seperating hyperplane that discriminates between labels.} & Linear & Interpratable & $^{37}$\\ \hline
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal gaussian \\seperating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
+Decision tree & \makecell[l]{A classifier that sorts samples down from the 
+root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline
+Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ & $^{40-41}$  \\ \hline
+XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ & $^{42-43}$ \\ \hline
+\end{tabular}\begin{tablenotes}\footnotesize
+\item[1]{$^*$Explainable models are not inherently interpretable but can be explained with post-hoc analyses.}
+\end{tablenotes}\newpage
 

### 2019-07-18 submission/Table1.Rmd
add Table 1 and 2

--- 2019-07-12 submission/manuscript.Rmd
@@
 ---
-title: '**Evaluation of machine learning methods for 16S rRNA gene data**'
-csl: mbio.csl
-fontsize: 11pt
+title: ""
 output:
   pdf_document:
+    keep_tex: false
     includes:
       in_header: header.tex
-    keep_tex: yes
-  word_document: default
 geometry: margin=1.0in
-bibliography: references.bib
+fontsize: 11pt
 ---
 
-```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
-######################################################################
-#----------------- Read in necessary libraries -------------------#
-######################################################################
-deps = c("reshape2","knitr","rmarkdown","vegan","gtools", "tidyverse");
-for (dep in deps){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
-  }
-  library(dep, verbose=FALSE, character.only=TRUE)
-}
-######################################################################
-#-------------- Define the chunk options ----------------#
-######################################################################
-opts_chunk$set("tidy" = TRUE)
-opts_chunk$set("echo" = FALSE)
-opts_chunk$set("eval" = TRUE)
-opts_chunk$set("warning" = FALSE)
-opts_chunk$set("cache" = FALSE)
+\newpage
+\captionof{table}{Characteristics of the machine learning models in our comparative study.}
+\small
+\begin{tabular}{|l|l|l|l|l|}
+\hline
 
-inline_hook <- function(x){
-	print(x)
+\rowcolor{lightgray}
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} & \textbf{Refs.} \\ \hline
 
-	if(is.list(x)){
-		x <- unlist(x)
-	}
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable & $^{36}$\\ \hline
 
-	if(is.numeric(x)){
-		if(abs(x - round(x)) < .Machine$double.eps^0.5){
-			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
-		} else {
-			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
-		}
-	} else {
-    	paste(x)      
-	}
-}
-knitr::knit_hooks$set(inline=inline_hook)
-```
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\seperating hyperplane that discriminates between labels.} & Linear & Interpratable & $^{37}$\\ \hline
 
-```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
-######################################################################
-#----------------- Read in important functions -----------------#
-######################################################################
-source("../code/learning/functions.R")
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal gaussian \\seperating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
+Decision tree & \makecell[l]{A classifier that sorts samples down from the 
+root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline
 
-#####################################################################
+Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ & $^{40-41}$  \\ \hline
 
-######################################################################
-#----------------- Load OTU table and MetaData -----------------#
-######################################################################
-source("../code/learning/load_data.R")
+XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ & $^{42-43}$ \\ \hline
 
-######################################################################
-#----------------- Load feature importance data -----------------#
-# Median of base testing AUC for each model
-# 100 datasplit permutation importances new testing AUC for top 5 features
-######################################################################
-
-source("../code/learning/load_imp_data.R")
-
-######################################################################
-#-------------------------- Load traintime data -----------------#
-######################################################################
-source("../code/learning/load_traintime.R")
-
-######################################################################
-#-------- Load AUROC performance data and Wilcoxon test ------------#
-
-######################################################################
-source("../code/learning/load_AUROC_data.R")
-```
-
-\vspace{35mm}
-
-Running title: Machine learning methods in microbiome studies
-
-\vspace{35mm}
-
-Begüm D. Topçuoğlu${^1}$, Nick Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^
-
-\vspace{40mm}
-
-$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
-
-1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
-
-2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109
-
-3\. Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA
-
+\end{tabular}
+\begin{tablenotes}\footnotesize
+\item[1]{$^*$Explainable models are not inherently interpretable but can be explained with post-hoc analyses.}
+\end{tablenotes}
 \newpage
-\linenumbers
-
-## Abstract
-
-Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed ML models into question.
-Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers.
-To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
-To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
-The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
-Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
-In this study, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.
-Additionally, we established standards for the development of modeling pipelines for microbiome-associated ML models
-
-\newpage
-## Importance (needs work)
-
-Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies.
-However, the developed ML models so far are overoptimistic in terms of validity and predictive performance.
-Without rigorous ML pipelines, we cannot trust ML models.
-Before we can speed up progress, we need to slow down, define and start using good ML practices.
-\newpage
-
-## Background
-
-Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications.
-Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health.
-However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease.
-It is unlikely that a single species can explain a disease.
-Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.
-
-Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
-They learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction.
-Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-However, the field's use of ML lacks clarity and consistency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019].
-More notably, we commonly see flawed ML practices such as (1) using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or (2) reporting few or only the best outcomes of cross-validation.
-Even when there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
-
-Moreover, there is a lack of discussion on why a particular ML model is utilized.
-Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-Black box machine learning models are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction.
-These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018].
-The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].
-
-The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliy.
-We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
-We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of `r paste(nrow(shared))` patients.
-The human-associated human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
-We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
-The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples.
-We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.
-We used the area under the receiver operating characteristic curve (AUROC) as a metric for determining the predictibve performance of each model.
-The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance.
-The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`.
-Random forest had the highest median AUROC for detecting SRN.
-Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.
-In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`).
-We also found that depending on how the data is split to create a held-out test set, the AUROC values of a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs.
-This study established standards for microbiome-associated ML models and underscored the importance of model selection.
-
-## Results
-
-__Model selection and pipeline construction__
-
-We used a cohort of 490 patients with 261 cases of SRN.
-For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (having colorectal lesions that are identified as SRN or normal).
-All the cases were independently labeled through colonoscopies.
-We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost) to emphasize the differences in  model accuracy, complexity, interpretibility and computational efficiency due to  model selection.
-
-For regularized logistic regression and SVM with linear kernel we used L2 regularization to keep all potentially important features.
-For comparison, we also trained an L1 regularized SVM model with linear kernel.
-L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero).
-Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGboost, as well as an SVM with non-linear kernel.
-
-We established a ML pipeline where we train and validate each of the seven models [Figure 1].
-We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
-Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets.
-Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
-The training/validation data was used for training purposes and validation of hyperparameter selection (i.e.
-model selection), and the test set was used for evaluation purposes.
-Validation of hyperparameter selection was performed using repeated five-fold cross-validation on the training/validation set [Figure 1].
-Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets.
-We validated the cross-validation performances of each hyperparameter setting over 100 randomizations and selected the best performing hyperparameter setting in terms of AUROC metric and trained the full training/validation dataset [Figures S1 and S2].
-We then used the held-out test set to evaluate the prediction performance of each ML model.
-The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].
-
-__Discriminative performance and generalizability of the seven models.__
-
-We evaluated the predictive performances of seven binary classification models when applied to held-out test data using AUROC metric [Figure 2].
-Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01).
-The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another.
-They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively.
-L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
-
-For each model, we compared the median cross-validation AUROC to the median testing AUROC.
-The difference between the two should be low to suggest the model is not overfitting despite the large number of features.
-The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1 SVM with linear kernel,  followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
-
-We reported the testing AUROC values over 100 randomizations of the initial data-split.
-The results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
-The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
-For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.
-
-__Interpretation of each ML model.__
-
-The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity.
-We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the absolute feature weights of the trained models.
-We ranked the absolute weights of all the OTUs for each data-split [Figure 3].
-We calculated the median ranks of these features over the 100 data-splits.
-In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belonged to families *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3].
-Some of the OTUs with the highest ranks were shared among the linear models.
-
-We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.
-Permutation importance analysis is where we randomly permute non-correlated features individually and groups of highly correlated features together.
-We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly.
-We ranked the OTUs based on how much they decreased the median testing AUROC; the OTU with the largest decrease ranking highest.
-The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
-Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure 4].
-Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
-
-To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3].
-L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in the two interpretation methods; feature rankings based on weights [Figure 3] and permutation importance [Figure S3].
-Similarly,  L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively.
-However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.
-
-__The computational efficiency of each ML model.__
-
-As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5].
-Linear models trained faster than non-linear models.
-L1 and L2 SVM with linear kernel and L2 logistic regression had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
-Whereas, a decision tree, SVM with radial basis function kernel, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
-
-## Discussion
-
-In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
-We set-up standards for developing and evaluating ML models for microbiome data.
-First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
-When the difference between cross-validation and test performance is low, this suggets the models are not overfit and that they will perform similar with similar data.
-In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data.
-Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
-The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics for a ML model.
-Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model.
-In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
-Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits.
-Third, we used the AUROC metric instead of accuracy in our study to evaluate the predictive performance of the ML models.
-AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
-We also performed a full grid search for hyperparameter settings when building a ML model.
-Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model.
-In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
-
-In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden.
-Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not, to new data, but also to learn which features are important to discriminate between labels.
-If the goal of a study is not just to predict a health outcome but also to learn the ecology behind a disease and to identify microbial biomarkers of a disease, then the ML model has to be interpretable.
-In terms of predictive performance, random forest model had testing AUROC values statistically significantly larger than others.
-However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015.
-In terms of interpretability, random forest was a complex ML model and it could only be explained using methods such as permutation importance.
-On the other hand, L2-regularized logistic regression was easier to interpret (i.e.
-ranking absolute regression coefficients of the trained model).
-
-Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models.
-In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
-We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset.
-Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique.
-The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model.
-To avoid misinterpreting the models we can; (1) use ML models that has highly correlated OTUs removed, (2) use the highly ranked OTUs that are correlated with other OTUs to generate hypotheses about the ecology of the disease and test them with followup experiments.
-
-There are other criteria when choosing ML models such as the computational burden of developing it and sample size.
-In terms of computational burden, random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
-The generalization performance of ML models depends on sample size.
-The more complex the model, the more data it will need.
-The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.
-
-This study highlights the need to make educated choices at every step of developing a ML model with microbiome data.
-Model selection, setting up a rigorous ML pipeline and choosing the right interpretation method contribute to achieving the level of validity and accountability we want from models built for patient health.
-
-## Materials and Methods
-
-**Data collection and study population.**
-  The data used for this analysis are stool bacterial abundances and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study.
-These data were obtained from Sze et al [@sze_leveraging_2018].
-  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscopy.
-Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
-Patients' colonic health was labeled by colonoscopy with adequate preparation and tissue histopathology of all resected lesions.
-Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma.
-Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies, `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas.
-Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas.
-Stool provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances.
-The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].
-
-**Data definitions and pre-processing.**
-
-  The colorectal health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs).
-Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas.
-The bacterial abundances are the features used to predict colonic health of the patients.
-Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
-OTU counts were set to the size of our smallest sample and were subsampled at the same distances.
-They were then transformd by scaling to a [0-1] range.
-
-**Model training and evaluation.**
-
-Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).
-Within the caret package, we have made modifications to L2-regularized SVM with linear kernel function __svmLinear3__ and developed a L1-regularized SVM with linear kernel function __svmLinear4__ to calculate decision values instead of predicted probabilities.
-These changes are available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.
-
-For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
-For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
-For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has.
-For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
-For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.For hyperparameter selection, we started with a granular grid search.
-Then we narrowed and fine-tuned the range of each hyperparameter.
-The range of the grid depends on the ML task and ML model.
-A full grid search needs to be performed to avoid variability in testing performance.
-We can use hyper-band to help us with our hyperparameter selection [@li_hyperband:_2016].
-
-The computational burden during model training due to model complexity was reduced by parallelizing segments of the ML pipeline.
-In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
-We can further parallelize the cross-validation step for each hyperparameter setting.
-
-**Permutation importance workflow.**
-We created a Spearman's rank-order correlation matrix, corrected for multiple pairwise comparisons.
-We then defined correlated OTUs as having perfect correlation (correlation coef=1 and p<0.01).Other OTUs were permuted individually to get permutation importance but the correlated ones are grouped together and permuted at the same time.
-The reported OTUs that have impact on testing AUROC are reported in Figures xxx.
-If we want we can decrease the correlation coefficient to consider OTUs that are correlated with ecological consequences but this was out of the scope of this study but will be followed up in further analyses.
-
-**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1).
-We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best discriminative performance.
-
-**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.
-
-\newpage
-\includegraphics{Figure_1}
-**Figure 1.
-Machine learning pipeline showing predictive model training and evaluation flowchart.
-**  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data.
-The model was evaluated on a held-out set of data (not used in selecting the model).
-Abbreviations: AUROC, area under the receiver operating characteristic curve
-\newpage
-\includegraphics{Figure_2.png}
-
-**Figure 2.
-Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.
-**   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-Discriminative performance of random forest model was higher than other ML models.
-The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
-The whiskers show the farthest points that are not outliers.
-Outliers are data points that are not within 3/2 times the interquartile ranges.
-Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
-\newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.png}
-
-**Figure 3.
-Interpretation of the linear ML models.** The absolute feature weights of (A) L2 logistic regression coefficients (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel were ranked from highest rank 1 to 100 for each data-split.
-The feature ranks of the highest ranked five OTUs based on their median ranks are shown here.
-Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
-\newpage
-\includegraphics{Figure_4.png}
-
-**Figure 4.
-Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set.
-The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
-The colors of the boxplots stand for the unique OTUs that are shared among the different models; pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for OTU00882.
-For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
-\newpage
-\includegraphics{Figure_5.png}
-
-**Figure 5.
-Computational efficiency of seven ML models.** The training times for of each data-split showed the differences in computational efficieny of the seven models.
-The median training time in hours was the highest for XGBoost and shortest for L1-regularized SVM with linear kernel.
-The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
-The whiskers show the farthest points that are not outliers.
-Outliers are data points that are not within 3/2 times the interquartile ranges.
-Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
-\newpage
-\includegraphics{Figure_S1.png}
-
-**Figure S1.
-Hyperparameter setting performances for linear models.** (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters are used in training the model.
-The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
-
-\newpage
-\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
-
-**Figure S2.
-Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGoost mean cross-validation AUROC values when different hyperparameters are used in training the model.
-The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
-
-**Figure S3.
-Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel (B) L2-regularized SVM with linear kernel and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
-The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
-
-\newpage
-\includegraphics{TableS1.pdf}
-**Table S1. ML model complexity and hyperparameter range.** 
-\newpage
-
-## References
-
-<div id="refs"></div>

### 2019-07-18 submission/Table2.Rmd
add Table 1 and 2

---
title: ""
output:
  pdf_document:
    keep_tex: false
    includes:
      in_header: header.tex
geometry: margin=1.0in
fontsize: 11pt
---

\captionof{table}{An aspirational rubric for evaluating the rigor of ML practices.}
\small
\begin{tabular}{|l|l|l|l|}
\hline

\rowcolor{lightgray}
\textbf{Practice} & \textbf{Good} & \textbf{Better} & \textbf{Best} \\ \hline

\makecell[l]{Problem \\ definition} & \makecell[l]{Have we clearly stated \\ the ML task? \\ Do we have a priori hypotheses? \\Do we know the predictions \\ a domain expert would \\ make manually?} & \makecell[l]{Do we know the motivation \\ for solving the problem? \\ How much interpretability \\ does the problem need? } & \makecell[l]{Do we know our data? \\ Do we know the confounding \\ variables? } \\ \hline

\makecell[l]{Model \\ selection} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem? \\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretibility \\ does the problem need? \\ How much each \\ candidate algorithm \\ can provide?} \\ \hline

\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Do we have an held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\ held-out datasets?} & \makecell[l]{Have we tuned our model \\ hyperparameters in \\ cross-validation?} \\ \hline

\makecell[l]{Hyperparameter \\ selection} & \makecell[l]{Do we know the different \\ hyperparameters each model \\ can use and why?} & \makecell[l]{Did we use historically \\ effective hyperparameters? } & \makecell[l]{Did we search the \\ full grid space and \\optimized our model? } \\ \hline

\makecell[l]{Model \\ evaluation} & \makecell[l]{Have we chosen an appropriate \\ metric to evaluate \\ predictive performance?} & \makecell[l]{Have we reported \\ the predictive performance \\ on a held-out test data?} & \makecell[l]{Have we provided an average \\ predictive performance \\ of many model runs? } \\ \hline

\makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it? \\ Have we checked for \\ the effect of confounding \\ variables?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results? } \\ \hline

\end{tabular}

### 2019-07-18 submission/manuscript.Rmd
add Table 1 and 2

@@
 \hline
+
 \rowcolor{lightgray}
 \textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} & \textbf{Refs.} \\ \hline
+
 \makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable & $^{36}$\\ \hline
+
 \makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\seperating hyperplane that discriminates between labels.} & Linear & Interpratable & $^{37}$\\ \hline
+
 \makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal gaussian \\seperating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
@@
 root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline
+
 Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ & $^{40-41}$  \\ \hline
+
 XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ & $^{42-43}$ \\ \hline
+
 \end{tabular}
@@
 
+\captionof{table}{An aspirational rubric for evaluating the rigor of ML practices.}
+\small
+\begin{tabular}{|l|l|l|l|}
+\hline
+
+\rowcolor{lightgray}
+\textbf{Practice} & \textbf{Good} & \textbf{Better} & \textbf{Best} \\ \hline
+
+\makecell[l]{Problem \\ definition} & \makecell[l]{Have we clearly stated \\ the ML task?
+\\ Do we have a priori hypotheses?
+\\Do we know the predictions \\ a domain expert would \\ make manually?} & \makecell[l]{Do we know the motivation \\ for solving the problem?
+\\ How much interpretability \\ does the problem need?
+} & \makecell[l]{Do we know our data?
+\\ Do we know the confounding \\ variables?
+} \\ \hline
+
+\makecell[l]{Model \\ selection} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem?
+\\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretibility \\ does the problem need?
+\\ How much each \\ candidate algorithm \\ can provide?} \\ \hline
+
+\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Do we have an held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\ held-out datasets?} & \makecell[l]{Have we tuned our model \\ hyperparameters in \\ cross-validation?} \\ \hline
+
+\makecell[l]{Hyperparameter \\ selection} & \makecell[l]{Do we know the different \\ hyperparameters each model \\ can use and why?} & \makecell[l]{Did we use historically \\ effective hyperparameters?
+} & \makecell[l]{Did we search the \\ full grid space and \\optimized our model?
+} \\ \hline
+
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Have we chosen an appropriate \\ metric to evaluate \\ predictive performance?} & \makecell[l]{Have we reported \\ the predictive performance \\ on a held-out test data?} & \makecell[l]{Have we provided an average \\ predictive performance \\ of many model runs?
+} \\ \hline
+
+\makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it?
+\\ Have we checked for \\ the effect of confounding \\ variables?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
+} \\ \hline
+
+\end{tabular}
+\newpage
+
 ## References

### 2019-07-18 submission/manuscript.tex
add Table 1 and 2

@@
 root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline
+
 Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ & $^{40-41}$  \\ \hline
+
 XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ & $^{42-43}$ \\ \hline
+
 \end{tabular}\begin{tablenotes}\footnotesize
@@
 \end{tablenotes}\newpage
+
+\captionof{table}{An aspirational rubric for evaluating the rigor of ML practices.}
+\small
+
+\begin{tabular}{|l|l|l|l|}
+\hline
+
+\rowcolor{lightgray}
+\textbf{Practice} & \textbf{Good} & \textbf{Better} & \textbf{Best} \\ \hline
+
+\makecell[l]{Problem \\ definition} & \makecell[l]{Have we clearly stated \\ the ML task?
+\\ Do we have a priori hypotheses?
+\\Do we know the predictions \\ a domain expert would \\ make manually?} & \makecell[l]{Do we know the motivation \\ for solving the problem?
+\\ How much interpretability \\ does the problem need?
+} & \makecell[l]{Do we know our data?
+\\ Do we know the confounding \\ variables?
+} \\ \hline
+
+\makecell[l]{Model \\ selection} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem?
+\\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretibility \\ does the problem need?
+\\ How much each \\ candidate algorithm \\ can provide?} \\ \hline
+
+\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Do we have an held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\ held-out datasets?} & \makecell[l]{Have we tuned our model \\ hyperparameters in \\ cross-validation?} \\ \hline
+
+\makecell[l]{Hyperparameter \\ selection} & \makecell[l]{Do we know the different \\ hyperparameters each model \\ can use and why?} & \makecell[l]{Did we use historically \\ effective hyperparameters?
+} & \makecell[l]{Did we search the \\ full grid space and \\optimized our model?
+} \\ \hline
+
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Have we chosen an appropriate \\ metric to evaluate \\ predictive performance?} & \makecell[l]{Have we reported \\ the predictive performance \\ on a held-out test data?} & \makecell[l]{Have we provided an average \\ predictive performance \\ of many model runs?
+} \\ \hline
+
+\makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it?
+\\ Have we checked for \\ the effect of confounding \\ variables?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
+} \\ \hline
+
+\end{tabular}\newpage
 

### 2019-07-19 submission/manuscript.Rmd
update paper for table

@@
 
-In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
-We built on others' work in the microbiome field to set-up standards for developing and evaluating ML models for microbiome data [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019].
-First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
+Microbiome studies use ML models, often with a classification task to predict a phenotype but also to learn which microbes are important to discriminate that phenotype from others [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+Achieving either of these tasks in the context of a disease have far-reaching impact on human health, however ML as a tool in microbiome studies is still at its infancy.
+There is a need for a framework for developing rigorous ML models, identifying and overcoming potential pitfalls.
+Previous studies generated robust workflows to allow ML to be widely used by the microbiome researchers [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019] and in this study we set-up standards for developing and evaluating ML models for microbiome data [Table 1].
+
+We benchmarked seven ML models with different classification algorithms to show that the goal of the microbiome study and the clearly defined ML problem should inform our model selection.
+Our results showed that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we could create ML models that are inherently interpretable without losing predictive power.
+In terms of predictive performance, random forest model had the best testing AUROC values compared to the other 6 models.
+However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
+In terms of interpretability, random forest was a complex ML model and could only be explained using methods such as permutation importance.
+On the other hand, L2-regularized logistic regression was easier to interpret by ranking absolute regression coefficients of the trained model.
+
+Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models.
+If we have a priori hypotheses based on knowledge from domain experts and we know the confounding variables in our dataset, we should incorporate that to our model interpretation.
+In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
+We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset.
+Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique.
+The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model.
+To avoid misinterpreting the models we should use the highly ranked correlated OTUs to generate hypotheses about the ecology of the disease and test them with follow-up experiments.
+
+Another criteria when choosing the ML models was the computational time it would take to develop it.
+Random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
+
+We established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
 When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data.
@@
 
-In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretability and computational burden.
-Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not to new data, but also to learn which features are important to discriminate between labels [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-Our results show that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we could create ML models that are inherently interpretable without losing predictive power.
-In terms of predictive performance, random forest model had the best testing AUROC values compared to the other 6 models.
-However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
-In terms of interpretability, random forest was a complex ML model and could only be explained using methods such as permutation importance.
-On the other hand, L2-regularized logistic regression was easier to interpret (i.e.
-ranking absolute regression coefficients of the trained model).
-
-Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models.
-In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
-We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset.
-Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique.
-The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model.
-To avoid misinterpreting the models we should use the highly ranked correlated OTUs to generate hypotheses about the ecology of the disease and test them with follow-up experiments.
-
-There are other criteria when choosing ML models such as the computational burden of developing it and the sample size.
-In terms of computational burden, random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
 The generalization performance of ML models depends on sample size.

### 2019-07-22 submission/manuscript.Rmd
update manuscript and typo fix in Makefile

@@
 
-Microbiome studies use ML models, often with a classification task to predict a phenotype but also to learn which microbes are important to discriminate that phenotype from others [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+Microbiome studies use ML models, often with a classification task, to predict a phenotype but also to learn which microbes are important to discriminate that phenotype from others [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
 Achieving either of these tasks in the context of a disease have far-reaching impact on human health, however ML as a tool in microbiome studies is still at its infancy.

### 2019-07-22 submission/manuscript.Rmd
update manuscript for Pat

--- 2019-07-18 submission/manuscript.Rmd
@@
 Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed ML models into question.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed models into question.
 Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers.
@@
 Moreover, there is a lack of discussion on why a particular ML model is utilized and a lack of emphasis on the strengths and weaknesses of it.
-Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
+Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretability is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
 These models are also called black box ML models because they are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction.
@@
 However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.
+Collinearity in a microbial dataset is when one OTU is dependant on another OTU.
+The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret models using feature weights.
+Our interpretation of L2-regularized logistic regression based on feature weight rankings, showed that an OTU that belongs to *Lachnospiraceae* (OTU00056) was not among the most important OTUs in making a prediction [Figure 3C].
+However, it was picked out as important when the same model was interpreted using permutation importance [Figure S3C].
+This is due to collinearity in the dataset, where *Lachnospiraceae* (OTU00056) is significantly correlated with *Feacalibacterium* (OTU00015) which has the highest ranked feature weights [Figure 3C].
+We need to investigate the relationships among features to identify the true underlying factors when making a prediction.
 
@@
 We compared the training times of the seven ML models.
-As the complexity of a ML model [Table S1] and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5].
+As the complexity of a ML model and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5].
 Linear models trained faster than non-linear models.
-L1 and L2 SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
+L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
 Whereas, a decision tree, SVM with radial basis function kernel, random forest and XGBoost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
@@
 
-In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
-We built on others' work in the microbiome field to set-up standards for developing and evaluating ML models for microbiome data [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019].
-First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
+Microbiome studies use ML models, often with a classification task, to predict a disease but also to learn which microbes are indicators of that disease [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+Achieving either of these tasks have far-reaching impact on human health, however ML as a tool in microbiome studies is still at its infancy.
+A framework is needed to develop rigorous ML models, to identify and overcome potential pitfalls.
+Previous studies generated workflows to allow ML to be widely used by the microbiome researchers [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
+This study sets-up standards for developing and evaluating rigorous ML models for microbiome data [Table 1].
+
+We benchmarked seven ML models with different classification algorithms to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection.
+Our results showed that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we can create ML models that are inherently interpretable and easily trained without losing predictive power.
+In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models.
+However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
+While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes.
+In terms of interpretability, random forest was a more complex ML model and could only be explained using methods such as permutation importance.
+On the other hand, L2-regularized logistic regression was easier to interpret by ranking absolute feature weights of the trained model.
+
+Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease.
+As domain experts, we know that human-associated microbial communities have complex correlation structures that creates collinearity in the dataset.
+Collinearity is a severe problem and needs to be addressed to reliably interpret ML models [@dormann_collinearity:_2013].
+In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
+We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset.
+To avoid misinterpreting the models, once we identify the most important microbes, we should check for their relationships with other microbes as well.
+These relationships will help us generate new hypotheses about the ecology of the disease.
+These hypotheses needs to be tested with follow-up experiments to identify the true biomarkers of a disease.
+
+In this study, we also established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
+First of all, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
 When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data.
-In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data.
-Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
-The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics.
+In all the models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data.
+Furthermore, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
 Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model.
-In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
+We showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
 Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits.
-Third, we used the AUROC metric in our study to evaluate the predictive performance of the ML models.
-AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
-We also performed a full grid search for hyperparameter settings when building a ML model.
+The randomization and resampling of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics.
+We also performed a full grid search for hyperparameter settings when training our ML models.
 Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model.
 In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
+And finally, we used the AUROC metric in our study to evaluate the predictive performance of the ML models.
+AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
 
-In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretability and computational burden.
-Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not to new data, but also to learn which features are important to discriminate between labels [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-Our results show that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we could create ML models that are inherently interpretable without losing predictive power.
-In terms of predictive performance, random forest model had the best testing AUROC values compared to the other 6 models.
-However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
-In terms of interpretability, random forest was a complex ML model and could only be explained using methods such as permutation importance.
-On the other hand, L2-regularized logistic regression was easier to interpret (i.e.
-ranking absolute regression coefficients of the trained model).
-
-Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models.
-In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
-We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset.
-Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique.
-The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model.
-To avoid misinterpreting the models we should use the highly ranked correlated OTUs to generate hypotheses about the ecology of the disease and test them with follow-up experiments.
-
-There are other criteria when choosing ML models such as the computational burden of developing it and the sample size.
-In terms of computational burden, random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.
+We used a balanced CRC dataset to develop ML models with a binary classification task.
+We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes.
+However, the principles highlighted throughout this study [Table 1] apply to all ML modeling tasks with microbiome data.
+The models we built were generalizable despite the high number of features microbiome datasets usually have.
 The generalization performance of ML models depends on sample size.
 The more complex the model, the more data it will need.
-The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.
+Our dataset had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models.
+Our analysis was limited to shallow learning methods and did not explore deep learning methods such as neural networks.
+Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting.
+There are studies that address overcoming these challenges in biomedical datasets [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], however sudies that estalish frameworks with microbiome data are lacking.
+This would be an interesting direction for future work in microbiome studies.
 
-This study highlights the need to make educated choices at every step of developing a ML model with microbiome data.
+This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data.
 Model selection should be done with a solid understanding of model complexity and interpretability, rigorous ML pipelines should be built with cross-validation for hyperparameter tuning and with a held-out test set for evaluating predictive performance and models should be interpreted while considering collinearity in datasets.
@@
 
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\seperating hyperplane that discriminates between labels.} & Linear & Interpratable & $^{37}$\\ \hline
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable & $^{37}$\\ \hline
 
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal gaussian \\seperating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
 Decision tree & \makecell[l]{A classifier that sorts samples down from the 
@@
 \makecell[l]{Model \\ selection} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem?
-\\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretibility \\ does the problem need?
+\\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretability \\ does the problem need?
 \\ How much each \\ candidate algorithm \\ can provide?} \\ \hline

### 2019-07-22 submission/manuscript.tex
update manuscript for Pat

@@
 \usepackage{caption}
-\usepackage{makecell}
-\usepackage{threeparttable}
 
@@
 models are scarce due to the flawed methods that call the validity of
-developed ML models into question. Furthermore, the use of black box ML
+developed models into question. Furthermore, the use of black box ML
 models has hindered the validation of microbial biomarkers. To overcome
@@
 random forest, extreme gradient boosting (XGBoost) and neural networks
-without a discussion on if and how much model interpretibility is
+without a discussion on if and how much model interpretability is
 necessary for the study (11, 21--23). These models are also called black
@@
 linear models, the rankings of these features were different due to the
-collinearity in microbial communities.
+collinearity in microbial communities. Collinearity in a microbial
+dataset is when one OTU is dependant on another OTU. The feature weights
+of correlated OTUs are influenced by one another which makes it
+difficult to interpret models using feature weights. Our interpretation
+of L2-regularized logistic regression based on feature weight rankings,
+showed that an OTU that belongs to \emph{Lachnospiraceae} (OTU00056) was
+not among the most important OTUs in making a prediction {[}Figure
+3C{]}. However, it was picked out as important when the same model was
+interpreted using permutation importance {[}Figure S3C{]}. This is due
+to collinearity in the dataset, where \emph{Lachnospiraceae} (OTU00056)
+is significantly correlated with \emph{Feacalibacterium} (OTU00015)
+which has the highest ranked feature weights {[}Figure 3C{]}. We need to
+investigate the relationships among features to identify the true
+underlying factors when making a prediction.
 
@@
 We compared the training times of the seven ML models. As the complexity
-of a ML model {[}Table S1{]} and the number of tuned hyperparameter
-settings increased {[}Figures S1-S2{]}, its training times increased as
-well {[}Figure 5{]}. Linear models trained faster than non-linear
-models. L1 and L2 SVM with linear kernel and L2-regularized logistic
+of a ML model and the number of tuned hyperparameter settings increased
+{[}Figures S1-S2{]}, its training times increased as well {[}Figure
+5{]}. Linear models trained faster than non-linear models. L1 and
+L2-regularized SVM with linear kernel and L2-regularized logistic
 regression had the shortest training times with 0.2 hours, (std ± 0.03),
@@
 
-In this study we established a rigorous ML pipeline to use 16S rRNA
-sequence counts to predict a binary health outcome. We built on others'
-work in the microbiome field to set-up standards for developing and
-evaluating ML models for microbiome data (28--30). First, we used a
-held-out test set to illustrate the difference between cross-validation
-and testing AUROC values. When the difference between cross-validation
-and test performance is low, this suggest the models are not overfit and
-that they will perform similar with similar data. In all seven models,
-the difference between median cross-validation and testing AUROC values
-did not exceed 0.021 which suggests that these models are generalizable
-and can be used to test similar new data. Second, we performed the
-initial 80\%-20\% random datasplit 100 times in our ML pipeline. The
-randomization of the initial data-split to create a held-out test set is
-a crucial step in the ML pipeline to develop robust ML models and to
-report reliable performance metrics. Depending on how the data is split,
-there is the chance of being overoptimistic about the predictive
-performance of a model. In our study, we showed that there was
-variability in AUROC values between different random data-splits in each
-of the models we tested. Our results showed that the testing AUROC
-values varied 0.23 on average between different data-splits. Third, we
-used the AUROC metric in our study to evaluate the predictive
-performance of the ML models. AUROC is always random at the value 0.5
-and is a robust metric when a dataset is imbalanced. We also performed a
-full grid search for hyperparameter settings when building a ML model.
-Default hyperparameter settings in previously developed ML packages in
-R, Python, and Matlab programming languages are inadequate for effective
+Microbiome studies use ML models, often with a classification task, to
+predict a disease but also to learn which microbes are indicators of
+that disease (2--11). Achieving either of these tasks have far-reaching
+impact on human health, however ML as a tool in microbiome studies is
+still at its infancy. A framework is needed to develop rigorous ML
+models, to identify and overcome potential pitfalls. Previous studies
+generated workflows to allow ML to be widely used by the microbiome
+researchers (28--31). This study sets-up standards for developing and
+evaluating rigorous ML models for microbiome data {[}Table 1{]}.
+
+We benchmarked seven ML models with different classification algorithms
+to show that a clearly defined ML problem that is based on the goal of
+the microbiome study should inform our model selection. Our results
+showed that if the goal of a study is to learn the ecology behind a
+disease and to identify microbial biomarkers, we can create ML models
+that are inherently interpretable and easily trained without losing
+predictive power. In terms of predictive performance, random forest
+model had the best testing AUROC values compared to the other six
+models. However, the second best model was L2-regularized logistic
+regression with a median AUROC difference of only 0.015 compared to
+random forest. While random forest took 83.2 hours to train,
+L2-regularized logistic regression trained in 12 minutes. In terms of
+interpretability, random forest was a more complex ML model and could
+only be explained using methods such as permutation importance. On the
+other hand, L2-regularized logistic regression was easier to interpret
+by ranking absolute feature weights of the trained model.
+
+Even with interpretable models such as L2-regularized logistic
+regression, there are potential pitfalls when it comes to identifying
+biomarkers of a disease. As domain experts, we know that
+human-associated microbial communities have complex correlation
+structures that creates collinearity in the dataset. Collinearity is a
+severe problem and needs to be addressed to reliably interpret ML models
+(32). In this study we used two different methods to interpret our
+linear models; ranking each OTU by (1) their absolute weights in the
+trained models and (2) their impact on the predictive performance based
+on permutation importance. We observed differences in the OTU rankings
+between the two interpretation methods due to collinearity in the
+dataset. To avoid misinterpreting the models, once we identify the most
+important microbes, we should check for their relationships with other
+microbes as well. These relationships will help us generate new
+hypotheses about the ecology of the disease. These hypotheses needs to
+be tested with follow-up experiments to identify the true biomarkers of
+a disease.
+
+In this study, we also established a rigorous ML pipeline to use 16S
+rRNA sequence counts to predict a binary health outcome. First of all,
+we used a held-out test set to illustrate the difference between
+cross-validation and testing AUROC values. When the difference between
+cross-validation and test performance is low, this suggest the models
+are not overfit and that they will perform similar with similar data. In
+all the models, the difference between median cross-validation and
+testing AUROC values did not exceed 0.021 which suggests that these
+models are generalizable and can be used to test similar new data.
+Furthermore, we performed the initial 80\%-20\% random datasplit 100
+times in our ML pipeline. Depending on how the data is split, there is
+the chance of being overoptimistic about the predictive performance of a
+model. We showed that there was variability in AUROC values between
+different random data-splits in each of the models we tested. Our
+results showed that the testing AUROC values varied 0.23 on average
+between different data-splits. The randomization and resampling of the
+initial data-split to create a held-out test set is a crucial step in
+the ML pipeline to develop robust ML models and to report reliable
+performance metrics. We also performed a full grid search for
+hyperparameter settings when training our ML models. Default
+hyperparameter settings in previously developed ML packages in R,
+Python, and Matlab programming languages are inadequate for effective
 application of classification algorithms and need to be optimized for
@@
 was susceptible to performing poorly if the wrong regularization
-coefficient was assigned to the model by default.
-
-In this study, we benchmarked seven ML models with different
-classification algorithms to show that we should use ML models based on
-the goal of the study and our expectations of predictive performance,
-interpretability and computational burden. Microbiome studies use ML
-models with a classification task to learn the training data to assign
-labels, such as healthy or not to new data, but also to learn which
-features are important to discriminate between labels (2--11). Our
-results show that if the goal of a study is to learn the ecology behind
-a disease and to identify microbial biomarkers, we could create ML
-models that are inherently interpretable without losing predictive
-power. In terms of predictive performance, random forest model had the
-best testing AUROC values compared to the other 6 models. However, the
-second best model was L2-regularized logistic regression with a median
-AUROC difference of only 0.015 compared to random forest. In terms of
-interpretability, random forest was a complex ML model and could only be
-explained using methods such as permutation importance. On the other
-hand, L2-regularized logistic regression was easier to interpret
-(i.e.~ranking absolute regression coefficients of the trained model).
-
-Even with interpretable models such as L2-regularized logistic
-regression, we need to be careful with how we treat the information we
-get from the models. In this study we used two different methods to
-interpret our linear models; ranking each OTU by (1) their absolute
-weights in the trained models and (2) their impact on the predictive
-performance based on permutation importance. We observed differences in
-the OTU rankings between the two interpretation methods due to
-collinearity in the dataset. Collinearity in a microbial dataset is when
-one OTU is highly correlated with another OTU which causes feature
-weights to not be unique. The feature weights of correlated OTUs are
-influenced by one another which makes it difficult to interpret the ML
-model. To avoid misinterpreting the models we should use the highly
-ranked correlated OTUs to generate hypotheses about the ecology of the
-disease and test them with follow-up experiments.
-
-There are other criteria when choosing ML models such as the
-computational burden of developing it and the sample size. In terms of
-computational burden, random forest model trained each data-split in
-83.2 hours whereas L2-regularized logistic regression trained in 12
-minutes. The generalization performance of ML models depends on sample
-size. The more complex the model, the more data it will need. The
-dataset we used for our study had 490 samples, however microbiome
-studies that have smaller sample sizes would benefit from using less
-complex models such as L2-regularized logistic regression.
-
-This study highlights the need to make educated choices at every step of
-developing a ML model with microbiome data. Model selection should be
+coefficient was assigned to the model by default. And finally, we used
+the AUROC metric in our study to evaluate the predictive performance of
+the ML models. AUROC is always random at the value 0.5 and is a robust
+metric when a dataset is imbalanced.
+
+We used a balanced CRC dataset to develop ML models with a binary
+classification task. We did not evaluate multicategory classification
+methods or regression analyses to predict non-binary outcomes. However,
+the principles highlighted throughout this study {[}Table 1{]} apply to
+all ML modeling tasks with microbiome data. The models we built were
+generalizable despite the high number of features microbiome datasets
+usually have. The generalization performance of ML models depends on
+sample size. The more complex the model, the more data it will need. Our
+dataset had 490 samples, however microbiome studies that have smaller
+sample sizes would benefit from using less complex models. Our analysis
+was limited to shallow learning methods and did not explore deep
+learning methods such as neural networks. Microbiome datasets often
+suffer from having high dimensionality but low sample sizes which makes
+deep learning models prone to overfitting. There are studies that
+address overcoming these challenges in biomedical datasets (11, 33, 34),
+however sudies that estalish frameworks with microbiome data are
+lacking. This would be an interesting direction for future work in
+microbiome studies.
+
+This study highlighted the need to make educated choices at every step
+of developing a ML model with microbiome data. Model selection should be
 done with a solid understanding of model complexity and
@@
 patients recruited by Great Lakes-New England Early Detection Research
-Network study. These data were obtained from Sze et al (31). The stool
+Network study. These data were obtained from Sze et al (35). The stool
 samples were provided by recruited adult participants who were
@@
 assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(32--34).
+(36--38).
 
@@
 testing performance. We can use hyper-band to help us with our
-hyperparameter selection (35).
+hyperparameter selection (39).
 
@@
 
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\seperating hyperplane that discriminates between labels.} & Linear & Interpratable & $^{37}$\\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal gaussian \\seperating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable & $^{37}$\\ \hline
+
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
 Decision tree & \makecell[l]{A classifier that sorts samples down from the 
@@
 
+\hypertarget{ref-vangay_microbiome_2019}{}
+31. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
+Microbiome learning repo (ML repo): A public repository of microbiome
+regression and classification tasks. Gigascience \textbf{8}.
+doi:\href{https://doi.org/10.1093/gigascience/giz042}{10.1093/gigascience/giz042}.
+
+\hypertarget{ref-dormann_collinearity:_2013}{}
+32. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
+JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
+\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
+\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
+\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
+of methods to deal with it and a simulation study evaluating their
+performance. Ecography \textbf{36}:27--46.
+doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.
+
+\hypertarget{ref-kocheturov_massive_2019}{}
+33. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+2019. Massive datasets and machine learning for computational
+biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
+doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.
+
+\hypertarget{ref-kim_improved_2018}{}
+34. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+method for prediction of cancer prognosis by network learning. Genes
+\textbf{9}:478.
+doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.
+
 \hypertarget{ref-sze_leveraging_2018}{}
-31. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+35. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-32. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+36. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+37. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-34. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+38. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-35. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+39. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-08-01 submission/manuscript.Rmd
fix a couple of sentences
before Pat's edits
update

@@
 
-Advances in sequencing technology and decreases in the costs of generating 16S rRNA gene sequences have allowed rapid exploration of the human microbiome and its health implications.
-Currently, the human microbiome field is growing at an unprecedented rate and there is an increasing demand for methods that identify associations between microbiome members and human health.
-However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease.
-It is unlikely that a single species can explain a disease.
-Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.
+As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease.
+However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
+This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals.
+Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease.
+Instead, subsets of the microbiome account for differences in health.
+Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually.
+Recently, machine learning models have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.
 
-Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets.
-They learn from existing data to predict the outcomes of new data and allow us to infer on the reasons underlying that prediction.
-Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-The high-stake task of predicting the correct diagnosis with high confidence relies on a ML model that is built with rigorous methods.
-However, the field's use of ML lacks clarity and consistency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019].
-More notably, we commonly see flawed ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate predictive performance or reporting few or only the best outcomes of cross-validation.
-Even when there are seperate testing sets to evaluate predictive performance, there are large differences between cross-validation and testing performances and large confidence intervals in testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017].
-These practices prevent the development of generalizable models, where the model makes accurate predictions with newly acquired data as well as it does with the training data.
+ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data.
+Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods.
+However,  ML methodology in the microbiome field has flaws that have not been addressed previously.
+There is a lack of transparency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019]; models are being developed and evaluated without a seperate held-out test data [@flemer_oral_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @montassier_pretreatment_2016; @ai_systematic_2017]; there is large variation between the cross-validation outcomes [@dai_multi-cohort_2018; @ai_systematic_2017] and between cross-validation and testing performances [@mossotto_classification_2017] .
+These practices limit reproducibility, cause overoptimism for model performance and prevent the development of generalizable models, where the model makes accurate predictions with newly acquired data as well as it does with the training data.
+Nevertheless, the microbiome field is making progress to avoid potential pitfalls of ML.
+More and more studies are now validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and they are introducing frameworks to accurately use ML tools [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
 
-Moreover, there is a lack of discussion on why a particular ML model is utilized and a lack of emphasis on the strengths and weaknesses of it.
-Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting (XGBoost) and neural networks without a discussion on if and how much model interpretability is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].
-These models are also called black box ML models because they are not inherently interpretable and require posthoc explanations to determine the feature importances in making a prediction.
-These explanations can be misleading and at times unreliable when making high-stake decisions about patient health [@rudin_please_2018].
-The models we develop for healthcare, both to predict disease and to understand underlying reasons behind that prediction, should be transparent and accountable [@rudin_optimized_2018].
+Among microbiome researchers, there has been a trend towards using more complex ML models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @@wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
+There is an implicit assumption that the more complex models are better because they are more complex.
+Although these models may be better at incorporating non-linear relationships or yield better predictions, they are also called black box ML models because they are not inherently interpretable.
+These models require post hoc explanations to quantify the importance of each feature in making a prediction.
+Depending on the application of the model researchers may chose to use different modeling approaches.
+For example, a researcher trying to identify the populations causing a disease would likely want a more interpretable model whereas a clinician may emphasize performance.
+Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018].
 
-The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibility.
-We need to strive toward better machine learning practices by (1) implementing rigorous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of predictive performance, complexity, interpretability and computational efficiency.
-To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline.
-We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences of CRC patients.
-The human microbiome is hypothesized to directly contribute to the development of CRC and fecal 16S rRNA gene sequences have been used to detect it [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
-We built seven ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN).
-We established modeling pipelines for three linear models with different forms of regularization; L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear kernel.
-We also developed four non-linear models; SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
+The lack of transparency in model selection and using flawed modelling and interpretation methods negatively impacts model validity.
+To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches using the same dataset and  pipeline.
+We established modeling pipelines for three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
+We also developed four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
 We compared the predictive performance, interpretability and computational efficiency of the seven models to highlight the importance of model selection.
-We established standards for ML pipeline construction, predictive performance evaluation and model interpretation for microbiome-associated ML models that have a binary prediction task.
+To demonstrate the performance of these modelling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
+This dataset was selected because it is a relatively large collection of samples (N=490) connected to clinically significant disease where there is ample evidence that it is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
+With this dataset, we established standards for ML pipeline construction, evaluated predictive performance, and demonstrated model interpretation for models.
+This framework can be easily applied to other host-associated and environmental microbiome problems.
 
@@
 
-__Model selection and pipeline construction__
+__Model selection and pipeline construction__ We established a ML pipeline where we train and validate each of the seven models [Figure 1].
+We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
+Since the cases are not uniformly represented in the data, the data-split was stratified to maintain the overall label distribution in both the training and test sets.
+For example, our example dataset represented data from 490 individuals, 261 had normal colons and 229 had a screen relevant neoplasia (SRN).
+After the data-split, the training set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
+The training data was used for hyperparameter selection and the test set was used for evaluating predictive performance.
 
-We first determined the dataset and the ML models to use in our study.
-We used a previously published study on a CRC cohort of 490 patients with 261 cases of SRN.
-For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (normal or SRN colorectal lesions as defined by colonoscopies).
-We established modeling pipelines for a binary prediction task with L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosted decision tree (XGBoost).
-
-We used ML models with different classification algorithms and regularization parameters.
+We used ML models with different classification algorithms and regularization methods.
+Regularization is a technique to discourage overfitting by penalizing the model for learning the training data too well.
 For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features.
@@
 
-We established a ML pipeline where we train and validate each of the seven models [Figure 1].
-We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
-Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets.
-Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
-The training/validation data was used for training purposes and validation of hyperparameter selection, and the test set was used for evaluation purposes.
-
 Hyperparameters are the rules that are learned from the training data in a classification algorithm.
-When they are tuned over a full grid search and selected by validation, they make the ML model better are predicting.
-We selected hyperparameter settings using repeated five-fold cross-validation on the training/validation set [Figure 1].
-We chose the best hyperparameter setting for each model based on its predictive performance on the validation set using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2].
+For example regularization term (C) that decides how big the penalty for overfitting will be, is an hyperparameter.
+The model becomes more generalizable when we tune for the optimal C value.
+Similar to regularization term C, other hyperparameters can be tuned over a full grid search and selected by validation to build better models.
+We selected hyperparameter settings by performing repeated five-fold cross-validation (CV) on the training set [Figure 1].
+Similar to the initial data-split, five-fold CV was also stratified to maintain the overall label distribution.
+We chose the best hyperparameter setting for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2].
 The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance.
-Similar to the initial data-split, five-fold cross-validation for hyperparameter selection was also stratified to maintain the overall label distribution on the training and validation sets.
 The cross-validation of each hyperparameter setting was repeated over 100 randomizations to get a robust reading of predictive performance.
 
-We then trained the full training/validation dataset with the selected hyperparameters.
-We used the held-out test set to evaluate the predictive performance of each ML model.
+We then trained the full training dataset with the selected hyperparameters.
+We used the held-out test set to evaluate the testing predictive performance of each ML model.
 The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].
@@
 The randomization and resampling of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics.
-We also performed a full grid search for hyperparameter settings when training our ML models.
+Additionally, we performed a full grid search for hyperparameter settings when training our ML models.
 Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model.
@@
 The study had 261 normal and 229 SRN samples.
-The bacterial abundances are the features used to predict colonic health of the patients.
-Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
+The bacterial abundances are the features used to predict colorectal health of the patients.
+For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (normal or SRN colorectal lesions as defined by colonoscopies).
+We established modeling pipelines for a binary prediction task Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
 OTU counts were set to the size of our smallest sample and were subsampled at the same distances.

### 2019-08-02 submission/manuscript.Rmd
changes before meeting 2 with Pat

@@
 output:
-  pdf_document:
-    includes:
-      in_header: header.tex
-    keep_tex: yes
   word_document: default
@@
 
-__Predictive performance and generalizability of the seven models.__
-
-We evaluated the predictive performances of seven binary classification models when applied to held-out test data using the AUROC metric [Figure 2].
+__Predictive performance and generalizability of the seven models.__ We evaluated the predictive performances of seven binary classification models when applied to held-out test data using the AUROC metric [Figure 2].
 Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01).
@@
 
-__Interpretation of each ML model.__
-
-Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017].
+__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017].
 Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies.

### 2019-08-02 submission/manuscript.tex
changes before meeting 2 with Pat

@@
 
-Advances in sequencing technology and decreases in the costs of
-generating 16S rRNA gene sequences have allowed rapid exploration of the
-human microbiome and its health implications. Currently, the human
-microbiome field is growing at an unprecedented rate and there is an
-increasing demand for methods that identify associations between
-microbiome members and human health. However, this is difficult as human
-associated microbial communities are remarkably complex,
-high-dimensional and uneven within and between individuals with the same
-disease. It is unlikely that a single species can explain a disease.
-Instead, subsets of those communities, in relation to one another and to
-their host, account for differences in health outcomes.
-
-Machine learning (ML) methods are effective at recognizing and
-highlighting patterns in complex microbial datasets. They learn from
-existing data to predict the outcomes of new data and allow us to infer
-on the reasons underlying that prediction. Therefore, researchers have
-started to explore the utility of ML models that use microbiota
-associated biomarkers to predict human health and to understand the
+As the number of people represented in human microbiome datasets grow,
+there is an increasing desire to use microbiome data to diagnose
+disease. However, the structure of the human microbiome is remarkably
+variable between individuals to the point where it is often difficult to
+identify the bacterial populations that are associated with diseases
+using traditional statistical models. This variation is likely due to
+the ability of many bacterial populations to fill the same niche such
+that different populations cause the same disease in different
+individuals. Furthermore, a growing number of studies have shown that it
+is rare for a single bacterial species to be associated with a disease.
+Instead, subsets of the microbiome account for differences in health.
+Traditional statistical approaches do not adequately account for the
+variation in the human microbiome and typically consider the protective
+or risk effects of each bacterial population individually. Recently,
+machine learning models have grown in popularity among microbiome
+researchers because of the large amount of data that can now be
+generated and because the models are effective at accounting for the
+interpersonal microbiome variation and the ecology of the disease.
+
+ML models are useful for understanding the variation in the structure of
+existing data and to apply that knowledge to make predictions about new
+data. Researchers have used ML models to diagnose and understand the
 ecological basis of diseases such as liver cirrhosis, colorectal cancer,
 inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
-The high-stake task of predicting the correct diagnosis with high
-confidence relies on a ML model that is built with rigorous methods.
-However, the field's use of ML lacks clarity and consistency in which
-methods are used and how these methods are implemented (12, 13). More
-notably, we commonly see flawed ML practices such as using ML pipelines
-where there is no seperate held-out test dataset to evaluate predictive
-performance or reporting few or only the best outcomes of
-cross-validation. Even when there are seperate testing sets to evaluate
-predictive performance, there are large differences between
-cross-validation and testing performances and large confidence intervals
-in testing performances (4, 14--20). These practices prevent the
-development of generalizable models, where the model makes accurate
-predictions with newly acquired data as well as it does with the
-training data.
-
-Moreover, there is a lack of discussion on why a particular ML model is
-utilized and a lack of emphasis on the strengths and weaknesses of it.
-Recently, there is a trend towards using more complex ML models such as
-random forest, extreme gradient boosting (XGBoost) and neural networks
-without a discussion on if and how much model interpretability is
-necessary for the study (11, 21--23). These models are also called black
-box ML models because they are not inherently interpretable and require
-posthoc explanations to determine the feature importances in making a
-prediction. These explanations can be misleading and at times unreliable
-when making high-stake decisions about patient health (24). The models
-we develop for healthcare, both to predict disease and to understand
-underlying reasons behind that prediction, should be transparent and
-accountable (25).
-
-The lack of transparency on model selection and interpretation as well
-as flawed modeling methods negatively impact model validity and
-reproducibility. We need to strive toward better machine learning
-practices by (1) implementing rigorous machine learning pipelines and
-(2) selecting ML models that reflect the goal of the study as it will
-inform our expectations of predictive performance, complexity,
-interpretability and computational efficiency. To showcase a rigorous ML
-pipeline and to shed light on how ML model selection can affect modeling
-results, we performed an empirical analysis comparing several different
-ML models using the same dataset and the same ML pipeline. We used a
-previously published colorectal cancer (CRC) study (3) which had fecal
-16S rRNA gene sequences of CRC patients. The human microbiome is
-hypothesized to directly contribute to the development of CRC and fecal
-16S rRNA gene sequences have been used to detect it (1, 3, 4, 26). We
-built seven ML models using fecal 16S rRNA gene sequences to predict
-healthy patients versus patients with colorectal lesions that were
-identified by colonoscopy as screen relevant neoplasias (SRN). We
-established modeling pipelines for three linear models with different
-forms of regularization; L2-regularized logistic regression, L1 and
-L2-regularized support vector machines (SVM) with linear kernel. We also
-developed four non-linear models; SVM with radial basis function kernel,
-a decision tree, random forest and XGBoost. We compared the predictive
-performance, interpretability and computational efficiency of the seven
-models to highlight the importance of model selection. We established
-standards for ML pipeline construction, predictive performance
-evaluation and model interpretation for microbiome-associated ML models
-that have a binary prediction task.
+The task of diagnosing an individual with high confidence relies on a ML
+model that is built with rigorous methods. However, ML methodology in
+the microbiome field has flaws that have not been addressed previously.
+There is a lack of transparency in which methods are used and how these
+methods are implemented (12, 13); models are being developed and
+evaluated without a seperate held-out test data (3, 4, 14--16); there is
+large variation between the cross-validation outcomes (16, 17) and
+between cross-validation and testing performances (18) . These practices
+limit reproducibility, cause overoptimism for model performance and
+prevent the development of generalizable models, where the model makes
+accurate predictions with newly acquired data as well as it does with
+the training data. Nevertheless, the microbiome field is making progress
+to avoid potential pitfalls of ML. More and more studies are now
+validating their models on independent datasets (7, 18, 19) and they are
+introducing frameworks to accurately use ML tools (20--23).
+
+Among microbiome researchers, there has been a trend towards using more
+complex ML models such as random forest and neural networks (2, 11,
+24--26) over simpler models such as logistic regression or other linear
+models (18, 22, 27). There is an implicit assumption that the more
+complex models are better because they are more complex. Although these
+models may be better at incorporating non-linear relationships or yield
+better predictions, they are also called black box ML models because
+they are not inherently interpretable. These models require post hoc
+explanations to quantify the importance of each feature in making a
+prediction. Depending on the application of the model researchers may
+chose to use different modeling approaches. For example, a researcher
+trying to identify the populations causing a disease would likely want a
+more interpretable model whereas a clinician may emphasize performance.
+Although one may feel that they are sacrificing interpretability for
+performance, that tradeoff may be minimal (28, 29).
+
+The lack of transparency in model selection and using flawed modelling
+and interpretation methods negatively impacts model validity. To
+showcase a rigorous ML pipeline and to shed light on how ML model
+selection can affect modeling results, we performed an empirical
+analysis comparing 7 modeling approaches using the same dataset and
+pipeline. We established modeling pipelines for three linear models with
+different forms of regularization: L2-regularized logistic regression
+and L1 and L2-regularized support vector machines (SVM) with a linear
+kernel. We also developed four non-linear models: SVM with radial basis
+function kernel, a decision tree, random forest and XGBoost. We compared
+the predictive performance, interpretability and computational
+efficiency of the seven models to highlight the importance of model
+selection. To demonstrate the performance of these modelling approaches
+and our pipeline, we used data from a previously published study that
+sought to classifiy individuals as having normal colons or colonic
+lesions based on the 16S rRNA gene sequences collected from fecal
+samples (3). This dataset was selected because it is a relatively large
+collection of samples (N=490) connected to clinically significant
+disease where there is ample evidence that it is driven by variation in
+the microbiome (1, 3, 4, 30). With this dataset, we established
+standards for ML pipeline construction, evaluated predictive
+performance, and demonstrated model interpretation for models. This
+framework can be easily applied to other host-associated and
+environmental microbiome problems.
 
@@
 
-\textbf{Model selection and pipeline construction}
-
-We first determined the dataset and the ML models to use in our study.
-We used a previously published study on a CRC cohort of 490 patients
-with 261 cases of SRN. For each patient, we had 6920 features (fecal
-bacterial abundances) and a two-class label that defines their
-colorectal health (normal or SRN colorectal lesions as defined by
-colonoscopies). We established modeling pipelines for a binary
-prediction task with L2-regularized logistic regression, L1 and L2
-support vector machines (SVM) with linear and radial basis function
-kernels, a decision tree, random forest and extreme gradient boosted
-decision tree (XGBoost).
+\textbf{Model selection and pipeline construction} We established a ML
+pipeline where we train and validate each of the seven models {[}Figure
+1{]}. We randomly split the data into training and test sets so that the
+training set consisted of 80\% of the full dataset while the test set
+was composed of the remaining data {[}Figure 1{]}. Since the cases are
+not uniformly represented in the data, the data-split was stratified to
+maintain the overall label distribution in both the training and test
+sets. For example, our example dataset represented data from 490
+individuals, 261 had normal colons and 229 had a screen relevant
+neoplasia (SRN). After the data-split, the training set consisted of 393
+patients (209 SRN), while the test set was composed of 97 patients (52
+SRN). The training data was used for hyperparameter selection and the
+test set was used for evaluating predictive performance.
 
 We used ML models with different classification algorithms and
-regularization parameters. For regularized logistic regression and SVM
-with linear kernel, we used L2 regularization to keep all potentially
-important features. For comparison, we also trained an L1 regularized
-SVM model with linear kernel. L1-regularization on microbiome data lead
-to a sparser solution (i.e., force many coefficients to zero). Finally,
-to explore the potential for non-linear relationships among features and
-the outcome of interest, we trained tree based models, decision tree,
-random forest and XGBoost, as well as an SVM with non-linear kernel.
-
-We established a ML pipeline where we train and validate each of the
-seven models {[}Figure 1{]}. We randomly split the data into
-training/validation and test sets so that the training/validation set
-consisted of 80\% of the full dataset while the test set was composed of
-the remaining data {[}Figure 1{]}. Since the cases are not uniformly
-represented in the data, the initial data-split was stratified to
-maintain the overall label distribution in both the training/validation
-and test sets. Training/validation set consisted of 393 patients (209
-SRN), while the test set was composed of 97 patients (52 SRN). The
-training/validation data was used for training purposes and validation
-of hyperparameter selection, and the test set was used for evaluation
-purposes.
+regularization methods. Regularization is a technique to discourage
+overfitting by penalizing the model for learning the training data too
+well. For regularized logistic regression and SVM with linear kernel, we
+used L2 regularization to keep all potentially important features. For
+comparison, we also trained an L1 regularized SVM model with linear
+kernel. L1-regularization on microbiome data lead to a sparser solution
+(i.e., force many coefficients to zero). Finally, to explore the
+potential for non-linear relationships among features and the outcome of
+interest, we trained tree based models, decision tree, random forest and
+XGBoost, as well as an SVM with non-linear kernel.
 
 Hyperparameters are the rules that are learned from the training data in
-a classification algorithm. When they are tuned over a full grid search
-and selected by validation, they make the ML model better are
-predicting. We selected hyperparameter settings using repeated five-fold
-cross-validation on the training/validation set {[}Figure 1{]}. We chose
-the best hyperparameter setting for each model based on its predictive
-performance on the validation set using the area under the receiver
-operating characteristic curve (AUROC) metric {[}Figure S1 and S2{]}.
-The AUROC ranges from 1.0, where the model perfectly distinguishes
-between cases and controls, to 0.50, where the model's predictions are
-no different from random chance. Similar to the initial data-split,
-five-fold cross-validation for hyperparameter selection was also
-stratified to maintain the overall label distribution on the training
-and validation sets. The cross-validation of each hyperparameter setting
-was repeated over 100 randomizations to get a robust reading of
-predictive performance.
-
-We then trained the full training/validation dataset with the selected
-hyperparameters. We used the held-out test set to evaluate the
+a classification algorithm. For example regularization term (C) that
+decides how big the penalty for overfitting will be, is an
+hyperparameter. The model becomes more generalizable when we tune for
+the optimal C value. Similar to regularization term C, other
+hyperparameters can be tuned over a full grid search and selected by
+validation to build better models. We selected hyperparameter settings
+by performing repeated five-fold cross-validation (CV) on the training
+set {[}Figure 1{]}. Similar to the initial data-split, five-fold CV was
+also stratified to maintain the overall label distribution. We chose the
+best hyperparameter setting for each model based on its CV predictive
+performance using the area under the receiver operating characteristic
+curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC ranges from 1.0,
+where the model perfectly distinguishes between cases and controls, to
+0.50, where the model's predictions are no different from random chance.
+The cross-validation of each hyperparameter setting was repeated over
+100 randomizations to get a robust reading of predictive performance.
+
+We then trained the full training dataset with the selected
+hyperparameters. We used the held-out test set to evaluate the testing
 predictive performance of each ML model. The data-split, hyperparameter
@@
 \textbf{Predictive performance and generalizability of the seven
-models.}
-
-We evaluated the predictive performances of seven binary classification
-models when applied to held-out test data using the AUROC metric
-{[}Figure 2{]}. Random forest had significantly higher test AUROC values
-than the other models for detecting SRNs when AUROC values were compared
-to the other six by Wilcoxon rank sum test (p \textless{} 0.01). The
-median AUROC of the random forest model was 0.695 (IQR 0.044).
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with
-linear and radial basis function kernel AUROC values were not
+models.} We evaluated the predictive performances of seven binary
+classification models when applied to held-out test data using the AUROC
+metric {[}Figure 2{]}. Random forest had significantly higher test AUROC
+values than the other models for detecting SRNs when AUROC values were
+compared to the other six by Wilcoxon rank sum test (p \textless{}
+0.01). The median AUROC of the random forest model was 0.695 (IQR
+0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
+with linear and radial basis function kernel AUROC values were not
 significantly different from one another. They had median AUROC values
@@
 
-\textbf{Interpretation of each ML model.}
-
-Interpretability is the degree to which humans can understand the
-reasons behind a model prediction (27). Because we often use ML models
-not just to predict a health outcome but also to learn the ecology
-behind a disease, model interpretation becomes crucial for microbiome
-studies. The ML models we built using L2-regularized logistic
-regression, L1 and L2 support vector machines (SVM) with linear and
-radial basis function kernels, a decision tree, random forest and
-XGBoost decrease in interpretability as they increase in complexity. In
-this study we highlighted two methods to interpret models with varying
-complexity.
+\textbf{Interpretation of each ML model.} Interpretability is the degree
+to which humans can understand the reasons behind a model prediction
+(31). Because we often use ML models not just to predict a health
+outcome but also to learn the ecology behind a disease, model
+interpretation becomes crucial for microbiome studies. The ML models we
+built using L2-regularized logistic regression, L1 and L2 support vector
+machines (SVM) with linear and radial basis function kernels, a decision
+tree, random forest and XGBoost decrease in interpretability as they
+increase in complexity. In this study we highlighted two methods to
+interpret models with varying complexity.
 
@@
 generated workflows to allow ML to be widely used by the microbiome
-researchers (28--31). This study sets-up standards for developing and
+researchers (20--23). This study sets-up standards for developing and
 evaluating rigorous ML models for microbiome data {[}Table 1{]}.
@@
 the ML pipeline to develop robust ML models and to report reliable
-performance metrics. We also performed a full grid search for
+performance metrics. Additionally, we performed a full grid search for
 hyperparameter settings when training our ML models. Default
@@
 study had 261 normal and 229 SRN samples. The bacterial abundances are
-the features used to predict colonic health of the patients. Bacterial
-abundances are discrete data in the form of Operational Taxonomic Unit
-(OTU) counts. OTU counts were set to the size of our smallest sample and
-were subsampled at the same distances. They were then transformed by
-scaling to a {[}0-1{]} range.
+the features used to predict colorectal health of the patients. For each
+patient, we had 6920 features (fecal bacterial abundances) and a
+two-class label that defines their colorectal health (normal or SRN
+colorectal lesions as defined by colonoscopies). We established modeling
+pipelines for a binary prediction task Bacterial abundances are discrete
+data in the form of Operational Taxonomic Unit (OTU) counts. OTU counts
+were set to the size of our smallest sample and were subsampled at the
+same distances. They were then transformed by scaling to a {[}0-1{]}
+range.
 
@@
 area under the receiver operating characteristic curve; SVM, support
-vector machine; XGBoost, extreme gradient boosting.\\
-\newpage
+vector machine; XGBoost, extreme gradient boosting. \newpage
 \includegraphics{Figure_S1.png}
@@
 
+\hypertarget{ref-montassier_pretreatment_2016}{}
+15. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
+\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
+P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
+2016. Pretreatment gut microbiome predicts chemotherapy-related
+bloodstream infection. Genome Medicine \textbf{8}:49.
+doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.
+
+\hypertarget{ref-ai_systematic_2017}{}
+16. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
+\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
+supervised classifiers for fecal microbiota-based prediction of
+colorectal cancer. Oncotarget \textbf{8}:9546--9556.
+doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.
+
 \hypertarget{ref-dai_multi-cohort_2018}{}
-15. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
+17. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
 WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
@@
 
-\hypertarget{ref-montassier_pretreatment_2016}{}
-16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
-\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
-P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
-2016. Pretreatment gut microbiome predicts chemotherapy-related
-bloodstream infection. Genome Medicine \textbf{8}:49.
-doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.
-
-\hypertarget{ref-papa_non-invasive_2012}{}
-17. \textbf{Papa E}, \textbf{Docktor M}, \textbf{Smillie C},
-\textbf{Weber S}, \textbf{Preheim SP}, \textbf{Gevers D},
-\textbf{Giannoukos G}, \textbf{Ciulla D}, \textbf{Tabbaa D},
-\textbf{Ingram J}, \textbf{Schauer DB}, \textbf{Ward DV},
-\textbf{Korzenik JR}, \textbf{Xavier RJ}, \textbf{Bousvaros A},
-\textbf{Alm EJ}. 2012. Non-invasive mapping of the gastrointestinal
-microbiota identifies children with inflammatory bowel disease. PLOS ONE
-\textbf{7}:e39242.
-doi:\href{https://doi.org/10.1371/journal.pone.0039242}{10.1371/journal.pone.0039242}.
-
 \hypertarget{ref-mossotto_classification_2017}{}
@@
 
-\hypertarget{ref-ai_systematic_2017}{}
-19. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
-\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
-supervised classifiers for fecal microbiota-based prediction of
-colorectal cancer. Oncotarget \textbf{8}:9546--9556.
-doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.
-
 \hypertarget{ref-wong_quantitation_2017}{}
-20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
+19. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
 AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
@@
 
-\hypertarget{ref-galkin_human_2018}{}
-21. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
-\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
-2018. Human microbiome aging clocks based on deep learning and tandem of
-permutation feature importance and accumulated local effects. bioRxiv.
-doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
-
-\hypertarget{ref-reiman_using_2017}{}
-22. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
-convolutional neural networks to explore the microbiome, pp. 4269--4272.
-\emph{In} 2017 39th annual international conference of the IEEE
-engineering in medicine and biology society (EMBC).
-
-\hypertarget{ref-fioravanti_phylogenetic_2017}{}
-23. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
-\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
-\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
-in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
-
-\hypertarget{ref-rudin_please_2018}{}
-24. \textbf{Rudin C}. 2018. Please stop explaining black box models for
-high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
-
-\hypertarget{ref-rudin_optimized_2018}{}
-25. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
-Toward trust in machine learning for healthcare and criminal justice.
-Interfaces \textbf{48}:449--466.
-doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.
-
-\hypertarget{ref-knights_human-associated_2011}{}
-26. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
-\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
-signatures: Examining their predictive value. Cell Host Microbe
-\textbf{10}:292--296.
-doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.
-
-\hypertarget{ref-miller_explanation_2017}{}
-27. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
-Insights from the social sciences. arXiv:170607269 {[}cs{]}.
-
 \hypertarget{ref-statnikov_comprehensive_2013}{}
-28. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
+20. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
 \textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
@@
 \hypertarget{ref-knights_supervised_2011}{}
-29. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
+21. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
 Supervised classification of human microbiota. FEMS Microbiology Reviews
@@
 \hypertarget{ref-wirbel_meta-analysis_2019}{}
-30. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
+22. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
 K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
@@
 \hypertarget{ref-vangay_microbiome_2019}{}
-31. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
+23. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
 Microbiome learning repo (ML repo): A public repository of microbiome
@@
 doi:\href{https://doi.org/10.1093/gigascience/giz042}{10.1093/gigascience/giz042}.
+
+\hypertarget{ref-galkin_human_2018}{}
+24. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
+2018. Human microbiome aging clocks based on deep learning and tandem of
+permutation feature importance and accumulated local effects. bioRxiv.
+doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
+
+\hypertarget{ref-reiman_using_2017}{}
+25. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+convolutional neural networks to explore the microbiome, pp. 4269--4272.
+\emph{In} 2017 39th annual international conference of the IEEE
+engineering in medicine and biology society (EMBC).
+
+\hypertarget{ref-fioravanti_phylogenetic_2017}{}
+26. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
+\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
+in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
+
+\hypertarget{ref-thomas_metagenomic_2019}{}
+27. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
+\textbf{Pasolli E}, \textbf{Armanini F}, \textbf{Zolfo M},
+\textbf{Beghini F}, \textbf{Manara S}, \textbf{Karcher N}, \textbf{Pozzi
+C}, \textbf{Gandini S}, \textbf{Serrano D}, \textbf{Tarallo S},
+\textbf{Francavilla A}, \textbf{Gallo G}, \textbf{Trompetto M},
+\textbf{Ferrero G}, \textbf{Mizutani S}, \textbf{Shiroma H},
+\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
+T}, \textbf{Wirbel J}, \textbf{Schrotz-King P}, \textbf{Ulrich CM},
+\textbf{Brenner H}, \textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller
+G}, \textbf{Cordero F}, \textbf{Dias-Neto E}, \textbf{Setubal JC},
+\textbf{Tett A}, \textbf{Pardini B}, \textbf{Rescigno M},
+\textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N}. 2019.
+Metagenomic analysis of colorectal cancer datasets identifies
+cross-cohort microbial diagnostic signatures and a link with choline
+degradation. Nature Medicine \textbf{25}:667.
+doi:\href{https://doi.org/10.1038/s41591-019-0405-7}{10.1038/s41591-019-0405-7}.
+
+\hypertarget{ref-rudin_please_2018}{}
+28. \textbf{Rudin C}. 2018. Please stop explaining black box models for
+high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
+
+\hypertarget{ref-rudin_optimized_2018}{}
+29. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
+Toward trust in machine learning for healthcare and criminal justice.
+Interfaces \textbf{48}:449--466.
+doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.
+
+\hypertarget{ref-knights_human-associated_2011}{}
+30. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
+\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
+signatures: Examining their predictive value. Cell Host Microbe
+\textbf{10}:292--296.
+doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.
+
+\hypertarget{ref-miller_explanation_2017}{}
+31. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
+Insights from the social sciences. arXiv:170607269 {[}cs{]}.
 

### 2019-08-02 submission/manuscript.Rmd
Changes after meeing 2 with Pat

@@
 
-As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease.
+As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to.
 However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
@@
 ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data.
-Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
+Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019, @flemer_oral_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
 The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods.
-However,  ML methodology in the microbiome field has flaws that have not been addressed previously.
-There is a lack of transparency in which methods are used and how these methods are implemented [@thaiss_persistent_2016; @dadkhah_gut_2019]; models are being developed and evaluated without a seperate held-out test data [@flemer_oral_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @montassier_pretreatment_2016; @ai_systematic_2017]; there is large variation between the cross-validation outcomes [@dai_multi-cohort_2018; @ai_systematic_2017] and between cross-validation and testing performances [@mossotto_classification_2017] .
-These practices limit reproducibility, cause overoptimism for model performance and prevent the development of generalizable models, where the model makes accurate predictions with newly acquired data as well as it does with the training data.
-Nevertheless, the microbiome field is making progress to avoid potential pitfalls of ML.
-More and more studies are now validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and they are introducing frameworks to accurately use ML tools [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
+However, there are common methodological problems across many of these studies that need to be addressed as the field progresses.
+These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances.
+Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing analysis frameworks to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
+More work is needed to further improve reproducibility and minimize over-optimism for model performance.
 
-Among microbiome researchers, there has been a trend towards using more complex ML models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @@wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
-There is an implicit assumption that the more complex models are better because they are more complex.
-Although these models may be better at incorporating non-linear relationships or yield better predictions, they are also called black box ML models because they are not inherently interpretable.
-These models require post hoc explanations to quantify the importance of each feature in making a prediction.
-Depending on the application of the model researchers may chose to use different modeling approaches.
-For example, a researcher trying to identify the populations causing a disease would likely want a more interpretable model whereas a clinician may emphasize performance.
+Among microbiome researchers the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex.
+This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @@wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
+Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable.
+These models require post hoc explanations to quantify the importance of each feature in making a prediction and they do not show the structure of how the features are used.
+Depending on the application of the model, researchers may choose to use different modeling approaches.
+For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize performance.
 Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018].
+Regardless, it is important for researchers to articulate why they have selected a specific modelling approach or even compare multiple approaches in the same study.
 
-The lack of transparency in model selection and using flawed modelling and interpretation methods negatively impacts model validity.
-To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches using the same dataset and  pipeline.
-We established modeling pipelines for three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
-We also developed four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
-We compared the predictive performance, interpretability and computational efficiency of the seven models to highlight the importance of model selection.
-To demonstrate the performance of these modelling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
-This dataset was selected because it is a relatively large collection of samples (N=490) connected to clinically significant disease where there is ample evidence that it is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
-With this dataset, we established standards for ML pipeline construction, evaluated predictive performance, and demonstrated model interpretation for models.
-This framework can be easily applied to other host-associated and environmental microbiome problems.
+To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches with the same dataset and pipeline.
+We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
+We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
+We compared the predictive performance, interpretability, and computational efficiency.
+To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
+This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
+With this dataset we developed a framework that implementf a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models.
+This framework can be easily applied to other host-associated and environmental microbiome datasets.
 
@@
 
-__Model selection and pipeline construction__ We established a ML pipeline where we train and validate each of the seven models [Figure 1].
-We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1].
-Since the cases are not uniformly represented in the data, the data-split was stratified to maintain the overall label distribution in both the training and test sets.
-For example, our example dataset represented data from 490 individuals, 261 had normal colons and 229 had a screen relevant neoplasia (SRN).
-After the data-split, the training set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN).
-The training data was used for hyperparameter selection and the test set was used for evaluating predictive performance.
+__Model selection and pipeline construction__ We established a ML pipeline where we trained and validated each of the seven models using a common approach that is based on standard methods within the ML community (REFS)[Figure 1].
 
-We used ML models with different classification algorithms and regularization methods.
-Regularization is a technique to discourage overfitting by penalizing the model for learning the training data too well.
+First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining 20% of the data [Figure 1].
+To maintain the distribution of controls and cases that was found with the full dataset, we performed stratified splits.
+For example, our full dataset included 490 individuals.
+Of these, 261 had normal colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%).
+A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%).
+The training data was used to build the models and the test set was used for evaluating predictive performance.
+
+Second, we trained seven different models using the training data.
+We selected models with different classification algorithms and regularization methods.
+Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well.
 For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features.
 For comparison, we also trained an L1 regularized SVM model with linear kernel.
-L1-regularization on microbiome data lead to a sparser solution (i.e., force many coefficients to zero).
-Finally, to explore the potential for non-linear relationships among features and the outcome of interest, we trained tree based models, decision tree, random forest and XGBoost, as well as an SVM with non-linear kernel.
+L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero).
+To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with non-linear kernel.
 
-Hyperparameters are the rules that are learned from the training data in a classification algorithm.
-For example regularization term (C) that decides how big the penalty for overfitting will be, is an hyperparameter.
-The model becomes more generalizable when we tune for the optimal C value.
-Similar to regularization term C, other hyperparameters can be tuned over a full grid search and selected by validation to build better models.
-We selected hyperparameter settings by performing repeated five-fold cross-validation (CV) on the training set [Figure 1].
-Similar to the initial data-split, five-fold CV was also stratified to maintain the overall label distribution.
-We chose the best hyperparameter setting for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2].
-The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model's predictions are no different from random chance.
-The cross-validation of each hyperparameter setting was repeated over 100 randomizations to get a robust reading of predictive performance.
+Third, fitting of these models require selecting appropriate hyperparameters.
+Hyperparameters are the rules that are learned from the training set in a classification algorithm.
+For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting.
+Similar to regularization term C, all hyperparameters are tuned to find the best model.
+We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1].
+The five-fold CV was also stratified to maintain the overall case and control distribution.
+We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2].
+The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance.
 
-We then trained the full training dataset with the selected hyperparameters.
-We used the held-out test set to evaluate the testing predictive performance of each ML model.
+Finally, we trained the full training dataset with the selected hyperparameter values and applied the model to the held-out data to evaluate the testing predictive performance of each model.
 The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].
 
-__Predictive performance and generalizability of the seven models.__ We evaluated the predictive performances of seven binary classification models when applied to held-out test data using the AUROC metric [Figure 2].
-Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p < 0.01).
+__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of seven models to classify individuals as having normal colons or SRNs [Figure 2].
+The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01).
 The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another.
-They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively.
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively.
 L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
-Random forest had the highest median AUROC for detecting SRN.
-Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance.
+Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularlized logistic regression and linear SVM) performed nearly as well as random forest.
 
 To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC.
-The difference between the two should be low to suggest the model is not overfitting despite the large number of features.
-The largest difference between the two was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
-We also reported the testing AUROC values over 100 randomizations of the initial data-split.
-The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
+If the difference between the cross-validation and testing AUROCs is large, then that would indicate that the models were overfit.
+difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.
+
+To evaluate the risk for over-optimism of each model, we calculated the range for each model using 100 splits.
+The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
+If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model.
 For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.
 These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
+Therefore, it is important to employ the hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.
 
@@
 Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies.
-The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretability as they increase in complexity.
-In this study we highlighted two methods to interpret models with varying complexity.
+ML models decrease in interpretability as they increase in complexity.
+In this study we used two methods to interpret models with varying complexity.
 
 We interpreted linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the absolute feature weights of the trained models.
-We ranked the absolute weights of all the OTUs for each data-split [Figure 3].
+We ranked the absolute values of the weights of all the OTUs for each data-split [Figure 3].
 We calculated the median ranks of these features over the 100 data-splits.
-In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belonged to families *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3].
-Some of the OTUs with the highest ranks were shared among the linear models.
+In the three linear models, the OTUs that had the largest median rank and drove the detection of SRNs contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU00822, OTU00768), Clostridium (OTU00609), Blautia (OTU00659, OTU00012), Faecalibacterium (OTU00015), Gamella (OTU00426) and Peptostreptococcus (OTU00367) [Figure 3].
+Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that … were shared among both of the L2-regularized models.
 

### 2019-08-07 submission/manuscript.Rmd
changes after meeting 2

@@
 output:
-  word_document: default
+  word_document:
 geometry: margin=1.0in
@@
 
-As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to.
+As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases.
 However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
@@
 ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data.
-Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019, @flemer_oral_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
+Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
 The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods.
@@
 Among microbiome researchers the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex.
-This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @@wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
+This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
 Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable.
@@
 This would be an interesting direction for future work in microbiome studies.
+
+Paragraph about deploying a model.
 

### 2019-08-07 submission/manuscript.tex
changes after meeting 2

@@
 there is an increasing desire to use microbiome data to diagnose
-disease. However, the structure of the human microbiome is remarkably
+diseases. However, the structure of the human microbiome is remarkably
 variable between individuals to the point where it is often difficult to
@@
 ecological basis of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
-The task of diagnosing an individual with high confidence relies on a ML
-model that is built with rigorous methods. However, ML methodology in
-the microbiome field has flaws that have not been addressed previously.
-There is a lack of transparency in which methods are used and how these
-methods are implemented (12, 13); models are being developed and
-evaluated without a seperate held-out test data (3, 4, 14--16); there is
-large variation between the cross-validation outcomes (16, 17) and
-between cross-validation and testing performances (18) . These practices
-limit reproducibility, cause overoptimism for model performance and
-prevent the development of generalizable models, where the model makes
-accurate predictions with newly acquired data as well as it does with
-the training data. Nevertheless, the microbiome field is making progress
-to avoid potential pitfalls of ML. More and more studies are now
-validating their models on independent datasets (7, 18, 19) and they are
-introducing frameworks to accurately use ML tools (20--23).
-
-Among microbiome researchers, there has been a trend towards using more
-complex ML models such as random forest and neural networks (2, 11,
-24--26) over simpler models such as logistic regression or other linear
-models (18, 22, 27). There is an implicit assumption that the more
-complex models are better because they are more complex. Although these
-models may be better at incorporating non-linear relationships or yield
-better predictions, they are also called black box ML models because
-they are not inherently interpretable. These models require post hoc
-explanations to quantify the importance of each feature in making a
-prediction. Depending on the application of the model researchers may
-chose to use different modeling approaches. For example, a researcher
-trying to identify the populations causing a disease would likely want a
-more interpretable model whereas a clinician may emphasize performance.
+inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--16,
+16--18). The task of diagnosing an individual with high confidence
+relies on a ML model that is built with rigorous methods. However, there
+are common methodological problems across many of these studies that
+need to be addressed as the field progresses. These include a lack of
+transparency in which methods are used and how these methods are
+implemented; developing and evaluating models without a separate
+held-out test data; large variation between the predictive performance
+on different folds of cross-validation; and large variation between
+cross-validation and testing performances. Nevertheless, the microbiome
+field is making progress to avoid some of these pitfalls including
+validating their models on independent datasets (7, 18, 19) and
+introducing analysis frameworks to better use ML tools (20--23). More
+work is needed to further improve reproducibility and minimize
+over-optimism for model performance.
+
+Among microbiome researchers the lack of transparency in justifying a
+modelling approach has been due to an implicit assumption that more
+complex models are better because they are more complex. This has
+resulted in a trend towards using models such as random forest and
+neural networks (2, 11, 24--26) over simpler models such as logistic
+regression or other linear models (18, 22, 27). Although the more
+complex models may be better at incorporating non-linear relationships
+or yield better predictions, they are considered to be black box models
+because they are not inherently interpretable. These models require post
+hoc explanations to quantify the importance of each feature in making a
+prediction and they do not show the structure of how the features are
+used. Depending on the application of the model, researchers may choose
+to use different modeling approaches. For example, researchers trying to
+identify the populations causing a disease would likely want a more
+interpretable model whereas clinicians may emphasize performance.
 Although one may feel that they are sacrificing interpretability for
-performance, that tradeoff may be minimal (28, 29).
-
-The lack of transparency in model selection and using flawed modelling
-and interpretation methods negatively impacts model validity. To
-showcase a rigorous ML pipeline and to shed light on how ML model
+performance, that tradeoff may be minimal (28, 29). Regardless, it is
+important for researchers to articulate why they have selected a
+specific modelling approach or even compare multiple approaches in the
+same study.
+
+To showcase a rigorous ML pipeline and to shed light on how ML model
 selection can affect modeling results, we performed an empirical
-analysis comparing 7 modeling approaches using the same dataset and
-pipeline. We established modeling pipelines for three linear models with
-different forms of regularization: L2-regularized logistic regression
-and L1 and L2-regularized support vector machines (SVM) with a linear
-kernel. We also developed four non-linear models: SVM with radial basis
-function kernel, a decision tree, random forest and XGBoost. We compared
-the predictive performance, interpretability and computational
-efficiency of the seven models to highlight the importance of model
-selection. To demonstrate the performance of these modelling approaches
-and our pipeline, we used data from a previously published study that
-sought to classifiy individuals as having normal colons or colonic
-lesions based on the 16S rRNA gene sequences collected from fecal
-samples (3). This dataset was selected because it is a relatively large
-collection of samples (N=490) connected to clinically significant
-disease where there is ample evidence that it is driven by variation in
-the microbiome (1, 3, 4, 30). With this dataset, we established
-standards for ML pipeline construction, evaluated predictive
-performance, and demonstrated model interpretation for models. This
-framework can be easily applied to other host-associated and
-environmental microbiome problems.
+analysis comparing 7 modeling approaches with the same dataset and
+pipeline. We built three linear models with different forms of
+regularization: L2-regularized logistic regression and L1 and
+L2-regularized support vector machines (SVM) with a linear kernel. We
+also built four non-linear models: SVM with radial basis function
+kernel, a decision tree, random forest and XGBoost. We compared the
+predictive performance, interpretability, and computational efficiency.
+To demonstrate the performance of these modeling approaches and our
+pipeline, we used data from a previously published study that sought to
+classifiy individuals as having normal colons or colonic lesions based
+on the 16S rRNA gene sequences collected from fecal samples (3). This
+dataset was selected because it is a relatively large collection of
+individuals (N=490) connected to a clinically significant disease where
+there is ample evidence that the disease is driven by variation in the
+microbiome (1, 3, 4, 30). With this dataset we developed a framework
+that implementf a ML pipeline that can be used for any modeling
+approach, evaluates predictive performance, and demonstrates how to
+interpret these models. This framework can be easily applied to other
+host-associated and environmental microbiome datasets.
 
@@
 \textbf{Model selection and pipeline construction} We established a ML
-pipeline where we train and validate each of the seven models {[}Figure
-1{]}. We randomly split the data into training and test sets so that the
-training set consisted of 80\% of the full dataset while the test set
-was composed of the remaining data {[}Figure 1{]}. Since the cases are
-not uniformly represented in the data, the data-split was stratified to
-maintain the overall label distribution in both the training and test
-sets. For example, our example dataset represented data from 490
-individuals, 261 had normal colons and 229 had a screen relevant
-neoplasia (SRN). After the data-split, the training set consisted of 393
-patients (209 SRN), while the test set was composed of 97 patients (52
-SRN). The training data was used for hyperparameter selection and the
-test set was used for evaluating predictive performance.
-
-We used ML models with different classification algorithms and
-regularization methods. Regularization is a technique to discourage
+pipeline where we trained and validated each of the seven models using a
+common approach that is based on standard methods within the ML
+community (REFS){[}Figure 1{]}.
+
+First, we randomly split the data into training and test sets so that
+the training set consisted of 80\% of the full dataset while the test
+set was composed of the remaining 20\% of the data {[}Figure 1{]}. To
+maintain the distribution of controls and cases that was found with the
+full dataset, we performed stratified splits. For example, our full
+dataset included 490 individuals. Of these, 261 had normal colons (53\%)
+and 229 had a screen relevant neoplasia (SRN; 46.7\%). A training set
+included 393 individuals, of which 209 had an SRN (53\%), while the test
+set was composed of 97 individuals of which 52 had an SRN (54\%). The
+training data was used to build the models and the test set was used for
+evaluating predictive performance.
+
+Second, we trained seven different models using the training data. We
+selected models with different classification algorithms and
+regularization methods. Regularization is a technique that discourages
 overfitting by penalizing the model for learning the training data too
@@
 comparison, we also trained an L1 regularized SVM model with linear
-kernel. L1-regularization on microbiome data lead to a sparser solution
-(i.e., force many coefficients to zero). Finally, to explore the
-potential for non-linear relationships among features and the outcome of
-interest, we trained tree based models, decision tree, random forest and
-XGBoost, as well as an SVM with non-linear kernel.
-
-Hyperparameters are the rules that are learned from the training data in
-a classification algorithm. For example regularization term (C) that
-decides how big the penalty for overfitting will be, is an
-hyperparameter. The model becomes more generalizable when we tune for
-the optimal C value. Similar to regularization term C, other
-hyperparameters can be tuned over a full grid search and selected by
-validation to build better models. We selected hyperparameter settings
-by performing repeated five-fold cross-validation (CV) on the training
-set {[}Figure 1{]}. Similar to the initial data-split, five-fold CV was
-also stratified to maintain the overall label distribution. We chose the
-best hyperparameter setting for each model based on its CV predictive
-performance using the area under the receiver operating characteristic
-curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC ranges from 1.0,
-where the model perfectly distinguishes between cases and controls, to
-0.50, where the model's predictions are no different from random chance.
-The cross-validation of each hyperparameter setting was repeated over
-100 randomizations to get a robust reading of predictive performance.
-
-We then trained the full training dataset with the selected
-hyperparameters. We used the held-out test set to evaluate the testing
-predictive performance of each ML model. The data-split, hyperparameter
-selection, training and testing steps were repeated 100 times to get a
-reliable and robust reading of model performance {[}Figure 1{]}.
+kernel. L1-regularization on microbiome data led to a sparser solution
+(i.e., force many coefficients to zero). To explore the potential for
+non-linear relationships among features to improve classification, we
+trained tree-based models including decision tree, random forest, and
+XGBoost and we trained an SVM model with non-linear kernel.
+
+Third, fitting of these models require selecting appropriate
+hyperparameters. Hyperparameters are the rules that are learned from the
+training set in a classification algorithm. For example, in the linear
+models the regularization term (C) is a hyperparameter that indicates
+the penalty for overfitting. Similar to regularization term C, all
+hyperparameters are tuned to find the best model. We selected
+hyperparameters by performing 100 five-fold cross-validation (CV)
+repeats on the training set {[}Figure 1{]}. The five-fold CV was also
+stratified to maintain the overall case and control distribution. We
+chose the best hyperparameter values for each model based on its CV
+predictive performance using the area under the receiver operating
+characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
+ranges from 1.0, where the model perfectly distinguishes between cases
+and controls, to 0.50, where the model's predictions are no different
+from random chance.
+
+Finally, we trained the full training dataset with the selected
+hyperparameter values and applied the model to the held-out data to
+evaluate the testing predictive performance of each model. The
+data-split, hyperparameter selection, training and testing steps were
+repeated 100 times to get a reliable and robust reading of model
+performance {[}Figure 1{]}.
 
 \textbf{Predictive performance and generalizability of the seven
-models.} We evaluated the predictive performances of seven binary
-classification models when applied to held-out test data using the AUROC
-metric {[}Figure 2{]}. Random forest had significantly higher test AUROC
-values than the other models for detecting SRNs when AUROC values were
-compared to the other six by Wilcoxon rank sum test (p \textless{}
+models.} We evaluated the predictive performance of seven models to
+classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
+random forest model had significantly higher test AUROC values than the
+other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
 0.01). The median AUROC of the random forest model was 0.695 (IQR
@@
 with linear and radial basis function kernel AUROC values were not
-significantly different from one another. They had median AUROC values
-of 0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
+significantly different from one another and had median AUROC values of
+0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
 0.056) respectively. L1-regularized SVM with linear kernel and decision
@@
 median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
-{[}Figure 2{]}. Random forest had the highest median AUROC for detecting
-SRN. Despite its simplicity, the L2-regularized logistic regression was
-second best in predictive performance.
+{[}Figure 2{]}. Interestingly, these results demonstrate that the most
+complex model (XGBoost) did not have the best performance and that the
+most interpretable models (L2-regularlized logistic regression and
+linear SVM) performed nearly as well as random forest.
 
 To evaluate the generalizability of each model, we compared the median
-cross-validation AUROC to the median testing AUROC. The difference
-between the two should be low to suggest the model is not overfitting
-despite the large number of features. The largest difference between the
-two was 0.021 in L1-regularized SVM with linear kernel, followed by SVM
-with radial basis function kernel and decision tree with a difference of
-0.007 and 0.006, respectively {[}Figure 2{]}. We also reported the
-testing AUROC values over 100 randomizations of the initial data-split.
-The testing AUROC values within each model varied 0.23 on average across
-the seven models. For instance, the lowest AUROC value of the random
-forest model was 0.59 whereas the highest was 0.81. These results showed
-that depending on the data-split, the testing AUROC values showed great
-variability {[}Figure 2{]}.
+cross-validation AUROC to the median testing AUROC. If the difference
+between the cross-validation and testing AUROCs is large, then that
+would indicate that the models were overfit. difference in median AUROCs
+was 0.021 in L1-regularized SVM with linear kernel, followed by SVM with
+radial basis function kernel and decision tree with a difference of
+0.007 and 0.006, respectively {[}Figure 2{]}; however, these differences
+are relatively small and would not indicate a problem with overfitting.
+
+To evaluate the risk for over-optimism of each model, we calculated the
+range for each model using 100 splits. The range among the testing AUROC
+values within each model varied by 0.23 on average across the seven
+models. If we had only done a single split, then there is a risk that we
+could gotten lucky or unlucky with the performance of the model. For
+instance, the lowest AUROC value of the random forest model was 0.59
+whereas the highest was 0.81. These results showed that depending on the
+data-split, the testing AUROC values showed great variability {[}Figure
+2{]}. Therefore, it is important to employ the hierarchical data splits
+that were included in our pipeline to minimize the risk of
+over-optimism.
 
@@
 outcome but also to learn the ecology behind a disease, model
-interpretation becomes crucial for microbiome studies. The ML models we
-built using L2-regularized logistic regression, L1 and L2 support vector
-machines (SVM) with linear and radial basis function kernels, a decision
-tree, random forest and XGBoost decrease in interpretability as they
-increase in complexity. In this study we highlighted two methods to
-interpret models with varying complexity.
+interpretation becomes crucial for microbiome studies. ML models
+decrease in interpretability as they increase in complexity. In this
+study we used two methods to interpret models with varying complexity.
 
@@
 kernel and L2-regularized logistic regression) using the absolute
-feature weights of the trained models. We ranked the absolute weights of
-all the OTUs for each data-split {[}Figure 3{]}. We calculated the
-median ranks of these features over the 100 data-splits. In the three
-linear models, OTUs that had the largest median ranks and drove the
-detection of SRNs belonged to families \emph{Lachnospiraceae}, and
-\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
-OTU00015, OTU00768, OTU00822, OTU00609), genera \emph{Gamella}
-(OTU00426) and genera \emph{Peptostreptococcus} (OTU00367) {[}Figure
-3{]}. Some of the OTUs with the highest ranks were shared among the
-linear models.
+feature weights of the trained models. We ranked the absolute values of
+the weights of all the OTUs for each data-split {[}Figure 3{]}. We
+calculated the median ranks of these features over the 100 data-splits.
+In the three linear models, the OTUs that had the largest median rank
+and drove the detection of SRNs contained sequences that were most
+similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia
+(OTU00822, OTU00768), Clostridium (OTU00609), Blautia (OTU00659,
+OTU00012), Faecalibacterium (OTU00015), Gamella (OTU00426) and
+Peptostreptococcus (OTU00367) {[}Figure 3{]}. Among the five highest
+ranked OTUs across these three models, it was encouraging that OTU 1239
+was shared among each of the models and that \ldots{} were shared among
+both of the L2-regularized models.
 
@@
 
+Paragraph about deploying a model.
+
 This study highlighted the need to make educated choices at every step

### 2019-08-07 submission/manuscript.Rmd
update after meeting with Pat 3

@@
 To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC.
-If the difference between the cross-validation and testing AUROCs is large, then that would indicate that the models were overfit.
-difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.
+If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit.
+The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.
+Subsample experiment results to show what overfitting looks like goes here…
 
-To evaluate the risk for over-optimism of each model, we calculated the range for each model using 100 splits.
-The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models.
+To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 splits.
+The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` on average across the seven models.
 If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model.
-For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.
+For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`.
 These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
@@
 ML models decrease in interpretability as they increase in complexity.
-In this study we used two methods to interpret models with varying complexity.
+In this study we used two methods to help interpret our models.
 
-We interpreted linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the absolute feature weights of the trained models.
-We ranked the absolute values of the weights of all the OTUs for each data-split [Figure 3].
-We calculated the median ranks of these features over the 100 data-splits.
-In the three linear models, the OTUs that had the largest median rank and drove the detection of SRNs contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU00822, OTU00768), Clostridium (OTU00609), Blautia (OTU00659, OTU00012), Faecalibacterium (OTU00015), Gamella (OTU00426) and Peptostreptococcus (OTU00367) [Figure 3].
-Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that … were shared among both of the L2-regularized models.
+We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median absolute feature weights for each OTU in the models [Figure 3].
+The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) and Peptostreptococcus (OTU 367) [Figure 3].
+Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that 4 OTUs were shared among the five highest ranked OTUs for both of the L2-regularized models.
 
-We explained the feature importances in non-linear models; SVM with radial basis kernel, decision tree, random forest and XGBoost, using a method called permutation importance on the held-out test data.
-Permutation importance analysis is a posthoc explanation of the model where we randomly permute non-correlated features individually and groups of highly correlated features together.
-We then calculate how much the predictive performance of the model (i.e AUROC values) decrease when each OTU or group of OTUs is permuted randomly.
-We ranked the OTUs based on how much they decreased the median testing AUROC; the OTU with the largest decrease ranking highest.
-The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4].
-Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most in all tree-based methods [Figure 4].
-Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` median AUROC to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].
+We interpreted the feature importance in the non-linear models using permutation importance.
+Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups.
+We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
+We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest.
+The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) [Figure 4].
+Among the twenty OTUs with the largest impact across these four models, there was only one OTU (OTU 822) that was shared among all of the models; however, among the tree-based models, we found three OTUs (OTU 367, OTU 58 and OTU 110) to be consistently important.
+Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs were shared between the models.
 
-To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models as well [Figure S3].
-L1-regularized SVM with linear kernel picked out some of the same OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings based on weights [Figure 3] and permutation importance [Figure S3].
-Similarly, L2-regularized SVM and L2-regularized logistic regression picked out some of the same OTUs in both interpretation methods, OTU00659 and OTU00012, respectively.
-However, for all the linear models, the rankings of these features were different due to the collinearity in microbial communities.
-Collinearity in a microbial dataset is when one OTU is dependant on another OTU.
-The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret models using feature weights.
-Our interpretation of L2-regularized logistic regression based on feature weight rankings, showed that an OTU that belongs to *Lachnospiraceae* (OTU00056) was not among the most important OTUs in making a prediction [Figure 3C].
-However, it was picked out as important when the same model was interpreted using permutation importance [Figure S3C].
-This is due to collinearity in the dataset, where *Lachnospiraceae* (OTU00056) is significantly correlated with *Feacalibacterium* (OTU00015) which has the highest ranked feature weights [Figure 3C].
-We need to investigate the relationships among features to identify the true underlying factors when making a prediction.
+When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out many of the same OTUs (e.g.
+OTU00822, OTU01239, OTU00609) that were important when we interpreted the model using feature rankings based on weights [Figure 3].
+Similarly, L2-regularized SVM and L2-regularized logistic regression  each identified many OTUs that were selected by both interpretation methods.
+
+Interestingly, for the all the models, except the poorly performing L1-regularized SVM and decision tree, the variation in the impact of these important OTUs was minimal.
+This supports the findings of others that a classification requires numerous OTUs, rather the presence of a single or a few of the OTUs [@sze_leveraging_2018].
 
@@
 We compared the training times of the seven ML models.
-As the complexity of a ML model and the number of tuned hyperparameter settings increased [Figures S1-S2], its training times increased as well [Figure 5].
-Linear models trained faster than non-linear models.
-L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression had the shortest training times with `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively.
-Whereas, a decision tree, SVM with radial basis function kernel, random forest and XGBoost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].
+As expected, the training times increased as the complexity of a ML model and the number of tuned hyperparameter settings increased and the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
+When we subsetted the size of the training dataset, we observed a [linear, quadratic, etc] relationship between the size of the dataset and the training time for the random forest model.
 
 ## Discussion
+
+Overview
+•	Establish pipeline that can be generalized to any modeling method
+•	Linear/non-linear not a bfd
+•	Interpretability tradeoff
+Point 1
+•	Importance for transparency, picking the most complex model is not necessarilty the best model
+•	Good / better / best:  Table S1
+•	“best” has multiple meanings
+Point 2
+•	Our paper is concerned with developing models to gain greater biological insights into what’s going on
+•	Still need biological validation and further experimentation
+Point 3
+•	These models show great opportunity to use microbiome to make diagnoses
+•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
+•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
+•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
+•	We also have a problem with OTU-based approaches
+Conclusion/Future direction
+•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out.
+We encourage reseraches to apply multiple models to their data and report the results
+•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret
 

### 2019-08-16 submission/manuscript.Rmd
changes to fig 3

@@
 
-We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median absolute feature weights for each OTU in the models [Figure 3].
+We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU in the models [Figure 3].
 The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) and Peptostreptococcus (OTU 367) [Figure 3].

### 2019-08-16 code/learning/Figure3.R

@@ -47,7 +47,7 @@ plot_feature_ranks <- function(data){
     # Rank 1 is the highest rank
     plot <- ggplot(data, aes(reorder(data$key, -data$rank, FUN = median), data$rank)) +
       geom_point(aes(colour= factor(data$sign)), size=2) + # datapoints lighter color
-      scale_color_manual(values=c("red3", "#56B4E9", "#999999")) +
+      scale_color_manual(values=c("#56B4E9","red3", "#999999")) +
       stat_summary(fun.y = function(x) median(x), colour = 'black', geom = "point", size = 3) + # Median darker
       coord_flip() +
       scale_y_continuous(limits=c(0, 100)) + # Only keep the first 100 ranks (truncated)
@@ -86,6 +86,11 @@ get_taxa_info_as_labels <- function(data){
     select(-OTU)
 
   taxa_otus <- inner_join(otus, taxa_info, by="key") %>% 
+    mutate_if(is.character, str_to_upper) %>% 
+    #mutate(key=str_replace(key, "0000", " ")) %>% 
+    #mutate(key=str_replace(key, "000", " ")) %>% 
+    #mutate(key=str_replace(key, "00", " ")) %>% 
+    #mutate(key=str_replace(key, "0", " ")) %>% 
     mutate(taxa=gsub("(.*);.*","\\1",Taxonomy)) %>% 
     mutate(taxa=gsub("(.*)_.*","\\1",Taxonomy)) %>% 
     mutate(taxa=gsub("(.*);.*","\\1",Taxonomy)) %>% 

### 2019-08-16 code/learning/compare_sumbsampled_perf.R

@@ -53,6 +53,9 @@ rf_performance <- map_df(rf_files, read_files) %>%
   unite_("model_number", c("model","number")) %>% 
   filter(Performance != "testing")
 
+rf_performance$model_number <-  as.factor(rf_performance$model_number)
+
+rf_performance$model_number <- factor(rf_performance$model_number , c("Random_Forest_490", "Random_Forest_245", "Random_Forest_120", "Random_Forest_60", "Random_Forest_30", "Random_Forest_15"))
 ######################################################################
 #Plot the AUC values for cross validation and testing for each model #
 ######################################################################
@@ -62,7 +65,7 @@ logit_plot <- ggplot(logit_performance, aes(x = model_number, y = AUC)) +
   geom_boxplot(alpha=0.5, fatten = 4, fill="blue4") +
   geom_hline(yintercept = 0.5, linetype="dashed") +
   coord_flip() +
-  scale_y_continuous(name = "AUROC",
+  scale_y_continuous(name = "cross-validation AUROC",
                      breaks = seq(0, 1, 0.1),
                      limits=c(0, 1),
                      expand=c(0,0)) +
@@ -102,6 +105,11 @@ logit_plot <- ggplot(logit_performance, aes(x = model_number, y = AUC)) +
 ggsave("Figure_S5.png", plot = logit_plot, device = 'png', path = 'submission', width = 12, height = 9)
 
 
+
+
+
+
+
 rf_plot <- ggplot(rf_performance, aes(x = model_number, y = AUC)) +
   geom_boxplot(alpha=0.5, fatten = 4, fill="blue4") +
   geom_hline(yintercept = 0.5, linetype="dashed") +
@@ -110,7 +118,6 @@ rf_plot <- ggplot(rf_performance, aes(x = model_number, y = AUC)) +
                      breaks = seq(0, 1, 0.1),
                      limits=c(0, 1),
                      expand=c(0,0)) +
-  
   scale_x_discrete(name = "Random forest", 
                    labels=c("n=490",
                             "n=245", 

### 2019-08-16 code/learning/main.R

@@ -93,7 +93,7 @@ data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
     Dx_Bin== "adv Adenoma" ~ "cancer",
     Dx_Bin== "Cancer" ~ "cancer"
   )) %>%
-  select(-sample, -Dx_Bin, -fit_result) %>%
+  select(-sample, -Dx_Bin) %>%
   drop_na() 
 
 # We want the diagnosis column to be a factor

### 2019-08-19 submission/manuscript.Rmd
changes to interp results

--- 2019-08-07 submission/manuscript.Rmd
@@
 output:
-  word_document:
-geometry: margin=1.0in
+  pdf_document:
+    includes:
+      in_header: header.tex
+    keep_tex: yes
 bibliography: references.bib
@@
 The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.
-Subsample experiment results to show what overfitting looks like goes here…
 
@@
 
+To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our sample size from 490 to 245, 120, 60, 30 and 15 [Figure S4].
+The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when full sample size was used to train and validate the models.
+These results showed that because microbiome data has many features (6920 OTUs), it is important to train models using appropriate sample sizes to avoid generalizability issues.
+
 __Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017].
@@
 
-We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median absolute feature weights for each OTU in the models [Figure 3].
-The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) and Peptostreptococcus (OTU 367) [Figure 3].
-Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that 4 OTUs were shared among the five highest ranked OTUs for both of the L2-regularized models.
+We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights of each OTU [Figure 3].
+The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) Peptostreptococcus (OTU 367) and others [Figure 3].
+Among the highest ranked OTUs across these three models, it was encouraging that many OTUs (e.g.
+OTU 822, OTU 1239, OTU 426, OTU 609, OTU 50) were shared among all the models.
 
 We interpreted the feature importance in the non-linear models using permutation importance.
+Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
 Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups.
@@
 We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest.
-The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) [Figure 4].
+The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) and others [Figure 4].
 Among the twenty OTUs with the largest impact across these four models, there was only one OTU (OTU 822) that was shared among all of the models; however, among the tree-based models, we found three OTUs (OTU 367, OTU 58 and OTU 110) to be consistently important.
-Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs were shared between the models.
+Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs (OTU 361, 2, 477, 12) were shared between the models.
 
-When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out many of the same OTUs (e.g.
-OTU00822, OTU01239, OTU00609) that were important when we interpreted the model using feature rankings based on weights [Figure 3].
-Similarly, L2-regularized SVM and L2-regularized logistic regression  each identified many OTUs that were selected by both interpretation methods.
+To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models [Figure S3].
+When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out 17 out of 20 top OTUs (e.g.
+OTU 822, OTU 1239, OTU 609) that were deemed important when we interpreted the model using feature rankings based on weights [Figure 3].
+Similarly, L2-regularized SVM and L2-regularized logistic regression each identified many OTUs that were selected by both interpretation methods, 9 out of 20 and 12 out of 20 respectively.
 

### 2019-08-19 submission/manuscript.tex
changes to interp results

@@
 }{}
-\usepackage[margin=1.0in]{geometry}
+\usepackage[margin=1in]{geometry}
 \usepackage{hyperref}
@@
 cross-validation AUROC to the median testing AUROC. If the difference
-between the cross-validation and testing AUROCs is large, then that
-would indicate that the models were overfit. difference in median AUROCs
-was 0.021 in L1-regularized SVM with linear kernel, followed by SVM with
-radial basis function kernel and decision tree with a difference of
-0.007 and 0.006, respectively {[}Figure 2{]}; however, these differences
-are relatively small and would not indicate a problem with overfitting.
+between the cross-validation and testing AUROCs was large, then that
+would indicate that the models were overfit. The difference in median
+AUROCs was 0.021 in L1-regularized SVM with linear kernel, followed by
+SVM with radial basis function kernel and decision tree with a
+difference of 0.007 and 0.006, respectively {[}Figure 2{]}; however,
+these differences are relatively small and would not indicate a problem
+with overfitting.
 
 To evaluate the risk for over-optimism of each model, we calculated the
-range for each model using 100 splits. The range among the testing AUROC
-values within each model varied by 0.23 on average across the seven
-models. If we had only done a single split, then there is a risk that we
-could gotten lucky or unlucky with the performance of the model. For
-instance, the lowest AUROC value of the random forest model was 0.59
-whereas the highest was 0.81. These results showed that depending on the
-data-split, the testing AUROC values showed great variability {[}Figure
-2{]}. Therefore, it is important to employ the hierarchical data splits
-that were included in our pipeline to minimize the risk of
-over-optimism.
+range of AUROC values for each model using 100 splits. The range among
+the testing AUROC values within each model varied by 0.23 on average
+across the seven models. If we had only done a single split, then there
+is a risk that we could gotten lucky or unlucky with the performance of
+the model. For instance, the lowest AUROC value of the random forest
+model was 0.593 whereas the highest was 0.81. These results showed that
+depending on the data-split, the testing AUROC values showed great
+variability {[}Figure 2{]}. Therefore, it is important to employ the
+hierarchical data splits that were included in our pipeline to minimize
+the risk of over-optimism.
+
+To show the effect of sample size on model generalizability, we compared
+cross-validation AUROC values of L2-regularized logistic regression and
+random forest models when we subsetted our sample size from 490 to 245,
+120, 60, 30 and 15 {[}Figure S4{]}. The range among the cross-validation
+AUROC values within both models at lower sample sizes were much larger
+than when full sample size was used to train and validate the models.
+These results showed that because microbiome data has many features
+(6920 OTUs), it is important to train models using appropriate sample
+sizes to avoid generalizability issues.
 
@@
 decrease in interpretability as they increase in complexity. In this
-study we used two methods to interpret models with varying complexity.
-
-We interpreted linear models (L1 and L2-regularized SVM with linear
-kernel and L2-regularized logistic regression) using the absolute
-feature weights of the trained models. We ranked the absolute values of
-the weights of all the OTUs for each data-split {[}Figure 3{]}. We
-calculated the median ranks of these features over the 100 data-splits.
-In the three linear models, the OTUs that had the largest median rank
-and drove the detection of SRNs contained sequences that were most
-similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia
-(OTU00822, OTU00768), Clostridium (OTU00609), Blautia (OTU00659,
-OTU00012), Faecalibacterium (OTU00015), Gamella (OTU00426) and
-Peptostreptococcus (OTU00367) {[}Figure 3{]}. Among the five highest
-ranked OTUs across these three models, it was encouraging that OTU 1239
-was shared among each of the models and that \ldots{} were shared among
-both of the L2-regularized models.
-
-We explained the feature importances in non-linear models; SVM with
-radial basis kernel, decision tree, random forest and XGBoost, using a
-method called permutation importance on the held-out test data.
-Permutation importance analysis is a posthoc explanation of the model
-where we randomly permute non-correlated features individually and
-groups of highly correlated features together. We then calculate how
-much the predictive performance of the model (i.e AUROC values) decrease
-when each OTU or group of OTUs is permuted randomly. We ranked the OTUs
-based on how much they decreased the median testing AUROC; the OTU with
-the largest decrease ranking highest. The top 5 OTUs with the largest
-negative impact on testing AUROC overlapped in tree-based models
-{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
-(OTU00367) abundances randomly, dropped the predictive performances the
-most in all tree-based methods {[}Figure 4{]}. Decision tree, random
-forest and XGBoost models' predictive performance dropped from 0.6
-median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
-respectively {[}Figure 4{]}.
+study we used two methods to help interpret our models.
+
+We interpreted the feature importance in the linear models (L1 and
+L2-regularized SVM with linear kernel and L2-regularized logistic
+regression) using the median rank of absolute feature weights of each
+OTU {[}Figure 3{]}. The OTUs that had the largest median rank and drove
+the detection of SRNs across the linear models contained sequences that
+were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742),
+Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659,
+OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) Peptostreptococcus
+(OTU 367) and others {[}Figure 3{]}. Among the highest ranked OTUs
+across these three models, it was encouraging that many OTUs (e.g.~OTU
+822, OTU 1239, OTU 426, OTU 609, OTU 50) were shared among all the
+models.
+
+We interpreted the feature importance in the non-linear models using
+permutation importance. Whereas the absolute feature weights were
+determined from the trained models, here we measured importance using
+the held-out test data. Permutation importance analysis is a posthoc
+explanation of the model where we randomly permuted non-correlated
+features individually and groups of perfectly correlated features across
+the two groups. We then calculated how much the predictive performance
+of the model (i.e testing AUROC values) decreased when each OTU or group
+of OTUs was randomly permuted. We ranked the OTUs based on how much the
+median testing AUROC decreased when it was permuted; the OTU with the
+largest decrease ranked highest. The OTUs that had the largest impact
+and drove the detection of SRNs across the non-linear models contained
+sequences that were most similar to members of the Lachnospiraceae (OTU
+50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629),
+Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58),
+Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477),
+Parvimonas (OTU 361), Roseburia (OTU 822) and others {[}Figure 4{]}.
+Among the twenty OTUs with the largest impact across these four models,
+there was only one OTU (OTU 822) that was shared among all of the
+models; however, among the tree-based models, we found three OTUs (OTU
+367, OTU 58 and OTU 110) to be consistently important. Similarly, in the
+random forest and XGBoost models, four of the twenty most important OTUs
+(OTU 361, 2, 477, 12) were shared between the models.
 
 To highlight the differences between the two interpretation methods, we
-used permutation importance to interpret linear models as well {[}Figure
-S3{]}. L1-regularized SVM with linear kernel picked out some of the same
-OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings
-based on weights {[}Figure 3{]} and permutation importance {[}Figure
-S3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
-regression picked out some of the same OTUs in both interpretation
-methods, OTU00659 and OTU00012, respectively. However, for all the
-linear models, the rankings of these features were different due to the
-collinearity in microbial communities. Collinearity in a microbial
-dataset is when one OTU is dependant on another OTU. The feature weights
-of correlated OTUs are influenced by one another which makes it
-difficult to interpret models using feature weights. Our interpretation
-of L2-regularized logistic regression based on feature weight rankings,
-showed that an OTU that belongs to \emph{Lachnospiraceae} (OTU00056) was
-not among the most important OTUs in making a prediction {[}Figure
-3C{]}. However, it was picked out as important when the same model was
-interpreted using permutation importance {[}Figure S3C{]}. This is due
-to collinearity in the dataset, where \emph{Lachnospiraceae} (OTU00056)
-is significantly correlated with \emph{Feacalibacterium} (OTU00015)
-which has the highest ranked feature weights {[}Figure 3C{]}. We need to
-investigate the relationships among features to identify the true
-underlying factors when making a prediction.
+used permutation importance to interpret linear models {[}Figure S3{]}.
+When we analyzed the L1-regularized SVM with linear kernel model using
+permutation importance, the approach picked out 17 out of 20 top OTUs
+(e.g.~OTU 822, OTU 1239, OTU 609) that were deemed important when we
+interpreted the model using feature rankings based on weights {[}Figure
+3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
+regression each identified many OTUs that were selected by both
+interpretation methods, 9 out of 20 and 12 out of 20 respectively.
+
+Interestingly, for the all the models, except the poorly performing
+L1-regularized SVM and decision tree, the variation in the impact of
+these important OTUs was minimal. This supports the findings of others
+that a classification requires numerous OTUs, rather the presence of a
+single or a few of the OTUs (32).
 
@@
 
-We compared the training times of the seven ML models. As the complexity
-of a ML model and the number of tuned hyperparameter settings increased
-{[}Figures S1-S2{]}, its training times increased as well {[}Figure
-5{]}. Linear models trained faster than non-linear models. L1 and
-L2-regularized SVM with linear kernel and L2-regularized logistic
-regression had the shortest training times with 0.2 hours, (std ± 0.03),
-0.2 hours, (std ± 0.02), and 0.2 hours, (std ± 0.02), respectively.
-Whereas, a decision tree, SVM with radial basis function kernel, random
-forest and XGBoost had training times of 4.4 hours, (std ± 0.3), 59.6
-hours, (std ± 8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1),
-respectively {[}Figure 5{]}.
+We compared the training times of the seven ML models. As expected, the
+training times increased as the complexity of a ML model and the number
+of tuned hyperparameter settings increased and the linear models trained
+faster than non-linear models {[}Figures S1-S2; Figure 5{]}. When we
+subsetted the size of the training dataset, we observed a {[}linear,
+quadratic, etc{]} relationship between the size of the dataset and the
+training time for the random forest model.
 
 \subsection{Discussion}\label{discussion}
+
+Overview • Establish pipeline that can be generalized to any modeling
+method • Linear/non-linear not a bfd • Interpretability tradeoff Point 1
+• Importance for transparency, picking the most complex model is not
+necessarilty the best model • Good / better / best: Table S1 • ``best''
+has multiple meanings Point 2 • Our paper is concerned with developing
+models to gain greater biological insights into what's going on • Still
+need biological validation and further experimentation Point 3 • These
+models show great opportunity to use microbiome to make diagnoses • We
+talked about classification, but others have used it in microbiome data
+to do regression, mulit-class -- pipeline is the same • Others have used
+this on shotgun metagenomic data and collinearity is expected to be an
+even larger problem since 5000 genes are all correlated with each other
+• Deployment is a different question that requires independent datasets
+to test on; we currently lack decent sized parallel datasets where the
+same type of sequence data and clinical data have been collected • We
+also have a problem with OTU-based approaches Conclusion/Future
+direction • ML is awesome and there's always new methods; our framework
+gives structure to investigators wanting to try these out. We encourage
+reseraches to apply multiple models to their data and report the results
+• Things like neural nets are attractive and hold great promise, but we
+need a ton more data to apply and they are impossible to interpret
 
@@
 severe problem and needs to be addressed to reliably interpret ML models
-(32). In this study we used two different methods to interpret our
+(33). In this study we used two different methods to interpret our
 linear models; ranking each OTU by (1) their absolute weights in the
@@
 deep learning models prone to overfitting. There are studies that
-address overcoming these challenges in biomedical datasets (11, 33, 34),
+address overcoming these challenges in biomedical datasets (11, 34, 35),
 however sudies that estalish frameworks with microbiome data are
@@
 patients recruited by Great Lakes-New England Early Detection Research
-Network study. These data were obtained from Sze et al (35). The stool
+Network study. These data were obtained from Sze et al (32). The stool
 samples were provided by recruited adult participants who were
@@
 
+\hypertarget{ref-sze_leveraging_2018}{}
+32. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+rRNA gene surveys to identify reproducible biomarkers in individuals
+with colorectal tumors. mBio \textbf{9}:e00630--18.
+doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
+
 \hypertarget{ref-dormann_collinearity:_2013}{}
-32. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+33. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 \hypertarget{ref-kocheturov_massive_2019}{}
-33. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+34. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \hypertarget{ref-kim_improved_2018}{}
-34. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+35. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.
-
-\hypertarget{ref-sze_leveraging_2018}{}
-35. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
-rRNA gene surveys to identify reproducible biomarkers in individuals
-with colorectal tumors. mBio \textbf{9}:e00630--18.
-doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
 

### 2019-08-23 submission/manuscript.Rmd
add interpretation data and fix after mmeting 4

--- 2019-08-07 submission/manuscript.Rmd
@@
 The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance.
+To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models.
+Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model.
+In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
 
@@
 
+To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4].
+The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models.
+These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism.
+Furthermore, it was encouraging that even for a small number of samples, the 95% confidence interval included the median AUROC values for the larger subsetted datasets.
+
 __Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017].
@@
 
-We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median absolute feature weights for each OTU in the models [Figure 3].
-The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) and Peptostreptococcus (OTU 367) [Figure 3].
-Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that 4 OTUs were shared among the five highest ranked OTUs for both of the L2-regularized models.
+First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3].
+We also reviewed at the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having SRN.
+Among the highest ranked OTUs across these three models, it was encouraging that many of the same OTUs (e.g.
+OTU 50, 426, 609, 822, 1239) were important across all of the models.
+The benefit of this approach is that the results of the analysis are based on the trained model parameters and provide information regarding the sign and magnitude of the impact of each OTU.
+However, this approach is only possible with linear models.
 
-We interpreted the feature importance in the non-linear models using permutation importance.
-Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups.
+Second, to analyze non-linear models we interpreted the feature importance.
+Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
+Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data.
 We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
-We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest.
-The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) [Figure 4].
-Among the twenty OTUs with the largest impact across these four models, there was only one OTU (OTU 822) that was shared among all of the models; however, among the tree-based models, we found three OTUs (OTU 367, OTU 58 and OTU 110) to be consistently important.
-Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs were shared between the models.
+We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4].
+Among the twenty OTUs with the largest impact for each of these models, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models.
+Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477).
+Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance.
+For example, if OTU 367 were permuted across the diagnoses from the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`.
+In contrast, if we permuted the same OTU from the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`.
+Effectively, the complexity of the communities was more fully represented in the better performing models[@sze_leveraging_2018; @wirbel_meta-analysis_2019].
+At least in this case, it was not possible to distinguish between health and disease using a single OTU.
+While permutation analysis allowed us to gauge the importance of an OTU, the analysis is post-hoc (i.e.
+done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.
 
-When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out many of the same OTUs (e.g.
-OTU00822, OTU01239, OTU00609) that were important when we interpreted the model using feature rankings based on weights [Figure 3].
-Similarly, L2-regularized SVM and L2-regularized logistic regression  each identified many OTUs that were selected by both interpretation methods.
+To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S3].
+When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S3], 17 out of the 20 top OTUs (e.g.
+OTU 609, 822, 1239) were deemed important by both approaches.
+Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two approaches.
+These results indicate that both approaches are consistent in selecting the most important OTUs.
 
-Interestingly, for the all the models, except the poorly performing L1-regularized SVM and decision tree, the variation in the impact of these important OTUs was minimal.
-This supports the findings of others that a classification requires numerous OTUs, rather the presence of a single or a few of the OTUs [@sze_leveraging_2018].
-
-__The computational efficiency of each ML model.__
-
-We compared the training times of the seven ML models.
-As expected, the training times increased as the complexity of a ML model and the number of tuned hyperparameter settings increased and the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
+__The computational efficiency of each ML model.__ We compared the training times of the seven ML models.
+As expected, the training times increased with the complexity of the model and the number of tuned hyperparameter settings.
+Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
 When we subsetted the size of the training dataset, we observed a [linear, quadratic, etc] relationship between the size of the dataset and the training time for the random forest model.
@@
 •	Still need biological validation and further experimentation
+
 Point 3
@@
 
-Microbiome studies use ML models, often with a classification task, to predict a disease but also to learn which microbes are indicators of that disease [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].
-Achieving either of these tasks have far-reaching impact on human health, however ML as a tool in microbiome studies is still at its infancy.
-A framework is needed to develop rigorous ML models, to identify and overcome potential pitfalls.
-Previous studies generated workflows to allow ML to be widely used by the microbiome researchers [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
-This study sets-up standards for developing and evaluating rigorous ML models for microbiome data [Table 1].
+There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations.
+Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases.
+If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification.
+Because application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models [Table 1].
 
-We benchmarked seven ML models with different classification algorithms to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection.
-Our results showed that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we can create ML models that are inherently interpretable and easily trained without losing predictive power.
+To test this framework, we benchmarked seven ML models to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection.
+Our results showed that if the goal of a study to identify microbial biomarkers of disease, we can create ML models that are inherently interpretable and easily trained without losing predictive power.
 In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models.
@@
 While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest was a more complex ML model and could only be explained using methods such as permutation importance.
-On the other hand, L2-regularized logistic regression was easier to interpret by ranking absolute feature weights of the trained model.
+In terms of interpretability, random forest was a more complex ML model and could only be explained using post-hoc methods such as permutation importance.
+On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights.
 
 Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease.
-As domain experts, we know that human-associated microbial communities have complex correlation structures that creates collinearity in the dataset.
-Collinearity is a severe problem and needs to be addressed to reliably interpret ML models [@dormann_collinearity:_2013].
-In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
-We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset.
-To avoid misinterpreting the models, once we identify the most important microbes, we should check for their relationships with other microbes as well.
-These relationships will help us generate new hypotheses about the ecology of the disease.
-These hypotheses needs to be tested with follow-up experiments to identify the true biomarkers of a disease.
+Human-associated microbial communities have complex correlation structures.
+This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013].
+In this study we used two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
+The feature weights of correlated OTUs are influenced by and dependent on one another.
+To avoid misinterpreting the models, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well.
+These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments.
+To avoid misinterpreting models due to correlation structures of the communities when we used permutation importance, we grouped correlated OTUs to determine their impact as a group.
+We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features to identify the true underlying factors when making a classification.
 
-In this study, we also established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.
-First of all, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values.
-When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data.
-In all the models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data.
-Furthermore, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline.
-Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model.
-We showed that there was variability in AUROC values between different random data-splits in each of the models we tested.
-Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits.
-The randomization and resampling of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics.
-Additionally, we performed a full grid search for hyperparameter settings when training our ML models.
-Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model.
-In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
-And finally, we used the AUROC metric in our study to evaluate the predictive performance of the ML models.
-AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.
+In this study, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome.
+We used a held-out test set and we performed the initial 80%-20% random datasplit to create the held-out test set 100 times in our ML pipeline.
+The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.
 
-We used a balanced CRC dataset to develop ML models with a binary classification task.
+We used a CRC dataset to develop ML models with a binary classification task.
 We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes.
-However, the principles highlighted throughout this study [Table 1] apply to all ML modeling tasks with microbiome data.
-The models we built were generalizable despite the high number of features microbiome datasets usually have.
-The generalization performance of ML models depends on sample size.
-The more complex the model, the more data it will need.
-Our dataset had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models.
-Our analysis was limited to shallow learning methods and did not explore deep learning methods such as neural networks.
+However, the principles highlighted throughout this study [Table 2] apply to all ML modeling tasks with microbiome data.
+Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks.
 Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting.
+
 There are studies that address overcoming these challenges in biomedical datasets [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], however sudies that estalish frameworks with microbiome data are lacking.

### 2019-08-23 submission/manuscript.tex
add interpretation data and fix after mmeting 4

@@
 and controls, to 0.50, where the model's predictions are no different
-from random chance.
+from random chance. To select the best performing hyperparameter, we
+performed a full grid search for hyperparameter settings when training
+our models. Default hyperparameter settings in previously developed ML
+packages in R, Python, and Matlab programming languages are inadequate
+for effective application of classification algorithms and need to be
+optimized for each new dataset used to generate a model. In the example
+of L1-regularized SVM with linear kernel {[}Figure S1{]}, the model
+showed large variability between different regularization coefficients
+(C) and was susceptible to performing poorly if the wrong regularization
+coefficient was assigned to the model by default.
 
@@
 cross-validation AUROC values of L2-regularized logistic regression and
-random forest models when we subsetted our sample size from 490 to 245,
-120, 60, 30 and 15 {[}Figure S4{]}. The range among the cross-validation
-AUROC values within both models at lower sample sizes were much larger
-than when full sample size was used to train and validate the models.
-These results showed that because microbiome data has many features
-(6920 OTUs), it is important to train models using appropriate sample
-sizes to avoid generalizability issues.
+random forest models when we subsetted our original study design with
+490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
+range among the cross-validation AUROC values within both models at
+lower sample sizes were much larger than when the full collection of
+samples was used to train and validate the models. These results showed
+that because the microbiome data had many features (6920 OTUs), it was
+important to train the models using appropriate sample sizes to avoid
+problems with generalizability and over-optimism. Furthermore, it was
+encouraging that even for a small number of samples, the 95\% confidence
+interval included the median AUROC values for the larger subsetted
+datasets.
 
@@
 
-We interpreted the feature importance in the linear models (L1 and
-L2-regularized SVM with linear kernel and L2-regularized logistic
-regression) using the median rank of absolute feature weights of each
-OTU {[}Figure 3{]}. The OTUs that had the largest median rank and drove
-the detection of SRNs across the linear models contained sequences that
-were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742),
-Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659,
-OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) Peptostreptococcus
-(OTU 367) and others {[}Figure 3{]}. Among the highest ranked OTUs
-across these three models, it was encouraging that many OTUs (e.g.~OTU
-822, OTU 1239, OTU 426, OTU 609, OTU 50) were shared among all the
-models.
-
-We interpreted the feature importance in the non-linear models using
-permutation importance. Whereas the absolute feature weights were
-determined from the trained models, here we measured importance using
-the held-out test data. Permutation importance analysis is a posthoc
-explanation of the model where we randomly permuted non-correlated
-features individually and groups of perfectly correlated features across
-the two groups. We then calculated how much the predictive performance
-of the model (i.e testing AUROC values) decreased when each OTU or group
-of OTUs was randomly permuted. We ranked the OTUs based on how much the
-median testing AUROC decreased when it was permuted; the OTU with the
-largest decrease ranked highest. The OTUs that had the largest impact
-and drove the detection of SRNs across the non-linear models contained
-sequences that were most similar to members of the Lachnospiraceae (OTU
-50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629),
-Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58),
-Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477),
-Parvimonas (OTU 361), Roseburia (OTU 822) and others {[}Figure 4{]}.
-Among the twenty OTUs with the largest impact across these four models,
-there was only one OTU (OTU 822) that was shared among all of the
-models; however, among the tree-based models, we found three OTUs (OTU
-367, OTU 58 and OTU 110) to be consistently important. Similarly, in the
-random forest and XGBoost models, four of the twenty most important OTUs
-(OTU 361, 2, 477, 12) were shared between the models.
-
-To highlight the differences between the two interpretation methods, we
-used permutation importance to interpret linear models {[}Figure S3{]}.
-When we analyzed the L1-regularized SVM with linear kernel model using
-permutation importance, the approach picked out 17 out of 20 top OTUs
-(e.g.~OTU 822, OTU 1239, OTU 609) that were deemed important when we
-interpreted the model using feature rankings based on weights {[}Figure
-3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
-regression each identified many OTUs that were selected by both
-interpretation methods, 9 out of 20 and 12 out of 20 respectively.
-
-Interestingly, for the all the models, except the poorly performing
-L1-regularized SVM and decision tree, the variation in the impact of
-these important OTUs was minimal. This supports the findings of others
-that a classification requires numerous OTUs, rather the presence of a
-single or a few of the OTUs (32).
-
-\textbf{The computational efficiency of each ML model.}
-
-We compared the training times of the seven ML models. As expected, the
-training times increased as the complexity of a ML model and the number
-of tuned hyperparameter settings increased and the linear models trained
-faster than non-linear models {[}Figures S1-S2; Figure 5{]}. When we
-subsetted the size of the training dataset, we observed a {[}linear,
-quadratic, etc{]} relationship between the size of the dataset and the
-training time for the random forest model.
+First, we interpreted the feature importance of the linear models (L1
+and L2-regularized SVM with linear kernel and L2-regularized logistic
+regression) using the median rank of absolute feature weights for each
+OTU {[}Figure 3{]}. We also reviewed at the signs of feature weights to
+determine whether an OTU is associated with classifying a subject as
+being healthy or having an SRN - negative sign indicated being healthy
+and positive sign indicated having SRN. Among the highest ranked OTUs
+across these three models, it was encouraging that many of the same OTUs
+(e.g.~OTU 50, 426, 609, 822, 1239) were important across all of the
+models. The benefit of this approach is that the results of the analysis
+are based on the trained model parameters and provide information
+regarding the sign and magnitude of the impact of each OTU. However,
+this approach is only possible with linear models.
+
+Second, to analyze non-linear models we interpreted the feature
+importance. Whereas the absolute feature weights were determined from
+the trained models, here we measured importance using the held-out test
+data. Permutation importance analysis is a posthoc explanation of the
+model where we randomly permuted non-correlated features individually
+and groups of perfectly correlated features across the two groups in the
+held-out test data. We then calculated how much the predictive
+performance of the model (i.e testing AUROC values) decreased when each
+OTU or group of OTUs was randomly permuted. We ranked the OTUs based on
+how much the median testing AUROC decreased when it was permuted; the
+OTU with the largest decrease ranked highest {[}Figure 4{]}. Among the
+twenty OTUs with the largest impact for each of these models, there was
+only one OTU (OTU 822) that was shared among all of the models; however,
+we found three OTUs (OTU 58, 110, 367) that were important in each of
+the tree-based models. Similarly, the random forest and XGBoost models,
+shared four of the most important OTUs (OTU 2, 12, 361, 477).
+Permutation analysis results also revealed that with the exception of
+the decision tree model, removal of any individual OTU had a minimal
+impact on model performance. For example, if OTU 367 were permuted
+across the diagnoses from the decision tree model, the median AUROC
+dropped from 0.601 to 0.525. In contrast, if we permuted the same OTU
+from the random forest model, the AUROC only dropped from 0.695 to
+0.6799. Effectively, the complexity of the communities was more fully
+represented in the better performing models(22, 32). At least in this
+case, it was not possible to distinguish between health and disease
+using a single OTU. While permutation analysis allowed us to gauge the
+importance of an OTU, the analysis is post-hoc (i.e.~done using the test
+data) and these results did not allow us to directly interrogate the
+models to know whether an OTU is associated with classifying a subject
+as being healthy or having an SRN.
+
+To further highlight the differences between the two interpretation
+methods, we used permutation importance to interpret the linear models
+{[}Figure S3{]}. When we analyzed the L1-regularized SVM with linear
+kernel model using feature rankings based on weights {[}Figure 3{]} and
+permutation importance {[}Figure S3{]}, 17 out of the 20 top OTUs
+(e.g.~OTU 609, 822, 1239) were deemed important by both approaches.
+Similarly, for the L2-regularized SVM and L2-regularized logistic
+regression, 9 and 12 OTUs, respectively, were shared among the two
+approaches. These results indicate that both approaches are consistent
+in selecting the most important OTUs.
+
+\textbf{The computational efficiency of each ML model.} We compared the
+training times of the seven ML models. As expected, the training times
+increased with the complexity of the model and the number of tuned
+hyperparameter settings. Also, the linear models trained faster than
+non-linear models {[}Figures S1-S2; Figure 5{]}. When we subsetted the
+size of the training dataset, we observed a {[}linear, quadratic, etc{]}
+relationship between the size of the dataset and the training time for
+the random forest model.
 
@@
 models to gain greater biological insights into what's going on • Still
-need biological validation and further experimentation Point 3 • These
-models show great opportunity to use microbiome to make diagnoses • We
-talked about classification, but others have used it in microbiome data
-to do regression, mulit-class -- pipeline is the same • Others have used
-this on shotgun metagenomic data and collinearity is expected to be an
-even larger problem since 5000 genes are all correlated with each other
-• Deployment is a different question that requires independent datasets
-to test on; we currently lack decent sized parallel datasets where the
-same type of sequence data and clinical data have been collected • We
-also have a problem with OTU-based approaches Conclusion/Future
-direction • ML is awesome and there's always new methods; our framework
-gives structure to investigators wanting to try these out. We encourage
-reseraches to apply multiple models to their data and report the results
-• Things like neural nets are attractive and hold great promise, but we
-need a ton more data to apply and they are impossible to interpret
-
-Microbiome studies use ML models, often with a classification task, to
-predict a disease but also to learn which microbes are indicators of
-that disease (2--11). Achieving either of these tasks have far-reaching
-impact on human health, however ML as a tool in microbiome studies is
-still at its infancy. A framework is needed to develop rigorous ML
-models, to identify and overcome potential pitfalls. Previous studies
-generated workflows to allow ML to be widely used by the microbiome
-researchers (20--23). This study sets-up standards for developing and
-evaluating rigorous ML models for microbiome data {[}Table 1{]}.
-
-We benchmarked seven ML models with different classification algorithms
-to show that a clearly defined ML problem that is based on the goal of
-the microbiome study should inform our model selection. Our results
-showed that if the goal of a study is to learn the ecology behind a
-disease and to identify microbial biomarkers, we can create ML models
-that are inherently interpretable and easily trained without losing
-predictive power. In terms of predictive performance, random forest
-model had the best testing AUROC values compared to the other six
-models. However, the second best model was L2-regularized logistic
-regression with a median AUROC difference of only 0.015 compared to
-random forest. While random forest took 83.2 hours to train,
-L2-regularized logistic regression trained in 12 minutes. In terms of
-interpretability, random forest was a more complex ML model and could
-only be explained using methods such as permutation importance. On the
-other hand, L2-regularized logistic regression was easier to interpret
-by ranking absolute feature weights of the trained model.
+need biological validation and further experimentation
+
+Point 3 • These models show great opportunity to use microbiome to make
+diagnoses • We talked about classification, but others have used it in
+microbiome data to do regression, mulit-class -- pipeline is the same •
+Others have used this on shotgun metagenomic data and collinearity is
+expected to be an even larger problem since 5000 genes are all
+correlated with each other • Deployment is a different question that
+requires independent datasets to test on; we currently lack decent sized
+parallel datasets where the same type of sequence data and clinical data
+have been collected • We also have a problem with OTU-based approaches
+Conclusion/Future direction • ML is awesome and there's always new
+methods; our framework gives structure to investigators wanting to try
+these out. We encourage reseraches to apply multiple models to their
+data and report the results • Things like neural nets are attractive and
+hold great promise, but we need a ton more data to apply and they are
+impossible to interpret
+
+There is a growing awareness that many human diseases and environmental
+processes are not driven by single organism, but are the product of
+multiple bacterial populations. Whereas traditional statistical
+approaches are useful for identifying those cases where a single
+organism can drive a process, ML methods offer the ability to use many
+features of these communities to identify those cases. If it is possible
+to classify communities reliably, then it is hoped that we could
+identify those populations that are responsible for the classification.
+Because application of ML in microbiome studies is still in its infancy,
+we developed a framework to train rigorous, transparent and reproducible
+ML models {[}Table 1{]}.
+
+To test this framework, we benchmarked seven ML models to show that a
+clearly defined ML problem that is based on the goal of the microbiome
+study should inform our model selection. Our results showed that if the
+goal of a study is to learn the ecology behind a disease and to identify
+microbial biomarkers, we can create ML models that are inherently
+interpretable and easily trained without losing predictive power. In
+terms of predictive performance, random forest model had the best
+testing AUROC values compared to the other six models. However, the
+second best model was L2-regularized logistic regression with a median
+AUROC difference of only 0.015 compared to random forest. While random
+forest took 83.2 hours to train, L2-regularized logistic regression
+trained in 12 minutes. In terms of interpretability, random forest was a
+more complex ML model and could only be explained using post-hoc methods
+such as permutation importance. On the other hand, L2-regularized
+logistic regression was easier to interpret by ranking absolute feature
+weights of the trained model.
 
@@
 regression, there are potential pitfalls when it comes to identifying
-biomarkers of a disease. As domain experts, we know that
-human-associated microbial communities have complex correlation
-structures that creates collinearity in the dataset. Collinearity is a
-severe problem and needs to be addressed to reliably interpret ML models
-(33). In this study we used two different methods to interpret our
-linear models; ranking each OTU by (1) their absolute weights in the
-trained models and (2) their impact on the predictive performance based
-on permutation importance. We observed differences in the OTU rankings
-between the two interpretation methods due to collinearity in the
-dataset. To avoid misinterpreting the models, once we identify the most
-important microbes, we should check for their relationships with other
-microbes as well. These relationships will help us generate new
-hypotheses about the ecology of the disease. These hypotheses needs to
-be tested with follow-up experiments to identify the true biomarkers of
-a disease.
-
-In this study, we also established a rigorous ML pipeline to use 16S
-rRNA sequence counts to predict a binary health outcome. First of all,
-we used a held-out test set to illustrate the difference between
-cross-validation and testing AUROC values. When the difference between
-cross-validation and test performance is low, this suggest the models
-are not overfit and that they will perform similar with similar data. In
-all the models, the difference between median cross-validation and
-testing AUROC values did not exceed 0.021 which suggests that these
-models are generalizable and can be used to test similar new data.
-Furthermore, we performed the initial 80\%-20\% random datasplit 100
-times in our ML pipeline. Depending on how the data is split, there is
-the chance of being overoptimistic about the predictive performance of a
-model. We showed that there was variability in AUROC values between
-different random data-splits in each of the models we tested. Our
-results showed that the testing AUROC values varied 0.23 on average
-between different data-splits. The randomization and resampling of the
-initial data-split to create a held-out test set is a crucial step in
-the ML pipeline to develop robust ML models and to report reliable
-performance metrics. Additionally, we performed a full grid search for
-hyperparameter settings when training our ML models. Default
-hyperparameter settings in previously developed ML packages in R,
-Python, and Matlab programming languages are inadequate for effective
-application of classification algorithms and need to be optimized for
-each new dataset used to generate a model. In the example of
-L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
-large variability between different regularization coefficients (C) and
-was susceptible to performing poorly if the wrong regularization
-coefficient was assigned to the model by default. And finally, we used
-the AUROC metric in our study to evaluate the predictive performance of
-the ML models. AUROC is always random at the value 0.5 and is a robust
-metric when a dataset is imbalanced.
-
-We used a balanced CRC dataset to develop ML models with a binary
-classification task. We did not evaluate multicategory classification
-methods or regression analyses to predict non-binary outcomes. However,
-the principles highlighted throughout this study {[}Table 1{]} apply to
-all ML modeling tasks with microbiome data. The models we built were
+biomarkers of a disease. Human-associated microbial communities have
+complex correlation structures. This can hinder our ability to reliably
+interpret models (33). In this study we used two different
+interpretation methods; ranking each OTU by (1) their absolute weights
+in the trained models and (2) their impact on the predictive performance
+based on permutation importance. The feature weights of correlated OTUs
+are influenced by and dependent on one another, which makes it harder to
+identify the importance of unique OTUs. To avoid misinterpreting the
+models, once we identify highly ranked OTUs, we should check for their
+relationships with other OTUs as well. These relationships will help us
+generate new hypotheses about the ecology of the disease and test with
+follow-up experiments. On the other hand, when we used permutation
+importance, we grouped correlated OTUs to determine their impact as a
+group. In this study, we grouped OTUs that had a perfect correlation
+with each other however, we can reduce the correlation threshold to
+further investigate the relationships among features to identify the
+true underlying factors when making a classification.
+
+In this study, we established a pipeline that can be generalized to any
+modeling method that uses 16S rRNA sequence counts to predict a binary
+health outcome. We used a held-out test set and we performed the initial
+80\%-20\% random datasplit to create the held-out test set 100 times in
+our ML pipeline. The models we built with this pipeline were
 generalizable despite the high number of features microbiome datasets
-usually have. The generalization performance of ML models depends on
-sample size. The more complex the model, the more data it will need. Our
-dataset had 490 samples, however microbiome studies that have smaller
-sample sizes would benefit from using less complex models. Our analysis
-was limited to shallow learning methods and did not explore deep
-learning methods such as neural networks. Microbiome datasets often
-suffer from having high dimensionality but low sample sizes which makes
-deep learning models prone to overfitting. There are studies that
-address overcoming these challenges in biomedical datasets (11, 34, 35),
-however sudies that estalish frameworks with microbiome data are
-lacking. This would be an interesting direction for future work in
-microbiome studies.
+usually have.
+
+We used a CRC dataset to develop ML models with a binary classification
+task. We did not evaluate multicategory classification methods or
+regression analyses to predict non-binary outcomes. However, the
+principles highlighted throughout this study {[}Table 2{]} apply to all
+ML modeling tasks with microbiome data. Our analysis was also limited to
+shallow learning methods and did not explore deep learning methods such
+as neural networks. Microbiome datasets often suffer from having high
+dimensionality but low sample sizes which makes deep learning models
+prone to overfitting.
+
+There are studies that address overcoming these challenges in biomedical
+datasets (11, 34, 35), however sudies that estalish frameworks with
+microbiome data are lacking. This would be an interesting direction for
+future work in microbiome studies.
 

### 2019-08-26 submission/manuscript.Rmd
fixed figures

@@
 
-Second, we trained seven different models using the training data.
+Second, we trained seven different models using the training data [Table 1].
 We selected models with different classification algorithms and regularization methods.
@@
 
-To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4].
+To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3].
 The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models.
@@
 
-To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S3].
-When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S3], 17 out of the 20 top OTUs (e.g.
+To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4].
+When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 out of the 20 top OTUs (e.g.
 OTU 609, 822, 1239) were deemed important by both approaches.

### 2019-08-26 submission/manuscript.Rmd
add training time subsampling resutls

@@
 Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
-When we subsetted the size of the training dataset, we observed a [linear, quadratic, etc] relationship between the size of the dataset and the training time for the random forest model.
+When we subsetted the size of the training dataset, we observed an exponential relationship between the size of the dataset and the training time for the random forest model [Figure S5].
 

### 2019-08-26 code/learning/Figure_S5.R

@@ -1,0 +1,95 @@
+######################################################################
+#---------- Compare training time with subsampling results -------#
+######################################################################
+
+# Load in needed functions and libraries
+source('code/learning/functions.R')
+# detach("package:randomForest", unload=TRUE) to run
+
+
+# -------------------- Read files ------------------------------------>
+grab_number <- function(file){
+  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
+    str_extract("[^_]*$")
+}
+
+# Read in files as delim that are saved in a list with a pattern
+read_files <- function(filenames){
+  for(file in filenames){
+    number <- grab_number(file)
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>% 
+      mutate(number = number)
+  }
+  return(data)
+}
+# -------------------------------------------------------------------->
+
+# Read in the traintime 100 splits.
+logit_files <- list.files(path= 'data/process/', pattern='combined_traintime_L2_Logistic_Regression_.*', full.names = TRUE)
+rf_files <- list.files(path= 'data/process/', pattern='combined_traintime_Random_Forest_.*', full.names = TRUE)
+
+
+logit_performance <- map_df(logit_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+logit_performance$number <-  as.factor(logit_performance$number)
+logit_performance$number <- factor(logit_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+rf_performance <- map_df(rf_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+rf_performance$number <-  as.factor(rf_performance$number)
+rf_performance$number <- factor(rf_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+
+
+logit_plot <- ggplot(logit_performance, aes(x = number, y = x)) +
+  geom_boxplot(alpha=0.6, fatten = 4, fill="darkred") +
+  geom_smooth(method="lm", aes(color="Exp Model"), formula= (y ~ exp(x)), se=FALSE, linetype = 1) +
+  scale_y_continuous(name = "L2-regularized logistic regression 
+training time (hours)",
+                     breaks = seq(0, 0.2, 0.05),
[…]

### 2019-08-26 code/learning/load_traintime.R

@@ -36,3 +36,4 @@ traintime_df <- bind_rows(min_fixed_result) %>%
 
 # Get the sorted mean traintime for each model small to large
 traintime_index <- order(traintime_df$mean_traintime) 
+

### 2019-08-26 submission/manuscript.Rmd
add changes after 4th meeting

@@
 If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification.
-Because application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models [Table 1].
+Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models where we pointed out potential pitfalls and ways to avoid them [Table 2].
 
@@
 In this study we used two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
-The feature weights of correlated OTUs are influenced by and dependent on one another.
-To avoid misinterpreting the models, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well.
+The feature weights of correlated OTUs are influenced by one another.
+For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight in the regression model.
+To avoid misinterpreting the models and missing important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well.
 These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments.
-To avoid misinterpreting models due to correlation structures of the communities when we used permutation importance, we grouped correlated OTUs to determine their impact as a group.
-We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features to identify the true underlying factors when making a classification.
+When we used permutation importance, we avoided misinterpreting models due to correlation structures of the communities by grouping correlated OTUs to determine their impact as a group.
+We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.
 
-In this study, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome.
-We used a held-out test set and we performed the initial 80%-20% random datasplit to create the held-out test set 100 times in our ML pipeline.
+We established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome.
+We performed an initial 80%-20% random datasplit to create an held-out test set which we used to evaluate model performance.
+We repeated this datasplit 100 times to get a robust reading of predictive performance.
+We tuned our model hyperparameters in a five-fold cross-validation.
 The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.

### 2019-08-29 submission/manuscript.Rmd
mini review

@@
 These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism.
-Furthermore, it was encouraging that even for a small number of samples, the 95% confidence interval included the median AUROC values for the larger subsetted datasets.
+Furthermore, it was encouraging that even for a small number of samples, the interquartile range included the median AUROC values for the larger subsetted datasets.
 
@@
 First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3].
-We also reviewed at the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having SRN.
-Among the highest ranked OTUs across these three models, it was encouraging that many of the same OTUs (e.g.
-OTU 50, 426, 609, 822, 1239) were important across all of the models.
-The benefit of this approach is that the results of the analysis are based on the trained model parameters and provide information regarding the sign and magnitude of the impact of each OTU.
-However, this approach is only possible with linear models.
+We also reviewed the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having an SRN.
+It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g.
+OTU 50, 426, 609, 822, 1239).
+The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU.
+However, this approach was only possible with linear models.
 
-Second, to analyze non-linear models we interpreted the feature importance.
+Second, to analyze non-linear models we interpreted the feature importance using permutation importance.
 Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
@@
 In contrast, if we permuted the same OTU from the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`.
-Effectively, the complexity of the communities was more fully represented in the better performing models[@sze_leveraging_2018; @wirbel_meta-analysis_2019].
+Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019].
 At least in this case, it was not possible to distinguish between health and disease using a single OTU.
-While permutation analysis allowed us to gauge the importance of an OTU, the analysis is post-hoc (i.e.
+While permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e.
 done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.
@@
 Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
-When we subsetted the size of the training dataset, we observed an exponential relationship between the size of the dataset and the training time for the random forest model [Figure S5].
+When we subsetted the size of the training dataset, we observed a linear relationship between the size of the dataset and the training times for L2-regularized logistic regression and random forest models [Figure S5].
 
@@
 •	Interpretability tradeoff
-Point 1
-•	Importance for transparency, picking the most complex model is not necessarilty the best model
-•	Good / better / best:  Table S1
-•	“best” has multiple meanings
-Point 2
-•	Our paper is concerned with developing models to gain greater biological insights into what’s going on
-•	Still need biological validation and further experimentation
 
@@
 •	We also have a problem with OTU-based approaches
+
 Conclusion/Future direction
@@
 If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification.
-Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models where we pointed out potential pitfalls and ways to avoid them [Table 2].
+Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models and we highlighted potential pitfalls and ways to avoid them [Table S1].
 
-To test this framework, we benchmarked seven ML models to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection.
-Our results showed that if the goal of a study to identify microbial biomarkers of disease, we can create ML models that are inherently interpretable and easily trained without losing predictive power.
+The foundation of this framework was to clearly define the ML task and to choose the best model for that task [Table S1].
+We benchmarked seven ML models and showed that if the ML task is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power.
 In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models.
@@
 While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest was a more complex ML model and could only be explained using post-hoc methods such as permutation importance.
+In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance.
 On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights.
+Comparing many different models showed us that the most complex model is not necessarilty the best model for our ML task.
 
-Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease.
-Human-associated microbial communities have complex correlation structures.
-This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013].
-In this study we used two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
-The feature weights of correlated OTUs are influenced by one another.
-For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight in the regression model.
-To avoid misinterpreting the models and missing important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well.
-These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments.
-When we used permutation importance, we avoided misinterpreting models due to correlation structures of the communities by grouping correlated OTUs to determine their impact as a group.
-We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.
-
-We established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome.
+As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome [Table S1].
 We performed an initial 80%-20% random datasplit to create an held-out test set which we used to evaluate model performance.
@@
 
-We used a CRC dataset to develop ML models with a binary classification task.
+We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases.
+In this study we showcased two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
+Human-associated microbial communities have complex correlation structures.
+This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013].
+The feature weights of correlated OTUs are influenced by one another.
+For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight thus hiding the importance of its correlated OTU.
+To avoid misinterpreting the models and to capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well.
+These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments.
+When we used permutation importance, we avoided misinterpreting models by grouping correlated OTUs to determine their impact as a group.
+We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.
+To identify the true underlying microbial factors of a disease, it is crucial to do further experimentation for biological validation.
+
 We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes.
-However, the principles highlighted throughout this study [Table 2] apply to all ML modeling tasks with microbiome data.
+However, the principles highlighted throughout this study [Table S1] apply to all ML modeling tasks with microbiome data.
 Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks.

### 2019-09-02 submission/manuscript.Rmd
edit the whole thing

@@
 This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
-With this dataset we developed a framework that implementf a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models.
+With this dataset we developed a framework that implements a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models.
 This framework can be easily applied to other host-associated and environmental microbiome datasets.
@@
 The foundation of this framework was to clearly define the ML task and to choose the best model for that task [Table S1].
-We benchmarked seven ML models and showed that if the ML task is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power.
+We benchmarked seven ML models and showed that if the goal is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power.
 In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models.

### 2019-09-03 submission/manuscript.Rmd
changese after R5 with Pat

@@
 
-__Model selection and pipeline construction__ We established a ML pipeline where we trained and validated each of the seven models using a common approach that is based on standard methods within the ML community (REFS)[Figure 1].
+__Model selection and pipeline construction__ We established a ML pipeline where we trained and validated each of the seven models using a common approach that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].
 
@@
 
-Overview
-•	Establish pipeline that can be generalized to any modeling method
-•	Linear/non-linear not a bfd
-•	Interpretability tradeoff
+There is a growing awareness that many human diseases and environmental processes are not driven by a single organism but are the product of multiple bacterial populations.
+Traditional statistical approaches are useful for identifying those cases where a single organism is associated with a process.
+In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole to classify them into different categories such as coming from a patient who is healthy or has SRNs.
+If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification.
+However, the application of ML in microbiome studies is still in its infancy and the field still needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.
 
-Point 3
-•	These models show great opportunity to use microbiome to make diagnoses
-•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
-•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
-•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
-•	We also have a problem with OTU-based approaches
-
-Conclusion/Future direction
-•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out.
-We encourage reseraches to apply multiple models to their data and report the results
-•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret
-
-There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations.
-Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases.
-If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification.
-Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models and we highlighted potential pitfalls and ways to avoid them [Table S1].
-
-The foundation of this framework was to clearly define the ML task and to choose the best model for that task [Table S1].
-We benchmarked seven ML models and showed that if the goal is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power.
-In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models.
-However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
+To address these needs, we developed a framework to train rigorous, transparent, and reproducible models.
+We benchmarked seven ML models and showed that we can create models that are inherently interpretable and easily trained without losing predictive performance.
+In terms of predictive performance, the random forest model had the best testing AUROC values compared to the other six models.
+However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
 While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes.
@@
 On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights.
-Comparing many different models showed us that the most complex model is not necessarilty the best model for our ML task.
+Comparing many different models showed us that the most complex model is not necessarily the best model for our ML task.
 
-As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome [Table S1].
-We performed an initial 80%-20% random datasplit to create an held-out test set which we used to evaluate model performance.
-We repeated this datasplit 100 times to get a robust reading of predictive performance.
-We tuned our model hyperparameters in a five-fold cross-validation.
-The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.
+As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome.
+We performed a random datasplit to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance.
+We repeated this datasplit 100 times to measure predictive performance.
+During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation.
+Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the similar AUROC values from the cross-validation and testing.
 
 We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases.
-In this study we showcased two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance.
-Human-associated microbial communities have complex correlation structures.
-This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013].
-The feature weights of correlated OTUs are influenced by one another.
-For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight thus hiding the importance of its correlated OTU.
-To avoid misinterpreting the models and to capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well.
+In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance.
+Human-associated microbial communities have complex correlation structures which create collinearity in the datasets we work with.
+This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013].
+For example if one OTU is highly correlated with another OTU, only one of these OTUs would have a large feature weight thus hiding the importance of its correlated OTU.
+To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs.
 These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments.
-When we used permutation importance, we avoided misinterpreting models by grouping correlated OTUs to determine their impact as a group.
+When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group.
 We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.
-To identify the true underlying microbial factors of a disease, it is crucial to do further experimentation for biological validation.
+It is important to know the correlation strutures of the data to avoid missinterpreting the models.
+This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronunced due to many genes being correlated with one another because they come from the same chromosome.
+Therefore, to identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.
 
-We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes.
-However, the principles highlighted throughout this study [Table S1] apply to all ML modeling tasks with microbiome data.
+In this study, we did not consider all possible modeling approaches.
+However, the principles highlighted throughout this study apply to all ML modeling tasks with microbiome data.
+For example, we did not evaluate multicategory classification methods to predict non-binary outcomes.
+For example, we could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories).
+We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs.
+Furthermore, as number of categories to classify increase, more samples are required for each category to train a model.
+We also did not use regression-based analyses to predict a non-categorical outcome.
+We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019].
 Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks.
-Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting.
+Deep learning methods hold great promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which makes the deep learning models prone to overfitting.
+These methods are even more complex than random forest and XGBoost and are considered uninterpretable.
+There is great potential for applying ML approaches to microbiome data.
 
-There are studies that address overcoming these challenges in biomedical datasets [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], however sudies that estalish frameworks with microbiome data are lacking.
-This would be an interesting direction for future work in microbiome studies.
-
-Paragraph about deploying a model.
+Our framework gives structure to investigators wanting to train, evaluate, and interpret their own ML models to identify OTUs that might be biologically relevant.
+However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly harder and distinct undertaking.
+For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data.
+We are also challenged by the practical constraints of OTU-based approaches.
+The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added.
+Finally, we also need independent validation cohorts to test the performance of a diagnostic model.
+To realize the potential for using ML approaches with microbiome data, it is necessary we direct our efforts to overcome these challenges.
 
 This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data.
-Model selection should be done with a solid understanding of model complexity and interpretability, rigorous ML pipelines should be built with cross-validation for hyperparameter tuning and with a held-out test set for evaluating predictive performance and models should be interpreted while considering collinearity in datasets.
-The right methods will help us achieve the level of validity and accountability we want from models built for patient health.
+We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table 2].
+We have highlighted the tradeoffs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
+Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.
 

### 2019-09-10 submission/manuscript.Rmd
changes after Pat meeting 6

@@
 ---
-title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
+title: '**Effective application of machine learning to bacterial 16S rRNA gene sequencing data**'
 csl: mbio.csl
@@
 
-Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^
+Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack Ruffin${^3}$, Jenna Wiens${^2}$, Patrick D. Schloss^1$\dagger$^
 
@@
 
-Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed models into question.
-Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers.
-To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases).
+Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes, and colorectal cancer.
+Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but inconsistent adoption of methods call the validity of these models into question.
+Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability.
+To overcome these challenges, we trained seven models that used fecal 16S rRNA sequences to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
+We developed a generalizable pipeline to train, validate and interpret the models.
 To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
-The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
-Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min).
-In this study, we established standards for the development of modeling pipelines for microbiome-associated ML models.
-Additionally, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.
+The random forest model was best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
+Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, it trained much faster (12 min) and was more interpretable.
+Our analysis showed that ML models should be chosen based on the goal of the study as it will inform our expectations of predictive performance and interpretability.
 
 \newpage
-## Importance (needs work)
+## Importance
 
-Prediction of health outcomes using ML is rapidly being adopted by human microbiome studies.
-However, the developed ML models so far are overoptimistic in terms of validity and predictive performance.
-Without rigorous ML pipelines, we cannot trust ML models.
-Before we can speed up progress, we need to slow down, define and implement good ML practices.
+Prediction of health outcomes using machine learning (ML) is rapidly being adopted in human microbiome studies.
+However, these ML models are likely overoptimistic in terms of quantifying predictive performance.
+Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease.
+This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study.
+These concepts are not particular to study of health outcomes but can also be applied to environmental microbiology studies as well.
+
 \newpage
@@
 Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
-The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods.
+The task of diagnosing an individual with high confidence relies on a model that is built with rigorous methods.
 However, there are common methodological problems across many of these studies that need to be addressed as the field progresses.
@@
 
-Among microbiome researchers the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex.
+Among microbiome researchers, the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex.
 This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
@@
 Depending on the application of the model, researchers may choose to use different modeling approaches.
-For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize performance.
+For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize predictive performance.
 Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018].
@@
 We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
-We compared the predictive performance, interpretability, and computational efficiency.
+We compared their predictive performance, interpretability, and computational efficiency.
 To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
@@
 For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features.
-For comparison, we also trained an L1 regularized SVM model with linear kernel.
+For comparison, we also trained an L1-regularized SVM model with linear kernel.
 L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero).
@@
 To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models.
-Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model.
+Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
 In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
@@
 The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively.
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively.
 L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
-Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularlized logistic regression and linear SVM) performed nearly as well as random forest.
+Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well as random forest.
 
@@
 
-To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 splits.
+To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 data-splits.
 The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` on average across the seven models.
@@
 We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4].
-Among the twenty OTUs with the largest impact for each of these models, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models.
+Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models.
 Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477).
 Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance.
-For example, if OTU 367 were permuted across the diagnoses from the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`.
-In contrast, if we permuted the same OTU from the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`.
+For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`.
+In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`.
 Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019].
@@
 When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 out of the 20 top OTUs (e.g.
-OTU 609, 822, 1239) were deemed important by both approaches.
-Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two approaches.
-These results indicate that both approaches are consistent in selecting the most important OTUs.
+OTU 609, 822, 1239) were deemed important by both interpretation methods.
+Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods.
+These results indicate that both methods are consistent in selecting the most important OTUs.
 
@@
 This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronunced due to many genes being correlated with one another because they come from the same chromosome.
-Therefore, to identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.
+To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.
 
@@
 These methods are even more complex than random forest and XGBoost and are considered uninterpretable.
-There is great potential for applying ML approaches to microbiome data.
+There is great potential for applying ML approaches to microbiome data and we can choose which ML approach is best for the study.
 
@@
 Finally, we also need independent validation cohorts to test the performance of a diagnostic model.
-To realize the potential for using ML approaches with microbiome data, it is necessary we direct our efforts to overcome these challenges.
+To realize the potential for using ML approaches with microbiome data, it is necessary that we direct our efforts to overcome these challenges.
 
 This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data.
-We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table 2].
+We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1].
 We have highlighted the tradeoffs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
@@
 
-**Data collection and study population.**
-  The data used for this analysis are stool bacterial abundances and clinical information of the patients recruited by Great Lakes-New England Early Detection Research Network study.
-These data were obtained from Sze et al [@sze_leveraging_2018].
-  The stool samples were provided by recruited adult participants who were undergoing scheduled screening or surveillance colonoscopy.
-Colonoscopies were performed and fecal samples were collected from participants in four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
-Patients' colonic health was labeled by colonoscopy with adequate preparation and tissue histopathology of all resected lesions.
-Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as advanced adenoma.
-Study had `r paste(sum(meta$dx=="normal"))` patients with normal colonoscopies, `r paste(sum(meta$dx=="adenoma"))` with adenomas and `r paste(sum(meta$dx=="cancer"))` with carcinomas.
-Of the `r paste(sum(meta$dx=="adenoma"))` adenomas, `r paste(sum(meta$Dx_Bin=="adv Adenoma"))` were identified as advanced adenomas.
-Stool provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances.
-The bacterial abundance data was generated by Sze et al, by processing 16S rRNA sequences in Mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016].
+**Data collection and study population.** The original stool samples described in our analysis were obtained from patients recruited by Great Lakes-New England Early Detection Research Network (Reference from Mack?).
+Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy.
+Participants were recruited from four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
+Patients’ colonic health was visually assessed by colonoscopy with bowel preparation and tissue histopathology of all resected lesions.
+We assigned patients into two classes: those with normal colons and those with screen relevant neoplasias (SRNs).
+The normal class included patients with normal colons or non-advanced adenomas whereas the SRN class included patients with advanced adenomas or carcinomas (reference for SRN?).
+Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas.
+The study had 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
+Of the 198 adenomas, 109 were identified as advanced adenomas.
+The study collected samples from 261 patients classified as normal and 229 patients classified as having a SRN.
 
-**Data definitions and pre-processing.**
+**16S rRNA gene sequencing data.** Stool samples provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances.
+The sequence data used in our analyses were originally generated by Baxter et al.
+(available through NCBI Sequence Read Archive [SRP062005], 2015).
+The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm ([@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; GitHub link to table).
+These OTU abundances were the features we used to predict colorectal health of the patients.
+There were 6920 OTUs.
+OTU abundances were subsampled to the size of the smallest sample and normalized across samples such that the highest abundance of each OTU would be 1 and lowest would be 0.
 
-  The colorectal health of the patient was defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs).
-Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas.
-The study had 261 normal and 229 SRN samples.
-The bacterial abundances are the features used to predict colorectal health of the patients.
-For each patient, we had `r paste(ncol(data)-1)` features (fecal bacterial abundances) and a two-class label that defines their colorectal health (normal or SRN colorectal lesions as defined by colonoscopies).
-We established modeling pipelines for a binary prediction task Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts.
-OTU counts were set to the size of our smallest sample and were subsampled at the same distances.
-They were then transformed by scaling to a [0-1] range.
+**Model training and evaluation.** Models were trained using the caret package (v.6.0.81) in R (v.3.5.0).
+We modified the caret code to calculate decision values instead of predicted probabilities for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel.
+These changes were necessary to calculate AUROC values for SVMs.
+The code for these changes are available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.XX_2019/.
 
-**Model training and evaluation.**
-
-Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).
-Within the caret package, we have made modifications to L2-regularized SVM with linear kernel function __svmLinear3__ and developed a L1-regularized SVM with linear kernel function __svmLinear4__ to calculate decision values instead of predicted probabilities.
-These changes are available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.
-
-For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
-For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
-For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has.
-For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
-For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.For hyperparameter selection, we started with a granular grid search.
+For hyperparameter selection, we started with a granular grid search.
 Then we narrowed and fine-tuned the range of each hyperparameter.
-The range of the grid depends on the ML task and ML model.
-A full grid search needs to be performed to avoid variability in testing performance.
-We can use hyper-band to help us with our hyperparameter selection [@li_hyperband:_2016].
+The range of the grid changes with different ML tasks and ML models.
+A full grid search was needed to avoid large variation in prediction performance.
+For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the _cost_ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
+For SVM with radial basis function kernel we also tuned _sigma_ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
+For the decision tree model, we tuned the _depth of the tree_ where deeper the tree, the more splits it has.
+For random forest, we tuned the _number of features_ to consider when looking for the best tree split.
+For XGBoost, we tuned for _learning rate_ and the _fraction of samples_ to be used for fitting the individual base learners.
+There are newly developed tools such as Hyperband that help researcher with hyperparameter selection [@li_hyperband:_2016] that can be incorporated to microbiome studies.
 
 The computational burden during model training due to model complexity was reduced by parallelizing segments of the ML pipeline.
-In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
-We can further parallelize the cross-validation step for each hyperparameter setting.
+We parallelized the training of each data-split.
+This allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
+It is possible to further parallelize the cross-validation step for each hyperparameter setting if limited by computational resources.
 
 **Permutation importance workflow.**
-We created a Spearman's rank-order correlation matrix, corrected for multiple pairwise comparisons.
-We then defined correlated OTUs as having perfect correlation (correlation coef=1 and p<0.01).
+We created a Spearman's rank-order correlation matrix and defined correlated OTUs as having perfect correlation (correlation coefficient = 1 and p < 0.01).
 Non-correlated OTUs were permuted individually whereas correlated ones were grouped together and permuted at the same time.
@@
 **Figure 1.
-Machine learning pipeline showing predictive model training and evaluation flowchart.
-**  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data.
-The model was evaluated on a held-out set of data (not used in selecting the model).
-Abbreviations: cvAUROC, cross-validation area under the receiver operating characteristic curve
+Machine learning pipeline.
+**   We split the data to create a training (80%) and held-out test set (20%).
+The splits were stratified to maintain the overall label distribution.
+We performed five-fold cross-validation on the training data to select the best hyperparameter setting and then used these hyperparameters to train the models.
+The model was evaluated on the held-out data set.
+Abbreviations: cvAUC, cross-validation area under the receiver operating characteristic curve.
 \newpage
@@
 Predictive performance of random forest model was higher than other ML models.
-The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
-The whiskers show the farthest points that are not outliers.
-Outliers are data points that are not within 3/2 times the interquartile ranges.
-Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
+The boxplot shows quartiles at the box ends and the median as the horizontal line in the box.
+The whiskers show the farthest points that were not outliers.
+Outliers were defined as those data points that are not within 1.5 times the interquartile ranges.
 \newpage
@@
 **Figure 3.
-Interpretation of the linear ML models.** The absolute feature weights of (A) L2 logistic regression coefficients (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel were ranked from highest rank 1 to 100 for each data-split.
-The feature ranks of the highest ranked five OTUs based on their median ranks are shown here.
-Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
+Interpretation of the linear ML models.** The absolute feature weights of (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel were ranked from highest rank, 1, to lowest rank, 100, for each data-split.
+The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.
+OTUs that are associated with classifying a subject as being healthy or having an SRN are in blue and red, respectively.
+Some of the same OTUs were identified as important in all of the linear models.
 \newpage
@@
 **Figure 4.
-Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGBoost feature importances were explained using permutation importance using held-out test set.
-The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
-The colors of the box plots stand for the unique OTUs that are shared among the different models; pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for OTU00882.
-For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel, (B) decision tree, (C) random forest, and (D) XGBoost feature importances were explained using permutation importance on the held-out test data set.
+The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation.
+The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, salmon were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only.
+For all of the tree-based models, a Peptostreptococcus species (OTU00367) had the largest impact on predictive performance.
 \newpage
@@
 
-**Figure 5.
-Computational efficiency of seven ML models.** The training times for of each data-split showed the differences in computational efficiency of the seven models.
-The median training time in hours was the highest for XGBoost and shortest for L1-regularized SVM with linear kernel.
-The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box.
-The whiskers show the farthest points that are not outliers.
-Outliers are data points that are not within 3/2 times the interquartile ranges.
-Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.
+**Figure 5. Computational efficiency of seven ML models.** The median training time was the highest for XGBoost and shortest for L2-regularized logistic regression. 
 \newpage
@@
 **Figure S1.
-Hyperparameter setting performances for linear models.** (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters are used in training the model.
+Hyperparameter setting performances for linear models.** (A) L2 logistic regression, (B) L1 SVM with linear kernel, and (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model.
 The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
@@
 **Figure S2.
-Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGBoost mean cross-validation AUROC values when different hyperparameters are used in training the model.
+Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) Random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.
 The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
@@
 **Figure S3.
-Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel (B) L2-regularized SVM with linear kernel and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
-The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+Classification performance of ML models across cross validation when dataset is subsetted .** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
+The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
+
+\newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
+
+**Figure S4.
+Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
+
+\newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
+
+**Figure S5.
+Training times of ML models when dataset is subsetted .** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
+As the size of the dataset increased, the training times for L2-regularized logistic regression and random forest models increased linearly.
 
@@
 
-\captionof{table}{An aspirational rubric for evaluating the rigor of ML practices.}
+\captionof{table}{Table S1. An aspirational rubric for evaluating the rigor of ML practices.}
 \small
@@
 } & \makecell[l]{Do we know our data?
-\\ Do we know the confounding \\ variables?
+\\ Do we know the correlated \\ variables?
 } \\ \hline
@@
 
-\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Do we have an held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\ held-out datasets?} & \makecell[l]{Have we tuned our model \\ hyperparameters in \\ cross-validation?} \\ \hline
+\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Did we do \\ cross-validation} & \makecell[l]{Do we have an held-out \\ test dataset?
+\\ Have we tuned our model \\ hyperparameters in \\ cross-validation?} & \makecell[l]{Have we tested our model \\ on many different \\  datasets?} \\ \hline
 
@@
 \makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it?
-\\ Have we checked for \\ the effect of confounding \\ variables?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
+\\ Have we checked for \\ the effect of correlated \\ variables?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
 } \\ \hline

### 2019-09-10 submission/manuscript.tex
changes after Pat meeting 6

@@
 \hypersetup{unicode=true,
-            pdftitle={Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data},
+            pdftitle={Effective model selection and implementation when applying machine learning to bacterial 16S rRNA gene sequencing data},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Best practices for applying machine learning to bacterial 16S
-rRNA gene sequencing data}}
+  \title{\textbf{Effective model selection and implementation when applying
+machine learning to bacterial 16S rRNA gene sequencing data}}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Jenna
-Wiens\({^2}\), Mack Ruffin\({^3}\), Patrick D.
+Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Mack
+Ruffin\({^3}\), Jenna Wiens\({^2}\), Patrick D.
 Schloss\textsuperscript{1\(\dagger\)}
@@
 to identify the microbial biomarkers and aid in diagnosis of many
-chronic diseases such as inflammatory bowel disease, diabetes and
+chronic diseases such as inflammatory bowel disease, diabetes, and
 colorectal cancer. Progress has been made towards developing ML models
-that predict health outcomes from bacterial abundances, but rigourous ML
-models are scarce due to the flawed methods that call the validity of
-developed models into question. Furthermore, the use of black box ML
-models has hindered the validation of microbial biomarkers. To overcome
-these challenges, we benchmarked seven different ML models that use
-fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions
-(n=490 patients, 261 controls and 229 cases). To show the effect of
-model selection, we assessed the predictive performance,
-interpretability, and computational efficiency of the following models:
-L2-regularized logistic regression, L1 and L2-regularized support vector
-machines (SVM) with linear and radial basis function kernels, a decision
-tree, random forest, and extreme gradient boosting (XGBoost). The random
-forest model was best at detecting CRC lesions with an AUROC of 0.695
-but it was slow to train (83.2 h) and hard to interpret. Despite its
-simplicity, L2-regularized logistic regression followed random forest in
-predictive performance with an AUROC of 0.680, and it trained much
-faster (12 min). In this study, we established standards for the
-development of modeling pipelines for microbiome-associated ML models.
-Additionally, we showed that ML models should be chosen based on
-expectations of predictive performance, interpretability and available
-computational resources.
+that predict health outcomes from bacterial abundances, but inconsistent
+adoption of methods call the validity of these models into question.
+Furthermore, there appears to be a preference by many researchers to
+favor increased model complexity over interpretability. To overcome
+these challenges, we trained seven models that used fecal 16S rRNA
+sequences to predict the presence of colonic screen relevant neoplasias
+(SRNs; n=490 patients, 261 controls and 229 cases). We developed a
+generalizable pipeline to train, validate and interpret the models. To
+show the effect of model selection, we assessed the predictive
+performance, interpretability, and computational efficiency of the
+following models: L2-regularized logistic regression, L1 and
+L2-regularized support vector machines (SVM) with linear and radial
+basis function kernels, a decision tree, random forest, and extreme
+gradient boosting (XGBoost). The random forest model was best at
+detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h)
+and hard to interpret. Despite its simplicity, L2-regularized logistic
+regression followed random forest in predictive performance with an
+AUROC of 0.680, it trained much faster (12 min) and was more
+interpretable. Our analysis showed that ML models should be chosen based
+on the goal of the study as it will inform our expectations of
+predictive performance and interpretability.
 
@@
 
-\subsection{Importance (needs work)}\label{importance-needs-work}
-
-Prediction of health outcomes using ML is rapidly being adopted by human
-microbiome studies. However, the developed ML models so far are
-overoptimistic in terms of validity and predictive performance. Without
-rigorous ML pipelines, we cannot trust ML models. Before we can speed up
-progress, we need to slow down, define and implement good ML practices.
+\subsection{Importance}\label{importance}
+
+Prediction of health outcomes using machine learning (ML) is rapidly
+being adopted in human microbiome studies. However, these ML models are
+likely overoptimistic in terms of quantifying predictive performance.
+Moreover, there is a trend towards using black box models such as random
+forest and neural networks without a discussion of the difficulty of
+interpreting such models when trying to identify microbial biomarkers of
+disease. This work represents a step towards developing better ML
+practices in microbiome research by implementing a rigorous pipeline and
+emphasizing the importance of selecting ML models that reflect the goal
+of the study. These concepts are not particular to study of health
+outcomes but can also be applied to environmental microbiology studies
+as well.
+
 \newpage
@@
 16--18). The task of diagnosing an individual with high confidence
-relies on a ML model that is built with rigorous methods. However, there
+relies on a model that is built with rigorous methods. However, there
 are common methodological problems across many of these studies that
@@
 
-Among microbiome researchers the lack of transparency in justifying a
+Among microbiome researchers, the lack of transparency in justifying a
 modelling approach has been due to an implicit assumption that more
@@
 identify the populations causing a disease would likely want a more
-interpretable model whereas clinicians may emphasize performance.
-Although one may feel that they are sacrificing interpretability for
-performance, that tradeoff may be minimal (28, 29). Regardless, it is
-important for researchers to articulate why they have selected a
-specific modelling approach or even compare multiple approaches in the
-same study.
+interpretable model whereas clinicians may emphasize predictive
+performance. Although one may feel that they are sacrificing
+interpretability for performance, that tradeoff may be minimal (28, 29).
+Regardless, it is important for researchers to articulate why they have
+selected a specific modelling approach or even compare multiple
+approaches in the same study.
 
@@
 also built four non-linear models: SVM with radial basis function
-kernel, a decision tree, random forest and XGBoost. We compared the
+kernel, a decision tree, random forest and XGBoost. We compared their
 predictive performance, interpretability, and computational efficiency.
@@
 microbiome (1, 3, 4, 30). With this dataset we developed a framework
-that implementf a ML pipeline that can be used for any modeling
+that implements a ML pipeline that can be used for any modeling
 approach, evaluates predictive performance, and demonstrates how to
@@
 common approach that is based on standard methods within the ML
-community (REFS){[}Figure 1{]}.
+community (REFS by Jenna?){[}Figure 1{]}.
 
@@
 
-Second, we trained seven different models using the training data. We
-selected models with different classification algorithms and
-regularization methods. Regularization is a technique that discourages
-overfitting by penalizing the model for learning the training data too
-well. For regularized logistic regression and SVM with linear kernel, we
-used L2 regularization to keep all potentially important features. For
-comparison, we also trained an L1 regularized SVM model with linear
-kernel. L1-regularization on microbiome data led to a sparser solution
-(i.e., force many coefficients to zero). To explore the potential for
-non-linear relationships among features to improve classification, we
-trained tree-based models including decision tree, random forest, and
-XGBoost and we trained an SVM model with non-linear kernel.
+Second, we trained seven different models using the training data
+{[}Table 1{]}. We selected models with different classification
+algorithms and regularization methods. Regularization is a technique
+that discourages overfitting by penalizing the model for learning the
+training data too well. For regularized logistic regression and SVM with
+linear kernel, we used L2 regularization to keep all potentially
+important features. For comparison, we also trained an L1-regularized
+SVM model with linear kernel. L1-regularization on microbiome data led
+to a sparser solution (i.e., force many coefficients to zero). To
+explore the potential for non-linear relationships among features to
+improve classification, we trained tree-based models including decision
+tree, random forest, and XGBoost and we trained an SVM model with
+non-linear kernel.
 
@@
 for effective application of classification algorithms and need to be
-optimized for each new dataset used to generate a model. In the example
-of L1-regularized SVM with linear kernel {[}Figure S1{]}, the model
-showed large variability between different regularization coefficients
-(C) and was susceptible to performing poorly if the wrong regularization
-coefficient was assigned to the model by default.
+optimized for each new ML task. In the example of L1-regularized SVM
+with linear kernel {[}Figure S1{]}, the model showed large variability
+between different regularization coefficients (C) and was susceptible to
+performing poorly if the wrong regularization coefficient was assigned
+to the model by default.
 
@@
 0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
-0.056) respectively. L1-regularized SVM with linear kernel and decision
+0.056), respectively. L1-regularized SVM with linear kernel and decision
 tree had significantly lower AUROC values than the other ML models with
@@
 complex model (XGBoost) did not have the best performance and that the
-most interpretable models (L2-regularlized logistic regression and
-linear SVM) performed nearly as well as random forest.
+most interpretable models (L2-regularized logistic regression and SVM
+with linear kernel) performed nearly as well as random forest.
 
@@
 To evaluate the risk for over-optimism of each model, we calculated the
-range of AUROC values for each model using 100 splits. The range among
-the testing AUROC values within each model varied by 0.23 on average
-across the seven models. If we had only done a single split, then there
-is a risk that we could gotten lucky or unlucky with the performance of
-the model. For instance, the lowest AUROC value of the random forest
-model was 0.593 whereas the highest was 0.81. These results showed that
-depending on the data-split, the testing AUROC values showed great
-variability {[}Figure 2{]}. Therefore, it is important to employ the
-hierarchical data splits that were included in our pipeline to minimize
-the risk of over-optimism.
+range of AUROC values for each model using 100 data-splits. The range
+among the testing AUROC values within each model varied by 0.23 on
+average across the seven models. If we had only done a single split,
+then there is a risk that we could gotten lucky or unlucky with the
+performance of the model. For instance, the lowest AUROC value of the
+random forest model was 0.593 whereas the highest was 0.81. These
+results showed that depending on the data-split, the testing AUROC
+values showed great variability {[}Figure 2{]}. Therefore, it is
+important to employ the hierarchical data splits that were included in
+our pipeline to minimize the risk of over-optimism.
 
@@
 random forest models when we subsetted our original study design with
-490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
+490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S3{]}. The
 range among the cross-validation AUROC values within both models at
@@
 problems with generalizability and over-optimism. Furthermore, it was
-encouraging that even for a small number of samples, the 95\% confidence
-interval included the median AUROC values for the larger subsetted
+encouraging that even for a small number of samples, the interquartile
+range included the median AUROC values for the larger subsetted
 datasets.
@@
 regression) using the median rank of absolute feature weights for each
-OTU {[}Figure 3{]}. We also reviewed at the signs of feature weights to
+OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
 determine whether an OTU is associated with classifying a subject as
 being healthy or having an SRN - negative sign indicated being healthy
-and positive sign indicated having SRN. Among the highest ranked OTUs
-across these three models, it was encouraging that many of the same OTUs
-(e.g.~OTU 50, 426, 609, 822, 1239) were important across all of the
-models. The benefit of this approach is that the results of the analysis
-are based on the trained model parameters and provide information
-regarding the sign and magnitude of the impact of each OTU. However,
-this approach is only possible with linear models.
+and positive sign indicated having an SRN. It was encouraging that many
+of the highest ranked OTUs were shared across these three models,
+(e.g.~OTU 50, 426, 609, 822, 1239). The benefit of this approach was
+that the results of the analysis were based on the trained model
+parameters and provided information regarding the sign and magnitude of
+the impact of each OTU. However, this approach was only possible with
+linear models.
 
 Second, to analyze non-linear models we interpreted the feature
-importance. Whereas the absolute feature weights were determined from
-the trained models, here we measured importance using the held-out test
-data. Permutation importance analysis is a posthoc explanation of the
-model where we randomly permuted non-correlated features individually
-and groups of perfectly correlated features across the two groups in the
-held-out test data. We then calculated how much the predictive
-performance of the model (i.e testing AUROC values) decreased when each
-OTU or group of OTUs was randomly permuted. We ranked the OTUs based on
-how much the median testing AUROC decreased when it was permuted; the
-OTU with the largest decrease ranked highest {[}Figure 4{]}. Among the
-twenty OTUs with the largest impact for each of these models, there was
-only one OTU (OTU 822) that was shared among all of the models; however,
-we found three OTUs (OTU 58, 110, 367) that were important in each of
-the tree-based models. Similarly, the random forest and XGBoost models,
-shared four of the most important OTUs (OTU 2, 12, 361, 477).
-Permutation analysis results also revealed that with the exception of
-the decision tree model, removal of any individual OTU had a minimal
-impact on model performance. For example, if OTU 367 were permuted
-across the diagnoses from the decision tree model, the median AUROC
-dropped from 0.601 to 0.525. In contrast, if we permuted the same OTU
-from the random forest model, the AUROC only dropped from 0.695 to
-0.6799. Effectively, the complexity of the communities was more fully
-represented in the better performing models(22, 32). At least in this
+importance using permutation importance. Whereas the absolute feature
+weights were determined from the trained models, here we measured
+importance using the held-out test data. Permutation importance analysis
+is a posthoc explanation of the model where we randomly permuted
+non-correlated features individually and groups of perfectly correlated
+features across the two groups in the held-out test data. We then
+calculated how much the predictive performance of the model (i.e testing
+AUROC values) decreased when each OTU or group of OTUs was randomly
+permuted. We ranked the OTUs based on how much the median testing AUROC
+decreased when it was permuted; the OTU with the largest decrease ranked
+highest {[}Figure 4{]}. Among the twenty OTUs with the largest impact,
+there was only one OTU (OTU 822) that was shared among all of the
+models; however, we found three OTUs (OTU 58, 110, 367) that were
+important in each of the tree-based models. Similarly, the random forest
+and XGBoost models, shared four of the most important OTUs (OTU 2, 12,
+361, 477). Permutation analysis results also revealed that with the
+exception of the decision tree model, removal of any individual OTU had
+a minimal impact on model performance. For example, if OTU 367 was
+permuted across the samples in the decision tree model, the median AUROC
+dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
+in the random forest model, the AUROC only dropped from 0.695 to 0.68.
+Effectively, the complexity of the communities was more fully
+represented in the better performing models (22, 32). At least in this
 case, it was not possible to distinguish between health and disease
 using a single OTU. While permutation analysis allowed us to gauge the
-importance of an OTU, the analysis is post-hoc (i.e.~done using the test
-data) and these results did not allow us to directly interrogate the
-models to know whether an OTU is associated with classifying a subject
-as being healthy or having an SRN.
+importance of an OTU, the analysis was post-hoc (i.e.~done using the
+test data) and these results did not allow us to directly interrogate
+the models to know whether an OTU is associated with classifying a
+subject as being healthy or having an SRN.
 
@@
 methods, we used permutation importance to interpret the linear models
-{[}Figure S3{]}. When we analyzed the L1-regularized SVM with linear
+{[}Figure S4{]}. When we analyzed the L1-regularized SVM with linear
 kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S3{]}, 17 out of the 20 top OTUs
-(e.g.~OTU 609, 822, 1239) were deemed important by both approaches.
-Similarly, for the L2-regularized SVM and L2-regularized logistic
-regression, 9 and 12 OTUs, respectively, were shared among the two
-approaches. These results indicate that both approaches are consistent
-in selecting the most important OTUs.
+permutation importance {[}Figure S4{]}, 17 out of the 20 top OTUs
+(e.g.~OTU 609, 822, 1239) were deemed important by both interpretation
+methods. Similarly, for the L2-regularized SVM and L2-regularized
+logistic regression, 9 and 12 OTUs, respectively, were shared among the
+two interpretation methods. These results indicate that both methods are
+consistent in selecting the most important OTUs.
 
@@
 non-linear models {[}Figures S1-S2; Figure 5{]}. When we subsetted the
-size of the training dataset, we observed a {[}linear, quadratic, etc{]}
-relationship between the size of the dataset and the training time for
-the random forest model.
+size of the training dataset, we observed a linear relationship between
+the size of the dataset and the training times for L2-regularized
+logistic regression and random forest models {[}Figure S5{]}.
 
@@
 
-Overview • Establish pipeline that can be generalized to any modeling
-method • Linear/non-linear not a bfd • Interpretability tradeoff Point 1
-• Importance for transparency, picking the most complex model is not
-necessarilty the best model • Good / better / best: Table S1 • ``best''
-has multiple meanings Point 2 • Our paper is concerned with developing
-models to gain greater biological insights into what's going on • Still
-need biological validation and further experimentation
-
-Point 3 • These models show great opportunity to use microbiome to make
-diagnoses • We talked about classification, but others have used it in
-microbiome data to do regression, mulit-class -- pipeline is the same •
-Others have used this on shotgun metagenomic data and collinearity is
-expected to be an even larger problem since 5000 genes are all
-correlated with each other • Deployment is a different question that
-requires independent datasets to test on; we currently lack decent sized
-parallel datasets where the same type of sequence data and clinical data
-have been collected • We also have a problem with OTU-based approaches
-Conclusion/Future direction • ML is awesome and there's always new
-methods; our framework gives structure to investigators wanting to try
-these out. We encourage reseraches to apply multiple models to their
-data and report the results • Things like neural nets are attractive and
-hold great promise, but we need a ton more data to apply and they are
-impossible to interpret
-
 There is a growing awareness that many human diseases and environmental
-processes are not driven by single organism, but are the product of
-multiple bacterial populations. Whereas traditional statistical
-approaches are useful for identifying those cases where a single
-organism can drive a process, ML methods offer the ability to use many
-features of these communities to identify those cases. If it is possible
-to classify communities reliably, then it is hoped that we could
-identify those populations that are responsible for the classification.
-Because application of ML in microbiome studies is still in its infancy,
-we developed a framework to train rigorous, transparent and reproducible
-ML models {[}Table 1{]}.
-
-To test this framework, we benchmarked seven ML models to show that a
-clearly defined ML problem that is based on the goal of the microbiome
-study should inform our model selection. Our results showed that if the
-goal of a study is to learn the ecology behind a disease and to identify
-microbial biomarkers, we can create ML models that are inherently
-interpretable and easily trained without losing predictive power. In
-terms of predictive performance, random forest model had the best
-testing AUROC values compared to the other six models. However, the
-second best model was L2-regularized logistic regression with a median
-AUROC difference of only 0.015 compared to random forest. While random
-forest took 83.2 hours to train, L2-regularized logistic regression
-trained in 12 minutes. In terms of interpretability, random forest was a
-more complex ML model and could only be explained using post-hoc methods
-such as permutation importance. On the other hand, L2-regularized
-logistic regression was easier to interpret by ranking absolute feature
-weights of the trained model.
-
-Even with interpretable models such as L2-regularized logistic
-regression, there are potential pitfalls when it comes to identifying
-biomarkers of a disease. Human-associated microbial communities have
-complex correlation structures. This can hinder our ability to reliably
-interpret models (33). In this study we used two different
-interpretation methods; ranking each OTU by (1) their absolute weights
-in the trained models and (2) their impact on the predictive performance
-based on permutation importance. The feature weights of correlated OTUs
-are influenced by and dependent on one another, which makes it harder to
-identify the importance of unique OTUs. To avoid misinterpreting the
-models, once we identify highly ranked OTUs, we should check for their
-relationships with other OTUs as well. These relationships will help us
-generate new hypotheses about the ecology of the disease and test with
-follow-up experiments. On the other hand, when we used permutation
-importance, we grouped correlated OTUs to determine their impact as a
-group. In this study, we grouped OTUs that had a perfect correlation
-with each other however, we can reduce the correlation threshold to
-further investigate the relationships among features to identify the
-true underlying factors when making a classification.
-
-In this study, we established a pipeline that can be generalized to any
-modeling method that uses 16S rRNA sequence counts to predict a binary
-health outcome. We used a held-out test set and we performed the initial
-80\%-20\% random datasplit to create the held-out test set 100 times in
-our ML pipeline. The models we built with this pipeline were
-generalizable despite the high number of features microbiome datasets
-usually have.
-
-We used a CRC dataset to develop ML models with a binary classification
-task. We did not evaluate multicategory classification methods or
-regression analyses to predict non-binary outcomes. However, the
-principles highlighted throughout this study {[}Table 2{]} apply to all
-ML modeling tasks with microbiome data. Our analysis was also limited to
-shallow learning methods and did not explore deep learning methods such
-as neural networks. Microbiome datasets often suffer from having high
-dimensionality but low sample sizes which makes deep learning models
-prone to overfitting.
-
-There are studies that address overcoming these challenges in biomedical
-datasets (11, 34, 35), however sudies that estalish frameworks with
-microbiome data are lacking. This would be an interesting direction for
-future work in microbiome studies.
-
-Paragraph about deploying a model.
+processes are not driven by a single organism but are the product of
+multiple bacterial populations. Traditional statistical approaches are
+useful for identifying those cases where a single organism is associated
+with a process. In contrast, ML methods offer the ability to incorporate
+the structure of the microbial communities as a whole to classify them
+into different categories such as coming from a patient who is healthy
+or has SRNs. If it is possible to classify communities reliably, then ML
+methods also offer the ability to identify those microbial populations
+within the communities that are responsible for the classification.
+However, the application of ML in microbiome studies is still in its
+infancy and the field still needs to develop a better understanding of
+different ML methods, their strengths and weaknesses, and how to
+implement them.
+
+To address these needs, we developed a framework to train rigorous,
+transparent, and reproducible models. We benchmarked seven ML models and
+showed that we can create models that are inherently interpretable and
+easily trained without losing predictive performance. In terms of
+predictive performance, the random forest model had the best testing
+AUROC values compared to the other six models. However, the second-best
+model was L2-regularized logistic regression with a median AUROC
+difference of only 0.015 compared to random forest. While random forest
+took 83.2 hours to train, L2-regularized logistic regression trained in
+12 minutes. In terms of interpretability, random forest was a complex ML
+model and could only be explained using post-hoc methods such as
+permutation importance. On the other hand, L2-regularized logistic
+regression was easier to interpret by ranking the OTUs based on their
+feature weights and reviewing the signs of these weights. Comparing many
+different models showed us that the most complex model is not
+necessarily the best model for our ML task.
+
+As we set out to select the best model, we established a pipeline that
+can be generalized to any modeling method that uses 16S rRNA sequence
+counts to predict a binary health outcome. We performed a random
+datasplit to create a training set (80\% of the data) and a held-out
+test set (20\% of the data), which we used to evaluate predictive
+performance. We repeated this datasplit 100 times to measure predictive
+performance. During the training, we tuned the model hyperparameters
+with a repeated five-fold cross-validation. Despite the high number of
+features microbiome datasets typically have, the models we built with
+this pipeline were generalizable as shown by the similar AUROC values
+from the cross-validation and testing.
+
+We highlighted the importance of model interpretation to gain greater
+biological insights into microbiota-associated diseases. In this study
+we showcased two different interpretation methods: ranking each OTU by
+(i) their absolute weights in the trained models and (ii) their impact
+on the predictive performance based on permutation importance.
+Human-associated microbial communities have complex correlation
+structures which create collinearity in the datasets we work with. This
+can hinder our ability to reliably interpret models because the feature
+weights of correlated OTUs are influenced by one another (33). For
+example if one OTU is highly correlated with another OTU, only one of
+these OTUs would have a large feature weight thus hiding the importance
+of its correlated OTU. To capture all important features, once we
+identify highly ranked OTUs, we should review their relationships with
+other OTUs. These relationships will help us generate new hypotheses
+about the ecology of the disease and test them with follow-up
+experiments. When we used permutation importance, we took collinearity
+into consideration by grouping correlated OTUs to determine their impact
+as a group. We grouped OTUs that had a perfect correlation with each
+other however, we can reduce the correlation threshold to further
+investigate the relationships among features. It is important to know
+the correlation strutures of the data to avoid missinterpreting the
+models. This is likely to be a particular problem with shotgun
+metagenomic datasets where collinearity will be more pronunced due to
+many genes being correlated with one another because they come from the
+same chromosome. To identify the true underlying microbial factors of a
+disease, it is crucial to do correlation analyses and further
+experimentation for biological validation.
+
+In this study, we did not consider all possible modeling approaches.
+However, the principles highlighted throughout this study apply to all
+ML modeling tasks with microbiome data. For example, we did not evaluate
+multicategory classification methods to predict non-binary outcomes. For
+example, we could have trained models to differentiate between people
+with normal colons and those with adenomas or carcinomas (k=3
+categories). We did not perform this analysis because the clinically
+relevant diagnosis grouping was between patients with normal colons and
+those with SRNs. Furthermore, as number of categories to classify
+increase, more samples are required for each category to train a model.
+We also did not use regression-based analyses to predict a
+non-categorical outcome. We have previously used such an approach to
+train random forest models to predict fecal short-chain fatty acid
+concentrations based on microbiome data (34). Our analysis was also
+limited to shallow learning methods and did not explore deep learning
+methods such as neural networks. Deep learning methods hold great
+promise (11, 35, 36) but microbiome datasets often suffer from having
+many features and small sample sizes, which makes the deep learning
+models prone to overfitting. These methods are even more complex than
+random forest and XGBoost and are considered uninterpretable. There is
+great potential for applying ML approaches to microbiome data and we can
+choose which ML approach is best for the study.
+
+Our framework gives structure to investigators wanting to train,
+evaluate, and interpret their own ML models to identify OTUs that might
+be biologically relevant. However, deploying microbiome-based models to
+make clinical diagnoses or predictions is a significantly harder and
+distinct undertaking. For example, we currently lack standardized
+methods to collect patient samples, generate sequence data, and report
+clinical data. We are also challenged by the practical constraints of
+OTU-based approaches. The de novo algorithms commonly in use are slow,
+require considerable memory, and result in different OTU assignments as
+new data are added. Finally, we also need independent validation cohorts
+to test the performance of a diagnostic model. To realize the potential
+for using ML approaches with microbiome data, it is necessary that we
+direct our efforts to overcome these challenges.
 
 This study highlighted the need to make educated choices at every step
-of developing a ML model with microbiome data. Model selection should be
-done with a solid understanding of model complexity and
-interpretability, rigorous ML pipelines should be built with
-cross-validation for hyperparameter tuning and with a held-out test set
-for evaluating predictive performance and models should be interpreted
-while considering collinearity in datasets. The right methods will help
-us achieve the level of validity and accountability we want from models
-built for patient health.
+of developing a ML model with microbiome data. We created an
+aspirational rubric that researchers can use to identify potential
+pitfalls when using ML in microbiome studies and ways to avoid them
+{[}Table S1{]}. We have highlighted the tradeoffs between model
+complexity and interpretability, the need for cross-validation to tune
+hyperparameters, the utility of held-out test sets for evaluating
+predictive performance, and the importance of considering correlation
+structures in datasets for reliable interpretation. Furthermore, we
+underscored the importance of proper experimental design and methods to
+help us achieve the level of validity and accountability we want from
+models built for patient health.
 
@@
 
-\textbf{Data collection and study population.} The data used for this
-analysis are stool bacterial abundances and clinical information of the
-patients recruited by Great Lakes-New England Early Detection Research
-Network study. These data were obtained from Sze et al (32). The stool
-samples were provided by recruited adult participants who were
-undergoing scheduled screening or surveillance colonoscopy.
-Colonoscopies were performed and fecal samples were collected from
-participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
+\textbf{Data collection and study population.} The original stool
+samples described in our analysis were obtained from patients recruited
+by Great Lakes-New England Early Detection Research Network (Reference
+from Mack?). Stool samples were provided by adults who were undergoing a
+scheduled screening or surveillance colonoscopy. Participants were
+recruited from four locations: Toronto (ON, Canada), Boston (MA, USA),
 Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
-labeled by colonoscopy with adequate preparation and tissue
-histopathology of all resected lesions. Patients with an adenoma greater
-than 1 cm, more than three adenomas of any size, or an adenoma with
-villous histology were classified as advanced adenoma. Study had 172
-patients with normal colonoscopies, 198 with adenomas and 120 with
-carcinomas. Of the 198 adenomas, 109 were identified as advanced
-adenomas. Stool provided by the patients was used for 16S rRNA gene
-sequencing to measure bacterial population abundances. The bacterial
-abundance data was generated by Sze et al, by processing 16S rRNA
-sequences in Mothur (v1.39.3) using the default quality filtering
-methods, identifying and removing chimeric sequences using VSEARCH and
-assigning to OTUs at 97\% similarity using the OptiClust algorithm
-(36--38).
-
-\textbf{Data definitions and pre-processing.}
-
-The colorectal health of the patient was defined as two encompassing
-classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class
-includes patients with non-advanced adenomas or normal colons whereas
-SRN class includes patients with advanced adenomas or carcinomas. The
-study had 261 normal and 229 SRN samples. The bacterial abundances are
-the features used to predict colorectal health of the patients. For each
-patient, we had 6920 features (fecal bacterial abundances) and a
-two-class label that defines their colorectal health (normal or SRN
-colorectal lesions as defined by colonoscopies). We established modeling
-pipelines for a binary prediction task Bacterial abundances are discrete
-data in the form of Operational Taxonomic Unit (OTU) counts. OTU counts
-were set to the size of our smallest sample and were subsampled at the
-same distances. They were then transformed by scaling to a {[}0-1{]}
-range.
-
-\textbf{Model training and evaluation.}
-
-Models were trained using the machine learning wrapper caret package
-(v.6.0.81) in R (v.3.5.0). Within the caret package, we have made
-modifications to L2-regularized SVM with linear kernel function
-\textbf{svmLinear3} and developed a L1-regularized SVM with linear
-kernel function \textbf{svmLinear4} to calculate decision values instead
-of predicted probabilities. These changes are available at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/}.
-
-For L2-regularized logistic regression, L1 and L2 support vector
-machines (SVM) with linear and radial basis function kernels we tuned
-the \textbf{cost} hyperparameter which determines the regularization
+visually assessed by colonoscopy with bowel preparation and tissue
+histopathology of all resected lesions. We assigned patients into two
+classes: those with normal colons and those with screen relevant
+neoplasias (SRNs). The normal class included patients with normal colons
+or non-advanced adenomas whereas the SRN class included patients with
+advanced adenomas or carcinomas (reference for SRN?). Patients with an
+adenoma greater than 1 cm, more than three adenomas of any size, or an
+adenoma with villous histology were classified as having advanced
+adenomas. The study had 172 patients with normal colonoscopies, 198 with
+adenomas, and 120 with carcinomas. Of the 198 adenomas, 109 were
+identified as advanced adenomas. The study collected samples from 261
+patients classified as normal and 229 patients classified as having a
+SRN.
+
+\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
+patients was used for 16S rRNA gene sequencing to measure bacterial
+population abundances. The sequence data used in our analyses were
+originally generated by Baxter et al. (available through NCBI Sequence
+Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
+generated by Sze et al (32), who processed the 16S rRNA sequences in
+mothur (v1.39.3) using the default quality filtering methods,
+identifying and removing chimeric sequences using VSEARCH, and assigning
+to OTUs at 97\% similarity using the OptiClust algorithm ((37--39);
+GitHub link to table). These OTU abundances were the features we used to
+predict colorectal health of the patients. There were 6920 OTUs. OTU
+abundances were subsampled to the size of the smallest sample and
+normalized across samples such that the highest abundance of each OTU
+would be 1 and lowest would be 0.
+
+\textbf{Model training and evaluation.} Models were trained using the
+caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
+calculate decision values instead of predicted probabilities for models
+generated using L2-regularized SVM with linear kernel and L1-regularized
+SVM with linear kernel. These changes were necessary to calculate AUROC
+values for SVMs. The code for these changes are available at
+\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.XX_2019/}.
+
+For hyperparameter selection, we started with a granular grid search.
+Then we narrowed and fine-tuned the range of each hyperparameter. The
+range of the grid changes with different ML tasks and ML models. A full
+grid search was needed to avoid large variation in prediction
+performance. For L2-regularized logistic regression, L1 and
+L2-regularized SVM with linear and radial basis function kernels, we
+tuned the \emph{cost} hyperparameter which determines the regularization
 strength where smaller values specify stronger regularization. For SVM
-with radial basis function kernel we also tuned \textbf{sigma}
+with radial basis function kernel we also tuned \emph{sigma}
 hyperparameter which determines the reach of a single training instance
@@
 dependent on the points that are closest to the decision boundary. For
-the decision tree model, we tuned the \textbf{depth of the tree} where
+the decision tree model, we tuned the \emph{depth of the tree} where
 deeper the tree, the more splits it has. For random forest, we tuned the
-\textbf{number of features} to consider when looking for the best tree
-split. For XGBoost, we tuned for \textbf{learning rate} and the
-\textbf{fraction of samples} to be used for fitting the individual base
-learners.For hyperparameter selection, we started with a granular grid
-search. Then we narrowed and fine-tuned the range of each
-hyperparameter. The range of the grid depends on the ML task and ML
-model. A full grid search needs to be performed to avoid variability in
-testing performance. We can use hyper-band to help us with our
-hyperparameter selection (39).
+\emph{number of features} to consider when looking for the best tree
+split. For XGBoost, we tuned for \emph{learning rate} and the
+\emph{fraction of samples} to be used for fitting the individual base
+learners. There are newly developed tools such as Hyperband that help
+researcher with hyperparameter selection (40) that can be incorporated
+to microbiome studies.
 
 The computational burden during model training due to model complexity
-was reduced by parallelizing segments of the ML pipeline. In this study
-we have parallelized each data-split which allowed 100 data-splits to be
-processed through the ML pipeline at the same time for each model. We
-can further parallelize the cross-validation step for each
-hyperparameter setting.
+was reduced by parallelizing segments of the ML pipeline. We
+parallelized the training of each data-split. This allowed 100
+data-splits to be processed through the ML pipeline at the same time for
+each model. It is possible to further parallelize the cross-validation
+step for each hyperparameter setting if limited by computational
+resources.
 
 \textbf{Permutation importance workflow.} We created a Spearman's
-rank-order correlation matrix, corrected for multiple pairwise
-comparisons. We then defined correlated OTUs as having perfect
-correlation (correlation coef=1 and p\textless{}0.01). Non-correlated
-OTUs were permuted individually whereas correlated ones were grouped
-together and permuted at the same time.
+rank-order correlation matrix and defined correlated OTUs as having
+perfect correlation (correlation coefficient = 1 and p \textless{}
+0.01). Non-correlated OTUs were permuted individually whereas correlated
+ones were grouped together and permuted at the same time.
 
@@
 
-\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
-showing predictive model training and evaluation flowchart. } We split
-the data 80\%/20\% stratified to maintain the overall label
-distribution, performed five-fold cross-validation on the training data
-to select the best hyperparameter setting and then using these
-hyperparameters to train all of the training data. The model was
-evaluated on a held-out set of data (not used in selecting the model).
-Abbreviations: cvAUROC, cross-validation area under the receiver
-operating characteristic curve \newpage
+\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline.
+} We split the data to create a training (80\%) and held-out test set
+(20\%). The splits were stratified to maintain the overall label
+distribution. We performed five-fold cross-validation on the training
+data to select the best hyperparameter setting and then used these
+hyperparameters to train the models. The model was evaluated on the
+held-out data set. Abbreviations: cvAUC, cross-validation area under the
+receiver operating characteristic curve. \newpage
 \includegraphics{Figure_2.png}
@@
 of random forest model was higher than other ML models. The boxplot
-shows quartiles at the box ends and the statistical median as the
-horizontal line in the box. The whiskers show the farthest points that
-are not outliers. Outliers are data points that are not within 3/2 times
-the interquartile ranges. Abbreviations: SRN, screen-relevant
-neoplasias; AUROC, area under the receiver operating characteristic
-curve; SVM, support vector machine; XGBoost, extreme gradient boosting
+shows quartiles at the box ends and the median as the horizontal line in
+the box. The whiskers show the farthest points that were not outliers.
+Outliers were defined as those data points that are not within 1.5 times
+the interquartile ranges. \newpage
+\includegraphics[height=17cm, width=12cm]{Figure_3.png}
+
+\textbf{Figure 3. Interpretation of the linear ML models.} The absolute
+feature weights of (A) L2-regularized logistic regression, (B)
+L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with
+linear kernel were ranked from highest rank, 1, to lowest rank, 100, for
+each data-split. The feature ranks of the 20 highest ranked OTUs based
+on their median ranks (median shown in black) are reported here. OTUs
+that are associated with classifying a subject as being healthy or
+having an SRN are in blue and red, respectively. Some of the same OTUs
+were identified as important in all of the linear models. \newpage
+\includegraphics{Figure_4.png}
+
+\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
+with radial basis kernel, (B) decision tree, (C) random forest, and (D)
+XGBoost feature importances were explained using permutation importance
+on the held-out test data set. The gray rectangle and the dashed line
+show the IQR range and median of the base testing AUROC without any
+permutation. The colors of the box plots represent the OTUs that were
+shared among the different models; yellow were OTUs that were shared
+among all the non-linear models, salmon were OTUs that were shared among
+the tree-based models, green were the OTUs shared among SVM with radial
+basis kernel, decision tree and XGBoost, pink were the OTUs shared among
+SVM with radial basis kernel and XGBoost only, red were the OTUs shared
+among random forest and XGBoost only and blue were the OTUs shared among
+decision tree and random forest only. For all of the tree-based models,
+a Peptostreptococcus species (OTU00367) had the largest impact on
+predictive performance. \newpage
+\includegraphics{Figure_5.png}
+
+\textbf{Figure 5. Computational efficiency of seven ML models.} The
+median training time was the highest for XGBoost and shortest for
+L2-regularized logistic regression. \newpage
+\includegraphics{Figure_S1.png}
+
+\textbf{Figure S1. Hyperparameter setting performances for linear
+models.} (A) L2 logistic regression, (B) L1 SVM with linear kernel, and
+(C) L2 SVM with linear kernel mean cross-validation AUROC values when
+different hyperparameters were used in training the model. The
+differences in AUROC values when hyperparameters change show that
+hyperparameter tuning is a crucial step in building a ML model.
+
 \newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.png}
-
-\textbf{Figure 3. Interpretation of the linear ML models.} The absolute
-feature weights of (A) L2 logistic regression coefficients (B) L1 SVM
-with linear kernel (C) L2 SVM with linear kernel were ranked from
-highest rank 1 to 100 for each data-split. The feature ranks of the
-highest ranked five OTUs based on their median ranks are shown here.
-Similar OTUs had the largest impact on the predictive performance of L2
-logistic regression and L2 SVM with linear kernel. Abbreviations: SVM,
-support vector machine; OTU, Operational Taxonomic Unit. \newpage
-\includegraphics{Figure_4.png}
-
-\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
-with radial basis kernel (B) decision tree (C) random forest (D) XGBoost
-feature importances were explained using permutation importance using
-held-out test set. The gray rectangle and the dashed line show the IQR
-range and median of the base testing AUROC without any permutation
-performed. The colors of the box plots stand for the unique OTUs that
-are shared among the different models; pink for OTU0008, salmon for
-OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and
-red for OTU00882. For all the tree-based models, a
-\emph{Peptostreptococcus} species (OTU00367) had the largest impact on
-predictive performance of the model. Abbreviations: SVM, support vector
-machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
-Operational Taxonomic Unit. \newpage
-\includegraphics{Figure_5.png}
-
-\textbf{Figure 5. Computational efficiency of seven ML models.} The
-training times for of each data-split showed the differences in
-computational efficiency of the seven models. The median training time
-in hours was the highest for XGBoost and shortest for L1-regularized SVM
-with linear kernel. The boxplot shows quartiles at the box ends and the
-statistical median as the horizontal line in the box. The whiskers show
-the farthest points that are not outliers. Outliers are data points that
-are not within 3/2 times the interquartile ranges. Abbreviations: AUROC,
-area under the receiver operating characteristic curve; SVM, support
-vector machine; XGBoost, extreme gradient boosting. \newpage
-\includegraphics{Figure_S1.png}
-
-\textbf{Figure S1. Hyperparameter setting performances for linear
-models.} (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2
-SVM with linear kernel mean cross-validation AUROC values when different
-hyperparameters are used in training the model. The differences in AUROC
-values when hyperparameters change show that hyperparameter tuning is a
-crucial step in building a ML model.
+
+\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
+
+\textbf{Figure S2. Hyperparameter setting performances for non-linear
+models.} (A) Decision tree, (B) Random forest, (C) SVM with radial basis
+kernel, and (D) XGBoost mean cross-validation AUROC values when
+different hyperparameters were used in training the model. The
+differences in AUROC values when hyperparameters change show that
+hyperparameter tuning is a crucial step in building a ML model. \newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
+
+\textbf{Figure S3. Classification performance of ML models across cross
+validation when dataset is subsetted .} (A) L2-regularized logistic
+regression and (B) Random forest models were trained using the original
+study design with 490 subjects and subsets of it with 15, 30, 60, 120,
+and 245 subjects. The range among the cross-validation AUROC values
+within both models at lower sample sizes were much larger than when the
+full collection of samples was used to train and validate the models,
+but included the ranges observed with the more complete datasets.
 
@@
 
-\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
-
-\textbf{Figure S2. Hyperparameter setting performances for non-linear
-models.} (A) Decision tree (B) Random forest (C) SVM with radial basis
-kernel (D) XGBoost mean cross-validation AUROC values when different
-hyperparameters are used in training the model. The differences in AUROC
-values when hyperparameters change show that hyperparameter tuning is a
-crucial step in building a ML model. \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
-
-\textbf{Figure S3. Interpretation of the linear ML models with
-permutation importance.} (A) L1-regularized SVM with linear kernel (B)
-L2-regularized SVM with linear kernel and (C) L2-regularized logistic
+\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
+
+\textbf{Figure S4. Interpretation of the linear ML models with
+permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
+L2-regularized SVM with linear kernel, and (C) L2-regularized logistic
 regression were interpreted using permutation importance using held-out
-test set. The gray rectangle and the dashed line show the IQR range and
-median of the base testing AUROC without any permutation performed.
-Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
-Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
+test set.
+
+\newpage
+
+\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
+
+\textbf{Figure S5. Training times of ML models when dataset is subsetted
+.} (A) L2-regularized logistic regression and (B) Random forest models
+were trained using the original study design with 490 subjects and
+subsets of it with 15, 30, 60, 120, and 245 subjects. As the size of the
+dataset increased, the training times for L2-regularized logistic
+regression and random forest models increased linearly.
 
@@
 
+\hypertarget{ref-sze_fecal_2019}{}
+34. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
+acids are not predictive of colonic tumor status and cannot be predicted
+based on bacterial community structure. mBio \textbf{10}:e01454--19.
+doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.
+
 \hypertarget{ref-kocheturov_massive_2019}{}
-34. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+35. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \hypertarget{ref-kim_improved_2018}{}
-35. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+36. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-36. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+37. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-37. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+38. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-38. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+39. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-39. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+40. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-09-10 submission/manuscript.Rmd
add links

@@
 (available through NCBI Sequence Read Archive [SRP062005], 2015).
-The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm ([@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; GitHub link to table).
+The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
 These OTU abundances were the features we used to predict colorectal health of the patients.
@@
 These changes were necessary to calculate AUROC values for SVMs.
-The code for these changes are available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.XX_2019/.
+The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R}, respectively.
 
@@
 A full grid search was needed to avoid large variation in prediction performance.
-For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the _cost_ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
-For SVM with radial basis function kernel we also tuned _sigma_ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
-For the decision tree model, we tuned the _depth of the tree_ where deeper the tree, the more splits it has.
-For random forest, we tuned the _number of features_ to consider when looking for the best tree split.
-For XGBoost, we tuned for _learning rate_ and the _fraction of samples_ to be used for fitting the individual base learners.
+For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
+For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
+For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has.
+For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
+For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
 There are newly developed tools such as Hyperband that help researcher with hyperparameter selection [@li_hyperband:_2016] that can be incorporated to microbiome studies.
@@
 
-**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.
+**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
 

### 2019-09-10 submission/manuscript.tex
add links

@@
 \hypersetup{unicode=true,
-            pdftitle={Effective model selection and implementation when applying machine learning to bacterial 16S rRNA gene sequencing data},
+            pdftitle={Effective application of machine learning to bacterial 16S rRNA gene sequencing data},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Effective model selection and implementation when applying
-machine learning to bacterial 16S rRNA gene sequencing data}}
+  \title{\textbf{Effective application of machine learning to bacterial 16S rRNA
+gene sequencing data}}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 \usepackage{caption}
+\usepackage{hyperref}
 
@@
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm ((37--39);
-GitHub link to table). These OTU abundances were the features we used to
-predict colorectal health of the patients. There were 6920 OTUs. OTU
-abundances were subsampled to the size of the smallest sample and
-normalized across samples such that the highest abundance of each OTU
-would be 1 and lowest would be 0.
+to OTUs at 97\% similarity using the OptiClust algorithm (37--39);
+(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
+These OTU abundances were the features we used to predict colorectal
+health of the patients. There were 6920 OTUs. OTU abundances were
+subsampled to the size of the smallest sample and normalized across
+samples such that the highest abundance of each OTU would be 1 and
+lowest would be 0.
 
@@
 SVM with linear kernel. These changes were necessary to calculate AUROC
-values for SVMs. The code for these changes are available at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.XX_2019/}.
+values for SVMs. The code for these changes on L2-regularized SVM with
+linear kernel and L1-regularized SVM with linear kernel are available at
+\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
+and at
+\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
+respectively.
 
@@
 L2-regularized SVM with linear and radial basis function kernels, we
-tuned the \emph{cost} hyperparameter which determines the regularization
-strength where smaller values specify stronger regularization. For SVM
-with radial basis function kernel we also tuned \emph{sigma}
-hyperparameter which determines the reach of a single training instance
-where for a high value of sigma, the SVM decision boundary will be
-dependent on the points that are closest to the decision boundary. For
-the decision tree model, we tuned the \emph{depth of the tree} where
-deeper the tree, the more splits it has. For random forest, we tuned the
-\emph{number of features} to consider when looking for the best tree
-split. For XGBoost, we tuned for \emph{learning rate} and the
-\emph{fraction of samples} to be used for fitting the individual base
-learners. There are newly developed tools such as Hyperband that help
-researcher with hyperparameter selection (40) that can be incorporated
-to microbiome studies.
+tuned the \textbf{cost} hyperparameter which determines the
+regularization strength where smaller values specify stronger
+regularization. For SVM with radial basis function kernel we also tuned
+\textbf{sigma} hyperparameter which determines the reach of a single
+training instance where for a high value of sigma, the SVM decision
+boundary will be dependent on the points that are closest to the
+decision boundary. For the decision tree model, we tuned the
+\textbf{depth of the tree} where deeper the tree, the more splits it
+has. For random forest, we tuned the \textbf{number of features} to
+consider when looking for the best tree split. For XGBoost, we tuned for
+\textbf{learning rate} and the \textbf{fraction of samples} to be used
+for fitting the individual base learners. There are newly developed
+tools such as Hyperband that help researcher with hyperparameter
+selection (40) that can be incorporated to microbiome studies.
 
@@
 analysis steps including an Rmarkdown version of this manuscript is
-available at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/}.
+available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
 

### 2019-09-11 submission/manuscript.Rmd
version sent to Jenna

@@
 # Median of base testing AUC for each model
-# 100 datasplit permutation importances new testing AUC for top 5 features
+# 100 data-split permutation importances new testing AUC for top 5 features
 ######################################################################
@@
 
-Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes, and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but inconsistent adoption of methods call the validity of these models into question.
+Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer.
+Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training methods call the validity of these models into question.
 Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability.
-To overcome these challenges, we trained seven models that used fecal 16S rRNA sequences to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
+To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
 We developed a generalizable pipeline to train, validate and interpret the models.
-To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
-The random forest model was best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
-Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, it trained much faster (12 min) and was more interpretable.
-Our analysis showed that ML models should be chosen based on the goal of the study as it will inform our expectations of predictive performance and interpretability.
+To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
+The random forest model performed best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
+Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, trained much faster (12 min), and was more interpretable.
+Our analysis showed that ML models should be chosen based on the goal of the study as the choice will inform expectations of performance and interpretability.
 
@@
 This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study.
-These concepts are not particular to study of health outcomes but can also be applied to environmental microbiology studies as well.
+These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.
 
@@
 Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually.
-Recently, machine learning models have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.
+Recently, machine learning models (ML) have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.
 
 ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data.
-Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
+Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
 The task of diagnosing an individual with high confidence relies on a model that is built with rigorous methods.
@@
 These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances.
-Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing analysis frameworks to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
+Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
 More work is needed to further improve reproducibility and minimize over-optimism for model performance.
@@
 
-__Model selection and pipeline construction__ We established a ML pipeline where we trained and validated each of the seven models using a common approach that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].
+__Model selection and pipeline construction__ We established a common ML pipeline to train and validate each of the seven models that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].
 
@@
 Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well.
-For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features.
+For regularized logistic regression and SVM with linear kernel, we used L2-regularization to keep all potentially important features.
 For comparison, we also trained an L1-regularized SVM model with linear kernel.
-L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero).
-To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with non-linear kernel.
+L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero).
+To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with a non-linear kernel.
 
-Third, fitting of these models require selecting appropriate hyperparameters.
+Third, fitting of these models required selecting appropriate hyperparameters.
 Hyperparameters are the rules that are learned from the training set in a classification algorithm.
@@
 Finally, we trained the full training dataset with the selected hyperparameter values and applied the model to the held-out data to evaluate the testing predictive performance of each model.
-The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].
+The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a reliable and robust interpretation of model performance [Figure 1].
 
-__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of seven models to classify individuals as having normal colons or SRNs [Figure 2].
+__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2].
 The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01).
 The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively.
-L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(format(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively.
+L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
 Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well as random forest.
@@
 If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit.
-The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.
+The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
+These differences are relatively small and would not indicate a problem with overfitting.
 
 To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 data-splits.
-The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` on average across the seven models.
+The range among the testing AUROC values within each model varied by `r paste(format(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3), nsmall=3))` on average across the seven models.
 If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model.
-For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`.
+For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(format(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3), nsmall=3))`.
 These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
-Therefore, it is important to employ the hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.
+Therefore, it is important to employ the 100 hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.
 
@@
 __Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017].
-Because we often use ML models not just to predict a health outcome but also to learn the ecology behind a disease, model interpretation becomes crucial for microbiome studies.
+Because we often use ML models not just to predict a health outcome but also to identify the biomarkers for a disease, model interpretation becomes crucial for microbiome studies.
 ML models decrease in interpretability as they increase in complexity.
@@
 First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3].
-We also reviewed the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having an SRN.
+We also reviewed the signs of feature weights to determine whether an OTU was associated with classifying a subject as being healthy or having an SRN.
 It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g.
@@
 The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU.
-However, this approach was only possible with linear models.
+However, this approach is only possible with linear models.
 
@@
 For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`.
-In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`.
+In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.
 Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019].
 At least in this case, it was not possible to distinguish between health and disease using a single OTU.
-While permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e.
-done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.
+Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e.
+done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.
 
 To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4].
-When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 out of the 20 top OTUs (e.g.
+When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 of the 20 top OTUs (e.g.
 OTU 609, 822, 1239) were deemed important by both interpretation methods.
 Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods.
-These results indicate that both methods are consistent in selecting the most important OTUs.
+Although permutation analysis does not not allow us to determine the weight or the sign of the features, these results indicate that both methods are consistent in selecting the most important OTUs.
 
@@
 If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification.
-However, the application of ML in microbiome studies is still in its infancy and the field still needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.
+However, the application of ML in microbiome studies is still in its infancy and the field needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.
 
@@
 While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance.
+In terms of interpretability, random forest is a complex ML model and could only be explained using post-hoc methods such as permutation importance.
 On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights.
-Comparing many different models showed us that the most complex model is not necessarily the best model for our ML task.
+Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.
 
 As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome.
-We performed a random datasplit to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance.
-We repeated this datasplit 100 times to measure predictive performance.
+We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance.
+We repeated this data-split 100 times to measure the possible variation in predictive performance.
 During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation.
@@
 In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance.
-Human-associated microbial communities have complex correlation structures which create collinearity in the datasets we work with.
+Human-associated microbial communities have complex correlation structures which create collinearity in the datasets.
 This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013].
@@
 We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.
-It is important to know the correlation strutures of the data to avoid missinterpreting the models.
-This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronunced due to many genes being correlated with one another because they come from the same chromosome.
+It is important to know the correlation structures of the data to avoid misinterpreting the models.
+This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome.
 To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.
@@
 In this study, we did not consider all possible modeling approaches.
-However, the principles highlighted throughout this study apply to all ML modeling tasks with microbiome data.
+However, the principles highlighted throughout this study apply to other ML modeling tasks with microbiome data.
 For example, we did not evaluate multicategory classification methods to predict non-binary outcomes.
-For example, we could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories).
+We could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories).
 We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs.
@@
 
-This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data.
+Our study highlights the need to make educated choices at every step of developing a ML model with microbiome data.
 We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1].
-We have highlighted the tradeoffs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
+We have highlighted the trade-offs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
 Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.
@@
 Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy.
-Participants were recruited from four locations: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
+Participants were recruited from: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
 Patients’ colonic health was visually assessed by colonoscopy with bowel preparation and tissue histopathology of all resected lesions.
@@
 Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas.
-The study had 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
+There were 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
 Of the 198 adenomas, 109 were identified as advanced adenomas.
-The study collected samples from 261 patients classified as normal and 229 patients classified as having a SRN.
+Together 261 patients were classified as normal and 229 patients were classified as having a SRN.
 
-**16S rRNA gene sequencing data.** Stool samples provided by the patients was used for 16S rRNA gene sequencing to measure bacterial population abundances.
+**16S rRNA gene sequencing data.** Stool samples provided by the patients were used for 16S rRNA gene sequencing to measure bacterial population abundances.
 The sequence data used in our analyses were originally generated by Baxter et al.
@@
 These changes were necessary to calculate AUROC values for SVMs.
-The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R}, respectively.
+The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel models are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R}, respectively.
 
@@
 Then we narrowed and fine-tuned the range of each hyperparameter.
-The range of the grid changes with different ML tasks and ML models.
 A full grid search was needed to avoid large variation in prediction performance.
@@
 For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
-There are newly developed tools such as Hyperband that help researcher with hyperparameter selection [@li_hyperband:_2016] that can be incorporated to microbiome studies.
+Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016] .
 
@@
 We parallelized the training of each data-split.
-This allowed 100 data-splits to be processed through the ML pipeline at the same time for each model.
+This allowed the 100 data-splits to be processed through the ML pipeline simultaneously at the same time for each model.
 It is possible to further parallelize the cross-validation step for each hyperparameter setting if limited by computational resources.
 
-**Permutation importance workflow.**
-We created a Spearman's rank-order correlation matrix and defined correlated OTUs as having perfect correlation (correlation coefficient = 1 and p < 0.01).
+**Permutation importance workflow.** We calculated a Spearman's rank-order correlation matrix and defined correlated OTUs as having perfect correlation (correlation coefficient = 1 and p < 0.01).
 Non-correlated OTUs were permuted individually whereas correlated ones were grouped together and permuted at the same time.
@@
 The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.
-OTUs that are associated with classifying a subject as being healthy or having an SRN are in blue and red, respectively.
+OTUs that are associated with classifying a subject as being healthy had negative signs and were shown in blue.
+OTUs that are associated with classifying a subject having an SRN had positive signs and were shown in red.
 Some of the same OTUs were identified as important in all of the linear models.
@@
 \newpage
-\captionof{table}{Characteristics of the machine learning models in our comparative study.}
+\captionsetup{labelformat=empty}
+\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
 \small
@@
 \rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} & \textbf{Refs.} \\ \hline
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} \\ \hline
 
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable & $^{36}$\\ \hline
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable \\ \hline
 
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable & $^{37}$\\ \hline
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable \\ \hline
 
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ \\ \hline
 Decision tree & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline
+root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear & Interpretable \\ \hline
 
-Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ & $^{40-41}$  \\ \hline
+Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ \\ \hline
 
-XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ & $^{42-43}$ \\ \hline
+XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ \\ \hline
 
@@
 
-\captionof{table}{Table S1. An aspirational rubric for evaluating the rigor of ML practices.}
+\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices.}
 \small

### 2019-09-11 submission/manuscript.tex
version sent to Jenna

@@
 Machine learning (ML) modeling of the human microbiome has the potential
-to identify the microbial biomarkers and aid in diagnosis of many
-chronic diseases such as inflammatory bowel disease, diabetes, and
-colorectal cancer. Progress has been made towards developing ML models
-that predict health outcomes from bacterial abundances, but inconsistent
-adoption of methods call the validity of these models into question.
+to identify microbial biomarkers and aid in diagnosis of many diseases
+such as inflammatory bowel disease, diabetes, and colorectal cancer.
+Progress has been made towards developing ML models that predict health
+outcomes using bacterial abundances, but inconsistent adoption of
+training methods call the validity of these models into question.
 Furthermore, there appears to be a preference by many researchers to
@@
 these challenges, we trained seven models that used fecal 16S rRNA
-sequences to predict the presence of colonic screen relevant neoplasias
-(SRNs; n=490 patients, 261 controls and 229 cases). We developed a
-generalizable pipeline to train, validate and interpret the models. To
-show the effect of model selection, we assessed the predictive
-performance, interpretability, and computational efficiency of the
-following models: L2-regularized logistic regression, L1 and
-L2-regularized support vector machines (SVM) with linear and radial
-basis function kernels, a decision tree, random forest, and extreme
-gradient boosting (XGBoost). The random forest model was best at
-detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h)
-and hard to interpret. Despite its simplicity, L2-regularized logistic
-regression followed random forest in predictive performance with an
-AUROC of 0.680, it trained much faster (12 min) and was more
-interpretable. Our analysis showed that ML models should be chosen based
-on the goal of the study as it will inform our expectations of
-predictive performance and interpretability.
+sequence data to predict the presence of colonic screen relevant
+neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We
+developed a generalizable pipeline to train, validate and interpret the
+models. To show the effect of model selection, we assessed the
+predictive performance, interpretability, and computational efficiency
+of L2-regularized logistic regression, L1 and L2-regularized support
+vector machines (SVM) with linear and radial basis function kernels, a
+decision tree, random forest, and extreme gradient boosting (XGBoost).
+The random forest model performed best at detecting SRNs with an AUROC
+of 0.695 but it was slow to train (83.2 h) and hard to interpret.
+Despite its simplicity, L2-regularized logistic regression followed
+random forest in predictive performance with an AUROC of 0.680, trained
+much faster (12 min), and was more interpretable. Our analysis showed
+that ML models should be chosen based on the goal of the study as the
+choice will inform expectations of performance and interpretability.
 
@@
 emphasizing the importance of selecting ML models that reflect the goal
-of the study. These concepts are not particular to study of health
-outcomes but can also be applied to environmental microbiology studies
-as well.
+of the study. These concepts are not particular to the study of health
+outcomes but can also be applied to environmental microbiology studies.
 
@@
 or risk effects of each bacterial population individually. Recently,
-machine learning models have grown in popularity among microbiome
+machine learning models (ML) have grown in popularity among microbiome
 researchers because of the large amount of data that can now be
@@
 ecological basis of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--16,
+inflammatory bowel diseases, obesity, and type 2 diabetes (1--16,
 16--18). The task of diagnosing an individual with high confidence
@@
 validating their models on independent datasets (7, 18, 19) and
-introducing analysis frameworks to better use ML tools (20--23). More
-work is needed to further improve reproducibility and minimize
-over-optimism for model performance.
+introducing ways to better use ML tools (20--23). More work is needed to
+further improve reproducibility and minimize over-optimism for model
+performance.
 
@@
 
-\textbf{Model selection and pipeline construction} We established a ML
-pipeline where we trained and validated each of the seven models using a
-common approach that is based on standard methods within the ML
-community (REFS by Jenna?){[}Figure 1{]}.
+\textbf{Model selection and pipeline construction} We established a
+common ML pipeline to train and validate each of the seven models that
+is based on standard methods within the ML community (REFS by
+Jenna?){[}Figure 1{]}.
 
@@
 training data too well. For regularized logistic regression and SVM with
-linear kernel, we used L2 regularization to keep all potentially
+linear kernel, we used L2-regularization to keep all potentially
 important features. For comparison, we also trained an L1-regularized
 SVM model with linear kernel. L1-regularization on microbiome data led
-to a sparser solution (i.e., force many coefficients to zero). To
+to a sparser solution (i.e., forced many coefficients to zero). To
 explore the potential for non-linear relationships among features to
 improve classification, we trained tree-based models including decision
-tree, random forest, and XGBoost and we trained an SVM model with
+tree, random forest, and XGBoost and we trained an SVM model with a
 non-linear kernel.
 
-Third, fitting of these models require selecting appropriate
+Third, fitting of these models required selecting appropriate
 hyperparameters. Hyperparameters are the rules that are learned from the
@@
 data-split, hyperparameter selection, training and testing steps were
-repeated 100 times to get a reliable and robust reading of model
-performance {[}Figure 1{]}.
+repeated 100 times to obtain a reliable and robust interpretation of
+model performance {[}Figure 1{]}.
 
 \textbf{Predictive performance and generalizability of the seven
-models.} We evaluated the predictive performance of seven models to
+models.} We evaluated the predictive performance of the seven models to
 classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
@@
 significantly different from one another and had median AUROC values of
-0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
+0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
 0.056), respectively. L1-regularized SVM with linear kernel and decision
 tree had significantly lower AUROC values than the other ML models with
-median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
+median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
 {[}Figure 2{]}. Interestingly, these results demonstrate that the most
@@
 SVM with radial basis function kernel and decision tree with a
-difference of 0.007 and 0.006, respectively {[}Figure 2{]}; however,
-these differences are relatively small and would not indicate a problem
-with overfitting.
+difference of 0.007 and 0.006, respectively {[}Figure 2{]}. These
+differences are relatively small and would not indicate a problem with
+overfitting.
 
@@
 range of AUROC values for each model using 100 data-splits. The range
-among the testing AUROC values within each model varied by 0.23 on
+among the testing AUROC values within each model varied by 0.230 on
 average across the seven models. If we had only done a single split,
@@
 performance of the model. For instance, the lowest AUROC value of the
-random forest model was 0.593 whereas the highest was 0.81. These
+random forest model was 0.593 whereas the highest was 0.810. These
 results showed that depending on the data-split, the testing AUROC
 values showed great variability {[}Figure 2{]}. Therefore, it is
-important to employ the hierarchical data splits that were included in
-our pipeline to minimize the risk of over-optimism.
+important to employ the 100 hierarchical data splits that were included
+in our pipeline to minimize the risk of over-optimism.
 
@@
 (31). Because we often use ML models not just to predict a health
-outcome but also to learn the ecology behind a disease, model
+outcome but also to identify the biomarkers for a disease, model
 interpretation becomes crucial for microbiome studies. ML models
@@
 OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
-determine whether an OTU is associated with classifying a subject as
-being healthy or having an SRN - negative sign indicated being healthy
-and positive sign indicated having an SRN. It was encouraging that many
-of the highest ranked OTUs were shared across these three models,
-(e.g.~OTU 50, 426, 609, 822, 1239). The benefit of this approach was
-that the results of the analysis were based on the trained model
-parameters and provided information regarding the sign and magnitude of
-the impact of each OTU. However, this approach was only possible with
-linear models.
+determine whether an OTU was associated with classifying a subject as
+being healthy or having an SRN. It was encouraging that many of the
+highest ranked OTUs were shared across these three models, (e.g.~OTU 50,
+426, 609, 822, 1239). The benefit of this approach was that the results
+of the analysis were based on the trained model parameters and provided
+information regarding the sign and magnitude of the impact of each OTU.
+However, this approach is only possible with linear models.
 
@@
 dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
-in the random forest model, the AUROC only dropped from 0.695 to 0.68.
+in the random forest model, the AUROC only dropped from 0.695 to 0.680.
 Effectively, the complexity of the communities was more fully
@@
 case, it was not possible to distinguish between health and disease
-using a single OTU. While permutation analysis allowed us to gauge the
-importance of an OTU, the analysis was post-hoc (i.e.~done using the
+using a single OTU. Although permutation analysis allowed us to gauge
+the importance of an OTU, the analysis was post-hoc (i.e.~done using the
 test data) and these results did not allow us to directly interrogate
-the models to know whether an OTU is associated with classifying a
+the models to know whether an OTU was associated with classifying a
 subject as being healthy or having an SRN.
@@
 kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S4{]}, 17 out of the 20 top OTUs
-(e.g.~OTU 609, 822, 1239) were deemed important by both interpretation
-methods. Similarly, for the L2-regularized SVM and L2-regularized
-logistic regression, 9 and 12 OTUs, respectively, were shared among the
-two interpretation methods. These results indicate that both methods are
-consistent in selecting the most important OTUs.
+permutation importance {[}Figure S4{]}, 17 of the 20 top OTUs (e.g.~OTU
+609, 822, 1239) were deemed important by both interpretation methods.
+Similarly, for the L2-regularized SVM and L2-regularized logistic
+regression, 9 and 12 OTUs, respectively, were shared among the two
+interpretation methods. Although permutation analysis does not not allow
+us to determine the weight or the sign of the features, these results
+indicate that both methods are consistent in selecting the most
+important OTUs.
 
@@
 However, the application of ML in microbiome studies is still in its
-infancy and the field still needs to develop a better understanding of
+infancy and the field needs to develop a better understanding of
 different ML methods, their strengths and weaknesses, and how to
@@
 took 83.2 hours to train, L2-regularized logistic regression trained in
-12 minutes. In terms of interpretability, random forest was a complex ML
+12 minutes. In terms of interpretability, random forest is a complex ML
 model and could only be explained using post-hoc methods such as
@@
 feature weights and reviewing the signs of these weights. Comparing many
-different models showed us that the most complex model is not
+different models showed us that the most complex model was not
 necessarily the best model for our ML task.
@@
 counts to predict a binary health outcome. We performed a random
-datasplit to create a training set (80\% of the data) and a held-out
+data-split to create a training set (80\% of the data) and a held-out
 test set (20\% of the data), which we used to evaluate predictive
-performance. We repeated this datasplit 100 times to measure predictive
-performance. During the training, we tuned the model hyperparameters
-with a repeated five-fold cross-validation. Despite the high number of
-features microbiome datasets typically have, the models we built with
-this pipeline were generalizable as shown by the similar AUROC values
-from the cross-validation and testing.
+performance. We repeated this data-split 100 times to measure the
+possible variation in predictive performance. During the training, we
+tuned the model hyperparameters with a repeated five-fold
+cross-validation. Despite the high number of features microbiome
+datasets typically have, the models we built with this pipeline were
+generalizable as shown by the similar AUROC values from the
+cross-validation and testing.
 
@@
 Human-associated microbial communities have complex correlation
-structures which create collinearity in the datasets we work with. This
-can hinder our ability to reliably interpret models because the feature
-weights of correlated OTUs are influenced by one another (33). For
-example if one OTU is highly correlated with another OTU, only one of
-these OTUs would have a large feature weight thus hiding the importance
-of its correlated OTU. To capture all important features, once we
-identify highly ranked OTUs, we should review their relationships with
-other OTUs. These relationships will help us generate new hypotheses
-about the ecology of the disease and test them with follow-up
-experiments. When we used permutation importance, we took collinearity
-into consideration by grouping correlated OTUs to determine their impact
-as a group. We grouped OTUs that had a perfect correlation with each
-other however, we can reduce the correlation threshold to further
-investigate the relationships among features. It is important to know
-the correlation strutures of the data to avoid missinterpreting the
-models. This is likely to be a particular problem with shotgun
-metagenomic datasets where collinearity will be more pronunced due to
-many genes being correlated with one another because they come from the
-same chromosome. To identify the true underlying microbial factors of a
-disease, it is crucial to do correlation analyses and further
-experimentation for biological validation.
+structures which create collinearity in the datasets. This can hinder
+our ability to reliably interpret models because the feature weights of
+correlated OTUs are influenced by one another (33). For example if one
+OTU is highly correlated with another OTU, only one of these OTUs would
+have a large feature weight thus hiding the importance of its correlated
+OTU. To capture all important features, once we identify highly ranked
+OTUs, we should review their relationships with other OTUs. These
+relationships will help us generate new hypotheses about the ecology of
+the disease and test them with follow-up experiments. When we used
+permutation importance, we took collinearity into consideration by
+grouping correlated OTUs to determine their impact as a group. We
+grouped OTUs that had a perfect correlation with each other however, we
+can reduce the correlation threshold to further investigate the
+relationships among features. It is important to know the correlation
+structures of the data to avoid misinterpreting the models. This is
+likely to be a particular problem with shotgun metagenomic datasets
+where collinearity will be more pronounced due to many genes being
+correlated with one another because they come from the same chromosome.
+To identify the true underlying microbial factors of a disease, it is
+crucial to do correlation analyses and further experimentation for
+biological validation.
 
 In this study, we did not consider all possible modeling approaches.
-However, the principles highlighted throughout this study apply to all
+However, the principles highlighted throughout this study apply to other
 ML modeling tasks with microbiome data. For example, we did not evaluate
-multicategory classification methods to predict non-binary outcomes. For
-example, we could have trained models to differentiate between people
-with normal colons and those with adenomas or carcinomas (k=3
-categories). We did not perform this analysis because the clinically
-relevant diagnosis grouping was between patients with normal colons and
-those with SRNs. Furthermore, as number of categories to classify
-increase, more samples are required for each category to train a model.
-We also did not use regression-based analyses to predict a
-non-categorical outcome. We have previously used such an approach to
-train random forest models to predict fecal short-chain fatty acid
-concentrations based on microbiome data (34). Our analysis was also
-limited to shallow learning methods and did not explore deep learning
-methods such as neural networks. Deep learning methods hold great
-promise (11, 35, 36) but microbiome datasets often suffer from having
-many features and small sample sizes, which makes the deep learning
-models prone to overfitting. These methods are even more complex than
-random forest and XGBoost and are considered uninterpretable. There is
-great potential for applying ML approaches to microbiome data and we can
-choose which ML approach is best for the study.
+multicategory classification methods to predict non-binary outcomes. We
+could have trained models to differentiate between people with normal
+colons and those with adenomas or carcinomas (k=3 categories). We did
+not perform this analysis because the clinically relevant diagnosis
+grouping was between patients with normal colons and those with SRNs.
+Furthermore, as number of categories to classify increase, more samples
+are required for each category to train a model. We also did not use
+regression-based analyses to predict a non-categorical outcome. We have
+previously used such an approach to train random forest models to
+predict fecal short-chain fatty acid concentrations based on microbiome
+data (34). Our analysis was also limited to shallow learning methods and
+did not explore deep learning methods such as neural networks. Deep
+learning methods hold great promise (11, 35, 36) but microbiome datasets
+often suffer from having many features and small sample sizes, which
+makes the deep learning models prone to overfitting. These methods are
+even more complex than random forest and XGBoost and are considered
+uninterpretable. There is great potential for applying ML approaches to
+microbiome data and we can choose which ML approach is best for the
+study.
 
@@
 
-This study highlighted the need to make educated choices at every step
-of developing a ML model with microbiome data. We created an
-aspirational rubric that researchers can use to identify potential
-pitfalls when using ML in microbiome studies and ways to avoid them
-{[}Table S1{]}. We have highlighted the tradeoffs between model
-complexity and interpretability, the need for cross-validation to tune
-hyperparameters, the utility of held-out test sets for evaluating
-predictive performance, and the importance of considering correlation
-structures in datasets for reliable interpretation. Furthermore, we
-underscored the importance of proper experimental design and methods to
-help us achieve the level of validity and accountability we want from
-models built for patient health.
+Our study highlights the need to make educated choices at every step of
+developing a ML model with microbiome data. We created an aspirational
+rubric that researchers can use to identify potential pitfalls when
+using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
+have highlighted the trade-offs between model complexity and
+interpretability, the need for cross-validation to tune hyperparameters,
+the utility of held-out test sets for evaluating predictive performance,
+and the importance of considering correlation structures in datasets for
+reliable interpretation. Furthermore, we underscored the importance of
+proper experimental design and methods to help us achieve the level of
+validity and accountability we want from models built for patient
+health.
 
@@
 scheduled screening or surveillance colonoscopy. Participants were
-recruited from four locations: Toronto (ON, Canada), Boston (MA, USA),
-Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
-visually assessed by colonoscopy with bowel preparation and tissue
-histopathology of all resected lesions. We assigned patients into two
-classes: those with normal colons and those with screen relevant
-neoplasias (SRNs). The normal class included patients with normal colons
-or non-advanced adenomas whereas the SRN class included patients with
-advanced adenomas or carcinomas (reference for SRN?). Patients with an
-adenoma greater than 1 cm, more than three adenomas of any size, or an
-adenoma with villous histology were classified as having advanced
-adenomas. The study had 172 patients with normal colonoscopies, 198 with
-adenomas, and 120 with carcinomas. Of the 198 adenomas, 109 were
-identified as advanced adenomas. The study collected samples from 261
-patients classified as normal and 229 patients classified as having a
-SRN.
+recruited from: Toronto (ON, Canada), Boston (MA, USA), Houston (TX,
+USA), and Ann Arbor (MI, USA). Patients' colonic health was visually
+assessed by colonoscopy with bowel preparation and tissue histopathology
+of all resected lesions. We assigned patients into two classes: those
+with normal colons and those with screen relevant neoplasias (SRNs). The
+normal class included patients with normal colons or non-advanced
+adenomas whereas the SRN class included patients with advanced adenomas
+or carcinomas (reference for SRN?). Patients with an adenoma greater
+than 1 cm, more than three adenomas of any size, or an adenoma with
+villous histology were classified as having advanced adenomas. There
+were 172 patients with normal colonoscopies, 198 with adenomas, and 120
+with carcinomas. Of the 198 adenomas, 109 were identified as advanced
+adenomas. Together 261 patients were classified as normal and 229
+patients were classified as having a SRN.
 
 \textbf{16S rRNA gene sequencing data.} Stool samples provided by the
-patients was used for 16S rRNA gene sequencing to measure bacterial
+patients were used for 16S rRNA gene sequencing to measure bacterial
 population abundances. The sequence data used in our analyses were
@@
 values for SVMs. The code for these changes on L2-regularized SVM with
-linear kernel and L1-regularized SVM with linear kernel are available at
+linear kernel and L1-regularized SVM with linear kernel models are
+available at
 \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
@@
 For hyperparameter selection, we started with a granular grid search.
-Then we narrowed and fine-tuned the range of each hyperparameter. The
-range of the grid changes with different ML tasks and ML models. A full
+Then we narrowed and fine-tuned the range of each hyperparameter. A full
 grid search was needed to avoid large variation in prediction
@@
 \textbf{learning rate} and the \textbf{fraction of samples} to be used
-for fitting the individual base learners. There are newly developed
-tools such as Hyperband that help researcher with hyperparameter
-selection (40) that can be incorporated to microbiome studies.
+for fitting the individual base learners. Recently developed tools such
+as Hyperband help researchers with hyperparameter selection that can be
+incorporated to microbiome studies (40) .
 
@@
 was reduced by parallelizing segments of the ML pipeline. We
-parallelized the training of each data-split. This allowed 100
-data-splits to be processed through the ML pipeline at the same time for
-each model. It is possible to further parallelize the cross-validation
-step for each hyperparameter setting if limited by computational
-resources.
-
-\textbf{Permutation importance workflow.} We created a Spearman's
+parallelized the training of each data-split. This allowed the 100
+data-splits to be processed through the ML pipeline simultaneously at
+the same time for each model. It is possible to further parallelize the
+cross-validation step for each hyperparameter setting if limited by
+computational resources.
+
+\textbf{Permutation importance workflow.} We calculated a Spearman's
 rank-order correlation matrix and defined correlated OTUs as having
@@
 on their median ranks (median shown in black) are reported here. OTUs
-that are associated with classifying a subject as being healthy or
-having an SRN are in blue and red, respectively. Some of the same OTUs
-were identified as important in all of the linear models. \newpage
+that are associated with classifying a subject as being healthy had
+negative signs and were shown in blue. OTUs that are associated with
+classifying a subject having an SRN had positive signs and were shown in
+red. Some of the same OTUs were identified as important in all of the
+linear models. \newpage
 \includegraphics{Figure_4.png}
@@
 
-\captionof{table}{Characteristics of the machine learning models in our comparative study.}
+\captionsetup{labelformat=empty}
+\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
 \small
@@
 \rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} & \textbf{Refs.} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable & $^{36}$\\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable & $^{37}$\\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ & $^{38}$\\ \hline
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} \\ \hline
+
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable \\ \hline
+
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable \\ \hline
+
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ \\ \hline
 Decision tree & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels} & Non-linear & Interpretable & $^{39}$\\ \hline
-
-Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ & $^{40-41}$  \\ \hline
-
-XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ & $^{42-43}$ \\ \hline
+root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear & Interpretable \\ \hline
+
+Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ \\ \hline
+
+XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ \\ \hline
 

### 2019-10-03 submission/manuscript.Rmd
jennas edits

@@
 
-2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109
+2\. Department of Electrical Engineering and Computer Science, University or Michigan, Ann Arbor, MI 48109
 
@@
 Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer.
-Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training methods call the validity of these models into question.
+Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question.
 Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability.
@@
 We developed a generalizable pipeline to train, validate and interpret the models.
-To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost).
-The random forest model performed best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret.
-Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, trained much faster (12 min), and was more interpretable.
-Our analysis showed that ML models should be chosen based on the goal of the study as the choice will inform expectations of performance and interpretability.
+To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and extreme gradient boosted trees (XGBoost).
+The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and difficult to interpret.
+Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was more interpretable.
+Our analysis showed that ML models should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.
 
@@
 As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases.
-However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
+However, the structure of the human microbiome is remarkably variable AMONG individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
 This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals.
@@
 
-ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data.
+ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data.
 Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
-The task of diagnosing an individual with high confidence relies on a model that is built with rigorous methods.
-However, there are common methodological problems across many of these studies that need to be addressed as the field progresses.
-These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances.
+The task of diagnosing an individual relies on a rigorously validated model.
+However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress.
+These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances.
 Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
-More work is needed to further improve reproducibility and minimize over-optimism for model performance.
+More work is needed to further improve reproducibility and minimize overestimating for model performance.
 
-Among microbiome researchers, the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex.
-This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
-Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable.
-These models require post hoc explanations to quantify the importance of each feature in making a prediction and they do not show the structure of how the features are used.
-Depending on the application of the model, researchers may choose to use different modeling approaches.
-For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize predictive performance.
-Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018].
-Regardless, it is important for researchers to articulate why they have selected a specific modelling approach or even compare multiple approaches in the same study.
+Among microbiome researchers, the lack of justification when selecting a modelling approach has been due to an implicit assumption that more complex models are better.
+This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
+Although complex models may in some cases, capture important non-linear relationships and therefore yield better predictions, they may result in black boxes that lack interpretibility.
+Such models require post hoc explanations to quantify the importance of each feature in making predictions.
+Depending on the goal of the modeling, researchers may choose to use different approaches.
+For example, researchers trying to identify the populations of microbiota associated wwith disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance.
+Nonetheless, it is important to understand the tradeoff and accepting that this tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018].
+It is important for researchers to justify their choice of modelling approach.
 
-To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches with the same dataset and pipeline.
+To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline.
 We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
-We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
-We compared their predictive performance, interpretability, and computational efficiency.
+We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
+We compared their predictive performance, interpretability, and training time.
 To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
 This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
-With this dataset we developed a framework that implements a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models.
+With this dataset we developed a ML pipeline that can be used in many different scenarios for training and evaluating models.
 This framework can be easily applied to other host-associated and environmental microbiome datasets.
@@
 
-__Model selection and pipeline construction__ We established a common ML pipeline to train and validate each of the seven models that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].
+__Model selection and pipeline construction__ We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].
 
-First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining 20% of the data [Figure 1].
-To maintain the distribution of controls and cases that was found with the full dataset, we performed stratified splits.
+First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1].
+To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits.
 For example, our full dataset included 490 individuals.
@@
 A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%).
-The training data was used to build the models and the test set was used for evaluating predictive performance.
+The training data was used to build and select the models and the test set was used for evaluating the model.
 
-Second, we trained seven different models using the training data [Table 1].
-We selected models with different classification algorithms and regularization methods.
-Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well.
-For regularized logistic regression and SVM with linear kernel, we used L2-regularization to keep all potentially important features.
-For comparison, we also trained an L1-regularized SVM model with linear kernel.
+We trained seven different models using the training data [Table 1].
+We focused on different classification algorithms and regularization methods.
+Regularization helps to prevent overfitting by penalizing a model that fits the training data too well.
+For regularized logistic regression and SVM with a linear kernel, we used L2-regularization to keep all potentially important features.
+For comparison, we also trained an L1-regularized SVM with a linear kernel.
 L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero).
-To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with a non-linear kernel.
+To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including a decision tree, a random forest, and XGBoost and an SVM with a non-linear kernel.
 
-Third, fitting of these models required selecting appropriate hyperparameters.
-Hyperparameters are the rules that are learned from the training set in a classification algorithm.
-For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting.
-Similar to regularization term C, all hyperparameters are tuned to find the best model.
-We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1].
+Model selection required tuning hyperparameters.
+Hyperparameters are parameters which need to be specified or tuned by the user, in order to train a model for a specific modeling problem.
+For example, when using regularization, (C) is a hyperparameter that indicates the penalty for overfitting.
+Hyperparameters are tuned using the training data to find the best model.
+We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1].
 The five-fold CV was also stratified to maintain the overall case and control distribution.
-We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2].
-The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance.
-To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models.
-Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
-In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.
+We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2].
+The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0, where the model’s predictions are perfectly incorrect.
+An AUROC value of 0.5 indicates that model's predictions are no differect than random.
+To select hyperparameters, we performed a full grid search for hyperparameter settings when training our models.
+Default hyperparameter settings in previously developed ML packages in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
+In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization strengths (C).
 
-Finally, we trained the full training dataset with the selected hyperparameter values and applied the model to the held-out data to evaluate the testing predictive performance of each model.
+Once hyperparameters were selected, we trained the model using the full training dataset and applied the final model to the held-out data to evaluate the testing predictive performance of each model.
 The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a reliable and robust interpretation of model performance [Figure 1].
@@
 L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
-Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well as random forest.
+Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well non-linear models.
 
 To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC.
-If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit.
+If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data.
 The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
-These differences are relatively small and would not indicate a problem with overfitting.
+These differences are relatively small and gives us confidence in our estimate of the generalization performance of the models.
 
-To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 data-splits.
+To evaluate the variation in the estimated performance, we calculated the range of AUROC values for each model using 100 data-splits.
 The range among the testing AUROC values within each model varied by `r paste(format(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3), nsmall=3))` on average across the seven models.
-If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model.
+If we had only done a single split, then there is a risk that we could have gotten lucky or unlucky in estimating model performance.
 For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(format(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3), nsmall=3))`.
-These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2].
-Therefore, it is important to employ the 100 hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.
+These results showed that depending on the data-split, the testing performance can vary [Figure 2].
+Therefore, it is important to employ multiple data splits when estimating generalization performance.
 
 To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3].
-The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models.
-These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism.
-Furthermore, it was encouraging that even for a small number of samples, the interquartile range included the median AUROC values for the larger subsetted datasets.
+The variation in cross-validation performance within both models at lower sample sizes was larger than when the full collection of samples was used to train and validate the models.
+Because of the high dimensionality of the microbiome data (6920 OTUs), large sample sizes can lead to better models.
 
-__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017].
+__Interpretation of each ML model.__ Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019].
 Because we often use ML models not just to predict a health outcome but also to identify the biomarkers for a disease, model interpretation becomes crucial for microbiome studies.
@@
 Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
-Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data.
-We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
+Permutation importance analysis is a posthoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data.
+We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
 We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4].
@@
 Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477).
-Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance.
+Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance.
 For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`.
 In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.
-Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019].
-At least in this case, it was not possible to distinguish between health and disease using a single OTU.
+In this case, similar to previous studies [@sze_leveraging_2018; @wirbel_meta-analysis_2019], it was not possible to distinguish between health and disease using a single OTU.
 Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e.
@@
 __The computational efficiency of each ML model.__ We compared the training times of the seven ML models.
-As expected, the training times increased with the complexity of the model and the number of tuned hyperparameter settings.
+As expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations.
 Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
-When we subsetted the size of the training dataset, we observed a linear relationship between the size of the dataset and the training times for L2-regularized logistic regression and random forest models [Figure S5].
 
@@
 Traditional statistical approaches are useful for identifying those cases where a single organism is associated with a process.
-In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole to classify them into different categories such as coming from a patient who is healthy or has SRNs.
+In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole and identify associations between community structure and disease state.
 If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification.
@@
 
-To address these needs, we developed a framework to train rigorous, transparent, and reproducible models.
-We benchmarked seven ML models and showed that we can create models that are inherently interpretable and easily trained without losing predictive performance.
-In terms of predictive performance, the random forest model had the best testing AUROC values compared to the other six models.
+To address these needs, we developed an open-sourced framework to ML models.
+using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized.
+In terms of predictive performance, the random forest model had the best AUROC compared to the other six models.
 However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
-While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a complex ML model and could only be explained using post-hoc methods such as permutation importance.
-On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights.
+While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes.
+In terms of interpretability, random forest is a non-linear ML model and is most often explained using post-hoc methods such as permutation importance.
+On the other hand, using L2-regularized logistic regression we ranked the importance of each OTU based on their feature weights.
 Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.
 
-As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome.
+As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that predicts a binary health outcome.
 We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance.
@@
 During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation.
-Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the similar AUROC values from the cross-validation and testing.
+Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the good test AUROCs.
 
@@
 This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013].
-For example if one OTU is highly correlated with another OTU, only one of these OTUs would have a large feature weight thus hiding the importance of its correlated OTU.
 To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs.
@@
 When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group.
-We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.
+We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among correlated features.
 It is important to know the correlation structures of the data to avoid misinterpreting the models.
@@
 We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs.
-Furthermore, as number of categories to classify increase, more samples are required for each category to train a model.
+Furthermore, as number of classes increases, more samples are required for each category to train an accurate model.
 We also did not use regression-based analyses to predict a non-categorical outcome.
@@
 Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks.
-Deep learning methods hold great promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which makes the deep learning models prone to overfitting.
-These methods are even more complex than random forest and XGBoost and are considered uninterpretable.
-There is great potential for applying ML approaches to microbiome data and we can choose which ML approach is best for the study.
+Deep learning methods hold promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which can result in overfitting.
 
-Our framework gives structure to investigators wanting to train, evaluate, and interpret their own ML models to identify OTUs that might be biologically relevant.
+Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant.
 However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly harder and distinct undertaking.
@@
 We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1].
-We have highlighted the trade-offs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
+We have highlighted the trade-offs between model complexity and interpretability, the need for tuning hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
 Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.
@@
 \newpage
-\includegraphics{Figure_1}
+\includegraphics{Figure_1.pdf}
 **Figure 1.
@@
 
-**Figure 5. Computational efficiency of seven ML models.** The median training time was the highest for XGBoost and shortest for L2-regularized logistic regression. 
+**Figure 5. Training times of seven ML models.** The median training time was the highest for XGBoost and shortest for L2-regularized logistic regression. 
 \newpage
@@
 Hyperparameter setting performances for linear models.** (A) L2 logistic regression, (B) L1 SVM with linear kernel, and (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model.
-The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
+The stars represent the highest performing hyperparameter setting for each model.
 
@@
 Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) Random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.
-The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
+The stars represent the highest performing hyperparameter setting for the models.
 \newpage
@@
 **Figure S3.
-Classification performance of ML models across cross validation when dataset is subsetted .** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
+Classification performance of ML models across cross validation when trained on a subset of th dataset.** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
 The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
@@
 \small
-\begin{tabular}{|l|l|l|l|l|}
+\begin{tabular}{|l|l|l|l|}
 \hline
@@
 \rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} \\ \hline
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
 
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable \\ \hline
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
 
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable \\ \hline
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
 
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ \\ \hline
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
 Decision tree & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear & Interpretable \\ \hline
+root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
 
-Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ \\ \hline
+Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
 
-XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ \\ \hline
+XGBoost & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
 
@@
 \\ How much interpretability \\ does the problem need?
-} & \makecell[l]{Do we know our data?
-\\ Do we know the correlated \\ variables?
+} & \makecell[l]{Do we know the correlated \\ features?
 } \\ \hline
 
-\makecell[l]{Model \\ selection} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem?
+\makecell[l]{Classification \\ algorithm} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem?
 \\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretability \\ does the problem need?
@@
 
-\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Did we do \\ cross-validation} & \makecell[l]{Do we have an held-out \\ test dataset?
-\\ Have we tuned our model \\ hyperparameters in \\ cross-validation?} & \makecell[l]{Have we tested our model \\ on many different \\  datasets?} \\ \hline
+\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Did we do \\ cross-validation?} & \makecell[l]{Do we have a held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\  datasets?} \\ \hline
 
@@
 \makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it?
-\\ Have we checked for \\ the effect of correlated \\ variables?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
+\\ Have we checked for \\ the effect of correlated \\ features?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
 } \\ \hline

### 2019-10-03 submission/manuscript.tex
jennas edits

@@
 
-2. Department of Computer Science and Engineering, University or
-Michigan, Ann Arbor, MI 49109
+2. Department of Electrical Engineering and Computer Science, University
+or Michigan, Ann Arbor, MI 48109
 
@@
 outcomes using bacterial abundances, but inconsistent adoption of
-training methods call the validity of these models into question.
-Furthermore, there appears to be a preference by many researchers to
-favor increased model complexity over interpretability. To overcome
-these challenges, we trained seven models that used fecal 16S rRNA
-sequence data to predict the presence of colonic screen relevant
-neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We
-developed a generalizable pipeline to train, validate and interpret the
-models. To show the effect of model selection, we assessed the
-predictive performance, interpretability, and computational efficiency
-of L2-regularized logistic regression, L1 and L2-regularized support
-vector machines (SVM) with linear and radial basis function kernels, a
-decision tree, random forest, and extreme gradient boosting (XGBoost).
-The random forest model performed best at detecting SRNs with an AUROC
-of 0.695 but it was slow to train (83.2 h) and hard to interpret.
-Despite its simplicity, L2-regularized logistic regression followed
-random forest in predictive performance with an AUROC of 0.680, trained
-much faster (12 min), and was more interpretable. Our analysis showed
-that ML models should be chosen based on the goal of the study as the
-choice will inform expectations of performance and interpretability.
+training and evaluation methods call the validity of these models into
+question. Furthermore, there appears to be a preference by many
+researchers to favor increased model complexity over interpretability.
+To overcome these challenges, we trained seven models that used fecal
+16S rRNA sequence data to predict the presence of colonic screen
+relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
+We developed a generalizable pipeline to train, validate and interpret
+the models. To show the effect of model selection, we assessed the
+predictive performance, interpretability, and training time of
+L2-regularized logistic regression, L1 and L2-regularized support vector
+machines (SVM) with linear and radial basis function kernels, decision
+trees, random forest, and extreme gradient boosted trees (XGBoost). The
+random forest model performed best at detecting SRNs with an AUROC of
+0.695 {[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and difficult
+to interpret. Despite its simplicity, L2-regularized logistic regression
+followed random forest in predictive performance with an AUROC of 0.680
+{[}IQR 0.625-0.735{]}, trained faster (12 min), and was more
+interpretable. Our analysis showed that ML models should be chosen based
+on the goal of the study, as the choice will inform expectations of
+performance and interpretability.
 
@@
 diseases. However, the structure of the human microbiome is remarkably
-variable between individuals to the point where it is often difficult to
+variable AMONG individuals to the point where it is often difficult to
 identify the bacterial populations that are associated with diseases
@@
 
-ML models are useful for understanding the variation in the structure of
-existing data and to apply that knowledge to make predictions about new
-data. Researchers have used ML models to diagnose and understand the
+ML models can be used to increase our understanding of the variation in
+the structure of existing data and in making predictions about new data.
+Researchers have used ML models to diagnose and understand the
 ecological basis of diseases such as liver cirrhosis, colorectal cancer,
 inflammatory bowel diseases, obesity, and type 2 diabetes (1--16,
-16--18). The task of diagnosing an individual with high confidence
-relies on a model that is built with rigorous methods. However, there
-are common methodological problems across many of these studies that
-need to be addressed as the field progresses. These include a lack of
+16--18). The task of diagnosing an individual relies on a rigorously
+validated model. However, there are common methodological and reporting
+problems that arise when applying ML to such data, that need to be
+addressed for the field to progress. These problems include a lack of
 transparency in which methods are used and how these methods are
-implemented; developing and evaluating models without a separate
-held-out test data; large variation between the predictive performance
-on different folds of cross-validation; and large variation between
-cross-validation and testing performances. Nevertheless, the microbiome
-field is making progress to avoid some of these pitfalls including
-validating their models on independent datasets (7, 18, 19) and
-introducing ways to better use ML tools (20--23). More work is needed to
-further improve reproducibility and minimize over-optimism for model
-performance.
-
-Among microbiome researchers, the lack of transparency in justifying a
+implemented; evaluating models without a separate held-out test data;
+large variation between the predictive performance on different folds of
+cross-validation; and large variation between cross-validation and
+testing performances. Nevertheless, the microbiome field is making
+progress to avoid some of these pitfalls including validating their
+models on independent datasets (7, 18, 19) and introducing ways to
+better use ML tools (20--23). More work is needed to further improve
+reproducibility and minimize overestimating for model performance.
+
+Among microbiome researchers, the lack of justification when selecting a
 modelling approach has been due to an implicit assumption that more
-complex models are better because they are more complex. This has
-resulted in a trend towards using models such as random forest and
-neural networks (2, 11, 24--26) over simpler models such as logistic
-regression or other linear models (18, 22, 27). Although the more
-complex models may be better at incorporating non-linear relationships
-or yield better predictions, they are considered to be black box models
-because they are not inherently interpretable. These models require post
-hoc explanations to quantify the importance of each feature in making a
-prediction and they do not show the structure of how the features are
-used. Depending on the application of the model, researchers may choose
-to use different modeling approaches. For example, researchers trying to
-identify the populations causing a disease would likely want a more
-interpretable model whereas clinicians may emphasize predictive
-performance. Although one may feel that they are sacrificing
-interpretability for performance, that tradeoff may be minimal (28, 29).
-Regardless, it is important for researchers to articulate why they have
-selected a specific modelling approach or even compare multiple
-approaches in the same study.
+complex models are better. This has resulted in a trend towards using
+models such as random forest and deep neural networks (2, 11, 24--26)
+over simpler models such as logistic regression or other linear models
+(18, 22, 27). Although complex models may in some cases, capture
+important non-linear relationships and therefore yield better
+predictions, they may result in black boxes that lack interpretibility.
+Such models require post hoc explanations to quantify the importance of
+each feature in making predictions. Depending on the goal of the
+modeling, researchers may choose to use different approaches. For
+example, researchers trying to identify the populations of microbiota
+associated wwith disease may desire a more interpretable model, whereas
+clinicians may emphasize predictive performance. Nonetheless, it is
+important to understand the tradeoff and accepting that this tradeoff
+may be minimal (28, 29). It is important for researchers to justify
+their choice of modelling approach.
 
@@
 selection can affect modeling results, we performed an empirical
-analysis comparing 7 modeling approaches with the same dataset and
+analysis comparing seven modeling approaches with the same dataset and
 pipeline. We built three linear models with different forms of
@@
 L2-regularized support vector machines (SVM) with a linear kernel. We
-also built four non-linear models: SVM with radial basis function
+also trained four non-linear models: SVM with radial basis function
 kernel, a decision tree, random forest and XGBoost. We compared their
-predictive performance, interpretability, and computational efficiency.
-To demonstrate the performance of these modeling approaches and our
+predictive performance, interpretability, and training time. To
+demonstrate the performance of these modeling approaches and our
 pipeline, we used data from a previously published study that sought to
@@
 there is ample evidence that the disease is driven by variation in the
-microbiome (1, 3, 4, 30). With this dataset we developed a framework
-that implements a ML pipeline that can be used for any modeling
-approach, evaluates predictive performance, and demonstrates how to
-interpret these models. This framework can be easily applied to other
-host-associated and environmental microbiome datasets.
+microbiome (1, 3, 4, 30). With this dataset we developed a ML pipeline
+that can be used in many different scenarios for training and evaluating
+models. This framework can be easily applied to other host-associated
+and environmental microbiome datasets.
 
@@
 \textbf{Model selection and pipeline construction} We established a
-common ML pipeline to train and validate each of the seven models that
-is based on standard methods within the ML community (REFS by
-Jenna?){[}Figure 1{]}.
+reusable ML pipeline for model selection and evaluation, focusing on
+seven different commonly used supervised learning algorithms {[}Figure
+1{]}.
 
 First, we randomly split the data into training and test sets so that
-the training set consisted of 80\% of the full dataset while the test
-set was composed of the remaining 20\% of the data {[}Figure 1{]}. To
-maintain the distribution of controls and cases that was found with the
-full dataset, we performed stratified splits. For example, our full
-dataset included 490 individuals. Of these, 261 had normal colons (53\%)
-and 229 had a screen relevant neoplasia (SRN; 46.7\%). A training set
-included 393 individuals, of which 209 had an SRN (53\%), while the test
-set was composed of 97 individuals of which 52 had an SRN (54\%). The
-training data was used to build the models and the test set was used for
-evaluating predictive performance.
-
-Second, we trained seven different models using the training data
-{[}Table 1{]}. We selected models with different classification
-algorithms and regularization methods. Regularization is a technique
-that discourages overfitting by penalizing the model for learning the
-training data too well. For regularized logistic regression and SVM with
-linear kernel, we used L2-regularization to keep all potentially
-important features. For comparison, we also trained an L1-regularized
-SVM model with linear kernel. L1-regularization on microbiome data led
-to a sparser solution (i.e., forced many coefficients to zero). To
-explore the potential for non-linear relationships among features to
-improve classification, we trained tree-based models including decision
-tree, random forest, and XGBoost and we trained an SVM model with a
-non-linear kernel.
-
-Third, fitting of these models required selecting appropriate
-hyperparameters. Hyperparameters are the rules that are learned from the
-training set in a classification algorithm. For example, in the linear
-models the regularization term (C) is a hyperparameter that indicates
-the penalty for overfitting. Similar to regularization term C, all
-hyperparameters are tuned to find the best model. We selected
-hyperparameters by performing 100 five-fold cross-validation (CV)
-repeats on the training set {[}Figure 1{]}. The five-fold CV was also
-stratified to maintain the overall case and control distribution. We
-chose the best hyperparameter values for each model based on its CV
-predictive performance using the area under the receiver operating
-characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
+the training set consisted of 80\% of the full dataset, while the test
+set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
+distribution of controls and cases found in the full dataset, we
+performed stratified splits. For example, our full dataset included 490
+individuals. Of these, 261 had normal colons (53\%) and 229 had a screen
+relevant neoplasia (SRN; 46.7\%). A training set included 393
+individuals, of which 209 had an SRN (53\%), while the test set was
+composed of 97 individuals of which 52 had an SRN (54\%). The training
+data was used to build and select the models and the test set was used
+for evaluating the model.
+
+We trained seven different models using the training data {[}Table 1{]}.
+We focused on different classification algorithms and regularization
+methods. Regularization helps to prevent overfitting by penalizing a
+model that fits the training data too well. For regularized logistic
+regression and SVM with a linear kernel, we used L2-regularization to
+keep all potentially important features. For comparison, we also trained
+an L1-regularized SVM with a linear kernel. L1-regularization on
+microbiome data led to a sparser solution (i.e., forced many
+coefficients to zero). To explore the potential for non-linear
+relationships among features to improve classification, we trained
+tree-based models including a decision tree, a random forest, and
+XGBoost and an SVM with a non-linear kernel.
+
+Model selection required tuning hyperparameters. Hyperparameters are
+parameters which need to be specified or tuned by the user, in order to
+train a model for a specific modeling problem. For example, when using
+regularization, (C) is a hyperparameter that indicates the penalty for
+overfitting. Hyperparameters are tuned using the training data to find
+the best model. We selected hyperparameters by performing repeated
+five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
+five-fold CV was also stratified to maintain the overall case and
+control distribution. We chose the hyperparameter values that led to the
+best average CV predictive performance using the area under the receiver
+operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
 ranges from 1.0, where the model perfectly distinguishes between cases
-and controls, to 0.50, where the model's predictions are no different
-from random chance. To select the best performing hyperparameter, we
-performed a full grid search for hyperparameter settings when training
-our models. Default hyperparameter settings in previously developed ML
-packages in R, Python, and Matlab programming languages are inadequate
-for effective application of classification algorithms and need to be
+and controls, to 0, where the model's predictions are perfectly
+incorrect. An AUROC value of 0.5 indicates that model's predictions are
+no differect than random. To select hyperparameters, we performed a full
+grid search for hyperparameter settings when training our models.
+Default hyperparameter settings in previously developed ML packages in
+R, Python, and MATLAB programming languages may be inadequate for
+effective application of classification algorithms and need to be
 optimized for each new ML task. In the example of L1-regularized SVM
 with linear kernel {[}Figure S1{]}, the model showed large variability
-between different regularization coefficients (C) and was susceptible to
-performing poorly if the wrong regularization coefficient was assigned
-to the model by default.
-
-Finally, we trained the full training dataset with the selected
-hyperparameter values and applied the model to the held-out data to
+between different regularization strengths (C).
+
+Once hyperparameters were selected, we trained the model using the full
+training dataset and applied the final model to the held-out data to
 evaluate the testing predictive performance of each model. The
@@
 most interpretable models (L2-regularized logistic regression and SVM
-with linear kernel) performed nearly as well as random forest.
+with linear kernel) performed nearly as well non-linear models.
 
@@
 between the cross-validation and testing AUROCs was large, then that
-would indicate that the models were overfit. The difference in median
-AUROCs was 0.021 in L1-regularized SVM with linear kernel, followed by
-SVM with radial basis function kernel and decision tree with a
-difference of 0.007 and 0.006, respectively {[}Figure 2{]}. These
-differences are relatively small and would not indicate a problem with
-overfitting.
-
-To evaluate the risk for over-optimism of each model, we calculated the
-range of AUROC values for each model using 100 data-splits. The range
-among the testing AUROC values within each model varied by 0.230 on
-average across the seven models. If we had only done a single split,
-then there is a risk that we could gotten lucky or unlucky with the
-performance of the model. For instance, the lowest AUROC value of the
-random forest model was 0.593 whereas the highest was 0.810. These
-results showed that depending on the data-split, the testing AUROC
-values showed great variability {[}Figure 2{]}. Therefore, it is
-important to employ the 100 hierarchical data splits that were included
-in our pipeline to minimize the risk of over-optimism.
+could indicate that the models were overfit to the training data. The
+difference in median AUROCs was 0.021 in L1-regularized SVM with linear
+kernel, followed by SVM with radial basis function kernel and decision
+tree with a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
+These differences are relatively small and gives us confidence in our
+estimate of the generalization performance of the models.
+
+To evaluate the variation in the estimated performance, we calculated
+the range of AUROC values for each model using 100 data-splits. The
+range among the testing AUROC values within each model varied by 0.230
+on average across the seven models. If we had only done a single split,
+then there is a risk that we could have gotten lucky or unlucky in
+estimating model performance. For instance, the lowest AUROC value of
+the random forest model was 0.593 whereas the highest was 0.810. These
+results showed that depending on the data-split, the testing performance
+can vary {[}Figure 2{]}. Therefore, it is important to employ multiple
+data splits when estimating generalization performance.
 
@@
 490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S3{]}. The
-range among the cross-validation AUROC values within both models at
-lower sample sizes were much larger than when the full collection of
-samples was used to train and validate the models. These results showed
-that because the microbiome data had many features (6920 OTUs), it was
-important to train the models using appropriate sample sizes to avoid
-problems with generalizability and over-optimism. Furthermore, it was
-encouraging that even for a small number of samples, the interquartile
-range included the median AUROC values for the larger subsetted
-datasets.
-
-\textbf{Interpretation of each ML model.} Interpretability is the degree
-to which humans can understand the reasons behind a model prediction
-(31). Because we often use ML models not just to predict a health
-outcome but also to identify the biomarkers for a disease, model
-interpretation becomes crucial for microbiome studies. ML models
+variation in cross-validation performance within both models at lower
+sample sizes was larger than when the full collection of samples was
+used to train and validate the models. Because of the high
+dimensionality of the microbiome data (6920 OTUs), large sample sizes
+can lead to better models.
+
+\textbf{Interpretation of each ML model.} Interpretability is related to
+the degree to which humans can understand the reasons behind a model
+prediction (31--33). Because we often use ML models not just to predict
+a health outcome but also to identify the biomarkers for a disease,
+model interpretation becomes crucial for microbiome studies. ML models
 decrease in interpretability as they increase in complexity. In this
@@
 importance using the held-out test data. Permutation importance analysis
-is a posthoc explanation of the model where we randomly permuted
-non-correlated features individually and groups of perfectly correlated
-features across the two groups in the held-out test data. We then
-calculated how much the predictive performance of the model (i.e testing
-AUROC values) decreased when each OTU or group of OTUs was randomly
-permuted. We ranked the OTUs based on how much the median testing AUROC
-decreased when it was permuted; the OTU with the largest decrease ranked
-highest {[}Figure 4{]}. Among the twenty OTUs with the largest impact,
-there was only one OTU (OTU 822) that was shared among all of the
-models; however, we found three OTUs (OTU 58, 110, 367) that were
-important in each of the tree-based models. Similarly, the random forest
-and XGBoost models, shared four of the most important OTUs (OTU 2, 12,
-361, 477). Permutation analysis results also revealed that with the
-exception of the decision tree model, removal of any individual OTU had
-a minimal impact on model performance. For example, if OTU 367 was
+is a posthoc explanation of the model, in which we randomly permuted
+groups of perfectly correlated features together and other features
+individually across the two groups in the held-out test data. We then
+calculated how much the predictive performance of the model (i.e,
+testing AUROC values) decreased when each OTU or group of OTUs was
+randomly permuted. We ranked the OTUs based on how much the median
+testing AUROC decreased when it was permuted; the OTU with the largest
+decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
+largest impact, there was only one OTU (OTU 822) that was shared among
+all of the models; however, we found three OTUs (OTU 58, 110, 367) that
+were important in each of the tree-based models. Similarly, the random
+forest and XGBoost models, shared four of the most important OTUs (OTU
+2, 12, 361, 477). Permutation analysis results also revealed that with
+the exception of the decision tree model, removal of any individual OTU
+had minimal impact on model performance. For example, if OTU 367 was
 permuted across the samples in the decision tree model, the median AUROC
@@
 in the random forest model, the AUROC only dropped from 0.695 to 0.680.
-Effectively, the complexity of the communities was more fully
-represented in the better performing models (22, 32). At least in this
-case, it was not possible to distinguish between health and disease
-using a single OTU. Although permutation analysis allowed us to gauge
-the importance of an OTU, the analysis was post-hoc (i.e.~done using the
-test data) and these results did not allow us to directly interrogate
-the models to know whether an OTU was associated with classifying a
-subject as being healthy or having an SRN.
+In this case, similar to previous studies (22, 34), it was not possible
+to distinguish between health and disease using a single OTU. Although
+permutation analysis allowed us to gauge the importance of an OTU, the
+analysis was post-hoc (i.e.~done using the test data) and these results
+did not allow us to directly interrogate the models to know whether an
+OTU was associated with classifying a subject as being healthy or having
+an SRN.
 
@@
 training times of the seven ML models. As expected, the training times
-increased with the complexity of the model and the number of tuned
-hyperparameter settings. Also, the linear models trained faster than
-non-linear models {[}Figures S1-S2; Figure 5{]}. When we subsetted the
-size of the training dataset, we observed a linear relationship between
-the size of the dataset and the training times for L2-regularized
-logistic regression and random forest models {[}Figure S5{]}.
+increased with the complexity of the model and the number of potential
+hyperparameter combinations. Also, the linear models trained faster than
+non-linear models {[}Figures S1-S2; Figure 5{]}.
 
@@
 with a process. In contrast, ML methods offer the ability to incorporate
-the structure of the microbial communities as a whole to classify them
-into different categories such as coming from a patient who is healthy
-or has SRNs. If it is possible to classify communities reliably, then ML
-methods also offer the ability to identify those microbial populations
-within the communities that are responsible for the classification.
-However, the application of ML in microbiome studies is still in its
-infancy and the field needs to develop a better understanding of
-different ML methods, their strengths and weaknesses, and how to
-implement them.
-
-To address these needs, we developed a framework to train rigorous,
-transparent, and reproducible models. We benchmarked seven ML models and
-showed that we can create models that are inherently interpretable and
-easily trained without losing predictive performance. In terms of
-predictive performance, the random forest model had the best testing
-AUROC values compared to the other six models. However, the second-best
-model was L2-regularized logistic regression with a median AUROC
-difference of only 0.015 compared to random forest. While random forest
-took 83.2 hours to train, L2-regularized logistic regression trained in
-12 minutes. In terms of interpretability, random forest is a complex ML
-model and could only be explained using post-hoc methods such as
-permutation importance. On the other hand, L2-regularized logistic
-regression was easier to interpret by ranking the OTUs based on their
-feature weights and reviewing the signs of these weights. Comparing many
-different models showed us that the most complex model was not
-necessarily the best model for our ML task.
+the structure of the microbial communities as a whole and identify
+associations between community structure and disease state. If it is
+possible to classify communities reliably, then ML methods also offer
+the ability to identify those microbial populations within the
+communities that are responsible for the classification. However, the
+application of ML in microbiome studies is still in its infancy and the
+field needs to develop a better understanding of different ML methods,
+their strengths and weaknesses, and how to implement them.
+
+To address these needs, we developed an open-sourced framework to ML
+models. using this pipeline, we benchmarked seven ML models and showed
+that the tradeoff between model complexity and performance may be less
+severe than originally hypothesized. In terms of predictive performance,
+the random forest model had the best AUROC compared to the other six
+models. However, the second-best model was L2-regularized logistic
+regression with a median AUROC difference of only 0.015 compared to
+random forest. While our implementation of random forest took 83.2 hours
+to train, our L2-regularized logistic regression trained in 12 minutes.
+In terms of interpretability, random forest is a non-linear ML model and
+is most often explained using post-hoc methods such as permutation
+importance. On the other hand, using L2-regularized logistic regression
+we ranked the importance of each OTU based on their feature weights.
+Comparing many different models showed us that the most complex model
+was not necessarily the best model for our ML task.
 
 As we set out to select the best model, we established a pipeline that
-can be generalized to any modeling method that uses 16S rRNA sequence
-counts to predict a binary health outcome. We performed a random
-data-split to create a training set (80\% of the data) and a held-out
-test set (20\% of the data), which we used to evaluate predictive
-performance. We repeated this data-split 100 times to measure the
-possible variation in predictive performance. During the training, we
-tuned the model hyperparameters with a repeated five-fold
-cross-validation. Despite the high number of features microbiome
-datasets typically have, the models we built with this pipeline were
-generalizable as shown by the similar AUROC values from the
-cross-validation and testing.
+can be generalized to any modeling method that predicts a binary health
+outcome. We performed a random data-split to create a training set (80\%
+of the data) and a held-out test set (20\% of the data), which we used
+to evaluate predictive performance. We repeated this data-split 100
+times to measure the possible variation in predictive performance.
+During the training, we tuned the model hyperparameters with a repeated
+five-fold cross-validation. Despite the high number of features
+microbiome datasets typically have, the models we built with this
+pipeline were generalizable as shown by the good test AUROCs.
 
@@
 our ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (33). For example if one
-OTU is highly correlated with another OTU, only one of these OTUs would
-have a large feature weight thus hiding the importance of its correlated
-OTU. To capture all important features, once we identify highly ranked
-OTUs, we should review their relationships with other OTUs. These
-relationships will help us generate new hypotheses about the ecology of
-the disease and test them with follow-up experiments. When we used
-permutation importance, we took collinearity into consideration by
-grouping correlated OTUs to determine their impact as a group. We
-grouped OTUs that had a perfect correlation with each other however, we
-can reduce the correlation threshold to further investigate the
-relationships among features. It is important to know the correlation
-structures of the data to avoid misinterpreting the models. This is
-likely to be a particular problem with shotgun metagenomic datasets
-where collinearity will be more pronounced due to many genes being
-correlated with one another because they come from the same chromosome.
-To identify the true underlying microbial factors of a disease, it is
-crucial to do correlation analyses and further experimentation for
-biological validation.
+correlated OTUs are influenced by one another (35). To capture all
+important features, once we identify highly ranked OTUs, we should
+review their relationships with other OTUs. These relationships will
+help us generate new hypotheses about the ecology of the disease and
+test them with follow-up experiments. When we used permutation
+importance, we took collinearity into consideration by grouping
+correlated OTUs to determine their impact as a group. We grouped OTUs
+that had a perfect correlation with each other however, we can reduce
+the correlation threshold to further investigate the relationships among
+correlated features. It is important to know the correlation structures
+of the data to avoid misinterpreting the models. This is likely to be a
+particular problem with shotgun metagenomic datasets where collinearity
+will be more pronounced due to many genes being correlated with one
+another because they come from the same chromosome. To identify the true
+underlying microbial factors of a disease, it is crucial to do
+correlation analyses and further experimentation for biological
+validation.
 
@@
 grouping was between patients with normal colons and those with SRNs.
-Furthermore, as number of categories to classify increase, more samples
-are required for each category to train a model. We also did not use
+Furthermore, as number of classes increases, more samples are required
+for each category to train an accurate model. We also did not use
 regression-based analyses to predict a non-categorical outcome. We have
@@
 predict fecal short-chain fatty acid concentrations based on microbiome
-data (34). Our analysis was also limited to shallow learning methods and
+data (36). Our analysis was also limited to shallow learning methods and
 did not explore deep learning methods such as neural networks. Deep
-learning methods hold great promise (11, 35, 36) but microbiome datasets
-often suffer from having many features and small sample sizes, which
-makes the deep learning models prone to overfitting. These methods are
-even more complex than random forest and XGBoost and are considered
-uninterpretable. There is great potential for applying ML approaches to
-microbiome data and we can choose which ML approach is best for the
-study.
-
-Our framework gives structure to investigators wanting to train,
-evaluate, and interpret their own ML models to identify OTUs that might
-be biologically relevant. However, deploying microbiome-based models to
-make clinical diagnoses or predictions is a significantly harder and
-distinct undertaking. For example, we currently lack standardized
-methods to collect patient samples, generate sequence data, and report
-clinical data. We are also challenged by the practical constraints of
-OTU-based approaches. The de novo algorithms commonly in use are slow,
-require considerable memory, and result in different OTU assignments as
-new data are added. Finally, we also need independent validation cohorts
-to test the performance of a diagnostic model. To realize the potential
-for using ML approaches with microbiome data, it is necessary that we
-direct our efforts to overcome these challenges.
+learning methods hold promise (11, 37, 38) but microbiome datasets often
+suffer from having many features and small sample sizes, which can
+result in overfitting.
+
+Our framework provides a reproducible structure to investigators wanting
+to train, evaluate, and interpret their own ML models to generate
+hypotheses regarding which OTUs might be biologically relevant. However,
+deploying microbiome-based models to make clinical diagnoses or
+predictions is a significantly harder and distinct undertaking. For
+example, we currently lack standardized methods to collect patient
+samples, generate sequence data, and report clinical data. We are also
+challenged by the practical constraints of OTU-based approaches. The de
+novo algorithms commonly in use are slow, require considerable memory,
+and result in different OTU assignments as new data are added. Finally,
+we also need independent validation cohorts to test the performance of a
+diagnostic model. To realize the potential for using ML approaches with
+microbiome data, it is necessary that we direct our efforts to overcome
+these challenges.
 
@@
 have highlighted the trade-offs between model complexity and
-interpretability, the need for cross-validation to tune hyperparameters,
-the utility of held-out test sets for evaluating predictive performance,
-and the importance of considering correlation structures in datasets for
+interpretability, the need for tuning hyperparameters, the utility of
+held-out test sets for evaluating predictive performance, and the
+importance of considering correlation structures in datasets for
 reliable interpretation. Furthermore, we underscored the importance of
@@
 Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
-generated by Sze et al (32), who processed the 16S rRNA sequences in
+generated by Sze et al (34), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (37--39);
+to OTUs at 97\% similarity using the OptiClust algorithm (39--41);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 as Hyperband help researchers with hyperparameter selection that can be
-incorporated to microbiome studies (40) .
+incorporated to microbiome studies (42) .
 
@@
 
-\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline.
-} We split the data to create a training (80\%) and held-out test set
-(20\%). The splits were stratified to maintain the overall label
-distribution. We performed five-fold cross-validation on the training
-data to select the best hyperparameter setting and then used these
-hyperparameters to train the models. The model was evaluated on the
-held-out data set. Abbreviations: cvAUC, cross-validation area under the
-receiver operating characteristic curve. \newpage
+\includegraphics{Figure_1.pdf} \textbf{Figure 1. Machine learning
+pipeline. } We split the data to create a training (80\%) and held-out
+test set (20\%). The splits were stratified to maintain the overall
+label distribution. We performed five-fold cross-validation on the
+training data to select the best hyperparameter setting and then used
+these hyperparameters to train the models. The model was evaluated on
+the held-out data set. Abbreviations: cvAUC, cross-validation area under
+the receiver operating characteristic curve. \newpage
 \includegraphics{Figure_2.png}
@@
 
-\textbf{Figure 5. Computational efficiency of seven ML models.} The
-median training time was the highest for XGBoost and shortest for
+\textbf{Figure 5. Training times of seven ML models.} The median
+training time was the highest for XGBoost and shortest for
 L2-regularized logistic regression. \newpage
@@
 (C) L2 SVM with linear kernel mean cross-validation AUROC values when
-different hyperparameters were used in training the model. The
-differences in AUROC values when hyperparameters change show that
-hyperparameter tuning is a crucial step in building a ML model.
+different hyperparameters were used in training the model. The stars
+represent the highest performing hyperparameter setting for each model.
 
@@
 kernel, and (D) XGBoost mean cross-validation AUROC values when
-different hyperparameters were used in training the model. The
-differences in AUROC values when hyperparameters change show that
-hyperparameter tuning is a crucial step in building a ML model. \newpage
+different hyperparameters were used in training the model. The stars
+represent the highest performing hyperparameter setting for the models.
+\newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
@@
 \textbf{Figure S3. Classification performance of ML models across cross
-validation when dataset is subsetted .} (A) L2-regularized logistic
-regression and (B) Random forest models were trained using the original
-study design with 490 subjects and subsets of it with 15, 30, 60, 120,
-and 245 subjects. The range among the cross-validation AUROC values
-within both models at lower sample sizes were much larger than when the
-full collection of samples was used to train and validate the models,
-but included the ranges observed with the more complete datasets.
+validation when trained on a subset of th dataset.} (A) L2-regularized
+logistic regression and (B) Random forest models were trained using the
+original study design with 490 subjects and subsets of it with 15, 30,
+60, 120, and 245 subjects. The range among the cross-validation AUROC
+values within both models at lower sample sizes were much larger than
+when the full collection of samples was used to train and validate the
+models, but included the ranges observed with the more complete
+datasets.
 
@@
 
-\begin{tabular}{|l|l|l|l|l|}
+\begin{tabular}{|l|l|l|l|}
 \hline
@@
 \rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} & \textbf{Interpretability} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear & Interpretable \\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear & Interpretable \\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear & Explainable$^*$ \\ \hline
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
+
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
+
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
+
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
 Decision tree & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear & Interpretable \\ \hline
-
-Random forest & \makecell[l]{A classifier that is a decision tree ensemble that \\grow randomly with subsampled data.} & Non-linear & Explainable$^*$ \\ \hline
-
-XGBoost & \makecell[l]{A classifier that is a decision tree ensemble that \\grow with additive training.} & Non-linear & Explainable$^*$ \\ \hline
+root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
+
+Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
+
+XGBoost & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
 
@@
 
+\hypertarget{ref-ribeiro_why_2016}{}
+32. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
+``Why should i trust you?'': Explaining the predictions of any
+classifier. arXiv:160204938 {[}cs, stat{]}.
+
+\hypertarget{ref-nori_interpretml:_2019}{}
+33. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
+\textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
+learning interpretability. arXiv:190909223 {[}cs, stat{]}.
+
 \hypertarget{ref-sze_leveraging_2018}{}
-32. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+34. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-dormann_collinearity:_2013}{}
-33. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+35. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 \hypertarget{ref-sze_fecal_2019}{}
-34. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+36. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
@@
 \hypertarget{ref-kocheturov_massive_2019}{}
-35. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+37. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \hypertarget{ref-kim_improved_2018}{}
-36. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+38. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-37. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+39. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-38. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+40. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-39. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+41. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-40. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+42. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-10-03 code/learning/FigureS1.R

@@ -58,6 +58,24 @@ linear_plots <- fig_data %>%
                               expression(paste(L[2], "-regularized logistic regression"))),
                      breaks=c("L1_Linear_SVM", "L2_Linear_SVM", "L2_Logistic_Regression"),
                      values=c("#00AFBB", "#E7B800", "#FC4E07"))+
+  annotate("point", # best hp for L1 Linear SVM - highest mean cv AUROC
+           x = 0.5, 
+           y = 0.59, 
+           colour = "#00AFBB", 
+           size = 3,
+           shape = 8) +
+  annotate("point", # best hp for L2 Linear SVM - highest mean cv AUROC
+           x = 0.5, 
+           y = 0.71, 
+           colour = "#E7B800", 
+           size = 3,
+           shape = 8) +
+  annotate("point", # best hp for L2 logistic regression - highest mean cv AUROC
+           x = 0.25, 
+           y = 0.71, 
+           colour = "#FC4E07", 
+           size = 3,
+           shape = 8) +
   theme_bw() +
   theme(legend.background = element_rect(linetype="solid", color="black", size=0.5),
         legend.box.margin=margin(c(12,12,12, 12)),

### 2019-10-03 code/learning/FigureS2.R

@@ -98,6 +98,12 @@ rbf_plot <- ggplot(rbf_data, aes(x = sigma, y = C, fill = mean_AUC)) +
   scale_fill_gradient(name= "SVM RBF mean cvAUROC",
                       low = "#FFFFFF",
                       high = "#012345") +
+  annotate("point", # best hp for rbf svm - highest mean cv AUROC
+           x = 0.000001, 
+           y = 0.01, 
+           colour = "#FC4E07", 
+           size = 3,
+           shape = 8) +
   #coord_fixed(ratio = 0.5) +
   #coord_equal() +
   scale_y_log10(name="regularization penalty 
@@ -133,6 +139,12 @@ xgboost_plot <- ggplot(xgboost_data, aes(x = eta, y = subsample, fill = mean_AUC
   scale_fill_gradient(name= "XGBoost mean cvAUROC",
                       low = "#FFFFFF",
                       high = "#012345") +
+  annotate("point", # best hp for xgboost - highest mean cv AUROC
+           x = 0.01, 
+           y = 0.5, 
+           colour = "#FC4E07", 
+           size = 3,
+           shape = 8) +
   scale_y_continuous(name="ratio of the training data
 (subsample)",
     breaks = c(0.4, 0.5, 0.6, 0.7),

### 2019-10-03 code/learning/functions.R

@@ -145,3 +145,46 @@ wilcoxon_test <- function(data, model_name_1, model_name_2){
   wilcox_result <- wilcox.test((data %>% filter(model==model_name_1))$test_aucs, (data %>% filter(model==model_name_2))$test_aucs)
   return(wilcox_result)
 }
+
+# Function to compare 2 models based on histogram
+histogram_p_value <- function(data, model_name_1, model_name_2){
+  differences <- (data %>% filter(model==model_name_2))$test_aucs - (data %>% filter(model==model_name_1))$test_aucs
+  p_value <- 2*min(sum(differences>=0)/100, sum(differences<=0)/100)
+  
+  plot <- ggplot() + 
+    aes(differences) +
+    geom_histogram(aes(y=..density..), colour="black", fill="white", bins=30)+
+    geom_density(alpha=.2, fill="blue") +
+    geom_vline(xintercept = 0, color="red", size=1) +
+    geom_hline(yintercept = 0, color="red", size=1) +
+    #geom_vline(aes(xintercept=mean(differences)),
+    #           color="blue", linetype="dashed") + 
+    coord_cartesian(xlim=c(-0.2,0.2)) +
+    scale_x_continuous(name= "The difference between random forest and L2-regularized
+    logistic regression AUROC values of each datasplit") +
+    scale_y_continuous(expand=c(0,0)) +
+    theme(axis.line = element_blank(),
+          panel.background = element_blank(),
+          axis.text.y = element_blank(),
+          axis.title.y = element_blank(),
+          axis.ticks.y = element_blank(), 
+          axis.text.x=element_text(size = 12, colour='black'),
+          axis.title.x=element_text(size = 14, colour='black')) +
+    annotate("text", x= 0.15, y= 10, label = paste0("p-value = ", p_value))
+  
+  return(plot)
+}
+
+# Permutation test of means from two groups 
+set.seed(200889) 
+detach("package:mosaic", unload=TRUE)
+perm_p_value <- function(data, model_name_1, model_name_2){
+  test_subsample <- all %>% 
+    select(-cv_aucs) %>% 
+    filter(model == model_name_1 | model == model_name_2)
+  library(mosaic)
+  obs <- abs(diff(mosaic::mean(test_aucs ~ model, data = test_subsample)))
+  auc.null <- do(1000) * diff(mosaic::mean(test_aucs ~ shuffle(model), data = test_subsample))
[…]

### 2019-10-03 code/learning/load_AUROC_data.R

@@ -44,7 +44,6 @@ difference_model <- data.frame(difference, test_all$model)
 # Get the order index from small to large of the differences
 difference_index <- order(abs(difference_model$difference))
 
-
 # Compute what happens when OTU367 is permuted in random forest and decision tree
 
 # These will be the 5 OTUs that have effect the testing AUROC the most when permuted

### 2019-10-03 code/learning/significance.R

@@ -1,0 +1,62 @@
+library(tidyverse)
+source('code/learning/functions.R')
+######################################################################
+#--------- Load cv and test AUROCs of 7 models for 100 datasplits--------#
+######################################################################
+
+# Read in the cvAUCs, testAUCs for 100 splits.
+best_files <- list.files(path= 'data/process', pattern='combined_best.*', full.names = TRUE)
+all <- map_df(best_files, read_files) 
+
+histogram_p_value(all, "L2_Logistic_Regression","Random_Forest")
+
+# Resampling test of means from two groups for L2-logistic regression and random forest
+
+model_name_1 <- "Decision_Tree"
+model_name_2 <- "Random_Forest"
+histogram_p_value <- function(data, model_name_1, model_name_2){
+  data_1 <- all %>% filter(model==model_name_1)
+  data_2 <- all %>% filter(model==model_name_2)
+  library(mosaic)
+  observed_diff <- base::mean(data_2$test_aucs) - mean(data_1$test_aucs)
+  resampled_data1 <- do(1000) * resample(data_1) 
+  resampled_data2 <- do(1000) * resample(data_2)
+  detach("package:mosaic", unload=TRUE)
+  d1 <- resampled_data1 %>% 
+    spread(.index, test_aucs) %>% 
+    select(-cv_aucs, -model, -orig.id, -.row) %>% 
+    colMeans(na.rm=TRUE)
+  d2 <- resampled_data2 %>% 
+    spread(.index, test_aucs) %>% 
+    select(-cv_aucs, -model, -orig.id, -.row) %>% 
+    colMeans(na.rm=TRUE)
+  resampled_diff <- d2-d1
+  ggplot() + aes(resampled_diff, color=resampled_diff>=observed_diff) + geom_histogram(fill="white", alpha=0.5, position="identity") + scale_color_brewer(palette="Dark2")
+  
+  
+  plot <- ggplot() + 
+    aes(resampled_diff, color=resamples_diff>=observed_diff) +
+    geom_histogram(aes(y=..density..), colour="black", fill="white", bins=30)+
+    geom_density(alpha=.2, fill="blue") +
+    geom_vline(xintercept = 0, color="red", size=1) +
+    geom_hline(yintercept = 0, color="red", size=1) +
+    #geom_vline(aes(xintercept=mean(differences)),
+    #           color="blue", linetype="dashed") + 
+    coord_cartesian(xlim=c(-0.2,0.2)) +
[…]

### 2019-10-07 submission/manuscript.Rmd
updates after Pats 2nd final read

@@
 ---
-title: '**Effective application of machine learning to bacterial 16S rRNA gene sequencing data**'
+title: '**Effective application of machine learning to microbiome based classification problems**'
 csl: mbio.csl
@@
 
-\vspace{40mm}
+\vspace{38mm}
 
@@
 
-Prediction of health outcomes using machine learning (ML) is rapidly being adopted in human microbiome studies.
+Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies.
 However, these ML models are likely overoptimistic in terms of quantifying predictive performance.
@@
 As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases.
-However, the structure of the human microbiome is remarkably variable AMONG individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
+However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
 This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals.
@@
 Instead, subsets of the microbiome account for differences in health.
-Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually.
-Recently, machine learning models (ML) have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.
+Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population seperately [@segata_metagenomic_2011].
+Recently, machine learning models (ML) have grown in popularity among microbiome researchers because we are able to sample large numbers of individuals and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.
 
 ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data.
-Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017].
+Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017].
 The task of diagnosing an individual relies on a rigorously validated model.
@@
 This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
-Although complex models may in some cases, capture important non-linear relationships and therefore yield better predictions, they may result in black boxes that lack interpretibility.
+Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they result in black boxes that lack interpretibility.
 Such models require post hoc explanations to quantify the importance of each feature in making predictions.
 Depending on the goal of the modeling, researchers may choose to use different approaches.
-For example, researchers trying to identify the populations of microbiota associated wwith disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance.
-Nonetheless, it is important to understand the tradeoff and accepting that this tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018].
+For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance.
+Nonetheless, it is important to understand the tradeoff and accept that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
 It is important for researchers to justify their choice of modelling approach.
@@
 Hyperparameters are parameters which need to be specified or tuned by the user, in order to train a model for a specific modeling problem.
-For example, when using regularization, (C) is a hyperparameter that indicates the penalty for overfitting.
+For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting.
 Hyperparameters are tuned using the training data to find the best model.
@@
 We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2].
-The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0, where the model’s predictions are perfectly incorrect.
+The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls.
 An AUROC value of 0.5 indicates that model's predictions are no differect than random.
-To select hyperparameters, we performed a full grid search for hyperparameter settings when training our models.
-Default hyperparameter settings in previously developed ML packages in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
-In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization strengths (C).
+To select hyperparameters, we performed a full grid search for hyperparameter settings when training the models.
+Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
+For example, of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization strengths (C).
 
@@
 __Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2].
-The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01).
+The random forest model had higher test AUROC values than the other models for detecting SRNs (resampling test from two groups, p < 0.05).
 The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
@@
 L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
-Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well non-linear models.
+Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.
+To support our claim, we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3).
+Our results underscored that L2-regularized logistic regression perform nearly as well as random forest.
 
@@
 The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
-These differences are relatively small and gives us confidence in our estimate of the generalization performance of the models.
+These differences are relatively small and give us confidence in our estimate of the generalization performance of the models.
 
@@
 
-To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3].
+To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4].
 The variation in cross-validation performance within both models at lower sample sizes was larger than when the full collection of samples was used to train and validate the models.
@@
 Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
-Permutation importance analysis is a posthoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data.
+Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data.
 We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
 We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4].
-Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models.
-Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477).
+Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models.
+Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTUs 2, 12, 361, 477).
 Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance.
@@
 In this case, similar to previous studies [@sze_leveraging_2018; @wirbel_meta-analysis_2019], it was not possible to distinguish between health and disease using a single OTU.
-Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e.
+Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e.
 done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.
 
-To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4].
-When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 of the 20 top OTUs (e.g.
+To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S5].
+When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g.
 OTU 609, 822, 1239) were deemed important by both interpretation methods.
@@
 To address these needs, we developed an open-sourced framework to ML models.
-using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized.
+Using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized.
 In terms of predictive performance, the random forest model had the best AUROC compared to the other six models.
-However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest.
+However, the second-best model was L2-regularized logistic regression with a median AUROC difference less than 0.015 compared to random forest.
 While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model and is most often explained using post-hoc methods such as permutation importance.
+In terms of interpretability, random forest is a non-linear ML model and is most often explained using post hoc methods such as permutation importance.
 On the other hand, using L2-regularized logistic regression we ranked the importance of each OTU based on their feature weights.
@@
 
-As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that predicts a binary health outcome.
+We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome.
 We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance.
@@
 When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group.
-We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among correlated features.
+We grouped OTUs that had a perfect correlation with each other, however, we could reduce the correlation threshold to further investigate the relationships among correlated features.
 It is important to know the correlation structures of the data to avoid misinterpreting the models.
@@
 We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs.
-Furthermore, as number of classes increases, more samples are required for each category to train an accurate model.
+Furthermore, as the number of classes increases, more samples are required for each category to train an accurate model.
 We also did not use regression-based analyses to predict a non-categorical outcome.
@@
 Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks.
-Deep learning methods hold promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which can result in overfitting.
+Deep learning methods hold promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which result in overfitting.
 
@@
 
-**Data collection and study population.** The original stool samples described in our analysis were obtained from patients recruited by Great Lakes-New England Early Detection Research Network (Reference from Mack?).
+**Data collection and study population.** The original stool samples described in our analysis were obtained from patients recruited by Great Lakes-New England Early Detection Research Network [@baxter_microbiota-based_2016].
 Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy.
@@
 We assigned patients into two classes: those with normal colons and those with screen relevant neoplasias (SRNs).
-The normal class included patients with normal colons or non-advanced adenomas whereas the SRN class included patients with advanced adenomas or carcinomas (reference for SRN?).
+The normal class included patients with normal colons or non-advanced adenomas whereas the SRN class included patients with advanced adenomas or carcinomas [@redwood_stool_2016].
 Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas.
@@
 A full grid search was needed to avoid large variation in prediction performance.
-For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
-For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
-For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has.
-For random forest, we tuned the __number of features__ to consider when looking for the best tree split.
-For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.
+For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
+For SVM with radial basis function kernel we also tuned sigma hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
+For the decision tree model, we tuned the depth of the tree where deeper the tree, the more splits it has.
+For random forest, we tuned the number of features to consider when looking for the best tree split.
+For XGBoost, we tuned for learning rate and the fraction of samples to be used for fitting the individual base learners.
 Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016] .
@@
 **Permutation importance workflow.** We calculated a Spearman's rank-order correlation matrix and defined correlated OTUs as having perfect correlation (correlation coefficient = 1 and p < 0.01).
-Non-correlated OTUs were permuted individually whereas correlated ones were grouped together and permuted at the same time.
+OTUs without a perfect correlation to each other were permuted individually whereas correlated ones were grouped together and permuted at the same time.
 
 **Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1).
-We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best predictive performance.
+We compared the AUROC values of the models by performing a resampling test.
+Our H0 model assumed that the AUROC values of two models were equal.
+To create the stocasthic model following H0, we permuted the AUROC values of the two models 10,000 times to be resampled to produce artificial datasets.
+We than compared the difference of means of the artificial resampled data and means of the real data.
 
 **Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
+
+**Acknowlegements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network.
+We would like to thank the members of the Schloss lab for their valuable feedback during the formulation of the manuscript.
+Salary support for M.R came from NIH grant 1R01CA215574.
+Salary support for P.D.S.
+came from NIH grants P30DK034933 and 1R01CA215574.
 
@@
 ** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-Predictive performance of random forest model was higher than other ML models.
+The predictive performance of random forest model was higher than other ML models.
 The boxplot shows quartiles at the box ends and the median as the horizontal line in the box.
@@
 The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, salmon were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only.
-For all of the tree-based models, a Peptostreptococcus species (OTU00367) had the largest impact on predictive performance.
+For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
 \newpage
@@
 **Figure S3.
-Classification performance of ML models across cross validation when trained on a subset of th dataset.** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
-The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
-
+Histogram of AUROC differences between L2-regularized logistic regression and random forest for each datasplit.** The percentage of dataplits where the difference between random forest and L2-regularized logistic regression was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25.
+The p-value was calculated for a double tail event.
 \newpage
@@
 **Figure S4.
-Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
+Classification performance of ML models across cross validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
+The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
 
@@
 **Figure S5.
-Training times of ML models when dataset is subsetted .** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
-As the size of the dataset increased, the training times for L2-regularized logistic regression and random forest models increased linearly.
+Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
 
@@
 \begin{tablenotes}\footnotesize
-\item[1]{$^*$Explainable models are not inherently interpretable but can be explained with post-hoc analyses.}
 \end{tablenotes}

### 2019-10-07 submission/manuscript.tex
updates after Pats 2nd final read

@@
 diseases. However, the structure of the human microbiome is remarkably
-variable AMONG individuals to the point where it is often difficult to
+variable among individuals to the point where it is often difficult to
 identify the bacterial populations that are associated with diseases
@@
 classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
-random forest model had significantly higher test AUROC values than the
-other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
-0.01). The median AUROC of the random forest model was 0.695 (IQR
+random forest model had higher test AUROC values than the other models
+for detecting SRNs (Resampling test from two groups, p \textless{}
+0.05). The median AUROC of the random forest model was 0.695 (IQR
 0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
@@
 complex model (XGBoost) did not have the best performance and that the
-most interpretable models (L2-regularized logistic regression and SVM
-with linear kernel) performed nearly as well non-linear models.
+most interpretable models (L2-regularized logistic regression and
+L2-regularized SVM with linear kernel) performed nearly as well
+non-linear models. To support our claim, we investigated the
+distribution of differences between random forest and L2-regularized
+logistic regression AUROC values for each data-split (Figure S3). Our
+results underscored that L2-regularized logistic regression perform
+nearly as well as random forest.
 
@@
 
-\textbf{Figure S3. Classification performance of ML models across cross
-validation when trained on a subset of th dataset.} (A) L2-regularized
+\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
+logistic regression and random forest for each datasplit.} The
+percentage of dataplits where the difference between random forest and
+L2-regularized logistic regression was higher than or equal to 0 were
+0.75, lower than or equal to 0 were 0.25. \newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
+
+\textbf{Figure S4. Classification performance of ML models across cross
+validation when trained on a subset of the dataset.} (A) L2-regularized
 logistic regression and (B) Random forest models were trained using the
@@
 
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
-
-\textbf{Figure S4. Interpretation of the linear ML models with
+\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
+
+\textbf{Figure S5. Interpretation of the linear ML models with
 permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
@@
 
-\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
-
-\textbf{Figure S5. Training times of ML models when dataset is subsetted
+\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
+
+\textbf{Figure S6. Training times of ML models when dataset is subsetted
 .} (A) L2-regularized logistic regression and (B) Random forest models

### 2019-10-07 code/learning/Figure2.R

@@ -27,7 +27,35 @@ best_performance <- map_df(best_files, read_files) %>%
 ######################################################################
 #Plot the AUC values for cross validation and testing for each model #
 ######################################################################
+# Permutation test of means from two groups 
+perm_p_value <- function(data, model_name_1, model_name_2){
+  test_subsample <- all %>% 
+    select(-cv_aucs) %>% 
+    filter(model == model_name_1 | model == model_name_2)
+  library(mosaic)
+  obs <- abs(diff(mosaic::mean(test_aucs ~ model, data = test_subsample)))
+  auc.null <- do(10000) * diff(mosaic::mean(test_aucs ~ shuffle(model), data = test_subsample))
+  n <- length(auc.null[,1])
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(auc.null[,1]) >= obs)  
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
+  p.value=(r+1)/(n+1)
+  plot <- ggplot(auc.null, aes(x=auc.null[,1], color=auc.null[,1]>=obs)) + geom_histogram(fill="white", alpha=0.5, position="identity") + scale_color_brewer(palette="Dark2")
+  return(p.value)
+}
 
+set.seed(101)
+#resampling method 
+rf_logit <- perm_p_value(all, "L2_Logistic_Regression","Random_Forest") # different p=0.00049995
+perm_p_value(all,"XGBoost", "L2_Logistic_Regression") # not different
+perm_p_value(all, "XGBoost", "L2_Linear_SVM") # not different
+perm_p_value(all, "RBF_SVM", "L2_Linear_SVM") # not different
+perm_p_value(all, "L1_Linear_SVM", "RBF_SVM") # different p = 0.00069993
+perm_p_value(all, "Decision_Tree", "RBF_SVM") # different p = 9.999e-05
+perm_p_value(all, "L1_Linear_SVM", "Decision_Tree") # different p = 9.999e-05
+
+library(ggpubr)
+my_comparisons = c("Random_Forest", "L2_Logistic_Regression")
 
 performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC, fill = Performance)) +
   geom_boxplot(alpha=0.5, fatten = 4) +
@@ -47,6 +75,7 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
                             "XGBoost",
[…]

### 2019-10-07 code/learning/FigureS3.R

@@ -1,131 +1,83 @@
-# Author: Begum Topcuoglu
-# Date: 2018-08-18
-#
-######################################################################
-# This script plots Figure S3:
-#   1. cvAUC (medians of 100 repeats for the best hp) of 100 datasplits
-#   2. How subsampling changes the performance
-######################################################################
-
-######################################################################
-# Load in needed functions and libraries
+library(tidyverse)
 source('code/learning/functions.R')
-# detach("package:randomForest", unload=TRUE) to run
 ######################################################################
-# -------------------- Read files ------------------------------------>
-grab_number <- function(file){
-  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
-    str_extract("[^_]*$")
-}
-# Read in files as delim that are saved in a list with a pattern
-read_files <- function(filenames){
-  for(file in filenames){
-    number <- grab_number(file)
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',') %>% 
-      mutate(number = number)
-  }
-  return(data)
-}
-######################################################################
-# Load .csv data generated with modeling pipeline
+#--------- Load cv and test AUROCs of 7 models for 100 datasplits--------#
 ######################################################################
 
 # Read in the cvAUCs, testAUCs for 100 splits.
-logit_files <- list.files(path= 'data/process/subsampling', pattern='combined_best_hp_results_L2_Logistic_Regression_.*', full.names = TRUE)
+best_files <- list.files(path= 'data/process', pattern='combined_best.*', full.names = TRUE)
+all <- map_df(best_files, read_files) 
 
-logit_performance <- map_df(logit_files, read_files) %>% 
-  melt_data() %>% 
-  unite_("model_number", c("model","number")) %>% 
-  filter(Performance != "Testing") %>% 
-  group_by(model_number) 
+
+# Function to take differences for each data-split and compare 2 models based on histogram
+histogram_p_value <- function(data, model_name_1, model_name_2){
[…]

### 2019-10-07 code/learning/FigureS4.R

@@ -1,223 +1,131 @@
 # Author: Begum Topcuoglu
-# Date: 2018-02-13
+# Date: 2018-08-18
 #
 ######################################################################
-# This script plots permutation importance results of linear models
+# This script plots Figure S3:
+#   1. cvAUC (medians of 100 repeats for the best hp) of 100 datasplits
+#   2. How subsampling changes the performance
 ######################################################################
 
-
 ######################################################################
-#----------------- Read in necessary libraries -------------------#
+# Load in needed functions and libraries
+source('code/learning/functions.R')
+# detach("package:randomForest", unload=TRUE) to run
 ######################################################################
-deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
-for (dep in deps){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
+# -------------------- Read files ------------------------------------>
+grab_number <- function(file){
+  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
+    str_extract("[^_]*$")
+}
+# Read in files as delim that are saved in a list with a pattern
+read_files <- function(filenames){
+  for(file in filenames){
+    number <- grab_number(file)
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>% 
+      mutate(number = number)
   }
-  library(dep, verbose=FALSE, character.only=TRUE)
+  return(data)
 }
 ######################################################################
-#----------------- Call the functions we will use -----------------#
+# Load .csv data generated with modeling pipeline
 ######################################################################
 
-source("code/learning/functions.R")
+# Read in the cvAUCs, testAUCs for 100 splits.
+logit_files <- list.files(path= 'data/process/subsampling', pattern='combined_best_hp_results_L2_Logistic_Regression_.*', full.names = TRUE)
 
+logit_performance <- map_df(logit_files, read_files) %>% 
+  melt_data() %>% 
+  unite_("model_number", c("model","number")) %>% 
[…]

### 2019-10-07 code/learning/FigureS5.R

@@ -1,95 +1,223 @@
+# Author: Begum Topcuoglu
+# Date: 2018-02-13
+#
 ######################################################################
-#---------- Compare training time with subsampling results -------#
+# This script plots permutation importance results of linear models
 ######################################################################
 
-# Load in needed functions and libraries
-source('code/learning/functions.R')
-# detach("package:randomForest", unload=TRUE) to run
 
-
-# -------------------- Read files ------------------------------------>
-grab_number <- function(file){
-  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
-    str_extract("[^_]*$")
-}
-
-# Read in files as delim that are saved in a list with a pattern
-read_files <- function(filenames){
-  for(file in filenames){
-    number <- grab_number(file)
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',') %>% 
-      mutate(number = number)
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
+deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
   }
-  return(data)
+  library(dep, verbose=FALSE, character.only=TRUE)
 }
+######################################################################
+#----------------- Call the functions we will use -----------------#
+######################################################################
+
+source("code/learning/functions.R")
+
+######################################################################
+#--------------Run the functions and plot importance ----------#
+######################################################################
+
+# ----------- Read in saved combined feature importances ---------->
+# List the important features files by defining a pattern in the path
+
+# Correlated files are:
+# For linear models these files have the feature weights and coefficients.
[…]

### 2019-10-07 code/learning/FigureS6.R

@@ -1,0 +1,95 @@
+######################################################################
+#---------- Compare training time with subsampling results -------#
+######################################################################
+
+# Load in needed functions and libraries
+source('code/learning/functions.R')
+# detach("package:randomForest", unload=TRUE) to run
+
+
+# -------------------- Read files ------------------------------------>
+grab_number <- function(file){
+  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
+    str_extract("[^_]*$")
+}
+
+# Read in files as delim that are saved in a list with a pattern
+read_files <- function(filenames){
+  for(file in filenames){
+    number <- grab_number(file)
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>% 
+      mutate(number = number)
+  }
+  return(data)
+}
+# -------------------------------------------------------------------->
+
+# Read in the traintime 100 splits.
+logit_files <- list.files(path= 'data/process/subsampling', pattern='combined_traintime_L2_Logistic_Regression_.*', full.names = TRUE)
+rf_files <- list.files(path= 'data/process/subsampling', pattern='combined_traintime_Random_Forest_.*', full.names = TRUE)
+
+
+logit_performance <- map_df(logit_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+logit_performance$number <-  as.factor(logit_performance$number)
+logit_performance$number <- factor(logit_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+rf_performance <- map_df(rf_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+rf_performance$number <-  as.factor(rf_performance$number)
+rf_performance$number <- factor(rf_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+
+
+logit_plot <- ggplot(logit_performance, aes(x = number, y = x)) +
+  geom_boxplot(alpha=0.6, fatten = 4, fill="darkred") +
+  geom_smooth(method="lm", aes(color="Exp Model"), formula= (y ~ exp(x)), se=FALSE, linetype = 1) +
+  scale_y_continuous(name = "L2-regularized logistic regression 
+training time (hours)",
+                     breaks = seq(0, 0.2, 0.05),
[…]

### 2019-10-07 code/learning/functions.R

@@ -145,46 +145,3 @@ wilcoxon_test <- function(data, model_name_1, model_name_2){
   wilcox_result <- wilcox.test((data %>% filter(model==model_name_1))$test_aucs, (data %>% filter(model==model_name_2))$test_aucs)
   return(wilcox_result)
 }
-
-# Function to compare 2 models based on histogram
-histogram_p_value <- function(data, model_name_1, model_name_2){
-  differences <- (data %>% filter(model==model_name_2))$test_aucs - (data %>% filter(model==model_name_1))$test_aucs
-  p_value <- 2*min(sum(differences>=0)/100, sum(differences<=0)/100)
-  
-  plot <- ggplot() + 
-    aes(differences) +
-    geom_histogram(aes(y=..density..), colour="black", fill="white", bins=30)+
-    geom_density(alpha=.2, fill="blue") +
-    geom_vline(xintercept = 0, color="red", size=1) +
-    geom_hline(yintercept = 0, color="red", size=1) +
-    #geom_vline(aes(xintercept=mean(differences)),
-    #           color="blue", linetype="dashed") + 
-    coord_cartesian(xlim=c(-0.2,0.2)) +
-    scale_x_continuous(name= "The difference between random forest and L2-regularized
-    logistic regression AUROC values of each datasplit") +
-    scale_y_continuous(expand=c(0,0)) +
-    theme(axis.line = element_blank(),
-          panel.background = element_blank(),
-          axis.text.y = element_blank(),
-          axis.title.y = element_blank(),
-          axis.ticks.y = element_blank(), 
-          axis.text.x=element_text(size = 12, colour='black'),
-          axis.title.x=element_text(size = 14, colour='black')) +
-    annotate("text", x= 0.15, y= 10, label = paste0("p-value = ", p_value))
-  
-  return(plot)
-}
-
-# Permutation test of means from two groups 
-set.seed(200889) 
-detach("package:mosaic", unload=TRUE)
-perm_p_value <- function(data, model_name_1, model_name_2){
-  test_subsample <- all %>% 
-    select(-cv_aucs) %>% 
-    filter(model == model_name_1 | model == model_name_2)
-  library(mosaic)
-  obs <- abs(diff(mosaic::mean(test_aucs ~ model, data = test_subsample)))
-  auc.null <- do(1000) * diff(mosaic::mean(test_aucs ~ shuffle(model), data = test_subsample))
[…]

### 2019-10-07 code/learning/significance.R

@@ -8,55 +8,8 @@ source('code/learning/functions.R')
 best_files <- list.files(path= 'data/process', pattern='combined_best.*', full.names = TRUE)
 all <- map_df(best_files, read_files) 
 
-histogram_p_value(all, "L2_Logistic_Regression","Random_Forest")
+plot <- histogram_p_value(all, "L2_Logistic_Regression","Random_Forest")
 
-# Resampling test of means from two groups for L2-logistic regression and random forest
+#resampling method shows that Random forest is best
+perm_p_value(all, "L2_Logistic_Regression","Random_Forest")
 
-model_name_1 <- "Decision_Tree"
-model_name_2 <- "Random_Forest"
-histogram_p_value <- function(data, model_name_1, model_name_2){
-  data_1 <- all %>% filter(model==model_name_1)
-  data_2 <- all %>% filter(model==model_name_2)
-  library(mosaic)
-  observed_diff <- base::mean(data_2$test_aucs) - mean(data_1$test_aucs)
-  resampled_data1 <- do(1000) * resample(data_1) 
-  resampled_data2 <- do(1000) * resample(data_2)
-  detach("package:mosaic", unload=TRUE)
-  d1 <- resampled_data1 %>% 
-    spread(.index, test_aucs) %>% 
-    select(-cv_aucs, -model, -orig.id, -.row) %>% 
-    colMeans(na.rm=TRUE)
-  d2 <- resampled_data2 %>% 
-    spread(.index, test_aucs) %>% 
-    select(-cv_aucs, -model, -orig.id, -.row) %>% 
-    colMeans(na.rm=TRUE)
-  resampled_diff <- d2-d1
-  ggplot() + aes(resampled_diff, color=resampled_diff>=observed_diff) + geom_histogram(fill="white", alpha=0.5, position="identity") + scale_color_brewer(palette="Dark2")
-  
-  
-  plot <- ggplot() + 
-    aes(resampled_diff, color=resamples_diff>=observed_diff) +
-    geom_histogram(aes(y=..density..), colour="black", fill="white", bins=30)+
-    geom_density(alpha=.2, fill="blue") +
-    geom_vline(xintercept = 0, color="red", size=1) +
-    geom_hline(yintercept = 0, color="red", size=1) +
-    #geom_vline(aes(xintercept=mean(differences)),
-    #           color="blue", linetype="dashed") + 
-    coord_cartesian(xlim=c(-0.2,0.2)) +
-    scale_x_continuous(name= "The difference between random forest and L2-regularized
[…]

### 2019-10-09 submission/manuscript.Rmd
edits before sending to Jenna for 2nd time

@@
 
-Among microbiome researchers, the lack of justification when selecting a modelling approach has been due to an implicit assumption that more complex models are better.
+Among microbiome researchers, the lack of justification when selecting a modeling approach has been due to an implicit assumption that more complex models are better.
 This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
-Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they result in black boxes that lack interpretibility.
+Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they result in black boxes that lack interpretability.
 Such models require post hoc explanations to quantify the importance of each feature in making predictions.
@@
 Nonetheless, it is important to understand the tradeoff and accept that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
-It is important for researchers to justify their choice of modelling approach.
+It is important for researchers to justify their choice of modeling approach.
 
@@
 We compared their predictive performance, interpretability, and training time.
-To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
+To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classify individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
 This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
@@
 The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls.
-An AUROC value of 0.5 indicates that model's predictions are no differect than random.
+An AUROC value of 0.5 indicates that model's predictions are no different than random.
 To select hyperparameters, we performed a full grid search for hyperparameter settings when training the models.
 Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
-For example, of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization strengths (C).
+For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and needed to be tuned to find the optimal C value [Figure S1].
 
@@
 __Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2].
-The random forest model had higher test AUROC values than the other models for detecting SRNs (resampling test from two groups, p < 0.05).
-The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`).
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(format(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively.
-L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].
+The random forest model had higher test AUROC values than the other models for detecting SRNs (resampling test between means of two groups, p < 0.01).
+The median AUROC of the random forest model was `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`].
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively.
+L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2].
 Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.
@@
 It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g.
-OTU 50, 426, 609, 822, 1239).
+OTUs 50, 426, 609, 822, 1239).
 The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU.
@@
 We are also challenged by the practical constraints of OTU-based approaches.
-The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added.
+The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added [@westcott_opticlust_2017].
 Finally, we also need independent validation cohorts to test the performance of a diagnostic model.
@@
 Our H0 model assumed that the AUROC values of two models were equal.
-To create the stocasthic model following H0, we permuted the AUROC values of the two models 10,000 times to be resampled to produce artificial datasets.
-We than compared the difference of means of the artificial resampled data and means of the real data.
+To create the stochastic model following H0, we permuted the AUROC values of the two models 10,000 times to be resampled to produce artificial datasets.
+We then compared the difference between the means of the artificial resampled data and the real data.
 
@@
 **Figure 1.
-Machine learning pipeline.
-**   We split the data to create a training (80%) and held-out test set (20%).
+Machine learning pipeline.**   We split the data to create a training (80%) and held-out test set (20%).
 The splits were stratified to maintain the overall label distribution.
@@
 **Figure 2.
-Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.
-** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-The predictive performance of random forest model was higher than other ML models.
+Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
+The predictive performance of random forest model was higher than other ML models (resampling test p-value < 0.01).
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another (pairwise resampling p-value > 0.01).
 The boxplot shows quartiles at the box ends and the median as the horizontal line in the box.
@@
 The stars represent the highest performing hyperparameter setting for each model.
-
 \newpage
@@
 **Figure S3.
-Histogram of AUROC differences between L2-regularized logistic regression and random forest for each datasplit.** The percentage of dataplits where the difference between random forest and L2-regularized logistic regression was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25.
+Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred datasplits.** This histogram shows number of datasplits in each bin.
+The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25.
+The vertical red line highlights the bins where there AUROC difference between the two model is 0.
 The p-value was calculated for a double tail event.
@@
 The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
-
 \newpage
@@
 \small
-\begin{tabular}{|l|l|l|l|}
+\begin{tabular}{|l|l|l|}
 \hline
@@
 \makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-Decision tree & \makecell[l]{A classifier that sorts samples down from the
+
+\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
 root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
@@
 \end{tabular}
-\begin{tablenotes}\footnotesize
-\end{tablenotes}
 \newpage

### 2019-10-09 submission/manuscript.tex
edits before sending to Jenna for 2nd time

@@
 \hypersetup{unicode=true,
-            pdftitle={Effective application of machine learning to bacterial 16S rRNA gene sequencing data},
+            pdftitle={Effective application of machine learning to microbiome based classification problems},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Effective application of machine learning to bacterial 16S rRNA
-gene sequencing data}}
+  \title{\textbf{Effective application of machine learning to microbiome based
+classification problems}}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 Prediction of health outcomes using machine learning (ML) is rapidly
-being adopted in human microbiome studies. However, these ML models are
-likely overoptimistic in terms of quantifying predictive performance.
-Moreover, there is a trend towards using black box models such as random
-forest and neural networks without a discussion of the difficulty of
+being adopted in microbiome studies. However, these ML models are likely
+overoptimistic in terms of quantifying predictive performance. Moreover,
+there is a trend towards using black box models such as random forest
+and neural networks without a discussion of the difficulty of
 interpreting such models when trying to identify microbial biomarkers of
@@
 variation in the human microbiome and typically consider the protective
-or risk effects of each bacterial population individually. Recently,
+or risk effects of each bacterial population seperately (1). Recently,
 machine learning models (ML) have grown in popularity among microbiome
-researchers because of the large amount of data that can now be
-generated and because the models are effective at accounting for the
-interpersonal microbiome variation and the ecology of the disease.
+researchers because we are able to sample large numbers of individuals
+and because the models are effective at accounting for the interpersonal
+microbiome variation and the ecology of the disease.
 
@@
 ecological basis of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases, obesity, and type 2 diabetes (1--16,
-16--18). The task of diagnosing an individual relies on a rigorously
-validated model. However, there are common methodological and reporting
-problems that arise when applying ML to such data, that need to be
-addressed for the field to progress. These problems include a lack of
-transparency in which methods are used and how these methods are
-implemented; evaluating models without a separate held-out test data;
-large variation between the predictive performance on different folds of
-cross-validation; and large variation between cross-validation and
-testing performances. Nevertheless, the microbiome field is making
-progress to avoid some of these pitfalls including validating their
-models on independent datasets (7, 18, 19) and introducing ways to
-better use ML tools (20--23). More work is needed to further improve
-reproducibility and minimize overestimating for model performance.
+inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
+task of diagnosing an individual relies on a rigorously validated model.
+However, there are common methodological and reporting problems that
+arise when applying ML to such data, that need to be addressed for the
+field to progress. These problems include a lack of transparency in
+which methods are used and how these methods are implemented; evaluating
+models without a separate held-out test data; large variation between
+the predictive performance on different folds of cross-validation; and
+large variation between cross-validation and testing performances.
+Nevertheless, the microbiome field is making progress to avoid some of
+these pitfalls including validating their models on independent datasets
+(8, 19, 20) and introducing ways to better use ML tools (21--24). More
+work is needed to further improve reproducibility and minimize
+overestimating for model performance.
 
 Among microbiome researchers, the lack of justification when selecting a
-modelling approach has been due to an implicit assumption that more
+modeling approach has been due to an implicit assumption that more
 complex models are better. This has resulted in a trend towards using
-models such as random forest and deep neural networks (2, 11, 24--26)
+models such as random forest and deep neural networks (3, 12, 25--27)
 over simpler models such as logistic regression or other linear models
-(18, 22, 27). Although complex models may in some cases, capture
+(19, 23, 28). Although in some cases complex models may capture
 important non-linear relationships and therefore yield better
-predictions, they may result in black boxes that lack interpretibility.
-Such models require post hoc explanations to quantify the importance of
-each feature in making predictions. Depending on the goal of the
-modeling, researchers may choose to use different approaches. For
-example, researchers trying to identify the populations of microbiota
-associated wwith disease may desire a more interpretable model, whereas
-clinicians may emphasize predictive performance. Nonetheless, it is
-important to understand the tradeoff and accepting that this tradeoff
-may be minimal (28, 29). It is important for researchers to justify
-their choice of modelling approach.
+predictions, they result in black boxes that lack interpretability. Such
+models require post hoc explanations to quantify the importance of each
+feature in making predictions. Depending on the goal of the modeling,
+researchers may choose to use different approaches. For example,
+researchers trying to identify the microbiota associated with disease
+may desire a more interpretable model, whereas clinicians may emphasize
+predictive performance. Nonetheless, it is important to understand the
+tradeoff and accept that the benefit of more complex, less interpretable
+models may be minimal (29, 30). It is important for researchers to
+justify their choice of modeling approach.
 
@@
 pipeline, we used data from a previously published study that sought to
-classifiy individuals as having normal colons or colonic lesions based
-on the 16S rRNA gene sequences collected from fecal samples (3). This
+classify individuals as having normal colons or colonic lesions based on
+the 16S rRNA gene sequences collected from fecal samples (4). This
 dataset was selected because it is a relatively large collection of
@@
 there is ample evidence that the disease is driven by variation in the
-microbiome (1, 3, 4, 30). With this dataset we developed a ML pipeline
+microbiome (2, 4, 5, 31). With this dataset we developed a ML pipeline
 that can be used in many different scenarios for training and evaluating
@@
 train a model for a specific modeling problem. For example, when using
-regularization, (C) is a hyperparameter that indicates the penalty for
+regularization, C is a hyperparameter that indicates the penalty for
 overfitting. Hyperparameters are tuned using the training data to find
@@
 operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
-ranges from 1.0, where the model perfectly distinguishes between cases
-and controls, to 0, where the model's predictions are perfectly
-incorrect. An AUROC value of 0.5 indicates that model's predictions are
-no differect than random. To select hyperparameters, we performed a full
-grid search for hyperparameter settings when training our models.
-Default hyperparameter settings in previously developed ML packages in
-R, Python, and MATLAB programming languages may be inadequate for
-effective application of classification algorithms and need to be
-optimized for each new ML task. In the example of L1-regularized SVM
-with linear kernel {[}Figure S1{]}, the model showed large variability
-between different regularization strengths (C).
+ranges from 0, where the model's predictions are perfectly incorrect, to
+1.0, where the model perfectly distinguishes between cases and controls.
+An AUROC value of 0.5 indicates that model's predictions are no
+different than random. To select hyperparameters, we performed a full
+grid search for hyperparameter settings when training the models.
+Default hyperparameter settings in previously developed ML packages
+available in R, Python, and MATLAB programming languages may be
+inadequate for effective application of classification algorithms and
+need to be optimized for each new ML task. For example, L1-regularized
+SVM with linear kernel showed large variability between different
+regularization strengths (C) and needed to be tuned to find the optimal
+C value {[}Figure S1{]}.
 
@@
 random forest model had higher test AUROC values than the other models
-for detecting SRNs (Resampling test from two groups, p \textless{}
-0.05). The median AUROC of the random forest model was 0.695 (IQR
-0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
-with linear and radial basis function kernel AUROC values were not
-significantly different from one another and had median AUROC values of
-0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
-0.056), respectively. L1-regularized SVM with linear kernel and decision
-tree had significantly lower AUROC values than the other ML models with
-median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
-{[}Figure 2{]}. Interestingly, these results demonstrate that the most
-complex model (XGBoost) did not have the best performance and that the
-most interpretable models (L2-regularized logistic regression and
-L2-regularized SVM with linear kernel) performed nearly as well
-non-linear models. To support our claim, we investigated the
-distribution of differences between random forest and L2-regularized
-logistic regression AUROC values for each data-split (Figure S3). Our
-results underscored that L2-regularized logistic regression perform
-nearly as well as random forest.
+for detecting SRNs (resampling test between means of two groups, p
+\textless{} 0.01). The median AUROC of the random forest model was 0.695
+{[}IQR 0.650-0.739{]}. L2-regularized logistic regression, XGBoost,
+L2-regularized SVM with linear and radial basis function kernel AUROC
+values were not significantly different from one another and had median
+AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
+0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
+0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
+decision tree had significantly lower AUROC values than the other ML
+models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
+0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
+results demonstrate that the most complex model (XGBoost) did not have
+the best performance and that the most interpretable models
+(L2-regularized logistic regression and L2-regularized SVM with linear
+kernel) performed nearly as well non-linear models. To support our
+claim, we investigated the distribution of differences between random
+forest and L2-regularized logistic regression AUROC values for each
+data-split (Figure S3). Our results underscored that L2-regularized
+logistic regression perform nearly as well as random forest.
 
@@
 tree with a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
-These differences are relatively small and gives us confidence in our
+These differences are relatively small and give us confidence in our
 estimate of the generalization performance of the models.
@@
 random forest models when we subsetted our original study design with
-490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S3{]}. The
+490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
 variation in cross-validation performance within both models at lower
@@
 the degree to which humans can understand the reasons behind a model
-prediction (31--33). Because we often use ML models not just to predict
+prediction (32--34). Because we often use ML models not just to predict
 a health outcome but also to identify the biomarkers for a disease,
@@
 being healthy or having an SRN. It was encouraging that many of the
-highest ranked OTUs were shared across these three models, (e.g.~OTU 50,
-426, 609, 822, 1239). The benefit of this approach was that the results
-of the analysis were based on the trained model parameters and provided
-information regarding the sign and magnitude of the impact of each OTU.
-However, this approach is only possible with linear models.
+highest ranked OTUs were shared across these three models, (e.g.~OTUs
+50, 426, 609, 822, 1239). The benefit of this approach was that the
+results of the analysis were based on the trained model parameters and
+provided information regarding the sign and magnitude of the impact of
+each OTU. However, this approach is only possible with linear models.
 
@@
 importance using the held-out test data. Permutation importance analysis
-is a posthoc explanation of the model, in which we randomly permuted
+is a post hoc explanation of the model, in which we randomly permuted
 groups of perfectly correlated features together and other features
@@
 largest impact, there was only one OTU (OTU 822) that was shared among
-all of the models; however, we found three OTUs (OTU 58, 110, 367) that
+all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
 were important in each of the tree-based models. Similarly, the random
-forest and XGBoost models, shared four of the most important OTUs (OTU
+forest and XGBoost models, shared four of the most important OTUs (OTUs
 2, 12, 361, 477). Permutation analysis results also revealed that with
@@
 in the random forest model, the AUROC only dropped from 0.695 to 0.680.
-In this case, similar to previous studies (22, 34), it was not possible
+In this case, similar to previous studies (23, 35), it was not possible
 to distinguish between health and disease using a single OTU. Although
 permutation analysis allowed us to gauge the importance of an OTU, the
-analysis was post-hoc (i.e.~done using the test data) and these results
+analysis was post hoc (i.e.~done using the test data) and these results
 did not allow us to directly interrogate the models to know whether an
@@
 methods, we used permutation importance to interpret the linear models
-{[}Figure S4{]}. When we analyzed the L1-regularized SVM with linear
+{[}Figure S5{]}. When we analyzed the L1-regularized SVM with linear
 kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S4{]}, 17 of the 20 top OTUs (e.g.~OTU
+permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g.~OTU
 609, 822, 1239) were deemed important by both interpretation methods.
@@
 To address these needs, we developed an open-sourced framework to ML
-models. using this pipeline, we benchmarked seven ML models and showed
+models. Using this pipeline, we benchmarked seven ML models and showed
 that the tradeoff between model complexity and performance may be less
@@
 models. However, the second-best model was L2-regularized logistic
-regression with a median AUROC difference of only 0.015 compared to
+regression with a median AUROC difference less than 0.015 compared to
 random forest. While our implementation of random forest took 83.2 hours
@@
 In terms of interpretability, random forest is a non-linear ML model and
-is most often explained using post-hoc methods such as permutation
+is most often explained using post hoc methods such as permutation
 importance. On the other hand, using L2-regularized logistic regression
@@
 
-As we set out to select the best model, we established a pipeline that
-can be generalized to any modeling method that predicts a binary health
-outcome. We performed a random data-split to create a training set (80\%
-of the data) and a held-out test set (20\% of the data), which we used
-to evaluate predictive performance. We repeated this data-split 100
-times to measure the possible variation in predictive performance.
-During the training, we tuned the model hyperparameters with a repeated
-five-fold cross-validation. Despite the high number of features
-microbiome datasets typically have, the models we built with this
-pipeline were generalizable as shown by the good test AUROCs.
+We established a pipeline that can be generalized to any modeling method
+that predicts a binary health outcome. We performed a random data-split
+to create a training set (80\% of the data) and a held-out test set
+(20\% of the data), which we used to evaluate predictive performance. We
+repeated this data-split 100 times to measure the possible variation in
+predictive performance. During the training, we tuned the model
+hyperparameters with a repeated five-fold cross-validation. Despite the
+high number of features microbiome datasets typically have, the models
+we built with this pipeline were generalizable as shown by the good test
+AUROCs.
 
@@
 our ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (35). To capture all
+correlated OTUs are influenced by one another (36). To capture all
 important features, once we identify highly ranked OTUs, we should
@@
 correlated OTUs to determine their impact as a group. We grouped OTUs
-that had a perfect correlation with each other however, we can reduce
+that had a perfect correlation with each other, however, we could reduce
 the correlation threshold to further investigate the relationships among
@@
 grouping was between patients with normal colons and those with SRNs.
-Furthermore, as number of classes increases, more samples are required
-for each category to train an accurate model. We also did not use
-regression-based analyses to predict a non-categorical outcome. We have
-previously used such an approach to train random forest models to
+Furthermore, as the number of classes increases, more samples are
+required for each category to train an accurate model. We also did not
+use regression-based analyses to predict a non-categorical outcome. We
+have previously used such an approach to train random forest models to
 predict fecal short-chain fatty acid concentrations based on microbiome
-data (36). Our analysis was also limited to shallow learning methods and
+data (37). Our analysis was also limited to shallow learning methods and
 did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (11, 37, 38) but microbiome datasets often
-suffer from having many features and small sample sizes, which can
-result in overfitting.
+learning methods hold promise (12, 38, 39) but microbiome datasets often
+suffer from having many features and small sample sizes, which result in
+overfitting.
 
@@
 novo algorithms commonly in use are slow, require considerable memory,
-and result in different OTU assignments as new data are added. Finally,
-we also need independent validation cohorts to test the performance of a
-diagnostic model. To realize the potential for using ML approaches with
-microbiome data, it is necessary that we direct our efforts to overcome
-these challenges.
+and result in different OTU assignments as new data are added (40).
+Finally, we also need independent validation cohorts to test the
+performance of a diagnostic model. To realize the potential for using ML
+approaches with microbiome data, it is necessary that we direct our
+efforts to overcome these challenges.
 
@@
 samples described in our analysis were obtained from patients recruited
-by Great Lakes-New England Early Detection Research Network (Reference
-from Mack?). Stool samples were provided by adults who were undergoing a
-scheduled screening or surveillance colonoscopy. Participants were
-recruited from: Toronto (ON, Canada), Boston (MA, USA), Houston (TX,
-USA), and Ann Arbor (MI, USA). Patients' colonic health was visually
-assessed by colonoscopy with bowel preparation and tissue histopathology
-of all resected lesions. We assigned patients into two classes: those
-with normal colons and those with screen relevant neoplasias (SRNs). The
-normal class included patients with normal colons or non-advanced
-adenomas whereas the SRN class included patients with advanced adenomas
-or carcinomas (reference for SRN?). Patients with an adenoma greater
-than 1 cm, more than three adenomas of any size, or an adenoma with
-villous histology were classified as having advanced adenomas. There
-were 172 patients with normal colonoscopies, 198 with adenomas, and 120
-with carcinomas. Of the 198 adenomas, 109 were identified as advanced
-adenomas. Together 261 patients were classified as normal and 229
-patients were classified as having a SRN.
+by Great Lakes-New England Early Detection Research Network (5). Stool
+samples were provided by adults who were undergoing a scheduled
+screening or surveillance colonoscopy. Participants were recruited from:
+Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
+(MI, USA). Patients' colonic health was visually assessed by colonoscopy
+with bowel preparation and tissue histopathology of all resected
+lesions. We assigned patients into two classes: those with normal colons
+and those with screen relevant neoplasias (SRNs). The normal class
+included patients with normal colons or non-advanced adenomas whereas
+the SRN class included patients with advanced adenomas or carcinomas
+(41). Patients with an adenoma greater than 1 cm, more than three
+adenomas of any size, or an adenoma with villous histology were
+classified as having advanced adenomas. There were 172 patients with
+normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the
+198 adenomas, 109 were identified as advanced adenomas. Together 261
+patients were classified as normal and 229 patients were classified as
+having a SRN.
 
@@
 Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
-generated by Sze et al (34), who processed the 16S rRNA sequences in
+generated by Sze et al (35), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (39--41);
+to OTUs at 97\% similarity using the OptiClust algorithm (40, 42, 43);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 L2-regularized SVM with linear and radial basis function kernels, we
-tuned the \textbf{cost} hyperparameter which determines the
-regularization strength where smaller values specify stronger
-regularization. For SVM with radial basis function kernel we also tuned
-\textbf{sigma} hyperparameter which determines the reach of a single
-training instance where for a high value of sigma, the SVM decision
-boundary will be dependent on the points that are closest to the
-decision boundary. For the decision tree model, we tuned the
-\textbf{depth of the tree} where deeper the tree, the more splits it
-has. For random forest, we tuned the \textbf{number of features} to
+tuned the cost hyperparameter which determines the regularization
+strength where smaller values specify stronger regularization. For SVM
+with radial basis function kernel we also tuned sigma hyperparameter
+which determines the reach of a single training instance where for a
+high value of sigma, the SVM decision boundary will be dependent on the
+points that are closest to the decision boundary. For the decision tree
+model, we tuned the depth of the tree where deeper the tree, the more
+splits it has. For random forest, we tuned the number of features to
 consider when looking for the best tree split. For XGBoost, we tuned for
-\textbf{learning rate} and the \textbf{fraction of samples} to be used
-for fitting the individual base learners. Recently developed tools such
-as Hyperband help researchers with hyperparameter selection that can be
-incorporated to microbiome studies (42) .
+learning rate and the fraction of samples to be used for fitting the
+individual base learners. Recently developed tools such as Hyperband
+help researchers with hyperparameter selection that can be incorporated
+to microbiome studies (44) .
 
@@
 perfect correlation (correlation coefficient = 1 and p \textless{}
-0.01). Non-correlated OTUs were permuted individually whereas correlated
-ones were grouped together and permuted at the same time.
+0.01). OTUs without a perfect correlation to each other were permuted
+individually whereas correlated ones were grouped together and permuted
+at the same time.
 
@@
 the tidyverse package (v.1.2.1). We compared the AUROC values of the
-seven ML models by Wilcoxon rank sum tests to determine the best
-predictive performance.
+models by performing a resampling test. Our H0 model assumed that the
+AUROC values of two models were equal. To create the stochastic model
+following H0, we permuted the AUROC values of the two models 10,000
+times to be resampled to produce artificial datasets. We then compared
+the difference between the means of the artificial resampled data and
+the real data.
 
@@
 
+\textbf{Acknowlegements.} We thank all the study participants of Great
+Lakes-New England Early Detection Research Network. We would like to
+thank the members of the Schloss lab for their valuable feedback during
+the formulation of the manuscript. Salary support for M.R came from NIH
+grant 1R01CA215574. Salary support for P.D.S. came from NIH grants
+P30DK034933 and 1R01CA215574.
+
 \newpage
@@
 \includegraphics{Figure_1.pdf} \textbf{Figure 1. Machine learning
-pipeline. } We split the data to create a training (80\%) and held-out
+pipeline.} We split the data to create a training (80\%) and held-out
 test set (20\%). The splits were stratified to maintain the overall
@@
 models using AUROC values of all cross validation and testing
-performances. } The median AUROC for diagnosing individuals with SRN
+performances.} The median AUROC for diagnosing individuals with SRN
 using bacterial abundances was higher than chance (depicted by
-horizontal line at 0.50) for all the ML models. Predictive performance
-of random forest model was higher than other ML models. The boxplot
+horizontal line at 0.50) for all the ML models. The predictive
+performance of random forest model was higher than other ML models
+(resampling test p-value \textless{} 0.01). L2-regularized logistic
+regression, XGBoost, L2-regularized SVM with linear and radial basis
+function kernel performances were not significantly different from one
+another (pairwise resampling p-value \textgreater{} 0.01). The boxplot
 shows quartiles at the box ends and the median as the horizontal line in
@@
 decision tree and random forest only. For all of the tree-based models,
-a Peptostreptococcus species (OTU00367) had the largest impact on
+a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
 predictive performance. \newpage
@@
 represent the highest performing hyperparameter setting for each model.
-
 \newpage
-
 \includegraphics[height=30cm, width=15cm]{Figure_S2.png}
@@
 \textbf{Figure S3. Histogram of AUROC differences between L2-regularized
-logistic regression and random forest for each datasplit.} The
+logistic regression and random forest for each of the hundred
+datasplits.} This histogram shows number of datasplits in each bin. The
 percentage of dataplits where the difference between random forest and
-L2-regularized logistic regression was higher than or equal to 0 were
-0.75, lower than or equal to 0 were 0.25. \newpage
+L2-regularized logistic regression AUROC values was higher than or equal
+to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical red
+line highlights the bins where there AUROC difference between the two
+model is 0. The p-value was calculated for a double tail event. \newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
@@
 models, but included the ranges observed with the more complete
-datasets.
-
-\newpage
-
+datasets. \newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
@@
 
-\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
-
-\textbf{Figure S6. Training times of ML models when dataset is subsetted
-.} (A) L2-regularized logistic regression and (B) Random forest models
-were trained using the original study design with 490 subjects and
-subsets of it with 15, 30, 60, 120, and 245 subjects. As the size of the
-dataset increased, the training times for L2-regularized logistic
-regression and random forest models increased linearly.
-
-\newpage
-
 \captionsetup{labelformat=empty}
@@
 
-\begin{tabular}{|l|l|l|l|}
+\begin{tabular}{|l|l|l|}
 \hline
@@
 
-\end{tabular}\begin{tablenotes}\footnotesize
-\item[1]{$^*$Explainable models are not inherently interpretable but can be explained with post-hoc analyses.}
-\end{tablenotes}\newpage
+\end{tabular}\newpage
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-segata_metagenomic_2011}{}
+1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
+\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
+\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
+explanation. Genome Biol \textbf{12}:R60.
+doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.
+
 \hypertarget{ref-zeller_potential_2014}{}
-1. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
 \textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
@@
 \hypertarget{ref-zackular_human_2014}{}
-2. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
 \textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
@@
 \hypertarget{ref-baxter_dna_2016}{}
-3. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
+4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
@@
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-4. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
 \textbf{Schloss PD}. 2016. Microbiota-based model improves the
@@
 \hypertarget{ref-hale_shifts_2017}{}
-5. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
+6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
 \textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
@@
 \hypertarget{ref-pasolli_machine_2016}{}
-6. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
+7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
 \textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
@@
 \hypertarget{ref-sze_looking_2016}{}
-7. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
+8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
 the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
@@
 \hypertarget{ref-walters_meta-analyses_2014}{}
-8. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
+9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
 Meta-analyses of human gut microbes associated with obesity and IBD.
@@
 \hypertarget{ref-vazquez-baeza_guiding_2018}{}
-9. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
+10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
 \textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
@@
 \hypertarget{ref-qin_alterations_2014}{}
-10. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
+11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
 \textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
@@
 \hypertarget{ref-geman_deep_2018}{}
-11. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
+12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
 \textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
@@
 \hypertarget{ref-thaiss_persistent_2016}{}
-12. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
+13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
 \textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
@@
 \hypertarget{ref-dadkhah_gut_2019}{}
-13. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
+14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
 \textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
@@
 \hypertarget{ref-flemer_oral_2018}{}
-14. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
+15. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
 \textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
@@
 \hypertarget{ref-ai_systematic_2017}{}
-16. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
+17. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
 \textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
@@
 \hypertarget{ref-dai_multi-cohort_2018}{}
-17. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
+18. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
 WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
@@
 \hypertarget{ref-mossotto_classification_2017}{}
-18. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
+19. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
 \textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
@@
 \hypertarget{ref-statnikov_comprehensive_2013}{}
-20. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
+21. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
 \textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
@@
 \hypertarget{ref-knights_supervised_2011}{}
-21. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
+22. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
 Supervised classification of human microbiota. FEMS Microbiology Reviews
@@
 \hypertarget{ref-wirbel_meta-analysis_2019}{}
-22. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
+23. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
 K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
@@
 \hypertarget{ref-vangay_microbiome_2019}{}
-23. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
+24. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
 Microbiome learning repo (ML repo): A public repository of microbiome
@@
 \hypertarget{ref-galkin_human_2018}{}
-24. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+25. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
 \textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
@@
 \hypertarget{ref-reiman_using_2017}{}
-25. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+26. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
 convolutional neural networks to explore the microbiome, pp. 4269--4272.
@@
 \hypertarget{ref-fioravanti_phylogenetic_2017}{}
-26. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+27. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
 \textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
@@
 \hypertarget{ref-thomas_metagenomic_2019}{}
-27. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
+28. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
 \textbf{Pasolli E}, \textbf{Armanini F}, \textbf{Zolfo M},
@@
 \hypertarget{ref-rudin_please_2018}{}
-28. \textbf{Rudin C}. 2018. Please stop explaining black box models for
+29. \textbf{Rudin C}. 2018. Please stop explaining black box models for
 high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
@@
 \hypertarget{ref-rudin_optimized_2018}{}
-29. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
+30. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
 Toward trust in machine learning for healthcare and criminal justice.
@@
 \hypertarget{ref-knights_human-associated_2011}{}
-30. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
+31. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
 \textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
@@
 \hypertarget{ref-miller_explanation_2017}{}
-31. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
+32. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
 Insights from the social sciences. arXiv:170607269 {[}cs{]}.
@@
 \hypertarget{ref-ribeiro_why_2016}{}
-32. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
+33. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
 ``Why should i trust you?'': Explaining the predictions of any
@@
 \hypertarget{ref-nori_interpretml:_2019}{}
-33. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
+34. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
 \textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
@@
 \hypertarget{ref-sze_leveraging_2018}{}
-34. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+35. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-dormann_collinearity:_2013}{}
-35. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+36. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 \hypertarget{ref-sze_fecal_2019}{}
-36. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+37. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
@@
 \hypertarget{ref-kocheturov_massive_2019}{}
-37. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+38. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \hypertarget{ref-kim_improved_2018}{}
-38. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+39. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 
+\hypertarget{ref-westcott_opticlust_2017}{}
+40. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+Improved Method for Assigning Amplicon-Based Sequence Data to
+Operational Taxonomic Units. mSphere \textbf{2}.
+doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
+
+\hypertarget{ref-redwood_stool_2016}{}
+41. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
+\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
+\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
+TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
+for screening detection of colorectal neoplasia in alaska native people.
+Mayo Clin Proc \textbf{91}:61--70.
+doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.
+
 \hypertarget{ref-schloss_introducing_2009}{}
-39. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+42. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 
-\hypertarget{ref-westcott_opticlust_2017}{}
-40. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
-Improved Method for Assigning Amplicon-Based Sequence Data to
-Operational Taxonomic Units. mSphere \textbf{2}.
-doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
-
 \hypertarget{ref-rognes_vsearch_2016}{}
-41. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+43. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-42. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+44. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-10-09 code/learning/Figure2.R

@@ -28,22 +28,7 @@ best_performance <- map_df(best_files, read_files) %>%
 #Plot the AUC values for cross validation and testing for each model #
 ######################################################################
 # Permutation test of means from two groups 
-perm_p_value <- function(data, model_name_1, model_name_2){
-  test_subsample <- all %>% 
-    select(-cv_aucs) %>% 
-    filter(model == model_name_1 | model == model_name_2)
-  library(mosaic)
-  obs <- abs(diff(mosaic::mean(test_aucs ~ model, data = test_subsample)))
-  auc.null <- do(10000) * diff(mosaic::mean(test_aucs ~ shuffle(model), data = test_subsample))
-  n <- length(auc.null[,1])
-  # r = #replications at least as extreme as observed effect
-  r <- sum(abs(auc.null[,1]) >= obs)  
-  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
-  p.value=(r+1)/(n+1)
-  plot <- ggplot(auc.null, aes(x=auc.null[,1], color=auc.null[,1]>=obs)) + geom_histogram(fill="white", alpha=0.5, position="identity") + scale_color_brewer(palette="Dark2")
-  return(p.value)
-}
-
+all <- map_df(best_files, read_files) 
 set.seed(101)
 #resampling method 
 rf_logit <- perm_p_value(all, "L2_Logistic_Regression","Random_Forest") # different p=0.00049995
@@ -53,9 +38,6 @@ perm_p_value(all, "RBF_SVM", "L2_Linear_SVM") # not different
 perm_p_value(all, "L1_Linear_SVM", "RBF_SVM") # different p = 0.00069993
 perm_p_value(all, "Decision_Tree", "RBF_SVM") # different p = 9.999e-05
 perm_p_value(all, "L1_Linear_SVM", "Decision_Tree") # different p = 9.999e-05
-
-library(ggpubr)
-my_comparisons = c("Random_Forest", "L2_Logistic_Regression")
 
 performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC, fill = Performance)) +
   geom_boxplot(alpha=0.5, fatten = 4) +
@@ -74,8 +56,13 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
                             expression(paste(L[2], "-regularized linear SVM")),
                             "XGBoost",
[…]

### 2019-10-09 code/learning/FigureS3.R

@@ -16,8 +16,8 @@ histogram_p_value <- function(data, model_name_1, model_name_2){
   
   plot <- ggplot() + 
     aes(differences) +
-    geom_histogram(aes(y=..density..), colour="black", fill="white", bins=30)+
-    geom_density(alpha=.2, fill="blue") +
+    geom_histogram(aes(y=..density..), colour="black", fill="white", bins=100)+
+    #geom_density(alpha=.2, fill="blue") +
     geom_vline(xintercept = 0, color="red", size=1) +
     geom_hline(yintercept = 0, color="red", size=1) +
     #geom_vline(aes(xintercept=mean(differences)),
@@ -66,18 +66,9 @@ compute.p.value <- function(results, observed.effect, precision=3) {
 #  ggplot() + aes(resampled_diff, color=resampled_diff>=observed_diff) + geom_histogram(fill="white", alpha=0.5, position="identity") + scale_color_brewer(palette="Dark2")
   #}
 
-
+set.seed(101)
 
 plot <- histogram_p_value(all, "L2_Logistic_Regression","Random_Forest")
-set.seed(101)
-#resampling method shows that Random forest is best
-perm_p_value(all, "L2_Logistic_Regression","Random_Forest") # different p=0.00049995
-perm_p_value(all,"XGBoost", "L2_Logistic_Regression") # not different
-perm_p_value(all, "XGBoost", "L2_Linear_SVM") # not different
-perm_p_value(all, "RBF_SVM", "L2_Linear_SVM") # not different
-perm_p_value(all, "L1_Linear_SVM", "RBF_SVM") # different p = 0.00069993
-perm_p_value(all, "Decision_Tree", "RBF_SVM") # different p = 9.999e-05
-perm_p_value(all, "L1_Linear_SVM", "Decision_Tree") # different p =
 
 
 ggsave("Figure_S3.png", plot = plot, device = 'png', path = 'submission', width = 12, height = 9)

### 2019-10-09 code/learning/functions.R

@@ -5,7 +5,7 @@
 # Place to store useful functions that will be used repeatedly throughout
 ######################################################################
 
-deps = c("reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+deps = c("reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse", "ggsignif");
 for (dep in deps){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), quiet=TRUE);
@@ -145,3 +145,32 @@ wilcoxon_test <- function(data, model_name_1, model_name_2){
   wilcox_result <- wilcox.test((data %>% filter(model==model_name_1))$test_aucs, (data %>% filter(model==model_name_2))$test_aucs)
   return(wilcox_result)
 }
+
+
+# subsampling test
+perm_p_value <- function(data, model_name_1, model_name_2){
+  test_subsample <- all %>% 
+    select(-cv_aucs) %>% 
+    filter(model == model_name_1 | model == model_name_2)
+  install.packages("mosaic")
+  library(mosaic)
+  obs <- abs(diff(mosaic::mean(test_aucs ~ model, data = test_subsample)))
+  auc.null <- do(10000) * diff(mosaic::mean(test_aucs ~ shuffle(model), data = test_subsample))
+  n <- length(auc.null[,1])
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(auc.null[,1]) >= obs)  
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
+  p.value=(r+1)/(n+1)
+ # plot <- ggplot(auc.null, aes(x=auc.null[,1], color=auc.null[,1]>=obs)) + geom_histogram(fill="white", alpha=0.5, position="identity") + scale_color_brewer(palette="Dark2")
+  detach("package:mosaic", unload=TRUE)
+  return(p.value)
+}
+
+# Find median and IQRs for each AUROC datasplit for each model
+median_iqr <- function(data, model_name){
+  median <- format(round(median((data %>% filter(model==model_name))$test_aucs), 3), nsmall=3)
+  min_iqr <- format(round(rf_median - IQR((data %>% filter(model==model_name))$test_aucs), 3), nsmall=3)
+  max_iqr <- format(round(rf_median + IQR((data %>% filter(model==model_name))$test_aucs), 3), nsmall=3)
+  return(list(median, min_iqr, max_iqr))
+}
+

### 2019-10-14 submission/manuscript.Rmd
Jenna's edits last round

@@
 
-Running title: Machine learning methods in microbiome studies
+Running title: Machine learning methods applied to microbiome studies
 
@@
 
-$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
+$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu, wiensj@umich.edu
 
@@
 To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
-We developed a generalizable pipeline to train, validate and interpret the models.
-To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and extreme gradient boosted trees (XGBoost).
-The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and difficult to interpret.
-Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was more interpretable.
-Our analysis showed that ML models should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.
+We developed a reusable open-source pipeline to train, validate and interpret the models.
+To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost).
+The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and not immidiately interpretable.
+Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and yielded a model that was inherently interpretable.
+Our analysis highlights the importance of choosing an ML approach should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.
 
@@
 Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies.
-However, these ML models are likely overoptimistic in terms of quantifying predictive performance.
-Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease.
-This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study.
+However, estimated performance associated with these ML models is likely overoptimistic.
+Moreover, there is a trend towards using black box models without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease.
+This work represents a step towards developing more reproducble ML practices in applying ML to microbiome research.
+We implement a rigorous pipeline and emphasize the importance of selecting ML models that reflect the goal of the study.
 These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.
@@
 
-As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases.
+As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease.
 However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
@@
 Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population seperately [@segata_metagenomic_2011].
-Recently, machine learning models (ML) have grown in popularity among microbiome researchers because we are able to sample large numbers of individuals and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.
+Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease.
 
@@
 However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress.
-These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances.
-Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019].
-More work is needed to further improve reproducibility and minimize overestimating for model performance.
+These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances.
+Though, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to further improve reproducibility and minimize overestimating for model performance.
 
-Among microbiome researchers, the lack of justification when selecting a modeling approach has been due to an implicit assumption that more complex models are better.
-This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
-Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they result in black boxes that lack interpretability.
+Among microbiome researchers, the lack of justification when selecting a modeling approach has often been due to an implicit assumption that more complex models are better.
+This has resulted in a trend towards using non-linear models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
+Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability.
 Such models require post hoc explanations to quantify the importance of each feature in making predictions.
-Depending on the goal of the modeling, researchers may choose to use different approaches.
-For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance.
-Nonetheless, it is important to understand the tradeoff and accept that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
+Depending on the goal of the modeling, other approaches may be more appropriate.
+For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may (but not always) emphasize predictive performance.
+Nonetheless, it is important to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
 It is important for researchers to justify their choice of modeling approach.
@@
 We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
-We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost.
+We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and gradient boosted trees.
 We compared their predictive performance, interpretability, and training time.
@@
 This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
-With this dataset we developed a ML pipeline that can be used in many different scenarios for training and evaluating models.
+With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models.
 This framework can be easily applied to other host-associated and environmental microbiome datasets.
@@
 A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%).
-The training data was used to build and select the models and the test set was used for evaluating the model.
+The training data were used to build and select the models and the test set was used for evaluating the model.
 
@@
 We focused on different classification algorithms and regularization methods.
-Regularization helps to prevent overfitting by penalizing a model that fits the training data too well.
+Regularization helps to prevent overfitting by penalizing a model that fits the training data too well [@Ng:2004:FSL:1015330.1015435].
 For regularized logistic regression and SVM with a linear kernel, we used L2-regularization to keep all potentially important features.
@@
 L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero).
-To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including a decision tree, a random forest, and XGBoost and an SVM with a non-linear kernel.
+To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including a decision tree, a random forest, and gradient boosted trees (XGBoost) and an SVM with a non-linear kernel.
 
-Model selection required tuning hyperparameters.
-Hyperparameters are parameters which need to be specified or tuned by the user, in order to train a model for a specific modeling problem.
+Model selection requires tuning hyperparameters.
+Hyperparameters are parameters that need to be specified or tuned by the user, in order to train a model for a specific modeling problem.
 For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting.
@@
 An AUROC value of 0.5 indicates that model's predictions are no different than random.
-To select hyperparameters, we performed a full grid search for hyperparameter settings when training the models.
-Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
-For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and needed to be tuned to find the optimal C value [Figure S1].
+To select hyperparameters, we performed a grid search for hyperparameter settings when training the models.
+Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and should to be optimized for each new ML task.
+For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning [Figure S1].
 
 Once hyperparameters were selected, we trained the model using the full training dataset and applied the final model to the held-out data to evaluate the testing predictive performance of each model.
-The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a reliable and robust interpretation of model performance [Figure 1].
+The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a robust interpretation of model performance, less likely to be affected by a "lucky" or "unlucky" split [Figure 1].
 
 __Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2].
-The random forest model had higher test AUROC values than the other models for detecting SRNs (resampling test between means of two groups, p < 0.01).
-The median AUROC of the random forest model was `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`].
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively.
+The random forest model had a median AUROC value of `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`] which was higher than the other models for detecting SRNs.
+However, we determined that the random forest model performance was not statistically significantly higher (p = 0.5) when we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3).
+Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively.
 L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2].
 Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.
-To support our claim, we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3).
-Our results underscored that L2-regularized logistic regression perform nearly as well as random forest.
 
@@
 __Interpretation of each ML model.__ Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019].
-Because we often use ML models not just to predict a health outcome but also to identify the biomarkers for a disease, model interpretation becomes crucial for microbiome studies.
-ML models decrease in interpretability as they increase in complexity.
-In this study we used two methods to help interpret our models.
+Because we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease, model interpretation becomes crucial for microbiome studies.
+ML models often decrease in interpretability as they increase in complexity.
+In this study, we used two methods to help interpret our models.
 
@@
 The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU.
-However, this approach is only possible with linear models.
+However, this approach is limited to linear models or models with prespecified interaction terms.
 
-Second, to analyze non-linear models we interpreted the feature importance using permutation importance.
+Second, to analyze non-linear models we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134].
 Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
@@
 In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.
-In this case, similar to previous studies [@sze_leveraging_2018; @wirbel_meta-analysis_2019], it was not possible to distinguish between health and disease using a single OTU.
-Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e.
-done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.
+Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e., done using the test data).
 
@@
 Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods.
-Although permutation analysis does not not allow us to determine the weight or the sign of the features, these results indicate that both methods are consistent in selecting the most important OTUs.
+These results indicate that both methods are consistent in selecting the most important OTUs.
 
 __The computational efficiency of each ML model.__ We compared the training times of the seven ML models.
-As expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations.
+We did not take advantage of parallelization in training nor was the code optimized for speed.
+Still, as expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations.
 Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
@@
 While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model and is most often explained using post hoc methods such as permutation importance.
-On the other hand, using L2-regularized logistic regression we ranked the importance of each OTU based on their feature weights.
+In terms of interpretability, random forest is a non-linear ML model while L2-regularized logistic regression, a linear model, is easily interpreted according to the feature weights.
 Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.
@@
 We repeated this data-split 100 times to measure the possible variation in predictive performance.
-During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation.
-Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the good test AUROCs.
+During training, we tuned the model hyperparameters with a repeated five-fold cross-validation.
+Despite the high number of features microbiome datasets typically have, the models we built with this pipeline generalized to the held-out test sets.
 
 We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases.
-In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance.
+In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance.
 Human-associated microbial communities have complex correlation structures which create collinearity in the datasets.
@@
 These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments.
-When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group.
+When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group.
 We grouped OTUs that had a perfect correlation with each other, however, we could reduce the correlation threshold to further investigate the relationships among correlated features.
@@
 This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome.
-To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.
+To identify the true underlying microbial factors of a disease, it is crucial to follow up any correlation analyses with further hypothesis testing and experimentation for biological validation.
 
@@
 Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant.
-However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly harder and distinct undertaking.
+However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking.
 For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data.
@@
 
-Our study highlights the need to make educated choices at every step of developing a ML model with microbiome data.
+Our study highlights the need to make educated choices at every step of developing an ML model with microbiome data.
 We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1].
@@
 **Model training and evaluation.** Models were trained using the caret package (v.6.0.81) in R (v.3.5.0).
-We modified the caret code to calculate decision values instead of predicted probabilities for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel.
-These changes were necessary to calculate AUROC values for SVMs.
+We modified the caret code to calculate decision values for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel.
 The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel models are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R}, respectively.
@@
 Then we narrowed and fine-tuned the range of each hyperparameter.
-A full grid search was needed to avoid large variation in prediction performance.
-For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter which determines the regularization strength where smaller values specify stronger regularization.
-For SVM with radial basis function kernel we also tuned sigma hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
-For the decision tree model, we tuned the depth of the tree where deeper the tree, the more splits it has.
+A grid search was needed to avoid large variation in prediction performance.
+Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016].
+For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization.
+For SVM with radial basis function kernel, we also tuned the sigma hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
+For the decision tree model, we tuned the depth of the tree where the deeper the tree, the more splits it has.
 For random forest, we tuned the number of features to consider when looking for the best tree split.
-For XGBoost, we tuned for learning rate and the fraction of samples to be used for fitting the individual base learners.
-Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016] .
+For XGBoost, we tuned the learning rate and the fraction of samples used for fitting the individual base learners.
 
@@
 This allowed the 100 data-splits to be processed through the ML pipeline simultaneously at the same time for each model.
-It is possible to further parallelize the cross-validation step for each hyperparameter setting if limited by computational resources.
+It is possible to further parallelize the cross-validation step for each hyperparameter setting.
 
@@
 **Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1).
-We compared the AUROC values of the models by performing a resampling test.
-Our H0 model assumed that the AUROC values of two models were equal.
-To create the stochastic model following H0, we permuted the AUROC values of the two models 10,000 times to be resampled to produce artificial datasets.
-We then compared the difference between the means of the artificial resampled data and the real data.
+We compared the performance of the models by
+taking the difference of AUROC values from the same datasplit.
+We plotted a histogram that showed the AUROC differences for 100 datasplits and calculated the empirical p-value of the histogram (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for double tail event (e.g., Figure S3).
 
@@
 Machine learning pipeline.**   We split the data to create a training (80%) and held-out test set (20%).
-The splits were stratified to maintain the overall label distribution.
+The splits were stratified to maintain the overall class distribution.
 We performed five-fold cross-validation on the training data to select the best hyperparameter setting and then used these hyperparameters to train the models.
@@
 Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-The predictive performance of random forest model was higher than other ML models (resampling test p-value < 0.01).
-L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another (pairwise resampling p-value > 0.01).
+The predictive performance of random forest model was higher than other ML models though not significantly (p=0.5).
+L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another.
 The boxplot shows quartiles at the box ends and the median as the horizontal line in the box.
@@
 **Figure 3.
-Interpretation of the linear ML models.** The absolute feature weights of (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel were ranked from highest rank, 1, to lowest rank, 100, for each data-split.
+Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel were ranked from highest rank, 1, to lowest rank, 100, for each data-split.
 The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.
@@
 
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
 
@@
 
-XGBoost & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
+\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
 

### 2019-10-14 submission/manuscript.tex
Jenna's edits last round

@@
 \usepackage{hyperref}
-
 \usepackage{helvet} % Helvetica font
@@
 \(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}},
+\href{mailto:wiensj@umich.edu}{\nolinkurl{wiensj@umich.edu}}
 
@@
 relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
-We developed a generalizable pipeline to train, validate and interpret
-the models. To show the effect of model selection, we assessed the
-predictive performance, interpretability, and training time of
+We developed a reusable open-source pipeline to train, validate and
+interpret the models. To show the effect of model selection, we assessed
+the predictive performance, interpretability, and training time of
 L2-regularized logistic regression, L1 and L2-regularized support vector
 machines (SVM) with linear and radial basis function kernels, decision
-trees, random forest, and extreme gradient boosted trees (XGBoost). The
-random forest model performed best at detecting SRNs with an AUROC of
-0.695 {[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and difficult
-to interpret. Despite its simplicity, L2-regularized logistic regression
-followed random forest in predictive performance with an AUROC of 0.680
-{[}IQR 0.625-0.735{]}, trained faster (12 min), and was more
-interpretable. Our analysis showed that ML models should be chosen based
-on the goal of the study, as the choice will inform expectations of
-performance and interpretability.
+trees, random forest, and gradient boosted trees (XGBoost). The random
+forest model performed best at detecting SRNs with an AUROC of 0.695
+{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immidiately
+interpretable. Despite its simplicity, L2-regularized logistic
+regression followed random forest in predictive performance with an
+AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and
+yielded a model that was inherently interpretable. Our analysis
+highlights the importance of choosing an ML approach should be chosen
+based on the goal of the study, as the choice will inform expectations
+of performance and interpretability.
 
@@
 Prediction of health outcomes using machine learning (ML) is rapidly
-being adopted in microbiome studies. However, these ML models are likely
-overoptimistic in terms of quantifying predictive performance. Moreover,
-there is a trend towards using black box models such as random forest
-and neural networks without a discussion of the difficulty of
-interpreting such models when trying to identify microbial biomarkers of
-disease. This work represents a step towards developing better ML
-practices in microbiome research by implementing a rigorous pipeline and
-emphasizing the importance of selecting ML models that reflect the goal
-of the study. These concepts are not particular to the study of health
-outcomes but can also be applied to environmental microbiology studies.
+being adopted in microbiome studies. However, estimated performance
+associated with these ML models is likely overoptimistic. Moreover,
+there is a trend towards using black box models without a discussion of
+the difficulty of interpreting such models when trying to identify
+microbial biomarkers of disease. This work represents a step towards
+developing more reproducble ML practices in applying ML to microbiome
+research. We implement a rigorous pipeline and emphasize the importance
+of selecting ML models that reflect the goal of the study. These
+concepts are not particular to the study of health outcomes but can also
+be applied to environmental microbiology studies.
 
@@
 there is an increasing desire to use microbiome data to diagnose
-diseases. However, the structure of the human microbiome is remarkably
+disease. However, the structure of the human microbiome is remarkably
 variable among individuals to the point where it is often difficult to
@@
 or risk effects of each bacterial population seperately (1). Recently,
-machine learning models (ML) have grown in popularity among microbiome
-researchers because we are able to sample large numbers of individuals
-and because the models are effective at accounting for the interpersonal
-microbiome variation and the ecology of the disease.
+machine learning (ML) models have grown in popularity among microbiome
+researchers as our ability to to sample large numbers of individuals has
+grown; such models can effectively account for the interpersonal
+microbiome variation and the ecology of disease.
 
@@
 which methods are used and how these methods are implemented; evaluating
-models without a separate held-out test data; large variation between
-the predictive performance on different folds of cross-validation; and
-large variation between cross-validation and testing performances.
-Nevertheless, the microbiome field is making progress to avoid some of
-these pitfalls including validating their models on independent datasets
-(8, 19, 20) and introducing ways to better use ML tools (21--24). More
-work is needed to further improve reproducibility and minimize
-overestimating for model performance.
+models without a separate held-out test data; unreported variation
+between the predictive performance on different folds of
+cross-validation; and unreported variation between cross-validation and
+testing performances. Though, the microbiome field is making progress to
+avoid some of these pitfalls including validating their models on
+independent datasets (8, 19, 20) and introducing ways to better use ML
+tools (21--24), more work is needed to further improve reproducibility
+and minimize overestimating for model performance.
 
 Among microbiome researchers, the lack of justification when selecting a
-modeling approach has been due to an implicit assumption that more
+modeling approach has often been due to an implicit assumption that more
 complex models are better. This has resulted in a trend towards using
-models such as random forest and deep neural networks (3, 12, 25--27)
-over simpler models such as logistic regression or other linear models
-(19, 23, 28). Although in some cases complex models may capture
+non-linear models such as random forest and deep neural networks (3, 12,
+25--27) over simpler models such as logistic regression or other linear
+models (19, 23, 28). Although in some cases complex models may capture
 important non-linear relationships and therefore yield better
-predictions, they result in black boxes that lack interpretability. Such
-models require post hoc explanations to quantify the importance of each
-feature in making predictions. Depending on the goal of the modeling,
-researchers may choose to use different approaches. For example,
-researchers trying to identify the microbiota associated with disease
-may desire a more interpretable model, whereas clinicians may emphasize
-predictive performance. Nonetheless, it is important to understand the
-tradeoff and accept that the benefit of more complex, less interpretable
-models may be minimal (29, 30). It is important for researchers to
-justify their choice of modeling approach.
+predictions, they can also result in black boxes that lack
+interpretability. Such models require post hoc explanations to quantify
+the importance of each feature in making predictions. Depending on the
+goal of the modeling, other approaches may be more appropriate. For
+example, researchers trying to identify the microbiota associated with
+disease may desire a more interpretable model, whereas clinicians may
+(but not always) emphasize predictive performance. Nonetheless, it is
+important to understand that the benefit of more complex, less
+interpretable models may be minimal (29, 30). It is important for
+researchers to justify their choice of modeling approach.
 
@@
 also trained four non-linear models: SVM with radial basis function
-kernel, a decision tree, random forest and XGBoost. We compared their
-predictive performance, interpretability, and training time. To
-demonstrate the performance of these modeling approaches and our
-pipeline, we used data from a previously published study that sought to
-classify individuals as having normal colons or colonic lesions based on
-the 16S rRNA gene sequences collected from fecal samples (4). This
+kernel, a decision tree, random forest and gradient boosted trees. We
+compared their predictive performance, interpretability, and training
+time. To demonstrate the performance of these modeling approaches and
+our pipeline, we used data from a previously published study that sought
+to classify individuals as having normal colons or colonic lesions based
+on the 16S rRNA gene sequences collected from fecal samples (4). This
 dataset was selected because it is a relatively large collection of
@@
 there is ample evidence that the disease is driven by variation in the
-microbiome (2, 4, 5, 31). With this dataset we developed a ML pipeline
+microbiome (2, 4, 5, 31). With this dataset, we developed an ML pipeline
 that can be used in many different scenarios for training and evaluating
@@
 composed of 97 individuals of which 52 had an SRN (54\%). The training
-data was used to build and select the models and the test set was used
+data were used to build and select the models and the test set was used
 for evaluating the model.
@@
 methods. Regularization helps to prevent overfitting by penalizing a
-model that fits the training data too well. For regularized logistic
-regression and SVM with a linear kernel, we used L2-regularization to
-keep all potentially important features. For comparison, we also trained
-an L1-regularized SVM with a linear kernel. L1-regularization on
-microbiome data led to a sparser solution (i.e., forced many
-coefficients to zero). To explore the potential for non-linear
-relationships among features to improve classification, we trained
-tree-based models including a decision tree, a random forest, and
-XGBoost and an SVM with a non-linear kernel.
-
-Model selection required tuning hyperparameters. Hyperparameters are
-parameters which need to be specified or tuned by the user, in order to
+model that fits the training data too well (32). For regularized
+logistic regression and SVM with a linear kernel, we used
+L2-regularization to keep all potentially important features. For
+comparison, we also trained an L1-regularized SVM with a linear kernel.
+L1-regularization on microbiome data led to a sparser solution (i.e.,
+forced many coefficients to zero). To explore the potential for
+non-linear relationships among features to improve classification, we
+trained tree-based models including a decision tree, a random forest,
+and gradient boosted trees (XGBoost) and an SVM with a non-linear
+kernel.
+
+Model selection requires tuning hyperparameters. Hyperparameters are
+parameters that need to be specified or tuned by the user, in order to
 train a model for a specific modeling problem. For example, when using
@@
 An AUROC value of 0.5 indicates that model's predictions are no
-different than random. To select hyperparameters, we performed a full
-grid search for hyperparameter settings when training the models.
-Default hyperparameter settings in previously developed ML packages
-available in R, Python, and MATLAB programming languages may be
-inadequate for effective application of classification algorithms and
-need to be optimized for each new ML task. For example, L1-regularized
-SVM with linear kernel showed large variability between different
-regularization strengths (C) and needed to be tuned to find the optimal
-C value {[}Figure S1{]}.
+different than random. To select hyperparameters, we performed a grid
+search for hyperparameter settings when training the models. Default
+hyperparameter settings in previously developed ML packages available in
+R, Python, and MATLAB programming languages may be inadequate for
+effective application of classification algorithms and should to be
+optimized for each new ML task. For example, L1-regularized SVM with
+linear kernel showed large variability between different regularization
+strengths (C) and benefited from tuning {[}Figure S1{]}.
 
@@
 data-split, hyperparameter selection, training and testing steps were
-repeated 100 times to obtain a reliable and robust interpretation of
-model performance {[}Figure 1{]}.
+repeated 100 times to obtain a robust interpretation of model
+performance, less likely to be affected by a ``lucky'' or ``unlucky''
+split {[}Figure 1{]}.
 
@@
 classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
-random forest model had higher test AUROC values than the other models
-for detecting SRNs (resampling test between means of two groups, p
-\textless{} 0.01). The median AUROC of the random forest model was 0.695
-{[}IQR 0.650-0.739{]}. L2-regularized logistic regression, XGBoost,
+random forest model had a median AUROC value of 0.695 {[}IQR
+0.650-0.739{]} which was higher than the other models for detecting
+SRNs. However, we determined that the random forest model performance
+was not statistically significantly higher (p = 0.5) when we
+investigated the distribution of differences between random forest and
+L2-regularized logistic regression AUROC values for each data-split
+(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
 L2-regularized SVM with linear and radial basis function kernel AUROC
@@
 (L2-regularized logistic regression and L2-regularized SVM with linear
-kernel) performed nearly as well non-linear models. To support our
-claim, we investigated the distribution of differences between random
-forest and L2-regularized logistic regression AUROC values for each
-data-split (Figure S3). Our results underscored that L2-regularized
-logistic regression perform nearly as well as random forest.
+kernel) performed nearly as well non-linear models.
 
@@
 the degree to which humans can understand the reasons behind a model
-prediction (32--34). Because we often use ML models not just to predict
-a health outcome but also to identify the biomarkers for a disease,
+prediction (33--35). Because we often use ML models not just to predict
+a health outcome, but also to identify potential biomarkers for disease,
 model interpretation becomes crucial for microbiome studies. ML models
-decrease in interpretability as they increase in complexity. In this
-study we used two methods to help interpret our models.
+often decrease in interpretability as they increase in complexity. In
+this study, we used two methods to help interpret our models.
 
@@
 provided information regarding the sign and magnitude of the impact of
-each OTU. However, this approach is only possible with linear models.
+each OTU. However, this approach is limited to linear models or models
+with prespecified interaction terms.
 
 Second, to analyze non-linear models we interpreted the feature
-importance using permutation importance. Whereas the absolute feature
-weights were determined from the trained models, here we measured
-importance using the held-out test data. Permutation importance analysis
-is a post hoc explanation of the model, in which we randomly permuted
-groups of perfectly correlated features together and other features
-individually across the two groups in the held-out test data. We then
-calculated how much the predictive performance of the model (i.e,
-testing AUROC values) decreased when each OTU or group of OTUs was
+importance using permutation importance (36). Whereas the absolute
+feature weights were determined from the trained models, here we
+measured importance using the held-out test data. Permutation importance
+analysis is a post hoc explanation of the model, in which we randomly
+permuted groups of perfectly correlated features together and other
+features individually across the two groups in the held-out test data.
+We then calculated how much the predictive performance of the model
+(i.e, testing AUROC values) decreased when each OTU or group of OTUs was
 randomly permuted. We ranked the OTUs based on how much the median
@@
 in the random forest model, the AUROC only dropped from 0.695 to 0.680.
-In this case, similar to previous studies (23, 35), it was not possible
-to distinguish between health and disease using a single OTU. Although
-permutation analysis allowed us to gauge the importance of an OTU, the
-analysis was post hoc (i.e.~done using the test data) and these results
-did not allow us to directly interrogate the models to know whether an
-OTU was associated with classifying a subject as being healthy or having
-an SRN.
+Although permutation analysis allowed us to gauge the importance of an
+OTU, the analysis was post hoc (i.e., done using the test data).
 
@@
 regression, 9 and 12 OTUs, respectively, were shared among the two
-interpretation methods. Although permutation analysis does not not allow
-us to determine the weight or the sign of the features, these results
-indicate that both methods are consistent in selecting the most
-important OTUs.
+interpretation methods. These results indicate that both methods are
+consistent in selecting the most important OTUs.
 
 \textbf{The computational efficiency of each ML model.} We compared the
-training times of the seven ML models. As expected, the training times
-increased with the complexity of the model and the number of potential
-hyperparameter combinations. Also, the linear models trained faster than
-non-linear models {[}Figures S1-S2; Figure 5{]}.
+training times of the seven ML models. We did not take advantage of
+parallelization in training nor was the code optimized for speed. Still,
+as expected, the training times increased with the complexity of the
+model and the number of potential hyperparameter combinations. Also, the
+linear models trained faster than non-linear models {[}Figures S1-S2;
+Figure 5{]}.
 
@@
 to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model and
-is most often explained using post hoc methods such as permutation
-importance. On the other hand, using L2-regularized logistic regression
-we ranked the importance of each OTU based on their feature weights.
-Comparing many different models showed us that the most complex model
-was not necessarily the best model for our ML task.
+In terms of interpretability, random forest is a non-linear ML model
+while L2-regularized logistic regression, a linear model, is easily
+interpreted according to the feature weights. Comparing many different
+models showed us that the most complex model was not necessarily the
+best model for our ML task.
 
@@
 repeated this data-split 100 times to measure the possible variation in
-predictive performance. During the training, we tuned the model
+predictive performance. During training, we tuned the model
 hyperparameters with a repeated five-fold cross-validation. Despite the
 high number of features microbiome datasets typically have, the models
-we built with this pipeline were generalizable as shown by the good test
-AUROCs.
+we built with this pipeline generalized to the held-out test sets.
 
 We highlighted the importance of model interpretation to gain greater
-biological insights into microbiota-associated diseases. In this study
+biological insights into microbiota-associated diseases. In this study,
 we showcased two different interpretation methods: ranking each OTU by
@@
 our ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (36). To capture all
+correlated OTUs are influenced by one another (37). To capture all
 important features, once we identify highly ranked OTUs, we should
@@
 test them with follow-up experiments. When we used permutation
-importance, we took collinearity into consideration by grouping
+importance, we partially accounted for collinearity by grouping
 correlated OTUs to determine their impact as a group. We grouped OTUs
@@
 another because they come from the same chromosome. To identify the true
-underlying microbial factors of a disease, it is crucial to do
-correlation analyses and further experimentation for biological
-validation.
+underlying microbial factors of a disease, it is crucial to follow up
+any correlation analyses with further hypothesis testing and
+experimentation for biological validation.
 
@@
 predict fecal short-chain fatty acid concentrations based on microbiome
-data (37). Our analysis was also limited to shallow learning methods and
+data (38). Our analysis was also limited to shallow learning methods and
 did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (12, 38, 39) but microbiome datasets often
+learning methods hold promise (12, 39, 40) but microbiome datasets often
 suffer from having many features and small sample sizes, which result in
@@
 deploying microbiome-based models to make clinical diagnoses or
-predictions is a significantly harder and distinct undertaking. For
-example, we currently lack standardized methods to collect patient
-samples, generate sequence data, and report clinical data. We are also
-challenged by the practical constraints of OTU-based approaches. The de
-novo algorithms commonly in use are slow, require considerable memory,
-and result in different OTU assignments as new data are added (40).
-Finally, we also need independent validation cohorts to test the
-performance of a diagnostic model. To realize the potential for using ML
-approaches with microbiome data, it is necessary that we direct our
-efforts to overcome these challenges.
+predictions is a significantly more challenging and distinct
+undertaking. For example, we currently lack standardized methods to
+collect patient samples, generate sequence data, and report clinical
+data. We are also challenged by the practical constraints of OTU-based
+approaches. The de novo algorithms commonly in use are slow, require
+considerable memory, and result in different OTU assignments as new data
+are added (41). Finally, we also need independent validation cohorts to
+test the performance of a diagnostic model. To realize the potential for
+using ML approaches with microbiome data, it is necessary that we direct
+our efforts to overcome these challenges.
 
 Our study highlights the need to make educated choices at every step of
-developing a ML model with microbiome data. We created an aspirational
+developing an ML model with microbiome data. We created an aspirational
 rubric that researchers can use to identify potential pitfalls when
@@
 the SRN class included patients with advanced adenomas or carcinomas
-(41). Patients with an adenoma greater than 1 cm, more than three
+(42). Patients with an adenoma greater than 1 cm, more than three
 adenomas of any size, or an adenoma with villous histology were
@@
 Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
-generated by Sze et al (35), who processed the 16S rRNA sequences in
+generated by Sze et al (43), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (40, 42, 43);
+to OTUs at 97\% similarity using the OptiClust algorithm (41, 44, 45);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
-calculate decision values instead of predicted probabilities for models
-generated using L2-regularized SVM with linear kernel and L1-regularized
-SVM with linear kernel. These changes were necessary to calculate AUROC
-values for SVMs. The code for these changes on L2-regularized SVM with
-linear kernel and L1-regularized SVM with linear kernel models are
-available at
+calculate decision values for models generated using L2-regularized SVM
+with linear kernel and L1-regularized SVM with linear kernel. The code
+for these changes on L2-regularized SVM with linear kernel and
+L1-regularized SVM with linear kernel models are available at
 \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
@@
 For hyperparameter selection, we started with a granular grid search.
-Then we narrowed and fine-tuned the range of each hyperparameter. A full
-grid search was needed to avoid large variation in prediction
-performance. For L2-regularized logistic regression, L1 and
-L2-regularized SVM with linear and radial basis function kernels, we
-tuned the cost hyperparameter which determines the regularization
-strength where smaller values specify stronger regularization. For SVM
-with radial basis function kernel we also tuned sigma hyperparameter
-which determines the reach of a single training instance where for a
-high value of sigma, the SVM decision boundary will be dependent on the
+Then we narrowed and fine-tuned the range of each hyperparameter. A grid
+search was needed to avoid large variation in prediction performance.
+Recently developed tools such as Hyperband help researchers with
+hyperparameter selection that can be incorporated to microbiome studies
+(46). For L2-regularized logistic regression, L1 and L2-regularized SVM
+with linear and radial basis function kernels, we tuned the cost
+hyperparameter, which controls the regularization strength, where
+smaller values specify stronger regularization. For SVM with radial
+basis function kernel, we also tuned the sigma hyperparameter which
+determines the reach of a single training instance where for a high
+value of sigma, the SVM decision boundary will be dependent on the
 points that are closest to the decision boundary. For the decision tree
-model, we tuned the depth of the tree where deeper the tree, the more
-splits it has. For random forest, we tuned the number of features to
-consider when looking for the best tree split. For XGBoost, we tuned for
-learning rate and the fraction of samples to be used for fitting the
-individual base learners. Recently developed tools such as Hyperband
-help researchers with hyperparameter selection that can be incorporated
-to microbiome studies (44) .
+model, we tuned the depth of the tree where the deeper the tree, the
+more splits it has. For random forest, we tuned the number of features
+to consider when looking for the best tree split. For XGBoost, we tuned
+the learning rate and the fraction of samples used for fitting the
+individual base learners.
 
@@
 the same time for each model. It is possible to further parallelize the
-cross-validation step for each hyperparameter setting if limited by
-computational resources.
+cross-validation step for each hyperparameter setting.
 
@@
 analysis, and data visualizations were performed using R (v.3.5.0) with
-the tidyverse package (v.1.2.1). We compared the AUROC values of the
-models by performing a resampling test. Our H0 model assumed that the
-AUROC values of two models were equal. To create the stochastic model
-following H0, we permuted the AUROC values of the two models 10,000
-times to be resampled to produce artificial datasets. We then compared
-the difference between the means of the artificial resampled data and
-the real data.
+the tidyverse package (v.1.2.1). We compared the performance of the
+models by taking the difference of AUROC values from the same datasplit.
+We plotted a histogram that showed the AUROC differences for 100
+datasplits and calculated the empirical p-value of the histogram (2 x
+min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
+0) for double tail event (e.g., Figure S3).
 
@@
 test set (20\%). The splits were stratified to maintain the overall
-label distribution. We performed five-fold cross-validation on the
+class distribution. We performed five-fold cross-validation on the
 training data to select the best hyperparameter setting and then used
@@
 performance of random forest model was higher than other ML models
-(resampling test p-value \textless{} 0.01). L2-regularized logistic
-regression, XGBoost, L2-regularized SVM with linear and radial basis
-function kernel performances were not significantly different from one
-another (pairwise resampling p-value \textgreater{} 0.01). The boxplot
-shows quartiles at the box ends and the median as the horizontal line in
-the box. The whiskers show the farthest points that were not outliers.
-Outliers were defined as those data points that are not within 1.5 times
-the interquartile ranges. \newpage
+though not significantly (p=0.5). L2-regularized logistic regression,
+XGBoost, L2-regularized SVM with linear and radial basis function kernel
+performances were not significantly different from one another. The
+boxplot shows quartiles at the box ends and the median as the horizontal
+line in the box. The whiskers show the farthest points that were not
+outliers. Outliers were defined as those data points that are not within
+1.5 times the interquartile ranges. \newpage
 \includegraphics[height=17cm, width=12cm]{Figure_3.png}
 
-\textbf{Figure 3. Interpretation of the linear ML models.} The absolute
-feature weights of (A) L2-regularized logistic regression, (B)
+\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
+absolute feature weights of (A) L2-regularized logistic regression, (B)
 L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with
@@
 
+\hypertarget{ref-Ng:2004:FSL:1015330.1015435}{}
+32. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
+and rotational invariance, pp. 78. \emph{In} Proceedings of the
+twenty-first international conference on machine learning. ACM, New
+York, NY, USA.
+
 \hypertarget{ref-miller_explanation_2017}{}
-32. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
+33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
 Insights from the social sciences. arXiv:170607269 {[}cs{]}.
@@
 \hypertarget{ref-ribeiro_why_2016}{}
-33. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
+34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
 ``Why should i trust you?'': Explaining the predictions of any
@@
 \hypertarget{ref-nori_interpretml:_2019}{}
-34. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
+35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
 \textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
@@
 
-\hypertarget{ref-sze_leveraging_2018}{}
-35. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
-rRNA gene surveys to identify reproducible biomarkers in individuals
-with colorectal tumors. mBio \textbf{9}:e00630--18.
-doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
+\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
+36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
+\textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
+importance measure. Bioinformatics \textbf{26}:1340--1347.
+doi:\href{https://doi.org/10.1093/bioinformatics/btq134}{10.1093/bioinformatics/btq134}.
 
 \hypertarget{ref-dormann_collinearity:_2013}{}
-36. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 \hypertarget{ref-sze_fecal_2019}{}
-37. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+38. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
@@
 \hypertarget{ref-kocheturov_massive_2019}{}
-38. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+39. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \hypertarget{ref-kim_improved_2018}{}
-39. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+40. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-40. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+41. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-redwood_stool_2016}{}
-41. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+42. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
 \textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
@@
 
+\hypertarget{ref-sze_leveraging_2018}{}
+43. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+rRNA gene surveys to identify reproducible biomarkers in individuals
+with colorectal tumors. mBio \textbf{9}:e00630--18.
+doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
+
 \hypertarget{ref-schloss_introducing_2009}{}
-42. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-43. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+45. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-44. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+46. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-10-14 code/learning/Figure2.R

@@ -57,12 +57,9 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
                             "XGBoost",
                             expression(paste(L[2], "-regularized logistic regression")),
                             "Random forest")) + 
-  geom_signif(comparisons=list(c("Random_Forest", "L2_Logistic_Regression")), annotations="",
-              y_position = 0.85, tip_length = 0.02, vjust=0.3) +
-  annotate("text", x= 6.5, y= 0.9, size=4, label = paste0("P = ", round(rf_logit, 5))) + 
-  geom_signif(comparisons=list(c("RBF_SVM", "L2_Logistic_Regression")), annotations="",
+  geom_signif(comparisons=list(c("RBF_SVM", "Random_Forest")), annotations="",
               y_position = 0.83, tip_length = 0.02, vjust=0.3) +
-  annotate("text", x= 4.5, y= 0.85, size=4, label = paste0("N.S.")) + 
+  annotate("text", x= 5, y= 0.85, size=4, label = paste0("N.S.")) + 
   theme_bw() +
   theme(plot.margin=unit(c(1.1,1.1,1.1,1.1),"cm"),
         legend.justification=c(1,0),

### 2019-10-14 code/learning/FigureS3.R

@@ -16,24 +16,23 @@ histogram_p_value <- function(data, model_name_1, model_name_2){
   
   plot <- ggplot() + 
     aes(differences) +
-    geom_histogram(aes(y=..density..), colour="black", fill="white", bins=100)+
+    geom_histogram(colour="black", fill="white", bins=100)+
     #geom_density(alpha=.2, fill="blue") +
     geom_vline(xintercept = 0, color="red", size=1) +
-    geom_hline(yintercept = 0, color="red", size=1) +
     #geom_vline(aes(xintercept=mean(differences)),
     #           color="blue", linetype="dashed") + 
-    coord_cartesian(xlim=c(-0.2,0.2)) +
+    coord_cartesian(xlim=c(-0.2,0.2), ylim=c(0,8)) +
     scale_x_continuous(name= "The difference between random forest and L2-regularized
     logistic regression AUROC values of each datasplit") +
-    scale_y_continuous(expand=c(0,0)) +
-    theme(axis.line = element_blank(),
-          panel.background = element_blank(),
-          axis.text.y = element_blank(),
-          axis.title.y = element_blank(),
-          axis.ticks.y = element_blank(), 
-          axis.text.x=element_text(size = 20, colour='black'),
-          axis.title.x=element_text(size = 24, colour='black')) +
-    annotate("text", x= 0.15, y= 10, size=10, label = paste0("p-value = ", p_value))
+    scale_y_continuous(name="Number of datasplits",
+                       breaks = seq(0, 8, 1),
+                       expand=c(0,0)) +
+    theme(panel.background = element_blank(),
+          axis.text.x=element_text(size = 18, colour='black'),
+          axis.title.x=element_text(size = 20, colour='black'),
+          axis.text.y=element_text(size = 18, colour='black'),
+          axis.title.y=element_text(size = 20, colour='black')) +
+    annotate("text", x= 0.15, y= 6, size=6, label = paste0("p-value = ", p_value))
   
   return(plot)
 }

### 2019-10-14 submission/manuscript.Rmd
Jenna's edits 3rd round

@@
 Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and yielded a model that was inherently interpretable.
-Our analysis highlights the importance of choosing an ML approach should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.
+Our analysis highlights the importance of choosing an ML approach based on the goal of the study, as the choice will inform expectations of performance and interpretability.
 
@@
 
-__Model selection and pipeline construction__ We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].
+__Model selection and pipeline construction__.
+We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].
 
@@
 __Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2].
-The random forest model had a median AUROC value of `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`] which was higher than the other models for detecting SRNs.
-However, we determined that the random forest model performance was not statistically significantly higher (p = 0.5) when we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3).
+The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5) (Figure S3).
 Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively.
@@
 For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`.
-In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.
-Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e., done using the test data).
+In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))` which indicated high degree of collinearity in the dataset.
+Permutation analysis allowed us to gauge the importance of each OTU in non-linear models and partially account for collinearity by grouping correlated OTUs to determine their impact as a group.
 
@@
 Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant.
-However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking.
+However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking [@wiens_no_2019].
 For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data.
@@
 A grid search was needed to avoid large variation in prediction performance.
-Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016].
 For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization.
@@
 For XGBoost, we tuned the learning rate and the fraction of samples used for fitting the individual base learners.
+Performing a grid search for hyperparameter selection might not be feasible for when there are more than two hyperparameters to tune for.
+In such cases, it is more efficient to use random search or recently developed tools such as Hyperband to identify good hyperparameter configurations [@li_hyperband:_2016].)
 
@@
 **Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1).
-We compared the performance of the models by
-taking the difference of AUROC values from the same datasplit.
-We plotted a histogram that showed the AUROC differences for 100 datasplits and calculated the empirical p-value of the histogram (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for double tail event (e.g., Figure S3).
+We compared the performance of the models pairwise by calculating the difference between AUROC values from the same datasplit (for 100 datasplits).
+We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S3).
 
@@
 **Acknowlegements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network.
-We would like to thank the members of the Schloss lab for their valuable feedback during the formulation of the manuscript.
+We would like to thank the members of the Schloss lab for their valuable feedback.
 Salary support for M.R came from NIH grant 1R01CA215574.
@@
 Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-The predictive performance of random forest model was higher than other ML models though not significantly (p=0.5).
+The predictive performance of random forest model was higher than other ML models,s though not significantly (p=0.5).
 L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another.
@@
 **Figure 3.
-Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel were ranked from highest rank, 1, to lowest rank, 100, for each data-split.
+Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each datasplit.
 The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.
-OTUs that are associated with classifying a subject as being healthy had negative signs and were shown in blue.
-OTUs that are associated with classifying a subject having an SRN had positive signs and were shown in red.
+OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue.
+OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red.
 Some of the same OTUs were identified as important in all of the linear models.
@@
 The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation.
+The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here.
 The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, salmon were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only.
@@
 **Figure S1.
-Hyperparameter setting performances for linear models.** (A) L2 logistic regression, (B) L1 SVM with linear kernel, and (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model.
+Hyperparameter setting performances for linear models.** (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model.
 The stars represent the highest performing hyperparameter setting for each model.
@@
 **Figure S2.
-Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) Random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.
+Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.
 The stars represent the highest performing hyperparameter setting for the models.
@@
 **Figure S4.
-Classification performance of ML models across cross validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
+Classification performance of ML models across cross validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
 The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
@@
 \rowcolor{lightgray}
-\textbf{Practice} & \textbf{Good} & \textbf{Better} & \textbf{Best} \\ \hline
+\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
 
-\makecell[l]{Problem \\ definition} & \makecell[l]{Have we clearly stated \\ the ML task?
-\\ Do we have a priori hypotheses?
-\\Do we know the predictions \\ a domain expert would \\ make manually?} & \makecell[l]{Do we know the motivation \\ for solving the problem?
-\\ How much interpretability \\ does the problem need?
-} & \makecell[l]{Do we know the correlated \\ features?
-} \\ \hline
+\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application.
+\\ (e.g.
+input data can be \\ used for diagnosis \\ but not prognosis)} & \makecell[l]{Data are generated according \\ to study design and appropriate \\ for intended application} & \makecell[l]{Data are available \\ to predict intended \\ outcomes and will be \\ available in the future } \\ \hline
 
-\makecell[l]{Classification \\ algorithm} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem?
-\\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretability \\ does the problem need?
-\\ How much each \\ candidate algorithm \\ can provide?} \\ \hline
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance} & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to} & \makecell[l]{Model tested on \\ multiple cohorts \\ with different \\ class balances} \\ \hline
 
-\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Did we do \\ cross-validation?} & \makecell[l]{Do we have a held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\  datasets?} \\ \hline
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method} & \makecell[l]{Model choice is justified \\ for intended application} & \makecell[l]{Different modeling \\ choices tested} \\ \hline
 
-\makecell[l]{Hyperparameter \\ selection} & \makecell[l]{Do we know the different \\ hyperparameters each model \\ can use and why?} & \makecell[l]{Did we use historically \\ effective hyperparameters?
-} & \makecell[l]{Did we search the \\ full grid space and \\optimized our model?
-} \\ \hline
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning} & \makecell[l]{Tested across multiple \\ hyperparameter settings} & \makecell[l]{Cross-validation performed \\ on training set \\ with hyperparameter \\ grid search} \\ \hline
 
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Have we chosen an appropriate \\ metric to evaluate \\ predictive performance?} & \makecell[l]{Have we reported \\ the predictive performance \\ on a held-out test data?} & \makecell[l]{Have we provided an average \\ predictive performance \\ of many model runs?
-} \\ \hline
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the data \\ used to train the model} & \makecell[l]{Performance reported on \\ held-out test data} & \makecell[l]{Performance reported on \\ multiple held-out test data } \\ \hline
 
-\makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it?
-\\ Have we checked for \\ the effect of correlated \\ features?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
-} \\ \hline
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single metric that is \\ not appropriate for intended \\ application  (e.g.
+when predicting \\ rare outcome, accuracy metric is \\ not reliable)} & \makecell[l]{Reported multiple metrics} & \makecell[l]{Reported multiple metrics \\ with confidence intervals} \\ \hline
 
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested} \\ \hline
+
+\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported} & \makecell[l]{Model described \\ (e.g.
+feature weights)} & \makecell[l]{How to use the model \\ is explained} \\ \hline
 \end{tabular}

### 2019-10-14 submission/manuscript.tex
Jenna's edits 3rd round

@@
 yielded a model that was inherently interpretable. Our analysis
-highlights the importance of choosing an ML approach should be chosen
-based on the goal of the study, as the choice will inform expectations
-of performance and interpretability.
+highlights the importance of choosing an ML approach based on the goal
+of the study, as the choice will inform expectations of performance and
+interpretability.
 
@@
 
-\textbf{Model selection and pipeline construction} We established a
+\textbf{Model selection and pipeline construction}. We established a
 reusable ML pipeline for model selection and evaluation, focusing on
@@
 classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
-random forest model had a median AUROC value of 0.695 {[}IQR
-0.650-0.739{]} which was higher than the other models for detecting
-SRNs. However, we determined that the random forest model performance
-was not statistically significantly higher (p = 0.5) when we
-investigated the distribution of differences between random forest and
-L2-regularized logistic regression AUROC values for each data-split
-(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
-L2-regularized SVM with linear and radial basis function kernel AUROC
-values were not significantly different from one another and had median
-AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
-0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
-0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
-decision tree had significantly lower AUROC values than the other ML
-models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
-0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
-results demonstrate that the most complex model (XGBoost) did not have
-the best performance and that the most interpretable models
+predictive performance of random forest model was higher than other ML
+models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
+significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
+regression, XGBoost, L2-regularized SVM with linear and radial basis
+function kernel AUROC values were not significantly different from one
+another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
+0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
+{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
+kernel and decision tree had significantly lower AUROC values than the
+other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
+0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
+these results demonstrate that the most complex model (XGBoost) did not
+have the best performance and that the most interpretable models
 (L2-regularized logistic regression and L2-regularized SVM with linear
@@
 dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
-in the random forest model, the AUROC only dropped from 0.695 to 0.680.
-Although permutation analysis allowed us to gauge the importance of an
-OTU, the analysis was post hoc (i.e., done using the test data).
+in the random forest model, the AUROC only dropped from 0.695 to 0.680
+which indicated high degree of collinearity in the dataset. Permutation
+analysis allowed us to gauge the importance of each OTU in non-linear
+models and partially account for collinearity by grouping correlated
+OTUs to determine their impact as a group.
 
@@
 deploying microbiome-based models to make clinical diagnoses or
-predictions is a significantly more challenging and distinct
-undertaking. For example, we currently lack standardized methods to
-collect patient samples, generate sequence data, and report clinical
-data. We are also challenged by the practical constraints of OTU-based
+predictions is a significantly more challenging and distinct undertaking
+(41). For example, we currently lack standardized methods to collect
+patient samples, generate sequence data, and report clinical data. We
+are also challenged by the practical constraints of OTU-based
 approaches. The de novo algorithms commonly in use are slow, require
 considerable memory, and result in different OTU assignments as new data
-are added (41). Finally, we also need independent validation cohorts to
+are added (42). Finally, we also need independent validation cohorts to
 test the performance of a diagnostic model. To realize the potential for
@@
 the SRN class included patients with advanced adenomas or carcinomas
-(42). Patients with an adenoma greater than 1 cm, more than three
+(43). Patients with an adenoma greater than 1 cm, more than three
 adenomas of any size, or an adenoma with villous histology were
@@
 Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
-generated by Sze et al (43), who processed the 16S rRNA sequences in
+generated by Sze et al (44), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (41, 44, 45);
+to OTUs at 97\% similarity using the OptiClust algorithm (42, 45, 46);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 search was needed to avoid large variation in prediction performance.
-Recently developed tools such as Hyperband help researchers with
-hyperparameter selection that can be incorporated to microbiome studies
-(46). For L2-regularized logistic regression, L1 and L2-regularized SVM
-with linear and radial basis function kernels, we tuned the cost
+For L2-regularized logistic regression, L1 and L2-regularized SVM with
+linear and radial basis function kernels, we tuned the cost
 hyperparameter, which controls the regularization strength, where
@@
 the learning rate and the fraction of samples used for fitting the
-individual base learners.
+individual base learners. Performing a grid search for hyperparameter
+selection might not be feasible for when there are more than two
+hyperparameters to tune for. In such cases, it is more efficient to use
+random search or recently developed tools such as Hyperband to identify
+good hyperparameter configurations (47).)
 
@@
 the tidyverse package (v.1.2.1). We compared the performance of the
-models by taking the difference of AUROC values from the same datasplit.
-We plotted a histogram that showed the AUROC differences for 100
-datasplits and calculated the empirical p-value of the histogram (2 x
+models pairwise by calculating the difference between AUROC values from
+the same datasplit (for 100 datasplits). We determined if the models
+were significantly different by calculating the empirical p-value (2 x
 min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
-0) for double tail event (e.g., Figure S3).
+0) for the double tail event (e.g., Figure S3).
 
@@
 Lakes-New England Early Detection Research Network. We would like to
-thank the members of the Schloss lab for their valuable feedback during
-the formulation of the manuscript. Salary support for M.R came from NIH
-grant 1R01CA215574. Salary support for P.D.S. came from NIH grants
-P30DK034933 and 1R01CA215574.
+thank the members of the Schloss lab for their valuable feedback. Salary
+support for M.R came from NIH grant 1R01CA215574. Salary support for
+P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.
 
@@
 horizontal line at 0.50) for all the ML models. The predictive
-performance of random forest model was higher than other ML models
+performance of random forest model was higher than other ML models,s
 though not significantly (p=0.5). L2-regularized logistic regression,
@@
 \textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
-absolute feature weights of (A) L2-regularized logistic regression, (B)
-L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with
-linear kernel were ranked from highest rank, 1, to lowest rank, 100, for
-each data-split. The feature ranks of the 20 highest ranked OTUs based
-on their median ranks (median shown in black) are reported here. OTUs
-that are associated with classifying a subject as being healthy had
-negative signs and were shown in blue. OTUs that are associated with
-classifying a subject having an SRN had positive signs and were shown in
-red. Some of the same OTUs were identified as important in all of the
-linear models. \newpage
+absolute feature weights of (A) L1-regularized SVM with linear kernel,
+(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
+logistic regression, were ranked from highest rank, 1, to lowest rank,
+100, for each datasplit. The feature ranks of the 20 highest ranked OTUs
+based on their median ranks (median shown in black) are reported here.
+OTUs that were associated with classifying a subject as being healthy
+had negative signs and were shown in blue. OTUs that were associated
+with classifying a subject having an SRN had positive signs and were
+shown in red. Some of the same OTUs were identified as important in all
+of the linear models. \newpage
 \includegraphics{Figure_4.png}
@@
 show the IQR range and median of the base testing AUROC without any
-permutation. The colors of the box plots represent the OTUs that were
-shared among the different models; yellow were OTUs that were shared
-among all the non-linear models, salmon were OTUs that were shared among
-the tree-based models, green were the OTUs shared among SVM with radial
-basis kernel, decision tree and XGBoost, pink were the OTUs shared among
-SVM with radial basis kernel and XGBoost only, red were the OTUs shared
-among random forest and XGBoost only and blue were the OTUs shared among
-decision tree and random forest only. For all of the tree-based models,
-a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
-predictive performance. \newpage
+permutation. The 20 OTUs that caused the largest decrease in the AUROC
+when permuted are reported here. The colors of the box plots represent
+the OTUs that were shared among the different models; yellow were OTUs
+that were shared among all the non-linear models, salmon were OTUs that
+were shared among the tree-based models, green were the OTUs shared
+among SVM with radial basis kernel, decision tree and XGBoost, pink were
+the OTUs shared among SVM with radial basis kernel and XGBoost only, red
+were the OTUs shared among random forest and XGBoost only and blue were
+the OTUs shared among decision tree and random forest only. For all of
+the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
+had the largest impact on predictive performance. \newpage
 \includegraphics{Figure_5.png}
@@
 \textbf{Figure S1. Hyperparameter setting performances for linear
-models.} (A) L2 logistic regression, (B) L1 SVM with linear kernel, and
-(C) L2 SVM with linear kernel mean cross-validation AUROC values when
-different hyperparameters were used in training the model. The stars
-represent the highest performing hyperparameter setting for each model.
-\newpage
+models.} (A) L2-regularized logistic regression, (B) L1-regularized SVM
+with linear kernel, and (C) L2-regularized SVM with linear kernel mean
+cross-validation AUROC values when different hyperparameters were used
+in training the model. The stars represent the highest performing
+hyperparameter setting for each model. \newpage
 \includegraphics[height=30cm, width=15cm]{Figure_S2.png}
@@
 \textbf{Figure S2. Hyperparameter setting performances for non-linear
-models.} (A) Decision tree, (B) Random forest, (C) SVM with radial basis
+models.} (A) Decision tree, (B) random forest, (C) SVM with radial basis
 kernel, and (D) XGBoost mean cross-validation AUROC values when
@@
 validation when trained on a subset of the dataset.} (A) L2-regularized
-logistic regression and (B) Random forest models were trained using the
+logistic regression and (B) random forest models were trained using the
 original study design with 490 subjects and subsets of it with 15, 30,
@@
 \rowcolor{lightgray}
-\textbf{Practice} & \textbf{Good} & \textbf{Better} & \textbf{Best} \\ \hline
-
-\makecell[l]{Problem \\ definition} & \makecell[l]{Have we clearly stated \\ the ML task?
-\\ Do we have a priori hypotheses?
-\\Do we know the predictions \\ a domain expert would \\ make manually?} & \makecell[l]{Do we know the motivation \\ for solving the problem?
-\\ How much interpretability \\ does the problem need?
-} & \makecell[l]{Do we know the correlated \\ features?
-} \\ \hline
-
-\makecell[l]{Classification \\ algorithm} & \makecell[l]{Do we know the \\ candidate algorithms \\ for the ML problem?
-\\} & \makecell[l]{Do we know our \\ computational resources \\ to fully train each model?} & \makecell[l]{How much interpretability \\ does the problem need?
-\\ How much each \\ candidate algorithm \\ can provide?} \\ \hline
-
-\makecell[l]{ML pipeline \\ preparation} & \makecell[l]{Did we do \\ cross-validation?} & \makecell[l]{Do we have a held-out \\ test dataset?} & \makecell[l]{Have we tested our model \\ on many different \\  datasets?} \\ \hline
-
-\makecell[l]{Hyperparameter \\ selection} & \makecell[l]{Do we know the different \\ hyperparameters each model \\ can use and why?} & \makecell[l]{Did we use historically \\ effective hyperparameters?
-} & \makecell[l]{Did we search the \\ full grid space and \\optimized our model?
-} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Have we chosen an appropriate \\ metric to evaluate \\ predictive performance?} & \makecell[l]{Have we reported \\ the predictive performance \\ on a held-out test data?} & \makecell[l]{Have we provided an average \\ predictive performance \\ of many model runs?
-} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{Do we know if \\ our model is \\ interpretable?} & \makecell[l]{If the model is not \\ interpretable, do we know \\how to  explain it?
-\\ Have we checked for \\ the effect of correlated \\ features?} & \makecell[l]{Have we generated new \\ hypotheses based on \\ model interpretation \\ to test model results?
-} \\ \hline
-
+\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
+
+\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application.
+\\ (e.g.
+input data can be \\ used for diagnosis \\ but not prognosis)} & \makecell[l]{Data are generated according \\ to study design and appropriate \\ for intended application} & \makecell[l]{Data are available \\ to predict intended \\ outcomes and will be \\ available in the future } \\ \hline
+
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance} & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to} & \makecell[l]{Model tested on \\ multiple cohorts \\ with different \\ class balances} \\ \hline
+
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method} & \makecell[l]{Model choice is justified \\ for intended application} & \makecell[l]{Different modeling \\ choices tested} \\ \hline
+
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning} & \makecell[l]{Tested across multiple \\ hyperparameter settings} & \makecell[l]{Cross-validation performed \\ on training set \\ with hyperparameter \\ grid search} \\ \hline
+
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the data \\ used to train the model} & \makecell[l]{Performance reported on \\ held-out test data} & \makecell[l]{Performance reported on \\ multiple held-out test data } \\ \hline
+
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single metric that is \\ not appropriate for intended \\ application  (e.g.
+when predicting \\ rare outcome, accuracy metric is \\ not reliable)} & \makecell[l]{Reported multiple metrics} & \makecell[l]{Reported multiple metrics \\ with confidence intervals} \\ \hline
+
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested} \\ \hline
+
+\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported} & \makecell[l]{Model described \\ (e.g.
+feature weights)} & \makecell[l]{How to use the model \\ is explained} \\ \hline
 \end{tabular}\newpage
@@
 
+\hypertarget{ref-wiens_no_2019}{}
+41. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
+\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
+\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
+\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
+2019. Do no harm: A roadmap for responsible machine learning for health
+care. Nat Med \textbf{25}:1337--1340.
+doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.
+
 \hypertarget{ref-westcott_opticlust_2017}{}
-41. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-redwood_stool_2016}{}
-42. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+43. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
 \textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
@@
 \hypertarget{ref-sze_leveraging_2018}{}
-43. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+44. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-44. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+45. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-45. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+46. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-46. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+47. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2019-10-15 submission/manuscript.Rmd
edit typos and table S1

@@
 To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost).
-The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and not immidiately interpretable.
-Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and yielded a model that was inherently interpretable.
+The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739] but was slow to train (83.2 h) and not immediately interpretable.
+Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was inherently interpretable.
 Our analysis highlights the importance of choosing an ML approach based on the goal of the study, as the choice will inform expectations of performance and interpretability.
@@
 Instead, subsets of the microbiome account for differences in health.
-Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population seperately [@segata_metagenomic_2011].
-Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease.
+Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population separately [@segata_metagenomic_2011].
+Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease.
 
@@
 Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data.
-We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
+We then calculated how much the predictive performance of the model (i.e., testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
 We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4].
@@
 Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-The predictive performance of random forest model was higher than other ML models,s though not significantly (p=0.5).
+The predictive performance of random forest model was higher than other ML models, though not significantly (p=0.5).
 L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another.
@@
 
-\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application.
+\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application (e.g.
+input \\ data provided only from \\ patients with carcinomas but \\ model is expected to predict \\ advanced adenomas.)} & \makecell[l]{Input data are generated \\ according to study design and \\ they are appropriate for intended \\ application.} & \makecell[l]{Input data are available \\ to predict intended outcomes \\ and will be available \\ in the future (e.g.
+same OTU \\ assignments for new fecal \\ samples)} \\ \hline
+
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance.
 \\ (e.g.
-input data can be \\ used for diagnosis \\ but not prognosis)} & \makecell[l]{Data are generated according \\ to study design and appropriate \\ for intended application} & \makecell[l]{Data are available \\ to predict intended \\ outcomes and will be \\ available in the future } \\ \hline
+test data only includes \\ patients with carcinomas but \\ model is trained to predict \\ healthy colons as well.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with different \\ class balances.} \\ \hline
 
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance} & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to} & \makecell[l]{Model tested on \\ multiple cohorts \\ with different \\ class balances} \\ \hline
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling \\ choices tested.} \\ \hline
 
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method} & \makecell[l]{Model choice is justified \\ for intended application} & \makecell[l]{Different modeling \\ choices tested} \\ \hline
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Model is tested across multiple \\ hyperparameter settings.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
 
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning} & \makecell[l]{Tested across multiple \\ hyperparameter settings} & \makecell[l]{Cross-validation performed \\ on training set \\ with hyperparameter \\ grid search} \\ \hline
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test data.} \\ \hline
 
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the data \\ used to train the model} & \makecell[l]{Performance reported on \\ held-out test data} & \makecell[l]{Performance reported on \\ multiple held-out test data } \\ \hline
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single performance \\ metric that is not appropriate for \\ intended application  (e.g.
+when \\ predicting rare outcome, \\ accuracy metric is not \\ reliable).} & \makecell[l]{Reported a single performance \\ metric that is appropriate \\ for intended application with \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
 
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single metric that is \\ not appropriate for intended \\ application  (e.g.
-when predicting \\ rare outcome, accuracy metric is \\ not reliable)} & \makecell[l]{Reported multiple metrics} & \makecell[l]{Reported multiple metrics \\ with confidence intervals} \\ \hline
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested.} \\ \hline
 
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested} \\ \hline
-
-\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported} & \makecell[l]{Model described \\ (e.g.
-feature weights)} & \makecell[l]{How to use the model \\ is explained} \\ \hline
+\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported.} & \makecell[l]{Model described \\ (e.g.
+feature weights).} & \makecell[l]{How to use the model \\ is explained.} \\ \hline
 \end{tabular}

### 2019-10-15 submission/manuscript.tex
edit typos and table S1

@@
 forest model performed best at detecting SRNs with an AUROC of 0.695
-{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immidiately
+{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
 interpretable. Despite its simplicity, L2-regularized logistic
 regression followed random forest in predictive performance with an
-AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and
-yielded a model that was inherently interpretable. Our analysis
-highlights the importance of choosing an ML approach based on the goal
-of the study, as the choice will inform expectations of performance and
-interpretability.
+AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
+inherently interpretable. Our analysis highlights the importance of
+choosing an ML approach based on the goal of the study, as the choice
+will inform expectations of performance and interpretability.
 
@@
 variation in the human microbiome and typically consider the protective
-or risk effects of each bacterial population seperately (1). Recently,
+or risk effects of each bacterial population separately (1). Recently,
 machine learning (ML) models have grown in popularity among microbiome
-researchers as our ability to to sample large numbers of individuals has
+researchers as our ability to sample large numbers of individuals has
 grown; such models can effectively account for the interpersonal
@@
 We then calculated how much the predictive performance of the model
-(i.e, testing AUROC values) decreased when each OTU or group of OTUs was
-randomly permuted. We ranked the OTUs based on how much the median
+(i.e., testing AUROC values) decreased when each OTU or group of OTUs
+was randomly permuted. We ranked the OTUs based on how much the median
 testing AUROC decreased when it was permuted; the OTU with the largest
@@
 horizontal line at 0.50) for all the ML models. The predictive
-performance of random forest model was higher than other ML models,s
+performance of random forest model was higher than other ML models,
 though not significantly (p=0.5). L2-regularized logistic regression,
@@
 
-\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application.
+\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application (e.g.
+input \\ data provided only from \\ patients with carcinomas but \\ model is expected to predict \\ advanced adenomas.)} & \makecell[l]{Input data are generated \\ according to study design and \\ they are appropriate for intended \\ application.} & \makecell[l]{Input data are available \\ to predict intended outcomes \\ and will be available \\ in the future (e.g.
+same OTU \\ assignments for new fecal \\ samples)} \\ \hline
+
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance.
 \\ (e.g.
-input data can be \\ used for diagnosis \\ but not prognosis)} & \makecell[l]{Data are generated according \\ to study design and appropriate \\ for intended application} & \makecell[l]{Data are available \\ to predict intended \\ outcomes and will be \\ available in the future } \\ \hline
-
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance} & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to} & \makecell[l]{Model tested on \\ multiple cohorts \\ with different \\ class balances} \\ \hline
-
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method} & \makecell[l]{Model choice is justified \\ for intended application} & \makecell[l]{Different modeling \\ choices tested} \\ \hline
-
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning} & \makecell[l]{Tested across multiple \\ hyperparameter settings} & \makecell[l]{Cross-validation performed \\ on training set \\ with hyperparameter \\ grid search} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the data \\ used to train the model} & \makecell[l]{Performance reported on \\ held-out test data} & \makecell[l]{Performance reported on \\ multiple held-out test data } \\ \hline
-
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single metric that is \\ not appropriate for intended \\ application  (e.g.
-when predicting \\ rare outcome, accuracy metric is \\ not reliable)} & \makecell[l]{Reported multiple metrics} & \makecell[l]{Reported multiple metrics \\ with confidence intervals} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested} \\ \hline
-
-\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported} & \makecell[l]{Model described \\ (e.g.
-feature weights)} & \makecell[l]{How to use the model \\ is explained} \\ \hline
+test data only includes \\ patients with carcinomas but \\ model is trained to predict \\ healthy colons as well.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with different \\ class balances.} \\ \hline
+
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling \\ choices tested.} \\ \hline
+
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Model is tested across multiple \\ hyperparameter settings.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
+
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test data.} \\ \hline
+
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single performance \\ metric that is not appropriate for \\ intended application  (e.g.
+when \\ predicting rare outcome, \\ accuracy metric is not \\ reliable).} & \makecell[l]{Reported a single performance \\ metric that is appropriate \\ for intended application with \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
+
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested.} \\ \hline
+
+\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported.} & \makecell[l]{Model described \\ (e.g.
+feature weights).} & \makecell[l]{How to use the model \\ is explained.} \\ \hline
 \end{tabular}\newpage

### 2019-10-15 submission/manuscript.Rmd
edited typos and checked word count

@@
 Performing a grid search for hyperparameter selection might not be feasible for when there are more than two hyperparameters to tune for.
-In such cases, it is more efficient to use random search or recently developed tools such as Hyperband to identify good hyperparameter configurations [@li_hyperband:_2016].)
+In such cases, it is more efficient to use random search or recently developed tools such as Hyperband to identify good hyperparameter configurations [@li_hyperband:_2016].
 
@@
 Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
-The predictive performance of random forest model was higher than other ML models, though not significantly (p=0.5).
+The predictive performance of random forest model was higher than other ML models, though not significantly (p > 0.01).
 L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another.

### 2019-10-15 submission/manuscript.tex
edited typos and checked word count

@@
 random search or recently developed tools such as Hyperband to identify
-good hyperparameter configurations (47).)
+good hyperparameter configurations (47).
 
@@
 performance of random forest model was higher than other ML models,
-though not significantly (p=0.5). L2-regularized logistic regression,
-XGBoost, L2-regularized SVM with linear and radial basis function kernel
-performances were not significantly different from one another. The
-boxplot shows quartiles at the box ends and the median as the horizontal
-line in the box. The whiskers show the farthest points that were not
-outliers. Outliers were defined as those data points that are not within
-1.5 times the interquartile ranges. \newpage
+though not significantly (p \textgreater{} 0.01). L2-regularized
+logistic regression, XGBoost, L2-regularized SVM with linear and radial
+basis function kernel performances were not significantly different from
+one another. The boxplot shows quartiles at the box ends and the median
+as the horizontal line in the box. The whiskers show the farthest points
+that were not outliers. Outliers were defined as those data points that
+are not within 1.5 times the interquartile ranges. \newpage
 \includegraphics[height=17cm, width=12cm]{Figure_3.png}

### 2019-10-16 submission/manuscript.Rmd
changes after table edtits from jenna

@@
 ---
-title: '**Effective application of machine learning to microbiome based classification problems**'
+title: '**Effective application of machine learning to microbiome-based classification problems**'
 csl: mbio.csl
@@
 
-2\. Department of Electrical Engineering and Computer Science, University or Michigan, Ann Arbor, MI 48109
+2\. Department of Electrical Engineering and Computer Science, University of Michigan, Ann Arbor, MI 48109
 
@@
 
-Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer.
+Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in the diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer.
 Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question.
@@
 To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
-We developed a reusable open-source pipeline to train, validate and interpret the models.
+We developed a reusable open-source pipeline to train, validate, and interpret the models.
 To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost).
@@
 Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies.
-However, estimated performance associated with these ML models is likely overoptimistic.
+However, the estimated performance associated with these ML models is likely over-optimistic.
 Moreover, there is a trend towards using black box models without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease.
-This work represents a step towards developing more reproducble ML practices in applying ML to microbiome research.
+This work represents a step towards developing more reproducible ML practices in applying ML to microbiome research.
 We implement a rigorous pipeline and emphasize the importance of selecting ML models that reflect the goal of the study.
@@
 These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances.
-Though, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to further improve reproducibility and minimize overestimating for model performance.
+Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.
 
@@
 This has resulted in a trend towards using non-linear models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019].
-Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability.
+Although in some cases, complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability.
 Such models require post hoc explanations to quantify the importance of each feature in making predictions.
@@
 For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may (but not always) emphasize predictive performance.
-Nonetheless, it is important to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
+Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
 It is important for researchers to justify their choice of modeling approach.
@@
 We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
-We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and gradient boosted trees.
+We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest, and gradient boosted trees.
 We compared their predictive performance, interpretability, and training time.
-To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classify individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
+To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
 This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
@@
 For example, our full dataset included 490 individuals.
-Of these, 261 had normal colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%).
-A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%).
-The training data were used to build and select the models and the test set was used for evaluating the model.
+Of these, 261 had healthy colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%).
+A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals, of which 52 had an SRN (54%).
+The training data were used to build and select the models, and the test set was used for evaluating the model.
 
@@
 L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero).
-To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including a decision tree, a random forest, and gradient boosted trees (XGBoost) and an SVM with a non-linear kernel.
+To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including, a decision tree, a random forest, and gradient boosted trees (XGBoost) and an SVM with a non-linear kernel.
 
@@
 The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls.
-An AUROC value of 0.5 indicates that model's predictions are no different than random.
+An AUROC value of 0.5 indicates that the model's predictions are no different than random.
 To select hyperparameters, we performed a grid search for hyperparameter settings when training the models.
@@
 
-__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2].
+__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2].
 The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5) (Figure S3).
@@
 If we had only done a single split, then there is a risk that we could have gotten lucky or unlucky in estimating model performance.
-For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(format(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3), nsmall=3))`.
+For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`, whereas the highest was `r paste(format(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3), nsmall=3))`.
 These results showed that depending on the data-split, the testing performance can vary [Figure 2].
@@
 To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4].
-The variation in cross-validation performance within both models at lower sample sizes was larger than when the full collection of samples was used to train and validate the models.
+The variation in cross-validation performance within both models at smaller sample sizes was larger than when the full collection of samples was used to train and validate the models.
 Because of the high dimensionality of the microbiome data (6920 OTUs), large sample sizes can lead to better models.
@@
 We also reviewed the signs of feature weights to determine whether an OTU was associated with classifying a subject as being healthy or having an SRN.
-It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g.
-OTUs 50, 426, 609, 822, 1239).
+It was encouraging that many of the highest-ranked OTUs were shared across these three models (e.g., OTUs 50, 426, 609, 822, 1239).
 The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU.
@@
 
-Second, to analyze non-linear models we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134].
+Second, to analyze non-linear models, we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134].
 Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
@@
 Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models.
-Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTUs 2, 12, 361, 477).
+Similarly, the random forest and XGBoost models shared four of the most important OTUs (OTUs 2, 12, 361, 477).
 Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance.
 For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`.
-In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))` which indicated high degree of collinearity in the dataset.
+In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`, which indicated high degree of collinearity in the dataset.
 Permutation analysis allowed us to gauge the importance of each OTU in non-linear models and partially account for collinearity by grouping correlated OTUs to determine their impact as a group.
@@
 To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S5].
-When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g.
-OTU 609, 822, 1239) were deemed important by both interpretation methods.
+When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods.
 Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods.
@@
 __The computational efficiency of each ML model.__ We compared the training times of the seven ML models.
-We did not take advantage of parallelization in training nor was the code optimized for speed.
+We did not take advantage of parallelization in training, nor was the code optimized for speed.
 Still, as expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations.
@@
 If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification.
-However, the application of ML in microbiome studies is still in its infancy and the field needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.
+However, the application of ML in microbiome studies is still in its infancy, and the field needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.
 
-To address these needs, we developed an open-sourced framework to ML models.
+To address these needs, we developed an open-sourced framework for ML models.
 Using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized.
 In terms of predictive performance, the random forest model had the best AUROC compared to the other six models.
-However, the second-best model was L2-regularized logistic regression with a median AUROC difference less than 0.015 compared to random forest.
+However, the second-best model was L2-regularized logistic regression with a median AUROC difference of less than 0.015 compared to random forest.
 While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model while L2-regularized logistic regression, a linear model, is easily interpreted according to the feature weights.
+In terms of interpretability, random forest is a non-linear ML model, while L2-regularized logistic regression, a linear model, is easily interpreted according to the feature weights.
 Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.
@@
 In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance.
-Human-associated microbial communities have complex correlation structures which create collinearity in the datasets.
+Human-associated microbial communities have complex correlation structures that create collinearity in the datasets.
 This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013].
@@
 When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group.
-We grouped OTUs that had a perfect correlation with each other, however, we could reduce the correlation threshold to further investigate the relationships among correlated features.
+We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features.
 It is important to know the correlation structures of the data to avoid misinterpreting the models.
 This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome.
-To identify the true underlying microbial factors of a disease, it is crucial to follow up any correlation analyses with further hypothesis testing and experimentation for biological validation.
+To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.
 
@@
 For example, we did not evaluate multicategory classification methods to predict non-binary outcomes.
-We could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories).
-We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs.
+We could have trained models to differentiate between people with healthy colons and those with adenomas or carcinomas (k=3 categories).
+We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with healthy colons and those with SRNs.
 Furthermore, as the number of classes increases, more samples are required for each category to train an accurate model.
@@
 Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks.
-Deep learning methods hold promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which result in overfitting.
+Deep learning methods hold promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], but microbiome datasets often suffer from having many features and small sample sizes, which result in overfitting.
 
@@
 Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy.
-Participants were recruited from: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
+Participants were recruited from Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA).
 Patients’ colonic health was visually assessed by colonoscopy with bowel preparation and tissue histopathology of all resected lesions.
-We assigned patients into two classes: those with normal colons and those with screen relevant neoplasias (SRNs).
-The normal class included patients with normal colons or non-advanced adenomas whereas the SRN class included patients with advanced adenomas or carcinomas [@redwood_stool_2016].
+We assigned patients into two classes: those with healthy colons and those with screen relevant neoplasias (SRNs).
+The healthy class included patients with healthy colons or non-advanced adenomas, whereas the SRN class included patients with advanced adenomas or carcinomas [@redwood_stool_2016].
 Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas.
@@
 Of the 198 adenomas, 109 were identified as advanced adenomas.
-Together 261 patients were classified as normal and 229 patients were classified as having a SRN.
+Together 261 patients were classified as healthy and 229 patients were classified as having an SRN.
 
@@
 There were 6920 OTUs.
-OTU abundances were subsampled to the size of the smallest sample and normalized across samples such that the highest abundance of each OTU would be 1 and lowest would be 0.
+OTU abundances were subsampled to the size of the smallest sample and normalized across samples such that the highest abundance of each OTU would be 1, and the lowest would be 0.
 
@@
 For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization.
-For SVM with radial basis function kernel, we also tuned the sigma hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
+For SVM with radial basis function kernel, we also tuned the sigma hyperparameter, which determines the reach of a single training instance where, for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
 For the decision tree model, we tuned the depth of the tree where the deeper the tree, the more splits it has.
@@
 **Permutation importance workflow.** We calculated a Spearman's rank-order correlation matrix and defined correlated OTUs as having perfect correlation (correlation coefficient = 1 and p < 0.01).
-OTUs without a perfect correlation to each other were permuted individually whereas correlated ones were grouped together and permuted at the same time.
+OTUs without a perfect correlation to each other were permuted individually, whereas correlated ones were grouped together and permuted at the same time.
 
@@
 
-**Acknowlegements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network.
+**Acknowledgements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network.
 We would like to thank the members of the Schloss lab for their valuable feedback.
@@
 **Figure 2.
-Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models.
+Generalization and classification performance of ML models using AUROC values of all cross-validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by a horizontal line at 0.50) for all the ML models.
 The predictive performance of random forest model was higher than other ML models, though not significantly (p > 0.01).
@@
 **Figure S3.
-Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred datasplits.** This histogram shows number of datasplits in each bin.
+Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred datasplits.** This histogram shows the number of datasplits in each bin.
 The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25.
@@
 **Figure S4.
-Classification performance of ML models across cross validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
-The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models, but included the ranges observed with the more complete datasets.
+Classification performance of ML models across cross-validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
+The range among the cross-validation AUROC values within both models at smaller sample sizes were much larger than when the full collection of samples was used to train and validate the models but included the ranges observed with the more complete datasets.
 \newpage
@@
 
-\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application (e.g.
-input \\ data provided only from \\ patients with carcinomas but \\ model is expected to predict \\ advanced adenomas.)} & \makecell[l]{Input data are generated \\ according to study design and \\ they are appropriate for intended \\ application.} & \makecell[l]{Input data are available \\ to predict intended outcomes \\ and will be available \\ in the future (e.g.
-same OTU \\ assignments for new fecal \\ samples)} \\ \hline
+\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
 
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance.
-\\ (e.g.
-test data only includes \\ patients with carcinomas but \\ model is trained to predict \\ healthy colons as well.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with different \\ class balances.} \\ \hline
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
 
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling \\ choices tested.} \\ \hline
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
 
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Model is tested across multiple \\ hyperparameter settings.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
 
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test data.} \\ \hline
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
 
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single performance \\ metric that is not appropriate for \\ intended application  (e.g.
-when \\ predicting rare outcome, \\ accuracy metric is not \\ reliable).} & \makecell[l]{Reported a single performance \\ metric that is appropriate \\ for intended application with \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g.
+when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
 
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested.} \\ \hline
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
 
-\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported.} & \makecell[l]{Model described \\ (e.g.
-feature weights).} & \makecell[l]{How to use the model \\ is explained.} \\ \hline
 \end{tabular}

### 2019-10-16 submission/manuscript.tex
changes after table edtits from jenna

@@
 \hypersetup{unicode=true,
-            pdftitle={Effective application of machine learning to microbiome based classification problems},
+            pdftitle={Effective application of machine learning to microbiome-based classification problems},
             pdfborder={0 0 0},
@@
 
-  \title{\textbf{Effective application of machine learning to microbiome based
+  \title{\textbf{Effective application of machine learning to microbiome-based
 classification problems}}
@@
 2. Department of Electrical Engineering and Computer Science, University
-or Michigan, Ann Arbor, MI 48109
+of Michigan, Ann Arbor, MI 48109
 
@@
 Machine learning (ML) modeling of the human microbiome has the potential
-to identify microbial biomarkers and aid in diagnosis of many diseases
-such as inflammatory bowel disease, diabetes, and colorectal cancer.
-Progress has been made towards developing ML models that predict health
-outcomes using bacterial abundances, but inconsistent adoption of
+to identify microbial biomarkers and aid in the diagnosis of many
+diseases such as inflammatory bowel disease, diabetes, and colorectal
+cancer. Progress has been made towards developing ML models that predict
+health outcomes using bacterial abundances, but inconsistent adoption of
 training and evaluation methods call the validity of these models into
@@
 relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
-We developed a reusable open-source pipeline to train, validate and
+We developed a reusable open-source pipeline to train, validate, and
 interpret the models. To show the effect of model selection, we assessed
@@
 Prediction of health outcomes using machine learning (ML) is rapidly
-being adopted in microbiome studies. However, estimated performance
-associated with these ML models is likely overoptimistic. Moreover,
+being adopted in microbiome studies. However, the estimated performance
+associated with these ML models is likely over-optimistic. Moreover,
 there is a trend towards using black box models without a discussion of
@@
 microbial biomarkers of disease. This work represents a step towards
-developing more reproducble ML practices in applying ML to microbiome
+developing more reproducible ML practices in applying ML to microbiome
 research. We implement a rigorous pipeline and emphasize the importance
@@
 cross-validation; and unreported variation between cross-validation and
-testing performances. Though, the microbiome field is making progress to
+testing performances. Though the microbiome field is making progress to
 avoid some of these pitfalls including validating their models on
 independent datasets (8, 19, 20) and introducing ways to better use ML
-tools (21--24), more work is needed to further improve reproducibility
+tools (21--24), more work is needed to improve reproducibility further
 and minimize overestimating for model performance.
@@
 25--27) over simpler models such as logistic regression or other linear
-models (19, 23, 28). Although in some cases complex models may capture
+models (19, 23, 28). Although in some cases, complex models may capture
 important non-linear relationships and therefore yield better
@@
 (but not always) emphasize predictive performance. Nonetheless, it is
-important to understand that the benefit of more complex, less
+essential to understand that the benefit of more complex, less
 interpretable models may be minimal (29, 30). It is important for
@@
 also trained four non-linear models: SVM with radial basis function
-kernel, a decision tree, random forest and gradient boosted trees. We
+kernel, a decision tree, random forest, and gradient boosted trees. We
 compared their predictive performance, interpretability, and training
@@
 our pipeline, we used data from a previously published study that sought
-to classify individuals as having normal colons or colonic lesions based
-on the 16S rRNA gene sequences collected from fecal samples (4). This
-dataset was selected because it is a relatively large collection of
+to classify individuals as having healthy colons or colonic lesions
+based on the 16S rRNA gene sequences collected from fecal samples (4).
+This dataset was selected because it is a relatively large collection of
 individuals (N=490) connected to a clinically significant disease where
@@
 performed stratified splits. For example, our full dataset included 490
-individuals. Of these, 261 had normal colons (53\%) and 229 had a screen
-relevant neoplasia (SRN; 46.7\%). A training set included 393
+individuals. Of these, 261 had healthy colons (53\%) and 229 had a
+screen relevant neoplasia (SRN; 46.7\%). A training set included 393
 individuals, of which 209 had an SRN (53\%), while the test set was
-composed of 97 individuals of which 52 had an SRN (54\%). The training
-data were used to build and select the models and the test set was used
+composed of 97 individuals, of which 52 had an SRN (54\%). The training
+data were used to build and select the models, and the test set was used
 for evaluating the model.
@@
 non-linear relationships among features to improve classification, we
-trained tree-based models including a decision tree, a random forest,
+trained tree-based models including, a decision tree, a random forest,
 and gradient boosted trees (XGBoost) and an SVM with a non-linear
@@
 1.0, where the model perfectly distinguishes between cases and controls.
-An AUROC value of 0.5 indicates that model's predictions are no
+An AUROC value of 0.5 indicates that the model's predictions are no
 different than random. To select hyperparameters, we performed a grid
@@
 models.} We evaluated the predictive performance of the seven models to
-classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
-predictive performance of random forest model was higher than other ML
-models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
+classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
+The predictive performance of random forest model was higher than other
+ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
 significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
@@
 estimating model performance. For instance, the lowest AUROC value of
-the random forest model was 0.593 whereas the highest was 0.810. These
+the random forest model was 0.593, whereas the highest was 0.810. These
 results showed that depending on the data-split, the testing performance
@@
 490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
-variation in cross-validation performance within both models at lower
+variation in cross-validation performance within both models at smaller
 sample sizes was larger than when the full collection of samples was
@@
 being healthy or having an SRN. It was encouraging that many of the
-highest ranked OTUs were shared across these three models, (e.g.~OTUs
+highest-ranked OTUs were shared across these three models (e.g., OTUs
 50, 426, 609, 822, 1239). The benefit of this approach was that the
@@
 
-Second, to analyze non-linear models we interpreted the feature
+Second, to analyze non-linear models, we interpreted the feature
 importance using permutation importance (36). Whereas the absolute
@@
 were important in each of the tree-based models. Similarly, the random
-forest and XGBoost models, shared four of the most important OTUs (OTUs
+forest and XGBoost models shared four of the most important OTUs (OTUs
 2, 12, 361, 477). Permutation analysis results also revealed that with
@@
 dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
-in the random forest model, the AUROC only dropped from 0.695 to 0.680
+in the random forest model, the AUROC only dropped from 0.695 to 0.680,
 which indicated high degree of collinearity in the dataset. Permutation
@@
 kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g.~OTU
+permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g., OTU
 609, 822, 1239) were deemed important by both interpretation methods.
@@
 training times of the seven ML models. We did not take advantage of
-parallelization in training nor was the code optimized for speed. Still,
-as expected, the training times increased with the complexity of the
-model and the number of potential hyperparameter combinations. Also, the
-linear models trained faster than non-linear models {[}Figures S1-S2;
-Figure 5{]}.
+parallelization in training, nor was the code optimized for speed.
+Still, as expected, the training times increased with the complexity of
+the model and the number of potential hyperparameter combinations. Also,
+the linear models trained faster than non-linear models {[}Figures
+S1-S2; Figure 5{]}.
 
@@
 communities that are responsible for the classification. However, the
-application of ML in microbiome studies is still in its infancy and the
+application of ML in microbiome studies is still in its infancy, and the
 field needs to develop a better understanding of different ML methods,
@@
 
-To address these needs, we developed an open-sourced framework to ML
+To address these needs, we developed an open-sourced framework for ML
 models. Using this pipeline, we benchmarked seven ML models and showed
@@
 models. However, the second-best model was L2-regularized logistic
-regression with a median AUROC difference less than 0.015 compared to
+regression with a median AUROC difference of less than 0.015 compared to
 random forest. While our implementation of random forest took 83.2 hours
 to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model
+In terms of interpretability, random forest is a non-linear ML model,
 while L2-regularized logistic regression, a linear model, is easily
@@
 Human-associated microbial communities have complex correlation
-structures which create collinearity in the datasets. This can hinder
-our ability to reliably interpret models because the feature weights of
+structures that create collinearity in the datasets. This can hinder our
+ability to reliably interpret models because the feature weights of
 correlated OTUs are influenced by one another (37). To capture all
@@
 correlated OTUs to determine their impact as a group. We grouped OTUs
-that had a perfect correlation with each other, however, we could reduce
+that had a perfect correlation with each other; however, we could reduce
 the correlation threshold to further investigate the relationships among
@@
 another because they come from the same chromosome. To identify the true
-underlying microbial factors of a disease, it is crucial to follow up
+underlying microbial factors of a disease, it is crucial to follow up on
 any correlation analyses with further hypothesis testing and
@@
 multicategory classification methods to predict non-binary outcomes. We
-could have trained models to differentiate between people with normal
+could have trained models to differentiate between people with healthy
 colons and those with adenomas or carcinomas (k=3 categories). We did
 not perform this analysis because the clinically relevant diagnosis
-grouping was between patients with normal colons and those with SRNs.
+grouping was between patients with healthy colons and those with SRNs.
 Furthermore, as the number of classes increases, more samples are
@@
 did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (12, 39, 40) but microbiome datasets often
-suffer from having many features and small sample sizes, which result in
-overfitting.
+learning methods hold promise (12, 39, 40), but microbiome datasets
+often suffer from having many features and small sample sizes, which
+result in overfitting.
 
@@
 samples were provided by adults who were undergoing a scheduled
-screening or surveillance colonoscopy. Participants were recruited from:
+screening or surveillance colonoscopy. Participants were recruited from
 Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
@@
 with bowel preparation and tissue histopathology of all resected
-lesions. We assigned patients into two classes: those with normal colons
-and those with screen relevant neoplasias (SRNs). The normal class
-included patients with normal colons or non-advanced adenomas whereas
-the SRN class included patients with advanced adenomas or carcinomas
-(43). Patients with an adenoma greater than 1 cm, more than three
-adenomas of any size, or an adenoma with villous histology were
+lesions. We assigned patients into two classes: those with healthy
+colons and those with screen relevant neoplasias (SRNs). The healthy
+class included patients with healthy colons or non-advanced adenomas,
+whereas the SRN class included patients with advanced adenomas or
+carcinomas (43). Patients with an adenoma greater than 1 cm, more than
+three adenomas of any size, or an adenoma with villous histology were
 classified as having advanced adenomas. There were 172 patients with
@@
 198 adenomas, 109 were identified as advanced adenomas. Together 261
-patients were classified as normal and 229 patients were classified as
-having a SRN.
+patients were classified as healthy and 229 patients were classified as
+having an SRN.
 
@@
 subsampled to the size of the smallest sample and normalized across
-samples such that the highest abundance of each OTU would be 1 and
+samples such that the highest abundance of each OTU would be 1, and the
 lowest would be 0.
@@
 smaller values specify stronger regularization. For SVM with radial
-basis function kernel, we also tuned the sigma hyperparameter which
-determines the reach of a single training instance where for a high
+basis function kernel, we also tuned the sigma hyperparameter, which
+determines the reach of a single training instance where, for a high
 value of sigma, the SVM decision boundary will be dependent on the
@@
 0.01). OTUs without a perfect correlation to each other were permuted
-individually whereas correlated ones were grouped together and permuted
+individually, whereas correlated ones were grouped together and permuted
 at the same time.
@@
 
-\textbf{Acknowlegements.} We thank all the study participants of Great
+\textbf{Acknowledgements.} We thank all the study participants of Great
 Lakes-New England Early Detection Research Network. We would like to
@@
 \textbf{Figure 2. Generalization and classification performance of ML
-models using AUROC values of all cross validation and testing
+models using AUROC values of all cross-validation and testing
 performances.} The median AUROC for diagnosing individuals with SRN
-using bacterial abundances was higher than chance (depicted by
+using bacterial abundances was higher than chance (depicted by a
 horizontal line at 0.50) for all the ML models. The predictive
@@
 logistic regression and random forest for each of the hundred
-datasplits.} This histogram shows number of datasplits in each bin. The
-percentage of dataplits where the difference between random forest and
-L2-regularized logistic regression AUROC values was higher than or equal
-to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical red
-line highlights the bins where there AUROC difference between the two
-model is 0. The p-value was calculated for a double tail event. \newpage
+datasplits.} This histogram shows the number of datasplits in each bin.
+The percentage of dataplits where the difference between random forest
+and L2-regularized logistic regression AUROC values was higher than or
+equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical
+red line highlights the bins where there AUROC difference between the
+two model is 0. The p-value was calculated for a double tail event.
+\newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
 
-\textbf{Figure S4. Classification performance of ML models across cross
-validation when trained on a subset of the dataset.} (A) L2-regularized
-logistic regression and (B) random forest models were trained using the
-original study design with 490 subjects and subsets of it with 15, 30,
-60, 120, and 245 subjects. The range among the cross-validation AUROC
-values within both models at lower sample sizes were much larger than
-when the full collection of samples was used to train and validate the
-models, but included the ranges observed with the more complete
-datasets. \newpage
+\textbf{Figure S4. Classification performance of ML models across
+cross-validation when trained on a subset of the dataset.} (A)
+L2-regularized logistic regression and (B) random forest models were
+trained using the original study design with 490 subjects and subsets of
+it with 15, 30, 60, 120, and 245 subjects. The range among the
+cross-validation AUROC values within both models at smaller sample sizes
+were much larger than when the full collection of samples was used to
+train and validate the models but included the ranges observed with the
+more complete datasets. \newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
@@
 
-\makecell[l]{Source \\ of data} & \makecell[l]{Input data are not appropriate \\ for intended application (e.g.
-input \\ data provided only from \\ patients with carcinomas but \\ model is expected to predict \\ advanced adenomas.)} & \makecell[l]{Input data are generated \\ according to study design and \\ they are appropriate for intended \\ application.} & \makecell[l]{Input data are available \\ to predict intended outcomes \\ and will be available \\ in the future (e.g.
-same OTU \\ assignments for new fecal \\ samples)} \\ \hline
-
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to \\ remove class imbalance.
-\\ (e.g.
-test data only includes \\ patients with carcinomas but \\ model is trained to predict \\ healthy colons as well.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will eventually \\ be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with different \\ class balances.} \\ \hline
-
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling \\ choices tested.} \\ \hline
-
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Model is tested across multiple \\ hyperparameter settings.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test data.} \\ \hline
-
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported a single performance \\ metric that is not appropriate for \\ intended application  (e.g.
-when \\ predicting rare outcome, \\ accuracy metric is not \\ reliable).} & \makecell[l]{Reported a single performance \\ metric that is appropriate \\ for intended application with \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses on important \\ features are generated \\ and tested.} \\ \hline
-
-\makecell[l]{Model \\ specification} & \makecell[l]{Only predictive performance \\ reported.} & \makecell[l]{Model described \\ (e.g.
-feature weights).} & \makecell[l]{How to use the model \\ is explained.} \\ \hline
+\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
+
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
+
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
+
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
+
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
+
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g.
+when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
+
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
+
 \end{tabular}\newpage

### 2019-10-16 submission/manuscript.Rmd
sent to Pat for final review

@@
 To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
-We developed a reusable open-source pipeline to train, validate, and interpret the models.
+We developed a reusable open-source pipeline to train, validate, and interpret ML models.
 To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost).
@@
 If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data.
-The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
+The largest difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
 These differences are relatively small and give us confidence in our estimate of the generalization performance of the models.
@@
 We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance.
+We used the AUROC metric to evaluate predictive performance as our dataset was highly balanced.
 We repeated this data-split 100 times to measure the possible variation in predictive performance.
@@
 The healthy class included patients with healthy colons or non-advanced adenomas, whereas the SRN class included patients with advanced adenomas or carcinomas [@redwood_stool_2016].
-Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas.
+Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas [@redwood_stool_2016].
 There were 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
@@
 **Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1).
-We compared the performance of the models pairwise by calculating the difference between AUROC values from the same datasplit (for 100 datasplits).
+We compared the performance of the models pairwise by calculating the difference between AUROC values from the same data-split (for 100 data-splits).
 We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S3).
@@
 **Figure 3.
-Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each datasplit.
+Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each data-split.
 The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.
@@
 **Figure S3.
-Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred datasplits.** This histogram shows the number of datasplits in each bin.
+Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.** This histogram shows the number of data-splits in each bin.
 The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25.

### 2019-10-16 submission/manuscript.tex
sent to Pat for final review

@@
 We developed a reusable open-source pipeline to train, validate, and
-interpret the models. To show the effect of model selection, we assessed
+interpret ML models. To show the effect of model selection, we assessed
 the predictive performance, interpretability, and training time of
@@
 could indicate that the models were overfit to the training data. The
-difference in median AUROCs was 0.021 in L1-regularized SVM with linear
-kernel, followed by SVM with radial basis function kernel and decision
-tree with a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
-These differences are relatively small and give us confidence in our
-estimate of the generalization performance of the models.
+largest difference in median AUROCs was 0.021 in L1-regularized SVM with
+linear kernel, followed by SVM with radial basis function kernel and
+decision tree with a difference of 0.007 and 0.006, respectively
+{[}Figure 2{]}. These differences are relatively small and give us
+confidence in our estimate of the generalization performance of the
+models.
 
@@
 (20\% of the data), which we used to evaluate predictive performance. We
-repeated this data-split 100 times to measure the possible variation in
-predictive performance. During training, we tuned the model
-hyperparameters with a repeated five-fold cross-validation. Despite the
-high number of features microbiome datasets typically have, the models
-we built with this pipeline generalized to the held-out test sets.
+used the AUROC metric to evaluate predictive performance as our dataset
+was highly balanced. We repeated this data-split 100 times to measure
+the possible variation in predictive performance. During training, we
+tuned the model hyperparameters with a repeated five-fold
+cross-validation. Despite the high number of features microbiome
+datasets typically have, the models we built with this pipeline
+generalized to the held-out test sets.
 
@@
 three adenomas of any size, or an adenoma with villous histology were
-classified as having advanced adenomas. There were 172 patients with
-normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the
-198 adenomas, 109 were identified as advanced adenomas. Together 261
-patients were classified as healthy and 229 patients were classified as
-having an SRN.
+classified as having advanced adenomas (43). There were 172 patients
+with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
+Of the 198 adenomas, 109 were identified as advanced adenomas. Together
+261 patients were classified as healthy and 229 patients were classified
+as having an SRN.
 
@@
 models pairwise by calculating the difference between AUROC values from
-the same datasplit (for 100 datasplits). We determined if the models
+the same data-split (for 100 data-splits). We determined if the models
 were significantly different by calculating the empirical p-value (2 x
@@
 logistic regression, were ranked from highest rank, 1, to lowest rank,
-100, for each datasplit. The feature ranks of the 20 highest ranked OTUs
-based on their median ranks (median shown in black) are reported here.
-OTUs that were associated with classifying a subject as being healthy
-had negative signs and were shown in blue. OTUs that were associated
-with classifying a subject having an SRN had positive signs and were
-shown in red. Some of the same OTUs were identified as important in all
-of the linear models. \newpage
+100, for each data-split. The feature ranks of the 20 highest ranked
+OTUs based on their median ranks (median shown in black) are reported
+here. OTUs that were associated with classifying a subject as being
+healthy had negative signs and were shown in blue. OTUs that were
+associated with classifying a subject having an SRN had positive signs
+and were shown in red. Some of the same OTUs were identified as
+important in all of the linear models. \newpage
 \includegraphics{Figure_4.png}
@@
 logistic regression and random forest for each of the hundred
-datasplits.} This histogram shows the number of datasplits in each bin.
-The percentage of dataplits where the difference between random forest
-and L2-regularized logistic regression AUROC values was higher than or
-equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical
-red line highlights the bins where there AUROC difference between the
-two model is 0. The p-value was calculated for a double tail event.
-\newpage
+data-splits.} This histogram shows the number of data-splits in each
+bin. The percentage of dataplits where the difference between random
+forest and L2-regularized logistic regression AUROC values was higher
+than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The
+vertical red line highlights the bins where there AUROC difference
+between the two model is 0. The p-value was calculated for a double tail
+event. \newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

### 2019-10-21 submission/manuscript.Rmd
last edits before submission

@@
 However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress.
-These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances.
+These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances.
 Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.
@@
 Depending on the goal of the modeling, other approaches may be more appropriate.
-For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may (but not always) emphasize predictive performance.
+For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance.
 Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
@@
 __Model selection and pipeline construction__.
-We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].
+We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1, Table 1].
 
@@
 To select hyperparameters, we performed a grid search for hyperparameter settings when training the models.
-Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and should to be optimized for each new ML task.
-For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning [Figure S1].
+Default hyperparameter settings in developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
+For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning as the default C parameter was 1 [Figure S1].
 
@@
 L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2].
-Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.
+Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well as non-linear models.
 
@@
 The largest difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2].
-These differences are relatively small and give us confidence in our estimate of the generalization performance of the models.
+These differences were relatively small and gave us confidence in our estimate of the generalization performance of the models.
 
@@
 __The computational efficiency of each ML model.__ We compared the training times of the seven ML models.
-We did not take advantage of parallelization in training, nor was the code optimized for speed.
-Still, as expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations.
+The training times increased with the complexity of the model and the number of potential hyperparameter combinations.
 Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
@@
 We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance.
-We used the AUROC metric to evaluate predictive performance as our dataset was highly balanced.
+We used the AUROC metric to evaluate predictive performance as it is a clinically relevant evaluation metric for our study.
 We repeated this data-split 100 times to measure the possible variation in predictive performance.
@@
 The sequence data used in our analyses were originally generated by Baxter et al.
-(available through NCBI Sequence Read Archive [SRP062005], 2015).
+(available through NCBI Sequence Read Archive [SRP062005], @baxter_microbiota-based_2016).
 The OTU abundance table was generated by Sze et al [@sze_leveraging_2018], who processed the 16S rRNA sequences in mothur (v1.39.3) using the default quality filtering methods, identifying and removing chimeric sequences using VSEARCH, and assigning to OTUs at 97% similarity using the OptiClust algorithm [@schloss_introducing_2009; @westcott_opticlust_2017; @rognes_vsearch_2016]; (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 For hyperparameter selection, we started with a granular grid search.
-Then we narrowed and fine-tuned the range of each hyperparameter.
-A grid search was needed to avoid large variation in prediction performance.
-For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization.
+Then we narrowed and fine-tuned the range of each hyperparameter.For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter, which controls the regularization strength, where smaller values specify stronger regularization.
 For SVM with radial basis function kernel, we also tuned the sigma hyperparameter, which determines the reach of a single training instance where, for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary.
@@
 This allowed the 100 data-splits to be processed through the ML pipeline simultaneously at the same time for each model.
-It is possible to further parallelize the cross-validation step for each hyperparameter setting.
+It is possible to further parallelize the cross-validation step for each hyperparameter setting which we have not performed in this study.
 
@@
 We would like to thank the members of the Schloss lab for their valuable feedback.
-Salary support for M.R came from NIH grant 1R01CA215574.
-Salary support for P.D.S.
-came from NIH grants P30DK034933 and 1R01CA215574.
+Salary support for MR came from NIH grant 1R01CA215574.
+Salary support for PDS came from NIH grants P30DK034933 and 1R01CA215574.
+
+\newpage
+
+## References
+
+<div id="refs"></div>
+
+\newpage
+
+\captionsetup{labelformat=empty}
+\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
+\small
+\begin{tabular}{|l|l|l|}
+\hline
+
+\rowcolor{lightgray}
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
+
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
+
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
+
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
+
+\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
+root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
+
+Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
+
+\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
+
+\end{tabular}
+\newpage
 
@@
 OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red.
-Some of the same OTUs were identified as important in all of the linear models.
 \newpage
@@
 \newpage
+
+\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
+\small
+\begin{tabular}{|l|l|l|l|}
+\hline
+
+\rowcolor{lightgray}
+\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
+
+\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
+
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
+
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
+
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
+
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
+
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g.
+when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
+
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
+
+\end{tabular}
+\newpage
+
 \includegraphics{Figure_S1.png}
-
 **Figure S1.
@@
 Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
-
-\newpage
-\captionsetup{labelformat=empty}
-\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
-\small
-\begin{tabular}{|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-
-\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
-
-Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
-
-\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
-
-\end{tabular}
-\newpage
-
-\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices.}
-\small
-\begin{tabular}{|l|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
-
-\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
-
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
-
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
-
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
-
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g.
-when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
-
-\end{tabular}
-\newpage
-
-## References
-
-<div id="refs"></div>

### 2019-10-21 submission/manuscript.tex
last edits before submission

@@
 which methods are used and how these methods are implemented; evaluating
-models without a separate held-out test data; unreported variation
-between the predictive performance on different folds of
-cross-validation; and unreported variation between cross-validation and
-testing performances. Though the microbiome field is making progress to
-avoid some of these pitfalls including validating their models on
-independent datasets (8, 19, 20) and introducing ways to better use ML
-tools (21--24), more work is needed to improve reproducibility further
-and minimize overestimating for model performance.
+models without separate held-out test data; unreported variation between
+the predictive performance on different folds of cross-validation; and
+unreported variation between cross-validation and testing performances.
+Though the microbiome field is making progress to avoid some of these
+pitfalls including validating their models on independent datasets (8,
+19, 20) and introducing ways to better use ML tools (21--24), more work
+is needed to improve reproducibility further and minimize overestimating
+for model performance.
 
@@
 disease may desire a more interpretable model, whereas clinicians may
-(but not always) emphasize predictive performance. Nonetheless, it is
-essential to understand that the benefit of more complex, less
-interpretable models may be minimal (29, 30). It is important for
-researchers to justify their choice of modeling approach.
+emphasize predictive performance. Nonetheless, it is essential to
+understand that the benefit of more complex, less interpretable models
+may be minimal (29, 30). It is important for researchers to justify
+their choice of modeling approach.
 
@@
 seven different commonly used supervised learning algorithms {[}Figure
-1{]}.
+1, Table 1{]}.
 
@@
 search for hyperparameter settings when training the models. Default
-hyperparameter settings in previously developed ML packages available in
-R, Python, and MATLAB programming languages may be inadequate for
-effective application of classification algorithms and should to be
-optimized for each new ML task. For example, L1-regularized SVM with
-linear kernel showed large variability between different regularization
-strengths (C) and benefited from tuning {[}Figure S1{]}.
+hyperparameter settings in developed ML packages available in R, Python,
+and MATLAB programming languages may be inadequate for effective
+application of classification algorithms and need to be optimized for
+each new ML task. For example, L1-regularized SVM with linear kernel
+showed large variability between different regularization strengths (C)
+and benefited from tuning as the default C parameter was 1 {[}Figure
+S1{]}.
 
@@
 (L2-regularized logistic regression and L2-regularized SVM with linear
-kernel) performed nearly as well non-linear models.
+kernel) performed nearly as well as non-linear models.
 
@@
 decision tree with a difference of 0.007 and 0.006, respectively
-{[}Figure 2{]}. These differences are relatively small and give us
+{[}Figure 2{]}. These differences were relatively small and gave us
 confidence in our estimate of the generalization performance of the
@@
 \textbf{The computational efficiency of each ML model.} We compared the
-training times of the seven ML models. We did not take advantage of
-parallelization in training, nor was the code optimized for speed.
-Still, as expected, the training times increased with the complexity of
-the model and the number of potential hyperparameter combinations. Also,
-the linear models trained faster than non-linear models {[}Figures
-S1-S2; Figure 5{]}.
+training times of the seven ML models. The training times increased with
+the complexity of the model and the number of potential hyperparameter
+combinations. Also, the linear models trained faster than non-linear
+models {[}Figures S1-S2; Figure 5{]}.
 
@@
 (20\% of the data), which we used to evaluate predictive performance. We
-used the AUROC metric to evaluate predictive performance as our dataset
-was highly balanced. We repeated this data-split 100 times to measure
-the possible variation in predictive performance. During training, we
-tuned the model hyperparameters with a repeated five-fold
-cross-validation. Despite the high number of features microbiome
-datasets typically have, the models we built with this pipeline
-generalized to the held-out test sets.
+used the AUROC metric to evaluate predictive performance as it is a
+clinically relevant evaluation metric for our study. We repeated this
+data-split 100 times to measure the possible variation in predictive
+performance. During training, we tuned the model hyperparameters with a
+repeated five-fold cross-validation. Despite the high number of features
+microbiome datasets typically have, the models we built with this
+pipeline generalized to the held-out test sets.
 
@@
 originally generated by Baxter et al. (available through NCBI Sequence
-Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
+Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
 generated by Sze et al (44), who processed the 16S rRNA sequences in
@@
 For hyperparameter selection, we started with a granular grid search.
-Then we narrowed and fine-tuned the range of each hyperparameter. A grid
-search was needed to avoid large variation in prediction performance.
-For L2-regularized logistic regression, L1 and L2-regularized SVM with
+Then we narrowed and fine-tuned the range of each hyperparameter.For
+L2-regularized logistic regression, L1 and L2-regularized SVM with
 linear and radial basis function kernels, we tuned the cost
@@
 the same time for each model. It is possible to further parallelize the
-cross-validation step for each hyperparameter setting.
+cross-validation step for each hyperparameter setting which we have not
+performed in this study.
 
@@
 thank the members of the Schloss lab for their valuable feedback. Salary
-support for M.R came from NIH grant 1R01CA215574. Salary support for
-P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.
+support for MR came from NIH grant 1R01CA215574. Salary support for PDS
+came from NIH grants P30DK034933 and 1R01CA215574.
+
+\newpage
+
+\subsection{References}\label{references}
+
+\hypertarget{refs}{}
+\hypertarget{ref-segata_metagenomic_2011}{}
+1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
+\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
+\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
+explanation. Genome Biol \textbf{12}:R60.
+doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.
+
+\hypertarget{ref-zeller_potential_2014}{}
+2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
+\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
+\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
+\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
+\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
+J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
+\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
+\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
+for early-stage detection of colorectal cancer. Mol Syst Biol
+\textbf{10}.
+doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
+
+\hypertarget{ref-zackular_human_2014}{}
+3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
+
+\hypertarget{ref-baxter_dna_2016}{}
+4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
+\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
+immunochemical test can replace stool for detection of colonic lesions
+using a microbiota-based model. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
+
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\hypertarget{ref-hale_shifts_2017}{}
+6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
+\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
+\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
+\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
+\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
+microbiota associated with adenomatous polyps. Cancer Epidemiol
+Biomarkers Prev \textbf{26}:85--94.
+doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.
+
+\hypertarget{ref-pasolli_machine_2016}{}
+7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
+\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
+meta-analysis of large metagenomic datasets: Tools and biological
+insights. PLoS Comput Biol \textbf{12}.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.
+
+\hypertarget{ref-sze_looking_2016}{}
+8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
+the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
+doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.
+
+\hypertarget{ref-walters_meta-analyses_2014}{}
+9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
+Meta-analyses of human gut microbes associated with obesity and IBD.
+FEBS Lett \textbf{588}:4223--4233.
+doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
+
+\hypertarget{ref-vazquez-baeza_guiding_2018}{}
+10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
+\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
+\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
+Gut \textbf{67}:1743--1745.
+doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.
+
+\hypertarget{ref-qin_alterations_2014}{}
+11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
+\textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
+E}, \textbf{Yao J}, \textbf{Wu L}, \textbf{Zhou J}, \textbf{Ni S},
+\textbf{Liu L}, \textbf{Pons N}, \textbf{Batto JM}, \textbf{Kennedy SP},
+\textbf{Leonard P}, \textbf{Yuan C}, \textbf{Ding W}, \textbf{Chen Y},
+\textbf{Hu X}, \textbf{Zheng B}, \textbf{Qian G}, \textbf{Xu W},
+\textbf{Ehrlich SD}, \textbf{Zheng S}, \textbf{Li L}. 2014. Alterations
+of the human gut microbiome in liver cirrhosis. Nature
+\textbf{513}:59--64.
+doi:\href{https://doi.org/10.1038/nature13568}{10.1038/nature13568}.
+
+\hypertarget{ref-geman_deep_2018}{}
+12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
+\textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
+learning tools for human microbiome big data, pp. 265--275. \emph{In}
+Balas, VE, Jain, LC, Balas, MM (eds.), Soft computing applications.
+Springer International Publishing.
+
+\hypertarget{ref-thaiss_persistent_2016}{}
+13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
+\textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
+\textbf{Dohnalová L}, \textbf{Braverman S}, \textbf{Rozin S},
+\textbf{Malitsky S}, \textbf{Dori-Bachash M}, \textbf{Kuperman Y},
+\textbf{Biton I}, \textbf{Gertler A}, \textbf{Harmelin A},
+\textbf{Shapiro H}, \textbf{Halpern Z}, \textbf{Aharoni A},
+\textbf{Segal E}, \textbf{Elinav E}. 2016. Persistent microbiome
+alterations modulate the rate of post-dieting weight regain. Nature
+\textbf{540}:544--551.
+doi:\href{https://doi.org/10.1038/nature20796}{10.1038/nature20796}.
+
+\hypertarget{ref-dadkhah_gut_2019}{}
+14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
+\textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
+G}, \textbf{Kuehn T}, \textbf{Gillevet PM}. 2019. Gut microbiome
+identifies risk for colorectal polyps. BMJ Open Gastroenterology
+\textbf{6}:e000297.
+doi:\href{https://doi.org/10.1136/bmjgast-2019-000297}{10.1136/bmjgast-2019-000297}.
+
+\hypertarget{ref-flemer_oral_2018}{}
+15. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
+\textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
+E}, \textbf{O`Riordain M}, \textbf{Shanahan F}, \textbf{O`Toole PW}.
+2018. The oral microbiota in colorectal cancer is distinctive and
+predictive. Gut \textbf{67}:1454--1463.
+doi:\href{https://doi.org/10.1136/gutjnl-2017-314814}{10.1136/gutjnl-2017-314814}.
+
+\hypertarget{ref-montassier_pretreatment_2016}{}
+16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
+\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
+P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
+2016. Pretreatment gut microbiome predicts chemotherapy-related
+bloodstream infection. Genome Medicine \textbf{8}:49.
+doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.
+
+\hypertarget{ref-ai_systematic_2017}{}
+17. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
+\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
+supervised classifiers for fecal microbiota-based prediction of
+colorectal cancer. Oncotarget \textbf{8}:9546--9556.
+doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.
+
+\hypertarget{ref-dai_multi-cohort_2018}{}
+18. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
+WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
+\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
+J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
+identified altered bacteria across populations and universal bacterial
+markers. Microbiome \textbf{6}:70.
+doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.
+
+\hypertarget{ref-mossotto_classification_2017}{}
+19. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
+\textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
+Classification of paediatric inflammatory bowel disease using machine
+learning. Scientific Reports \textbf{7}.
+doi:\href{https://doi.org/10.1038/s41598-017-02606-2}{10.1038/s41598-017-02606-2}.
+
+\hypertarget{ref-wong_quantitation_2017}{}
+20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
+AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
+\textbf{Zhang L}, \textbf{Wu JCY}, \textbf{Chan FKL}, \textbf{Ng SSM},
+\textbf{Wong MCS}, \textbf{Ng SC}, \textbf{Wu WKK}, \textbf{Yu J},
+\textbf{Sung JJY}. 2017. Quantitation of faecal fusobacterium improves
+faecal immunochemical test in detecting advanced colorectal neoplasia.
+Gut \textbf{66}:1441--1448.
+doi:\href{https://doi.org/10.1136/gutjnl-2016-312766}{10.1136/gutjnl-2016-312766}.
+
+\hypertarget{ref-statnikov_comprehensive_2013}{}
+21. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
+\textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
+\textbf{Blaser MJ}, \textbf{Aliferis CF}, \textbf{Alekseyenko AV}. 2013.
+A comprehensive evaluation of multicategory classification methods for
+microbiomic data. Microbiome \textbf{1}:11.
+doi:\href{https://doi.org/10.1186/2049-2618-1-11}{10.1186/2049-2618-1-11}.
+
+\hypertarget{ref-knights_supervised_2011}{}
+22. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
+Supervised classification of human microbiota. FEMS Microbiology Reviews
+\textbf{35}:343--359.
+doi:\href{https://doi.org/10.1111/j.1574-6976.2010.00251.x}{10.1111/j.1574-6976.2010.00251.x}.
+
+\hypertarget{ref-wirbel_meta-analysis_2019}{}
+23. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
+K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
+\textbf{Voigt AY}, \textbf{Palleja A}, \textbf{Ponnudurai R},
+\textbf{Sunagawa S}, \textbf{Coelho LP}, \textbf{Schrotz-King P},
+\textbf{Vogtmann E}, \textbf{Habermann N}, \textbf{Niméus E},
+\textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Gandini S},
+\textbf{Serrano D}, \textbf{Mizutani S}, \textbf{Shiroma H},
+\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
+T}, \textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N},
+\textbf{Sinha R}, \textbf{Ulrich CM}, \textbf{Brenner H},
+\textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller G}. 2019.
+Meta-analysis of fecal metagenomes reveals global microbial signatures
+that are specific for colorectal cancer. Nature Medicine
+\textbf{25}:679.
+doi:\href{https://doi.org/10.1038/s41591-019-0406-6}{10.1038/s41591-019-0406-6}.
+
+\hypertarget{ref-vangay_microbiome_2019}{}
+24. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
+Microbiome learning repo (ML repo): A public repository of microbiome
+regression and classification tasks. Gigascience \textbf{8}.
+doi:\href{https://doi.org/10.1093/gigascience/giz042}{10.1093/gigascience/giz042}.
+
+\hypertarget{ref-galkin_human_2018}{}
+25. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
+\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
+2018. Human microbiome aging clocks based on deep learning and tandem of
+permutation feature importance and accumulated local effects. bioRxiv.
+doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
+
+\hypertarget{ref-reiman_using_2017}{}
+26. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
+convolutional neural networks to explore the microbiome, pp. 4269--4272.
+\emph{In} 2017 39th annual international conference of the IEEE
+engineering in medicine and biology society (EMBC).
+
+\hypertarget{ref-fioravanti_phylogenetic_2017}{}
+27. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
+\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
+\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
+in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
+
+\hypertarget{ref-thomas_metagenomic_2019}{}
+28. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
+\textbf{Pasolli E}, \textbf{Armanini F}, \textbf{Zolfo M},
+\textbf{Beghini F}, \textbf{Manara S}, \textbf{Karcher N}, \textbf{Pozzi
+C}, \textbf{Gandini S}, \textbf{Serrano D}, \textbf{Tarallo S},
+\textbf{Francavilla A}, \textbf{Gallo G}, \textbf{Trompetto M},
+\textbf{Ferrero G}, \textbf{Mizutani S}, \textbf{Shiroma H},
+\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
+T}, \textbf{Wirbel J}, \textbf{Schrotz-King P}, \textbf{Ulrich CM},
+\textbf{Brenner H}, \textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller
+G}, \textbf{Cordero F}, \textbf{Dias-Neto E}, \textbf{Setubal JC},
+\textbf{Tett A}, \textbf{Pardini B}, \textbf{Rescigno M},
+\textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N}. 2019.
+Metagenomic analysis of colorectal cancer datasets identifies
+cross-cohort microbial diagnostic signatures and a link with choline
+degradation. Nature Medicine \textbf{25}:667.
+doi:\href{https://doi.org/10.1038/s41591-019-0405-7}{10.1038/s41591-019-0405-7}.
+
+\hypertarget{ref-rudin_please_2018}{}
+29. \textbf{Rudin C}. 2018. Please stop explaining black box models for
+high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
+
+\hypertarget{ref-rudin_optimized_2018}{}
+30. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
+Toward trust in machine learning for healthcare and criminal justice.
+Interfaces \textbf{48}:449--466.
+doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.
+
+\hypertarget{ref-knights_human-associated_2011}{}
+31. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
+\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
+signatures: Examining their predictive value. Cell Host Microbe
+\textbf{10}:292--296.
+doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.
+
+\hypertarget{ref-Ng:2004:FSL:1015330.1015435}{}
+32. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
+and rotational invariance, pp. 78. \emph{In} Proceedings of the
+twenty-first international conference on machine learning. ACM, New
+York, NY, USA.
+
+\hypertarget{ref-miller_explanation_2017}{}
+33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
+Insights from the social sciences. arXiv:170607269 {[}cs{]}.
+
+\hypertarget{ref-ribeiro_why_2016}{}
+34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
+``Why should i trust you?'': Explaining the predictions of any
+classifier. arXiv:160204938 {[}cs, stat{]}.
+
+\hypertarget{ref-nori_interpretml:_2019}{}
+35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
+\textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
+learning interpretability. arXiv:190909223 {[}cs, stat{]}.
+
+\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
+36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
+\textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
+importance measure. Bioinformatics \textbf{26}:1340--1347.
+doi:\href{https://doi.org/10.1093/bioinformatics/btq134}{10.1093/bioinformatics/btq134}.
+
+\hypertarget{ref-dormann_collinearity:_2013}{}
+37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
+JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
+\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
+\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
+\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
+of methods to deal with it and a simulation study evaluating their
+performance. Ecography \textbf{36}:27--46.
+doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.
+
+\hypertarget{ref-sze_fecal_2019}{}
+38. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
+acids are not predictive of colonic tumor status and cannot be predicted
+based on bacterial community structure. mBio \textbf{10}:e01454--19.
+doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.
+
+\hypertarget{ref-kocheturov_massive_2019}{}
+39. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+2019. Massive datasets and machine learning for computational
+biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
+doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.
+
+\hypertarget{ref-kim_improved_2018}{}
+40. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+method for prediction of cancer prognosis by network learning. Genes
+\textbf{9}:478.
+doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.
+
+\hypertarget{ref-wiens_no_2019}{}
+41. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
+\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
+\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
+\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
+2019. Do no harm: A roadmap for responsible machine learning for health
+care. Nat Med \textbf{25}:1337--1340.
+doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.
+
+\hypertarget{ref-westcott_opticlust_2017}{}
+42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+Improved Method for Assigning Amplicon-Based Sequence Data to
+Operational Taxonomic Units. mSphere \textbf{2}.
+doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
+
+\hypertarget{ref-redwood_stool_2016}{}
+43. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
+\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
+\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
+TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
+for screening detection of colorectal neoplasia in alaska native people.
+Mayo Clin Proc \textbf{91}:61--70.
+doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.
+
+\hypertarget{ref-sze_leveraging_2018}{}
+44. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+rRNA gene surveys to identify reproducible biomarkers in individuals
+with colorectal tumors. mBio \textbf{9}:e00630--18.
+doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
+
+\hypertarget{ref-schloss_introducing_2009}{}
+45. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-Source, Platform-Independent,
+Community-Supported Software for Describing and Comparing Microbial
+Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.
+
+\hypertarget{ref-rognes_vsearch_2016}{}
+46. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
+source tool for metagenomics. PeerJ \textbf{4}:e2584.
+doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+
+\hypertarget{ref-li_hyperband:_2016}{}
+47. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
+bandit-based approach to hyperparameter optimization. arXiv:160306560
+{[}cs, stat{]}.
+
+\newpage
+
+\captionsetup{labelformat=empty}
+\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
+\small
+
+\begin{tabular}{|l|l|l|}
+\hline
+
+\rowcolor{lightgray}
+\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
+
+\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
+
+\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
+
+\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
+
+\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
+root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
+
+Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
+
+\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
+
+\end{tabular}\newpage
 
@@
 associated with classifying a subject having an SRN had positive signs
-and were shown in red. Some of the same OTUs were identified as
-important in all of the linear models. \newpage
+and were shown in red. \newpage
 \includegraphics{Figure_4.png}
@@
 L2-regularized logistic regression. \newpage
-\includegraphics{Figure_S1.png}
-
-\textbf{Figure S1. Hyperparameter setting performances for linear
-models.} (A) L2-regularized logistic regression, (B) L1-regularized SVM
-with linear kernel, and (C) L2-regularized SVM with linear kernel mean
-cross-validation AUROC values when different hyperparameters were used
-in training the model. The stars represent the highest performing
-hyperparameter setting for each model. \newpage
+
+\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
+\small
+
+\begin{tabular}{|l|l|l|l|}
+\hline
+
+\rowcolor{lightgray}
+\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
+
+\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
+
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
+
+\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
+
+\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
+
+\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
+
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g.
+when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
+
+\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
+
+\end{tabular}\newpage
+
+\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
+setting performances for linear models.} (A) L2-regularized logistic
+regression, (B) L1-regularized SVM with linear kernel, and (C)
+L2-regularized SVM with linear kernel mean cross-validation AUROC values
+when different hyperparameters were used in training the model. The
+stars represent the highest performing hyperparameter setting for each
+model. \newpage
 \includegraphics[height=30cm, width=15cm]{Figure_S2.png}
@@
 
-\newpage
-
-\captionsetup{labelformat=empty}
-\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
-\small
-
-\begin{tabular}{|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-
-\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
-
-Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
-
-\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
-
-\end{tabular}\newpage
-
-\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices.}
-\small
-
-\begin{tabular}{|l|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
-
-\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
-
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
-
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
-
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
-
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g.
-when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
-
-\end{tabular}\newpage
-
-\subsection{References}\label{references}
-
-\hypertarget{refs}{}
-\hypertarget{ref-segata_metagenomic_2011}{}
-1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
-\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
-\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
-explanation. Genome Biol \textbf{12}:R60.
-doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.
-
-\hypertarget{ref-zeller_potential_2014}{}
-2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
-\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
-\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
-\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
-\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
-\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
-J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
-\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
-\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
-for early-stage detection of colorectal cancer. Mol Syst Biol
-\textbf{10}.
-doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
-
-\hypertarget{ref-zackular_human_2014}{}
-3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
-\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
-for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
-doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
-
-\hypertarget{ref-baxter_dna_2016}{}
-4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
-\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
-immunochemical test can replace stool for detection of colonic lesions
-using a microbiota-based model. Microbiome \textbf{4}.
-doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
-
-\hypertarget{ref-baxter_microbiota-based_2016}{}
-5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\hypertarget{ref-hale_shifts_2017}{}
-6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
-\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
-\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
-\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
-\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
-microbiota associated with adenomatous polyps. Cancer Epidemiol
-Biomarkers Prev \textbf{26}:85--94.
-doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.
-
-\hypertarget{ref-pasolli_machine_2016}{}
-7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
-\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
-meta-analysis of large metagenomic datasets: Tools and biological
-insights. PLoS Comput Biol \textbf{12}.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.
-
-\hypertarget{ref-sze_looking_2016}{}
-8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
-the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
-doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.
-
-\hypertarget{ref-walters_meta-analyses_2014}{}
-9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
-Meta-analyses of human gut microbes associated with obesity and IBD.
-FEBS Lett \textbf{588}:4223--4233.
-doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
-
-\hypertarget{ref-vazquez-baeza_guiding_2018}{}
-10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
-\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
-\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
-Gut \textbf{67}:1743--1745.
-doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.
-
-\hypertarget{ref-qin_alterations_2014}{}
-11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
-\textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
-E}, \textbf{Yao J}, \textbf{Wu L}, \textbf{Zhou J}, \textbf{Ni S},
-\textbf{Liu L}, \textbf{Pons N}, \textbf{Batto JM}, \textbf{Kennedy SP},
-\textbf{Leonard P}, \textbf{Yuan C}, \textbf{Ding W}, \textbf{Chen Y},
-\textbf{Hu X}, \textbf{Zheng B}, \textbf{Qian G}, \textbf{Xu W},
-\textbf{Ehrlich SD}, \textbf{Zheng S}, \textbf{Li L}. 2014. Alterations
-of the human gut microbiome in liver cirrhosis. Nature
-\textbf{513}:59--64.
-doi:\href{https://doi.org/10.1038/nature13568}{10.1038/nature13568}.
-
-\hypertarget{ref-geman_deep_2018}{}
-12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
-\textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
-learning tools for human microbiome big data, pp. 265--275. \emph{In}
-Balas, VE, Jain, LC, Balas, MM (eds.), Soft computing applications.
-Springer International Publishing.
-
-\hypertarget{ref-thaiss_persistent_2016}{}
-13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
-\textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
-\textbf{Dohnalová L}, \textbf{Braverman S}, \textbf{Rozin S},
-\textbf{Malitsky S}, \textbf{Dori-Bachash M}, \textbf{Kuperman Y},
-\textbf{Biton I}, \textbf{Gertler A}, \textbf{Harmelin A},
-\textbf{Shapiro H}, \textbf{Halpern Z}, \textbf{Aharoni A},
-\textbf{Segal E}, \textbf{Elinav E}. 2016. Persistent microbiome
-alterations modulate the rate of post-dieting weight regain. Nature
-\textbf{540}:544--551.
-doi:\href{https://doi.org/10.1038/nature20796}{10.1038/nature20796}.
-
-\hypertarget{ref-dadkhah_gut_2019}{}
-14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
-\textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
-G}, \textbf{Kuehn T}, \textbf{Gillevet PM}. 2019. Gut microbiome
-identifies risk for colorectal polyps. BMJ Open Gastroenterology
-\textbf{6}:e000297.
-doi:\href{https://doi.org/10.1136/bmjgast-2019-000297}{10.1136/bmjgast-2019-000297}.
-
-\hypertarget{ref-flemer_oral_2018}{}
-15. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
-\textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
-E}, \textbf{O`Riordain M}, \textbf{Shanahan F}, \textbf{O`Toole PW}.
-2018. The oral microbiota in colorectal cancer is distinctive and
-predictive. Gut \textbf{67}:1454--1463.
-doi:\href{https://doi.org/10.1136/gutjnl-2017-314814}{10.1136/gutjnl-2017-314814}.
-
-\hypertarget{ref-montassier_pretreatment_2016}{}
-16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
-\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
-P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
-2016. Pretreatment gut microbiome predicts chemotherapy-related
-bloodstream infection. Genome Medicine \textbf{8}:49.
-doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.
-
-\hypertarget{ref-ai_systematic_2017}{}
-17. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
-\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
-supervised classifiers for fecal microbiota-based prediction of
-colorectal cancer. Oncotarget \textbf{8}:9546--9556.
-doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.
-
-\hypertarget{ref-dai_multi-cohort_2018}{}
-18. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
-WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
-\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
-J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
-identified altered bacteria across populations and universal bacterial
-markers. Microbiome \textbf{6}:70.
-doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.
-
-\hypertarget{ref-mossotto_classification_2017}{}
-19. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
-\textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
-Classification of paediatric inflammatory bowel disease using machine
-learning. Scientific Reports \textbf{7}.
-doi:\href{https://doi.org/10.1038/s41598-017-02606-2}{10.1038/s41598-017-02606-2}.
-
-\hypertarget{ref-wong_quantitation_2017}{}
-20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
-AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
-\textbf{Zhang L}, \textbf{Wu JCY}, \textbf{Chan FKL}, \textbf{Ng SSM},
-\textbf{Wong MCS}, \textbf{Ng SC}, \textbf{Wu WKK}, \textbf{Yu J},
-\textbf{Sung JJY}. 2017. Quantitation of faecal fusobacterium improves
-faecal immunochemical test in detecting advanced colorectal neoplasia.
-Gut \textbf{66}:1441--1448.
-doi:\href{https://doi.org/10.1136/gutjnl-2016-312766}{10.1136/gutjnl-2016-312766}.
-
-\hypertarget{ref-statnikov_comprehensive_2013}{}
-21. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
-\textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
-\textbf{Blaser MJ}, \textbf{Aliferis CF}, \textbf{Alekseyenko AV}. 2013.
-A comprehensive evaluation of multicategory classification methods for
-microbiomic data. Microbiome \textbf{1}:11.
-doi:\href{https://doi.org/10.1186/2049-2618-1-11}{10.1186/2049-2618-1-11}.
-
-\hypertarget{ref-knights_supervised_2011}{}
-22. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
-Supervised classification of human microbiota. FEMS Microbiology Reviews
-\textbf{35}:343--359.
-doi:\href{https://doi.org/10.1111/j.1574-6976.2010.00251.x}{10.1111/j.1574-6976.2010.00251.x}.
-
-\hypertarget{ref-wirbel_meta-analysis_2019}{}
-23. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
-K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
-\textbf{Voigt AY}, \textbf{Palleja A}, \textbf{Ponnudurai R},
-\textbf{Sunagawa S}, \textbf{Coelho LP}, \textbf{Schrotz-King P},
-\textbf{Vogtmann E}, \textbf{Habermann N}, \textbf{Niméus E},
-\textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Gandini S},
-\textbf{Serrano D}, \textbf{Mizutani S}, \textbf{Shiroma H},
-\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
-T}, \textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N},
-\textbf{Sinha R}, \textbf{Ulrich CM}, \textbf{Brenner H},
-\textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller G}. 2019.
-Meta-analysis of fecal metagenomes reveals global microbial signatures
-that are specific for colorectal cancer. Nature Medicine
-\textbf{25}:679.
-doi:\href{https://doi.org/10.1038/s41591-019-0406-6}{10.1038/s41591-019-0406-6}.
-
-\hypertarget{ref-vangay_microbiome_2019}{}
-24. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
-Microbiome learning repo (ML repo): A public repository of microbiome
-regression and classification tasks. Gigascience \textbf{8}.
-doi:\href{https://doi.org/10.1093/gigascience/giz042}{10.1093/gigascience/giz042}.
-
-\hypertarget{ref-galkin_human_2018}{}
-25. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
-\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
-2018. Human microbiome aging clocks based on deep learning and tandem of
-permutation feature importance and accumulated local effects. bioRxiv.
-doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
-
-\hypertarget{ref-reiman_using_2017}{}
-26. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
-convolutional neural networks to explore the microbiome, pp. 4269--4272.
-\emph{In} 2017 39th annual international conference of the IEEE
-engineering in medicine and biology society (EMBC).
-
-\hypertarget{ref-fioravanti_phylogenetic_2017}{}
-27. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
-\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
-\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
-in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
-
-\hypertarget{ref-thomas_metagenomic_2019}{}
-28. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
-\textbf{Pasolli E}, \textbf{Armanini F}, \textbf{Zolfo M},
-\textbf{Beghini F}, \textbf{Manara S}, \textbf{Karcher N}, \textbf{Pozzi
-C}, \textbf{Gandini S}, \textbf{Serrano D}, \textbf{Tarallo S},
-\textbf{Francavilla A}, \textbf{Gallo G}, \textbf{Trompetto M},
-\textbf{Ferrero G}, \textbf{Mizutani S}, \textbf{Shiroma H},
-\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
-T}, \textbf{Wirbel J}, \textbf{Schrotz-King P}, \textbf{Ulrich CM},
-\textbf{Brenner H}, \textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller
-G}, \textbf{Cordero F}, \textbf{Dias-Neto E}, \textbf{Setubal JC},
-\textbf{Tett A}, \textbf{Pardini B}, \textbf{Rescigno M},
-\textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N}. 2019.
-Metagenomic analysis of colorectal cancer datasets identifies
-cross-cohort microbial diagnostic signatures and a link with choline
-degradation. Nature Medicine \textbf{25}:667.
-doi:\href{https://doi.org/10.1038/s41591-019-0405-7}{10.1038/s41591-019-0405-7}.
-
-\hypertarget{ref-rudin_please_2018}{}
-29. \textbf{Rudin C}. 2018. Please stop explaining black box models for
-high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
-
-\hypertarget{ref-rudin_optimized_2018}{}
-30. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
-Toward trust in machine learning for healthcare and criminal justice.
-Interfaces \textbf{48}:449--466.
-doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.
-
-\hypertarget{ref-knights_human-associated_2011}{}
-31. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
-\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
-signatures: Examining their predictive value. Cell Host Microbe
-\textbf{10}:292--296.
-doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.
-
-\hypertarget{ref-Ng:2004:FSL:1015330.1015435}{}
-32. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
-and rotational invariance, pp. 78. \emph{In} Proceedings of the
-twenty-first international conference on machine learning. ACM, New
-York, NY, USA.
-
-\hypertarget{ref-miller_explanation_2017}{}
-33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
-Insights from the social sciences. arXiv:170607269 {[}cs{]}.
-
-\hypertarget{ref-ribeiro_why_2016}{}
-34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
-``Why should i trust you?'': Explaining the predictions of any
-classifier. arXiv:160204938 {[}cs, stat{]}.
-
-\hypertarget{ref-nori_interpretml:_2019}{}
-35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
-\textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
-learning interpretability. arXiv:190909223 {[}cs, stat{]}.
-
-\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
-36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
-\textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
-importance measure. Bioinformatics \textbf{26}:1340--1347.
-doi:\href{https://doi.org/10.1093/bioinformatics/btq134}{10.1093/bioinformatics/btq134}.
-
-\hypertarget{ref-dormann_collinearity:_2013}{}
-37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
-\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
-JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
-\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
-\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
-\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
-of methods to deal with it and a simulation study evaluating their
-performance. Ecography \textbf{36}:27--46.
-doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.
-
-\hypertarget{ref-sze_fecal_2019}{}
-38. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
-\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
-acids are not predictive of colonic tumor status and cannot be predicted
-based on bacterial community structure. mBio \textbf{10}:e01454--19.
-doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.
-
-\hypertarget{ref-kocheturov_massive_2019}{}
-39. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
-2019. Massive datasets and machine learning for computational
-biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
-doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.
-
-\hypertarget{ref-kim_improved_2018}{}
-40. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
-method for prediction of cancer prognosis by network learning. Genes
-\textbf{9}:478.
-doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.
-
-\hypertarget{ref-wiens_no_2019}{}
-41. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
-\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
-\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
-\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
-2019. Do no harm: A roadmap for responsible machine learning for health
-care. Nat Med \textbf{25}:1337--1340.
-doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.
-
-\hypertarget{ref-westcott_opticlust_2017}{}
-42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
-Improved Method for Assigning Amplicon-Based Sequence Data to
-Operational Taxonomic Units. mSphere \textbf{2}.
-doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
-
-\hypertarget{ref-redwood_stool_2016}{}
-43. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
-\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
-\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
-\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
-TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
-for screening detection of colorectal neoplasia in alaska native people.
-Mayo Clin Proc \textbf{91}:61--70.
-doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.
-
-\hypertarget{ref-sze_leveraging_2018}{}
-44. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
-rRNA gene surveys to identify reproducible biomarkers in individuals
-with colorectal tumors. mBio \textbf{9}:e00630--18.
-doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
-
-\hypertarget{ref-schloss_introducing_2009}{}
-45. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-Source, Platform-Independent,
-Community-Supported Software for Describing and Comparing Microbial
-Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.
-
-\hypertarget{ref-rognes_vsearch_2016}{}
-46. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
-\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
-source tool for metagenomics. PeerJ \textbf{4}:e2584.
-doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
-
-\hypertarget{ref-li_hyperband:_2016}{}
-47. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
-\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
-bandit-based approach to hyperparameter optimization. arXiv:160306560
-{[}cs, stat{]}.
-
 \end{document}

### 2019-10-21 code/learning/Figure3.R

@@ -133,6 +133,6 @@ logit_graph <- plot_feature_ranks(logit_imp) +
 
 linear <- plot_grid(l1_svm_graph, l2_svm_graph, logit_graph, labels = c("A", "B", "C"), align = 'v', ncol = 1)
 
-ggdraw(add_sub(linear, "Feature Ranks", vpadding=grid::unit(0,"lines"), y=5, x=0.7, vjust=4.75, size=15))
+ggdraw(add_sub(linear, "Feature ranks", vpadding=grid::unit(0,"lines"), y=5, x=0.7, vjust=4.75, size=15))
 
 ggsave("Figure_3.png", plot = last_plot(), device = 'png', path = 'submission', width = 6, height = 9.2)

### 2019-10-21 code/learning/Figure5.R

@@ -44,7 +44,7 @@ traintime_df <- bind_rows(min_fixed_result)
 traintime_plot <- ggplot(traintime_df, aes(x = fct_reorder(model, x), y = x)) +
   geom_boxplot(fill="#0072B2", alpha=0.5, fatten = 1.5) +
   coord_flip() +
-  scale_y_continuous(name = "training time (hours)") +
+  scale_y_continuous(name = "Training time (hours)") +
   scale_x_discrete(name = "",
                    labels=c(expression(paste(L[2], "-regularized logistic regression")),
                             expression(paste(L[1], "-regularized linear SVM")),

### 2019-10-21 code/learning/FigureS1.R

@@ -47,9 +47,9 @@ linear_plots <- fig_data %>%
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.001) +
   scale_x_log10(limits=c(min=1e-4,max=1),
                 breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1),
-                name="regularization penalty (C)",
+                name="Regularization penalty (C)",
                 labels=trans_format('log10',math_format(10^.x))) +
-  scale_y_continuous(name="mean cvAUROC",
+  scale_y_continuous(name="Mean cvAUROC",
                      limits=c(0.4,1.0),
                      breaks=seq(0.4,1,0.1)) +
   scale_color_manual(name=NULL,

### 2019-10-21 code/learning/FigureS2.R

@@ -67,7 +67,7 @@ dt <- dt_all %>%
   summarise(mean_AUC = mean(ROC), sd_AUC = sd(ROC))
 
 dt_plot <- base_plot(dt, dt$maxdepth, dt$mean_AUC) +
-scale_x_continuous(name="maximum depth of tree") +
+scale_x_continuous(name="Maximum depth of tree") +
   scale_y_continuous(name="Decision tree
 mean cvAUROC",
                      limits = c(0.30, 1),
@@ -79,7 +79,7 @@ rf <- rf_all %>%
   summarise(mean_AUC = mean(ROC), sd_AUC = sd(ROC))
 
 rf_plot <-  base_plot(rf, rf$mtry, rf$mean_AUC) +
-scale_x_continuous(name="number of features (mtry)",
+scale_x_continuous(name="Number of features (mtry)",
                    breaks=seq(0, 1500, 250), limits = c(0, 1500)) +
   scale_y_continuous(name="Random forest
 mean cvAUROC",
@@ -106,12 +106,12 @@ rbf_plot <- ggplot(rbf_data, aes(x = sigma, y = C, fill = mean_AUC)) +
            shape = 8) +
   #coord_fixed(ratio = 0.5) +
   #coord_equal() +
-  scale_y_log10(name="regularization penalty 
+  scale_y_log10(name="Regularization penalty 
 (C)",
                 breaks = c(0.0000001, 0.000001, 0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10), 
                 expand = c(0, 0), 
                 labels=trans_format('log10',math_format(10^.x))) +
-  scale_x_log10(name = "the reach of a single training instance (sigma)",
+  scale_x_log10(name = "The reach of a single training instance (sigma)",
                 breaks = c(0.00000001, 0.0000001, 0.000001, 0.00001, 0.0001, 0.001, 0.01, 0.1), 
                 expand = c(0, 0), 
                 labels=trans_format('log10',math_format(10^.x))) +
@@ -145,11 +145,11 @@ xgboost_plot <- ggplot(xgboost_data, aes(x = eta, y = subsample, fill = mean_AUC
            colour = "#FC4E07", 
            size = 3,
            shape = 8) +
-  scale_y_continuous(name="ratio of the training data
+  scale_y_continuous(name="Ratio of the training data
 (subsample)",
     breaks = c(0.4, 0.5, 0.6, 0.7),
     expand=c(0,0)) +
[…]

### 2020-01-17 submission/manuscript.Rmd
some additions for rebuttal

@@
 ---
-title: '**Effective application of machine learning to microbiome-based classification problems**'
+title: '**Effective application of machine learning to a microbiome-based classification problem**'
 csl: mbio.csl
@@
 
-Running title: Machine learning methods applied to microbiome studies
+Running title: Machine learning methods applied to a microbiome study
 
@@
 
-Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies.
+Prediction of diseases using machine learning (ML) is rapidly being adopted in microbiome studies.
 However, the estimated performance associated with these ML models is likely over-optimistic.
@@
 We implement a rigorous pipeline and emphasize the importance of selecting ML models that reflect the goal of the study.
-These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.
+These concepts are not particular to the study of human health but can also be applied to environmental microbiology studies.
 
@@
 For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance.
-Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018].
+Nonetheless, it is essential to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018; @Quinn847475].
 It is important for researchers to justify their choice of modeling approach.
 
+In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed.
 To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline.
@@
 We compared their predictive performance, interpretability, and training time.
-To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
+To demonstrate the performance of these modeling approaches and our pipeline on a case study, we used a data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
 This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
@@
 This framework can be easily applied to other host-associated and environmental microbiome datasets.
+We hope that this study ...
 
@@
 __Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2].
-The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5) (Figure S3).
+The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5; The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis) (Figure S3).
 Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively.
@@
 The vertical red line highlights the bins where there AUROC difference between the two model is 0.
-The p-value was calculated for a double tail event.
+The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis.
+We are testing how often random forest performs more accurately than L2-regularized logistic regression.
+We are assuming that the distribution of the differences between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore we are calculating the p-value for a double-tail event.
 \newpage

### 2020-01-17 submission/manuscript.tex
some additions for rebuttal

@@
 \IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\usepackage[]{microtype}
 \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
-\usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Effective application of machine learning to microbiome-based classification problems},
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\hypersetup{
+            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
             pdfborder={0 0 0},
@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage[margin=1in]{geometry}
 \usepackage{graphicx,grffile}
@@
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{\textbf{Effective application of machine learning to microbiome-based
-classification problems}}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
@@
 
+\title{\textbf{Effective application of machine learning to a microbiome-based
+classification problem}}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 
-Prediction of health outcomes using machine learning (ML) is rapidly
-being adopted in microbiome studies. However, the estimated performance
+Prediction of diseases using machine learning (ML) is rapidly being
+adopted in microbiome studies. However, the estimated performance
 associated with these ML models is likely over-optimistic. Moreover,
@@
 of selecting ML models that reflect the goal of the study. These
-concepts are not particular to the study of health outcomes but can also
-be applied to environmental microbiology studies.
+concepts are not particular to the study of human health but can also be
+applied to environmental microbiology studies.
 
@@
 understand that the benefit of more complex, less interpretable models
-may be minimal (29, 30). It is important for researchers to justify
+may be minimal (29--31). It is important for researchers to justify
 their choice of modeling approach.
 
-To showcase a rigorous ML pipeline and to shed light on how ML model
-selection can affect modeling results, we performed an empirical
-analysis comparing seven modeling approaches with the same dataset and
-pipeline. We built three linear models with different forms of
-regularization: L2-regularized logistic regression and L1 and
-L2-regularized support vector machines (SVM) with a linear kernel. We
-also trained four non-linear models: SVM with radial basis function
-kernel, a decision tree, random forest, and gradient boosted trees. We
-compared their predictive performance, interpretability, and training
-time. To demonstrate the performance of these modeling approaches and
-our pipeline, we used data from a previously published study that sought
-to classify individuals as having healthy colons or colonic lesions
-based on the 16S rRNA gene sequences collected from fecal samples (4).
-This dataset was selected because it is a relatively large collection of
-individuals (N=490) connected to a clinically significant disease where
-there is ample evidence that the disease is driven by variation in the
-microbiome (2, 4, 5, 31). With this dataset, we developed an ML pipeline
-that can be used in many different scenarios for training and evaluating
-models. This framework can be easily applied to other host-associated
-and environmental microbiome datasets.
+In this study, we provided steps toward standardization of machine
+learning methods for microbiome studies which are often poorly
+documented and executed. To showcase a rigorous ML pipeline and to shed
+light on how ML model selection can affect modeling results, we
+performed an empirical analysis comparing seven modeling approaches with
+the same dataset and pipeline. We built three linear models with
+different forms of regularization: L2-regularized logistic regression
+and L1 and L2-regularized support vector machines (SVM) with a linear
+kernel. We also trained four non-linear models: SVM with radial basis
+function kernel, a decision tree, random forest, and gradient boosted
+trees. We compared their predictive performance, interpretability, and
+training time. To demonstrate the performance of these modeling
+approaches and our pipeline on a case study, we used a data from a
+previously published study that sought to classify individuals as having
+healthy colons or colonic lesions based on the 16S rRNA gene sequences
+collected from fecal samples (4). This dataset was selected because it
+is a relatively large collection of individuals (N=490) connected to a
+clinically significant disease where there is ample evidence that the
+disease is driven by variation in the microbiome (2, 4, 5, 32). With
+this dataset, we developed an ML pipeline that can be used in many
+different scenarios for training and evaluating models. This framework
+can be easily applied to other host-associated and environmental
+microbiome datasets. We hope that this study \ldots{}
 
@@
 methods. Regularization helps to prevent overfitting by penalizing a
-model that fits the training data too well (32). For regularized
+model that fits the training data too well (33). For regularized
 logistic regression and SVM with a linear kernel, we used
@@
 ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
-significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
-regression, XGBoost, L2-regularized SVM with linear and radial basis
-function kernel AUROC values were not significantly different from one
-another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
-0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
-{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
-kernel and decision tree had significantly lower AUROC values than the
-other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
-0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
-these results demonstrate that the most complex model (XGBoost) did not
-have the best performance and that the most interpretable models
+significantly (p=0.5; The p-value was manually calculated using the
+sampling distribution of the test statistic under the null hypothesis)
+(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
+L2-regularized SVM with linear and radial basis function kernel AUROC
+values were not significantly different from one another and had median
+AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
+0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
+0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
+decision tree had significantly lower AUROC values than the other ML
+models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
+0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
+results demonstrate that the most complex model (XGBoost) did not have
+the best performance and that the most interpretable models
 (L2-regularized logistic regression and L2-regularized SVM with linear
@@
 the degree to which humans can understand the reasons behind a model
-prediction (33--35). Because we often use ML models not just to predict
+prediction (34--36). Because we often use ML models not just to predict
 a health outcome, but also to identify potential biomarkers for disease,
@@
 Second, to analyze non-linear models, we interpreted the feature
-importance using permutation importance (36). Whereas the absolute
+importance using permutation importance (37). Whereas the absolute
 feature weights were determined from the trained models, here we
@@
 ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (37). To capture all
+correlated OTUs are influenced by one another (38). To capture all
 important features, once we identify highly ranked OTUs, we should
@@
 predict fecal short-chain fatty acid concentrations based on microbiome
-data (38). Our analysis was also limited to shallow learning methods and
+data (39). Our analysis was also limited to shallow learning methods and
 did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (12, 39, 40), but microbiome datasets
+learning methods hold promise (12, 40, 41), but microbiome datasets
 often suffer from having many features and small sample sizes, which
@@
 predictions is a significantly more challenging and distinct undertaking
-(41). For example, we currently lack standardized methods to collect
+(42). For example, we currently lack standardized methods to collect
 patient samples, generate sequence data, and report clinical data. We
@@
 considerable memory, and result in different OTU assignments as new data
-are added (42). Finally, we also need independent validation cohorts to
+are added (43). Finally, we also need independent validation cohorts to
 test the performance of a diagnostic model. To realize the potential for
@@
 whereas the SRN class included patients with advanced adenomas or
-carcinomas (43). Patients with an adenoma greater than 1 cm, more than
+carcinomas (44). Patients with an adenoma greater than 1 cm, more than
 three adenomas of any size, or an adenoma with villous histology were
-classified as having advanced adenomas (43). There were 172 patients
+classified as having advanced adenomas (44). There were 172 patients
 with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
@@
 Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
-generated by Sze et al (44), who processed the 16S rRNA sequences in
+generated by Sze et al (45), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (42, 45, 46);
+to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 random search or recently developed tools such as Hyperband to identify
-good hyperparameter configurations (47).
+good hyperparameter configurations (48).
 
@@
 
+\hypertarget{ref-Quinn847475}{}
+31. \textbf{Quinn TP}, \textbf{Erb I}. 2019. Another look at
+microbemetabolite interactions: How scale invariant correlations can
+outperform a neural network. bioRxiv.
+doi:\href{https://doi.org/10.1101/847475}{10.1101/847475}.
+
 \hypertarget{ref-knights_human-associated_2011}{}
-31. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
+32. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
 \textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
@@
 \hypertarget{ref-Ng:2004:FSL:1015330.1015435}{}
-32. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
+33. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
 and rotational invariance, pp. 78. \emph{In} Proceedings of the
@@
 \hypertarget{ref-miller_explanation_2017}{}
-33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
+34. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
 Insights from the social sciences. arXiv:170607269 {[}cs{]}.
@@
 \hypertarget{ref-ribeiro_why_2016}{}
-34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
+35. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
 ``Why should i trust you?'': Explaining the predictions of any
@@
 \hypertarget{ref-nori_interpretml:_2019}{}
-35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
+36. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
 \textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
@@
 \hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
-36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
+37. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
 \textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
@@
 \hypertarget{ref-dormann_collinearity:_2013}{}
-37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+38. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 \hypertarget{ref-sze_fecal_2019}{}
-38. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
@@
 \hypertarget{ref-kocheturov_massive_2019}{}
-39. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \hypertarget{ref-kim_improved_2018}{}
-40. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 \hypertarget{ref-wiens_no_2019}{}
-41. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
+42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
 \textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-42. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-redwood_stool_2016}{}
-43. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
 \textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
@@
 \hypertarget{ref-sze_leveraging_2018}{}
-44. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-45. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-46. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-47. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
@@
 vertical red line highlights the bins where there AUROC difference
-between the two model is 0. The p-value was calculated for a double tail
-event. \newpage
+between the two model is 0. The p-value was manually calculated using
+the sampling distribution of the test statistic under the null
+hypothesis. We are testing how often random forest performs more
+accurately than L2-regularized logistic regression. We are assuming that
+the distribution of the differences between the AUROC values of random
+forest and L2 logistic regression is symmetric about 0, therefore we are
+calculating the p-value for a double-tail event. \newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

### 2020-01-20 submission/manuscript.Rmd
update for rebuttal

@@
 
-__Interpretation of each ML model.__ Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019].
-Because we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease, model interpretation becomes crucial for microbiome studies.
+__Interpretation of each ML model.__  To evaluate the overall structure of the CRC dataset and show that the structure of the human microbiome is remarkably variable among individuals with SRNs, we analyzed the Bray-Curtis distances with an NMDS ordination [Figure S5].
+The lack of seperation between healthy patients and patients with SRNs, show that traditional statistical approaches do not help us to adequately address disease ecology.
+This is why we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease.
+Therefore, model interpretation becomes crucial for microbiome studies.
+Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019].
 ML models often decrease in interpretability as they increase in complexity.
@@
 It was encouraging that many of the highest-ranked OTUs were shared across these three models (e.g., OTUs 50, 426, 609, 822, 1239).
-The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU.
+The great benefit of this approach was knowing the sign and magnitude of each OTU coefficient in the trained model.
+This allowed us to immidiately interpret negative and positive coefficient signs as protective and risk factors, respectively and the magnitude as the impact of these factors.
 However, this approach is limited to linear models or models with prespecified interaction terms.
@@
 
-To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S5].
-When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods.
+To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S6].
+When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S6], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods.
 Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods.
 These results indicate that both methods are consistent in selecting the most important OTUs.
+
+We also compared the top 20 OTUs selected by permutation importance in L2-regularized logistic regression and the highest performing tree-based models, random forest and XGBoost.
+2 and 5 OTUs, respectively, were shared among the models.
+These results indicate that we were able to identify a few more OTUs that are deemed important by both linear and non-linear models when we used permutation importance as our interpretation method.
+
+Figure S7:
 
@@
 We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features.
-It is important to know the correlation structures of the data to avoid misinterpreting the models.
+We are choosing to leave the decision to define correlation and setting up correlations thresholds to the researchers for their own microbiome questions.
+However, we would like to urge them to investigate the correlation structures of the data to avoid misinterpreting the models.
 This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome.
+True causal mechanisms (e.g.
+role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020].
 To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.
@@
 
-**Figure S5.
+**Figure S5. NMDS ordination with Bray-Curtis distances.**
+
+\newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
+
+**Figure S6.
 Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.

### 2020-01-20 submission/manuscript.tex
update for rebuttal

@@
 \IfFileExists{microtype.sty}{%
-\usepackage[]{microtype}
+\usepackage{microtype}
 \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-\hypersetup{
+\usepackage[margin=1in]{geometry}
+\usepackage{hyperref}
+\hypersetup{unicode=true,
             pdftitle={Effective application of machine learning to a microbiome-based classification problem},
@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage[margin=1in]{geometry}
 \usepackage{graphicx,grffile}
@@
 
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
+
+%%% Change title format to be more compact
+\usepackage{titling}
+
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
+
+\setlength{\droptitle}{-2em}
+
+  \title{\textbf{Effective application of machine learning to a microbiome-based
+classification problem}}
+    \pretitle{\vspace{\droptitle}\centering\huge}
+  \posttitle{\par}
+    \author{}
+    \preauthor{}\postauthor{}
+    \date{}
+    \predate{}\postdate{}
 
@@
 \floatplacement{figure}{H} % Keep the figure up top of the page
-
-\title{\textbf{Effective application of machine learning to a microbiome-based
-classification problem}}
-\author{}
-\date{\vspace{-2.5em}}
 
@@
 
-\textbf{Interpretation of each ML model.} Interpretability is related to
-the degree to which humans can understand the reasons behind a model
-prediction (34--36). Because we often use ML models not just to predict
-a health outcome, but also to identify potential biomarkers for disease,
-model interpretation becomes crucial for microbiome studies. ML models
-often decrease in interpretability as they increase in complexity. In
-this study, we used two methods to help interpret our models.
+\textbf{Interpretation of each ML model.} To evaluate the overall
+structure of the CRC dataset and show that the structure of the human
+microbiome is remarkably variable among individuals with SRNs, we
+analyzed the Bray-Curtis distances with an NMDS ordination {[}Figure
+S5{]}. The lack of seperation between healthy patients and patients with
+SRNs, show that traditional statistical approaches do not help us to
+adequately address disease ecology. This is why we often use ML models
+not just to predict a health outcome, but also to identify potential
+biomarkers for disease. Therefore, model interpretation becomes crucial
+for microbiome studies. Interpretability is related to the degree to
+which humans can understand the reasons behind a model prediction
+(34--36). ML models often decrease in interpretability as they increase
+in complexity. In this study, we used two methods to help interpret our
+models.
 
@@
 highest-ranked OTUs were shared across these three models (e.g., OTUs
-50, 426, 609, 822, 1239). The benefit of this approach was that the
-results of the analysis were based on the trained model parameters and
-provided information regarding the sign and magnitude of the impact of
-each OTU. However, this approach is limited to linear models or models
-with prespecified interaction terms.
+50, 426, 609, 822, 1239). The great benefit of this approach was knowing
+the sign and magnitude of each OTU coefficient in the trained model.
+This allowed us to immidiately interpret negative and positive
+coefficient signs as protective and risk factors, respectively and the
+magnitude as the impact of these factors. However, this approach is
+limited to linear models or models with prespecified interaction terms.
 
@@
 
+We also compared the top 20 OTUs selected by permutation importance in
+L2-regularized logistic regression and the highest performing tree-based
+models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
+shared among the models. These results indicate that we were able to
+identify a few more OTUs that are deemed important by both linear and
+non-linear models when we used permutation importance as our
+interpretation method.
+
 \textbf{The computational efficiency of each ML model.} We compared the
@@
 the correlation threshold to further investigate the relationships among
-correlated features. It is important to know the correlation structures
-of the data to avoid misinterpreting the models. This is likely to be a
-particular problem with shotgun metagenomic datasets where collinearity
-will be more pronounced due to many genes being correlated with one
-another because they come from the same chromosome. To identify the true
+correlated features. We are choosing to leave the decision to define
+correlation and setting up correlations thresholds to the researchers
+for their own microbiome questions. However, we would like to urge them
+to investigate the correlation structures of the data to avoid
+misinterpreting the models. This is likely to be a particular problem
+with shotgun metagenomic datasets where collinearity will be more
+pronounced due to many genes being correlated with one another because
+they come from the same chromosome. True causal mechanisms (e.g.~role of
+microbiome in colorectal cancer) cannot be explained solely by the
+highest performing machine learning model (39). To identify the true
 underlying microbial factors of a disease, it is crucial to follow up on
@@
 predict fecal short-chain fatty acid concentrations based on microbiome
-data (39). Our analysis was also limited to shallow learning methods and
+data (40). Our analysis was also limited to shallow learning methods and
 did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (12, 40, 41), but microbiome datasets
+learning methods hold promise (12, 41, 42), but microbiome datasets
 often suffer from having many features and small sample sizes, which
@@
 predictions is a significantly more challenging and distinct undertaking
-(42). For example, we currently lack standardized methods to collect
+(43). For example, we currently lack standardized methods to collect
 patient samples, generate sequence data, and report clinical data. We
@@
 considerable memory, and result in different OTU assignments as new data
-are added (43). Finally, we also need independent validation cohorts to
+are added (44). Finally, we also need independent validation cohorts to
 test the performance of a diagnostic model. To realize the potential for
@@
 whereas the SRN class included patients with advanced adenomas or
-carcinomas (44). Patients with an adenoma greater than 1 cm, more than
+carcinomas (45). Patients with an adenoma greater than 1 cm, more than
 three adenomas of any size, or an adenoma with villous histology were
-classified as having advanced adenomas (44). There were 172 patients
+classified as having advanced adenomas (45). There were 172 patients
 with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
@@
 Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
-generated by Sze et al (45), who processed the 16S rRNA sequences in
+generated by Sze et al (46), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
+to OTUs at 97\% similarity using the OptiClust algorithm (44, 47, 48);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 random search or recently developed tools such as Hyperband to identify
-good hyperparameter configurations (48).
+good hyperparameter configurations (49).
 
@@
 
+\hypertarget{ref-li_accurate_2020}{}
+39. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
+Accurate data-driven prediction does not mean high reproducibility. Nat
+Mach Intell \textbf{2}:13--15.
+doi:\href{https://doi.org/10.1038/s42256-019-0140-2}{10.1038/s42256-019-0140-2}.
+
 \hypertarget{ref-sze_fecal_2019}{}
-39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+40. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
@@
 \hypertarget{ref-kocheturov_massive_2019}{}
-40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+41. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \hypertarget{ref-kim_improved_2018}{}
-41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+42. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 \hypertarget{ref-wiens_no_2019}{}
-42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
+43. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
 \textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+44. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \hypertarget{ref-redwood_stool_2016}{}
-44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+45. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
 \textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
@@
 \hypertarget{ref-sze_leveraging_2018}{}
-45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+46. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+47. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+48. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \hypertarget{ref-li_hyperband:_2016}{}
-48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+49. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
@@
 
-\textbf{Figure S5. Interpretation of the linear ML models with
+\textbf{Figure S5. NMDS ordination with Bray-Curtis distances.}
+
+\newpage
+
+\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
+
+\textbf{Figure S6. Interpretation of the linear ML models with
 permutation importance.} (A) L1-regularized SVM with linear kernel, (B)

### 2020-01-20 code/learning/FigureS5.R

@@ -1,223 +1,56 @@
-# Author: Begum Topcuoglu
-# Date: 2018-02-13
+################################################################################
 #
-######################################################################
-# This script plots permutation importance results of linear models
-######################################################################
-
-
-######################################################################
-#----------------- Read in necessary libraries -------------------#
-######################################################################
-deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
-for (dep in deps){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
-  }
-  library(dep, verbose=FALSE, character.only=TRUE)
-}
-######################################################################
-#----------------- Call the functions we will use -----------------#
-######################################################################
-
-source("code/learning/functions.R")
-
-######################################################################
-#--------------Run the functions and plot importance ----------#
-######################################################################
-
-# ----------- Read in saved combined feature importances ---------->
-# List the important features files by defining a pattern in the path
-
-# Correlated files are:
-# For linear models these files have the feature weights and coefficients.
-# For non-linear models these files have the correlated OTUs grouped together
-cor_files <- list.files(path= 'data/process', pattern='combined_all_imp_features_cor_.*', full.names = TRUE)
-
-# Non-correlated files are:
-# For linear models these files have the non-correlated OTUs permutation importance results
-# For non-linear models these files have non-correlated OTUs permutation importance results
-non_cor_files <-  list.files(path= 'data/process', pattern='combined_all_imp_features_non_cor_.*', full.names = TRUE)
-# -------------------------------------------------------------------->
-
-# ----------- Loops to re-organize feature importance info ---------->
[…]

### 2020-01-20 code/learning/FigureS6.R

@@ -1,95 +1,223 @@
+# Author: Begum Topcuoglu
+# Date: 2018-02-13
+#
 ######################################################################
-#---------- Compare training time with subsampling results -------#
+# This script plots permutation importance results of linear models
 ######################################################################
 
-# Load in needed functions and libraries
-source('code/learning/functions.R')
-# detach("package:randomForest", unload=TRUE) to run
 
-
-# -------------------- Read files ------------------------------------>
-grab_number <- function(file){
-  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
-    str_extract("[^_]*$")
-}
-
-# Read in files as delim that are saved in a list with a pattern
-read_files <- function(filenames){
-  for(file in filenames){
-    number <- grab_number(file)
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',') %>% 
-      mutate(number = number)
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
+deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
   }
-  return(data)
+  library(dep, verbose=FALSE, character.only=TRUE)
 }
+######################################################################
+#----------------- Call the functions we will use -----------------#
+######################################################################
+
+source("code/learning/functions.R")
+
+######################################################################
+#--------------Run the functions and plot importance ----------#
+######################################################################
+
+# ----------- Read in saved combined feature importances ---------->
+# List the important features files by defining a pattern in the path
+
+# Correlated files are:
+# For linear models these files have the feature weights and coefficients.
[…]

### 2020-01-20 code/learning/FigureS7.R

@@ -1,0 +1,95 @@
+######################################################################
+#---------- Compare training time with subsampling results -------#
+######################################################################
+
+# Load in needed functions and libraries
+source('code/learning/functions.R')
+# detach("package:randomForest", unload=TRUE) to run
+
+
+# -------------------- Read files ------------------------------------>
+grab_number <- function(file){
+  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
+    str_extract("[^_]*$")
+}
+
+# Read in files as delim that are saved in a list with a pattern
+read_files <- function(filenames){
+  for(file in filenames){
+    number <- grab_number(file)
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>% 
+      mutate(number = number)
+  }
+  return(data)
+}
+# -------------------------------------------------------------------->
+
+# Read in the traintime 100 splits.
+logit_files <- list.files(path= 'data/process/subsampling', pattern='combined_traintime_L2_Logistic_Regression_.*', full.names = TRUE)
+rf_files <- list.files(path= 'data/process/subsampling', pattern='combined_traintime_Random_Forest_.*', full.names = TRUE)
+
+
+logit_performance <- map_df(logit_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+logit_performance$number <-  as.factor(logit_performance$number)
+logit_performance$number <- factor(logit_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+rf_performance <- map_df(rf_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+rf_performance$number <-  as.factor(rf_performance$number)
+rf_performance$number <- factor(rf_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+
+
+logit_plot <- ggplot(logit_performance, aes(x = number, y = x)) +
+  geom_boxplot(alpha=0.6, fatten = 4, fill="darkred") +
+  geom_smooth(method="lm", aes(color="Exp Model"), formula= (y ~ exp(x)), se=FALSE, linetype = 1) +
+  scale_y_continuous(name = "L2-regularized logistic regression 
+training time (hours)",
+                     breaks = seq(0, 0.2, 0.05),
[…]

### 2020-01-21 submission/manuscript.Rmd
updates for rebuttal

@@
 These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances.
-Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.
+Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing accessible and open-source ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.
 
@@
 In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed.
-To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline.
+To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing the predictive performance, interpretability and training times of seven modeling approaches with the same dataset and pipeline.
 We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
@@
 This framework can be easily applied to other host-associated and environmental microbiome datasets.
-We hope that this study ...
+We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Figure S1] to urge the audience to be diligent with with their study design, model selection, development, evaluation and interpretation.
 
@@
 The training data were used to build and select the models, and the test set was used for evaluating the model.
-
 We trained seven different models using the training data [Table 1].
-We focused on different classification algorithms and regularization methods.
-Regularization helps to prevent overfitting by penalizing a model that fits the training data too well [@Ng:2004:FSL:1015330.1015435].
-For regularized logistic regression and SVM with a linear kernel, we used L2-regularization to keep all potentially important features.
-For comparison, we also trained an L1-regularized SVM with a linear kernel.
-L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero).
-To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including, a decision tree, a random forest, and gradient boosted trees (XGBoost) and an SVM with a non-linear kernel.
 
@@
 Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data.
-Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data.
+Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data [Figure S6].
 We then calculated how much the predictive performance of the model (i.e., testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted.
@@
 
-To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S6].
+To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S7].
 When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S6], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods.
@@
 This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome.
-True causal mechanisms (e.g.
-role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020].
+True causal mechanisms (e.g., role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020].
 To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.
@@
 
-Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
+Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
 
@@
 The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here.
-The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, salmon were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only.
+The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, green were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only.
 For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
@@
 
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
+\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population to which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
 
@@
 
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g.
-when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
+\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g., when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
 
@@
 **Figure S3.
-Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.** This histogram shows the number of data-splits in each bin.
-The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25.
-The vertical red line highlights the bins where there AUROC difference between the two model is 0.
+Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.**  In 75% of data-splits, the AUROC of random forest was greater than that of L2-regularized logistic regression.
 The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis.
-We are testing how often random forest performs more accurately than L2-regularized logistic regression.
-We are assuming that the distribution of the differences between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore we are calculating the p-value for a double-tail event.
+We tested how often random forest performed more accurately than L2-regularized logistic regression.
+The null hypothesis is that the distribution of the difference between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore the p-value was calculated for a double-tail event.
 \newpage
@@
 
-**Figure S5. NMDS ordination with Bray-Curtis distances.**
+**Figure S5.
+NMDS ordination of Bray-Curtis distances.** NMDS ordination relating the community structures of the fecal microbiota from 490 patients (261 patients with normal colonoscopies and 229 patients who have screen relevant neoplasias; SRNs).
 
@@
 **Figure S6.
+Permutation importance analysis.** Permutation importance analysis measures the decrease in the predictive performance of the model after we permute (A) a feature’s or (B) a group of correlated features' values, which breaks the relationship between the feature and the diagnosis.
+
+\newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
+
+**Figure S7.
 Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
+
+\newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
+
+**Figure S8.
+Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.

### 2020-01-21 submission/manuscript.tex
updates for rebuttal

--- 2020-01-17 submission/manuscript.tex
@@
 pitfalls including validating their models on independent datasets (8,
-19, 20) and introducing ways to better use ML tools (21--24), more work
-is needed to improve reproducibility further and minimize overestimating
-for model performance.
+19, 20) and introducing accessible and open-source ML tools (21--24),
+more work is needed to improve reproducibility further and minimize
+overestimating for model performance.
 
@@
 light on how ML model selection can affect modeling results, we
-performed an empirical analysis comparing seven modeling approaches with
+performed an empirical analysis comparing the predictive performance,
+interpretability and training times of seven modeling approaches with
 the same dataset and pipeline. We built three linear models with
@@
 can be easily applied to other host-associated and environmental
-microbiome datasets. We hope that this study \ldots{}
+microbiome datasets. We also provided an aspirational rubric for
+evaluating the rigor of ML practices applied to microbiome data
+{[}Figure S1{]} to urge the audience to be diligent with with their
+study design, model selection, development, evaluation and
+interpretation.
 
@@
 data were used to build and select the models, and the test set was used
-for evaluating the model.
-
-We trained seven different models using the training data {[}Table 1{]}.
-We focused on different classification algorithms and regularization
-methods. Regularization helps to prevent overfitting by penalizing a
-model that fits the training data too well (33). For regularized
-logistic regression and SVM with a linear kernel, we used
-L2-regularization to keep all potentially important features. For
-comparison, we also trained an L1-regularized SVM with a linear kernel.
-L1-regularization on microbiome data led to a sparser solution (i.e.,
-forced many coefficients to zero). To explore the potential for
-non-linear relationships among features to improve classification, we
-trained tree-based models including, a decision tree, a random forest,
-and gradient boosted trees (XGBoost) and an SVM with a non-linear
-kernel.
+for evaluating the model. We trained seven different models using the
+training data {[}Table 1{]}.
 
@@
 
-\textbf{Interpretation of each ML model.} Interpretability is related to
-the degree to which humans can understand the reasons behind a model
-prediction (34--36). Because we often use ML models not just to predict
-a health outcome, but also to identify potential biomarkers for disease,
-model interpretation becomes crucial for microbiome studies. ML models
-often decrease in interpretability as they increase in complexity. In
-this study, we used two methods to help interpret our models.
+\textbf{Interpretation of each ML model.} To evaluate the overall
+structure of the CRC dataset and show that the structure of the human
+microbiome is remarkably variable among individuals with SRNs, we
+analyzed the Bray-Curtis distances with an NMDS ordination {[}Figure
+S5{]}. The lack of seperation between healthy patients and patients with
+SRNs, show that traditional statistical approaches do not help us to
+adequately address disease ecology. This is why we often use ML models
+not just to predict a health outcome, but also to identify potential
+biomarkers for disease. Therefore, model interpretation becomes crucial
+for microbiome studies. Interpretability is related to the degree to
+which humans can understand the reasons behind a model prediction
+(33--35). ML models often decrease in interpretability as they increase
+in complexity. In this study, we used two methods to help interpret our
+models.
 
@@
 highest-ranked OTUs were shared across these three models (e.g., OTUs
-50, 426, 609, 822, 1239). The benefit of this approach was that the
-results of the analysis were based on the trained model parameters and
-provided information regarding the sign and magnitude of the impact of
-each OTU. However, this approach is limited to linear models or models
-with prespecified interaction terms.
+50, 426, 609, 822, 1239). The great benefit of this approach was knowing
+the sign and magnitude of each OTU coefficient in the trained model.
+This allowed us to immidiately interpret negative and positive
+coefficient signs as protective and risk factors, respectively and the
+magnitude as the impact of these factors. However, this approach is
+limited to linear models or models with prespecified interaction terms.
 
 Second, to analyze non-linear models, we interpreted the feature
-importance using permutation importance (37). Whereas the absolute
+importance using permutation importance (36). Whereas the absolute
 feature weights were determined from the trained models, here we
@@
 permuted groups of perfectly correlated features together and other
-features individually across the two groups in the held-out test data.
-We then calculated how much the predictive performance of the model
-(i.e., testing AUROC values) decreased when each OTU or group of OTUs
-was randomly permuted. We ranked the OTUs based on how much the median
-testing AUROC decreased when it was permuted; the OTU with the largest
-decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
-largest impact, there was only one OTU (OTU 822) that was shared among
-all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
-were important in each of the tree-based models. Similarly, the random
-forest and XGBoost models shared four of the most important OTUs (OTUs
-2, 12, 361, 477). Permutation analysis results also revealed that with
-the exception of the decision tree model, removal of any individual OTU
-had minimal impact on model performance. For example, if OTU 367 was
-permuted across the samples in the decision tree model, the median AUROC
-dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
-in the random forest model, the AUROC only dropped from 0.695 to 0.680,
-which indicated high degree of collinearity in the dataset. Permutation
-analysis allowed us to gauge the importance of each OTU in non-linear
-models and partially account for collinearity by grouping correlated
-OTUs to determine their impact as a group.
+features individually across the two groups in the held-out test data
+{[}Figure S6{]}. We then calculated how much the predictive performance
+of the model (i.e., testing AUROC values) decreased when each OTU or
+group of OTUs was randomly permuted. We ranked the OTUs based on how
+much the median testing AUROC decreased when it was permuted; the OTU
+with the largest decrease ranked highest {[}Figure 4{]}. Among the
+twenty OTUs with the largest impact, there was only one OTU (OTU 822)
+that was shared among all of the models; however, we found three OTUs
+(OTUs 58, 110, 367) that were important in each of the tree-based
+models. Similarly, the random forest and XGBoost models shared four of
+the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis
+results also revealed that with the exception of the decision tree
+model, removal of any individual OTU had minimal impact on model
+performance. For example, if OTU 367 was permuted across the samples in
+the decision tree model, the median AUROC dropped from 0.601 to 0.525.
+In contrast, if the same OTU was permuted in the random forest model,
+the AUROC only dropped from 0.695 to 0.680, which indicated high degree
+of collinearity in the dataset. Permutation analysis allowed us to gauge
+the importance of each OTU in non-linear models and partially account
+for collinearity by grouping correlated OTUs to determine their impact
+as a group.
 
@@
 methods, we used permutation importance to interpret the linear models
-{[}Figure S5{]}. When we analyzed the L1-regularized SVM with linear
+{[}Figure S7{]}. When we analyzed the L1-regularized SVM with linear
 kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g., OTU
+permutation importance {[}Figure S6{]}, 17 of the 20 top OTUs (e.g., OTU
 609, 822, 1239) were deemed important by both interpretation methods.
@@
 consistent in selecting the most important OTUs.
+
+We also compared the top 20 OTUs selected by permutation importance in
+L2-regularized logistic regression and the highest performing tree-based
+models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
+shared among the models. These results indicate that we were able to
+identify a few more OTUs that are deemed important by both linear and
+non-linear models when we used permutation importance as our
+interpretation method.
+
+Figure S7:
 
@@
 ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (38). To capture all
+correlated OTUs are influenced by one another (37). To capture all
 important features, once we identify highly ranked OTUs, we should
@@
 the correlation threshold to further investigate the relationships among
-correlated features. It is important to know the correlation structures
-of the data to avoid misinterpreting the models. This is likely to be a
-particular problem with shotgun metagenomic datasets where collinearity
-will be more pronounced due to many genes being correlated with one
-another because they come from the same chromosome. To identify the true
+correlated features. We are choosing to leave the decision to define
+correlation and setting up correlations thresholds to the researchers
+for their own microbiome questions. However, we would like to urge them
+to investigate the correlation structures of the data to avoid
+misinterpreting the models. This is likely to be a particular problem
+with shotgun metagenomic datasets where collinearity will be more
+pronounced due to many genes being correlated with one another because
+they come from the same chromosome. True causal mechanisms (e.g., role
+of microbiome in colorectal cancer) cannot be explained solely by the
+highest performing machine learning model (38). To identify the true
 underlying microbial factors of a disease, it is crucial to follow up on
@@
 
-\hypertarget{ref-Ng:2004:FSL:1015330.1015435}{}
-33. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
-and rotational invariance, pp. 78. \emph{In} Proceedings of the
-twenty-first international conference on machine learning. ACM, New
-York, NY, USA.
-
 \hypertarget{ref-miller_explanation_2017}{}
-34. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
+33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
 Insights from the social sciences. arXiv:170607269 {[}cs{]}.
@@
 \hypertarget{ref-ribeiro_why_2016}{}
-35. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
+34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
 ``Why should i trust you?'': Explaining the predictions of any
@@
 \hypertarget{ref-nori_interpretml:_2019}{}
-36. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
+35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
 \textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
@@
 \hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
-37. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
+36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
 \textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
@@
 \hypertarget{ref-dormann_collinearity:_2013}{}
-38. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.
+
+\hypertarget{ref-li_accurate_2020}{}
+38. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
+Accurate data-driven prediction does not mean high reproducibility. Nat
+Mach Intell \textbf{2}:13--15.
+doi:\href{https://doi.org/10.1038/s42256-019-0140-2}{10.1038/s42256-019-0140-2}.
 
@@
 the OTUs that were shared among the different models; yellow were OTUs
-that were shared among all the non-linear models, salmon were OTUs that
+that were shared among all the non-linear models, green were OTUs that
 were shared among the tree-based models, green were the OTUs shared
@@
 logistic regression and random forest for each of the hundred
-data-splits.} This histogram shows the number of data-splits in each
-bin. The percentage of dataplits where the difference between random
-forest and L2-regularized logistic regression AUROC values was higher
-than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The
-vertical red line highlights the bins where there AUROC difference
-between the two model is 0. The p-value was manually calculated using
-the sampling distribution of the test statistic under the null
-hypothesis. We are testing how often random forest performs more
-accurately than L2-regularized logistic regression. We are assuming that
-the distribution of the differences between the AUROC values of random
-forest and L2 logistic regression is symmetric about 0, therefore we are
-calculating the p-value for a double-tail event. \newpage
+data-splits.} In 75\% of data-splits, the AUROC of random forest was
+greater than that of L2-regularized logistic regression. The p-value was
+manually calculated using the sampling distribution of the test
+statistic under the null hypothesis. We tested how often random forest
+performed more accurately than L2-regularized logistic regression. The
+null hypothesis is that the distribution of the difference between the
+AUROC values of random forest and L2 logistic regression is symmetric
+about 0, therefore the p-value was calculated for a double-tail event.
+\newpage
 \includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
@@
 
-\textbf{Figure S5. Interpretation of the linear ML models with
+\textbf{Figure S5. NMDS ordination with Bray-Curtis distances.}
+
+\newpage
+
+\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
+
+\textbf{Figure S6. Permutation importance analysis.} Permutation
+importance analysis measures the decrease in the predictive performance
+of the model after we permute (A) a feature's or (B) a group of
+correlated features' values, which breaks the relationship between the
+feature and the actual outcome.
+
+\newpage
+
+\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
+
+\textbf{Figure S7. Interpretation of the linear ML models with
 permutation importance.} (A) L1-regularized SVM with linear kernel, (B)

### 2020-01-21 code/learning/Figure4.R

@@ -165,8 +165,8 @@ base_nonlin_plot <-  function(data, name){
     mutate(common_otus = case_when(names %in% intersect_four ~ "all models",
                                    names %in% intersect_three ~ "tree-based models",
                                    names %in% intersect_rf_dt ~ "random forest and decision tree",
-                                   names %in% intersect_xgboost_dt ~ "XGBoost and decision tree",
-                                   names %in% intersect_xgboost_rf ~ "XGBoost and random forest",
+                                   names %in% intersect_xgboost_dt ~ "XGBoost, decision tree and RBF SVM",
+                                   names %in% intersect_xgboost_rf ~ "XGBoost, random forest and RBF SVM",
                                    names %in% intersect_rbf_xgboost ~ "XGBoost and RBF SVM",
                                    TRUE ~ "no overlap")) %>%
     group_by(names)
@@ -178,7 +178,7 @@ base_nonlin_plot <-  function(data, name){
     data.frame()
 
   # Define colors for overlapping OTUs
-  cols <- c("all models" = "orange", "tree-based models" = "lightsalmon", "random forest and decision tree" = "lightblue", "XGBoost and decision tree" = "darkseagreen3", "XGBoost and random forest"="red", "XGBoost and RBF SVM" = "hotpink" , "no overlap"="white")
+  cols <- c("all models" = "orange", "tree-based models" = "darkolivegreen3", "random forest and decision tree" = "deepskyblue2", "XGBoost, decision tree and RBF SVM" = "aquamarine2", "XGBoost, random forest and RBF SVM"="red", "XGBoost and RBF SVM" = "lightpink1" , "no overlap"="white")
 
   # Plot the figure
   plot <- ggplot() +
@@ -303,16 +303,16 @@ xgboost_plot <- base_nonlin_plot(xgboost, "XGBoost")+
 # ----------------------------------------------------------------------->
 
 # Extract legend
-legend <- get_legend(xgboost_plot + theme(legend.position="bottom", 
+legend <- get_legend(xgboost_plot + guides(fill=guide_legend(ncol=2)) + theme(legend.position="bottom", 
                                           legend.title = element_blank(),
                                           legend.direction = "horizontal",
-                                          legend.justification="center",
-                                          legend.box.just = "bottom",
-                                          legend.text=element_text(size=10),
-                                          legend.background = element_rect(linetype="solid", 
-                                                                           color="black", 
-                                                                           size=0.5),
-                                          legend.box.margin=margin(c(12,12,12,12))))
[…]

### 2020-01-21 code/learning/FigureS3.R

@@ -18,13 +18,13 @@ histogram_p_value <- function(data, model_name_1, model_name_2){
     aes(differences) +
     geom_histogram(colour="black", fill="white", bins=100)+
     #geom_density(alpha=.2, fill="blue") +
-    geom_vline(xintercept = 0, color="red", size=1) +
+    #geom_vline(xintercept = 0, color="red", size=1) +
     #geom_vline(aes(xintercept=mean(differences)),
     #           color="blue", linetype="dashed") + 
     coord_cartesian(xlim=c(-0.2,0.2), ylim=c(0,8)) +
     scale_x_continuous(name= "The difference between random forest and L2-regularized
-    logistic regression AUROC values of each datasplit") +
-    scale_y_continuous(name="Number of datasplits",
+    logistic regression AUROC values of each of the 100 data-splits") +
+    scale_y_continuous(name="Number of data-splits",
                        breaks = seq(0, 8, 1),
                        expand=c(0,0)) +
     theme(panel.background = element_blank(),

### 2020-01-21 code/learning/FigureS6.R

@@ -216,7 +216,7 @@ linear <- plot_grid(l1_plot, l2_plot, logit_plot, labels = c("A", "B", "C"), ali
 
 ggdraw(add_sub(linear, "AUROC with the OTU permuted randomly", size=18, vpadding=grid::unit(0,"lines"), y=5, x=0.65, vjust=4.75))
 
-ggsave("Figure_S6.png", plot = last_plot(), device = 'png', path = 'submission', width = 8, height = 12)
+ggsave("Figure_S7.png", plot = last_plot(), device = 'png', path = 'submission', width = 8, height = 12)
 
 
 

### 2020-01-21 code/learning/FigureS7.R

@@ -1,95 +1,223 @@
+# Author: Begum Topcuoglu
+# Date: 2018-02-13
+#
 ######################################################################
-#---------- Compare training time with subsampling results -------#
+# This script plots permutation importance results of linear models
 ######################################################################
 
-# Load in needed functions and libraries
-source('code/learning/functions.R')
-# detach("package:randomForest", unload=TRUE) to run
 
-
-# -------------------- Read files ------------------------------------>
-grab_number <- function(file){
-  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
-    str_extract("[^_]*$")
-}
-
-# Read in files as delim that are saved in a list with a pattern
-read_files <- function(filenames){
-  for(file in filenames){
-    number <- grab_number(file)
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',') %>% 
-      mutate(number = number)
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
+deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
   }
-  return(data)
+  library(dep, verbose=FALSE, character.only=TRUE)
 }
+######################################################################
+#----------------- Call the functions we will use -----------------#
+######################################################################
+
+source("code/learning/functions.R")
+
+######################################################################
+#--------------Run the functions and plot importance ----------#
+######################################################################
+
+# ----------- Read in saved combined feature importances ---------->
+# List the important features files by defining a pattern in the path
+
+# Correlated files are:
+# For linear models these files have the feature weights and coefficients.
[…]

### 2020-01-22 submission/manuscript.Rmd
updates for rebuttal

@@
 2 and 5 OTUs, respectively, were shared among the models.
-These results indicate that we were able to identify a few more OTUs that are deemed important by both linear and non-linear models when we used permutation importance as our interpretation method.
+These results indicate that we were able to identify important OTUs that are shared across the highest performing linear and non-linear models when we use permutation importance as our interpretation method.
 
-Figure S7:
+We then evaluated the difference in relative abundances of the top 20 OTUs identified in L2-regularized logistic regression and random forest models between healthy patients and patients with SRNs [Figure S8].
+There were minimal differences in relative abundances across OTUs between different diagnoses.
+This supports our claim that it is not possible to differentiate disease vs healthy states by focusing on individual taxa.
 
@@
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
+\includegraphics[height=20cm, width=13cm]{Figure_S8.pdf}
 
 **Figure S8.
-Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
+Relative abundances of the 20 most important OTUs in L2-regularized logistic regression and random forest models.** The most important 20 OTUs were chosen for (A) Random forest and (B) L2-regularized logistic regression models by permutation importance and ranking feature coefficients, respectively.
+The relative abundances of these OTUs were compared based on the diagnosis of the patients.
+The minimal differences betweeen relative abundances for these OTUs show that it is not possible to differentiate disease vs healthy states by focusing on individual taxa.

### 2020-01-22 submission/manuscript.tex
updates for rebuttal

@@
 shared among the models. These results indicate that we were able to
-identify a few more OTUs that are deemed important by both linear and
-non-linear models when we used permutation importance as our
+identify important OTUs that are shared across the highest performing
+linear and non-linear models when we use permutation importance as our
 interpretation method.
 
-Figure S7:
+We then evaluated the difference in relative abundances of the top 20
+OTUs identified in L2-regularized logistic regression and random forest
+models between healthy patients and patients with SRNs {[}Figure S8{]}.
+There were minimal differences in relative abundances across OTUs
+between different diagnoses. This supports our claim that it is not
+possible to differentiate disease vs healthy states by focusing on
+individual taxa.
 
@@
 
-\textbf{Figure S5. NMDS ordination with Bray-Curtis distances.}
+\textbf{Figure S5. NMDS ordination of Bray-Curtis distances.} NMDS
+ordination relating the community structures of the fecal microbiota
+from 490 patients (261 patients with normal colonoscopies and 229
+patients who have screen relevant neoplasias; SRNs).
 
@@
 correlated features' values, which breaks the relationship between the
-feature and the actual outcome.
+feature and the diagnosis.
 
@@
 
+\newpage
+
+\includegraphics[height=20cm, width=13cm]{Figure_S8.pdf}
+
+\textbf{Figure S8. Relative abundances of the 20 most important OTUs in
+L2-regularized logistic regression and random forest models.} The most
+important 20 OTUs were chosen for (A) Random forest and (B)
+L2-regularized logistic regression models by permutation importance and
+ranking feature coefficients, respectively. The relative abundances of
+these OTUs were compared based on the diagnosis of the patients. The
+minimal differences betweeen relative abundances for these OTUs show
+that it is not possible to differentiate disease vs healthy states by
+focusing on individual taxa.
+
 \end{document}

### 2020-01-22 code/learning/FigureS9.R

@@ -1,0 +1,95 @@
+######################################################################
+#---------- Compare training time with subsampling results -------#
+######################################################################
+
+# Load in needed functions and libraries
+source('code/learning/functions.R')
+# detach("package:randomForest", unload=TRUE) to run
+
+
+# -------------------- Read files ------------------------------------>
+grab_number <- function(file){
+  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
+    str_extract("[^_]*$")
+}
+
+# Read in files as delim that are saved in a list with a pattern
+read_files <- function(filenames){
+  for(file in filenames){
+    number <- grab_number(file)
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>% 
+      mutate(number = number)
+  }
+  return(data)
+}
+# -------------------------------------------------------------------->
+
+# Read in the traintime 100 splits.
+logit_files <- list.files(path= 'data/process/subsampling', pattern='combined_traintime_L2_Logistic_Regression_.*', full.names = TRUE)
+rf_files <- list.files(path= 'data/process/subsampling', pattern='combined_traintime_Random_Forest_.*', full.names = TRUE)
+
+
+logit_performance <- map_df(logit_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+logit_performance$number <-  as.factor(logit_performance$number)
+logit_performance$number <- factor(logit_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+rf_performance <- map_df(rf_files, read_files) %>% 
+  mutate(x=x/3600) %>% 
+  group_by(number) 
+rf_performance$number <-  as.factor(rf_performance$number)
+rf_performance$number <- factor(rf_performance$number , c("15", "30", "60", "120", "245", "490"))
+
+
+
+logit_plot <- ggplot(logit_performance, aes(x = number, y = x)) +
+  geom_boxplot(alpha=0.6, fatten = 4, fill="darkred") +
+  geom_smooth(method="lm", aes(color="Exp Model"), formula= (y ~ exp(x)), se=FALSE, linetype = 1) +
+  scale_y_continuous(name = "L2-regularized logistic regression 
+training time (hours)",
+                     breaks = seq(0, 0.2, 0.05),
[…]

### 2020-01-22 code/learning/Figure_S8.R

@@ -1,0 +1,203 @@
+# Author: Begum Topcuoglu
+# Date: 2018-06-06
+#
+######################################################################
+# This script plots the abundances of top 20 logistic and random forest featurs
+######################################################################
+
+
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
+deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+######################################################################
+#----------------- Call the functions we will use -----------------#
+######################################################################
+
+source("code/learning/functions.R")
+
+
+######################################################################
+#-------- Get the full abundance data and metadata -----------------#
+######################################################################
+meta <- read.delim('data/metadata.tsv', header=T, sep='\t') %>%
+  select(sample, Dx_Bin, fit_result)
+# Read in OTU table and remove label and numOtus columns
+shared <- read.delim('data/baxter.0.03.subsample.shared', header=T, sep='\t') %>%
+  select(-label, -numOtus)
+# Merge metadata and OTU table.
+# Group advanced adenomas and cancers together as cancer and normal, high risk normal and non-advanced adenomas as normal
+# Then remove the sample ID column
+data <- inner_join(meta, shared, by=c("sample"="Group")) %>%
+  mutate(dx = case_when(
+    Dx_Bin== "Adenoma" ~ "normal",
+    Dx_Bin== "Normal" ~ "normal",
[…]

### 2020-01-27 submission/manuscript.Rmd
tiff figures

@@
 ---
-title: '**Effective application of machine learning to a microbiome-based classification problem**'
+title: '**A framework for effective application of machine learning to a microbiome-based classification problem**'
 csl: mbio.csl
@@
 
-Running title: Machine learning methods applied to a microbiome study
+Running title: Machine learning framework applied to a microbiome study
 
@@
 However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
+For example it is not possible to classify individuals as having healthy colons or colorectal lesions using Bray-Curtis distances based on the 16S rRNA gene sequences collected from fecal samples [Figure S1].
 This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals.
@@
 This framework can be easily applied to other host-associated and environmental microbiome datasets.
-We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Figure S1] to urge the audience to be diligent with with their study design, model selection, development, evaluation and interpretation.
+We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Table S1] to urge the audience to be diligent with with their study design, model selection, development, evaluation and interpretation.
 
@@
 The five-fold CV was also stratified to maintain the overall case and control distribution.
-We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2].
+We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S2 and S3].
 The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls.
@@
 Default hyperparameter settings in developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task.
-For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning as the default C parameter was 1 [Figure S1].
+For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning as the default C parameter was 1 [Figure S2].
 
@@
 __Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2].
-The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5; The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis) (Figure S3).
+The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5; The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis) (Figure S4).
 Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively.
@@
 
-To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4].
+To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S5].
 The variation in cross-validation performance within both models at smaller sample sizes was larger than when the full collection of samples was used to train and validate the models.
@@
 
-__Interpretation of each ML model.__  To evaluate the overall structure of the CRC dataset and show that the structure of the human microbiome is remarkably variable among individuals with SRNs, we analyzed the Bray-Curtis distances with an NMDS ordination [Figure S5].
-The lack of seperation between healthy patients and patients with SRNs, show that traditional statistical approaches do not help us to adequately address disease ecology.
-This is why we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease.
+__Interpretation of each ML model.__  We often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease.
 Therefore, model interpretation becomes crucial for microbiome studies.
@@
 To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S7].
-When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S6], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods.
+When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S7], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods.
 Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods.
@@
 We compared the performance of the models pairwise by calculating the difference between AUROC values from the same data-split (for 100 data-splits).
-We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S3).
+We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S4).
 
@@
 \newpage
-\includegraphics{Figure_2.png}
+\includegraphics{Figure_2.tiff}
 
@@
 \newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.png}
+\includegraphics[height=17cm, width=12cm]{Figure_3.tiff}
 
@@
 \newpage
-\includegraphics{Figure_4.png}
+\includegraphics{Figure_4.tiff}
 
@@
 \newpage
-\includegraphics{Figure_5.png}
+\includegraphics{Figure_5.tiff}
 
@@
 \end{tabular}
+
 \newpage
+\includegraphics[height=17.5cm, width=13cm]{Figure_S1.tiff}
 
-\includegraphics{Figure_S1.png}
 **Figure S1.
+NMDS ordination of Bray-Curtis distances.** NMDS ordination relating the community structures of the fecal microbiota from 490 patients (261 patients with normal colonoscopies and 229 patients who have screen relevant neoplasias; SRNs).
+
+\newpage
+\includegraphics{Figure_S2.tiff}
+**Figure S2.
 Hyperparameter setting performances for linear models.** (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model.
 The stars represent the highest performing hyperparameter setting for each model.
+
 \newpage
-\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
-
-**Figure S2.
+\includegraphics[height=30cm, width=15cm]{Figure_S3.tiff}
+**Figure S3.
 Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.
 The stars represent the highest performing hyperparameter setting for the models.
+
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
-
-**Figure S3.
+\includegraphics[height=17.5cm, width=13cm]{Figure_S4.tiff}
+**Figure S4.
 Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.**  In 75% of data-splits, the AUROC of random forest was greater than that of L2-regularized logistic regression.
@@
 The null hypothesis is that the distribution of the difference between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore the p-value was calculated for a double-tail event.
+
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
-
-**Figure S4.
+\includegraphics[height=17.5cm, width=13cm]{Figure_S5.tiff}
+**Figure S5.
 Classification performance of ML models across cross-validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects.
 The range among the cross-validation AUROC values within both models at smaller sample sizes were much larger than when the full collection of samples was used to train and validate the models but included the ranges observed with the more complete datasets.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
-
-**Figure S5.
-NMDS ordination of Bray-Curtis distances.** NMDS ordination relating the community structures of the fecal microbiota from 490 patients (261 patients with normal colonoscopies and 229 patients who have screen relevant neoplasias; SRNs).
 
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
-
+\includegraphics[height=17.5cm, width=13cm]{Figure_S6.tiff}
 **Figure S6.
@@
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
-
+\includegraphics[height=17.5cm, width=13cm]{Figure_S7.tiff}
 **Figure S7.
@@
 \newpage
-\includegraphics[height=20cm, width=13cm]{Figure_S8.pdf}
-
+\includegraphics[height=20cm, width=13cm]{Figure_S8.tiff}
 **Figure S8.

### 2020-01-27 code/learning/Figure2.R

@@ -85,4 +85,4 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 
-ggsave("Figure_2.png", plot = performance, device = 'png', path = 'submission', width = 12, height = 9)
+ggsave("Figure_2.tiff", plot = performance, device = 'tiff', path = 'submission', width = 12, height = 9, dpi=300)

### 2020-01-27 code/learning/Figure3.R

@@ -135,4 +135,4 @@ linear <- plot_grid(l1_svm_graph, l2_svm_graph, logit_graph, labels = c("A", "B"
 
 ggdraw(add_sub(linear, "Feature ranks", vpadding=grid::unit(0,"lines"), y=5, x=0.7, vjust=4.75, size=15))
 
-ggsave("Figure_3.png", plot = last_plot(), device = 'png', path = 'submission', width = 6, height = 9.2)
+ggsave("Figure_3.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9.2, dpi=300)

### 2020-01-27 code/learning/Figure4.R

@@ -324,5 +324,5 @@ ggdraw(add_sub(perm_tree_based, "AUROC with the OTU permuted randomly", size=18,
 
 plot <- plot_grid(last_plot(), legend, ncol = 1, rel_heights=c(.92, .08))
 
-ggsave("Figure_4.png", plot = last_plot(), device = 'png', path = 'submission', width = 8, height = 18)
+ggsave("Figure_4.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 8, height = 18, dpi=300)
 

### 2020-01-27 code/learning/Figure5.R

@@ -69,7 +69,7 @@ traintime_plot <- ggplot(traintime_df, aes(x = fct_reorder(model, x), y = x)) +
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 
-ggsave("Figure_5.png", plot = traintime_plot, device = 'png', path = 'submission', width = 4, height = 2)
+ggsave("Figure_5.tiff", plot = traintime_plot, device = 'tiff', path = 'submission', width = 4, height = 2, dpi=300)
 
 
             

### 2020-01-27 code/learning/FigureS1.R

@@ -97,5 +97,5 @@ linear_plots <- fig_data %>%
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 
-ggsave("Figure_S1.png", plot = linear_plots, device = 'png', path = 'submission', width = 7, height = 5)
+ggsave("Figure_S2.tiff", plot = linear_plots, device = 'tiff', path = 'submission', width = 7, height = 5, dpi=300)
 

### 2020-01-27 code/learning/FigureS2.R

@@ -173,4 +173,4 @@ non_linear_models <- plot_grid(dt_plot, rf_plot, rbf_plot, xgboost_plot, labels 
 ######################################################################
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
-ggsave("Figure_S2.png", plot = non_linear_models, device = 'png', path = 'submission', width = 9, height = 7)
+ggsave("Figure_S3.tiff", plot = non_linear_models, device = 'png', path = 'submission', width = 9, height = 7, dpi=300)

### 2020-01-27 code/learning/FigureS3.R

@@ -70,4 +70,4 @@ set.seed(101)
 plot <- histogram_p_value(all, "L2_Logistic_Regression","Random_Forest")
 
 
-ggsave("Figure_S3.png", plot = plot, device = 'png', path = 'submission', width = 12, height = 9)
+ggsave("Figure_S4.tiff", plot = plot, device = 'tiff', path = 'submission', width = 12, height = 9, dpi=300)

### 2020-01-27 code/learning/FigureS4.R

@@ -127,5 +127,5 @@ plots <- plot_grid(logit_plot, rf_plot, labels = c("A", "B"), align = 'v', ncol 
 
 ggdraw(add_sub(plots, "Cross-validation AUROC", vpadding=grid::unit(0,"lines"), y=5, x=0.6, vjust=4.75, size=15))
 
-ggsave("Figure_S4.png", plot = last_plot(), device = 'png', path = 'submission', width = 6, height = 9.2)
+ggsave("Figure_S5.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9.2, dpi=300)
 

### 2020-01-27 code/learning/FigureS5.R

@@ -50,7 +50,7 @@ ggplot(data) +
     axis.text = element_text(size=12),
     panel.border = element_rect(colour = "black", fill=NA, size=1)
   ) +
-  ggsave("Figure_S5.png", plot = last_plot(), device = 'png', path = 'submission')
+  ggsave("Figure_S1.tiff", plot = last_plot(), device = 'tiff', path = 'submission', dpi=300)
 
 
 

### 2020-01-27 code/learning/FigureS6.R

@@ -216,7 +216,7 @@ linear <- plot_grid(l1_plot, l2_plot, logit_plot, labels = c("A", "B", "C"), ali
 
 ggdraw(add_sub(linear, "AUROC with the OTU permuted randomly", size=18, vpadding=grid::unit(0,"lines"), y=5, x=0.65, vjust=4.75))
 
-ggsave("Figure_S7.png", plot = last_plot(), device = 'png', path = 'submission', width = 8, height = 12)
+ggsave("Figure_S6.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 8, height = 12, dpi=300)
 
 
 

### 2020-01-27 code/learning/FigureS7.R

@@ -216,7 +216,7 @@ linear <- plot_grid(l1_plot, l2_plot, logit_plot, labels = c("A", "B", "C"), ali
 
 ggdraw(add_sub(linear, "AUROC with the OTU permuted randomly", size=18, vpadding=grid::unit(0,"lines"), y=5, x=0.65, vjust=4.75))
 
-ggsave("Figure_S7.png", plot = last_plot(), device = 'png', path = 'submission', width = 8, height = 12)
+ggsave("Figure_S7.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 8, height = 12, dpi=300)
 
 
 

### 2020-01-27 code/learning/FigureS8.R

@@ -1,95 +1,203 @@
+# Author: Begum Topcuoglu
+# Date: 2018-06-06
+#
 ######################################################################
-#---------- Compare training time with subsampling results -------#
+# This script plots the abundances of top 20 logistic and random forest featurs
 ######################################################################
 
-# Load in needed functions and libraries
-source('code/learning/functions.R')
-# detach("package:randomForest", unload=TRUE) to run
 
-
-# -------------------- Read files ------------------------------------>
-grab_number <- function(file){
-  regmatches(file, regexpr("[0-9].*[0-9]", file)) %>% 
-    str_extract("[^_]*$")
-}
-
-# Read in files as delim that are saved in a list with a pattern
-read_files <- function(filenames){
-  for(file in filenames){
-    number <- grab_number(file)
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',') %>% 
-      mutate(number = number)
+######################################################################
+#----------------- Read in necessary libraries -------------------#
+######################################################################
+deps = c("cowplot","reshape2", "cowplot", "ggplot2","knitr","rmarkdown","vegan","gtools", "tidyverse");
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
   }
-  return(data)
+  library(dep, verbose=FALSE, character.only=TRUE)
 }
-# -------------------------------------------------------------------->
+######################################################################
+#----------------- Call the functions we will use -----------------#
+######################################################################
 
-# Read in the traintime 100 splits.
[…]

### 2020-01-27 code/learning/FigureS9.R

@@ -92,4 +92,4 @@ plots <- plot_grid(logit_plot, rf_plot, labels = c("A", "B"), align = 'v', ncol 
 
 ggdraw(add_sub(plots, "Sample size (N)", vpadding=grid::unit(0,"lines"), y=5, x=0.6, vjust=4.75, size=15))
 
-ggsave("Figure_S8.png", plot = last_plot(), device = 'png', path = 'submission', width = 6, height = 9.5)
+ggsave("Figure_S9.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9.5, dpi=300)

### 2020-01-28 submission/manuscript.Rmd
Get to render on my laptop

@@
 \newpage
-\includegraphics{Figure_1.pdf}
 **Figure 1.
@@
 Abbreviations: cvAUC, cross-validation area under the receiver operating characteristic curve.
-\newpage
-\includegraphics{Figure_2.tiff}
 
@@
 Outliers were defined as those data points that are not within 1.5 times the interquartile ranges.
-\newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.tiff}
 
@@
 OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red.
-\newpage
-\includegraphics{Figure_4.tiff}
 
@@
 For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
-\newpage
-\includegraphics{Figure_5.tiff}
 
@@
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S1.tiff}
 
@@
 
-\newpage
-\includegraphics{Figure_S2.tiff}
 **Figure S2.
@@
 
-\newpage
-\includegraphics[height=30cm, width=15cm]{Figure_S3.tiff}
 **Figure S3.
@@
 
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.tiff}
 **Figure S4.
@@
 
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S5.tiff}
 **Figure S5.
@@
 
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S6.tiff}
 **Figure S6.
@@
 
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S7.tiff}
 **Figure S7.
@@
 
-\newpage
-\includegraphics[height=20cm, width=13cm]{Figure_S8.tiff}
 **Figure S8.

### 2020-01-28 submission/manuscript.tex
Get to render on my laptop

@@
-\documentclass[11pt,]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  11pt,
+]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
   \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={A framework for effective application of machine learning to a microbiome-based classification problem},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
 \usepackage{graphicx,grffile}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{\textbf{Effective application of machine learning to a microbiome-based
-classification problem}}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{booktabs}
@@
 \usepackage{multirow}
-\usepackage[table]{xcolor}
 \usepackage{wrapfig}
@@
 
+\title{\textbf{A framework for effective application of machine learning to a
+microbiome-based classification problem}}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 \newpage
-
 \linenumbers
 
-\subsection{Abstract}\label{abstract}
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
@@
 
-\subsection{Importance}\label{importance}
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
 
@@
 
-\subsection{Background}\label{background}
+\hypertarget{background}{%
+\subsection{Background}\label{background}}
 
@@
 identify the bacterial populations that are associated with diseases
-using traditional statistical models. This variation is likely due to
-the ability of many bacterial populations to fill the same niche such
-that different populations cause the same disease in different
+using traditional statistical models. For example it is not possible to
+classify individuals as having healthy colons or colorectal lesions
+using Bray-Curtis distances based on the 16S rRNA gene sequences
+collected from fecal samples {[}Figure S1{]}. This variation is likely
+due to the ability of many bacterial populations to fill the same niche
+such that different populations cause the same disease in different
 individuals. Furthermore, a growing number of studies have shown that it
@@
 microbiome datasets. We also provided an aspirational rubric for
-evaluating the rigor of ML practices applied to microbiome data
-{[}Figure S1{]} to urge the audience to be diligent with with their
-study design, model selection, development, evaluation and
-interpretation.
-
-\subsection{Results}\label{results}
+evaluating the rigor of ML practices applied to microbiome data {[}Table
+S1{]} to urge the audience to be diligent with with their study design,
+model selection, development, evaluation and interpretation.
+
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
 
@@
 best average CV predictive performance using the area under the receiver
-operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
+operating characteristic curve (AUROC) {[}Figure S2 and S3{]}. The AUROC
 ranges from 0, where the model's predictions are perfectly incorrect, to
@@
 and benefited from tuning as the default C parameter was 1 {[}Figure
-S1{]}.
+S2{]}.
 
@@
 sampling distribution of the test statistic under the null hypothesis)
-(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
+(Figure S4). Similarly, L2-regularized logistic regression, XGBoost,
 L2-regularized SVM with linear and radial basis function kernel AUROC
@@
 random forest models when we subsetted our original study design with
-490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
+490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S5{]}. The
 variation in cross-validation performance within both models at smaller
@@
 
-\textbf{Interpretation of each ML model.} To evaluate the overall
-structure of the CRC dataset and show that the structure of the human
-microbiome is remarkably variable among individuals with SRNs, we
-analyzed the Bray-Curtis distances with an NMDS ordination {[}Figure
-S5{]}. The lack of seperation between healthy patients and patients with
-SRNs, show that traditional statistical approaches do not help us to
-adequately address disease ecology. This is why we often use ML models
-not just to predict a health outcome, but also to identify potential
+\textbf{Interpretation of each ML model.} We often use ML models not
+just to predict a health outcome, but also to identify potential
 biomarkers for disease. Therefore, model interpretation becomes crucial
@@
 kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S6{]}, 17 of the 20 top OTUs (e.g., OTU
+permutation importance {[}Figure S7{]}, 17 of the 20 top OTUs (e.g., OTU
 609, 822, 1239) were deemed important by both interpretation methods.
@@
 
-\subsection{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 
@@
 
-\subsection{Materials and Methods}\label{materials-and-methods}
+\hypertarget{materials-and-methods}{%
+\subsection{Materials and Methods}\label{materials-and-methods}}
 
@@
 population abundances. The sequence data used in our analyses were
-originally generated by Baxter et al. (available through NCBI Sequence
+originally generated by Baxter et al.~(available through NCBI Sequence
 Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
@@
 min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
-0) for the double tail event (e.g., Figure S3).
+0) for the double tail event (e.g., Figure S4).
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
 \hypertarget{refs}{}
-\hypertarget{ref-segata_metagenomic_2011}{}
+\leavevmode\hypertarget{ref-segata_metagenomic_2011}{}%
 1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
@@
 
-\hypertarget{ref-zeller_potential_2014}{}
+\leavevmode\hypertarget{ref-zeller_potential_2014}{}%
 2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
@@
 
-\hypertarget{ref-zackular_human_2014}{}
+\leavevmode\hypertarget{ref-zackular_human_2014}{}%
 3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
@@
 
-\hypertarget{ref-baxter_dna_2016}{}
+\leavevmode\hypertarget{ref-baxter_dna_2016}{}%
 4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
@@
 
-\hypertarget{ref-baxter_microbiota-based_2016}{}
+\leavevmode\hypertarget{ref-baxter_microbiota-based_2016}{}%
 5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
@@
 
-\hypertarget{ref-hale_shifts_2017}{}
+\leavevmode\hypertarget{ref-hale_shifts_2017}{}%
 6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
@@
 
-\hypertarget{ref-pasolli_machine_2016}{}
+\leavevmode\hypertarget{ref-pasolli_machine_2016}{}%
 7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
@@
 
-\hypertarget{ref-sze_looking_2016}{}
+\leavevmode\hypertarget{ref-sze_looking_2016}{}%
 8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
@@
 
-\hypertarget{ref-walters_meta-analyses_2014}{}
+\leavevmode\hypertarget{ref-walters_meta-analyses_2014}{}%
 9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
@@
 
-\hypertarget{ref-vazquez-baeza_guiding_2018}{}
+\leavevmode\hypertarget{ref-vazquez-baeza_guiding_2018}{}%
 10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
@@
 
-\hypertarget{ref-qin_alterations_2014}{}
+\leavevmode\hypertarget{ref-qin_alterations_2014}{}%
 11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
@@
 
-\hypertarget{ref-geman_deep_2018}{}
+\leavevmode\hypertarget{ref-geman_deep_2018}{}%
 12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
@@
 
-\hypertarget{ref-thaiss_persistent_2016}{}
+\leavevmode\hypertarget{ref-thaiss_persistent_2016}{}%
 13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
@@
 
-\hypertarget{ref-dadkhah_gut_2019}{}
+\leavevmode\hypertarget{ref-dadkhah_gut_2019}{}%
 14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
@@
 
-\hypertarget{ref-flemer_oral_2018}{}
+\leavevmode\hypertarget{ref-flemer_oral_2018}{}%
 15. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
@@
 
-\hypertarget{ref-montassier_pretreatment_2016}{}
+\leavevmode\hypertarget{ref-montassier_pretreatment_2016}{}%
 16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
@@
 
-\hypertarget{ref-ai_systematic_2017}{}
+\leavevmode\hypertarget{ref-ai_systematic_2017}{}%
 17. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
@@
 
-\hypertarget{ref-dai_multi-cohort_2018}{}
+\leavevmode\hypertarget{ref-dai_multi-cohort_2018}{}%
 18. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
@@
 
-\hypertarget{ref-mossotto_classification_2017}{}
+\leavevmode\hypertarget{ref-mossotto_classification_2017}{}%
 19. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
@@
 
-\hypertarget{ref-wong_quantitation_2017}{}
+\leavevmode\hypertarget{ref-wong_quantitation_2017}{}%
 20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
@@
 
-\hypertarget{ref-statnikov_comprehensive_2013}{}
+\leavevmode\hypertarget{ref-statnikov_comprehensive_2013}{}%
 21. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
@@
 
-\hypertarget{ref-knights_supervised_2011}{}
+\leavevmode\hypertarget{ref-knights_supervised_2011}{}%
 22. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
@@
 
-\hypertarget{ref-wirbel_meta-analysis_2019}{}
+\leavevmode\hypertarget{ref-wirbel_meta-analysis_2019}{}%
 23. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
@@
 
-\hypertarget{ref-vangay_microbiome_2019}{}
+\leavevmode\hypertarget{ref-vangay_microbiome_2019}{}%
 24. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
@@
 
-\hypertarget{ref-galkin_human_2018}{}
+\leavevmode\hypertarget{ref-galkin_human_2018}{}%
 25. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
@@
 
-\hypertarget{ref-reiman_using_2017}{}
+\leavevmode\hypertarget{ref-reiman_using_2017}{}%
 26. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
@@
 
-\hypertarget{ref-fioravanti_phylogenetic_2017}{}
+\leavevmode\hypertarget{ref-fioravanti_phylogenetic_2017}{}%
 27. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
@@
 
-\hypertarget{ref-thomas_metagenomic_2019}{}
+\leavevmode\hypertarget{ref-thomas_metagenomic_2019}{}%
 28. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
@@
 
-\hypertarget{ref-rudin_please_2018}{}
+\leavevmode\hypertarget{ref-rudin_please_2018}{}%
 29. \textbf{Rudin C}. 2018. Please stop explaining black box models for
@@
 
-\hypertarget{ref-rudin_optimized_2018}{}
+\leavevmode\hypertarget{ref-rudin_optimized_2018}{}%
 30. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
@@
 
-\hypertarget{ref-Quinn847475}{}
+\leavevmode\hypertarget{ref-Quinn847475}{}%
 31. \textbf{Quinn TP}, \textbf{Erb I}. 2019. Another look at
@@
 
-\hypertarget{ref-knights_human-associated_2011}{}
+\leavevmode\hypertarget{ref-knights_human-associated_2011}{}%
 32. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
@@
 
-\hypertarget{ref-miller_explanation_2017}{}
+\leavevmode\hypertarget{ref-miller_explanation_2017}{}%
 33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
@@
 
-\hypertarget{ref-ribeiro_why_2016}{}
+\leavevmode\hypertarget{ref-ribeiro_why_2016}{}%
 34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
-``Why should i trust you?'': Explaining the predictions of any
-classifier. arXiv:160204938 {[}cs, stat{]}.
-
-\hypertarget{ref-nori_interpretml:_2019}{}
+"Why should i trust you?": Explaining the predictions of any classifier.
+arXiv:160204938 {[}cs, stat{]}.
+
+\leavevmode\hypertarget{ref-nori_interpretml:_2019}{}%
 35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
@@
 
-\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
+\leavevmode\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}%
 36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
@@
 
-\hypertarget{ref-dormann_collinearity:_2013}{}
+\leavevmode\hypertarget{ref-dormann_collinearity:_2013}{}%
 37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
@@
 
-\hypertarget{ref-li_accurate_2020}{}
+\leavevmode\hypertarget{ref-li_accurate_2020}{}%
 38. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
@@
 
-\hypertarget{ref-sze_fecal_2019}{}
+\leavevmode\hypertarget{ref-sze_fecal_2019}{}%
 39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
@@
 
-\hypertarget{ref-kocheturov_massive_2019}{}
+\leavevmode\hypertarget{ref-kocheturov_massive_2019}{}%
 40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
@@
 
-\hypertarget{ref-kim_improved_2018}{}
+\leavevmode\hypertarget{ref-kim_improved_2018}{}%
 41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
@@
 
-\hypertarget{ref-wiens_no_2019}{}
+\leavevmode\hypertarget{ref-wiens_no_2019}{}%
 42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
@@
 
-\hypertarget{ref-westcott_opticlust_2017}{}
+\leavevmode\hypertarget{ref-westcott_opticlust_2017}{}%
 43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
@@
 
-\hypertarget{ref-redwood_stool_2016}{}
+\leavevmode\hypertarget{ref-redwood_stool_2016}{}%
 44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
@@
 
-\hypertarget{ref-sze_leveraging_2018}{}
+\leavevmode\hypertarget{ref-sze_leveraging_2018}{}%
 45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
@@
 
-\hypertarget{ref-schloss_introducing_2009}{}
+\leavevmode\hypertarget{ref-schloss_introducing_2009}{}%
 46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 
-\hypertarget{ref-rognes_vsearch_2016}{}
+\leavevmode\hypertarget{ref-rognes_vsearch_2016}{}%
 47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
@@
 
-\hypertarget{ref-li_hyperband:_2016}{}
+\leavevmode\hypertarget{ref-li_hyperband:_2016}{}%
 48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
@@
 \small
-
 \begin{tabular}{|l|l|l|}
@@
 
-\end{tabular}\newpage
-
+\end{tabular}
 \newpage
 
-\includegraphics{Figure_1.pdf} \textbf{Figure 1. Machine learning
-pipeline.} We split the data to create a training (80\%) and held-out
-test set (20\%). The splits were stratified to maintain the overall
-class distribution. We performed five-fold cross-validation on the
-training data to select the best hyperparameter setting and then used
-these hyperparameters to train the models. The model was evaluated on
-the held-out data set. Abbreviations: cvAUC, cross-validation area under
-the receiver operating characteristic curve. \newpage
-\includegraphics{Figure_2.png}
+\newpage
+
+\textbf{Figure 1. Machine learning pipeline.} We split the data to
+create a training (80\%) and held-out test set (20\%). The splits were
+stratified to maintain the overall class distribution. We performed
+five-fold cross-validation on the training data to select the best
+hyperparameter setting and then used these hyperparameters to train the
+models. The model was evaluated on the held-out data set. Abbreviations:
+cvAUC, cross-validation area under the receiver operating characteristic
+curve.
 
@@
 that were not outliers. Outliers were defined as those data points that
-are not within 1.5 times the interquartile ranges. \newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.png}
+are not within 1.5 times the interquartile ranges.
 
@@
 associated with classifying a subject having an SRN had positive signs
-and were shown in red. \newpage
-\includegraphics{Figure_4.png}
+and were shown in red.
 
@@
 the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
-had the largest impact on predictive performance. \newpage
-\includegraphics{Figure_5.png}
+had the largest impact on predictive performance.
 
@@
 \small
-
 \begin{tabular}{|l|l|l|l|}
@@
 
-\end{tabular}\newpage
-
-\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
-setting performances for linear models.} (A) L2-regularized logistic
-regression, (B) L1-regularized SVM with linear kernel, and (C)
-L2-regularized SVM with linear kernel mean cross-validation AUROC values
-when different hyperparameters were used in training the model. The
-stars represent the highest performing hyperparameter setting for each
-model. \newpage
-\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
-
-\textbf{Figure S2. Hyperparameter setting performances for non-linear
+\end{tabular}
+
+\newpage
+
+\textbf{Figure S1. NMDS ordination of Bray-Curtis distances.} NMDS
+ordination relating the community structures of the fecal microbiota
+from 490 patients (261 patients with normal colonoscopies and 229
+patients who have screen relevant neoplasias; SRNs).
+
+\textbf{Figure S2. Hyperparameter setting performances for linear
+models.} (A) L2-regularized logistic regression, (B) L1-regularized SVM
+with linear kernel, and (C) L2-regularized SVM with linear kernel mean
+cross-validation AUROC values when different hyperparameters were used
+in training the model. The stars represent the highest performing
+hyperparameter setting for each model.
+
+\textbf{Figure S3. Hyperparameter setting performances for non-linear
 models.} (A) Decision tree, (B) random forest, (C) SVM with radial basis
@@
 represent the highest performing hyperparameter setting for the models.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
-
-\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
+
+\textbf{Figure S4. Histogram of AUROC differences between L2-regularized
 logistic regression and random forest for each of the hundred
@@
 about 0, therefore the p-value was calculated for a double-tail event.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
-
-\textbf{Figure S4. Classification performance of ML models across
+
+\textbf{Figure S5. Classification performance of ML models across
 cross-validation when trained on a subset of the dataset.} (A)
@@
 train and validate the models but included the ranges observed with the
-more complete datasets. \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
-
-\textbf{Figure S5. NMDS ordination of Bray-Curtis distances.} NMDS
-ordination relating the community structures of the fecal microbiota
-from 490 patients (261 patients with normal colonoscopies and 229
-patients who have screen relevant neoplasias; SRNs).
-
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
+more complete datasets.
 
@@
 
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
-
 \textbf{Figure S7. Interpretation of the linear ML models with
@@
 test set.
-
-\newpage
-
-\includegraphics[height=20cm, width=13cm]{Figure_S8.pdf}
 

### 2020-01-28 submission/Table_S1.Rmd
updates for rebuttal after pat meeting

--- 2020-01-21 submission/manuscript.tex
@@
-\documentclass[11pt,]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{\textbf{Effective application of machine learning to a microbiome-based
-classification problem}}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
-
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage[table]{xcolor}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{caption}
-\usepackage{hyperref}
-\usepackage{helvet} % Helvetica font
-\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
-\usepackage[T1]{fontenc}
-\usepackage[labelfont=bf]{caption}
-
-\usepackage[none]{hyphenat}
-
-\usepackage{setspace}
-\doublespacing
-\setlength{\parskip}{1em}
-
-\usepackage{lineno}
-
-\usepackage{pdfpages}
-\floatplacement{figure}{H} % Keep the figure up top of the page
-
-\begin{document}
-\maketitle
-
-\vspace{35mm}
-
-Running title: Machine learning methods applied to a microbiome study
-
-\vspace{35mm}
-
-Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Mack
-Ruffin\({^3}\), Jenna Wiens\({^2}\), Patrick D.
-Schloss\textsuperscript{1\(\dagger\)}
-
-\vspace{38mm}
-
-\(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}},
-\href{mailto:wiensj@umich.edu}{\nolinkurl{wiensj@umich.edu}}
-
-1. Department of Microbiology and Immunology, University of Michigan,
-Ann Arbor, MI 48109
-
-2. Department of Electrical Engineering and Computer Science, University
-of Michigan, Ann Arbor, MI 48109
-
-3. Department of Family Medicine and Community Medicine, Penn State
-Hershey Medical Center, Hershey, PA
+---
+csl: mbio.csl
+fontsize: 11pt
+output:
+  pdf_document:
+    includes:
+      in_header: header.tex
+    keep_tex: yes
+---
 
 \newpage
-
-\linenumbers
-
-\subsection{Abstract}\label{abstract}
-
-Machine learning (ML) modeling of the human microbiome has the potential
-to identify microbial biomarkers and aid in the diagnosis of many
-diseases such as inflammatory bowel disease, diabetes, and colorectal
-cancer. Progress has been made towards developing ML models that predict
-health outcomes using bacterial abundances, but inconsistent adoption of
-training and evaluation methods call the validity of these models into
-question. Furthermore, there appears to be a preference by many
-researchers to favor increased model complexity over interpretability.
-To overcome these challenges, we trained seven models that used fecal
-16S rRNA sequence data to predict the presence of colonic screen
-relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
-We developed a reusable open-source pipeline to train, validate, and
-interpret ML models. To show the effect of model selection, we assessed
-the predictive performance, interpretability, and training time of
-L2-regularized logistic regression, L1 and L2-regularized support vector
-machines (SVM) with linear and radial basis function kernels, decision
-trees, random forest, and gradient boosted trees (XGBoost). The random
-forest model performed best at detecting SRNs with an AUROC of 0.695
-{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
-interpretable. Despite its simplicity, L2-regularized logistic
-regression followed random forest in predictive performance with an
-AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
-inherently interpretable. Our analysis highlights the importance of
-choosing an ML approach based on the goal of the study, as the choice
-will inform expectations of performance and interpretability.
-
-\newpage
-
-\subsection{Importance}\label{importance}
-
-Prediction of diseases using machine learning (ML) is rapidly being
-adopted in microbiome studies. However, the estimated performance
-associated with these ML models is likely over-optimistic. Moreover,
-there is a trend towards using black box models without a discussion of
-the difficulty of interpreting such models when trying to identify
-microbial biomarkers of disease. This work represents a step towards
-developing more reproducible ML practices in applying ML to microbiome
-research. We implement a rigorous pipeline and emphasize the importance
-of selecting ML models that reflect the goal of the study. These
-concepts are not particular to the study of human health but can also be
-applied to environmental microbiology studies.
-
-\newpage
-
-\subsection{Background}\label{background}
-
-As the number of people represented in human microbiome datasets grow,
-there is an increasing desire to use microbiome data to diagnose
-disease. However, the structure of the human microbiome is remarkably
-variable among individuals to the point where it is often difficult to
-identify the bacterial populations that are associated with diseases
-using traditional statistical models. This variation is likely due to
-the ability of many bacterial populations to fill the same niche such
-that different populations cause the same disease in different
-individuals. Furthermore, a growing number of studies have shown that it
-is rare for a single bacterial species to be associated with a disease.
-Instead, subsets of the microbiome account for differences in health.
-Traditional statistical approaches do not adequately account for the
-variation in the human microbiome and typically consider the protective
-or risk effects of each bacterial population separately (1). Recently,
-machine learning (ML) models have grown in popularity among microbiome
-researchers as our ability to sample large numbers of individuals has
-grown; such models can effectively account for the interpersonal
-microbiome variation and the ecology of disease.
-
-ML models can be used to increase our understanding of the variation in
-the structure of existing data and in making predictions about new data.
-Researchers have used ML models to diagnose and understand the
-ecological basis of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
-task of diagnosing an individual relies on a rigorously validated model.
-However, there are common methodological and reporting problems that
-arise when applying ML to such data, that need to be addressed for the
-field to progress. These problems include a lack of transparency in
-which methods are used and how these methods are implemented; evaluating
-models without separate held-out test data; unreported variation between
-the predictive performance on different folds of cross-validation; and
-unreported variation between cross-validation and testing performances.
-Though the microbiome field is making progress to avoid some of these
-pitfalls including validating their models on independent datasets (8,
-19, 20) and introducing accessible and open-source ML tools (21--24),
-more work is needed to improve reproducibility further and minimize
-overestimating for model performance.
-
-Among microbiome researchers, the lack of justification when selecting a
-modeling approach has often been due to an implicit assumption that more
-complex models are better. This has resulted in a trend towards using
-non-linear models such as random forest and deep neural networks (3, 12,
-25--27) over simpler models such as logistic regression or other linear
-models (19, 23, 28). Although in some cases, complex models may capture
-important non-linear relationships and therefore yield better
-predictions, they can also result in black boxes that lack
-interpretability. Such models require post hoc explanations to quantify
-the importance of each feature in making predictions. Depending on the
-goal of the modeling, other approaches may be more appropriate. For
-example, researchers trying to identify the microbiota associated with
-disease may desire a more interpretable model, whereas clinicians may
-emphasize predictive performance. Nonetheless, it is essential to
-understand that the benefit of more complex, less interpretable models
-may be minimal (29--31). It is important for researchers to justify
-their choice of modeling approach.
-
-In this study, we provided steps toward standardization of machine
-learning methods for microbiome studies which are often poorly
-documented and executed. To showcase a rigorous ML pipeline and to shed
-light on how ML model selection can affect modeling results, we
-performed an empirical analysis comparing the predictive performance,
-interpretability and training times of seven modeling approaches with
-the same dataset and pipeline. We built three linear models with
-different forms of regularization: L2-regularized logistic regression
-and L1 and L2-regularized support vector machines (SVM) with a linear
-kernel. We also trained four non-linear models: SVM with radial basis
-function kernel, a decision tree, random forest, and gradient boosted
-trees. We compared their predictive performance, interpretability, and
-training time. To demonstrate the performance of these modeling
-approaches and our pipeline on a case study, we used a data from a
-previously published study that sought to classify individuals as having
-healthy colons or colonic lesions based on the 16S rRNA gene sequences
-collected from fecal samples (4). This dataset was selected because it
-is a relatively large collection of individuals (N=490) connected to a
-clinically significant disease where there is ample evidence that the
-disease is driven by variation in the microbiome (2, 4, 5, 32). With
-this dataset, we developed an ML pipeline that can be used in many
-different scenarios for training and evaluating models. This framework
-can be easily applied to other host-associated and environmental
-microbiome datasets. We also provided an aspirational rubric for
-evaluating the rigor of ML practices applied to microbiome data
-{[}Figure S1{]} to urge the audience to be diligent with with their
-study design, model selection, development, evaluation and
-interpretation.
-
-\subsection{Results}\label{results}
-
-\textbf{Model selection and pipeline construction}. We established a
-reusable ML pipeline for model selection and evaluation, focusing on
-seven different commonly used supervised learning algorithms {[}Figure
-1, Table 1{]}.
-
-First, we randomly split the data into training and test sets so that
-the training set consisted of 80\% of the full dataset, while the test
-set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
-distribution of controls and cases found in the full dataset, we
-performed stratified splits. For example, our full dataset included 490
-individuals. Of these, 261 had healthy colons (53\%) and 229 had a
-screen relevant neoplasia (SRN; 46.7\%). A training set included 393
-individuals, of which 209 had an SRN (53\%), while the test set was
-composed of 97 individuals, of which 52 had an SRN (54\%). The training
-data were used to build and select the models, and the test set was used
-for evaluating the model. We trained seven different models using the
-training data {[}Table 1{]}.
-
-Model selection requires tuning hyperparameters. Hyperparameters are
-parameters that need to be specified or tuned by the user, in order to
-train a model for a specific modeling problem. For example, when using
-regularization, C is a hyperparameter that indicates the penalty for
-overfitting. Hyperparameters are tuned using the training data to find
-the best model. We selected hyperparameters by performing repeated
-five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
-five-fold CV was also stratified to maintain the overall case and
-control distribution. We chose the hyperparameter values that led to the
-best average CV predictive performance using the area under the receiver
-operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
-ranges from 0, where the model's predictions are perfectly incorrect, to
-1.0, where the model perfectly distinguishes between cases and controls.
-An AUROC value of 0.5 indicates that the model's predictions are no
-different than random. To select hyperparameters, we performed a grid
-search for hyperparameter settings when training the models. Default
-hyperparameter settings in developed ML packages available in R, Python,
-and MATLAB programming languages may be inadequate for effective
-application of classification algorithms and need to be optimized for
-each new ML task. For example, L1-regularized SVM with linear kernel
-showed large variability between different regularization strengths (C)
-and benefited from tuning as the default C parameter was 1 {[}Figure
-S1{]}.
-
-Once hyperparameters were selected, we trained the model using the full
-training dataset and applied the final model to the held-out data to
-evaluate the testing predictive performance of each model. The
-data-split, hyperparameter selection, training and testing steps were
-repeated 100 times to obtain a robust interpretation of model
-performance, less likely to be affected by a ``lucky'' or ``unlucky''
-split {[}Figure 1{]}.
-
-\textbf{Predictive performance and generalizability of the seven
-models.} We evaluated the predictive performance of the seven models to
-classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
-The predictive performance of random forest model was higher than other
-ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
-significantly (p=0.5; The p-value was manually calculated using the
-sampling distribution of the test statistic under the null hypothesis)
-(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
-L2-regularized SVM with linear and radial basis function kernel AUROC
-values were not significantly different from one another and had median
-AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
-0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
-0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
-decision tree had significantly lower AUROC values than the other ML
-models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
-0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
-results demonstrate that the most complex model (XGBoost) did not have
-the best performance and that the most interpretable models
-(L2-regularized logistic regression and L2-regularized SVM with linear
-kernel) performed nearly as well as non-linear models.
-
-To evaluate the generalizability of each model, we compared the median
-cross-validation AUROC to the median testing AUROC. If the difference
-between the cross-validation and testing AUROCs was large, then that
-could indicate that the models were overfit to the training data. The
-largest difference in median AUROCs was 0.021 in L1-regularized SVM with
-linear kernel, followed by SVM with radial basis function kernel and
-decision tree with a difference of 0.007 and 0.006, respectively
-{[}Figure 2{]}. These differences were relatively small and gave us
-confidence in our estimate of the generalization performance of the
-models.
-
-To evaluate the variation in the estimated performance, we calculated
-the range of AUROC values for each model using 100 data-splits. The
-range among the testing AUROC values within each model varied by 0.230
-on average across the seven models. If we had only done a single split,
-then there is a risk that we could have gotten lucky or unlucky in
-estimating model performance. For instance, the lowest AUROC value of
-the random forest model was 0.593, whereas the highest was 0.810. These
-results showed that depending on the data-split, the testing performance
-can vary {[}Figure 2{]}. Therefore, it is important to employ multiple
-data splits when estimating generalization performance.
-
-To show the effect of sample size on model generalizability, we compared
-cross-validation AUROC values of L2-regularized logistic regression and
-random forest models when we subsetted our original study design with
-490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
-variation in cross-validation performance within both models at smaller
-sample sizes was larger than when the full collection of samples was
-used to train and validate the models. Because of the high
-dimensionality of the microbiome data (6920 OTUs), large sample sizes
-can lead to better models.
-
-\textbf{Interpretation of each ML model.} To evaluate the overall
-structure of the CRC dataset and show that the structure of the human
-microbiome is remarkably variable among individuals with SRNs, we
-analyzed the Bray-Curtis distances with an NMDS ordination {[}Figure
-S5{]}. The lack of seperation between healthy patients and patients with
-SRNs, show that traditional statistical approaches do not help us to
-adequately address disease ecology. This is why we often use ML models
-not just to predict a health outcome, but also to identify potential
-biomarkers for disease. Therefore, model interpretation becomes crucial
-for microbiome studies. Interpretability is related to the degree to
-which humans can understand the reasons behind a model prediction
-(33--35). ML models often decrease in interpretability as they increase
-in complexity. In this study, we used two methods to help interpret our
-models.
-
-First, we interpreted the feature importance of the linear models (L1
-and L2-regularized SVM with linear kernel and L2-regularized logistic
-regression) using the median rank of absolute feature weights for each
-OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
-determine whether an OTU was associated with classifying a subject as
-being healthy or having an SRN. It was encouraging that many of the
-highest-ranked OTUs were shared across these three models (e.g., OTUs
-50, 426, 609, 822, 1239). The great benefit of this approach was knowing
-the sign and magnitude of each OTU coefficient in the trained model.
-This allowed us to immidiately interpret negative and positive
-coefficient signs as protective and risk factors, respectively and the
-magnitude as the impact of these factors. However, this approach is
-limited to linear models or models with prespecified interaction terms.
-
-Second, to analyze non-linear models, we interpreted the feature
-importance using permutation importance (36). Whereas the absolute
-feature weights were determined from the trained models, here we
-measured importance using the held-out test data. Permutation importance
-analysis is a post hoc explanation of the model, in which we randomly
-permuted groups of perfectly correlated features together and other
-features individually across the two groups in the held-out test data
-{[}Figure S6{]}. We then calculated how much the predictive performance
-of the model (i.e., testing AUROC values) decreased when each OTU or
-group of OTUs was randomly permuted. We ranked the OTUs based on how
-much the median testing AUROC decreased when it was permuted; the OTU
-with the largest decrease ranked highest {[}Figure 4{]}. Among the
-twenty OTUs with the largest impact, there was only one OTU (OTU 822)
-that was shared among all of the models; however, we found three OTUs
-(OTUs 58, 110, 367) that were important in each of the tree-based
-models. Similarly, the random forest and XGBoost models shared four of
-the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis
-results also revealed that with the exception of the decision tree
-model, removal of any individual OTU had minimal impact on model
-performance. For example, if OTU 367 was permuted across the samples in
-the decision tree model, the median AUROC dropped from 0.601 to 0.525.
-In contrast, if the same OTU was permuted in the random forest model,
-the AUROC only dropped from 0.695 to 0.680, which indicated high degree
-of collinearity in the dataset. Permutation analysis allowed us to gauge
-the importance of each OTU in non-linear models and partially account
-for collinearity by grouping correlated OTUs to determine their impact
-as a group.
-
-To further highlight the differences between the two interpretation
-methods, we used permutation importance to interpret the linear models
-{[}Figure S7{]}. When we analyzed the L1-regularized SVM with linear
-kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S6{]}, 17 of the 20 top OTUs (e.g., OTU
-609, 822, 1239) were deemed important by both interpretation methods.
-Similarly, for the L2-regularized SVM and L2-regularized logistic
-regression, 9 and 12 OTUs, respectively, were shared among the two
-interpretation methods. These results indicate that both methods are
-consistent in selecting the most important OTUs.
-
-We also compared the top 20 OTUs selected by permutation importance in
-L2-regularized logistic regression and the highest performing tree-based
-models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
-shared among the models. These results indicate that we were able to
-identify a few more OTUs that are deemed important by both linear and
-non-linear models when we used permutation importance as our
-interpretation method.
-
-Figure S7:
-
-\textbf{The computational efficiency of each ML model.} We compared the
-training times of the seven ML models. The training times increased with
-the complexity of the model and the number of potential hyperparameter
-combinations. Also, the linear models trained faster than non-linear
-models {[}Figures S1-S2; Figure 5{]}.
-
-\subsection{Discussion}\label{discussion}
-
-There is a growing awareness that many human diseases and environmental
-processes are not driven by a single organism but are the product of
-multiple bacterial populations. Traditional statistical approaches are
-useful for identifying those cases where a single organism is associated
-with a process. In contrast, ML methods offer the ability to incorporate
-the structure of the microbial communities as a whole and identify
-associations between community structure and disease state. If it is
-possible to classify communities reliably, then ML methods also offer
-the ability to identify those microbial populations within the
-communities that are responsible for the classification. However, the
-application of ML in microbiome studies is still in its infancy, and the
-field needs to develop a better understanding of different ML methods,
-their strengths and weaknesses, and how to implement them.
-
-To address these needs, we developed an open-sourced framework for ML
-models. Using this pipeline, we benchmarked seven ML models and showed
-that the tradeoff between model complexity and performance may be less
-severe than originally hypothesized. In terms of predictive performance,
-the random forest model had the best AUROC compared to the other six
-models. However, the second-best model was L2-regularized logistic
-regression with a median AUROC difference of less than 0.015 compared to
-random forest. While our implementation of random forest took 83.2 hours
-to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model,
-while L2-regularized logistic regression, a linear model, is easily
-interpreted according to the feature weights. Comparing many different
-models showed us that the most complex model was not necessarily the
-best model for our ML task.
-
-We established a pipeline that can be generalized to any modeling method
-that predicts a binary health outcome. We performed a random data-split
-to create a training set (80\% of the data) and a held-out test set
-(20\% of the data), which we used to evaluate predictive performance. We
-used the AUROC metric to evaluate predictive performance as it is a
-clinically relevant evaluation metric for our study. We repeated this
-data-split 100 times to measure the possible variation in predictive
-performance. During training, we tuned the model hyperparameters with a
-repeated five-fold cross-validation. Despite the high number of features
-microbiome datasets typically have, the models we built with this
-pipeline generalized to the held-out test sets.
-
-We highlighted the importance of model interpretation to gain greater
-biological insights into microbiota-associated diseases. In this study,
-we showcased two different interpretation methods: ranking each OTU by
-(i) their absolute weights in the trained models and (ii) their impact
-on the predictive performance based on permutation importance.
-Human-associated microbial communities have complex correlation
-structures that create collinearity in the datasets. This can hinder our
-ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (37). To capture all
-important features, once we identify highly ranked OTUs, we should
-review their relationships with other OTUs. These relationships will
-help us generate new hypotheses about the ecology of the disease and
-test them with follow-up experiments. When we used permutation
-importance, we partially accounted for collinearity by grouping
-correlated OTUs to determine their impact as a group. We grouped OTUs
-that had a perfect correlation with each other; however, we could reduce
-the correlation threshold to further investigate the relationships among
-correlated features. We are choosing to leave the decision to define
-correlation and setting up correlations thresholds to the researchers
-for their own microbiome questions. However, we would like to urge them
-to investigate the correlation structures of the data to avoid
-misinterpreting the models. This is likely to be a particular problem
-with shotgun metagenomic datasets where collinearity will be more
-pronounced due to many genes being correlated with one another because
-they come from the same chromosome. True causal mechanisms (e.g., role
-of microbiome in colorectal cancer) cannot be explained solely by the
-highest performing machine learning model (38). To identify the true
-underlying microbial factors of a disease, it is crucial to follow up on
-any correlation analyses with further hypothesis testing and
-experimentation for biological validation.
-
-In this study, we did not consider all possible modeling approaches.
-However, the principles highlighted throughout this study apply to other
-ML modeling tasks with microbiome data. For example, we did not evaluate
-multicategory classification methods to predict non-binary outcomes. We
-could have trained models to differentiate between people with healthy
-colons and those with adenomas or carcinomas (k=3 categories). We did
-not perform this analysis because the clinically relevant diagnosis
-grouping was between patients with healthy colons and those with SRNs.
-Furthermore, as the number of classes increases, more samples are
-required for each category to train an accurate model. We also did not
-use regression-based analyses to predict a non-categorical outcome. We
-have previously used such an approach to train random forest models to
-predict fecal short-chain fatty acid concentrations based on microbiome
-data (39). Our analysis was also limited to shallow learning methods and
-did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (12, 40, 41), but microbiome datasets
-often suffer from having many features and small sample sizes, which
-result in overfitting.
-
-Our framework provides a reproducible structure to investigators wanting
-to train, evaluate, and interpret their own ML models to generate
-hypotheses regarding which OTUs might be biologically relevant. However,
-deploying microbiome-based models to make clinical diagnoses or
-predictions is a significantly more challenging and distinct undertaking
-(42). For example, we currently lack standardized methods to collect
-patient samples, generate sequence data, and report clinical data. We
-are also challenged by the practical constraints of OTU-based
-approaches. The de novo algorithms commonly in use are slow, require
-considerable memory, and result in different OTU assignments as new data
-are added (43). Finally, we also need independent validation cohorts to
-test the performance of a diagnostic model. To realize the potential for
-using ML approaches with microbiome data, it is necessary that we direct
-our efforts to overcome these challenges.
-
-Our study highlights the need to make educated choices at every step of
-developing an ML model with microbiome data. We created an aspirational
-rubric that researchers can use to identify potential pitfalls when
-using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
-have highlighted the trade-offs between model complexity and
-interpretability, the need for tuning hyperparameters, the utility of
-held-out test sets for evaluating predictive performance, and the
-importance of considering correlation structures in datasets for
-reliable interpretation. Furthermore, we underscored the importance of
-proper experimental design and methods to help us achieve the level of
-validity and accountability we want from models built for patient
-health.
-
-\subsection{Materials and Methods}\label{materials-and-methods}
-
-\textbf{Data collection and study population.} The original stool
-samples described in our analysis were obtained from patients recruited
-by Great Lakes-New England Early Detection Research Network (5). Stool
-samples were provided by adults who were undergoing a scheduled
-screening or surveillance colonoscopy. Participants were recruited from
-Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
-(MI, USA). Patients' colonic health was visually assessed by colonoscopy
-with bowel preparation and tissue histopathology of all resected
-lesions. We assigned patients into two classes: those with healthy
-colons and those with screen relevant neoplasias (SRNs). The healthy
-class included patients with healthy colons or non-advanced adenomas,
-whereas the SRN class included patients with advanced adenomas or
-carcinomas (44). Patients with an adenoma greater than 1 cm, more than
-three adenomas of any size, or an adenoma with villous histology were
-classified as having advanced adenomas (44). There were 172 patients
-with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
-Of the 198 adenomas, 109 were identified as advanced adenomas. Together
-261 patients were classified as healthy and 229 patients were classified
-as having an SRN.
-
-\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
-patients were used for 16S rRNA gene sequencing to measure bacterial
-population abundances. The sequence data used in our analyses were
-originally generated by Baxter et al. (available through NCBI Sequence
-Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
-generated by Sze et al (45), who processed the 16S rRNA sequences in
-mothur (v1.39.3) using the default quality filtering methods,
-identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
-(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
-These OTU abundances were the features we used to predict colorectal
-health of the patients. There were 6920 OTUs. OTU abundances were
-subsampled to the size of the smallest sample and normalized across
-samples such that the highest abundance of each OTU would be 1, and the
-lowest would be 0.
-
-\textbf{Model training and evaluation.} Models were trained using the
-caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
-calculate decision values for models generated using L2-regularized SVM
-with linear kernel and L1-regularized SVM with linear kernel. The code
-for these changes on L2-regularized SVM with linear kernel and
-L1-regularized SVM with linear kernel models are available at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
-and at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
-respectively.
-
-For hyperparameter selection, we started with a granular grid search.
-Then we narrowed and fine-tuned the range of each hyperparameter.For
-L2-regularized logistic regression, L1 and L2-regularized SVM with
-linear and radial basis function kernels, we tuned the cost
-hyperparameter, which controls the regularization strength, where
-smaller values specify stronger regularization. For SVM with radial
-basis function kernel, we also tuned the sigma hyperparameter, which
-determines the reach of a single training instance where, for a high
-value of sigma, the SVM decision boundary will be dependent on the
-points that are closest to the decision boundary. For the decision tree
-model, we tuned the depth of the tree where the deeper the tree, the
-more splits it has. For random forest, we tuned the number of features
-to consider when looking for the best tree split. For XGBoost, we tuned
-the learning rate and the fraction of samples used for fitting the
-individual base learners. Performing a grid search for hyperparameter
-selection might not be feasible for when there are more than two
-hyperparameters to tune for. In such cases, it is more efficient to use
-random search or recently developed tools such as Hyperband to identify
-good hyperparameter configurations (48).
-
-The computational burden during model training due to model complexity
-was reduced by parallelizing segments of the ML pipeline. We
-parallelized the training of each data-split. This allowed the 100
-data-splits to be processed through the ML pipeline simultaneously at
-the same time for each model. It is possible to further parallelize the
-cross-validation step for each hyperparameter setting which we have not
-performed in this study.
-
-\textbf{Permutation importance workflow.} We calculated a Spearman's
-rank-order correlation matrix and defined correlated OTUs as having
-perfect correlation (correlation coefficient = 1 and p \textless{}
-0.01). OTUs without a perfect correlation to each other were permuted
-individually, whereas correlated ones were grouped together and permuted
-at the same time.
-
-\textbf{Statistical analysis workflow.} Data summaries, statistical
-analysis, and data visualizations were performed using R (v.3.5.0) with
-the tidyverse package (v.1.2.1). We compared the performance of the
-models pairwise by calculating the difference between AUROC values from
-the same data-split (for 100 data-splits). We determined if the models
-were significantly different by calculating the empirical p-value (2 x
-min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
-0) for the double tail event (e.g., Figure S3).
-
-\textbf{Code availability.} The code for all sequence curation and
-analysis steps including an Rmarkdown version of this manuscript is
-available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
-
-\textbf{Acknowledgements.} We thank all the study participants of Great
-Lakes-New England Early Detection Research Network. We would like to
-thank the members of the Schloss lab for their valuable feedback. Salary
-support for MR came from NIH grant 1R01CA215574. Salary support for PDS
-came from NIH grants P30DK034933 and 1R01CA215574.
-
-\newpage
-
-\subsection{References}\label{references}
-
-\hypertarget{refs}{}
-\hypertarget{ref-segata_metagenomic_2011}{}
-1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
-\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
-\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
-explanation. Genome Biol \textbf{12}:R60.
-doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.
-
-\hypertarget{ref-zeller_potential_2014}{}
-2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
-\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
-\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
-\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
-\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
-\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
-J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
-\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
-\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
-for early-stage detection of colorectal cancer. Mol Syst Biol
-\textbf{10}.
-doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
-
-\hypertarget{ref-zackular_human_2014}{}
-3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
-\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
-for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
-doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
-
-\hypertarget{ref-baxter_dna_2016}{}
-4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
-\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
-immunochemical test can replace stool for detection of colonic lesions
-using a microbiota-based model. Microbiome \textbf{4}.
-doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
-
-\hypertarget{ref-baxter_microbiota-based_2016}{}
-5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\hypertarget{ref-hale_shifts_2017}{}
-6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
-\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
-\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
-\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
-\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
-microbiota associated with adenomatous polyps. Cancer Epidemiol
-Biomarkers Prev \textbf{26}:85--94.
-doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.
-
-\hypertarget{ref-pasolli_machine_2016}{}
-7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
-\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
-meta-analysis of large metagenomic datasets: Tools and biological
-insights. PLoS Comput Biol \textbf{12}.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.
-
-\hypertarget{ref-sze_looking_2016}{}
-8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
-the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
-doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.
-
-\hypertarget{ref-walters_meta-analyses_2014}{}
-9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
-Meta-analyses of human gut microbes associated with obesity and IBD.
-FEBS Lett \textbf{588}:4223--4233.
-doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
-
-\hypertarget{ref-vazquez-baeza_guiding_2018}{}
-10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
-\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
-\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
-Gut \textbf{67}:1743--1745.
-doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.
-
-\hypertarget{ref-qin_alterations_2014}{}
-11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
-\textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
-E}, \textbf{Yao J}, \textbf{Wu L}, \textbf{Zhou J}, \textbf{Ni S},
-\textbf{Liu L}, \textbf{Pons N}, \textbf{Batto JM}, \textbf{Kennedy SP},
-\textbf{Leonard P}, \textbf{Yuan C}, \textbf{Ding W}, \textbf{Chen Y},
-\textbf{Hu X}, \textbf{Zheng B}, \textbf{Qian G}, \textbf{Xu W},
-\textbf{Ehrlich SD}, \textbf{Zheng S}, \textbf{Li L}. 2014. Alterations
-of the human gut microbiome in liver cirrhosis. Nature
-\textbf{513}:59--64.
-doi:\href{https://doi.org/10.1038/nature13568}{10.1038/nature13568}.
-
-\hypertarget{ref-geman_deep_2018}{}
-12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
-\textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
-learning tools for human microbiome big data, pp. 265--275. \emph{In}
-Balas, VE, Jain, LC, Balas, MM (eds.), Soft computing applications.
-Springer International Publishing.
-
-\hypertarget{ref-thaiss_persistent_2016}{}
-13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
-\textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
-\textbf{Dohnalová L}, \textbf{Braverman S}, \textbf{Rozin S},
-\textbf{Malitsky S}, \textbf{Dori-Bachash M}, \textbf{Kuperman Y},
-\textbf{Biton I}, \textbf{Gertler A}, \textbf{Harmelin A},
-\textbf{Shapiro H}, \textbf{Halpern Z}, \textbf{Aharoni A},
-\textbf{Segal E}, \textbf{Elinav E}. 2016. Persistent microbiome
-alterations modulate the rate of post-dieting weight regain. Nature
-\textbf{540}:544--551.
-doi:\href{https://doi.org/10.1038/nature20796}{10.1038/nature20796}.
-
-\hypertarget{ref-dadkhah_gut_2019}{}
-14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
-\textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
-G}, \textbf{Kuehn T}, \textbf{Gillevet PM}. 2019. Gut microbiome
-identifies risk for colorectal polyps. BMJ Open Gastroenterology
-\textbf{6}:e000297.
-doi:\href{https://doi.org/10.1136/bmjgast-2019-000297}{10.1136/bmjgast-2019-000297}.
-
-\hypertarget{ref-flemer_oral_2018}{}
-15. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
-\textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
-E}, \textbf{O`Riordain M}, \textbf{Shanahan F}, \textbf{O`Toole PW}.
-2018. The oral microbiota in colorectal cancer is distinctive and
-predictive. Gut \textbf{67}:1454--1463.
-doi:\href{https://doi.org/10.1136/gutjnl-2017-314814}{10.1136/gutjnl-2017-314814}.
-
-\hypertarget{ref-montassier_pretreatment_2016}{}
-16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
-\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
-P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
-2016. Pretreatment gut microbiome predicts chemotherapy-related
-bloodstream infection. Genome Medicine \textbf{8}:49.
-doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.
-
-\hypertarget{ref-ai_systematic_2017}{}
-17. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
-\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
-supervised classifiers for fecal microbiota-based prediction of
-colorectal cancer. Oncotarget \textbf{8}:9546--9556.
-doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.
-
-\hypertarget{ref-dai_multi-cohort_2018}{}
-18. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
-WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
-\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
-J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
-identified altered bacteria across populations and universal bacterial
-markers. Microbiome \textbf{6}:70.
-doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.
-
-\hypertarget{ref-mossotto_classification_2017}{}
-19. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
-\textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
-Classification of paediatric inflammatory bowel disease using machine
-learning. Scientific Reports \textbf{7}.
-doi:\href{https://doi.org/10.1038/s41598-017-02606-2}{10.1038/s41598-017-02606-2}.
-
-\hypertarget{ref-wong_quantitation_2017}{}
-20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
-AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
-\textbf{Zhang L}, \textbf{Wu JCY}, \textbf{Chan FKL}, \textbf{Ng SSM},
-\textbf{Wong MCS}, \textbf{Ng SC}, \textbf{Wu WKK}, \textbf{Yu J},
-\textbf{Sung JJY}. 2017. Quantitation of faecal fusobacterium improves
-faecal immunochemical test in detecting advanced colorectal neoplasia.
-Gut \textbf{66}:1441--1448.
-doi:\href{https://doi.org/10.1136/gutjnl-2016-312766}{10.1136/gutjnl-2016-312766}.
-
-\hypertarget{ref-statnikov_comprehensive_2013}{}
-21. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
-\textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
-\textbf{Blaser MJ}, \textbf{Aliferis CF}, \textbf{Alekseyenko AV}. 2013.
-A comprehensive evaluation of multicategory classification methods for
-microbiomic data. Microbiome \textbf{1}:11.
-doi:\href{https://doi.org/10.1186/2049-2618-1-11}{10.1186/2049-2618-1-11}.
-
-\hypertarget{ref-knights_supervised_2011}{}
-22. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
-Supervised classification of human microbiota. FEMS Microbiology Reviews
-\textbf{35}:343--359.
-doi:\href{https://doi.org/10.1111/j.1574-6976.2010.00251.x}{10.1111/j.1574-6976.2010.00251.x}.
-
-\hypertarget{ref-wirbel_meta-analysis_2019}{}
-23. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
-K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
-\textbf{Voigt AY}, \textbf{Palleja A}, \textbf{Ponnudurai R},
-\textbf{Sunagawa S}, \textbf{Coelho LP}, \textbf{Schrotz-King P},
-\textbf{Vogtmann E}, \textbf{Habermann N}, \textbf{Niméus E},
-\textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Gandini S},
-\textbf{Serrano D}, \textbf{Mizutani S}, \textbf{Shiroma H},
-\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
-T}, \textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N},
-\textbf{Sinha R}, \textbf{Ulrich CM}, \textbf{Brenner H},
-\textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller G}. 2019.
-Meta-analysis of fecal metagenomes reveals global microbial signatures
-that are specific for colorectal cancer. Nature Medicine
-\textbf{25}:679.
-doi:\href{https://doi.org/10.1038/s41591-019-0406-6}{10.1038/s41591-019-0406-6}.
-
-\hypertarget{ref-vangay_microbiome_2019}{}
-24. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
-Microbiome learning repo (ML repo): A public repository of microbiome
-regression and classification tasks. Gigascience \textbf{8}.
-doi:\href{https://doi.org/10.1093/gigascience/giz042}{10.1093/gigascience/giz042}.
-
-\hypertarget{ref-galkin_human_2018}{}
-25. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
-\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
-2018. Human microbiome aging clocks based on deep learning and tandem of
-permutation feature importance and accumulated local effects. bioRxiv.
-doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
-
-\hypertarget{ref-reiman_using_2017}{}
-26. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
-convolutional neural networks to explore the microbiome, pp. 4269--4272.
-\emph{In} 2017 39th annual international conference of the IEEE
-engineering in medicine and biology society (EMBC).
-
-\hypertarget{ref-fioravanti_phylogenetic_2017}{}
-27. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
-\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
-\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
-in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
-
-\hypertarget{ref-thomas_metagenomic_2019}{}
-28. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
-\textbf{Pasolli E}, \textbf{Armanini F}, \textbf{Zolfo M},
-\textbf{Beghini F}, \textbf{Manara S}, \textbf{Karcher N}, \textbf{Pozzi
-C}, \textbf{Gandini S}, \textbf{Serrano D}, \textbf{Tarallo S},
-\textbf{Francavilla A}, \textbf{Gallo G}, \textbf{Trompetto M},
-\textbf{Ferrero G}, \textbf{Mizutani S}, \textbf{Shiroma H},
-\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
-T}, \textbf{Wirbel J}, \textbf{Schrotz-King P}, \textbf{Ulrich CM},
-\textbf{Brenner H}, \textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller
-G}, \textbf{Cordero F}, \textbf{Dias-Neto E}, \textbf{Setubal JC},
-\textbf{Tett A}, \textbf{Pardini B}, \textbf{Rescigno M},
-\textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N}. 2019.
-Metagenomic analysis of colorectal cancer datasets identifies
-cross-cohort microbial diagnostic signatures and a link with choline
-degradation. Nature Medicine \textbf{25}:667.
-doi:\href{https://doi.org/10.1038/s41591-019-0405-7}{10.1038/s41591-019-0405-7}.
-
-\hypertarget{ref-rudin_please_2018}{}
-29. \textbf{Rudin C}. 2018. Please stop explaining black box models for
-high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
-
-\hypertarget{ref-rudin_optimized_2018}{}
-30. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
-Toward trust in machine learning for healthcare and criminal justice.
-Interfaces \textbf{48}:449--466.
-doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.
-
-\hypertarget{ref-Quinn847475}{}
-31. \textbf{Quinn TP}, \textbf{Erb I}. 2019. Another look at
-microbemetabolite interactions: How scale invariant correlations can
-outperform a neural network. bioRxiv.
-doi:\href{https://doi.org/10.1101/847475}{10.1101/847475}.
-
-\hypertarget{ref-knights_human-associated_2011}{}
-32. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
-\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
-signatures: Examining their predictive value. Cell Host Microbe
-\textbf{10}:292--296.
-doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.
-
-\hypertarget{ref-miller_explanation_2017}{}
-33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
-Insights from the social sciences. arXiv:170607269 {[}cs{]}.
-
-\hypertarget{ref-ribeiro_why_2016}{}
-34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
-``Why should i trust you?'': Explaining the predictions of any
-classifier. arXiv:160204938 {[}cs, stat{]}.
-
-\hypertarget{ref-nori_interpretml:_2019}{}
-35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
-\textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
-learning interpretability. arXiv:190909223 {[}cs, stat{]}.
-
-\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
-36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
-\textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
-importance measure. Bioinformatics \textbf{26}:1340--1347.
-doi:\href{https://doi.org/10.1093/bioinformatics/btq134}{10.1093/bioinformatics/btq134}.
-
-\hypertarget{ref-dormann_collinearity:_2013}{}
-37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
-\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
-JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
-\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
-\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
-\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
-of methods to deal with it and a simulation study evaluating their
-performance. Ecography \textbf{36}:27--46.
-doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.
-
-\hypertarget{ref-li_accurate_2020}{}
-38. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
-Accurate data-driven prediction does not mean high reproducibility. Nat
-Mach Intell \textbf{2}:13--15.
-doi:\href{https://doi.org/10.1038/s42256-019-0140-2}{10.1038/s42256-019-0140-2}.
-
-\hypertarget{ref-sze_fecal_2019}{}
-39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
-\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
-acids are not predictive of colonic tumor status and cannot be predicted
-based on bacterial community structure. mBio \textbf{10}:e01454--19.
-doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.
-
-\hypertarget{ref-kocheturov_massive_2019}{}
-40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
-2019. Massive datasets and machine learning for computational
-biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
-doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.
-
-\hypertarget{ref-kim_improved_2018}{}
-41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
-method for prediction of cancer prognosis by network learning. Genes
-\textbf{9}:478.
-doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.
-
-\hypertarget{ref-wiens_no_2019}{}
-42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
-\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
-\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
-\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
-2019. Do no harm: A roadmap for responsible machine learning for health
-care. Nat Med \textbf{25}:1337--1340.
-doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.
-
-\hypertarget{ref-westcott_opticlust_2017}{}
-43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
-Improved Method for Assigning Amplicon-Based Sequence Data to
-Operational Taxonomic Units. mSphere \textbf{2}.
-doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
-
-\hypertarget{ref-redwood_stool_2016}{}
-44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
-\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
-\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
-\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
-TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
-for screening detection of colorectal neoplasia in alaska native people.
-Mayo Clin Proc \textbf{91}:61--70.
-doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.
-
-\hypertarget{ref-sze_leveraging_2018}{}
-45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
-rRNA gene surveys to identify reproducible biomarkers in individuals
-with colorectal tumors. mBio \textbf{9}:e00630--18.
-doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
-
-\hypertarget{ref-schloss_introducing_2009}{}
-46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-Source, Platform-Independent,
-Community-Supported Software for Describing and Comparing Microbial
-Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.
-
-\hypertarget{ref-rognes_vsearch_2016}{}
-47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
-\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
-source tool for metagenomics. PeerJ \textbf{4}:e2584.
-doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
-
-\hypertarget{ref-li_hyperband:_2016}{}
-48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
-\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
-bandit-based approach to hyperparameter optimization. arXiv:160306560
-{[}cs, stat{]}.
-
-\newpage
-
 \captionsetup{labelformat=empty}
-\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
-\small
-
-\begin{tabular}{|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-
-\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
-
-Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
-
-\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
-
-\end{tabular}\newpage
-
-\newpage
-
-\includegraphics{Figure_1.pdf} \textbf{Figure 1. Machine learning
-pipeline.} We split the data to create a training (80\%) and held-out
-test set (20\%). The splits were stratified to maintain the overall
-class distribution. We performed five-fold cross-validation on the
-training data to select the best hyperparameter setting and then used
-these hyperparameters to train the models. The model was evaluated on
-the held-out data set. Abbreviations: cvAUC, cross-validation area under
-the receiver operating characteristic curve. \newpage
-\includegraphics{Figure_2.png}
-
-\textbf{Figure 2. Generalization and classification performance of ML
-models using AUROC values of all cross-validation and testing
-performances.} The median AUROC for diagnosing individuals with SRN
-using bacterial abundances was higher than chance (depicted by a
-horizontal line at 0.50) for all the ML models. The predictive
-performance of random forest model was higher than other ML models,
-though not significantly (p \textgreater{} 0.01). L2-regularized
-logistic regression, XGBoost, L2-regularized SVM with linear and radial
-basis function kernel performances were not significantly different from
-one another. The boxplot shows quartiles at the box ends and the median
-as the horizontal line in the box. The whiskers show the farthest points
-that were not outliers. Outliers were defined as those data points that
-are not within 1.5 times the interquartile ranges. \newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.png}
-
-\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
-absolute feature weights of (A) L1-regularized SVM with linear kernel,
-(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
-logistic regression, were ranked from highest rank, 1, to lowest rank,
-100, for each data-split. The feature ranks of the 20 highest ranked
-OTUs based on their median ranks (median shown in black) are reported
-here. OTUs that were associated with classifying a subject as being
-healthy had negative signs and were shown in blue. OTUs that were
-associated with classifying a subject having an SRN had positive signs
-and were shown in red. \newpage
-\includegraphics{Figure_4.png}
-
-\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
-with radial basis kernel, (B) decision tree, (C) random forest, and (D)
-XGBoost feature importances were explained using permutation importance
-on the held-out test data set. The gray rectangle and the dashed line
-show the IQR range and median of the base testing AUROC without any
-permutation. The 20 OTUs that caused the largest decrease in the AUROC
-when permuted are reported here. The colors of the box plots represent
-the OTUs that were shared among the different models; yellow were OTUs
-that were shared among all the non-linear models, green were OTUs that
-were shared among the tree-based models, green were the OTUs shared
-among SVM with radial basis kernel, decision tree and XGBoost, pink were
-the OTUs shared among SVM with radial basis kernel and XGBoost only, red
-were the OTUs shared among random forest and XGBoost only and blue were
-the OTUs shared among decision tree and random forest only. For all of
-the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
-had the largest impact on predictive performance. \newpage
-\includegraphics{Figure_5.png}
-
-\textbf{Figure 5. Training times of seven ML models.} The median
-training time was the highest for XGBoost and shortest for
-L2-regularized logistic regression. \newpage
-
 \captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
 \small
-
 \begin{tabular}{|l|l|l|l|}
@@
 
-\end{tabular}\newpage
-
-\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
-setting performances for linear models.} (A) L2-regularized logistic
-regression, (B) L1-regularized SVM with linear kernel, and (C)
-L2-regularized SVM with linear kernel mean cross-validation AUROC values
-when different hyperparameters were used in training the model. The
-stars represent the highest performing hyperparameter setting for each
-model. \newpage
-\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
-
-\textbf{Figure S2. Hyperparameter setting performances for non-linear
-models.} (A) Decision tree, (B) random forest, (C) SVM with radial basis
-kernel, and (D) XGBoost mean cross-validation AUROC values when
-different hyperparameters were used in training the model. The stars
-represent the highest performing hyperparameter setting for the models.
+\end{tabular}
 \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
-
-\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
-logistic regression and random forest for each of the hundred
-data-splits.} In 75\% of data-splits, the AUROC of random forest was
-greater than that of L2-regularized logistic regression. The p-value was
-manually calculated using the sampling distribution of the test
-statistic under the null hypothesis. We tested how often random forest
-performed more accurately than L2-regularized logistic regression. The
-null hypothesis is that the distribution of the difference between the
-AUROC values of random forest and L2 logistic regression is symmetric
-about 0, therefore the p-value was calculated for a double-tail event.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
-
-\textbf{Figure S4. Classification performance of ML models across
-cross-validation when trained on a subset of the dataset.} (A)
-L2-regularized logistic regression and (B) random forest models were
-trained using the original study design with 490 subjects and subsets of
-it with 15, 30, 60, 120, and 245 subjects. The range among the
-cross-validation AUROC values within both models at smaller sample sizes
-were much larger than when the full collection of samples was used to
-train and validate the models but included the ranges observed with the
-more complete datasets. \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
-
-\textbf{Figure S5. NMDS ordination with Bray-Curtis distances.}
-
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
-
-\textbf{Figure S6. Permutation importance analysis.} Permutation
-importance analysis measures the decrease in the predictive performance
-of the model after we permute (A) a feature's or (B) a group of
-correlated features' values, which breaks the relationship between the
-feature and the actual outcome.
-
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
-
-\textbf{Figure S7. Interpretation of the linear ML models with
-permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
-L2-regularized SVM with linear kernel, and (C) L2-regularized logistic
-regression were interpreted using permutation importance using held-out
-test set.
-
-\end{document}

### 2020-01-28 submission/manuscript.Rmd
updates for rebuttal after pat meeting

@@
 ---
-title: '**A framework for effective application of machine learning to a microbiome-based classification problem**'
+title: '**A framework for effective application of machine learning to microbiome-based classification problems**'
 csl: mbio.csl
@@
 
-Running title: Machine learning framework applied to a microbiome study
+Running title: Machine learning framework to model microbiome data
 
@@
 
-We also compared the top 20 OTUs selected by permutation importance in L2-regularized logistic regression and the highest performing tree-based models, random forest and XGBoost.
+We also compared the top 20 OTUs selected by permutation importance in L2-regularized logistic regression [Figure S7] and the highest performing tree-based models, random forest and XGBoost [Figure 4].
 2 and 5 OTUs, respectively, were shared among the models.
@@
 The training times increased with the complexity of the model and the number of potential hyperparameter combinations.
-Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].
+Also, the linear models trained faster than non-linear models [Figures S2-S3; Figure 5].
 
@@
 
-\newpage
 **Figure 1.
@@
 Abbreviations: cvAUC, cross-validation area under the receiver operating characteristic curve.
+
+\hfill\break
 
@@
 
+\hfill\break
+
 **Figure 3.
@@
 OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red.
+
+\hfill\break
 
@@
 
+\hfill\break
+
 **Figure 5. Training times of seven ML models.** The median training time was the highest for XGBoost and shortest for L2-regularized logistic regression.
-\newpage
-
-\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
-\small
-\begin{tabular}{|l|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
-
-\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
-
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population to which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
-
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
-
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
-
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g., when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
-
-\end{tabular}
 
@@
 
+\hfill\break
+
 **Figure S2.
@@
 
+\hfill\break
+
 **Figure S3.
@@
 The stars represent the highest performing hyperparameter setting for the models.
+
+\hfill\break
 
@@
 
+\hfill\break
+
 **Figure S5.
@@
 
+\hfill\break
+
 **Figure S6.
@@
 
+\hfill\break
+
 **Figure S7.
 Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
+
+\hfill\break
 

### 2020-01-28 submission/manuscript.tex
updates for rebuttal after pat meeting

--- 2020-01-22 submission/manuscript.tex
@@
 \IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\usepackage[]{microtype}
 \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
-\usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\hypersetup{
+            pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
             pdfborder={0 0 0},
@@
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{\textbf{Effective application of machine learning to a microbiome-based
-classification problem}}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{}
-    \preauthor{}\postauthor{}
-    \date{}
-    \predate{}\postdate{}
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
@@
 
+\title{\textbf{A framework for effective application of machine learning to
+microbiome-based classification problems}}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 
-Running title: Machine learning methods applied to a microbiome study
+Running title: Machine learning framework to model microbiome data
 
@@
 microbiome datasets. We also provided an aspirational rubric for
-evaluating the rigor of ML practices applied to microbiome data
-{[}Figure S1{]} to urge the audience to be diligent with with their
-study design, model selection, development, evaluation and
-interpretation.
+evaluating the rigor of ML practices applied to microbiome data {[}Table
+S1{]} to urge the audience to be diligent with with their study design,
+model selection, development, evaluation and interpretation.
 
@@
 
-\newpage
-
-\includegraphics{Figure_1.pdf} \textbf{Figure 1. Machine learning
-pipeline.} We split the data to create a training (80\%) and held-out
-test set (20\%). The splits were stratified to maintain the overall
-class distribution. We performed five-fold cross-validation on the
-training data to select the best hyperparameter setting and then used
-these hyperparameters to train the models. The model was evaluated on
-the held-out data set. Abbreviations: cvAUC, cross-validation area under
-the receiver operating characteristic curve. \newpage
-\includegraphics{Figure_2.png}
+\textbf{Figure 1. Machine learning pipeline.} We split the data to
+create a training (80\%) and held-out test set (20\%). The splits were
+stratified to maintain the overall class distribution. We performed
+five-fold cross-validation on the training data to select the best
+hyperparameter setting and then used these hyperparameters to train the
+models. The model was evaluated on the held-out data set. Abbreviations:
+cvAUC, cross-validation area under the receiver operating characteristic
+curve.
+
+\hfill\break
 
@@
 that were not outliers. Outliers were defined as those data points that
-are not within 1.5 times the interquartile ranges. \newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.png}
+are not within 1.5 times the interquartile ranges.
+
+\hfill\break
 
@@
 associated with classifying a subject having an SRN had positive signs
-and were shown in red. \newpage
-\includegraphics{Figure_4.png}
+and were shown in red.
+
+\hfill\break
 
@@
 the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
-had the largest impact on predictive performance. \newpage
-\includegraphics{Figure_5.png}
+had the largest impact on predictive performance.
+
+\hfill\break
 
@@
 training time was the highest for XGBoost and shortest for
-L2-regularized logistic regression. \newpage
-
-\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
-\small
-
-\begin{tabular}{|l|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
-
-\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
-
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population to which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
-
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
-
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
-
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g., when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
-
-\end{tabular}\newpage
-
-\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
-setting performances for linear models.} (A) L2-regularized logistic
-regression, (B) L1-regularized SVM with linear kernel, and (C)
-L2-regularized SVM with linear kernel mean cross-validation AUROC values
-when different hyperparameters were used in training the model. The
-stars represent the highest performing hyperparameter setting for each
-model. \newpage
-\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
-
-\textbf{Figure S2. Hyperparameter setting performances for non-linear
+L2-regularized logistic regression.
+
+\newpage
+
+\textbf{Figure S1. NMDS ordination of Bray-Curtis distances.} NMDS
+ordination relating the community structures of the fecal microbiota
+from 490 patients (261 patients with normal colonoscopies and 229
+patients who have screen relevant neoplasias; SRNs).
+
+\hfill\break
+
+\textbf{Figure S2. Hyperparameter setting performances for linear
+models.} (A) L2-regularized logistic regression, (B) L1-regularized SVM
+with linear kernel, and (C) L2-regularized SVM with linear kernel mean
+cross-validation AUROC values when different hyperparameters were used
+in training the model. The stars represent the highest performing
+hyperparameter setting for each model.
+
+\hfill\break
+
+\textbf{Figure S3. Hyperparameter setting performances for non-linear
 models.} (A) Decision tree, (B) random forest, (C) SVM with radial basis
@@
 represent the highest performing hyperparameter setting for the models.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
-
-\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
+
+\hfill\break
+
+\textbf{Figure S4. Histogram of AUROC differences between L2-regularized
 logistic regression and random forest for each of the hundred
@@
 about 0, therefore the p-value was calculated for a double-tail event.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
-
-\textbf{Figure S4. Classification performance of ML models across
+
+\hfill\break
+
+\textbf{Figure S5. Classification performance of ML models across
 cross-validation when trained on a subset of the dataset.} (A)
@@
 train and validate the models but included the ranges observed with the
-more complete datasets. \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
-
-\textbf{Figure S5. NMDS ordination of Bray-Curtis distances.} NMDS
-ordination relating the community structures of the fecal microbiota
-from 490 patients (261 patients with normal colonoscopies and 229
-patients who have screen relevant neoplasias; SRNs).
-
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
+more complete datasets.
+
+\hfill\break
 
@@
 
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
+\hfill\break
 
@@
 
-\newpage
-
-\includegraphics[height=20cm, width=13cm]{Figure_S8.pdf}
+\hfill\break
 

### 2020-01-28 submission/Table_S1.tex

--- 2020-01-22 submission/manuscript.tex
@@
 \IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\usepackage[]{microtype}
 \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
-\usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\hypersetup{
             pdfborder={0 0 0},
@@
 
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
-\maketitle
-
-\vspace{35mm}
-
-Running title: Machine learning methods applied to a microbiome study
-
-\vspace{35mm}
-
-Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Mack
-Ruffin\({^3}\), Jenna Wiens\({^2}\), Patrick D.
-Schloss\textsuperscript{1\(\dagger\)}
-
-\vspace{38mm}
-
-\(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}},
-\href{mailto:wiensj@umich.edu}{\nolinkurl{wiensj@umich.edu}}
-
-1. Department of Microbiology and Immunology, University of Michigan,
-Ann Arbor, MI 48109
-
-2. Department of Electrical Engineering and Computer Science, University
-of Michigan, Ann Arbor, MI 48109
-
-3. Department of Family Medicine and Community Medicine, Penn State
-Hershey Medical Center, Hershey, PA
-
-\newpage
-
-\linenumbers
-
-\subsection{Abstract}\label{abstract}
-
-Machine learning (ML) modeling of the human microbiome has the potential
-to identify microbial biomarkers and aid in the diagnosis of many
-diseases such as inflammatory bowel disease, diabetes, and colorectal
-cancer. Progress has been made towards developing ML models that predict
-health outcomes using bacterial abundances, but inconsistent adoption of
-training and evaluation methods call the validity of these models into
-question. Furthermore, there appears to be a preference by many
-researchers to favor increased model complexity over interpretability.
-To overcome these challenges, we trained seven models that used fecal
-16S rRNA sequence data to predict the presence of colonic screen
-relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
-We developed a reusable open-source pipeline to train, validate, and
-interpret ML models. To show the effect of model selection, we assessed
-the predictive performance, interpretability, and training time of
-L2-regularized logistic regression, L1 and L2-regularized support vector
-machines (SVM) with linear and radial basis function kernels, decision
-trees, random forest, and gradient boosted trees (XGBoost). The random
-forest model performed best at detecting SRNs with an AUROC of 0.695
-{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
-interpretable. Despite its simplicity, L2-regularized logistic
-regression followed random forest in predictive performance with an
-AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
-inherently interpretable. Our analysis highlights the importance of
-choosing an ML approach based on the goal of the study, as the choice
-will inform expectations of performance and interpretability.
-
-\newpage
-
-\subsection{Importance}\label{importance}
-
-Prediction of diseases using machine learning (ML) is rapidly being
-adopted in microbiome studies. However, the estimated performance
-associated with these ML models is likely over-optimistic. Moreover,
-there is a trend towards using black box models without a discussion of
-the difficulty of interpreting such models when trying to identify
-microbial biomarkers of disease. This work represents a step towards
-developing more reproducible ML practices in applying ML to microbiome
-research. We implement a rigorous pipeline and emphasize the importance
-of selecting ML models that reflect the goal of the study. These
-concepts are not particular to the study of human health but can also be
-applied to environmental microbiology studies.
-
-\newpage
-
-\subsection{Background}\label{background}
-
-As the number of people represented in human microbiome datasets grow,
-there is an increasing desire to use microbiome data to diagnose
-disease. However, the structure of the human microbiome is remarkably
-variable among individuals to the point where it is often difficult to
-identify the bacterial populations that are associated with diseases
-using traditional statistical models. This variation is likely due to
-the ability of many bacterial populations to fill the same niche such
-that different populations cause the same disease in different
-individuals. Furthermore, a growing number of studies have shown that it
-is rare for a single bacterial species to be associated with a disease.
-Instead, subsets of the microbiome account for differences in health.
-Traditional statistical approaches do not adequately account for the
-variation in the human microbiome and typically consider the protective
-or risk effects of each bacterial population separately (1). Recently,
-machine learning (ML) models have grown in popularity among microbiome
-researchers as our ability to sample large numbers of individuals has
-grown; such models can effectively account for the interpersonal
-microbiome variation and the ecology of disease.
-
-ML models can be used to increase our understanding of the variation in
-the structure of existing data and in making predictions about new data.
-Researchers have used ML models to diagnose and understand the
-ecological basis of diseases such as liver cirrhosis, colorectal cancer,
-inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
-task of diagnosing an individual relies on a rigorously validated model.
-However, there are common methodological and reporting problems that
-arise when applying ML to such data, that need to be addressed for the
-field to progress. These problems include a lack of transparency in
-which methods are used and how these methods are implemented; evaluating
-models without separate held-out test data; unreported variation between
-the predictive performance on different folds of cross-validation; and
-unreported variation between cross-validation and testing performances.
-Though the microbiome field is making progress to avoid some of these
-pitfalls including validating their models on independent datasets (8,
-19, 20) and introducing accessible and open-source ML tools (21--24),
-more work is needed to improve reproducibility further and minimize
-overestimating for model performance.
-
-Among microbiome researchers, the lack of justification when selecting a
-modeling approach has often been due to an implicit assumption that more
-complex models are better. This has resulted in a trend towards using
-non-linear models such as random forest and deep neural networks (3, 12,
-25--27) over simpler models such as logistic regression or other linear
-models (19, 23, 28). Although in some cases, complex models may capture
-important non-linear relationships and therefore yield better
-predictions, they can also result in black boxes that lack
-interpretability. Such models require post hoc explanations to quantify
-the importance of each feature in making predictions. Depending on the
-goal of the modeling, other approaches may be more appropriate. For
-example, researchers trying to identify the microbiota associated with
-disease may desire a more interpretable model, whereas clinicians may
-emphasize predictive performance. Nonetheless, it is essential to
-understand that the benefit of more complex, less interpretable models
-may be minimal (29--31). It is important for researchers to justify
-their choice of modeling approach.
-
-In this study, we provided steps toward standardization of machine
-learning methods for microbiome studies which are often poorly
-documented and executed. To showcase a rigorous ML pipeline and to shed
-light on how ML model selection can affect modeling results, we
-performed an empirical analysis comparing the predictive performance,
-interpretability and training times of seven modeling approaches with
-the same dataset and pipeline. We built three linear models with
-different forms of regularization: L2-regularized logistic regression
-and L1 and L2-regularized support vector machines (SVM) with a linear
-kernel. We also trained four non-linear models: SVM with radial basis
-function kernel, a decision tree, random forest, and gradient boosted
-trees. We compared their predictive performance, interpretability, and
-training time. To demonstrate the performance of these modeling
-approaches and our pipeline on a case study, we used a data from a
-previously published study that sought to classify individuals as having
-healthy colons or colonic lesions based on the 16S rRNA gene sequences
-collected from fecal samples (4). This dataset was selected because it
-is a relatively large collection of individuals (N=490) connected to a
-clinically significant disease where there is ample evidence that the
-disease is driven by variation in the microbiome (2, 4, 5, 32). With
-this dataset, we developed an ML pipeline that can be used in many
-different scenarios for training and evaluating models. This framework
-can be easily applied to other host-associated and environmental
-microbiome datasets. We also provided an aspirational rubric for
-evaluating the rigor of ML practices applied to microbiome data
-{[}Figure S1{]} to urge the audience to be diligent with with their
-study design, model selection, development, evaluation and
-interpretation.
-
-\subsection{Results}\label{results}
-
-\textbf{Model selection and pipeline construction}. We established a
-reusable ML pipeline for model selection and evaluation, focusing on
-seven different commonly used supervised learning algorithms {[}Figure
-1, Table 1{]}.
-
-First, we randomly split the data into training and test sets so that
-the training set consisted of 80\% of the full dataset, while the test
-set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
-distribution of controls and cases found in the full dataset, we
-performed stratified splits. For example, our full dataset included 490
-individuals. Of these, 261 had healthy colons (53\%) and 229 had a
-screen relevant neoplasia (SRN; 46.7\%). A training set included 393
-individuals, of which 209 had an SRN (53\%), while the test set was
-composed of 97 individuals, of which 52 had an SRN (54\%). The training
-data were used to build and select the models, and the test set was used
-for evaluating the model. We trained seven different models using the
-training data {[}Table 1{]}.
-
-Model selection requires tuning hyperparameters. Hyperparameters are
-parameters that need to be specified or tuned by the user, in order to
-train a model for a specific modeling problem. For example, when using
-regularization, C is a hyperparameter that indicates the penalty for
-overfitting. Hyperparameters are tuned using the training data to find
-the best model. We selected hyperparameters by performing repeated
-five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
-five-fold CV was also stratified to maintain the overall case and
-control distribution. We chose the hyperparameter values that led to the
-best average CV predictive performance using the area under the receiver
-operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
-ranges from 0, where the model's predictions are perfectly incorrect, to
-1.0, where the model perfectly distinguishes between cases and controls.
-An AUROC value of 0.5 indicates that the model's predictions are no
-different than random. To select hyperparameters, we performed a grid
-search for hyperparameter settings when training the models. Default
-hyperparameter settings in developed ML packages available in R, Python,
-and MATLAB programming languages may be inadequate for effective
-application of classification algorithms and need to be optimized for
-each new ML task. For example, L1-regularized SVM with linear kernel
-showed large variability between different regularization strengths (C)
-and benefited from tuning as the default C parameter was 1 {[}Figure
-S1{]}.
-
-Once hyperparameters were selected, we trained the model using the full
-training dataset and applied the final model to the held-out data to
-evaluate the testing predictive performance of each model. The
-data-split, hyperparameter selection, training and testing steps were
-repeated 100 times to obtain a robust interpretation of model
-performance, less likely to be affected by a ``lucky'' or ``unlucky''
-split {[}Figure 1{]}.
-
-\textbf{Predictive performance and generalizability of the seven
-models.} We evaluated the predictive performance of the seven models to
-classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
-The predictive performance of random forest model was higher than other
-ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
-significantly (p=0.5; The p-value was manually calculated using the
-sampling distribution of the test statistic under the null hypothesis)
-(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
-L2-regularized SVM with linear and radial basis function kernel AUROC
-values were not significantly different from one another and had median
-AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
-0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
-0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
-decision tree had significantly lower AUROC values than the other ML
-models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
-0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
-results demonstrate that the most complex model (XGBoost) did not have
-the best performance and that the most interpretable models
-(L2-regularized logistic regression and L2-regularized SVM with linear
-kernel) performed nearly as well as non-linear models.
-
-To evaluate the generalizability of each model, we compared the median
-cross-validation AUROC to the median testing AUROC. If the difference
-between the cross-validation and testing AUROCs was large, then that
-could indicate that the models were overfit to the training data. The
-largest difference in median AUROCs was 0.021 in L1-regularized SVM with
-linear kernel, followed by SVM with radial basis function kernel and
-decision tree with a difference of 0.007 and 0.006, respectively
-{[}Figure 2{]}. These differences were relatively small and gave us
-confidence in our estimate of the generalization performance of the
-models.
-
-To evaluate the variation in the estimated performance, we calculated
-the range of AUROC values for each model using 100 data-splits. The
-range among the testing AUROC values within each model varied by 0.230
-on average across the seven models. If we had only done a single split,
-then there is a risk that we could have gotten lucky or unlucky in
-estimating model performance. For instance, the lowest AUROC value of
-the random forest model was 0.593, whereas the highest was 0.810. These
-results showed that depending on the data-split, the testing performance
-can vary {[}Figure 2{]}. Therefore, it is important to employ multiple
-data splits when estimating generalization performance.
-
-To show the effect of sample size on model generalizability, we compared
-cross-validation AUROC values of L2-regularized logistic regression and
-random forest models when we subsetted our original study design with
-490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S4{]}. The
-variation in cross-validation performance within both models at smaller
-sample sizes was larger than when the full collection of samples was
-used to train and validate the models. Because of the high
-dimensionality of the microbiome data (6920 OTUs), large sample sizes
-can lead to better models.
-
-\textbf{Interpretation of each ML model.} To evaluate the overall
-structure of the CRC dataset and show that the structure of the human
-microbiome is remarkably variable among individuals with SRNs, we
-analyzed the Bray-Curtis distances with an NMDS ordination {[}Figure
-S5{]}. The lack of seperation between healthy patients and patients with
-SRNs, show that traditional statistical approaches do not help us to
-adequately address disease ecology. This is why we often use ML models
-not just to predict a health outcome, but also to identify potential
-biomarkers for disease. Therefore, model interpretation becomes crucial
-for microbiome studies. Interpretability is related to the degree to
-which humans can understand the reasons behind a model prediction
-(33--35). ML models often decrease in interpretability as they increase
-in complexity. In this study, we used two methods to help interpret our
-models.
-
-First, we interpreted the feature importance of the linear models (L1
-and L2-regularized SVM with linear kernel and L2-regularized logistic
-regression) using the median rank of absolute feature weights for each
-OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
-determine whether an OTU was associated with classifying a subject as
-being healthy or having an SRN. It was encouraging that many of the
-highest-ranked OTUs were shared across these three models (e.g., OTUs
-50, 426, 609, 822, 1239). The great benefit of this approach was knowing
-the sign and magnitude of each OTU coefficient in the trained model.
-This allowed us to immidiately interpret negative and positive
-coefficient signs as protective and risk factors, respectively and the
-magnitude as the impact of these factors. However, this approach is
-limited to linear models or models with prespecified interaction terms.
-
-Second, to analyze non-linear models, we interpreted the feature
-importance using permutation importance (36). Whereas the absolute
-feature weights were determined from the trained models, here we
-measured importance using the held-out test data. Permutation importance
-analysis is a post hoc explanation of the model, in which we randomly
-permuted groups of perfectly correlated features together and other
-features individually across the two groups in the held-out test data
-{[}Figure S6{]}. We then calculated how much the predictive performance
-of the model (i.e., testing AUROC values) decreased when each OTU or
-group of OTUs was randomly permuted. We ranked the OTUs based on how
-much the median testing AUROC decreased when it was permuted; the OTU
-with the largest decrease ranked highest {[}Figure 4{]}. Among the
-twenty OTUs with the largest impact, there was only one OTU (OTU 822)
-that was shared among all of the models; however, we found three OTUs
-(OTUs 58, 110, 367) that were important in each of the tree-based
-models. Similarly, the random forest and XGBoost models shared four of
-the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis
-results also revealed that with the exception of the decision tree
-model, removal of any individual OTU had minimal impact on model
-performance. For example, if OTU 367 was permuted across the samples in
-the decision tree model, the median AUROC dropped from 0.601 to 0.525.
-In contrast, if the same OTU was permuted in the random forest model,
-the AUROC only dropped from 0.695 to 0.680, which indicated high degree
-of collinearity in the dataset. Permutation analysis allowed us to gauge
-the importance of each OTU in non-linear models and partially account
-for collinearity by grouping correlated OTUs to determine their impact
-as a group.
-
-To further highlight the differences between the two interpretation
-methods, we used permutation importance to interpret the linear models
-{[}Figure S7{]}. When we analyzed the L1-regularized SVM with linear
-kernel model using feature rankings based on weights {[}Figure 3{]} and
-permutation importance {[}Figure S6{]}, 17 of the 20 top OTUs (e.g., OTU
-609, 822, 1239) were deemed important by both interpretation methods.
-Similarly, for the L2-regularized SVM and L2-regularized logistic
-regression, 9 and 12 OTUs, respectively, were shared among the two
-interpretation methods. These results indicate that both methods are
-consistent in selecting the most important OTUs.
-
-We also compared the top 20 OTUs selected by permutation importance in
-L2-regularized logistic regression and the highest performing tree-based
-models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
-shared among the models. These results indicate that we were able to
-identify important OTUs that are shared across the highest performing
-linear and non-linear models when we use permutation importance as our
-interpretation method.
-
-We then evaluated the difference in relative abundances of the top 20
-OTUs identified in L2-regularized logistic regression and random forest
-models between healthy patients and patients with SRNs {[}Figure S8{]}.
-There were minimal differences in relative abundances across OTUs
-between different diagnoses. This supports our claim that it is not
-possible to differentiate disease vs healthy states by focusing on
-individual taxa.
-
-\textbf{The computational efficiency of each ML model.} We compared the
-training times of the seven ML models. The training times increased with
-the complexity of the model and the number of potential hyperparameter
-combinations. Also, the linear models trained faster than non-linear
-models {[}Figures S1-S2; Figure 5{]}.
-
-\subsection{Discussion}\label{discussion}
-
-There is a growing awareness that many human diseases and environmental
-processes are not driven by a single organism but are the product of
-multiple bacterial populations. Traditional statistical approaches are
-useful for identifying those cases where a single organism is associated
-with a process. In contrast, ML methods offer the ability to incorporate
-the structure of the microbial communities as a whole and identify
-associations between community structure and disease state. If it is
-possible to classify communities reliably, then ML methods also offer
-the ability to identify those microbial populations within the
-communities that are responsible for the classification. However, the
-application of ML in microbiome studies is still in its infancy, and the
-field needs to develop a better understanding of different ML methods,
-their strengths and weaknesses, and how to implement them.
-
-To address these needs, we developed an open-sourced framework for ML
-models. Using this pipeline, we benchmarked seven ML models and showed
-that the tradeoff between model complexity and performance may be less
-severe than originally hypothesized. In terms of predictive performance,
-the random forest model had the best AUROC compared to the other six
-models. However, the second-best model was L2-regularized logistic
-regression with a median AUROC difference of less than 0.015 compared to
-random forest. While our implementation of random forest took 83.2 hours
-to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model,
-while L2-regularized logistic regression, a linear model, is easily
-interpreted according to the feature weights. Comparing many different
-models showed us that the most complex model was not necessarily the
-best model for our ML task.
-
-We established a pipeline that can be generalized to any modeling method
-that predicts a binary health outcome. We performed a random data-split
-to create a training set (80\% of the data) and a held-out test set
-(20\% of the data), which we used to evaluate predictive performance. We
-used the AUROC metric to evaluate predictive performance as it is a
-clinically relevant evaluation metric for our study. We repeated this
-data-split 100 times to measure the possible variation in predictive
-performance. During training, we tuned the model hyperparameters with a
-repeated five-fold cross-validation. Despite the high number of features
-microbiome datasets typically have, the models we built with this
-pipeline generalized to the held-out test sets.
-
-We highlighted the importance of model interpretation to gain greater
-biological insights into microbiota-associated diseases. In this study,
-we showcased two different interpretation methods: ranking each OTU by
-(i) their absolute weights in the trained models and (ii) their impact
-on the predictive performance based on permutation importance.
-Human-associated microbial communities have complex correlation
-structures that create collinearity in the datasets. This can hinder our
-ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (37). To capture all
-important features, once we identify highly ranked OTUs, we should
-review their relationships with other OTUs. These relationships will
-help us generate new hypotheses about the ecology of the disease and
-test them with follow-up experiments. When we used permutation
-importance, we partially accounted for collinearity by grouping
-correlated OTUs to determine their impact as a group. We grouped OTUs
-that had a perfect correlation with each other; however, we could reduce
-the correlation threshold to further investigate the relationships among
-correlated features. We are choosing to leave the decision to define
-correlation and setting up correlations thresholds to the researchers
-for their own microbiome questions. However, we would like to urge them
-to investigate the correlation structures of the data to avoid
-misinterpreting the models. This is likely to be a particular problem
-with shotgun metagenomic datasets where collinearity will be more
-pronounced due to many genes being correlated with one another because
-they come from the same chromosome. True causal mechanisms (e.g., role
-of microbiome in colorectal cancer) cannot be explained solely by the
-highest performing machine learning model (38). To identify the true
-underlying microbial factors of a disease, it is crucial to follow up on
-any correlation analyses with further hypothesis testing and
-experimentation for biological validation.
-
-In this study, we did not consider all possible modeling approaches.
-However, the principles highlighted throughout this study apply to other
-ML modeling tasks with microbiome data. For example, we did not evaluate
-multicategory classification methods to predict non-binary outcomes. We
-could have trained models to differentiate between people with healthy
-colons and those with adenomas or carcinomas (k=3 categories). We did
-not perform this analysis because the clinically relevant diagnosis
-grouping was between patients with healthy colons and those with SRNs.
-Furthermore, as the number of classes increases, more samples are
-required for each category to train an accurate model. We also did not
-use regression-based analyses to predict a non-categorical outcome. We
-have previously used such an approach to train random forest models to
-predict fecal short-chain fatty acid concentrations based on microbiome
-data (39). Our analysis was also limited to shallow learning methods and
-did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (12, 40, 41), but microbiome datasets
-often suffer from having many features and small sample sizes, which
-result in overfitting.
-
-Our framework provides a reproducible structure to investigators wanting
-to train, evaluate, and interpret their own ML models to generate
-hypotheses regarding which OTUs might be biologically relevant. However,
-deploying microbiome-based models to make clinical diagnoses or
-predictions is a significantly more challenging and distinct undertaking
-(42). For example, we currently lack standardized methods to collect
-patient samples, generate sequence data, and report clinical data. We
-are also challenged by the practical constraints of OTU-based
-approaches. The de novo algorithms commonly in use are slow, require
-considerable memory, and result in different OTU assignments as new data
-are added (43). Finally, we also need independent validation cohorts to
-test the performance of a diagnostic model. To realize the potential for
-using ML approaches with microbiome data, it is necessary that we direct
-our efforts to overcome these challenges.
-
-Our study highlights the need to make educated choices at every step of
-developing an ML model with microbiome data. We created an aspirational
-rubric that researchers can use to identify potential pitfalls when
-using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
-have highlighted the trade-offs between model complexity and
-interpretability, the need for tuning hyperparameters, the utility of
-held-out test sets for evaluating predictive performance, and the
-importance of considering correlation structures in datasets for
-reliable interpretation. Furthermore, we underscored the importance of
-proper experimental design and methods to help us achieve the level of
-validity and accountability we want from models built for patient
-health.
-
-\subsection{Materials and Methods}\label{materials-and-methods}
-
-\textbf{Data collection and study population.} The original stool
-samples described in our analysis were obtained from patients recruited
-by Great Lakes-New England Early Detection Research Network (5). Stool
-samples were provided by adults who were undergoing a scheduled
-screening or surveillance colonoscopy. Participants were recruited from
-Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
-(MI, USA). Patients' colonic health was visually assessed by colonoscopy
-with bowel preparation and tissue histopathology of all resected
-lesions. We assigned patients into two classes: those with healthy
-colons and those with screen relevant neoplasias (SRNs). The healthy
-class included patients with healthy colons or non-advanced adenomas,
-whereas the SRN class included patients with advanced adenomas or
-carcinomas (44). Patients with an adenoma greater than 1 cm, more than
-three adenomas of any size, or an adenoma with villous histology were
-classified as having advanced adenomas (44). There were 172 patients
-with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
-Of the 198 adenomas, 109 were identified as advanced adenomas. Together
-261 patients were classified as healthy and 229 patients were classified
-as having an SRN.
-
-\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
-patients were used for 16S rRNA gene sequencing to measure bacterial
-population abundances. The sequence data used in our analyses were
-originally generated by Baxter et al. (available through NCBI Sequence
-Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
-generated by Sze et al (45), who processed the 16S rRNA sequences in
-mothur (v1.39.3) using the default quality filtering methods,
-identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
-(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
-These OTU abundances were the features we used to predict colorectal
-health of the patients. There were 6920 OTUs. OTU abundances were
-subsampled to the size of the smallest sample and normalized across
-samples such that the highest abundance of each OTU would be 1, and the
-lowest would be 0.
-
-\textbf{Model training and evaluation.} Models were trained using the
-caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
-calculate decision values for models generated using L2-regularized SVM
-with linear kernel and L1-regularized SVM with linear kernel. The code
-for these changes on L2-regularized SVM with linear kernel and
-L1-regularized SVM with linear kernel models are available at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
-and at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
-respectively.
-
-For hyperparameter selection, we started with a granular grid search.
-Then we narrowed and fine-tuned the range of each hyperparameter.For
-L2-regularized logistic regression, L1 and L2-regularized SVM with
-linear and radial basis function kernels, we tuned the cost
-hyperparameter, which controls the regularization strength, where
-smaller values specify stronger regularization. For SVM with radial
-basis function kernel, we also tuned the sigma hyperparameter, which
-determines the reach of a single training instance where, for a high
-value of sigma, the SVM decision boundary will be dependent on the
-points that are closest to the decision boundary. For the decision tree
-model, we tuned the depth of the tree where the deeper the tree, the
-more splits it has. For random forest, we tuned the number of features
-to consider when looking for the best tree split. For XGBoost, we tuned
-the learning rate and the fraction of samples used for fitting the
-individual base learners. Performing a grid search for hyperparameter
-selection might not be feasible for when there are more than two
-hyperparameters to tune for. In such cases, it is more efficient to use
-random search or recently developed tools such as Hyperband to identify
-good hyperparameter configurations (48).
-
-The computational burden during model training due to model complexity
-was reduced by parallelizing segments of the ML pipeline. We
-parallelized the training of each data-split. This allowed the 100
-data-splits to be processed through the ML pipeline simultaneously at
-the same time for each model. It is possible to further parallelize the
-cross-validation step for each hyperparameter setting which we have not
-performed in this study.
-
-\textbf{Permutation importance workflow.} We calculated a Spearman's
-rank-order correlation matrix and defined correlated OTUs as having
-perfect correlation (correlation coefficient = 1 and p \textless{}
-0.01). OTUs without a perfect correlation to each other were permuted
-individually, whereas correlated ones were grouped together and permuted
-at the same time.
-
-\textbf{Statistical analysis workflow.} Data summaries, statistical
-analysis, and data visualizations were performed using R (v.3.5.0) with
-the tidyverse package (v.1.2.1). We compared the performance of the
-models pairwise by calculating the difference between AUROC values from
-the same data-split (for 100 data-splits). We determined if the models
-were significantly different by calculating the empirical p-value (2 x
-min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
-0) for the double tail event (e.g., Figure S3).
-
-\textbf{Code availability.} The code for all sequence curation and
-analysis steps including an Rmarkdown version of this manuscript is
-available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
-
-\textbf{Acknowledgements.} We thank all the study participants of Great
-Lakes-New England Early Detection Research Network. We would like to
-thank the members of the Schloss lab for their valuable feedback. Salary
-support for MR came from NIH grant 1R01CA215574. Salary support for PDS
-came from NIH grants P30DK034933 and 1R01CA215574.
-
-\newpage
-
-\subsection{References}\label{references}
-
-\hypertarget{refs}{}
-\hypertarget{ref-segata_metagenomic_2011}{}
-1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
-\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
-\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
-explanation. Genome Biol \textbf{12}:R60.
-doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.
-
-\hypertarget{ref-zeller_potential_2014}{}
-2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
-\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
-\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
-\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
-\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
-\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
-J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
-\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
-\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
-for early-stage detection of colorectal cancer. Mol Syst Biol
-\textbf{10}.
-doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
-
-\hypertarget{ref-zackular_human_2014}{}
-3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
-\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
-for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
-doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
-
-\hypertarget{ref-baxter_dna_2016}{}
-4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
-\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
-immunochemical test can replace stool for detection of colonic lesions
-using a microbiota-based model. Microbiome \textbf{4}.
-doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
-
-\hypertarget{ref-baxter_microbiota-based_2016}{}
-5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\hypertarget{ref-hale_shifts_2017}{}
-6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
-\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
-\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
-\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
-\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the fecal
-microbiota associated with adenomatous polyps. Cancer Epidemiol
-Biomarkers Prev \textbf{26}:85--94.
-doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.
-
-\hypertarget{ref-pasolli_machine_2016}{}
-7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
-\textbf{Waldron L}, \textbf{Segata N}. 2016. Machine learning
-meta-analysis of large metagenomic datasets: Tools and biological
-insights. PLoS Comput Biol \textbf{12}.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1004977}{10.1371/journal.pcbi.1004977}.
-
-\hypertarget{ref-sze_looking_2016}{}
-8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
-the noise: Revisiting obesity and the microbiome. mBio \textbf{7}.
-doi:\href{https://doi.org/10.1128/mBio.01018-16}{10.1128/mBio.01018-16}.
-
-\hypertarget{ref-walters_meta-analyses_2014}{}
-9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
-Meta-analyses of human gut microbes associated with obesity and IBD.
-FEBS Lett \textbf{588}:4223--4233.
-doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
-
-\hypertarget{ref-vazquez-baeza_guiding_2018}{}
-10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
-\textbf{Washburne A}, \textbf{Herfarth HH}, \textbf{Sartor RB},
-\textbf{Knight R}. 2018. Guiding longitudinal sampling in IBD cohorts.
-Gut \textbf{67}:1743--1745.
-doi:\href{https://doi.org/10.1136/gutjnl-2017-315352}{10.1136/gutjnl-2017-315352}.
-
-\hypertarget{ref-qin_alterations_2014}{}
-11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
-\textbf{Chen Y}, \textbf{Shao L}, \textbf{Guo J}, \textbf{Le Chatelier
-E}, \textbf{Yao J}, \textbf{Wu L}, \textbf{Zhou J}, \textbf{Ni S},
-\textbf{Liu L}, \textbf{Pons N}, \textbf{Batto JM}, \textbf{Kennedy SP},
-\textbf{Leonard P}, \textbf{Yuan C}, \textbf{Ding W}, \textbf{Chen Y},
-\textbf{Hu X}, \textbf{Zheng B}, \textbf{Qian G}, \textbf{Xu W},
-\textbf{Ehrlich SD}, \textbf{Zheng S}, \textbf{Li L}. 2014. Alterations
-of the human gut microbiome in liver cirrhosis. Nature
-\textbf{513}:59--64.
-doi:\href{https://doi.org/10.1038/nature13568}{10.1038/nature13568}.
-
-\hypertarget{ref-geman_deep_2018}{}
-12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
-\textbf{Doloc C}, \textbf{Milici M-R}, \textbf{Milici L-D}. 2018. Deep
-learning tools for human microbiome big data, pp. 265--275. \emph{In}
-Balas, VE, Jain, LC, Balas, MM (eds.), Soft computing applications.
-Springer International Publishing.
-
-\hypertarget{ref-thaiss_persistent_2016}{}
-13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
-\textbf{Meijer MT}, \textbf{Levy M}, \textbf{Moresi C},
-\textbf{Dohnalová L}, \textbf{Braverman S}, \textbf{Rozin S},
-\textbf{Malitsky S}, \textbf{Dori-Bachash M}, \textbf{Kuperman Y},
-\textbf{Biton I}, \textbf{Gertler A}, \textbf{Harmelin A},
-\textbf{Shapiro H}, \textbf{Halpern Z}, \textbf{Aharoni A},
-\textbf{Segal E}, \textbf{Elinav E}. 2016. Persistent microbiome
-alterations modulate the rate of post-dieting weight regain. Nature
-\textbf{540}:544--551.
-doi:\href{https://doi.org/10.1038/nature20796}{10.1038/nature20796}.
-
-\hypertarget{ref-dadkhah_gut_2019}{}
-14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
-\textbf{Hardi R}, \textbf{Baybick J}, \textbf{Hanzel D}, \textbf{Kuehn
-G}, \textbf{Kuehn T}, \textbf{Gillevet PM}. 2019. Gut microbiome
-identifies risk for colorectal polyps. BMJ Open Gastroenterology
-\textbf{6}:e000297.
-doi:\href{https://doi.org/10.1136/bmjgast-2019-000297}{10.1136/bmjgast-2019-000297}.
-
-\hypertarget{ref-flemer_oral_2018}{}
-15. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
-\textbf{Cisek K}, \textbf{Das A}, \textbf{Jeffery IB}, \textbf{Hurley
-E}, \textbf{O`Riordain M}, \textbf{Shanahan F}, \textbf{O`Toole PW}.
-2018. The oral microbiota in colorectal cancer is distinctive and
-predictive. Gut \textbf{67}:1454--1463.
-doi:\href{https://doi.org/10.1136/gutjnl-2017-314814}{10.1136/gutjnl-2017-314814}.
-
-\hypertarget{ref-montassier_pretreatment_2016}{}
-16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
-\textbf{Corvec S}, \textbf{Gastinne T}, \textbf{Potel G}, \textbf{Moreau
-P}, \textbf{Cochetiere MF de la}, \textbf{Batard E}, \textbf{Knights D}.
-2016. Pretreatment gut microbiome predicts chemotherapy-related
-bloodstream infection. Genome Medicine \textbf{8}:49.
-doi:\href{https://doi.org/10.1186/s13073-016-0301-4}{10.1186/s13073-016-0301-4}.
-
-\hypertarget{ref-ai_systematic_2017}{}
-17. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
-\textbf{Xu J}, \textbf{Fang J-Y}. 2017. Systematic evaluation of
-supervised classifiers for fecal microbiota-based prediction of
-colorectal cancer. Oncotarget \textbf{8}:9546--9556.
-doi:\href{https://doi.org/10.18632/oncotarget.14488}{10.18632/oncotarget.14488}.
-
-\hypertarget{ref-dai_multi-cohort_2018}{}
-18. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
-WKK}, \textbf{Zhao L}, \textbf{Chen Z}, \textbf{Chan FKL},
-\textbf{Kristiansen K}, \textbf{Sung JJY}, \textbf{Wong SH}, \textbf{Yu
-J}. 2018. Multi-cohort analysis of colorectal cancer metagenome
-identified altered bacteria across populations and universal bacterial
-markers. Microbiome \textbf{6}:70.
-doi:\href{https://doi.org/10.1186/s40168-018-0451-2}{10.1186/s40168-018-0451-2}.
-
-\hypertarget{ref-mossotto_classification_2017}{}
-19. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
-\textbf{Beattie RM}, \textbf{MacArthur BD}, \textbf{Ennis S}. 2017.
-Classification of paediatric inflammatory bowel disease using machine
-learning. Scientific Reports \textbf{7}.
-doi:\href{https://doi.org/10.1038/s41598-017-02606-2}{10.1038/s41598-017-02606-2}.
-
-\hypertarget{ref-wong_quantitation_2017}{}
-20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
-AKC}, \textbf{Dai RZW}, \textbf{Nakatsu G}, \textbf{Lam TYT},
-\textbf{Zhang L}, \textbf{Wu JCY}, \textbf{Chan FKL}, \textbf{Ng SSM},
-\textbf{Wong MCS}, \textbf{Ng SC}, \textbf{Wu WKK}, \textbf{Yu J},
-\textbf{Sung JJY}. 2017. Quantitation of faecal fusobacterium improves
-faecal immunochemical test in detecting advanced colorectal neoplasia.
-Gut \textbf{66}:1441--1448.
-doi:\href{https://doi.org/10.1136/gutjnl-2016-312766}{10.1136/gutjnl-2016-312766}.
-
-\hypertarget{ref-statnikov_comprehensive_2013}{}
-21. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
-\textbf{Konganti K}, \textbf{Li Z}, \textbf{Yang L}, \textbf{Pei Z},
-\textbf{Blaser MJ}, \textbf{Aliferis CF}, \textbf{Alekseyenko AV}. 2013.
-A comprehensive evaluation of multicategory classification methods for
-microbiomic data. Microbiome \textbf{1}:11.
-doi:\href{https://doi.org/10.1186/2049-2618-1-11}{10.1186/2049-2618-1-11}.
-
-\hypertarget{ref-knights_supervised_2011}{}
-22. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
-Supervised classification of human microbiota. FEMS Microbiology Reviews
-\textbf{35}:343--359.
-doi:\href{https://doi.org/10.1111/j.1574-6976.2010.00251.x}{10.1111/j.1574-6976.2010.00251.x}.
-
-\hypertarget{ref-wirbel_meta-analysis_2019}{}
-23. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
-K}, \textbf{Kashani A}, \textbf{Milanese A}, \textbf{Fleck JS},
-\textbf{Voigt AY}, \textbf{Palleja A}, \textbf{Ponnudurai R},
-\textbf{Sunagawa S}, \textbf{Coelho LP}, \textbf{Schrotz-King P},
-\textbf{Vogtmann E}, \textbf{Habermann N}, \textbf{Niméus E},
-\textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Gandini S},
-\textbf{Serrano D}, \textbf{Mizutani S}, \textbf{Shiroma H},
-\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
-T}, \textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N},
-\textbf{Sinha R}, \textbf{Ulrich CM}, \textbf{Brenner H},
-\textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller G}. 2019.
-Meta-analysis of fecal metagenomes reveals global microbial signatures
-that are specific for colorectal cancer. Nature Medicine
-\textbf{25}:679.
-doi:\href{https://doi.org/10.1038/s41591-019-0406-6}{10.1038/s41591-019-0406-6}.
-
-\hypertarget{ref-vangay_microbiome_2019}{}
-24. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
-Microbiome learning repo (ML repo): A public repository of microbiome
-regression and classification tasks. Gigascience \textbf{8}.
-doi:\href{https://doi.org/10.1093/gigascience/giz042}{10.1093/gigascience/giz042}.
-
-\hypertarget{ref-galkin_human_2018}{}
-25. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
-\textbf{Kuznetsov I}, \textbf{Gladyshev VN}, \textbf{Zhavoronkov A}.
-2018. Human microbiome aging clocks based on deep learning and tandem of
-permutation feature importance and accumulated local effects. bioRxiv.
-doi:\href{https://doi.org/10.1101/507780}{10.1101/507780}.
-
-\hypertarget{ref-reiman_using_2017}{}
-26. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
-convolutional neural networks to explore the microbiome, pp. 4269--4272.
-\emph{In} 2017 39th annual international conference of the IEEE
-engineering in medicine and biology society (EMBC).
-
-\hypertarget{ref-fioravanti_phylogenetic_2017}{}
-27. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
-\textbf{Agostinelli C}, \textbf{Chierici M}, \textbf{Jurman G},
-\textbf{Furlanello C}. 2017. Phylogenetic convolutional neural networks
-in metagenomics. arXiv:170902268 {[}cs, q-bio{]}.
-
-\hypertarget{ref-thomas_metagenomic_2019}{}
-28. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
-\textbf{Pasolli E}, \textbf{Armanini F}, \textbf{Zolfo M},
-\textbf{Beghini F}, \textbf{Manara S}, \textbf{Karcher N}, \textbf{Pozzi
-C}, \textbf{Gandini S}, \textbf{Serrano D}, \textbf{Tarallo S},
-\textbf{Francavilla A}, \textbf{Gallo G}, \textbf{Trompetto M},
-\textbf{Ferrero G}, \textbf{Mizutani S}, \textbf{Shiroma H},
-\textbf{Shiba S}, \textbf{Shibata T}, \textbf{Yachida S}, \textbf{Yamada
-T}, \textbf{Wirbel J}, \textbf{Schrotz-King P}, \textbf{Ulrich CM},
-\textbf{Brenner H}, \textbf{Arumugam M}, \textbf{Bork P}, \textbf{Zeller
-G}, \textbf{Cordero F}, \textbf{Dias-Neto E}, \textbf{Setubal JC},
-\textbf{Tett A}, \textbf{Pardini B}, \textbf{Rescigno M},
-\textbf{Waldron L}, \textbf{Naccarati A}, \textbf{Segata N}. 2019.
-Metagenomic analysis of colorectal cancer datasets identifies
-cross-cohort microbial diagnostic signatures and a link with choline
-degradation. Nature Medicine \textbf{25}:667.
-doi:\href{https://doi.org/10.1038/s41591-019-0405-7}{10.1038/s41591-019-0405-7}.
-
-\hypertarget{ref-rudin_please_2018}{}
-29. \textbf{Rudin C}. 2018. Please stop explaining black box models for
-high stakes decisions. arXiv:181110154 {[}cs, stat{]}.
-
-\hypertarget{ref-rudin_optimized_2018}{}
-30. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
-Toward trust in machine learning for healthcare and criminal justice.
-Interfaces \textbf{48}:449--466.
-doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.
-
-\hypertarget{ref-Quinn847475}{}
-31. \textbf{Quinn TP}, \textbf{Erb I}. 2019. Another look at
-microbemetabolite interactions: How scale invariant correlations can
-outperform a neural network. bioRxiv.
-doi:\href{https://doi.org/10.1101/847475}{10.1101/847475}.
-
-\hypertarget{ref-knights_human-associated_2011}{}
-32. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
-\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
-signatures: Examining their predictive value. Cell Host Microbe
-\textbf{10}:292--296.
-doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.
-
-\hypertarget{ref-miller_explanation_2017}{}
-33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
-Insights from the social sciences. arXiv:170607269 {[}cs{]}.
-
-\hypertarget{ref-ribeiro_why_2016}{}
-34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
-``Why should i trust you?'': Explaining the predictions of any
-classifier. arXiv:160204938 {[}cs, stat{]}.
-
-\hypertarget{ref-nori_interpretml:_2019}{}
-35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
-\textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
-learning interpretability. arXiv:190909223 {[}cs, stat{]}.
-
-\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
-36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
-\textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
-importance measure. Bioinformatics \textbf{26}:1340--1347.
-doi:\href{https://doi.org/10.1093/bioinformatics/btq134}{10.1093/bioinformatics/btq134}.
-
-\hypertarget{ref-dormann_collinearity:_2013}{}
-37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
-\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
-JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
-\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
-\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
-\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
-of methods to deal with it and a simulation study evaluating their
-performance. Ecography \textbf{36}:27--46.
-doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.
-
-\hypertarget{ref-li_accurate_2020}{}
-38. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
-Accurate data-driven prediction does not mean high reproducibility. Nat
-Mach Intell \textbf{2}:13--15.
-doi:\href{https://doi.org/10.1038/s42256-019-0140-2}{10.1038/s42256-019-0140-2}.
-
-\hypertarget{ref-sze_fecal_2019}{}
-39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
-\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
-acids are not predictive of colonic tumor status and cannot be predicted
-based on bacterial community structure. mBio \textbf{10}:e01454--19.
-doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.
-
-\hypertarget{ref-kocheturov_massive_2019}{}
-40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
-2019. Massive datasets and machine learning for computational
-biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
-doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.
-
-\hypertarget{ref-kim_improved_2018}{}
-41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
-method for prediction of cancer prognosis by network learning. Genes
-\textbf{9}:478.
-doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.
-
-\hypertarget{ref-wiens_no_2019}{}
-42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
-\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
-\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
-\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
-2019. Do no harm: A roadmap for responsible machine learning for health
-care. Nat Med \textbf{25}:1337--1340.
-doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.
-
-\hypertarget{ref-westcott_opticlust_2017}{}
-43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
-Improved Method for Assigning Amplicon-Based Sequence Data to
-Operational Taxonomic Units. mSphere \textbf{2}.
-doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
-
-\hypertarget{ref-redwood_stool_2016}{}
-44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
-\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
-\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
-\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
-TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
-for screening detection of colorectal neoplasia in alaska native people.
-Mayo Clin Proc \textbf{91}:61--70.
-doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.
-
-\hypertarget{ref-sze_leveraging_2018}{}
-45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
-rRNA gene surveys to identify reproducible biomarkers in individuals
-with colorectal tumors. mBio \textbf{9}:e00630--18.
-doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
-
-\hypertarget{ref-schloss_introducing_2009}{}
-46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-Source, Platform-Independent,
-Community-Supported Software for Describing and Comparing Microbial
-Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.
-
-\hypertarget{ref-rognes_vsearch_2016}{}
-47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
-\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
-source tool for metagenomics. PeerJ \textbf{4}:e2584.
-doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
-
-\hypertarget{ref-li_hyperband:_2016}{}
-48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
-\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
-bandit-based approach to hyperparameter optimization. arXiv:160306560
-{[}cs, stat{]}.
 
@@
 \captionsetup{labelformat=empty}
-\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
-\small
-
-\begin{tabular}{|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-
-\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
-
-Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
-
-\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
-
-\end{tabular}\newpage
-
-\newpage
-
-\includegraphics{Figure_1.pdf} \textbf{Figure 1. Machine learning
-pipeline.} We split the data to create a training (80\%) and held-out
-test set (20\%). The splits were stratified to maintain the overall
-class distribution. We performed five-fold cross-validation on the
-training data to select the best hyperparameter setting and then used
-these hyperparameters to train the models. The model was evaluated on
-the held-out data set. Abbreviations: cvAUC, cross-validation area under
-the receiver operating characteristic curve. \newpage
-\includegraphics{Figure_2.png}
-
-\textbf{Figure 2. Generalization and classification performance of ML
-models using AUROC values of all cross-validation and testing
-performances.} The median AUROC for diagnosing individuals with SRN
-using bacterial abundances was higher than chance (depicted by a
-horizontal line at 0.50) for all the ML models. The predictive
-performance of random forest model was higher than other ML models,
-though not significantly (p \textgreater{} 0.01). L2-regularized
-logistic regression, XGBoost, L2-regularized SVM with linear and radial
-basis function kernel performances were not significantly different from
-one another. The boxplot shows quartiles at the box ends and the median
-as the horizontal line in the box. The whiskers show the farthest points
-that were not outliers. Outliers were defined as those data points that
-are not within 1.5 times the interquartile ranges. \newpage
-\includegraphics[height=17cm, width=12cm]{Figure_3.png}
-
-\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
-absolute feature weights of (A) L1-regularized SVM with linear kernel,
-(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
-logistic regression, were ranked from highest rank, 1, to lowest rank,
-100, for each data-split. The feature ranks of the 20 highest ranked
-OTUs based on their median ranks (median shown in black) are reported
-here. OTUs that were associated with classifying a subject as being
-healthy had negative signs and were shown in blue. OTUs that were
-associated with classifying a subject having an SRN had positive signs
-and were shown in red. \newpage
-\includegraphics{Figure_4.png}
-
-\textbf{Figure 4. Interpretation of the non-linear ML models.} (A) SVM
-with radial basis kernel, (B) decision tree, (C) random forest, and (D)
-XGBoost feature importances were explained using permutation importance
-on the held-out test data set. The gray rectangle and the dashed line
-show the IQR range and median of the base testing AUROC without any
-permutation. The 20 OTUs that caused the largest decrease in the AUROC
-when permuted are reported here. The colors of the box plots represent
-the OTUs that were shared among the different models; yellow were OTUs
-that were shared among all the non-linear models, green were OTUs that
-were shared among the tree-based models, green were the OTUs shared
-among SVM with radial basis kernel, decision tree and XGBoost, pink were
-the OTUs shared among SVM with radial basis kernel and XGBoost only, red
-were the OTUs shared among random forest and XGBoost only and blue were
-the OTUs shared among decision tree and random forest only. For all of
-the tree-based models, a \emph{Peptostreptococcus} species (OTU00367)
-had the largest impact on predictive performance. \newpage
-\includegraphics{Figure_5.png}
-
-\textbf{Figure 5. Training times of seven ML models.} The median
-training time was the highest for XGBoost and shortest for
-L2-regularized logistic regression. \newpage
-
 \captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
@@
 
-\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
-setting performances for linear models.} (A) L2-regularized logistic
-regression, (B) L1-regularized SVM with linear kernel, and (C)
-L2-regularized SVM with linear kernel mean cross-validation AUROC values
-when different hyperparameters were used in training the model. The
-stars represent the highest performing hyperparameter setting for each
-model. \newpage
-\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
-
-\textbf{Figure S2. Hyperparameter setting performances for non-linear
-models.} (A) Decision tree, (B) random forest, (C) SVM with radial basis
-kernel, and (D) XGBoost mean cross-validation AUROC values when
-different hyperparameters were used in training the model. The stars
-represent the highest performing hyperparameter setting for the models.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}
-
-\textbf{Figure S3. Histogram of AUROC differences between L2-regularized
-logistic regression and random forest for each of the hundred
-data-splits.} In 75\% of data-splits, the AUROC of random forest was
-greater than that of L2-regularized logistic regression. The p-value was
-manually calculated using the sampling distribution of the test
-statistic under the null hypothesis. We tested how often random forest
-performed more accurately than L2-regularized logistic regression. The
-null hypothesis is that the distribution of the difference between the
-AUROC values of random forest and L2 logistic regression is symmetric
-about 0, therefore the p-value was calculated for a double-tail event.
-\newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}
-
-\textbf{Figure S4. Classification performance of ML models across
-cross-validation when trained on a subset of the dataset.} (A)
-L2-regularized logistic regression and (B) random forest models were
-trained using the original study design with 490 subjects and subsets of
-it with 15, 30, 60, 120, and 245 subjects. The range among the
-cross-validation AUROC values within both models at smaller sample sizes
-were much larger than when the full collection of samples was used to
-train and validate the models but included the ranges observed with the
-more complete datasets. \newpage
-\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}
-
-\textbf{Figure S5. NMDS ordination of Bray-Curtis distances.} NMDS
-ordination relating the community structures of the fecal microbiota
-from 490 patients (261 patients with normal colonoscopies and 229
-patients who have screen relevant neoplasias; SRNs).
-
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}
-
-\textbf{Figure S6. Permutation importance analysis.} Permutation
-importance analysis measures the decrease in the predictive performance
-of the model after we permute (A) a feature's or (B) a group of
-correlated features' values, which breaks the relationship between the
-feature and the diagnosis.
-
-\newpage
-
-\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}
-
-\textbf{Figure S7. Interpretation of the linear ML models with
-permutation importance.} (A) L1-regularized SVM with linear kernel, (B)
-L2-regularized SVM with linear kernel, and (C) L2-regularized logistic
-regression were interpreted using permutation importance using held-out
-test set.
-
-\newpage
-
-\includegraphics[height=20cm, width=13cm]{Figure_S8.pdf}
-
-\textbf{Figure S8. Relative abundances of the 20 most important OTUs in
-L2-regularized logistic regression and random forest models.} The most
-important 20 OTUs were chosen for (A) Random forest and (B)
-L2-regularized logistic regression models by permutation importance and
-ranking feature coefficients, respectively. The relative abundances of
-these OTUs were compared based on the diagnosis of the patients. The
-minimal differences betweeen relative abundances for these OTUs show
-that it is not possible to differentiate disease vs healthy states by
-focusing on individual taxa.
-
 \end{document}

### 2020-01-28 submission/manuscript.tex

--- 2020-01-28 submission/manuscript.tex
@@
 \hypersetup{
-  pdftitle={A framework for effective application of machine learning to a microbiome-based classification problem},
+  pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
   hidelinks,
@@
 
-\title{\textbf{A framework for effective application of machine learning to a
-microbiome-based classification problem}}
+\title{\textbf{A framework for effective application of machine learning to
+microbiome-based classification problems}}
 \author{}
@@
 We also compared the top 20 OTUs selected by permutation importance in
-L2-regularized logistic regression and the highest performing tree-based
-models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
-shared among the models. These results indicate that we were able to
-identify important OTUs that are shared across the highest performing
-linear and non-linear models when we use permutation importance as our
-interpretation method.
+L2-regularized logistic regression {[}Figure S7{]} and the highest
+performing tree-based models, random forest and XGBoost {[}Figure 4{]}.
+2 and 5 OTUs, respectively, were shared among the models. These results
+indicate that we were able to identify important OTUs that are shared
+across the highest performing linear and non-linear models when we use
+permutation importance as our interpretation method.
 
@@
 combinations. Also, the linear models trained faster than non-linear
-models {[}Figures S1-S2; Figure 5{]}.
+models {[}Figures S2-S3; Figure 5{]}.
 
@@
 
-\newpage
-
 \textbf{Figure 1. Machine learning pipeline.} We split the data to
@@
 curve.
+
+\hfill\break
 
@@
 
+\hfill\break
+
 \textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
@@
 and were shown in red.
+
+\hfill\break
 
@@
 
+\hfill\break
+
 \textbf{Figure 5. Training times of seven ML models.} The median
 training time was the highest for XGBoost and shortest for
-L2-regularized logistic regression. \newpage
-
-\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
-\small
-\begin{tabular}{|l|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline
-
-\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline
-
-\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population to which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline
-
-\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline
-
-\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline
-
-\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline
-
-\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g., when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline
-
-\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
-
-\end{tabular}
+L2-regularized logistic regression.
 
@@
 patients who have screen relevant neoplasias; SRNs).
+
+\hfill\break
 
@@
 
+\hfill\break
+
 \textbf{Figure S3. Hyperparameter setting performances for non-linear
@@
 represent the highest performing hyperparameter setting for the models.
+
+\hfill\break
 
@@
 
+\hfill\break
+
 \textbf{Figure S5. Classification performance of ML models across
@@
 
+\hfill\break
+
 \textbf{Figure S6. Permutation importance analysis.} Permutation
@@
 
+\hfill\break
+
 \textbf{Figure S7. Interpretation of the linear ML models with
@@
 test set.
+
+\hfill\break
 

### 2020-01-28 submission/manuscript.Rmd
Pat's edits

@@
 
-\vspace{35mm}
+\vspace{30mm}
 
@@
 
-\vspace{35mm}
+\vspace{20mm}
 
@@
 
-\vspace{38mm}
+\vspace{30mm}
 
@@
 To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost).
-The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739] but was slow to train (83.2 h) and not immediately interpretable.
+The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739] but was slow to train (83.2 h) and not inherently interpretable.
 Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was inherently interpretable.
@@
 
-Prediction of diseases using machine learning (ML) is rapidly being adopted in microbiome studies.
-However, the estimated performance associated with these ML models is likely over-optimistic.
+Diagnosing diseases using machine learning (ML) is rapidly being adopted in microbiome studies.
+However, the estimated performance associated with these models is likely over-optimistic.
 Moreover, there is a trend towards using black box models without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease.
@@
 
-As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease.
+As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases.
 However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models.
-For example it is not possible to classify individuals as having healthy colons or colorectal lesions using Bray-Curtis distances based on the 16S rRNA gene sequences collected from fecal samples [Figure S1].
+For example it is not possible to classify individuals as having healthy colons or screen relevant neoplasia using Bray-Curtis distances based on the 16S rRNA gene sequences collected from fecal samples [Figure S1].
 This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals.
@@
 Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population separately [@segata_metagenomic_2011].
-Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease.
+Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease because they consider the relative abundance of each bacterial population in the context of others rather than in isolation.
 
@@
 The task of diagnosing an individual relies on a rigorously validated model.
-However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress.
+However, there are common methodological and reporting problems that arise when applying ML to such data that need to be addressed for the field to progress.
 These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances.
@@
 In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed.
-To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing the predictive performance, interpretability and training times of seven modeling approaches with the same dataset and pipeline.
+To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing the predictive performance, interpretability, data requirements, and training times of seven modeling approaches with the same dataset and pipeline.
 We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel.
@@
 We compared their predictive performance, interpretability, and training time.
-To demonstrate the performance of these modeling approaches and our pipeline on a case study, we used a data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
+To demonstrate the performance of these modeling approaches and our pipeline, we present a case study using data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016].
 This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011].
@@
 This framework can be easily applied to other host-associated and environmental microbiome datasets.
-We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Table S1] to urge the audience to be diligent with with their study design, model selection, development, evaluation and interpretation.
+We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Table S1] to urge the audience to be diligent in their study design and model selection, development, evaluation, and interpretation.
 
@@
 It was encouraging that many of the highest-ranked OTUs were shared across these three models (e.g., OTUs 50, 426, 609, 822, 1239).
-The great benefit of this approach was knowing the sign and magnitude of each OTU coefficient in the trained model.
+The benefit of this approach was knowing the sign and magnitude of each OTU coefficient in the trained model.
 This allowed us to immidiately interpret negative and positive coefficient signs as protective and risk factors, respectively and the magnitude as the impact of these factors.
@@
 We also compared the top 20 OTUs selected by permutation importance in L2-regularized logistic regression [Figure S7] and the highest performing tree-based models, random forest and XGBoost [Figure 4].
-2 and 5 OTUs, respectively, were shared among the models.
+Two and five OTUs, respectively, were shared among the models.
 These results indicate that we were able to identify important OTUs that are shared across the highest performing linear and non-linear models when we use permutation importance as our interpretation method.
@@
 We then evaluated the difference in relative abundances of the top 20 OTUs identified in L2-regularized logistic regression and random forest models between healthy patients and patients with SRNs [Figure S8].
-There were minimal differences in relative abundances across OTUs between different diagnoses.
-This supports our claim that it is not possible to differentiate disease vs healthy states by focusing on individual taxa.
+There were minimal differences in the median relative abundances across OTUs between different diagnoses.
+This supports our claim that it is not possible to differentiate disease versus healthy states by focusing on individual taxa.
+The ability for ML models to simultaneously consider the relative abundances of multiple OTUs and their context dependency is a great advantage over traditional statistical approaches that consider each OTU in isolation.
 
@@
 While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes.
-In terms of interpretability, random forest is a non-linear ML model, while L2-regularized logistic regression, a linear model, is easily interpreted according to the feature weights.
+In terms of interpretability, random forest is a non-linear ML model, while L2-regularized logistic regression, a linear model, was more easily interpreted because we could use the feature weights.
 Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.
@@
 We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features.
-We are choosing to leave the decision to define correlation and setting up correlations thresholds to the researchers for their own microbiome questions.
-However, we would like to urge them to investigate the correlation structures of the data to avoid misinterpreting the models.
-This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome.
-True causal mechanisms (e.g., role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020].
+By our approach, we identified 432 OTUs out of 6,920 that had perfect correlations with at least one other OTU.
+The decision to establish correlation thresholds is left to researchers to implement for their own analyses.
+Regardless of the threshold, undestanding the correlation structures within the data is critical to avoid misinterpreting the models.
+Such structures are likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome.
+Finally, true causal mechanisms (e.g., role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020].
 To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.
@@
 
-Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant.
+Our framework provides a reproducible structure for investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant.
 However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking [@wiens_no_2019].
@@
 Generalization and classification performance of ML models using AUROC values of all cross-validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by a horizontal line at 0.50) for all the ML models.
-The predictive performance of random forest model was higher than other ML models, though not significantly (p > 0.01).
+The predictive performance of random forest model was higher than other ML models, though not significantly (p > 0.05).
 L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another.

### 2020-01-28 submission/manuscript.tex
Pat's edits

@@
 forest model performed best at detecting SRNs with an AUROC of 0.695
-{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
+{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not inherently
 interpretable. Despite its simplicity, L2-regularized logistic
@@
 
-Prediction of diseases using machine learning (ML) is rapidly being
-adopted in microbiome studies. However, the estimated performance
-associated with these ML models is likely over-optimistic. Moreover,
-there is a trend towards using black box models without a discussion of
-the difficulty of interpreting such models when trying to identify
-microbial biomarkers of disease. This work represents a step towards
-developing more reproducible ML practices in applying ML to microbiome
-research. We implement a rigorous pipeline and emphasize the importance
-of selecting ML models that reflect the goal of the study. These
-concepts are not particular to the study of human health but can also be
-applied to environmental microbiology studies.
+Diagnosing diseases using machine learning (ML) is rapidly being adopted
+in microbiome studies. However, the estimated performance associated
+with these models is likely over-optimistic. Moreover, there is a trend
+towards using black box models without a discussion of the difficulty of
+interpreting such models when trying to identify microbial biomarkers of
+disease. This work represents a step towards developing more
+reproducible ML practices in applying ML to microbiome research. We
+implement a rigorous pipeline and emphasize the importance of selecting
+ML models that reflect the goal of the study. These concepts are not
+particular to the study of human health but can also be applied to
+environmental microbiology studies.
 
@@
 there is an increasing desire to use microbiome data to diagnose
-disease. However, the structure of the human microbiome is remarkably
+diseases. However, the structure of the human microbiome is remarkably
 variable among individuals to the point where it is often difficult to
@@
 using traditional statistical models. For example it is not possible to
-classify individuals as having healthy colons or colorectal lesions
-using Bray-Curtis distances based on the 16S rRNA gene sequences
-collected from fecal samples {[}Figure S1{]}. This variation is likely
-due to the ability of many bacterial populations to fill the same niche
-such that different populations cause the same disease in different
-individuals. Furthermore, a growing number of studies have shown that it
-is rare for a single bacterial species to be associated with a disease.
-Instead, subsets of the microbiome account for differences in health.
-Traditional statistical approaches do not adequately account for the
-variation in the human microbiome and typically consider the protective
-or risk effects of each bacterial population separately (1). Recently,
-machine learning (ML) models have grown in popularity among microbiome
-researchers as our ability to sample large numbers of individuals has
-grown; such models can effectively account for the interpersonal
-microbiome variation and the ecology of disease.
+classify individuals as having healthy colons or screen relevant
+neoplasia using Bray-Curtis distances based on the 16S rRNA gene
+sequences collected from fecal samples {[}Figure S1{]}. This variation
+is likely due to the ability of many bacterial populations to fill the
+same niche such that different populations cause the same disease in
+different individuals. Furthermore, a growing number of studies have
+shown that it is rare for a single bacterial species to be associated
+with a disease. Instead, subsets of the microbiome account for
+differences in health. Traditional statistical approaches do not
+adequately account for the variation in the human microbiome and
+typically consider the protective or risk effects of each bacterial
+population separately (1). Recently, machine learning (ML) models have
+grown in popularity among microbiome researchers as our ability to
+sample large numbers of individuals has grown; such models can
+effectively account for the interpersonal microbiome variation and the
+ecology of disease because they consider the relative abundance of each
+bacterial population in the context of others rather than in isolation.
 
@@
 However, there are common methodological and reporting problems that
-arise when applying ML to such data, that need to be addressed for the
+arise when applying ML to such data that need to be addressed for the
 field to progress. These problems include a lack of transparency in
@@
 performed an empirical analysis comparing the predictive performance,
-interpretability and training times of seven modeling approaches with
-the same dataset and pipeline. We built three linear models with
-different forms of regularization: L2-regularized logistic regression
-and L1 and L2-regularized support vector machines (SVM) with a linear
-kernel. We also trained four non-linear models: SVM with radial basis
-function kernel, a decision tree, random forest, and gradient boosted
-trees. We compared their predictive performance, interpretability, and
-training time. To demonstrate the performance of these modeling
-approaches and our pipeline on a case study, we used a data from a
-previously published study that sought to classify individuals as having
-healthy colons or colonic lesions based on the 16S rRNA gene sequences
-collected from fecal samples (4). This dataset was selected because it
-is a relatively large collection of individuals (N=490) connected to a
-clinically significant disease where there is ample evidence that the
-disease is driven by variation in the microbiome (2, 4, 5, 32). With
-this dataset, we developed an ML pipeline that can be used in many
-different scenarios for training and evaluating models. This framework
-can be easily applied to other host-associated and environmental
-microbiome datasets. We also provided an aspirational rubric for
-evaluating the rigor of ML practices applied to microbiome data {[}Table
-S1{]} to urge the audience to be diligent with with their study design,
-model selection, development, evaluation and interpretation.
+interpretability, data requirements, and training times of seven
+modeling approaches with the same dataset and pipeline. We built three
+linear models with different forms of regularization: L2-regularized
+logistic regression and L1 and L2-regularized support vector machines
+(SVM) with a linear kernel. We also trained four non-linear models: SVM
+with radial basis function kernel, a decision tree, random forest, and
+gradient boosted trees. We compared their predictive performance,
+interpretability, and training time. To demonstrate the performance of
+these modeling approaches and our pipeline, we present a case study
+using data from a previously published study that sought to classify
+individuals as having healthy colons or colonic lesions based on the 16S
+rRNA gene sequences collected from fecal samples (4). This dataset was
+selected because it is a relatively large collection of individuals
+(N=490) connected to a clinically significant disease where there is
+ample evidence that the disease is driven by variation in the microbiome
+(2, 4, 5, 32). With this dataset, we developed an ML pipeline that can
+be used in many different scenarios for training and evaluating models.
+This framework can be easily applied to other host-associated and
+environmental microbiome datasets. We also provided an aspirational
+rubric for evaluating the rigor of ML practices applied to microbiome
+data {[}Table S1{]} to urge the audience to be diligent in their study
+design and model selection, development, evaluation, and interpretation.
 
@@
 highest-ranked OTUs were shared across these three models (e.g., OTUs
-50, 426, 609, 822, 1239). The great benefit of this approach was knowing
-the sign and magnitude of each OTU coefficient in the trained model.
-This allowed us to immidiately interpret negative and positive
-coefficient signs as protective and risk factors, respectively and the
-magnitude as the impact of these factors. However, this approach is
-limited to linear models or models with prespecified interaction terms.
+50, 426, 609, 822, 1239). The benefit of this approach was knowing the
+sign and magnitude of each OTU coefficient in the trained model. This
+allowed us to immidiately interpret negative and positive coefficient
+signs as protective and risk factors, respectively and the magnitude as
+the impact of these factors. However, this approach is limited to linear
+models or models with prespecified interaction terms.
 
@@
 performing tree-based models, random forest and XGBoost {[}Figure 4{]}.
-2 and 5 OTUs, respectively, were shared among the models. These results
-indicate that we were able to identify important OTUs that are shared
-across the highest performing linear and non-linear models when we use
-permutation importance as our interpretation method.
+Two and five OTUs, respectively, were shared among the models. These
+results indicate that we were able to identify important OTUs that are
+shared across the highest performing linear and non-linear models when
+we use permutation importance as our interpretation method.
 
@@
 models between healthy patients and patients with SRNs {[}Figure S8{]}.
-There were minimal differences in relative abundances across OTUs
-between different diagnoses. This supports our claim that it is not
-possible to differentiate disease vs healthy states by focusing on
-individual taxa.
+There were minimal differences in the median relative abundances across
+OTUs between different diagnoses. This supports our claim that it is not
+possible to differentiate disease versus healthy states by focusing on
+individual taxa. The ability for ML models to simultaneously consider
+the relative abundances of multiple OTUs and their context dependency is
+a great advantage over traditional statistical approaches that consider
+each OTU in isolation.
 
@@
 In terms of interpretability, random forest is a non-linear ML model,
-while L2-regularized logistic regression, a linear model, is easily
-interpreted according to the feature weights. Comparing many different
-models showed us that the most complex model was not necessarily the
-best model for our ML task.
+while L2-regularized logistic regression, a linear model, was more
+easily interpreted because we could use the feature weights. Comparing
+many different models showed us that the most complex model was not
+necessarily the best model for our ML task.
 
@@
 the correlation threshold to further investigate the relationships among
-correlated features. We are choosing to leave the decision to define
-correlation and setting up correlations thresholds to the researchers
-for their own microbiome questions. However, we would like to urge them
-to investigate the correlation structures of the data to avoid
-misinterpreting the models. This is likely to be a particular problem
-with shotgun metagenomic datasets where collinearity will be more
-pronounced due to many genes being correlated with one another because
-they come from the same chromosome. True causal mechanisms (e.g., role
-of microbiome in colorectal cancer) cannot be explained solely by the
-highest performing machine learning model (38). To identify the true
-underlying microbial factors of a disease, it is crucial to follow up on
-any correlation analyses with further hypothesis testing and
-experimentation for biological validation.
+correlated features. By our approach, we identified 432 OTUs out of
+6,920 that had perfect correlations with at least one other OTU. The
+decision to establish correlation thresholds is left to researchers to
+implement for their own analyses. Regardless of the threshold,
+undestanding the correlation structures within the data is critical to
+avoid misinterpreting the models. Such structures are likely to be a
+particular problem with shotgun metagenomic datasets where collinearity
+will be more pronounced due to many genes being correlated with one
+another because they come from the same chromosome. Finally, true causal
+mechanisms (e.g., role of microbiome in colorectal cancer) cannot be
+explained solely by the highest performing machine learning model (38).
+To identify the true underlying microbial factors of a disease, it is
+crucial to follow up on any correlation analyses with further hypothesis
+testing and experimentation for biological validation.
 
@@
 
-Our framework provides a reproducible structure to investigators wanting
-to train, evaluate, and interpret their own ML models to generate
-hypotheses regarding which OTUs might be biologically relevant. However,
-deploying microbiome-based models to make clinical diagnoses or
+Our framework provides a reproducible structure for investigators
+wanting to train, evaluate, and interpret their own ML models to
+generate hypotheses regarding which OTUs might be biologically relevant.
+However, deploying microbiome-based models to make clinical diagnoses or
 predictions is a significantly more challenging and distinct undertaking
@@
 performance of random forest model was higher than other ML models,
-though not significantly (p \textgreater{} 0.01). L2-regularized
+though not significantly (p \textgreater{} 0.05). L2-regularized
 logistic regression, XGBoost, L2-regularized SVM with linear and radial

### 2020-01-28 submission/manuscript.Rmd
begum's edits

@@
 In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance.
+Previous studies have emphasized the difficulty of interpreting the feature coefficients in linear models [@breiman_statistical_2001] and the biases introduced by computing feature importance using built-in methods (e.g., gini drop) of tree-based models [@strobl_bias_2007].
+Therefore, we encourage our audience to use both interpretation methods highlighted in this study.
 Human-associated microbial communities have complex correlation structures that create collinearity in the datasets.

### 2020-01-28 submission/manuscript.tex
begum's edits

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  11pt,
-]{article}
+\documentclass[11pt,]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
   \fi
-% Use upquote if available, for straight quotes in verbatim environments
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
+% use microtype if available
+\IfFileExists{microtype.sty}{%
 \usepackage[]{microtype}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
 \hypersetup{
             pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[margin=1in]{geometry}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
 \makeatletter
@@
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+
 \usepackage{booktabs}
@@
 \newpage
+
 \linenumbers
 
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
+\subsection{Abstract}\label{abstract}
 
@@
 
-\hypertarget{importance}{%
-\subsection{Importance}\label{importance}}
+\subsection{Importance}\label{importance}
 
@@
 
-\hypertarget{background}{%
-\subsection{Background}\label{background}}
+\subsection{Background}\label{background}
 
@@
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+\subsection{Results}\label{results}
 
@@
 
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+\subsection{Discussion}\label{discussion}
 
@@
 (i) their absolute weights in the trained models and (ii) their impact
-on the predictive performance based on permutation importance.
-Human-associated microbial communities have complex correlation
-structures that create collinearity in the datasets. This can hinder our
-ability to reliably interpret models because the feature weights of
-correlated OTUs are influenced by one another (37). To capture all
-important features, once we identify highly ranked OTUs, we should
-review their relationships with other OTUs. These relationships will
-help us generate new hypotheses about the ecology of the disease and
-test them with follow-up experiments. When we used permutation
-importance, we partially accounted for collinearity by grouping
-correlated OTUs to determine their impact as a group. We grouped OTUs
-that had a perfect correlation with each other; however, we could reduce
-the correlation threshold to further investigate the relationships among
-correlated features. By our approach, we identified 432 OTUs out of
-6,920 that had perfect correlations with at least one other OTU. The
-decision to establish correlation thresholds is left to researchers to
-implement for their own analyses. Regardless of the threshold,
-undestanding the correlation structures within the data is critical to
-avoid misinterpreting the models. Such structures are likely to be a
-particular problem with shotgun metagenomic datasets where collinearity
-will be more pronounced due to many genes being correlated with one
-another because they come from the same chromosome. Finally, true causal
-mechanisms (e.g., role of microbiome in colorectal cancer) cannot be
-explained solely by the highest performing machine learning model (38).
-To identify the true underlying microbial factors of a disease, it is
-crucial to follow up on any correlation analyses with further hypothesis
-testing and experimentation for biological validation.
+on the predictive performance based on permutation importance. Previous
+studies have emphasized the difficulty of interpreting the feature
+coefficients in linear models (37) and the biases introduced by
+computing feature importance using built-in methods (e.g., gini drop) of
+tree-based models (38). Therefore, we encourage our audience to use both
+interpretation methods highlighted in this study. Human-associated
+microbial communities have complex correlation structures that create
+collinearity in the datasets. This can hinder our ability to reliably
+interpret models because the feature weights of correlated OTUs are
+influenced by one another (39). To capture all important features, once
+we identify highly ranked OTUs, we should review their relationships
+with other OTUs. These relationships will help us generate new
+hypotheses about the ecology of the disease and test them with follow-up
+experiments. When we used permutation importance, we partially accounted
+for collinearity by grouping correlated OTUs to determine their impact
+as a group. We grouped OTUs that had a perfect correlation with each
+other; however, we could reduce the correlation threshold to further
+investigate the relationships among correlated features. By our
+approach, we identified 432 OTUs out of 6,920 that had perfect
+correlations with at least one other OTU. The decision to establish
+correlation thresholds is left to researchers to implement for their own
+analyses. Regardless of the threshold, undestanding the correlation
+structures within the data is critical to avoid misinterpreting the
+models. Such structures are likely to be a particular problem with
+shotgun metagenomic datasets where collinearity will be more pronounced
+due to many genes being correlated with one another because they come
+from the same chromosome. Finally, true causal mechanisms (e.g., role of
+microbiome in colorectal cancer) cannot be explained solely by the
+highest performing machine learning model (40). To identify the true
+underlying microbial factors of a disease, it is crucial to follow up on
+any correlation analyses with further hypothesis testing and
+experimentation for biological validation.
 
@@
 predict fecal short-chain fatty acid concentrations based on microbiome
-data (39). Our analysis was also limited to shallow learning methods and
+data (41). Our analysis was also limited to shallow learning methods and
 did not explore deep learning methods such as neural networks. Deep
-learning methods hold promise (12, 40, 41), but microbiome datasets
+learning methods hold promise (12, 42, 43), but microbiome datasets
 often suffer from having many features and small sample sizes, which
@@
 predictions is a significantly more challenging and distinct undertaking
-(42). For example, we currently lack standardized methods to collect
+(44). For example, we currently lack standardized methods to collect
 patient samples, generate sequence data, and report clinical data. We
@@
 considerable memory, and result in different OTU assignments as new data
-are added (43). Finally, we also need independent validation cohorts to
+are added (45). Finally, we also need independent validation cohorts to
 test the performance of a diagnostic model. To realize the potential for
@@
 
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
+\subsection{Materials and Methods}\label{materials-and-methods}
 
@@
 whereas the SRN class included patients with advanced adenomas or
-carcinomas (44). Patients with an adenoma greater than 1 cm, more than
+carcinomas (46). Patients with an adenoma greater than 1 cm, more than
 three adenomas of any size, or an adenoma with villous histology were
-classified as having advanced adenomas (44). There were 172 patients
+classified as having advanced adenomas (46). There were 172 patients
 with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
@@
 population abundances. The sequence data used in our analyses were
-originally generated by Baxter et al.~(available through NCBI Sequence
+originally generated by Baxter et al. (available through NCBI Sequence
 Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
-generated by Sze et al (45), who processed the 16S rRNA sequences in
+generated by Sze et al (47), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
+to OTUs at 97\% similarity using the OptiClust algorithm (45, 48, 49);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 random search or recently developed tools such as Hyperband to identify
-good hyperparameter configurations (48).
+good hyperparameter configurations (50).
 
@@
 
-\hypertarget{references}{%
-\subsection{References}\label{references}}
+\subsection{References}\label{references}
 
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-segata_metagenomic_2011}{}%
+\hypertarget{ref-segata_metagenomic_2011}{}
 1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
@@
 
-\leavevmode\hypertarget{ref-zeller_potential_2014}{}%
+\hypertarget{ref-zeller_potential_2014}{}
 2. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
@@
 
-\leavevmode\hypertarget{ref-zackular_human_2014}{}%
+\hypertarget{ref-zackular_human_2014}{}
 3. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
@@
 
-\leavevmode\hypertarget{ref-baxter_dna_2016}{}%
+\hypertarget{ref-baxter_dna_2016}{}
 4. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
@@
 
-\leavevmode\hypertarget{ref-baxter_microbiota-based_2016}{}%
+\hypertarget{ref-baxter_microbiota-based_2016}{}
 5. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
@@
 
-\leavevmode\hypertarget{ref-hale_shifts_2017}{}%
+\hypertarget{ref-hale_shifts_2017}{}
 6. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
@@
 
-\leavevmode\hypertarget{ref-pasolli_machine_2016}{}%
+\hypertarget{ref-pasolli_machine_2016}{}
 7. \textbf{Pasolli E}, \textbf{Truong DT}, \textbf{Malik F},
@@
 
-\leavevmode\hypertarget{ref-sze_looking_2016}{}%
+\hypertarget{ref-sze_looking_2016}{}
 8. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
@@
 
-\leavevmode\hypertarget{ref-walters_meta-analyses_2014}{}%
+\hypertarget{ref-walters_meta-analyses_2014}{}
 9. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
@@
 
-\leavevmode\hypertarget{ref-vazquez-baeza_guiding_2018}{}%
+\hypertarget{ref-vazquez-baeza_guiding_2018}{}
 10. \textbf{Vázquez-Baeza Y}, \textbf{Gonzalez A}, \textbf{Xu ZZ},
@@
 
-\leavevmode\hypertarget{ref-qin_alterations_2014}{}%
+\hypertarget{ref-qin_alterations_2014}{}
 11. \textbf{Qin N}, \textbf{Yang F}, \textbf{Li A}, \textbf{Prifti E},
@@
 
-\leavevmode\hypertarget{ref-geman_deep_2018}{}%
+\hypertarget{ref-geman_deep_2018}{}
 12. \textbf{Geman O}, \textbf{Chiuchisan I}, \textbf{Covasa M},
@@
 
-\leavevmode\hypertarget{ref-thaiss_persistent_2016}{}%
+\hypertarget{ref-thaiss_persistent_2016}{}
 13. \textbf{Thaiss CA}, \textbf{Itav S}, \textbf{Rothschild D},
@@
 
-\leavevmode\hypertarget{ref-dadkhah_gut_2019}{}%
+\hypertarget{ref-dadkhah_gut_2019}{}
 14. \textbf{Dadkhah E}, \textbf{Sikaroodi M}, \textbf{Korman L},
@@
 
-\leavevmode\hypertarget{ref-flemer_oral_2018}{}%
+\hypertarget{ref-flemer_oral_2018}{}
 15. \textbf{Flemer B}, \textbf{Warren RD}, \textbf{Barrett MP},
@@
 
-\leavevmode\hypertarget{ref-montassier_pretreatment_2016}{}%
+\hypertarget{ref-montassier_pretreatment_2016}{}
 16. \textbf{Montassier E}, \textbf{Al-Ghalith GA}, \textbf{Ward T},
@@
 
-\leavevmode\hypertarget{ref-ai_systematic_2017}{}%
+\hypertarget{ref-ai_systematic_2017}{}
 17. \textbf{Ai L}, \textbf{Tian H}, \textbf{Chen Z}, \textbf{Chen H},
@@
 
-\leavevmode\hypertarget{ref-dai_multi-cohort_2018}{}%
+\hypertarget{ref-dai_multi-cohort_2018}{}
 18. \textbf{Dai Z}, \textbf{Coker OO}, \textbf{Nakatsu G}, \textbf{Wu
@@
 
-\leavevmode\hypertarget{ref-mossotto_classification_2017}{}%
+\hypertarget{ref-mossotto_classification_2017}{}
 19. \textbf{Mossotto E}, \textbf{Ashton JJ}, \textbf{Coelho T},
@@
 
-\leavevmode\hypertarget{ref-wong_quantitation_2017}{}%
+\hypertarget{ref-wong_quantitation_2017}{}
 20. \textbf{Wong SH}, \textbf{Kwong TNY}, \textbf{Chow T-C}, \textbf{Luk
@@
 
-\leavevmode\hypertarget{ref-statnikov_comprehensive_2013}{}%
+\hypertarget{ref-statnikov_comprehensive_2013}{}
 21. \textbf{Statnikov A}, \textbf{Henaff M}, \textbf{Narendra V},
@@
 
-\leavevmode\hypertarget{ref-knights_supervised_2011}{}%
+\hypertarget{ref-knights_supervised_2011}{}
 22. \textbf{Knights D}, \textbf{Costello EK}, \textbf{Knight R}. 2011.
@@
 
-\leavevmode\hypertarget{ref-wirbel_meta-analysis_2019}{}%
+\hypertarget{ref-wirbel_meta-analysis_2019}{}
 23. \textbf{Wirbel J}, \textbf{Pyl PT}, \textbf{Kartal E}, \textbf{Zych
@@
 
-\leavevmode\hypertarget{ref-vangay_microbiome_2019}{}%
+\hypertarget{ref-vangay_microbiome_2019}{}
 24. \textbf{Vangay P}, \textbf{Hillmann BM}, \textbf{Knights D}. 2019.
@@
 
-\leavevmode\hypertarget{ref-galkin_human_2018}{}%
+\hypertarget{ref-galkin_human_2018}{}
 25. \textbf{Galkin F}, \textbf{Aliper A}, \textbf{Putin E},
@@
 
-\leavevmode\hypertarget{ref-reiman_using_2017}{}%
+\hypertarget{ref-reiman_using_2017}{}
 26. \textbf{Reiman D}, \textbf{Metwally A}, \textbf{Dai Y}. 2017. Using
@@
 
-\leavevmode\hypertarget{ref-fioravanti_phylogenetic_2017}{}%
+\hypertarget{ref-fioravanti_phylogenetic_2017}{}
 27. \textbf{Fioravanti D}, \textbf{Giarratano Y}, \textbf{Maggio V},
@@
 
-\leavevmode\hypertarget{ref-thomas_metagenomic_2019}{}%
+\hypertarget{ref-thomas_metagenomic_2019}{}
 28. \textbf{Thomas AM}, \textbf{Manghi P}, \textbf{Asnicar F},
@@
 
-\leavevmode\hypertarget{ref-rudin_please_2018}{}%
+\hypertarget{ref-rudin_please_2018}{}
 29. \textbf{Rudin C}. 2018. Please stop explaining black box models for
@@
 
-\leavevmode\hypertarget{ref-rudin_optimized_2018}{}%
+\hypertarget{ref-rudin_optimized_2018}{}
 30. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
@@
 
-\leavevmode\hypertarget{ref-Quinn847475}{}%
+\hypertarget{ref-Quinn847475}{}
 31. \textbf{Quinn TP}, \textbf{Erb I}. 2019. Another look at
@@
 
-\leavevmode\hypertarget{ref-knights_human-associated_2011}{}%
+\hypertarget{ref-knights_human-associated_2011}{}
 32. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
@@
 
-\leavevmode\hypertarget{ref-miller_explanation_2017}{}%
+\hypertarget{ref-miller_explanation_2017}{}
 33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
@@
 
-\leavevmode\hypertarget{ref-ribeiro_why_2016}{}%
+\hypertarget{ref-ribeiro_why_2016}{}
 34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
-"Why should i trust you?": Explaining the predictions of any classifier.
-arXiv:160204938 {[}cs, stat{]}.
-
-\leavevmode\hypertarget{ref-nori_interpretml:_2019}{}%
+``Why should i trust you?'': Explaining the predictions of any
+classifier. arXiv:160204938 {[}cs, stat{]}.
+
+\hypertarget{ref-nori_interpretml:_2019}{}
 35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
@@
 
-\leavevmode\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}%
+\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
 36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
@@
 
-\leavevmode\hypertarget{ref-dormann_collinearity:_2013}{}%
-37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+\hypertarget{ref-breiman_statistical_2001}{}
+37. \textbf{Breiman L}. 2001. Statistical modeling: The two cultures
+(with comments and a rejoinder by the author). Statist Sci
+\textbf{16}:199--231.
+doi:\href{https://doi.org/10.1214/ss/1009213726}{10.1214/ss/1009213726}.
+
+\hypertarget{ref-strobl_bias_2007}{}
+38. \textbf{Strobl C}, \textbf{Boulesteix A-L}, \textbf{Zeileis A},
+\textbf{Hothorn T}. 2007. Bias in random forest variable importance
+measures: Illustrations, sources and a solution. BMC Bioinformatics
+\textbf{8}:25.
+doi:\href{https://doi.org/10.1186/1471-2105-8-25}{10.1186/1471-2105-8-25}.
+
+\hypertarget{ref-dormann_collinearity:_2013}{}
+39. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 
-\leavevmode\hypertarget{ref-li_accurate_2020}{}%
-38. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
+\hypertarget{ref-li_accurate_2020}{}
+40. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
 Accurate data-driven prediction does not mean high reproducibility. Nat
@@
 
-\leavevmode\hypertarget{ref-sze_fecal_2019}{}%
-39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+\hypertarget{ref-sze_fecal_2019}{}
+41. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
@@
 
-\leavevmode\hypertarget{ref-kocheturov_massive_2019}{}%
-40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+\hypertarget{ref-kocheturov_massive_2019}{}
+42. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 
-\leavevmode\hypertarget{ref-kim_improved_2018}{}%
-41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+\hypertarget{ref-kim_improved_2018}{}
+43. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 
-\leavevmode\hypertarget{ref-wiens_no_2019}{}%
-42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
+\hypertarget{ref-wiens_no_2019}{}
+44. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
 \textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
@@
 
-\leavevmode\hypertarget{ref-westcott_opticlust_2017}{}%
-43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+\hypertarget{ref-westcott_opticlust_2017}{}
+45. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 
-\leavevmode\hypertarget{ref-redwood_stool_2016}{}%
-44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+\hypertarget{ref-redwood_stool_2016}{}
+46. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
 \textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
@@
 
-\leavevmode\hypertarget{ref-sze_leveraging_2018}{}%
-45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+\hypertarget{ref-sze_leveraging_2018}{}
+47. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 
-\leavevmode\hypertarget{ref-schloss_introducing_2009}{}%
-46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\hypertarget{ref-schloss_introducing_2009}{}
+48. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 
-\leavevmode\hypertarget{ref-rognes_vsearch_2016}{}%
-47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+\hypertarget{ref-rognes_vsearch_2016}{}
+49. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 
-\leavevmode\hypertarget{ref-li_hyperband:_2016}{}%
-48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+\hypertarget{ref-li_hyperband:_2016}{}
+50. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
@@
 \small
+
 \begin{tabular}{|l|l|l|}
@@
 
-\end{tabular}
-\newpage
+\end{tabular}\newpage
 

### 2020-01-28 submission/manuscript.tex
Rerender

--- 2020-01-28 submission/manuscript.tex
@@
 Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
-generated by Sze et al (45), who processed the 16S rRNA sequences in
+generated by Sze et al (47), who processed the 16S rRNA sequences in
 mothur (v1.39.3) using the default quality filtering methods,
 identifying and removing chimeric sequences using VSEARCH, and assigning
-to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
+to OTUs at 97\% similarity using the OptiClust algorithm (45, 48, 49);
 (\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
@@
 
+\leavevmode\hypertarget{ref-breiman_statistical_2001}{}%
+37. \textbf{Breiman L}. 2001. Statistical modeling: The two cultures
+(with comments and a rejoinder by the author). Statist Sci
+\textbf{16}:199--231.
+doi:\href{https://doi.org/10.1214/ss/1009213726}{10.1214/ss/1009213726}.
+
+\leavevmode\hypertarget{ref-strobl_bias_2007}{}%
+38. \textbf{Strobl C}, \textbf{Boulesteix A-L}, \textbf{Zeileis A},
+\textbf{Hothorn T}. 2007. Bias in random forest variable importance
+measures: Illustrations, sources and a solution. BMC Bioinformatics
+\textbf{8}:25.
+doi:\href{https://doi.org/10.1186/1471-2105-8-25}{10.1186/1471-2105-8-25}.
+
 \leavevmode\hypertarget{ref-dormann_collinearity:_2013}{}%
-37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
+39. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
 \textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
@@
 \leavevmode\hypertarget{ref-li_accurate_2020}{}%
-38. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
+40. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
 Accurate data-driven prediction does not mean high reproducibility. Nat
@@
 \leavevmode\hypertarget{ref-sze_fecal_2019}{}%
-39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+41. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
 \textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
@@
 \leavevmode\hypertarget{ref-kocheturov_massive_2019}{}%
-40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
+42. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
 2019. Massive datasets and machine learning for computational
@@
 \leavevmode\hypertarget{ref-kim_improved_2018}{}%
-41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
+43. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
 method for prediction of cancer prognosis by network learning. Genes
@@
 \leavevmode\hypertarget{ref-wiens_no_2019}{}%
-42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
+44. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
 \textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
@@
 \leavevmode\hypertarget{ref-westcott_opticlust_2017}{}%
-43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+45. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
 Improved Method for Assigning Amplicon-Based Sequence Data to
@@
 \leavevmode\hypertarget{ref-redwood_stool_2016}{}%
-44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
+46. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
 \textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
@@
 \leavevmode\hypertarget{ref-sze_leveraging_2018}{}%
-45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+47. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
 rRNA gene surveys to identify reproducible biomarkers in individuals
@@
 \leavevmode\hypertarget{ref-schloss_introducing_2009}{}%
-46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+48. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \leavevmode\hypertarget{ref-rognes_vsearch_2016}{}%
-47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+49. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
@@
 \leavevmode\hypertarget{ref-li_hyperband:_2016}{}%
-48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
+50. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
 \textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel

### 2020-02-17 submission/manuscript.Rmd
added Jenna's comments

@@
 Previous studies have emphasized the difficulty of interpreting the feature coefficients in linear models [@breiman_statistical_2001] and the biases introduced by computing feature importance using built-in methods (e.g., gini drop) of tree-based models [@strobl_bias_2007].
-Therefore, we encourage our audience to use both interpretation methods highlighted in this study.
+Therefore, we encourage our audience to use both interpretation methods highlighted in this study as permutation importance is a model-agnostic tool that can be used to compared feature importance across different models.
 Human-associated microbial communities have complex correlation structures that create collinearity in the datasets.
@@
 We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1].
-We have highlighted the trade-offs between model complexity and interpretability, the need for tuning hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
+We highlighted the trade-offs between model complexity and interpretability, the need for tuning hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation.
+We showed the importance of interpretability for generating hypotheses to identify causal, biological relationships and for identifying inconsistencies in model setup.
 Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.

### 2020-02-17 submission/manuscript.tex
added Jenna's comments

--- 2020-01-28 submission/manuscript.tex
@@
 tree-based models (38). Therefore, we encourage our audience to use both
-interpretation methods highlighted in this study. Human-associated
-microbial communities have complex correlation structures that create
-collinearity in the datasets. This can hinder our ability to reliably
-interpret models because the feature weights of correlated OTUs are
-influenced by one another (39). To capture all important features, once
-we identify highly ranked OTUs, we should review their relationships
-with other OTUs. These relationships will help us generate new
-hypotheses about the ecology of the disease and test them with follow-up
-experiments. When we used permutation importance, we partially accounted
-for collinearity by grouping correlated OTUs to determine their impact
-as a group. We grouped OTUs that had a perfect correlation with each
-other; however, we could reduce the correlation threshold to further
-investigate the relationships among correlated features. By our
-approach, we identified 432 OTUs out of 6,920 that had perfect
-correlations with at least one other OTU. The decision to establish
-correlation thresholds is left to researchers to implement for their own
-analyses. Regardless of the threshold, undestanding the correlation
-structures within the data is critical to avoid misinterpreting the
-models. Such structures are likely to be a particular problem with
-shotgun metagenomic datasets where collinearity will be more pronounced
-due to many genes being correlated with one another because they come
-from the same chromosome. Finally, true causal mechanisms (e.g., role of
-microbiome in colorectal cancer) cannot be explained solely by the
-highest performing machine learning model (40). To identify the true
-underlying microbial factors of a disease, it is crucial to follow up on
-any correlation analyses with further hypothesis testing and
-experimentation for biological validation.
+interpretation methods highlighted in this study as permutation
+importance is a model-agnostic tool that can be used to compared feature
+importance across different models. Human-associated microbial
+communities have complex correlation structures that create collinearity
+in the datasets. This can hinder our ability to reliably interpret
+models because the feature weights of correlated OTUs are influenced by
+one another (39). To capture all important features, once we identify
+highly ranked OTUs, we should review their relationships with other
+OTUs. These relationships will help us generate new hypotheses about the
+ecology of the disease and test them with follow-up experiments. When we
+used permutation importance, we partially accounted for collinearity by
+grouping correlated OTUs to determine their impact as a group. We
+grouped OTUs that had a perfect correlation with each other; however, we
+could reduce the correlation threshold to further investigate the
+relationships among correlated features. By our approach, we identified
+432 OTUs out of 6,920 that had perfect correlations with at least one
+other OTU. The decision to establish correlation thresholds is left to
+researchers to implement for their own analyses. Regardless of the
+threshold, undestanding the correlation structures within the data is
+critical to avoid misinterpreting the models. Such structures are likely
+to be a particular problem with shotgun metagenomic datasets where
+collinearity will be more pronounced due to many genes being correlated
+with one another because they come from the same chromosome. Finally,
+true causal mechanisms (e.g., role of microbiome in colorectal cancer)
+cannot be explained solely by the highest performing machine learning
+model (40). To identify the true underlying microbial factors of a
+disease, it is crucial to follow up on any correlation analyses with
+further hypothesis testing and experimentation for biological
+validation.
 
@@
 using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
-have highlighted the trade-offs between model complexity and
+highlighted the trade-offs between model complexity and
 interpretability, the need for tuning hyperparameters, the utility of
@@
 importance of considering correlation structures in datasets for
-reliable interpretation. Furthermore, we underscored the importance of
-proper experimental design and methods to help us achieve the level of
-validity and accountability we want from models built for patient
-health.
+reliable interpretation. We showed the importance of interpretability
+for generating hypotheses to identify causal, biological relationships
+and for identifying inconsistencies in model setup. Furthermore, we
+underscored the importance of proper experimental design and methods to
+help us achieve the level of validity and accountability we want from
+models built for patient health.
 

### 2020-02-18 submission/manuscript.tex
Pat's edits

--- 2020-01-28 submission/manuscript.tex
@@
 Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Mack
-Ruffin\({^3}\), Jenna Wiens\({^2}\), Patrick D.
+Ruffin\({^3}\), Jenna Wiens\textsuperscript{2\(\dagger\)}, Patrick D.
 Schloss\textsuperscript{1\(\dagger\)}

### 2020-02-21 submission/MicrobeConference/Microbe_Abstract2019.Rmd
final edits before resubmission

--- 2018-11-29 submission/Microbe_Abstract2019.Rmd
@@
 
-As gut microbiome field continues to grow, there is an ever-increasing demand for reproducible machine learning methods to determine associations between the microbiome and a phenotype of interest.
-Currently, the use of machine learning in microbiome research lacks clarity and consistency over the modeling pipeline (training, validation and testing steps).
+As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and phenotypes.
+Currently, the field’s use of machine learning lacks clarity and consistency.
 There is a need for guidance on how to implement good machine learning practices to generate reproducible and robust models.
 
-Recently, there has been growing interest in using machine learning to identify colorectal tumors with microbiota-associated biomarkers.
+One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers.
 Colorectal cancer is one of the leading cause of death among cancers in the United States.
@@
 
-In this study, hemoglobin counts and 16S gene sequences in the stool were used to identify colorectal tumors of `r paste(nrow(meta))` patients as advanced tumors or not.
+We used the fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples to classify `r paste(nrow(meta))` patients as having advanced tumors or not.
 Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models.

### 2020-02-21 submission/MicrobeConference/Microbe_Abstract2019.tex
final edits before resubmission

--- 2018-11-28 submission/Microbe_Abstract2019.tex
@@
 \linenumbers
-\textbf{Evaluation of classification pipelines that predict colorectal
-cancer progression with microbiota-associated biomarkers}
+\textbf{Evaluation of machine learning methods that identify colorectal
+lesions with microbiota-associated biomarkers}
 
-Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Patrick D.
-Schloss\textsuperscript{1\(\dagger\)}
+Begüm D. Topçuoğlu, Jenna Wiens, Mack Ruffin, Patrick D. Schloss
 
-As gut microbiome field continues to grow, there will be an
-ever-increasing demand for reproducible machine learning methods to
-determine association of the microbiome with a continuous or categorical
-phenotype of interest. Currently, the use of machine learning in
-microbiome literature lack clarity over the training, validation and
-testing of the models used. There is a need to properly implement good
-machine learning practices to generate reproducible and robust models.
+As gut microbiome field continues to grow, there is an ever-increasing
+demand for reproducible machine learning methods to determine
+associations between the microbiome and a phenotype of interest.
+Currently, the use of machine learning in microbiome research lacks
+clarity and consistency over the modeling pipeline (training, validation
+and testing steps). There is a need for guidance on how to implement
+good machine learning practices to generate reproducible and robust
+models.
 
-Recently, there is an interest in using machine learning to predict
-colorectal cancer progression with microbiota-associated biomarkers.
-Colorectal cancer is one of the leading cause of death among cancers in
-the United States. Each person in the industrialized world has on
-average a one-in-twenty chance of developing colorectal cancer (CRC)
-(1--3). Colonoscopy as a screening tool is effective, however it is
-invasive, expensive and have a low rate of patient adherence. Therefore,
-gut microbiome-based biomarkers emerged as a non-invasive screening
-method.
+Recently, there has been growing interest in using machine learning to
+identify colorectal lesions that are precursors of coloractal cancer,
+with microbiota-associated biomarkers. Colorectal cancer is one of the
+leading cause of death among cancers in the United States. Colonoscopy
+as a screening tool is effective, however it is invasive, expensive and
+have a low rate of patient adherence. Previous studies have shown that
+bacterial population abundances in the stool can predict screen relevant
+lesions in the colon and can be used as a non-invasive screening tool.
+However, the prediction performance of these models vary greatly, with
+areas under the receiver operating characteristic curve (AUC) of 0.7-0.9
+(1--4). The variation in prediction performance is based in part on
+differences in the study populations, and in part on the differences in
+modeling pipelines.
 
-In this study, classification models that use human hemoglobin levels
-and bacterial population abundances in the stool were used to predict
-colorectal disease status as screen-relevant colonic growth or not.
-Training, validation and testing pipelines were established for
-L2-regularized Logistic Regression, L1 and L2 Linear Suppor Vector
-Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest
-and XGBoost classifiers. The generalization and prediction performance
-of these classifiers were evaluated and each classifier was examined for
-its reproducibility, robustness and susceptibility to overfitting.
-L2-regularized Logistic Regression had a mean AUC of 0.68 +/- 0.04, L1
-Linear SVM had a mean AUC of 0.76 +/- 0.05, L2 Linear SVM had a mean AUC
-of 0.68 +/- 0.05 and Radial Basis Function SVM had a mean AUC of 0.69
-+/- 0.05. Decision Tree had a mean AUC of 0.7 +/- 0.05, Random Forest
-had a mean AUC of 0.76 +/- 0.06 and XGBoost had a mean AUC of 0.76 +/-
-0.04. Tree-based models were less susceptible to overfiting and in
+In this study, hemoglobin levels and 16S rRNA gene sequences in the
+stool were used to identify colorectal lesions of 490 patients. The
+colorectal disease stage was defined as showing screen-relevant lesions
+or not. Modeling pipelines were established for L2-regularized Logistic
+Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis
+Function SVM, Decision Tree, Random Forest and XGBoost binary
+classification models. The mean AUCs of these models were 0.68 ± 0.04,
+0.76 ± 0.05, 0.68 ± 0.05, 0.69 ± 0.05, 0.71 ± 0.04, 0.82 ± 0.04, and
+0.76 ± 0.04, respectively. Tree-based methods, namely Decision Tree,
+Random Forest and XGBoost were less susceptible to overfitting and in
 general had higher sensitivity and specificity for colonic
-screen-relevant growth.
+screen-relevant lesions. Aside from evaluating generalization and
+classification performance of each model, this study established
+standards for modeling pipeline of the microbiome-associated machine
+learning models.
 
@@
 
-\subsection{References}\label{references}
+\textbf{References}
 
 \hypertarget{refs}{}
-\hypertarget{ref-seer_2016}{}
-1. \textbf{Howlader N KM \textnormal{Noone AM}}. SEER cancer statistics
-review, 1975-2013, (national cancer institute. bethesda, md).
+\hypertarget{ref-sze_leveraging_2018}{}
+1. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
+rRNA gene surveys to identify reproducible biomarkers in individuals
+with colorectal tumors. mBio \textbf{9}:e00630--18.
+doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.
 
-\hypertarget{ref-street_colorectal_nodate}{}
-2. \textbf{Street W}. Colorectal cancer facts \& figures 2017-2019 40.
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+2. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
 
-\hypertarget{ref-weir_past_2015}{}
-3. \textbf{Weir HK}, \textbf{Thompson TD}, \textbf{Soman A},
-\textbf{MÃžller B}, \textbf{Leadbetter S}. 2015. The past, present, and
-future of cancer incidence in the united states: 1975 through 2020.
-Cancer \textbf{121}:1827--1837.
-doi:\href{https://doi.org/10.1002/cncr.29258}{10.1002/cncr.29258}.
+\hypertarget{ref-baxter_dna_2016}{}
+3. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
+\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
+immunochemical test can replace stool for detection of colonic lesions
+using a microbiota-based model. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.
+
+\hypertarget{ref-zackular_human_2014}{}
+4. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
 

### 2020-02-24 submission/manuscript.Rmd
Pat's edits
Close to submitting

@@
 
-Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack Ruffin${^3}$, Jenna Wiens${^2}$, Patrick D. Schloss^1$\dagger$^
+Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack Ruffin${^3}$, Jenna Wiens^2$\dagger$^, Patrick D. Schloss^1$\dagger$^
 
@@
 The training times increased with the complexity of the model and the number of potential hyperparameter combinations.
-Also, the linear models trained faster than non-linear models [Figures S2-S3; Figure 5].
+Also, the linear models trained faster than non-linear models [Figure 5].
 
@@
 
-\captionsetup{labelformat=empty}
-\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
+**Table 1.** Characteristics of the machine learning models in our comparative study.
+
 \small
@@
 
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
+\makecell[l]{Logistic regression} & \makecell[l]{A predictive regression analysis when the dependent variable is binary.} & Linear \\ \hline
 
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
+\makecell[l]{SVM with linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear separating hyperplane that discriminates between labels.} & Linear  \\ \hline
 
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
+\makecell[l]{SVM with radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
 
 \makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
+root to the leaf node where an attribute is tested to discriminate between labels.} & Non-linear \\ \hline
 
-Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
+Random forest & \makecell[l]{A classifier that is an ensemble of decision trees that grows randomly with subsampled data.} & Non-linear \\ \hline
 
-\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
+\makecell[l]{Gradient Boosted Trees (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees that grows greedily.} & Non-linear \\ \hline
 

### 2020-04-19 submission/manuscript.Rmd
second rebuttal

--- 2020-02-17 submission/manuscript.Rmd
@@
 Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population separately [@segata_metagenomic_2011].
-Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease because they consider the relative abundance of each bacterial population in the context of others rather than in isolation.
+Recently, machine learning (ML) models have grown in popularity among microbiome researchers because ML models can effectively account for the interpersonal microbiome variation and the ecology of disease as they consider the relative abundance of each bacterial population in the context of others rather than in isolation.
 
@@
 We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4].
-Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models.
+Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models.
+We also found that three OTUs (OTUs 58, 110, 367) were important in each of the tree-based models.
 Similarly, the random forest and XGBoost models shared four of the most important OTUs (OTUs 2, 12, 361, 477).
@@
 
-Our framework provides a reproducible structure for investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant.
+Our framework provides a reproducible pipeline to train, evaluate, and interpret microbiome-based ML models and generate hypotheses to explain the underlying microbiology of the model prediction.
 However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking [@wiens_no_2019].
@@
 **Figure 2.
-Generalization and classification performance of ML models using AUROC values of all cross-validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by a horizontal line at 0.50) for all the ML models.
+Generalization and classification performance of ML models using AUROC values of all cross-validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by a vertical line at 0.50) for all the ML models.
 The predictive performance of random forest model was higher than other ML models, though not significantly (p > 0.05).
@@
 The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here.
-The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, green were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only.
+The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, green were OTUs that were shared among the tree-based models, turqoise were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only.
 For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.

### 2020-04-19 code/learning/Figure2.R

@@ -61,28 +61,28 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
               y_position = 0.83, tip_length = 0.02, vjust=0.3) +
   annotate("text", x= 5, y= 0.85, size=4, label = paste0("N.S.")) + 
   theme_bw() +
-  theme(plot.margin=unit(c(1.1,1.1,1.1,1.1),"cm"),
+  theme(plot.margin=unit(c(0,1.1,0,0),"cm"),
         legend.justification=c(1,0),
         legend.position=c(1,0),
         #legend.position="bottom",
         legend.title = element_blank(),
         legend.background = element_rect(linetype="solid", color="black", size=0.5),
-        legend.box.margin=margin(c(12,12,12, 12)),
-        legend.text=element_text(size=18),
+        legend.box.margin=margin(c(10,10,10, 10)),
+        legend.text=element_text(size=10),
         #legend.title=element_text(size=22),
         panel.grid.major.y = element_blank(),
         panel.grid.major.x = element_line( size=0.6),
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         text = element_text(size = 12),
-        axis.text.x=element_text(size = 20, colour='black'),
-        axis.text.y=element_text(size = 20, colour='black'),
-        axis.title.y=element_text(size = 24),
-        axis.title.x=element_text(size = 24),
+        axis.text.x=element_text(size = 10, colour='black'),
+        axis.text.y=element_text(size = 10, colour='black'),
+        axis.title.y=element_text(size = 10),
+        axis.title.x=element_text(size = 10),
         panel.border = element_rect(linetype="solid", colour = "black", fill=NA, size=1.5))
 
 ######################################################################
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 
-ggsave("Figure_2.tiff", plot = performance, device = 'tiff', path = 'submission', width = 12, height = 9, dpi=300)
+ggsave("Figure_2.tiff", compression = "lzw", plot = performance, device = 'tiff', path = 'submission', units="in",width=6.875, height=4.5)

### 2020-04-19 code/learning/Figure3.R

@@ -135,4 +135,4 @@ linear <- plot_grid(l1_svm_graph, l2_svm_graph, logit_graph, labels = c("A", "B"
 
 ggdraw(add_sub(linear, "Feature ranks", vpadding=grid::unit(0,"lines"), y=5, x=0.7, vjust=4.75, size=15))
 
-ggsave("Figure_3.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9.2, dpi=300)
+ggsave("Figure_3.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9, units="in", dpi=300)

### 2020-04-19 code/learning/Figure4.R

@@ -196,13 +196,13 @@ base_nonlin_plot <-  function(data, name){
                        expand=c(0,0)) +
     theme(plot.margin=unit(c(1.5,3,1.5,3),"mm"),
           legend.position="none",
-          axis.title = element_text(size=20),
-          axis.text = element_text(size=20),
+          axis.title = element_text(size=10),
+          axis.text = element_text(size=10),
           panel.border = element_rect(colour = "black", fill=NA, size=2),
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           panel.background = element_blank(),
-          axis.text.y=element_text(size = 12, colour='black'),
+          axis.text.y=element_text(size = 8, colour='black'),
           axis.title.x=element_blank(),
           axis.text.x=element_blank(),
           axis.ticks = element_line(colour = "black", size = 1.1))
@@ -299,7 +299,7 @@ rf_plot <- base_nonlin_plot(rf, "Random_Forest") +
 xgboost_plot <- base_nonlin_plot(xgboost, "XGBoost")+
   scale_x_discrete(name = "XGBoost ",
                    labels = get_taxa_info_as_labels("XGBoost")) +
-  theme(axis.text.x=element_text(size = 16, colour='black'))
+  theme(axis.text.x=element_text(size = 10, colour='black'))
 # ----------------------------------------------------------------------->
 
 # Extract legend
@@ -308,7 +308,7 @@ legend <- get_legend(xgboost_plot + guides(fill=guide_legend(ncol=2)) + theme(le
                                           legend.direction = "horizontal",
                                           legend.justification="left",
                                           #legend.box.just = "bottom",
-                                          legend.text=element_text(size=10)))
+                                          legend.text=element_text(size=8)))
                                           #legend.background = element_rect(linetype="solid", 
                                                                            #color="black", 
                                                                            #size=0.5)))
@@ -320,9 +320,9 @@ legend <- get_legend(xgboost_plot + guides(fill=guide_legend(ncol=2)) + theme(le
 #combine with cowplot
 perm_tree_based <- plot_grid(rbf_plot, dt_plot, rf_plot, xgboost_plot, labels = c("A", "B", "C", "D"), cols=1, scale = 0.97, align = "v")
 
[…]

### 2020-04-19 code/learning/Figure5.R

@@ -69,7 +69,7 @@ traintime_plot <- ggplot(traintime_df, aes(x = fct_reorder(model, x), y = x)) +
 #-----------------------Save figure as .pdf ------------------------ #
 ######################################################################
 
-ggsave("Figure_5.tiff", plot = traintime_plot, device = 'tiff', path = 'submission', width = 4, height = 2, dpi=300)
+ggsave("Figure_5.tiff", plot = traintime_plot, device = 'tiff', path = 'submission', width = 4, height = 2, units="in", dpi=300)
 
 
             

### 2020-04-19 code/learning/FigureS1.R

@@ -43,7 +43,7 @@ linear_plots <- fig_data %>%
   ggplot(aes(x=cost, y=mean_AUC, color=model)) +
   geom_hline(yintercept=0.5, linetype="dashed") +
   geom_line(size=1.5) +
-  geom_point(size=5) +
+  geom_point(size=3) +
   geom_errorbar(aes(ymin=mean_AUC-sd_AUC, ymax=mean_AUC+sd_AUC), width=.001) +
   scale_x_log10(limits=c(min=1e-4,max=1),
                 breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1),
@@ -62,34 +62,34 @@ linear_plots <- fig_data %>%
            x = 0.5, 
            y = 0.59, 
            colour = "#00AFBB", 
-           size = 3,
+           size = 2,
            shape = 8) +
   annotate("point", # best hp for L2 Linear SVM - highest mean cv AUROC
            x = 0.5, 
            y = 0.71, 
            colour = "#E7B800", 
-           size = 3,
+           size = 2,
            shape = 8) +
   annotate("point", # best hp for L2 logistic regression - highest mean cv AUROC
            x = 0.25, 
            y = 0.71, 
            colour = "#FC4E07", 
-           size = 3,
+           size = 2,
            shape = 8) +
   theme_bw() +
   theme(legend.background = element_rect(linetype="solid", color="black", size=0.5),
         legend.box.margin=margin(c(12,12,12, 12)),
-        legend.text=element_text(size=18),
+        legend.text=element_text(size=9),
         legend.title=element_blank(),
         legend.position=c(0.65, 0.85),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
-        text = element_text(size = 18),
-        axis.text.x=element_text(size = 20, colour='black'),
-        axis.text.y=element_text(size = 20, colour='black'),
-        axis.title.y=element_text(size = 24),
-        axis.title.x=element_text(size = 24),
+        text = element_text(size = 8),
+        axis.text.x=element_text(size = 10, colour='black'),
+        axis.text.y=element_text(size = 10, colour='black'),
+        axis.title.y=element_text(size = 12),
+        axis.title.x=element_text(size = 12),
         panel.border = element_rect(linetype="solid", colour = "black", fill=NA, size=1.5))
 
 
@@ -97,5 +97,5 @@ linear_plots <- fig_data %>%
[…]

### 2020-04-19 code/learning/FigureS2.R

@@ -49,16 +49,16 @@ base_plot <-  function(data, x_axis, y_axis){
   geom_point() +
   theme_bw() +
   geom_hline(yintercept = 0.5, linetype="dashed") +
-  theme(legend.text=element_text(size=10),
-        legend.title=element_text(size=10),
+  theme(legend.text=element_text(size=6),
+        legend.title=element_text(size=7),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
-        text = element_text(size = 10),
-        axis.text.x=element_text(size = 8, colour='black'),
-        axis.text.y=element_text(size = 8, colour='black'),
-        axis.title.y=element_text(size = 10),
-        axis.title.x=element_text(size = 10))
+        text = element_text(size = 6),
+        axis.text.x=element_text(size = 6, colour='black'),
+        axis.text.y=element_text(size = 6, colour='black'),
+        axis.title.y=element_text(size = 7),
+        axis.title.x=element_text(size = 7))
   return(plot)
 }
 
@@ -117,17 +117,17 @@ rbf_plot <- ggplot(rbf_data, aes(x = sigma, y = C, fill = mean_AUC)) +
                 labels=trans_format('log10',math_format(10^.x))) +
   theme(legend.background = element_rect(size=0.5, linetype="solid", color="black"),
         legend.box.margin=margin(c(1,1,1,1)),
-        legend.text=element_text(size=8),
-        legend.title=element_text(size=10), 
+        legend.text=element_text(size=5),
+        legend.title=element_text(size=5), 
         legend.position="bottom",
-        axis.title = element_text(size=10),
-        axis.text = element_text(size=10),
+        axis.title = element_text(size=6),
+        axis.text = element_text(size=6),
         panel.border = element_rect(colour = "black", fill=NA, size=3), 
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
-        axis.text.x=element_text(size = 8, colour='black'),
-        axis.text.y=element_text(size = 8, colour='black'))
+        axis.text.x=element_text(size = 6, colour='black'),
+        axis.text.y=element_text(size = 6, colour='black'))
 
 xgboost_data <- xgboost_all %>%
   group_by(eta, subsample) %>%
@@ -154,18 +154,18 @@ xgboost_plot <- ggplot(xgboost_data, aes(x = eta, y = subsample, fill = mean_AUC
                 expand = c(0, 0),
[…]

### 2020-04-19 code/learning/FigureS4.R

@@ -127,5 +127,5 @@ plots <- plot_grid(logit_plot, rf_plot, labels = c("A", "B"), align = 'v', ncol 
 
 ggdraw(add_sub(plots, "Cross-validation AUROC", vpadding=grid::unit(0,"lines"), y=5, x=0.6, vjust=4.75, size=15))
 
-ggsave("Figure_S5.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9.2, dpi=300)
+ggsave("Figure_S5.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9, dpi=300)
 

### 2020-04-19 code/learning/FigureS6.R

@@ -126,13 +126,13 @@ base_nonlin_plot <-  function(data, name){
                        expand=c(0,0)) +
     theme(plot.margin=unit(c(1.5,3,1.5,3),"mm"),
           legend.position="none",
-          axis.title = element_text(size=15),
-          axis.text = element_text(size=15),
+          axis.title = element_text(size=10),
+          axis.text = element_text(size=8),
           panel.border = element_rect(colour = "black", fill=NA, size=2), 
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           panel.background = element_blank(),
-          axis.text.y=element_text(size = 12, colour='black'), 
+          axis.text.y=element_text(size = 8, colour='black'), 
           axis.title.x=element_blank(),
           axis.text.x=element_blank(), 
           axis.ticks = element_line(colour = "black")) 
@@ -216,7 +216,7 @@ linear <- plot_grid(l1_plot, l2_plot, logit_plot, labels = c("A", "B", "C"), ali
 
 ggdraw(add_sub(linear, "AUROC with the OTU permuted randomly", size=18, vpadding=grid::unit(0,"lines"), y=5, x=0.65, vjust=4.75))
 
-ggsave("Figure_S6.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 8, height = 12, dpi=300)
+ggsave("Figure_S6.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 6, height = 9, dpi=300)
 
 
 

### 2020-04-19 code/learning/FigureS7.R

@@ -126,13 +126,13 @@ base_nonlin_plot <-  function(data, name){
                        expand=c(0,0)) +
     theme(plot.margin=unit(c(1.5,3,1.5,3),"mm"),
           legend.position="none",
-          axis.title = element_text(size=15),
-          axis.text = element_text(size=15),
+          axis.title = element_text(size=10),
+          axis.text = element_text(size=8),
           panel.border = element_rect(colour = "black", fill=NA, size=2), 
           panel.grid.major = element_blank(),
           panel.grid.minor = element_blank(),
           panel.background = element_blank(),
-          axis.text.y=element_text(size = 12, colour='black'), 
+          axis.text.y=element_text(size = 8, colour='black'), 
           axis.title.x=element_blank(),
           axis.text.x=element_blank(), 
           axis.ticks = element_line(colour = "black")) 
@@ -197,7 +197,7 @@ get_taxa_info_as_labels <- function(name){
 logit_plot <- base_nonlin_plot(logit, "L2_Logistic_Regression") +
   scale_x_discrete(name = expression(paste(L[2], "-regularized logistic regression")),
                    labels = get_taxa_info_as_labels("L2_Logistic_Regression")) +
-  theme(axis.text.x=element_text(size = 16, colour='black'))
+  theme(axis.text.x=element_text(size = 8, colour='black'))
 
 l1_plot <- base_nonlin_plot(l1svm, "L1_Linear_SVM") +
   scale_x_discrete(name = expression(paste(L[1], "-regularized linear SVM")),
@@ -214,9 +214,9 @@ l2_plot <- base_nonlin_plot(l2svm, "L2_Linear_SVM") +
 #combine with cowplot
 linear <- plot_grid(l1_plot, l2_plot, logit_plot, labels = c("A", "B", "C"), align = 'v', ncol = 1, scale = 0.97)
 
-ggdraw(add_sub(linear, "AUROC with the OTU permuted randomly", size=18, vpadding=grid::unit(0,"lines"), y=5, x=0.65, vjust=4.75))
+ggdraw(add_sub(linear, "AUROC with the OTU permuted randomly", size=8, vpadding=grid::unit(0,"lines"), y=5, x=0.65, vjust=4.75))
 
-ggsave("Figure_S7.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 8, height = 12, dpi=300)
[…]

### 2020-04-19 code/learning/FigureS8.R

@@ -165,7 +165,7 @@ rf_les_otus <- as.character(rf_otus$names)
 rf_les_tax <- rf_taxa_otus$names
 
 #Stripchart of abundances
-tiff('submission/Figure_S8.tiff', units="in", width=8, height=12, res=300)
+tiff('submission/Figure_S8.tiff', units="in", width=6.8, height=9, res=300)
 layout(matrix(2:1, nrow=2), widths=c(0.5,1))
 
 # L2-logistic-regression
@@ -182,7 +182,7 @@ for(i in logit_les_otus){
 legend("topright", legend=c("SRN", "Normal"), pch=c(21, 21), pt.bg=c("red3","lightskyblue"))
 axis(1, at=c(1e-4, 1e-3, 1e-2, 1e-1, 1), label=c("0", "0.1", "1", "10", "100"))
 mtext(logit_les_tax, at=1:length(logit_les_otus)*2-1, side=2, las=1, adj=1, line=0.5)
-mtext('B - L2 regularized logistic regression',cex=2, font=10, side=3, adj= 0, line=0.5)
+mtext('B - L2 logistic regression',cex=1.5, font=8, side=3, adj= 0, line=0.5)
 
 # random forest
 par(mar=c(4,12,1.8,1.8))
@@ -198,6 +198,6 @@ for(i in rf_les_otus){
 legend("topright", legend=c("SRN", "Normal"), pch=c(21, 21), pt.bg=c("red1","lightskyblue"))
 axis(1, at=c(1e-4, 1e-3, 1e-2, 1e-1, 1), label=c("0", "0.1", "1", "10", "100"))
 mtext(rf_les_tax, at=1:length(rf_les_otus)*2-1, side=2, las=1, adj=1, line=0.5)
-mtext('A - Random forest',cex=2, font=10, side=3, adj= 0, line=0.2)
+mtext('A - Random forest',cex=1.5, font=8, side=3, adj= 0, line=0.2)
 dev.off()
 

### 2020-04-28 submission/manuscript.Rmd
Final edits and cleaning

@@
 For example, our full dataset included 490 individuals.
-Of these, 261 had healthy colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%).
-A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals, of which 52 had an SRN (54%).
+Of these, 261 had healthy colons (53.3%) and 229 had a screen relevant neoplasia (SRN; 46.7%).
+A training set included 393 individuals, of which 184 had an SRN (46.8%), while the test set was composed of 97 individuals, of which 45 had an SRN (46.4%).
 The training data were used to build and select the models, and the test set was used for evaluating the model.
@@
 We modified the caret code to calculate decision values for models generated using L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel.
-The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel models are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R}, respectively.
+The code for these changes on L2-regularized SVM with linear kernel and L1-regularized SVM with linear kernel models are available at \url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear3.R} and at \url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear4.R}, respectively.
 
@@
 
-**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
+**Code availability.** The code for all sequence curation and analysis steps including an Rmarkdown version of this manuscript is available at \url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/}.
 
@@
 
-\small
-\begin{tabular}{|l|l|l|}
-\hline
+| **Model** | **Description** | **Linearity** |
+|-----------|-----------------|---------------|
+|  Logistic regression |  A predictive regression analysis when the dependent variable is binary | Linear  |
+|  SVM with linear kernel | A classifier that is defined by an optimal linear separating hyperplane that discriminates between labels  | Linear  |
+|  SVM with radial basis kernel | A classifier that is defined by an optimal non-linear separating hyperplane that discriminates between labels  | Non-linear  |
+|  Decision tree | A classifier that sorts samples down from the root to the leaf node where an attribute is tested to discriminate between labels  | Non-linear  |
+|  Random forest | A classifier that is an ensemble of decision trees that grows randomly with subsampled data  | Non-linear |
+|  Gradient Boosted Trees (XGBoost) | A classifier that is an ensembe of decision trees that grows greedily  | Non-linear  |
 
-\rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
-
-\makecell[l]{Logistic regression} & \makecell[l]{A predictive regression analysis when the dependent variable is binary.} & Linear \\ \hline
-
-\makecell[l]{SVM with linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear separating hyperplane that discriminates between labels.} & Linear  \\ \hline
-
-\makecell[l]{SVM with radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-
-\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the leaf node where an attribute is tested to discriminate between labels.} & Non-linear \\ \hline
-
-Random forest & \makecell[l]{A classifier that is an ensemble of decision trees that grows randomly with subsampled data.} & Non-linear \\ \hline
-
-\makecell[l]{Gradient Boosted Trees (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees that grows greedily.} & Non-linear \\ \hline
-
-\end{tabular}
 \newpage
@@
 The minimal differences betweeen relative abundances for these OTUs show that it is not possible to differentiate disease vs healthy states by focusing on individual taxa.
+
+\hfill\break
+
+**Table S1. An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.**

### 2020-04-28 submission/manuscript.tex
Final edits and cleaning

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx,grffile}
@@
 population separately (1). Recently, machine learning (ML) models have
-grown in popularity among microbiome researchers as our ability to
-sample large numbers of individuals has grown; such models can
+grown in popularity among microbiome researchers because ML models can
 effectively account for the interpersonal microbiome variation and the
-ecology of disease because they consider the relative abundance of each
+ecology of disease as they consider the relative abundance of each
 bacterial population in the context of others rather than in isolation.
@@
 performed stratified splits. For example, our full dataset included 490
-individuals. Of these, 261 had healthy colons (53\%) and 229 had a
+individuals. Of these, 261 had healthy colons (53.3\%) and 229 had a
 screen relevant neoplasia (SRN; 46.7\%). A training set included 393
-individuals, of which 209 had an SRN (53\%), while the test set was
-composed of 97 individuals, of which 52 had an SRN (54\%). The training
-data were used to build and select the models, and the test set was used
-for evaluating the model. We trained seven different models using the
-training data {[}Table 1{]}.
+individuals, of which 184 had an SRN (46.8\%), while the test set was
+composed of 97 individuals, of which 45 had an SRN (46.4\%). The
+training data were used to build and select the models, and the test set
+was used for evaluating the model. We trained seven different models
+using the training data {[}Table 1{]}.
 
@@
 twenty OTUs with the largest impact, there was only one OTU (OTU 822)
-that was shared among all of the models; however, we found three OTUs
-(OTUs 58, 110, 367) that were important in each of the tree-based
-models. Similarly, the random forest and XGBoost models shared four of
-the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis
-results also revealed that with the exception of the decision tree
-model, removal of any individual OTU had minimal impact on model
-performance. For example, if OTU 367 was permuted across the samples in
-the decision tree model, the median AUROC dropped from 0.601 to 0.525.
-In contrast, if the same OTU was permuted in the random forest model,
-the AUROC only dropped from 0.695 to 0.680, which indicated high degree
-of collinearity in the dataset. Permutation analysis allowed us to gauge
-the importance of each OTU in non-linear models and partially account
-for collinearity by grouping correlated OTUs to determine their impact
-as a group.
+that was shared among all of the models. We also found that three OTUs
+(OTUs 58, 110, 367) were important in each of the tree-based models.
+Similarly, the random forest and XGBoost models shared four of the most
+important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also
+revealed that with the exception of the decision tree model, removal of
+any individual OTU had minimal impact on model performance. For example,
+if OTU 367 was permuted across the samples in the decision tree model,
+the median AUROC dropped from 0.601 to 0.525. In contrast, if the same
+OTU was permuted in the random forest model, the AUROC only dropped from
+0.695 to 0.680, which indicated high degree of collinearity in the
+dataset. Permutation analysis allowed us to gauge the importance of each
+OTU in non-linear models and partially account for collinearity by
+grouping correlated OTUs to determine their impact as a group.
 
@@
 combinations. Also, the linear models trained faster than non-linear
-models {[}Figures S2-S3; Figure 5{]}.
+models {[}Figure 5{]}.
 
@@
 
-Our framework provides a reproducible structure for investigators
-wanting to train, evaluate, and interpret their own ML models to
-generate hypotheses regarding which OTUs might be biologically relevant.
-However, deploying microbiome-based models to make clinical diagnoses or
-predictions is a significantly more challenging and distinct undertaking
-(44). For example, we currently lack standardized methods to collect
-patient samples, generate sequence data, and report clinical data. We
-are also challenged by the practical constraints of OTU-based
-approaches. The de novo algorithms commonly in use are slow, require
-considerable memory, and result in different OTU assignments as new data
-are added (45). Finally, we also need independent validation cohorts to
-test the performance of a diagnostic model. To realize the potential for
-using ML approaches with microbiome data, it is necessary that we direct
-our efforts to overcome these challenges.
+Our framework provides a reproducible pipeline to train, evaluate, and
+interpret microbiome-based ML models and generate hypotheses to explain
+the underlying microbiology of the model prediction. However, deploying
+microbiome-based models to make clinical diagnoses or predictions is a
+significantly more challenging and distinct undertaking (44). For
+example, we currently lack standardized methods to collect patient
+samples, generate sequence data, and report clinical data. We are also
+challenged by the practical constraints of OTU-based approaches. The de
+novo algorithms commonly in use are slow, require considerable memory,
+and result in different OTU assignments as new data are added (45).
+Finally, we also need independent validation cohorts to test the
+performance of a diagnostic model. To realize the potential for using ML
+approaches with microbiome data, it is necessary that we direct our
+efforts to overcome these challenges.
 
@@
 L1-regularized SVM with linear kernel models are available at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
+\url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear3.R}
 and at
-\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
+\url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/blob/master/data/caret_models/svmLinear4.R},
 respectively.
@@
 analysis steps including an Rmarkdown version of this manuscript is
-available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.
+available at
+\url{https://github.com/SchlossLab/Topcuoglu_ML_mBio_2020/}.
 
@@
 
-\captionsetup{labelformat=empty}
-\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
-\small
-\begin{tabular}{|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-
-\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
-
-Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
-
-\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
-
-\end{tabular}
+\textbf{Table 1.} Characteristics of the machine learning models in our
+comparative study.
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule
+\begin{minipage}[b]{0.23\columnwidth}\raggedright
+\textbf{Model}\strut
+\end{minipage} & \begin{minipage}[b]{0.36\columnwidth}\raggedright
+\textbf{Description}\strut
+\end{minipage} & \begin{minipage}[b]{0.32\columnwidth}\raggedright
+\textbf{Linearity}\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.23\columnwidth}\raggedright
+Logistic regression\strut
+\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
+A predictive regression analysis when the dependent variable is
+binary\strut
+\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
+Linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright
+SVM with linear kernel\strut
+\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
+A classifier that is defined by an optimal linear separating hyperplane
+that discriminates between labels\strut
+\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
+Linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright
+SVM with radial basis kernel\strut
+\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
+A classifier that is defined by an optimal non-linear separating
+hyperplane that discriminates between labels\strut
+\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
+Non-linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright
+Decision tree\strut
+\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
+A classifier that sorts samples down from the root to the leaf node
+where an attribute is tested to discriminate between labels\strut
+\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
+Non-linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright
+Random forest\strut
+\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
+A classifier that is an ensemble of decision trees that grows randomly
+with subsampled data\strut
+\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
+Non-linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright
+Gradient Boosted Trees (XGBoost)\strut
+\end{minipage} & \begin{minipage}[t]{0.36\columnwidth}\raggedright
+A classifier that is an ensembe of decision trees that grows
+greedily\strut
+\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright
+Non-linear\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
 \newpage
@@
 using bacterial abundances was higher than chance (depicted by a
-horizontal line at 0.50) for all the ML models. The predictive
-performance of random forest model was higher than other ML models,
-though not significantly (p \textgreater{} 0.05). L2-regularized
-logistic regression, XGBoost, L2-regularized SVM with linear and radial
-basis function kernel performances were not significantly different from
-one another. The boxplot shows quartiles at the box ends and the median
-as the horizontal line in the box. The whiskers show the farthest points
+vertical line at 0.50) for all the ML models. The predictive performance
+of random forest model was higher than other ML models, though not
+significantly (p \textgreater{} 0.05). L2-regularized logistic
+regression, XGBoost, L2-regularized SVM with linear and radial basis
+function kernel performances were not significantly different from one
+another. The boxplot shows quartiles at the box ends and the median as
+the horizontal line in the box. The whiskers show the farthest points
 that were not outliers. Outliers were defined as those data points that
@@
 that were shared among all the non-linear models, green were OTUs that
-were shared among the tree-based models, green were the OTUs shared
+were shared among the tree-based models, turqoise were the OTUs shared
 among SVM with radial basis kernel, decision tree and XGBoost, pink were
@@
 
+\hfill\break
+
+\textbf{Table S1. An aspirational rubric for evaluating the rigor of ML
+practices applied to microbiome data.}
+
 \end{document}

### 2020-05-14 submission/manuscript.Rmd
new 2 figures

@@
 | **Model** | **Description** | **Linearity** |
-|-----------|-----------------|---------------|
+|--------------------|--------------------------------------------|---------------|
 | Logistic regression | A predictive regression analysis when the dependent variable is binary |  Linear  |

### 2020-05-14 submission/manuscript.tex
new 2 figures

--- 2020-02-17 submission/manuscript.tex
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs}
+% Fix footnotes in tables (requires footnote package)
+\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
 \usepackage{graphicx,grffile}
@@
 
-\captionsetup{labelformat=empty}
-\captionof{table}{\textbf{Table 1.} Characteristics of the machine learning models in our comparative study.}
-\small
-
-\begin{tabular}{|l|l|l|}
-\hline
-
-\rowcolor{lightgray}
-\textbf{Model} & \textbf{Description} & \textbf{Linearity} \\ \hline
-
-\makecell[l]{Logistic \\regression} & \makecell[l]{A predictive regression analysis when the dependent \\variable is binary.} & Linear \\ \hline
-
-\makecell[l]{SVM with \\linear kernel} & \makecell[l]{A classifier that is defined by an optimal linear \\separating hyperplane that discriminates between labels.} & Linear  \\ \hline
-
-\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
-
-\makecell[l]{Decision tree} & \makecell[l]{A classifier that sorts samples down from the
-root to the \\leaf node where an attribute is tested to discriminate \\between labels.} & Non-linear \\ \hline
-
-Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline
-
-\makecell[l]{Gradient Boosted Trees \\ (XGBoost)} & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows greedily.} & Non-linear \\ \hline
-
-\end{tabular}\newpage
+\textbf{Table 1.} Characteristics of the machine learning models in our
+comparative study.
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule
+\begin{minipage}[b]{0.23\columnwidth}\raggedright\strut
+\textbf{Model}\strut
+\end{minipage} & \begin{minipage}[b]{0.50\columnwidth}\raggedright\strut
+\textbf{Description}\strut
+\end{minipage} & \begin{minipage}[b]{0.18\columnwidth}\raggedright\strut
+\textbf{Linearity}\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
+Logistic regression\strut
+\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
+A predictive regression analysis when the dependent variable is
+binary\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
+Linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
+SVM with linear kernel\strut
+\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
+A classifier that is defined by an optimal linear separating hyperplane
+that discriminates between labels\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
+Linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
+SVM with radial basis kernel\strut
+\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
+A classifier that is defined by an optimal non-linear separating
+hyperplane that discriminates between labels\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
+Non-linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
+Decision tree\strut
+\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
+A classifier that sorts samples down from the root to the leaf node
+where an attribute is tested to discriminate between labels\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
+Non-linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
+Random forest\strut
+\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
+A classifier that is an ensemble of decision trees that grows randomly
+with subsampled data\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
+Non-linear\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
+Gradient Boosted Trees (XGBoost)\strut
+\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
+A classifier that is an ensembe of decision trees that grows
+greedily\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
+Non-linear\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\newpage
 

### 2020-05-14 code/learning/Figure2.R

@@ -59,7 +59,7 @@ performance <- ggplot(best_performance, aes(x = fct_reorder(model, AUC), y = AUC
                             "Random forest")) + 
   geom_signif(comparisons=list(c("RBF_SVM", "Random_Forest")), annotations="",
               y_position = 0.83, tip_length = 0.02, vjust=0.3) +
-  annotate("text", x= 5, y= 0.85, size=4, label = paste0("N.S.")) + 
+  annotate("text", x= 5, y= 0.86, size=4, label = paste0("N.S.")) + 
   theme_bw() +
   theme(plot.margin=unit(c(0,1.1,0,0),"cm"),
         legend.justification=c(1,0),

### 2020-05-14 code/learning/Figure4.R

@@ -324,5 +324,5 @@ ggdraw(add_sub(perm_tree_based, "AUROC with the OTU permuted randomly", size=10,
 
 #plot <- plot_grid(last_plot(), legend, ncol = 1, rel_heights=c(.92, .1))
 
-ggsave("Figure_4.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 4, height = 9, units="in", dpi=300)
+ggsave("Figure_4.tiff", plot = last_plot(), device = 'tiff', path = 'submission', width = 5, height = 9, units="in", dpi=300)
 
